### 2015-03-13 rebuttal.txt
previous version of the paper

The authors have significantly improved the quality of their paper. 
The changes on the paper are fine and contribute to increase readability and quality. 
I thank the authors for having considered my comments. I believe that most of them were answered.
The paper is now clear, and all the sections can be understood.
The improvements made to emphasize my previous positive feeling about the quality of the contributions.
The contribution has been formalised, and an important work has been done since the previous submission.
I recommend publication.

===============
We thank the reviewer for their positive comments and appreciate their time.

===============

Nevertheless, I have several minor remarks.
MINOR REMARKS
1. Reading the letter, I was glad to see that an industrial case study has been considered. However, even if this case is one from real life, I’m disappointed about its simplicity. In particular, it seems simpler than the case study of the paper.

==============

Further industrial case studies are being examined, including a UML-RT to Kiltera transformation.

==============

Additionally to this, I still have doubts about the replication of rules (Fig. 15). I would appreciate if the table 1 detailed the number of “disambiguation” made (corresponding to the section 4.2) and the number of “path condition combination” made for each one of the fourth possibilities (corresponding to the section 4.3). That way, it could justify the replication of rules containing only two different structures (in fact, the layer 1 describes one structure three times and layer 2 describes a second structure four times, Fig. 4)

========================
Can we do this? Possibly by hand?

========================

2. Several parts of the paper that were not clear or wrong in the previous submission have not been corrected but removed. I regret that choice for some of them, but I understand that the responses to reviewers (especially the formalisation) extend the paper. The paper has been deeply rewritten, and authors did choices that they have correctly justified.

===========

===========

3. Something is confused for me considering the Fig. 1 and its explanations in the text.
First, I think the section 4 could refer the Fig.1 to illustrate its explanations.
Second, I’m a bit confused with the difference between “a rule combination” and a “path condition” especially regarding the fig 1(b). Here is a list of remarks about this:
- Maybe the second paragraph of 2.3 could have a title in the same way than the following subsubsections “Disambiguating...” and “Generating...” because these two subsubsections are dedicated to figure 1(c) and 2 respectively, whereas the 2nd paragraph of 2.3 is dedicated to figure 1(b). Moreover, these subsubsections correspond to subsection 4.1 and 4.2 whereas the 2nd paragraph of 2.3 corresponds to subsection 4.3. I recommend cross-referring those (sub)subsections.
My confusion began with the transition between the two first paragraphs of the page 6, one ends producing “rule combination”, the other one begins consuming “path condition”.
My confusion is reinforced with the correspondences between the subsubsections of that section 2.3 and the subsection of section 4, especially the subsection 4.1. Is the Fig. 1(a) the result of the first part of section 4.1 or of the last part of section 4.1?
In the same way than in my previous review, I think that a sample of “path creation” is illustrated too late and without being dedicated to that illustration: Fig. 7(a). I understand that it is complicated section 2.3, but it could be done section 4.1.
- Section 2.3, first paragraph: “the first step ...” refers the Fig 1 and “the second step...” refers the Fig 2.
- 4th paragraph p.6, maybe it would be more appropriate to illustrate this paragraph with the Fig. 1(c) instead of Fig. 1(a)?
4. I appreciate that the Fig. 4 has been simplified. Unnecessary complexity has been removed. 
I still consider that a legend would provide an immediate understanding of the arrows.
5. Section 4.1 “Traceability Links” and Fig. 6: one can wonder if it is necessary to trace the links between elements?
6. Fig. 9 is not used in the text.
Result of Fig. 9(a) are letters different than the ones of the two source layers, but letters are the same in the two source layers and result in Fig. 9(b). Is a “working set” considering differently than a layer (I don’t think so)?
7. Section 4.3, the link between the enumeration (1., 2., 3., 4.) and the subsubsections could be emphasised.

=================

We thank the reviewer for their detailed comments. As we reorganized the paper, their suggestions have guided us in our editing process. Further precisions have been made in the areas highlighted in these comments. Due to the large changes made to the paper, we are unable to provide a point-by-point rebuttal.

=================

 8. I’m not convinced with the very small time to disprove a property. I understand that as soon as a counterexample is found the algorithm stops. However, I’m wondering if the result (0.003 second) is not due to the simplicity of the property. Maybe with a complex property or just a different one, the algorithm should study more path conditions before finding a counter-example?

==============

Future work will consider the effect of the size of the property and path condition on the execution time of the algorithm. However in the section on complexity we discuss that the main complexity in this step comes from graph matching. Therefore, an efficient implementation should scale well to larger properties.

What if we scanned all path conditions? How long does it take then?

==============

TYPOS
p.27 “Note that the fact that that”: maybe one useless “that”. 
p.29 “the the properties” -> “the properties”
p.41 v’ member of V -> v’ member of V’ ?
p.42 <V’, E’, t> -> <V’, E’, t’> maybe?

### 2015-03-13 presentations/poster_NECSIS_2014.tex
previous version of the paper

\documentclass[final,hyperref={pdfpagelabels=false}]{beamer}
\usepackage{grffile}
\usepackage{ragged2e}
\mode<presentation>
  {
  %  \usetheme{Berlin}
  \usetheme{I6pd2}
  }
  \usepackage{times}
  \usepackage{amsmath,amsthm, amssymb, latexsym}
  \boldmath
  \usepackage[english]{babel}
  \usepackage[latin1]{inputenc}
  \usepackage[orientation=portrait,size=custom,width=80,height=120,scale=1.4]{beamerposter}

  % my packages 
  \usepackage{multirow}

  %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%5
  \graphicspath{{figures/}}
  \title[Fancy Posters]{Verification of Translation Model Transformations\\\vspace{.5cm}Validity and Completeness Results}

  \author[Dreuw \& Deselaers]{Levi L\'ucio$^{\dagger}$, Bentley James Oakes$^{\dagger}$, and Hans Vangheluwe$^{\ddagger\dagger}$}

    \institute[McGill University]{$^{\dagger}$McGill University, Montreal,
    Canada\qquad $^{\ddagger}$University of Antwerp, Belgium}

  \date[June 14th, 2011]{\today}

  %%%%%%%%%%%%%%%%%%%%%%%%%%%%%
  \begin{document}
  \begin{frame}{} 
     \vspace{-1.5cm}
      \begin{columns}[t]
        \begin{column}{.993\linewidth}
        \begin{block}{\large Problem Statement}\vspace{.5cm}
 Our transformation verification approach employs a technique inspired from
 symbolic execution in order to automatically build a set of path
 conditions for a DSLTrans transformation~\cite{MODELS2010,SLE2010}. Each path condition abstractly represents a number of concrete
 transformation executions. Based on this formally defined abstraction relation, we show that \textbf{our path condition
 generation algorithm is both \emph{valid} and \emph{complete}}.

 Using \textbf{path conditions}, we are then able to prove model syntax relation properties over all transformation
 executions in a model-independent way.
 This is done by examining if the properties hold on the set of path conditions.
 Through the \textbf{abstraction relation}, if the property does not hold on a path
 condition, the property will not hold on the transformation executions
 abstracted by the path condition. We show \textbf{this property proving step is
 both \emph{valid} and \emph{complete}}.\vspace{1cm}
        \end{block}
        \end{column}
      \end{columns}

	 \begin{columns}[t]
	   \begin{column}{.71\linewidth}
        \begin{block}{Main Artifacts in Model Transformation Verification}
        \begin{columns}[c]
       \begin{column}{.50\linewidth}
       \begin{center}
         \textbf{DSLTrans Transformation:}\vspace{2cm}
         \includegraphics[width=.9\textwidth]{../figures/presentation/artifacts/transformation}\\
         \vspace{.7cm}\footnotesize{
         Six rules arranged in three layers that execute sequentially.  
         Each rule is composed of a Match Model and an Apply Model, analogous to the LHS and RHS in other transformation formalisms.\vspace{1cm}}
         \end{center}
         \end{column}
         \vrule{}
               \begin{column}{.24\linewidth}
               \begin{center}
          \vspace{-1cm}\textbf{Property:}\vspace{1cm}
            \includegraphics[width=\textwidth]{../figures/presentation/artifacts/property}\\
            %\vspace{.5cm}
            \small{Model Syntax Relation\\ \emph{AtomicContract} with\\pre- and post- conditions} \\\vspace{1cm}\footnotesize{``the mother and father in a family will always be transformed into a woman
            and a man''} 
            \end{center}
            \end{column} 
            \vrule{} 
        \begin{column}{.24\linewidth}
        \begin{center} 
          \vspace{-1.5cm}\textbf{Transformation\\ Execution:}\vspace{0.5cm}
          \includegraphics[width=.99\textwidth]{../figures/presentation/artifacts/trans_ex}\\
          \vspace{0.5cm}\small{Input and resulting Output}\\\vspace{1.2cm}\footnotesize{\textit{Traceability links} record which output elements were created from which input elements}
          \end{center} 
          \end{column} 
              \end{columns} 
        \end{block}
        \end{column}

        \begin{column}{.26\linewidth}

         \begin{block}{Proving Properties}
         \vspace{1.5cm}
        \small{
        The aim of our procedure is to prove \emph{model syntax relation}
        properties on all transformation executions. However, as there are
        \textbf{infinitely many transformation executions}, the proof requires
        an \textbf{abstraction}.\\~\\
		\vspace{.5cm}
        \textbf{Path conditions} abstract transformation executions. A path condition represents a combination of rules of the transformation that can execute. In a path condition each rule is represented only once, granting the required abstraction and ensuring the generation of a finite number of path conditions.} 
        \vspace{1.55cm}
%         Through the abstraction relation shown below \textbf{the set of path conditions generated for a DSLTrans transformation represents all possible executions of that transformation given an arbitrary input model}.
        \end{block}
        \end{column}
       \end{columns}

      \begin{columns}[t]
        \begin{column}{.993\linewidth}      
			\begin{block}{Abstraction Relation}
                \begin{columns}[t]
                \begin{column}{.24\linewidth}
                \begin{center}

                \textbf{Example Path Condition:}\\
              \includegraphics[width=0.8\textwidth]{../figures/presentation/artifacts/path_condition}\\

             \footnotesize{Path conditions are created by combining transformation rules~\cite{LuVa2013a}}\vspace{1cm}

%              \includegraphics[width=0.7\textwidth]{../figures/presentation/abstraction_relation}\\
%              One path condition can abstract many transformation executions

              \end{center}            
              \small{
              A \textbf{Path Condition} \emph{abstracts} a transformation execution when:}
              \vspace{.2cm}
              \begin{itemize}
                \item \small{An \textbf{\emph{Injective Typed Graph Homomorphism}} exists between the match part of each rule (component) in the path condition and the input of the transformation execution, \textbf{and}}\vspace{.5cm}
                \item \small{A \textbf{\emph{Surjective Typed Graph Homomorphism}} exists between the output of the transformation execution (including traceability links) and the apply graph of the path condition.}
              \end{itemize}

              \end{column}
              \vrule{}
              \begin{column}{.65\linewidth}
%               \large
%               \begin{center}
%              \textbf{\normalsize{Abstraction Holds}}\\
%               \end{center}
%              
%              \normalsize
%               
%               \begin{column}{.5\linewidth} 
%                \begin{center}
%                     \textbf{Empty Path Condition:}\\
%                  \includegraphics[width=\textwidth]{../figures/presentation/abstraction_relation/empty_pc}
%                 
%                                  
%                      \textbf{Non-Overlapping Rules:}\\
%                   \includegraphics[width=\textwidth]{../figures/presentation/abstraction_relation/non_overlapping2}
%                   
%                    \end{center} 
%                  \end{column}
%                  \begin{column}{.5\linewidth}
%                   \begin{center}
%                    \textbf{Non-Overlapping Rules:}\\
%                                      \includegraphics[width=\textwidth]{../figures/presentation/abstraction_relation/non_overlapping}
%                    
%                  
%                   \textbf{Overlapping Rules:}\\
%  	            \includegraphics[width=\textwidth]{../figures/presentation/abstraction_relation/overlapping}
%  	            
%  	            \end{center} 
%                 \end{column}
%                 
%                    \begin{center}
% %                   \textbf{Given this abstraction relation, properties that are proven on a path condition will hold on all transformation executions which the path condition abstracts over}
%                		\includegraphics[width=0.3\textwidth]{../figures/abstraction_relation/legend}                   
%                    \end{center}
%                  
%                    \end{column}
%                   \vrule{}
%                   \begin{column}{.24\linewidth}
%                    \begin{center}
%                    
%                    \large
%                    \textbf{\normalsize{Abstraction Does Not Hold}}
%                    \normalsize
%                    
%                   \textbf{Empty Path Condition:}\\
%                   \includegraphics[width=\textwidth]{../figures/presentation/abstraction_relation/empty_pc2}
%                   
%                   Surjective typed graph homomorphism does not hold
%                                          
%                     
%                         \textbf{Overlapping Rules:}\\
%         	            \includegraphics[width=\textwidth]{../figures/presentation/abstraction_relation/overlapping2}
%         	            
%         	            Injective typed graph homomorphism does not hold
%                         \end{center}

              \includegraphics[width=1\textwidth]{../figures/presentation/abstraction_relation/abstraction}\\

                    \end{column}
                 \end{columns}
%                 \vspace{1cm}
                \end{block}
      		\end{column}
        \end{columns}

%   \vspace{-1.5cm}
   \begin{columns}[t]
                   \begin{column}{.485\linewidth}

			\begin{block}{Path Condition Generation Validity/Completeness Theorems~\cite{LuVa2013a}}
                \begin{columns}[t]
                \begin{column}{.4\linewidth}
              \textbf{Validity:}
              \small{Every path condition abstracts at least one transformation execution.}

              \end{column}
              \hspace{-2cm}
			  \vrule{} 
              \hspace{1cm}

              \begin{column}{.485\linewidth}
                  \textbf{Completeness:}
                    \small{Every transformation execution is abstracted by one
                    path condition.}
                 \end{column}                            
                 \end{columns} 
                 \vspace{0.5cm} 
%			  	 \hrule{}  
                 \vspace{0.5cm} 
                 \small{\textbf{Uniqueness:}}
                 \footnotesize{Every transformation execution is abstracted by \textbf{\textit{exactly}} one
                 path condition.}

                \end{block}

                \end{column} 
                \begin{column}{.485\linewidth}
       \begin{block}{Property Proving Validity/Completeness Theorems~\cite{LuVa2013a}}
                       \begin{columns}[t]
                       \begin{column}{.485\linewidth}
                     \textbf{Validity:}
                     \small{Checking a property on a path condition and on a transformation execution that path condition abstracts produces the same result.}
                     \end{column}
              		 \hspace{-2cm}                     
                     \vrule{}
               		 \hspace{.3cm}
                     \begin{column}{.485\linewidth}
                          \textbf{Completeness:}
                        \small{A property can be proved to hold or not hold on all transformation executions.}
      					\vspace{1.93cm} 
                        \end{column}
                       \end{columns}
                       \end{block}                    
      \end{column}
      \end{columns}

   \begin{columns}[t] 
    \begin{column}{.485\linewidth}
    	\begin{block}{Future Work}
   	   \begin{itemize}
   	   \item Integrate in the theory the treatment of rules whose match graphs are subgraphs of other rules (already in the tool).
   	   \item Integrate contract negation in the theory (already in the tool).
   	   \item Integrate negative application conditions (NACs) into our symbolic verification approach.
   	   \item Expand expressiveness of the property language by allowing properties that involve multiple applications of the same rule.
   	   \end{itemize}
        \end{block}

    \end{column}
    \begin{column}{.485\linewidth}
    	\begin{block}{Bibliography}
    	   \footnotesize
    	   \begin{thebibliography}{10}   
    	   \bibitem{MODELS2010} Levi L\'ucio, Bruno Barroca, Vasco Amaral, {\em A
    	   Technique for Automatic Validation of Model Transformations},
    	   Proceedings of the MoDELS 2010 Conference, Springer, pp. 136-150.
    	   \bibitem{SLE2010} Bruno Barroca, Levi L\'ucio, Vasco Amaral, Roberto
    	   F{\'e}lix and Vasco Sousa, {\em DSLTrans: A Turing Incomplete
    	   Transformation Language}, Proceedings of the SLE 2010 conference,
    	   Springer 2010, pp. 296-305.
 		   \bibitem{LuVa2013a} Levi L\'ucio, Bentley Oakes and Hans Vangheluwe, {\em A Technique for Symbolically Verifying Properties of Graph-Based Model Transformations},  Technical Report SOCS-TR-2014.1, McGill University, 2013.
		\vspace{1.65cm}
		\end{thebibliography}	   
    	\end{block}        
    \end{column}    
   \end{columns}

  \end{frame}
\end{document}

### 2015-03-13 presentations/presentation.tex
previous version of the paper

\documentclass[xcolor=dvipsnames, 14pt]{beamer}
\useoutertheme{infolines} 
\usecolortheme[named=OliveGreen]{structure} 
\usetheme[height=9mm]{Rochester}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{epsfig}
\usepackage{caption}  
\usepackage{subcaption} 
\author[Lucio, Oakes, Vangheluwe]{Levi Lucio, Bentley James Oakes, and Hans Vangheluwe}
\title[Sym. Veri. of Translation Model Trans.]{Symbolic Verification of
Translation Model Transformations}

\AtBeginSection[]
{
   \begin{frame}
       \frametitle{Outline}
       \tableofcontents[currentsection]
   \end{frame}
}

%\setbeamercovered{transparent} 
\setbeamertemplate{navigation symbols}{} 
%\logo{} 
%\institute{} 
\date{\today} 
%\subject{} 
\begin{document}

\begin{frame}
\titlepage
\end{frame}

\begin{frame}
\tableofcontents
\end{frame}

\section{Overview}
\begin{frame}{Overview}
\begin{itemize}
\item Motivation is to formally prove properties on model transformations
\item A method to construct set of path conditions for a DSLTrans transformation
\item Set of path conditions represent all possible transformation executions
\item Prove structural properties on path conditions implies properties proved on transformation
\item Method is transformation-independent and graph-based
\end{itemize}
\end{frame}

\begin{frame}{Contributions}
\begin{itemize}
\item Algorithm to construct path conditions
\item Algorithm to prove properties on these path conditions
\item Validity and completeness proofs of these algorithms
\item Performance and scalability results
\end{itemize}
\end{frame}

\section{DSLTrans}
\begin{frame}{DSLTrans Overview}
\begin{itemize}

\item Transformation language
\item Turing-incomplete and outplace
\begin{itemize}
\item Avoids unbounded recursion or non-determinism
\item Allows construction of provably-finite set of path conditions
\end{itemize}
\item Transformation composed of rules in layers
\begin{itemize}
\item Rules in first layer are repeatedly executed in deterministic random order until exhausted
\item Next layer of rules is then executed
\end{itemize}
\end{itemize}
\end{frame}

\begin{frame}{Metamodels for Running Example}
\begin{figure}[htb]
        \centering
        \begin{subfigure}[b]{0.50\textwidth}
                \centering
                \includegraphics[width=1\textwidth]{../figures/policestation_dsltrans/squadagent.png}
                \caption{Chains of command}
                \label{fig:OrganizationLanguage}
        \end{subfigure}%
        ~~
        \begin{subfigure}[b]{0.50\textwidth}
                \centering
                \includegraphics[width=1\textwidth]{../figures/policestation_dsltrans/squadsex.png}
                \caption{Gender classification view}
                \label{fig:GenderLanguage}
        \end{subfigure}%
        %\caption{Metamodels for the Police Station Running Example}
        \label{fig:squadmetamodel}
\end{figure}
\end{frame}

\begin{frame}{Example DSLTrans Transformation}
\begin{figure}[h!]
	\centering
		\includegraphics[width=1\textwidth]{../figures/policestation_dsltrans/police_station_transformation.pdf}
	%\caption{A model transformation expressed in DSLTrans}
	\label{fig:dsltransformation}
\end{figure}
\end{frame}

\begin{frame}{Model Before/After Transformation}
Objective is to flatten a chain of command given in the Organization language language into two independent sets of male and female officers represented in the Gender language
\begin{figure}[htb]
        \centering
        \begin{subfigure}[b]{0.40\textwidth}
                \centering
                \includegraphics[width=1\textwidth]{../figures/policestation_dsltrans/model_police_hierarchy.pdf}
                \caption{Original model}
                \label{fig:police_hierarchy}
        \end{subfigure}%
        ~~
        \begin{subfigure}[b]{0.50\textwidth}
                \centering
                \includegraphics[width=1\textwidth]{../figures/policestation_dsltrans/model_police_gender.pdf}
                \caption{Transformed model}
                \label{fig:police_gender}
        \end{subfigure}%
        %\caption{Model before and after transformation}
        \label{fig:transformationexample}
\end{figure}
\end{frame}

\section{Path Condition Construction}

\begin{frame}{Path Conditions}
\begin{itemize}
\item Similar to symbolic path in symbolic execution
\item Formally represents concrete transformation executions
\item Composed of a match graph and an apply graph
\begin{itemize}
\item Path conditions look almost identical to DSLTrans rules
\end{itemize}
\end{itemize}
\end{frame}

\subsection{Building Path Constructions from Layers}

\begin{frame}{Rule Combinations}
\begin{itemize}
\item Take powerset of rules in first layer
\begin{itemize}
\item A, AB, ABC, BC
\item Abstract over number of times a rule executes
\end{itemize}
Also add traceability information to rules at this time
\end{itemize}

\end{frame}

\begin{frame}{Path Conditions}
\begin{itemize}
\item Take rule combination i.e. AC
\item Union match graphs together
\item Union apply graphs together
\end{itemize}

\end{frame}

\subsection{Disambiguating Path Conditions}

\begin{frame}{Disambiguating}

Match elements may refer to same element during transformation execution. Therefore, must disambiguate (recursively).
\begin{figure}[htb]
        \centering
        \begin{subfigure}[b]{0.40\textwidth}
                \centering
                \includegraphics[width=1\textwidth]{../figures/building_path_conditions/collapse_locate_match.pdf}
                \caption{Locating elements}
                \label{fig:collapse_locate_match}
        \end{subfigure}%
        ~~
        \begin{subfigure}[b]{0.40\textwidth}
                \centering
                \includegraphics[width=1\textwidth]{../figures/building_path_conditions/collapse_merge_match.pdf}
                \caption{Merging elements}
                \label{fig:collapse_merge_match}
        \end{subfigure}%
        %\caption{Selecting and merging match elements of the same type}
        \label{fig:merge_match_elements}
\end{figure}
\end{frame}

\subsection{Generating All Path Conditions}

\begin{frame}{Combination Overview}
\begin{figure}[htb]
        \centering
        \begin{subfigure}[b]{0.55\textwidth}
                \centering
                \includegraphics[width=1\textwidth]{../figures/building_path_conditions/combining_path_conditions.pdf}
                \caption{\textbf{Combining path conditions from two layers to produce working set of path conditions}}
                \label{fig:create_working_set}
        \end{subfigure}%
        ~
        \begin{subfigure}[b]{0.42\textwidth}
                \centering
                \includegraphics[width=1\textwidth]{../figures/building_path_conditions/next_layer.pdf}
                \caption{Combining the working set with the next layer's path conditions}
                \label{fig:combine_working_set}
        \end{subfigure}%
        %\caption{Combining path conditions from all layers}
        \label{fig:combining_path_conditions}
\end{figure}
\end{frame}

\begin{frame}{Combining Two Path Conditions}
\begin{itemize}
\item Take path conditions A from layer 1, and B from layer 2
\item Four different cases for combination
\begin{enumerate}
\item No interaction between A and B
\item A prevents B from holding
\item Both A and B \textit{may} hold
\item Both A and B \textit{will} hold
\end{enumerate}
\end{itemize}
Dependencies must be resolved by algorithm
\end{frame}

\begin{frame}{Path Condition Dependencies}
\begin{itemize}
\item Dependencies represented by backward link
\item Backward links are a construct in the DSLTrans rule language
\begin{itemize}
\item Enforce that a particular element in a apply graph was created from the connected element in the match graph
\end{itemize}
\end{itemize}
\begin{figure}[t]
                \centering
                \includegraphics[height=0.66\textheight]{../figures/building_path_conditions/collapse_merge_apply.pdf}
                %\caption{Merging elements}
                \label{fig:collapse_merge_match}
\end{figure}
\end{frame}

\begin{frame}{Case 1}
\begin{figure}[h!] \centering \includegraphics[width=0.7\textwidth]{../figures/building_path_conditions/no_dependencies.pdf}
	\caption{No dependencies between A and B}
	\label{fig:no_dependencies}
\end{figure}
\end{frame}

\begin{frame}{Case 2}
\begin{figure}[h!] \centering \includegraphics[width=0.7\textwidth]{../figures/building_path_conditions/non_satisfied_dependencies.pdf}
	\caption{B's dependencies are not satisfied by A}
	\label{fig:non_satisfied_dependencies}
\end{figure}
\end{frame}

\begin{frame}{Case 3}
\begin{figure}[h!] \centering \includegraphics[width=\textwidth]{../figures/building_path_conditions/partial_satisfied_dependencies.pdf}
	\caption{B's dependencies are partially satisfied by A}
	\label{fig:partial_satisfied_dependencies}
\end{figure}
\end{frame}

\begin{frame}{Case 4}
\begin{figure}[h!] \centering \includegraphics[width=\textwidth]{../figures/building_path_conditions/total_satisfied_dependencies.pdf}
	\caption{B's dependencies are fully satisfied by A}
	\label{fig:total_satisfied_dependencies}
\end{figure}
\end{frame}

\begin{frame}{Repeated Combinations}
\begin{figure}[htb]
        \centering
        \begin{subfigure}[b]{0.55\textwidth}
                \centering
                \includegraphics[width=1\textwidth]{../figures/building_path_conditions/combining_path_conditions.pdf}
                \caption{Combining path conditions from two layers to produce working set of path conditions}
                \label{fig:create_working_set}
        \end{subfigure}%
        ~
        \begin{subfigure}[b]{0.42\textwidth}
                \centering
                \includegraphics[width=1\textwidth]{../figures/building_path_conditions/next_layer.pdf}
                \caption{\textbf{Combining the working set with the next layer's path conditions}}
                \label{fig:combine_working_set}
        \end{subfigure}%
        %\caption{Combining path conditions from all layers}
        \label{fig:combining_path_conditions}
\end{figure}
\end{frame}

\section{Proving Properties}
\begin{frame}{Property Structure}
\begin{itemize}
\item Very similar to DSLTrans rules and path conditions
\item Composed of pre-condition and post-condition graphs
\end{itemize}

\end{frame}

\begin{frame}{Property Examples}
Property 1 -- Expected to hold: \small A model which includes a police station that has
both a male and female chief officers will be transformed into a model where the male
chief officer will exist in the male set and the female chief officer will exist in the female
set
\normalsize

\begin{figure}[htb]
        \centering
                \centering
                \includegraphics[height=.7\textheight]{../figures/policestation_dsltrans/satisfied.pdf}
                %\caption{Property 1 -- Expected to hold}
                \label{fig:dsltrans_prop1}

\end{figure}
\end{frame}
\begin{frame}{Property Examples}
Property 2 -- Not expected to hold: \small Any model which includes a female officer will
be transformed into a model where that female officer will always supervise another
female officer
\normalsize

\begin{figure}[htb]
        \centering
                \centering
                \includegraphics[height=.7\textheight]{../figures/policestation_dsltrans/unsatisfied.pdf}
                %\caption{Property 1 -- Expected to hold}
                \label{fig:dsltrans_prop1}

\end{figure}
\end{frame}

\begin{frame}{Repeated Combinations}
\begin{figure}[htb]
        \centering
        \begin{subfigure}[b]{0.55\textwidth}
                \centering
                \includegraphics[width=1\textwidth]{../figures/building_path_conditions/combining_path_conditions.pdf}
                \caption{Combining path conditions from two layers to produce working set of path conditions}
                \label{fig:create_working_set}
        \end{subfigure}%
        ~
        \begin{subfigure}[b]{0.42\textwidth}
                \centering
                \includegraphics[width=1\textwidth]{../figures/building_path_conditions/next_layer.pdf}
                \caption{\textbf{Combining the working set with the next layer's path conditions}}
                \label{fig:combine_working_set}
        \end{subfigure}%
        %\caption{Combining path conditions from all layers}
        \label{fig:combining_path_conditions}
\end{figure}
\end{frame}

\section{Implementation}

\begin{frame}{Implementation Overview}
\begin{itemize}
\item Prototype built in Python and T-Core
\begin{itemize}
\item T-Core is library for efficiently matching/rewriting graphs
\end{itemize}
\item Naive complexity is calculated to be larger than O($2^{2*r}$) where $r$ is the number of rules in the transformation
\begin{itemize}
\item Complexity made worse when disambiguation is required
\end{itemize}
\item Optimizations such as memoisation/caching/lazy disambiguation performed
\end{itemize}
\end{frame}

\begin{frame}{Metrics for PC Creation}
\begin{figure}[htb]
        \centering
        \begin{subfigure}[b]{0.40\textwidth}
                \centering
                \includegraphics[width=1\textwidth]{../figures/results/rules_vs_pcs.pdf}
                \caption{Number of rules vs. path conds. created}
                \label{fig:rules_vs_pcs}
        \end{subfigure}%
        ~~
        \begin{subfigure}[b]{0.40\textwidth}
                \centering
                \includegraphics[width=1\textwidth]{../figures/results/rules_vs_time.pdf}
                \caption{Number of rules vs. time taken}
                \label{fig:rules_vs_time}
        \end{subfigure}%

        %\caption{Metrics for the path condition creation process}
        %\label{fig:path_condition_building}
\end{figure}
\footnotesize
\begin{tabular}{|p{1.3cm}|c|c|c|c|c|c|c|c|c|}
\hline
Rules & 3&  5&  7&  10&  12&  14 & 17 & 19 & 21\\\hline
Path conds. created & 8&  16&  34&  272&  442&  1156 & 9248 & 15028 & 39304\\\hline
Time taken(s) & 0.01&  0.13&  0.39&  1.87&  2.68&  9.00 & 59.08 & 97.52 & 369.19\\
\hline
\end{tabular}
\normalsize
\end{frame}

\begin{frame}{Metrics for PC Creation}
\begin{figure}[htb]
        \centering
        \begin{subfigure}[b]{0.5\textwidth}
                \centering
                \includegraphics[width=1\textwidth]{../figures/results/rules_vs_memory.pdf}
                \caption{Number of rules vs. memory used}
                \label{fig:rules_vs_pcs}
        \end{subfigure}%

        %\caption{Metrics for the path condition creation process}
        %\label{fig:path_condition_building}
\end{figure}
\footnotesize
\begin{tabular}{|p{1.6cm}|c|c|c|c|c|c|c|c|c|}
\hline
Rules & 3&  5&  7&  10&  12&  14 & 17 & 19 & 21\\\hline
Memory used (KB) & 0.08&  0.096&  0.17&  1.24&  1.83& 4.98 & 38.01 & 60.10 & 156.79\\
\hline
\end{tabular}
\normalsize
\end{frame}

\begin{frame}{Property-Proving Time}
\begin{figure}[htb]
        \centering
        \begin{subfigure}[b]{0.5\textwidth}
                \centering
                \includegraphics[width=1\textwidth]{../figures/results/pcs_vs_prop1.pdf}
                \caption{Time required to prove the property that holds on all path conditions}
                \label{fig:rules_vs_pcs}
        \end{subfigure}%

        %\caption{Metrics for the path condition creation process}
        %\label{fig:path_condition_building}
\end{figure}
\footnotesize
\begin{tabular}{|p{1.6cm}|c|c|c|c|c|c|c|c|c|}
\hline
Rules & 3&  5&  7&  10&  12&  14 & 17 & 19 & 21\\\hline
Proof time & -&  0.19&  1.26&  2.40&  3.40& 8.38 & 73.51 & 140.77 & 412.02\\
\hline
\end{tabular}
\normalsize
\end{frame}

\begin{frame}{Property-Proving Time}

For property that was not expected to hold for all path conditions, proof time is constant
\begin{itemize}
\item Algorithm detects counter-example quickly and does not need to process further path conditions
\end{itemize}

\normalsize
\end{frame}

\section{Concluding Material}
\begin{frame}{Contributions}

Contributions of this work include:
\begin{itemize}
\item Algorithm to construct a set of path conditions for a DSLTrans transformation
\item Property-checking algorithm to prove model syntax relation properties over these path conditions, and therefore over all concrete transformation executions
\item Validity and completeness proofs for the above algorithms (not shown in this presentation)
\item Discussion of optimisation and scalability concerns
\begin{itemize}
\item Successfully used in industrial case-study
\end{itemize}
\end{itemize}

\end{frame}

\begin{frame}{Future Work}

\begin{itemize}
\item Extend property language with attributes 
\item Negative DSLTrans constructs such as NACs
\item Application to further case studies
\item Develop tools to automatically create artifacts needed in algorithms
\begin{itemize}
\item Requires development of HOTs
\item Bentley is currently working on this, involves model evolution concerns
\end{itemize}
\end{itemize}

\end{frame}

\begin{frame}{Thank You}

This work has been developed in the context of the NECSIS project, funded by Automotive Partnership Canada.

\end{frame}

\end{document}

### 2015-03-13 text/abstract.tex
previous version of the paper

\begin{abstract}

As model transformations are a required part of model-driven development, it is crucial to provide techniques that address their formal verification. One approach that has proven very successful in program verification is \emph{symbolic execution}. The symbolic abstraction in these techniques allows formal properties to be exhaustively proved for all executions of a given program.
In our approach we apply the same abstraction principle to verify model transformations. Our algorithm builds a finite set of path conditions which represents all concrete transformation executions through a formal abstraction relation. We are then able to prove properties over all transformation executions in a model-independent way. This is done by examining if any created path condition violates a given property, which will produce a counterexample if the property does not hold for the transformation. We demonstrate that this property proving approach is both valid and complete. Implementation results are also presented here which suggest that our approach is feasible and can scale to real-world transformations.

\end{abstract}

\begin{keywords}
Model Transformations, Symbolic Verification, Translation 
\end{keywords}

% \begin{abstract}
% In the work we present in this paper we apply the \emph{program verification}
% concept of \emph{symbolic execution} to the verification of model
% transformations written in the Turing-incomplete DSLTrans language. Our symbolic
% execution construction algorithms leverage on DSLTrans' reduced expressiveness
% in order to cope with the classical state space explosion problem typical to
% exhaustive verification techniques. By producing a symbolic execution for a
% DSLTrans model transformation we are able to prove precondition-postcondition
% axioms that regard the syntactic structure of our input/output models. Our
% property proofs are model independent, meaning a proof holds for any input model
% that satisfies an axiom's precondition. We are also able to produce a
% counterexample if a property does not hold. A counterexample consists of a set
% of transformation rules that, when executed, lead to the property being violated.
% 
% We have used model transformation technology to build the prototype we have used
% in the experiments with out technique we report in this paper. Metamodels and
% model transformations are used not only to produce the intermediate models we
% use for symbolic execution construction, but also at the core of our prototype
% in the symbolic execution construction algorithms. 
% \end{abstract}

% In this paper we present a novel technique and a prototype implementation for
% proving properties of model transformations expressed in the DSLTrans language.
% The approach is based on symbolic execution and the properties we are interested
% in concern relations between the structure of the input and output models. In
% particular, the properties are implications of the form `if a structural
% relation between some elements of the source model holds, then another
% structural relation between some elements of the target model must also hold'.
% Our technique is transformation dependent but model independent, meaning the
% proofs we produce hold for all executions of a given DSLTrans transformation 
% specification running on any instance of the transformation's input metamodel.
%  
% Our proof technique is based on (i) building the finite symbolic execution for a
% given DSLTrans transformation and (ii) on checking whether the property holds
% for all elements of the finite symbolic execution execution set. We explain how 
% a symbolic execution and a proof is built with our technique by using model
% transformations written using our T-Core framework.

### 2015-03-13 text/abstraction_relation.tex
previous version of the paper

\section{Abstraction Relation between Path Conditions and Transformation Executions}
\label{sec:abstraction_relation}

% More informally, there exists a typed graph injective homomorphism between
% $Match_{pc}$ and $Input_x$  \textbf{and} $\langle V_x,E_x\setminus
% E_{mx},\tau_x\rangle$ is a typed graph strict instance of $\langle
% V_{pc},E_{pc}\setminus E_{pc},\tau_{pc}\rangle$.

% \levi{It is possible that this relation is required to be a typed graph strict
% instance between the path condition and the execution. Such a constraint would
% oblige each set of executions for a path condition to be truly distinct from
% the set of executions for another path condition. With the current definition
% the executions for a path condition including rule 'A' will overlap with the
% executions for a path condition including rules 'AB'. If this is so validity
% needs to be proved again for the new relation between the match patterns of PC
% and ex. Elements that do not show up in the match part of the rules are
% excluded from the typed graph strict instance.}

In this section we define the abstraction relation between the execution of a
DSLTrans transformation and the path condition that represents it.
This abstraction relation allows us to prove properties on a finite set of representative
path conditions, as created by the path condition generation algorithm. As this set is finite, our technique is  guaranteed to be decidable.

This section also presents our arguments that our path condition building
algorithm is both \emph{valid} and \emph{complete}. In this context
\emph{validity} means that for each path condition there exists at least one
transformation execution that it abstracts.
In other words, no path conditions are produced that lack a concrete
transformation execution counterpart. \emph{Completeness} of the symbolic
execution means that every transformation execution is abstracted by at least
one path condition.

Let us start by formally defining the notion of abstraction of a transformation
execution by a path condition.

\setcounter{equation}{0}

\begin{definition}{Abstraction of a Transformation Execution by a Path Condition\\}
\label{def:abstraction_pc_ex}
\CatchFileBetweenTags{\abstractionrel}{text/definitions}{abstractionrel}{\abstractionrel}
\end{definition}

% \cref{def:abstraction_pc_ex} enforces the fact that the matcher part of
% components of the path condition are found through an injective typed graph
% homomorphism in the execution's input model. Additionally, the graph including
% the output of the execution (including traceability information), can be
% mapped by a surjective typed graph homomorphism onto the apply model (plus the
% symbolic traceability links) of the path condition.

To understand the abstraction relation in \cref{def:abstraction_pc_ex}, recall
that during the construction of a transformation execution rules are matched
injectively in the input model. This information is present in the first
condition of the abstraction relation (Proposition~\ref{eq:abstr_input_output})
via the injective typed graph homomorphism between the match part of the copies
of rules ``glued'' onto the path condition and the containment transitive
closure of the input part of the transformation execution. This relation
enforces the fact that certain parts of the execution were found, or matched, by
certain parts of the path condition.
On the other hand, the surjection from the output of the execution towards the
apply part of the path condition models the fact that the output of the
execution has been completely built by instantiating the apply parts of the
rules contained in the path condition.

The second condition of the abstraction relation
(Proposition~\ref{eq:abstr_trace}) checks for the fact that symbolic
traceability links in the path condition and traceability links in the execution
correctly correspond to each other. This is modeled by the fact that all
strongly connected components in the path condition, composed only of symbolic
traceability links, are injectively found on the execution. This injection
models the fact that traceability graphs between individual or combined rules in
the path condition are necessarily found in the execution. Note that components
of the path condition are considered because of the fact that disconnected rules
in the path condition may have matched over common elements of a particular
execution. As such, a full injection between the complete traceability graph in
the path condition and the execution would be incorrect. Additionally, in the
second part of Proposition~\ref{eq:abstr_trace} we check the fact that every
traceability link in the execution can be found in the path condition. This
additional sanity check enforces that no spurious traceability links that could
not have been created by the rules present in the path condition exist in the
transformation execution.

%Finally, the last clause of the abstraction relation states that rule copies
%that are repeated a number of times in the path condition need to be found at
%least a similar amount of times in the abstracted transformation execution.

It is important to mention that another abstraction relation, weaker or
stronger, could have been chosen. The abstraction relation presented in
\cref{def:abstraction_pc_ex} suits our needs in the sense that it allows us to
demonstrate the validity and completeness of our proof technique, as we will
show in the text follows. Additionally, it is particularly interesting because
it makes sure that, given a DSLTrans transformation, each of its transformation
executions is abstracted by one and only one of its path conditions. This result
adds to the consistency of our theory and is also exposed later in this section.

\subsection{Examples}

In this section, we provide a number of examples to demonstrate the workings of
the abstraction relation we chose to use. \cref{fig:legend} presents the legend
for the following figures.

\begin{figure}[htb]
 \centering
                \includegraphics[width=0.5\textwidth]{./figures/abstraction_relation/legend.pdf}
                \caption{Legend for abstraction relation figures}
                \label{fig:legend}
\end{figure}

\subsubsection{Example 1 -- Empty Path Condition}

We begin by defining which transformation executions an empty path condition
will abstract. \cref{fig:empty_pc} demonstrates two cases. In each, the path
condition is on the left-hand side, and a transformation execution is on the
right-hand side. Note that in \cref{fig:empty_pc_success}, the path condition
abstracts the transformation execution, while in \cref{fig:empty_pc_fail}, the
abstraction relation does not hold.

\begin{figure}[htb]
        \centering
        \begin{subfigure}[b]{0.40\textwidth}
                \centering
                \includegraphics[width=1\textwidth]{./figures/abstraction_relation/empty_pc.pdf}
               	\caption{Abstraction holds}
               	\label{fig:empty_pc_success}
        \end{subfigure}%
        ~~\\
        \begin{subfigure}[b]{0.40\textwidth}
                \centering
                \includegraphics[width=1\textwidth]{./figures/abstraction_relation/empty_pc2.pdf}
                \caption{Abstraction does not hold}
                \label{fig:empty_pc_fail}
        \end{subfigure}%
        \caption{Abstraction of transformation executions by the empty path condition}
        \label{fig:empty_pc}
\end{figure}

The match part of the path condition represents the pre-conditions for the path condition to be true, depending on which rules have symbolically executed in the transformation. For example, if the match graph is empty, this represents all executions where no rules have executed.

The first condition for the abstraction relation is to determine whether a typed graph injective homomorphism can be found between the match graph of the path condition, and an transformation executions. Note that in both \cref{fig:empty_pc_success} and \cref{fig:empty_pc_fail}, an empty typed graph homomorphism satisfies this condition, highlighted by blue arrows.

The second condition for the abstraction relation is whether a typed graph surjective homomorphism can be found from the transformation execution's output model to the apply graph of the path condition. This is represented by orange arrows in \cref{fig:empty_pc_success} and \cref{fig:empty_pc_fail}. This relation is surjective as there may not be any elements in the output model that are not represented by the path condition's apply graph. Note that multiple elements in the output model may match to the same element in the apply graph of the path condition. This is expected, as the structure found in the apply graph may be found multiple times in the output model.

The empty apply graph of the path condition defines no post-conditions on the output model, as no rules have executed. Note that there an empty surjective typed graph homomorphism can be found between the output model of the transformation execution in \cref{fig:empty_pc_success} and the path condition. This is intuitive, as the lack of elements in the output model means no rules have executed, which corresponds to the lack of post-conditions defined by the path condition.

In contrast, there is no surjection between the elements of the output model in \cref{fig:empty_pc_fail} and the path condition. Note that the transformation execution has elements in the output model and thus at least one rule must have executed. However, the path condition does not represent that a rule has executed. Therefore, the path condition shown does not represent this execution.

%The fact that this transformation execution will be abstracted by at least one path condition is demonstrated by \cref{prop:pc_completeness}.

\subsubsection{Example 2 -- Non-overlapping Rule Components}

This second example shows the abstraction relation between path conditions and transformation executions, when no match element of the same type appears in multiple rule components.

For these examples, we will represent the abstraction relation with two figures. The first will demonstrate the matching performed on match and apply graphs, while the second figure focuses on traceability link matching.

Let us first examine how the injection operates between the match elements in the path condition and the transformation executions in \cref{fig:non_overlapping_match_apply} and \cref{fig:non_overlapping2_match_apply}. Note that this injection can be found in both cases.

\begin{figure}[htb]
        \centering
        \begin{subfigure}[b]{0.40\textwidth}
                \centering
                \includegraphics[width=1\textwidth]{./figures/abstraction_relation/non_overlapping.pdf}
               	\caption{Abstraction holds on match and apply graphs}
               	\label{fig:non_overlapping_match_apply}
        \end{subfigure}%
        ~~\\
        \begin{subfigure}[b]{0.40\textwidth}
                \centering
                \includegraphics[width=1\textwidth]{./figures/abstraction_relation/non_overlapping_trace_links.pdf}
                \caption{Abstraction holds on traceability links}
                \label{fig:non_overlapping_trace_links}
        \end{subfigure}%
        \caption{Abstraction of transformation execution by non-overlapping rule components}
        \label{fig:non_overlapping}
\end{figure}

\begin{figure}[htb]
        \centering
        \begin{subfigure}[b]{0.40\textwidth}
                \centering
                \includegraphics[width=1\textwidth]{./figures/abstraction_relation/non_overlapping2.pdf}
               	\caption{Abstraction holds on match and apply graphs}
               	\label{fig:non_overlapping2_match_apply}
        \end{subfigure}%
        ~~\\
        \begin{subfigure}[b]{0.40\textwidth}
                \centering
                \includegraphics[width=1\textwidth]{./figures/abstraction_relation/non_overlapping2_trace_links.pdf}
                \caption{Abstraction holds on traceability links}
                \label{fig:non_overlapping2_trace_links}
        \end{subfigure}%
        \caption{Example of abstraction over multiple rule execution}
        \label{fig:non_overlapping2}
\end{figure}

Similarly, there is a surjection between the elements of the output model for both transformation execution and the apply graph of the respective path condition. Note that this surjective match also holds in \cref{fig:non_overlapping2_match_apply}, where examination of the transformation execution shows that one rule has executed twice. As mentioned before, the abstraction relation abstracts over the number of times that a rule has executed.

We also note that these matches must also match over associations between the elements, including association typing. This is not included in the figures for visual clarity.

We now examine \cref{fig:non_overlapping_trace_links} and \cref{fig:non_overlapping2_trace_links} to resolve whether the traceability links in the path condition can be found in the transformation execution. This matching is represented by the arrows from each component highlighted in a bold outline and differentiated by colour. We note that each component in the path condition can be successfully found in the transformation execution.

As well, there is a matching step from each individual traceability link in the transformation execution onto the path condition. Similar to the matching from the path condition, the bold components in the transformation execution figure are matched onto the path condition. We note that this matching is successful as well.

\subsubsection{Example 3 -- Overlapping Rule Components}
\label{subsubsec:abstraction_relation_overlapping}

For these examples, the path conditions contain overlapping rule components, i.e. separate rules share match elements of the same type. Our goal is to illustrate the interaction of rule elements, where the elements of non-dependent rules may match over the same or different elements in the transformation execution.

For example, the two rule components in \cref{fig:overlapping_match_apply} correctly match over the transformation execution shown. The abstraction relation holds due to the fact that, while match elements of the same component need to be found injectively in the execution, the injection constraint does not span multiple components. This allows the match elements from different rules to match to the same input model element.

\begin{figure}[htb]
        \centering
        \begin{subfigure}[b]{0.40\textwidth}
                \centering
                \includegraphics[width=1\textwidth]{./figures/abstraction_relation/overlapping.pdf}
               	\caption{Abstraction holds on match and apply graphs}
               	\label{fig:overlapping_match_apply}
        \end{subfigure}%
        ~~\\
        \begin{subfigure}[b]{0.40\textwidth}
                \centering
                \includegraphics[width=1\textwidth]{./figures/abstraction_relation/overlapping_trace_links.pdf}
                \caption{Abstraction holds on traceability links}
                \label{fig:overlapping_trace_links}
        \end{subfigure}%
        \caption{Abstraction of transformation execution by overlapping rule components}
        \label{fig:overlapping}
\end{figure}

As well, \cref{fig:overlapping_trace_links} shows the mapping from the path condition to the transformation execution. However, note that the pattern composed of the A, B, and Y elements, along with the traceability links, is to be matched as a whole. This is to ensure that the traceability links are found in the proper configuration in the transformation execution.

We also match the traceability links from the transformation execution back onto the path condition. Again, this is to ensure that no traceability links are found in the transformation execution that have not been represented in the path condition. Three matches are performed in this step, denoted by the three arrows in the bottom of \cref{fig:overlapping_trace_links}. Each match is composed of a traceability link as well as immediately connected elements.

\begin{figure}[htb]
        \centering
        \begin{subfigure}[b]{0.40\textwidth}
                \centering
                \includegraphics[width=1\textwidth]{./figures/abstraction_relation/overlapping2.pdf}
               	\caption{Elements cannot overlap within a component}
               	\label{fig:overlapping2_match_apply}
        \end{subfigure}%
        ~~\\
        \begin{subfigure}[b]{0.40\textwidth}
                \centering
                \includegraphics[width=1\textwidth]{./figures/abstraction_relation/overlapping2_trace_links.pdf}
                \caption{Traceability links cannot be found}
                \label{fig:overlapping2_trace_links}
        \end{subfigure}%
        \caption{Abstraction does not hold}
        \label{fig:overlapping2}
\end{figure}

In contrast to \cref{fig:overlapping}, \cref{fig:overlapping2} shows an example where the abstraction relation does not hold. Consider \cref{fig:overlapping2_match_apply}. Note that a component in the match graph of the path condition contains two B elements. Both of these elements must be found in the transformation execution, and thus it is not correct for them to injectively match to the same element in the input model.

As well, it is informative to examine \cref{fig:overlapping2_trace_links}. Note the one of the matches from the transformation execution attempts to match over 'a:A' and 'y:Y' elements, connected by a traceability link. Examination of the path condition shows that this traceability link is not present. Therefore, this path condition cannot accurately represent this transformation execution.

\subsubsection{Example 4 -- Indirect Links}

\begin{figure}[htb]
        \centering
        \begin{subfigure}[b]{0.40\textwidth}
                \centering
                \includegraphics[width=1\textwidth]{./figures/abstraction_relation/indirect.pdf}
               	\caption{Matching over match and apply graphs}
               	\label{fig:indirect_match_apply}
        \end{subfigure}%
        ~~\\
        \begin{subfigure}[b]{0.40\textwidth}
                \centering
                \includegraphics[width=1\textwidth]{./figures/abstraction_relation/indirect_trace_links.pdf}
                \caption{Matching over traceability links}
                \label{fig:indirect_trace_links}
        \end{subfigure}%
        \caption{Abstraction with indirect links}
        \label{fig:indirect}
\end{figure}

\begin{figure*}[htb]
        \centering
        \begin{subfigure}[b]{0.70\textwidth}
                \centering
                \includegraphics[width=1\textwidth]{./figures/abstraction_relation/combination.pdf}
               	\caption{Abstraction holds on match and apply graphs}
               	\label{fig:combination_match_apply}
        \end{subfigure}%
        ~~\\
        \begin{subfigure}[b]{0.70\textwidth}
                \centering
                \includegraphics[width=1\textwidth]{./figures/abstraction_relation/combination_trace_links.pdf}
                \caption{Abstraction holds on traceability links}
                \label{fig:combination_trace_links}
        \end{subfigure}%
        \caption{Abstraction where path condition has combined rules}
        \label{fig:combination}
\end{figure*}

We now present a path condition  in \cref{fig:indirect_match_apply} that includes indirect links. In this case, for the injective match to hold, the elements at both ends of the link must be found, and there must be an indirect link between the matched elements and between the elements in the transformation execution.

Note that the indirect link between elements $a:A$ and $b:B$ in the transformation execution is added by the containment transitive closure $Input^{*}$ in proposition~\ref{eq:abstr_input_output} of \cref{def:abstraction_pc_ex} to allow matching indirect links. Note also that, for the sake of our example, we are assuming that the links between $a:A$, $b:b$ and $c:C$ are containment relations.

\cref{fig:indirect_trace_links} highlights the structures involved in matching over traceability links. From the path condition, the structure contains the A, B, and Y elements with connected traceability links. From the transformation execution, there are two structures to be found in the path condition denoted in bold in the transformation execution. The matching of all structures can be successfully performed, and thus this abstraction relation holds.

%Note that only the match graph of the path condition may include an indirect link.

\subsubsection{Example 5 -- Combined Rules}

\cref{fig:combination} shows a path condition that is composed of a multitude of rules which have been combined in the path condition generation algorithm. Each individual rule is surrounded by dashed lines.

Note that the matching on the match and apply graphs in \cref{fig:combination_match_apply} is similar to other examples. The combined rules can be considered as a single graph for the abstraction relation.

\cref{fig:combination_trace_links} shows the matching when multiple traceability links are present in the transformation execution. Note that each individual traceability link and the connected elements are matched onto the path condition.

\subsection{Validity and Completeness}
\label{subsec:abstraction_relation_validity_completeness}

In this section, we discuss the validity and completeness of the abstraction relation. Only proof sketches are presented here for \cref{prop:pc_validity} and \cref{lem:uniqueness}, while more complete proofs are found in \cref{sec:val_complete_path_cond_gen}, in the proofs of \cref{prop:pc_validity_appendix} and \cref{lem:uniqueness_appendix} respectively.

%\newpage
\begin{proposition}{(Validity)}
\label{prop:pc_validity}
Every path condition abstracts at least one transformation execution.
\end{proposition}
\begin{ps}

Let $tr\in \textsc{Transf}^{sr}_{tg}$ be a DSLTans transformation. We wish to demonstrate that, for all path conditions $pc\in \textsc{Pathcond}(tr)$, there exists a transformation execution $ex\in \textsc{Exec}(tr)$ of the set of
rules used to build $pc$ such that $pc$ abstracts $ex$, as formally expressed in \cref{def:abstraction_pc_ex}. We can prove this property by induction on the set of transformations $\textsc{Transf}^{sr}_{tg}$ (see \cref{def:layer_transformation}), as follows:

\begin{itemize}
  \item \emph{Base case:} the base case is when $tr$ is the empty
  transformation. In this case, according to \cref{def:path_cond_gen} only the
  empty path condition $\epsilon_{pc}$ exists in the path condition set. We thus
  need to demonstrate that the empty path condition abstracts the empty
  transformation execution $\epsilon_{ex}$, as well as any execution for which
  the input model is never matched by any rule (consequently having an empty
  output model). For any of these transformation executions, Proposition~\ref{eq:abstr_input_output} of the abstraction relation definition is satisfied, as no rule copy exists in the path condition and the output of the transformation execution is empty -- empty typed graph homomorphisms thus satisfy the all the conditions of the proposition. Proposition~\ref{eq:abstr_trace} of the abstraction relation definition also trivially holds because no traceability links exist either in the path condition or in any of the considered executions.

%   This is so because:
%   1) the empty typed graph homomorphism is an injective typed graph homomorphism
%   between the match part of the empty path condition and any input model; and 2) as
%   well as a surjective typed graph homomorphism between an empty output model
%   and the apply part of the empty path condition.

\item \emph{Inductive case:} assuming every path condition generated for a transformation $tr$ abstracts at least one transformation execution, we need to show that every path condition generated for a transformation $tr'$, resulting from adding a layer $l\in \textsc{Layer}^{sr}_{tg}$ to $tr$, will also abstract at least one transformation execution. 
\end{itemize}

In order to demonstrate the inductive case we need to show the property holds for all path conditions resulting from combining the rules of layer $l$ with any path condition generated for $tr$. These path conditions for transformation $tr'$ are built as expressed in \cref{def:path_cond_layer_comb}. According to this definition, path conditions for $tr'$ are built by incrementally combining the path conditions generated for $tr$ with a rule of layer $l$, until all the rules in $l$ have been treated. We can thus again use induction for this proof, this time on the set of possible layers $\textsc{Layer}^{sr}_{tg}$.

\begin{itemize}
  \item \emph{Base case:} this is the case where layer $l$ contains no rules. In this case, by the base case of \cref{def:path_cond_layer_comb}, no new path condition is added to the set of path conditions generated for the transformation $tr$. As such the $tr=tr'$ and by induction hypothesis the property trivially holds for all path conditions generated for $tr'$.

  \item \emph{Inductive case:} for the inductive case (transitive case of \cref{def:path_cond_layer_comb}) we need to show that, assuming the property holds for all path conditions generated for a transformation $tr$, then the property will also hold for a transformation $tr'$ -- where $tr'$ results from adding a new rule $rl$ to the last layer of $tr$. We will thus need to consider the four cases of rule combination:\vspace{.2cm}

\begin{enumerate}
\item\label{lab:rule_case1} Rule $rl$ has no dependencies (\cref{def:rule_comb_no_dependencies}).
\item\label{lab:rule_case2} Rule $rl$ has dependencies and cannot execute (\cref{def:rule_comb_unsatisfied}).
\item\label{lab:rule_case3} Rule $rl$ has dependencies and may and/or will execute (\cref{def:rul_comb_partial_total}).
\end{enumerate}

The property trivially holds for case~\ref{lab:rule_case2}, given that no new path conditions are added to the path condition set generated for $tr$ and that the property holds for $tr$ by induction hypothesis. When a rule $rl$ is added to the last layer of $tr$ such that cases~\ref{lab:rule_case1} or~\ref{lab:rule_case3} occur, then the property can be shown to hold for $tr'$ as follows: 1) choose for a general path condition $pc$ generated for $tr$ an execution $ex$ such that $pc$ abstracts $ex$; 2) build an input model $m$ as the result of uniting the input model of $ex$ with a model that can be matched by $rl$; 3) execute $tr'$ having as input model $m$ to produce transformation execution $ex'$; and finally 4) demonstrate $ex'$ is abstracted by the path condition $pc'$ resulting from combining $pc$ with $rl$ whether rule $rl$ does not depend on $pc$ or rule $rl$ depends on $pc$ and may and/or will execute.
\end{itemize}
\end{ps}

\begin{proposition}{(Completeness)}
\label{prop:pc_completeness}
Every transformation execution is abstracted by one path condition.
\end{proposition}
\CatchFileBetweenTags{\pathgencompletenessproof}{text/definitions}{pathgencompletenessproof}{\pathgencompletenessproof}

% Our path conditions construction algorithm takes all possible interactions of rules
% into account:
% \begin{enumerate}
% \item When building the path conditions for one layer we have built all
% the combinations of rules, as well as their possible interactions through
% \emph{disambiguation};
% \item When combining the path conditions from different layer different layers
% we have considered the cases where no interactions exist, or where backward
% links in rules requires certain certain elements to already exist in the
% generated model thus far.
% \end{enumerate}
% Note that, given the considered subset of DSLTrans constructs as described
% in \cref{sec:dsltrans}, no additional cases of rule interaction other than the
% ones described above need to be considered.
% From \cref{prop:pc_validity} we also know that at least one concrete
% transformation execution exists per path condition. In the proof of
% \cref{prop:pc_validity} we have built, for each path condition $A$,
% the simplest transformation execution that is abstracted by $A$. These
% transformation executions produce one instance of each element and association
% present in the apply pattern of $A$. However, given
% \cref{def:instance_pc_ex} in \cref{sec:proofs} (abstraction of
% a transformation execution by a path condition), an infinite amount of
% transformation executions are always abstracted by $A$. These executions
% correspond to the generation of output models holding an arbitrarily large
% amount of the elements and associations while still being abstracted by $A$.
% Because we have considered all possibilities of execution in our path condition
% construction algorithm, the union of all transformation executions built for
% each path condition constitutes the complete, infinite set of transformation
% executions. We thus know that every possible transformation execution is
% abstracted by at least one path condition.

% Note that we do not provide any guarantee that the same transformation execution
% cannot be abstracted by two different path conditions. Although given our symbolic
% execution assumptions we believe this to be the case, further mathematical
% exploration of this issue is necessary. Given however the fact that our
% verification algorithm exhaustively explores all path conditions, it is
% sufficient for the property proof to provide a \emph{no} result in the analysis
% of a path condition for the result of the verification of the property to be
% \emph{no}. \levi{keep this?}

\begin{lemma} (Uniqueness) 
\label{lem:uniqueness}
A transformation execution is abstracted by exactly one path condition.
\end{lemma}
\begin{ps}
Let $tr\in \textsc{Transf}^{sr}_{tr}$ be a model transformation. We will demonstrate that two different path conditions\\$pc_1, pc_2\in \textsc{Pathcond}(tr)$ cannot exist such that we have a transformation execution $ex\in \textsc{Exec}(tr)$ where $ex\Vvdash pc_1$ and $ex\Vvdash pc_2$.

We will do so by attempting to to build an $ex\in \textsc{Exec}(tr)$ such that $ex\Vvdash pc_1$ and $ex\Vvdash pc_2$ and demonstrating that it is always the case that such is not possible. In order to structure our argumentation, we will consider two cases:
% \levi{the contradiction happens because at each point of the execution, when a rule is added any two path conditions from the existing set of path conditions will represent different executions. At all points if we consider two path conditions in the PC set we have that two different executions can be created if we find a model that satisfies the first condition of the abstraction relation, and it is always the case that none of those two executions can be abstracted by the two path conditions simultaneously.}
\begin{enumerate}
  \item\label{item:uniqueness_rule_no_dep} the case where no rules in $tr$ have dependencies.
  \item\label{item:uniqueness_rule_has_dep} the case where some rules in $tr$ have dependencies.
\end{enumerate}

We start by considering that $tr$ falls into case~\ref{item:uniqueness_rule_no_dep} above. By~\cref{def:path_cond_gen} of path condition generation, each rule appears at most once in a path condition. Also, by construction, each path condition always contains a different combination of rules. We additionally know from \cref{def:layer_transformation} that the rules that compose $tr$ necessarily have non-overlapping matchers. We can nonetheless build a model $m$ as the typed graph union of two input models $m_1$ and $m_2$, where injective typed graph morphisms can be found between the match parts of the rule copies that form $pc_1$, and $m_1$. Injective typed graph morphisms can be found as well between the match parts of the rules that form $pc_2$, and $m_2$. We thus know that injective typed graph morphisms can be found between the rule copies that compose $pc_1$ and $pc_2$, and $m$. This satisfies the first condition of \cref{eq:abstr_input_output} in ~\cref{def:abstraction_pc_ex} of abstraction relation.

Let us now consider that $ex_1$ and $ex_2$ are obtained by executing the transformations rules combined into $pc_1$ and $pc_2$, having $m$ as input model. As mentioned above, we know that the rules in $pc_1$ and $pc_2$ are not completely overlapping. This means that, due to the way in which $m$ is built (explained above), $m$ will always have at least one input that is matched by rules of $pc_1$ but not by rules of $pc_2$ (and vice-versa). Thus, when the transformation rules combined into $pc_1$ execute having $m$ as an input model, there will always exist a traceability link generated between an input and an output element of $m$ that is not generated when the transformation rules combined into $pc_2$ execute having $m$ as an input model (and vice-versa). As such, we have that $ex_1$ is always different from $ex_2$ by at least one traceability link. Given that this traceability link is symbolically represented in either $pc_1$ or $pc_2$ (but not in both), according to condition \cref{eq:abstr_trace} in \cref{def:abstraction_pc_ex} it cannot be that either $pc_1$ or $pc_2$ abstract $ex_1$ and $ex_2$ simultaneously.\\\\
We will now analyse the scenario where $tr$ falls into case~\ref{item:uniqueness_rule_has_dep} above, where some rules in $tr$ have dependencies. For this case, assume we have a path condition $pc_1$ contained in the set of path conditions generated for $tr$, considering layers up to layer $l$ of $tr$ have executed. Assume also we have a rule $rl$ of layer $l+1$ of $tr$ that has dependencies and can be combined with $pc$. If rule $rl$ is totally combined with path condition $pc_1$, according to \cref{def:rul_comb_partial_total} and \cref{fig:multiple_total_satisfied_dependencies}, then nothing needs to be shown as $pc_1$ is not kept in the path condition set but rather replaced by its combination with $rl$. However, in case rule $rl$ is partially combined with $pc$, as defined in \cref{def:rul_comb_partial_total} and \cref{def:comb_path_cond_rule_single}, then multiple path conditions are generated and additionally $pc_1$ is kept in the path condition set.\\The proof will be complete once it is shown that for the path conditions that are generated when rules are partially combined, it is also the case that no two path conditions can abstract the same execution. This last part of the proof can be built in an analog fashion to the construction of the proof for point~\ref{item:uniqueness_rule_no_dep}. As previously mentioned, the complete proof can be found as \cref{lem:uniqueness_appendix} in \cref{sec:val_complete_path_cond_gen}.

\end{ps}

### 2015-03-13 text/appendix.tex
previous version of the paper

\section{Formal Definitions and Proofs}
\label{sec:proofs}

\subsection{Graph definitions}

We begin this section by formally defining the graphs and graph instances used in our algorithm as fundamental structures.
\begin{definition}{Typed Graph}
\label{def:typed_graph}
A typed graph is a triple $\langle V,E,\tau\rangle$ where $V$ is a finite set of
vertices, $E\subseteq V\times V$ is a set of edges connecting the vertices and
$\tau:V\rightarrow Type$ is a typing function for the elements of V, where $Type$ is
a set of type names. Edges $(v,v')\in E$ are noted $v\rightarrow v'$. The set of
all typed graphs is called $TG$.
\end{definition}

Definition~\ref{def:typed_graph_homomorphism} describes when a typed graph is said to be homomorphic to another.

\begin{definition}{Typed Graph Homomorphism  / Typed Graph Instance}
\label{def:typed_graph_homomorphism}

Let $\langle V,E,\tau\rangle=g,\langle V',E',\tau'\rangle=g'\in TG$ be typed
graphs. There is a typed graph homomorphism between $g'$ and $g$ if and only if
there exists a graph homomorphism $f:V'\rightarrow V$ such that for all $v'\in
V$ we have that $\tau'(v')=\tau(f(v'))$. Note that, trivially, a typed graph
homomorphism is a formal homomorphism. We denote an \emph{injective} typed graph
homomorphism between typed graphs $g$ and $g'$ by $g' \vartriangleleft g$.
\end{definition}

In the text that follows we will use the terms \emph{typed graph homomorphism}
and \emph{typed graph instance} interchangeably.

Definition~\ref{def:typed_graph_strict_instance} is stricter than
definition~\ref{def:typed_graph_homomorphism}, and indicates that there is a
strict many-to-one type correspondence of the nodes of two graphs and their
relations.

\begin{definition}{Typed Graph Strict Instance}
\label{def:typed_graph_strict_instance}

Let $\langle V,E,\tau\rangle=g,\langle V',E',\tau\rangle=g'\in TG$ be typed
graphs. $g'$ is a typed graph strict instance of $g$, noted $g'
\blacktriangleleft g$, if and only if there exists a surjective
typed graph homomorphism between $g'$ and $g$.

\end{definition}

In order to simplify the subsequent development we will often
use the notation ${\langle V,E,\tau\rangle}^*$ to represent the transitive
closure of the edges of typed graph  $\langle V,E,\tau\rangle$.

\subsection{Required Component Definitions}

Following along from our definition of the graphs used in our algorithm, this section then builds these graphs into the required structures.

\begin{definition}{Metamodel}
\label{def:metamodel}

A metamodel $\langle V,E,\tau\rangle\in TG$ is a typed graph where $\tau$ is a bijective typing function. The set of all metamodels is called META.

\end{definition}

Definition~\ref{def:metamodel} formally states that a metamodel corresponds to a graph of typed elements where only one element for each type is represented. This definition is then used to state definition~\ref{def:pattern}, where a typed graph is combined with its metamodel.

\begin{definition}{Pattern\footnote{Note that we the definition of pattern corresponds to the formal definition of \emph{Model} (definition 5) in our previous work presented in~\cite{Lucio:10}. The renaming was done in an effort to construct a more appropriate vocabulary to better describe our technique.}}
\label{def:pattern}

A pattern is a 4-tuple $\langle V,E,\tau,M\rangle$, where $\langle
V,E,\tau\rangle$ is a typed graph. Moreover $M=\langle V',E',\tau'\rangle\in
META$ is a metamodel and the codomain of $\tau$ equals the codomain of $\tau'$.
We also have that $\langle V,E,\tau\rangle$ is a typed graph instance of a
metamodel $M$. The set of all patterns for a metamodel $M$ is called $PAT^{M}$.
\end{definition}

% \begin{figure}[h!] \centering \includegraphics[scale=.5]{./figures/model_instance_example.pdf}
% 	\caption{An Example of a Model, instance of a Metamodel}
% 	\label{fig:model_instance_example}
% \end{figure}

Match and apply graphs are constructs of DSLTrans and are informally described in section~\ref{subsec:DSLTrans_constructs} for DSLTrans rules and in section~\ref{subsubsec:path_condition_creation} for path conditions. These graphs are formalised in definition~\ref{def:match_apply_model}.

\begin{definition}{Match-Apply Pattern}
\label{def:match_apply_model}

A match-apply pattern is a 5-tuple $\langle V,E,\tau, Match,Apply\rangle$,
where $Match=\langle V',E',\tau',s\rangle$ and $Apply=\langle
V'',E'',\tau'',t\rangle$ are patterns, $V=V'\cup V''$, $E= E'\cup E''$ and
$\tau=\tau'\cup \tau''$. $s$ is called the \emph{source} metamodel and $t$ the
target metamodel. The set of all match-apply patterns for a source metamodel $s$
and a target metamodel $t$ is called $MAP^{s}_{t}$.
%Vertices in the $Apply$ model which are not connected to \emph{backward links} are called \emph{free vertices}.
%We additionally define the $back:MAM^{s}_{t}\rightarrow MAM^{s}_{t}$ function
%which extends a match-apply model by adding edges that connect all vertices in
%the $Match$ model to all \emph{free} vertices.

\end{definition}
%(##) as several models are applied
%	match model created

Definition~\ref{def:path_condition} then defines a path condition to be largely constructed from a match graph, an apply graph, and traceability links between the graphs.

\begin{definition}{Path Condition}
\label{def:path_condition}

A Path Condition is a 7-tuple $\langle V,E\cup Tl\cup
Il,\tau,Match,Apply,Tl,Il\rangle$, where\\$\langle V,E,\tau,Match,Apply\rangle
\in MAP^{s}_{t}$ is a match-apply pattern. $Match=\langle V',E', \tau',s\rangle$,
$Apply=\langle V'',E'', \tau'',t\rangle$, the edges $Tl\subseteq V'\times V''$
are called \emph{traceability links} and the edges $Il\subseteq V'\times V'$ are
called \emph{indirect links}. The set of all path conditions having source
metamodel $s$ and target metamodel $t$ is called $PC^{s}_{t}$.
%We additionally define the $strip:TR^{s}_{t}\rightarrow TR^{s}_{t}$
%function which removes from a transformation rule all free vertices and
%associated edges.

\end{definition}

% The language to describe properties is in fact very similar to the language to
% express transformations, with the additional possibility of expressing indirect
% links in the $apply$ pattern --- thus allowing more abstract patterns than the
% ones expressed in transformations. This is natural given that the properties of
% a transformation can be more abstract than the rules implementing them.

A transformation execution is formed from taking the input model, and executing
the transformation on it to produce the output model. During this
transformation, traceability links will be placed between match elements and the
apply elements they create. Definition~\ref{def:transf_ex} expresses this
formally. Note that we assume that transformation executions are built following
the algorithm described in~\cite{DBLP:conf/sle/BarrocaLAFS10}.

\begin{definition}{Transformation Execution}
\label{def:transf_ex}

Let $t$ be a DSLTrans transformation having source metamodel $s$ and target
metamodel $t$. A Transformation Execution is a 6-tuple $\langle V,E\cup
Tl,\tau,Match,Apply,Tl\rangle$, where $\langle V,E,\tau,Match,Apply\rangle \in
MAP^{s}_{t}$ is a match-apply pattern. $Match=\langle V',E', \tau',s\rangle$,
$Apply=\langle V'',E'', \tau'',t\rangle$ and the edges $Tl\subseteq V'\times
V''$ are called \emph{traceability links}. The set of all transformation
executions having source metamodel $s$ and target metamodel $t$ is called
$Exec^{s}_{t}$.
\end{definition}

Definition~\ref{def:instance_pc_ex} is crucial to our work as it says that a path condition formally represents some number of concrete transformation executions. This will be used in the next section as a building block to say that properties proved on the path condition are also proved on the transformation executions it represents.

\begin{definition}{Abstraction of a Transformation Execution by a Path Condition}
\label{def:instance_pc_ex}

Let $t$ be a DSLTrans transformation having source metamodel $s$ and target
metamodel $t$. Let also $pc = \langle
V_{pc},E_{pc},\tau_{pc},Match_{pc},Apply_{pc},Tl_{pc},Il_{pc}\rangle \in
PC^{s}_{t}$ of be a path condition and $ex = \langle
V_x,E_x,\tau_x,Match_x,Apply_x,Tl_x\rangle \in Exec^{s}_{t}$ be an execution of
$t$. Let also $Match_x = \langle V_{mx},E_{mx},\tau_{mx}\rangle$ and $Match_{pc}
= \langle V_{mpc},E_{mpc},\tau_{mpc}\rangle$. We have that $ex$ satisfies $pc$,
noted $ex\Vvdash pc$, if and only if:
$$Match_{pc} \vartriangleleft Match^*_x \;\land\; \langle V_x,E_x\setminus
E_{mx},\tau_x\rangle \blacktriangleleft \langle V_{pc},E_{pc}\setminus
E_{mpc},\tau_{pc}\rangle$$ More informally, there exists a typed graph injective
homomorphism between $Match_{pc}$ and $Match_x$ \levi{It is possible that this
relation is required to be a typed graph strict instance between the path
condition and the execution. Such a constraint would oblige each set of
executions for a path condition to be truly distinct from the set of executions
for another path condition. With the current definition the executions for a
path condition including rule 'A' will overlap with the executions for a path
condition including rules 'AB'. If this is so validity needs to be proved again
for the new relation between the match patterns of PC and ex. Elements that do
not show up in the match part of the rules are excluded from the typed graph
strict instance.}  \textbf{and} $\langle V_x,E_x\setminus E_{mx},\tau_x\rangle$
is a typed graph strict instance of $\langle V_{pc},E_{pc}\setminus
E_{pc},\tau_{pc}\rangle$.

\end{definition}

Note that in definition~\ref{def:instance_pc_ex} the match associations are naturally excluded from the typed strict instance relation between the apply pattern of the transformation execution and the apply pattern of the path condition. Vertices of match patterns are nonetheless kept. This is so because traceability links need to be considered in the typed graph strict instance relation as well as, together with those traceability links, the vertices of the match patterns.

%\levi{Is this where we want to mention one execution is covered by one/more than one path condition?}

% In figure~\levi{add a figure depicting the surjective graph homomorphism here}
% we depict the relation between a path condition and a transformation execution.

\subsection{Validity and Completeness of the Property Proving algorithm}

This section will use the above definitions in order to prove the validity and completeness of our property proof algorithm presented in section~\ref{sec:verif_dsltrans_props}. %Note that we make the following assumptions:

% \levi{Fix}
% \begin{itemize}
%   \item Edge types are disregarded
%   \item Transformation executions are correctly built given a transformation (as per SLE paper)
%   \item Path conditions are correctly built given a transformation (as described in this paper)
%   \item Mention notational differences with previous papers and why they were introduced
%   \item We do not present the formal definition of transformation here and refer the reader to the SLE paper
% \end{itemize}

The structure of a transformation property is formally defined in definition~\ref{def:trans_prop}. As with most other constructs in our algorithm, it is also formed from a match graph and an apply graph.

\begin{definition}{Property of a Transformation}
\label{def:trans_prop}

Let $t$ be a DSLTrans transformation having source metamodel $s$ and target
metamodel $t$. A property of $t$ is a 7-tuple $\langle V,E\cup
Il,\tau,Match,Apply,Bl,Il\rangle$, where \\$\langle V,E,\tau,Match,Apply\rangle
\in MAP^{s}_{t}$ is a match-apply pattern. $Match=\langle V',E',\tau',s\rangle$,
$Apply=\langle V'',E'',\tau'',t\rangle$, the edges $Tl\subseteq V'\times V''$
are called \emph{backward links} and the edges $Il\subseteq (V'\times V')\cup
(V''\times V'')$ are called \emph{indirect links}. The set of all properties
having source metamodel $s$ and target metamodel $t$ is called
$Property^{s}_{t}$.
\end{definition}

It is important to mention the fact that the property language, as defined
in definition~\ref{def:trans_prop}, is based solely on the source and target metamodels of
a DSLTrans model transformation. This property language is an over-approximation
of the property language that our algorithm practically verifies. This
over-approximation is due to the fact that certain properties that can be
expressed may refer to patterns that are never matched by the transformation. In
one of our previous articles on this topic~\cite{Lucio:10} we referred to these
properties as being \emph{non-provable}.
For our theoretical purposes we assume we only deal with \emph{provable} properties:
properties for which the pre-condition appears in the match pattern of at least
one path condition of the transformation being verified.

Definition~\ref{def:sat_prop_ex} details how a transformation execution is said to satisfy a property. Due to the common structure between properties and transformation executions, this satisfaction is based on whether the property can be isomorphically matched in the transformation execution.

\begin{definition}{Satisfaction of a Property by an Execution of a Transformation}
\label{def:sat_prop_ex}

Let $t$ be a DSLTrans transformation having source metamodel $s$ and target
metamodel $t$. Let also $p = \langle
V_p,E_p,\tau_p,Match_p,Apply_p,Bl_p,Il_p\rangle \in Property^{s}_{t}$ be a
property of $t$ and $ex = \langle V_x,E_x,\tau_x,Match_x,Apply_x,Tl_x\rangle \in
Exec^{s}_{t}$ be an execution of $t$. Execution $ex$ satisfies property $p$,
written $ex \models p$, if and only if:
$$Match_p \vartriangleleft Match^{*}_{x} \implies p \vartriangleleft ex^*$$
More informally, \textbf{if} there exists a typed graph
injective homomorphism between $Match_p$ and $Match^{*}_{x}$ \textbf{then} there
exists a typed graph injective homomorphism between $p$ and $ex^*$.
% \begin{enumerate}
% \item $\langle V_p,E_p\setminus Il_p,\tau_p\rangle$ is a typed subgraph of $\langle V_x, E_x,\tau_x\rangle$
% \item if $v_p\rightarrow v_{p}'\in Il_p$ then there exists $v_x\rightarrow
% v_{x}'\in E_{x}^{*}$ where $\tau(v_p)=\tau(v_x)$, $\tau(v_p')=\tau(v_x')$ and
% $E_{x}^{*}$ is obtained by the transitive closure of $E_{x}$.
% \end{enumerate}
\end{definition}

Note that in definition~\ref{def:sat_prop_ex} we use the typed graph injective
homomorphism relation `$\vartriangleleft$' between patterns, which are
structures that contain more information than typed graphs over which
relation `$\vartriangleleft$' is originally defined. In the text that follows, and with the aim of
simplifying our presentation, we will often make use of such notational abuse.

Definition~\ref{def:sat_prop_pc} is similar to that of property satisfaction for a transformation execution, but for path conditions. Again, the property is isomorphically matched onto the respective graphs in the path condition.

\begin{definition}{Satisfaction of a Property by a Path Condition}
\label{def:sat_prop_pc}

Let $t$ be a transformation, $p = \langle
V_p,E_p,\tau_p,Match_p,Apply_p,Bl,Il_p\rangle \in Property^{s}_{t}$ be a property of $T$ and $pc =
\langle V_{pc},E_{pc},\tau_{pc},Match_{pc},Apply_{pc},Tl,Il_{pc}\rangle \in PC^{s}_{t}$ be a path condition of
$t$. Path condition $pc$ satisfies property $p$, written $pc \vdash p$, if and only if:
 $$Match_p \vartriangleleft Match^{*}_{pc} \implies p \vartriangleleft pc^*$$
More informally, \textbf{if} there exists a typed graph injective homomorphism between $Match_p$ and $Match^{*}_{pc}$
\textbf{then} there exists a typed graph injective homomorphism between $p$ and $pc^*$.

\end{definition}

Given these definitions, we see that as all transformation executions are represented by path conditions, proving properties on the path conditions allows us to conclude property validity on all transformation executions.

\subsubsection{Validity}

In this section we prove the validity of our property proof algorithm by considering what it
formally means when a property is proved on a path condition. We argue that if
the property holds or not on the path condition, then it will accordingly hold
or not on for any transformation executions that the path condition represents.

% As well, we also prove that if a property holds or not on a specific transformation
% execution, it will hold or not on the path condition that represents that
% transformation execution.

\begin{proposition}{(Validity) The result of checking a property on a path condition and of checking the property on all the transformation executions that path condition abstracts will be the same.}
\label{prop:validity}

Let $t$ be a DSLTrans transformation having source metamodel $s$ and target
metamodel $t$. Let also $ex= \langle
V_x,E_x,\tau_x,Match_x,Apply_x,Tl_x\rangle\in Exec^{s}_{t}$ be a transformation
execution of a transformation $t$, $pc = \langle
V_{pc},E_{pc},\tau_{pc},Match_{pc},Apply_{pc},Tl_{pc},Il_{pc}\rangle \in
PC^{s}_{t}$ be a path condition of transformation $t$ and $p = \langle
V_{p},E_{p},\tau_{p},Match_{p},Apply_{p},Bl_{p},Il_{p}\rangle\in
Property^{s}_{t}$ be a property of $t$. This given, we have that:
$$ex\Vvdash pc \land\; pc\vdash p \;\Longleftrightarrow \; ex\models p$$
\end{proposition}

\begin{pf}
The proof of our proposition is divided in two lemmas: lemma~\ref{lemma:validity1}, the $ex\Vvdash pc
\land\; pc\vdash p \;\Longrightarrow \; ex\models p$ direction of the
equivalence; and lemma~\ref{lemma:validity2}, the $ex\models p \;\Longrightarrow\;ex\Vvdash pc\land\;
pc\vdash p$ direction of the equivalence.
\end{pf}

\begin{lemma}{If a property holds for a path condition then the property holds for any transformation execution that path condition abstracts.}
\label{lemma:validity1}
\end{lemma}
\begin{pf}
In order to prove that $ex\Vvdash pc
\land\; pc\vdash p \;\Longrightarrow \; ex\models p$ we have to demonstrate that, if there
exists a typed graph injective homomorphism between $Match_p$ and
$Match^{*}_{x}$ then there exists an typed graph injective homomorphism between
$p$ and $ex^*$. Throughout the proof we will make use of an example inspired by
the police station transformation. The example is given in
figure~\ref{fig:proof_example} and presents the three possible artifacts
involved in the premises of the proposition: a property $p$, a path condition
$pc$ and a transformation execution $ex$. The artifacts respect the relations in
the premise of the proposition: $ex\Vvdash pc$ and $pc\vdash p$. The example
will allow us to graphically portray how the several relations needed to
demonstrate that $ex\models p$ are built. As previously introduced in the paper,
the $p$, $pc$ and $ex$ in figure~\ref{fig:proof_example} are depicted as graphs
where the match (or pre-condition, for properties) part are represented on a
white background and the apply (or post-condition, for properties) are
represented on a grey background. Full arrows represent direct links, while
dashed arrows represent indirect links.

\begin{figure}[h!] \centering \includegraphics[scale=.35]{./figures/property_proving/proof_example.pdf}
	\caption{A Property, a Path Condition and a Transformation Execution for the Police Station Transformation}
	\label{fig:proof_example}
\end{figure}

This proof will be performed in two parts: in part 1) we will start by showing that the
premises of our proposition, $ex\Vvdash pc$ and $pc\vdash p$, imply that there
exists a typed graph injective homomorphism between $Match_p$ and
$Match^{*}_{x}$\;; in part 2) we then need to show that the premises of our proposition
and the fact that there exists a typed graph injective homomorphism between
$Match_p$ and $Match^{*}_{x}$ imply that there exists an typed graph injective
homomorphism between $p$ and $ex^*$.

Regarding part 1), let us show that $ex\Vvdash pc$ and $pc\vdash p$ imply
that there exists a typed graph injective homomorphism between $Match_p$ and
$Match^{*}_{x}$. Given that $pc\vdash p$, by 
definition~\ref{def:sat_prop_pc} we only have to consider the case where there
exists a typed graph injective homomorphism $f$ between $Match_p$ and
$Match^{*}_{pc}$. 
Given $ex\Vvdash pc$, by definition~\ref{def:instance_pc_ex} we only have to
consider the case where a typed graph injective homomorphism $g$ exists between  
$Match_{pc}$ and $Match^*_x$.
Given the above, we can conclude that there exists a typed graph injective
homomorphism between $Match_p$ and $Match^{*}_{x}$ that can be built as the
composition of $g$ and $f$, i.e. $g\circ f$. This part of the proof is
exemplified in figure~\ref{fig:proof_example_morphism_p_to_ex_match}. Note
that in the figure we have represented the transitive closure of $pc$ and $ex$ by including a different
type of dashed arrows between elements that are transitively linked.%\levi{careful about indirect links in $p$ and $pc$}

\begin{figure}[h!] \centering \includegraphics[scale=.35]{./figures/property_proving/proof_example_morphism_p_to_ex.pdf}
	\caption{Building the Injective Homomorphism between the Match part of a Property and a Transformation Execution}
	\label{fig:proof_example_morphism_p_to_ex_match}
\end{figure}

Regarding part 2), let us now show that the existence of a typed graph injective
homomorphism between $Match_p$ and $Match^{*}_{x}$ implies that there exists a
typed graph injective homomorphism between $p$ and $ex^*$. We have that
$pc\vdash p$ and as such by definition~\ref{def:sat_prop_pc} there exists a
typed graph injective homomorphism $k$ between $p$ and $pc^*$. We also have that
$ex\Vvdash pc$, which means that $\langle V_x,E_x\setminus E_{mx},\tau_x\rangle$
is a typed graph strict instance of $\langle V_{pc},E_{pc}\setminus
E_{mpc},\tau_{pc}\rangle$, where $Match_x = \langle V_{mx},E_{mx},\tau_{mx}\rangle$ and $Match_{pc}
= \langle V_{mpc},E_{mpc},\tau_{mpc}\rangle$.

By definition~\ref{def:typed_graph_strict_instance} this means there exists a
surjective graph homomorphism $h$ between $\langle V_x,E_x\setminus
E_{mx},\tau_x\rangle$ and $\langle V_{pc},E_{pc}\setminus
E_{pc},\tau_{pc}\rangle$. Relation $h^{-1}$, the inverse of $h$, necessarily
contains an injective typed graph homomorphism $i\subseteq h^{-1}$, given that $h$ is
type respecting. The relations involved in this part of the proof are
exemplified in figure~\ref{fig:proof_example_morphism_pc_to_ex_apply}. Note
that, since $i$ overlaps with part of $h^{-1}$, in order to make it distinct in
figure~\ref{fig:proof_example_morphism_pc_to_ex_apply}, $i$ is represented using
a thicker dashed line than the other homomorphisms in the picture.

\begin{figure}[h!] \centering \includegraphics[scale=.35]{./figures/property_proving/proof_example_morphism_inverse_surjection.pdf}
	\caption{Building the Injective Homomorphism between the Apply part of a Path Condition and a Transformation Execution}
	\label{fig:proof_example_morphism_pc_to_ex_apply}
\end{figure}
We can thus finally build an injective homomorphism $j = g\cup i$ between $pc^*$ and
$ex^*$. This morphism is exemplified in figure~\ref{fig:proof_example_morphism_pc_to_ex}.

\begin{figure}[h!] \centering \includegraphics[scale=.35]{./figures/property_proving/proof_example_morphism_pc_to_ex.pdf}
	\caption{Building the Injective Homomorphism between a Path Condition and a Transformation Execution}
	\label{fig:proof_example_morphism_pc_to_ex}
\end{figure}

From the above we can conclude that there exists a typed graph injective
homomorphism between $p$ and $ex^*$ that can be built as $j\circ k$, the
composition of injective typed graph isomorphisms $k$ and $j$. This homomorphism is exemplified in figure~\ref{fig:proof_example_morphism_composition}.

\begin{figure}[h!] \centering \includegraphics[scale=.35]{./figures/property_proving/proof_example_morphism_composition.pdf}
	\caption{Building the Injective Homomorphism between a Property and a Transformation Execution}
	\label{fig:proof_example_morphism_composition}
\end{figure}
\end{pf}

\begin{lemma}{If a property holds for a transformation execution then the property holds for the path condition that abstracts it.}
\label{lemma:validity2}
\end{lemma}
\begin{pf}
We will now prove the reverse direction of the implication, i.e. that:
$$ex\models p \;\Longrightarrow\;ex\Vvdash pc\land\; pc\vdash p$$
By using contraposition we can show that:
$$\neg(ex\Vvdash pc\land pc\vdash p) \;\Longrightarrow\; \neg(ex\models p)$$

We only need to worry about the case where we have  $ex\Vvdash pc\land
\neg(pc\vdash p)$ given our property proof results are meaningful only for
transformation executions which are abstracted by the path condition being
examined. In this case we have that, because $ex\Vvdash pc$, by
definition~\ref{def:instance_pc_ex} we know that there is a surjective typed
graph homomorphism between between the apply pattern of transformation execution
$ex$ and the apply pattern of the path condition $pc$. Given we know that
$\neg(pc\vdash p)$, then we know that an injective type graph homomorphism does
not exist between $p$ and $pc$. This means that, by the
definition~\ref{def:typed_graph_homomorphism} of injective typed graph
homomorphism, there exists a type $T$ which is instantiated in $pc$ but not in
$prop$. However, because of the fact that $ex\Vvdash pc$ we know that $T$ is
instantiated in $ex$ because a surjective typed graph homomorphism exists
between $ex$ and $pc$, implying type $T$ is instantiated at least once in $ex$.
We thus can deduce that we have $\neg(ex\models p)$ given that $ex\models p$
implies that there exists an injective type graph homomorphism between property
$p$ and execution $ex$. This injective type graph homomorphism cannot exist given the fact an instance
of $T$ exists in $ex$ but not in $p$.
\end{pf}

\subsubsection{Completeness}

\begin{proposition}{(Completeness) Proving that a property holds or does not
hold by examining all path conditions of the transformation is equivalent to
proving the property holds or does not hold for all transformation executions.}
\label{prop:PP_completeness}
\end{proposition}
\begin{ps}
This result follows from the combination of the validity and completeness of the
path condition algorithm, as seen in propositions~\ref{prop:pc_validity}
and~\ref{prop:completeness}, as well as the validity of the property proving
step on each path condition, as seen in proposition~\ref{prop:validity}.
\end{ps}

% \section{T-Core}
% \label{sec:t-core}
% In order to build our tool we have chosen the T-Core framework introduced by Syriani and Vangheluwe in~\cite{DBLP:journals/eceasst/SyrianiV10}. T-Core is a set of primitive model transformation blocks that can be used to replicate the behavior of existing transformation languages (e.g. in order to compare their expressiveness and provide a framework for interoperability), or to build new model transformation languages.
% 
% The framework includes five main primitive transformation blocks that exchange models and transformation information in messages called \emph{packets}. Those blocks are:
% 
% \begin{itemize}
% \item the \emph{Matcher}, that finds matches of a given pre-condition pattern within a model by running an efficient combination of the Ullmann's and VF2's subgraph isomorphism algorithm and collects those matches in a \emph{packet}
% 
% \item the \emph{Iterator} which allows selecting the next matched submodel from the set of matches gathered in a \emph{packet}
% 
% \item the \emph{Rewriter} consumes a matched subgraph from a model in a packet and changes the model according to a given post-condition pattern
% 
% \item the \emph{Rollbacker} which allows checkpointing and restoring \emph{packets} such that backtracking can be achieved
% 
% \item the \emph{Resolver} for solving potential conflicts
% between matches and rewritings
% \end{itemize}
% 
% 
% An additional T-Core construct is called the \emph{composer}, and is used to encapsulate compositions of the five primitive transformation blocks described above. The goal of the encapsulation mechanism is that complex transformation blocks, such as \emph{querying},\emph{rewriting one random match} or \emph{rewriting all matches found}, can be seamlessly created from the simplest transformation blocks.
% 
% Note that the transformation primitives described above execute transformations on models which are metamodel instances. Matching precondition patterns and rewriting postcondition patterns can only occur if the pre- and post-condition patterns are generated from the same metamodel as the models being treated. This is in order to insure coherence.

### 2015-03-13 text/appendix_DSLTrans.tex
previous version of the paper

\section{Formal Syntax and Semantics of DSLTrans}
\label{sec:DSLTrans_formal_appendix}

\counterwithin{definition}{section}
\setcounter{definition}{0}
\setcounter{subsection}{0}

\onehalfspacing

In this section we will formally introduce the syntax and the semantics of the DSLTrans language. The theory is based on the notion of typed graphs as described in \cref{sec:formal_background}.

We will start by introducing the notion of \emph{metamodel}, which in DSLTrans is used to type the input and output models of a DSLTrans transformation.

\begin{definition}{Metamodel}
\label{def:metamodel_appendix}

\CatchFileBetweenTags{\metamodel}{text/definitions}{metamodel}{\metamodel}

\end{definition}

\CatchFileBetweenTags{\metamodeltextappendix}{text/definitions}{metamodeltextappendix}{\metamodeltextappendix}

\begin{definition}{Expanded Metamodel}
\label{def:expanded_metamodel_appendix}

\CatchFileBetweenTags{\expandedmetamodel}{text/definitions}{expandedmetamodel}{\expandedmetamodel}

\end{definition}

\CatchFileBetweenTags{\expandedmetamodeltext}{text/definitions}{expandedmetamodeltext}{\expandedmetamodeltext}

\begin{definition}{Metamodel Instance}
\label{def:instance_appendix}

\CatchFileBetweenTags{\metamodelinstance}{text/definitions}{metamodelinstance}{\metamodelinstance}

\end{definition}

\CatchFileBetweenTags{\metamodelinstancetext}{text/definitions}{metamodelinstancetext}{\metamodelinstancetext}

\begin{definition}{Containment Transitive Closure}
\label{def:instance_closure_appendix}

\CatchFileBetweenTags{\instanceclosure}{text/definitions}{instanceclosure}{\instanceclosure}

\end{definition}

\CatchFileBetweenTags{\instanceclosuretextappendix}{text/definitions}{instanceclosuretextappendix}{\instanceclosuretextappendix}

\begin{definition}{Model}
\label{def:model_appendix}

\CatchFileBetweenTags{\model}{text/definitions}{model}{\model}

\end{definition}

\CatchFileBetweenTags{\modeltextappendix}{text/definitions}{modeltextappendix}{\modeltextappendix}

\begin{definition}{Input-Output Model\\}
\label{def:input_output_model_appendix} 
\CatchFileBetweenTags{\inputoutputmodel}{text/definitions}{inputoutputmodel}{\inputoutputmodel}

\end{definition}

\CatchFileBetweenTags{\inputoutputmodeltextappendix}{text/definitions}{inputoutputmodeltextappendix}{\inputoutputmodeltextappendix}

\begin{definition}{Metamodel Pattern and Indirect Metamodel Pattern}
\label{def:metamodel_pattern_appendix} 
\CatchFileBetweenTags{\metamodelpattern}{text/definitions}{metamodelpattern}{\metamodelpattern}

\end{definition}

\CatchFileBetweenTags{\metamodelpatterntextappendix}{text/definitions}{metamodelpatterntextappendix}{\metamodelpatterntextappendix}

\begin{definition}{Transformation Rule}
\label{def:transformation_rule_appendix}

\CatchFileBetweenTags{\transformationrule}{text/definitions}{transformationrule}{\transformationrule}
\end{definition}

\CatchFileBetweenTags{\transformationruletext}{text/definitions}{transformationruletext}{\transformationruletext}

\begin{definition}{Matcher of a Transformation Rule}
\label{def:back_match_transformation_rule_appendix}

 \CatchFileBetweenTags{\backmatchtransformationrule}{text/definitions}{backmatchtransformationrule}{\backmatchtransformationrule}
\end{definition}

\cref{def:back_match_transformation_rule_appendix} \CatchFileBetweenTags{\backmatchtransformationruletext}{text/definitions}{backmatchtransformationruletext}{\backmatchtransformationruletext}

\begin{definition}{Expanded Transformation Rule}
\label{def:transformation_rule_expansion_appendix}

\CatchFileBetweenTags{\expandedtransformationrule}{text/definitions}{expandedtransformationrule}{\expandedtransformationrule}

\end{definition}
\CatchFileBetweenTags{\expandedtransformationruletext}{text/definitions}{expandedtransformationruletext}{\expandedtransformationruletext}

\begin{definition}{Layer, Transformation}
\label{def:layer_transformation_appendix}

\CatchFileBetweenTags{\layertransformation}{text/definitions}{layertransformation}{\layertransformation}
\end{definition}

\CatchFileBetweenTags{\layertransformationtextappendix}{text/definitions}{layertransformationtextappendix}{\layertransformationtextappendix}

\subsubsection{\textbf{Notation:}}

\CatchFileBetweenTags{\notationtextappendix}{text/definitions}{notationtextappendix}{\notationtextappendix}

% We naturally extend the notion of union (definition~\ref{def:typed_graph_union})
% to models (definition~\ref{def:metamodelmodel}), match-apply models
% (definition~\ref{def:match_apply_model}) and transformation rules 
% (definition~\ref{def:transformation_rule}). We also extend the notion of
% indirect typed subgraph (definition~\ref{def:indirecttypedsubgraph}) to
% transformation rules (definition~\ref{def:transformation_rule}) and match-apply
% models (definition~\ref{def:match_apply_model}). Finally, we extend the notion
% of typed graph equivalence (definition~\ref{def:typed_graph_equivalence}) to
% transformation rules (definition~\ref{def:transformation_rule}).

\subsubsection{Transformation Language Semantics}

We will now address the semantics of the DSLTrans language. We will start by defining a match function that, given an input-output model and a transformation rule, returns all subgraphs of that input-output model where the rule's match pattern (including backward links) is found.

% \levi{Need to be careful throughout all this section to understand whether we model the match and apply functions as functions or relations. Because of the non-deterministic generation of new vertices in the definition of apply function~\ref{def:apply_function}, relations are used in the definitions that follow to evaluate a transformation. This is an issue of the formalism used to model, should we make it explicit?}

\begin{definition}{Match Function}
\label{def:match_function}

Let $m_{in}\in \textsc{Iom}^{sr}_{tr}$ be a input-output model and $rl\in \textsc{Rule}^{sr}_{tg}$ be a transformation
rule. The $match : \textsc{Iom}^{sr}_{tg}\times \textsc{Rule}^{sr}_{tg}\rightarrow
\mathcal{P}(\textsc{Iom}^{sr}_{tg})$ function is defined as follows: $$match_{rl}(m_{in})= \big\{ glue_{noind}\;|\;
glue\sqsubseteq m_{in}^{*} \land glue \cong \Vert rl\Vert\big\}$$

where $glue\in \textsc{Iom}^{sr}_{tr}$ is an input-output model and $glue_{noind}$ is a version of $glue$ where the indirect links have been removed.
% \levi{all backward links are replaced by trace links such that the matching using subgraph isomorphim can work.}
% 
% \levi{Where $rl^{-}$ is rule $rl$ without nodes in the apply part that are not connected to \emph{trace links}, as well as links adjacent to those nodes. $mg^{-ind}$ is the match graph without indirect links that come from the rule, given those cannot be part of the intermediate $MAM$ match result since that result is going to be united with the original $MAM$ (this was incorrect in the original SLE paper).}
% \levi{The matcher function returns all graphs of the input MAM that are matched by a rule, including its backward links. Note that to find a subgraph ($\sqsubseteq$) of the input MAM that is isomorphic ($\cong$) to the rule being matched because the graph rewriting is achieved by joint union of the match graph extended by the new rewrite node, and the original MAM.}
% \levi{TODO: review the $mg^{-ind}$ and the $rl^{-}$ notations.}

% Due to the fact that the $\cong$ relation is based on the notion of graph
% isomorphism, permutations of the same match result may exist in the
% $\big\{g\;|\; g\lhd m \land g \cong strip(tr)\big\}$ set. The --- undefined ---
% $remove:\mathcal{P}(TR^{s}_{t})\rightarrow \mathcal{P}(TR^{s}_{t})$ function is such that
% it removes such undesired permutations.

%The $strip:TR^{s}_{t}\rightarrow TR^{s}_{t}$ function is such that
%$$strip(\big\langle V,E,T,\langle V_m,E_m,T_m,s\rangle,\langle V_a,E_a,T_a,t\rangle,Bl,Il\big\rangle) = \big\langle V',E',T,\langle V_m,E_m,T_m\rangle,\langle V'_a,E'_a,T_a\rangle,Bl,Il\big\rangle$$
%\begin{center}
%where $(v'\in V'_a\Rightarrow v\rightarrow v'\in Bl)\;\land\;(v\rightarrow v'\in E'_a\Rightarrow (v\rightarrow v'\in E_m \land \{v,v'\}\subseteq V'_a))$
%\end{center}
\end{definition}

The match function in \cref{def:match_function} looks for subgraphs ($glue\sqsubseteq m_{in}^{*}$) of an input-output model that are isomorphic to the backward matcher of the given transformation rule ($glue \cong \Vert rl\Vert$). Note that the containment transitive closure of the input-output model ($m_{in}^{*}$) is considered such that indirect links in the rule can looked for in the input model. Additionally, indirect links need to be removed from the input-output models resulting from the match function ($glue_{noind}$). This is so because indirect links are not part of the original input model, but rather an auxiliary structure.

Let us now turn our attention to the apply function in \cref{def:apply_function}. Its role is extend all model fragments found by matching a rule on a given input-output model, such that each of those fragments becomes isomorphic to the complete rule (minus its backward links). This process effectively creates the new objects and relations specified in the apply part of the rule, for each of the fragments found when matching the rule.

\begin{definition}{Apply Function}
\label{def:apply_function}

Let $m_{glue}\in \textsc{Iom}^{sr}_{tg}$ be a input-output model and $rl\in \textsc{Rule}^{sr}_{tg}$ a
transformation rule. The $apply : \textsc{Iom}^{sr}_{tg}\times \textsc{Rule}^{sr}_{tg}\rightarrow \textsc{Iom}^{sr}_{tg}$ function
is defined as follows:

$$apply_{rl}(m_{in})=\bigsqcup_{m_{glue}\in match_{rl}(m_{in})}trace_{a_{\Delta}}(m_{glue}\sqcup a_{\Delta})$$

\begin{center}
where $a_{\Delta}\in \textsc{Iom}^{sr}_{tg}$ is such that $m_{glue} \sqcup a_{\Delta}\cong rl_{noind}$.\\
\end{center}

We impose that any instance of $a_\Delta$ is always disjoint from the $m_{in}$ input-output model and also that any two instances of $a_\Delta$ used in the large union are always disjoint.

Partial function $trace:\textsc{Iom}\times \textsc{Iom} \nrightarrow \textsc{Iom}$ is such that $trace_{\langle V_{\Delta},E_{\Delta},st_{\Delta},\tau_{\Delta}\rangle}(\langle V,E,st,\tau\rangle) = \langle V,E',st',\tau'\rangle$ where we have that $E\subseteq E'$, $st\subseteq st'$ (using a light notational abuse for the $s\subseteq s'$ and $t\subseteq t'$), $\tau \subseteq \tau'$ and if $v_1\xrightarrow{e} v_2\in E'\setminus E$ then $v_1\in Output(V_{\Delta})$, $v_2\in Input(V)$ and $\tau'(e)=trace$. Finally, $rl_{noind}$ is a version of $rl$ where indirect links have been removed.
% More concretely, if we have $mam = $\big\langle V,E,st,\tau, Input,Output\big\rangle\in MAM^{sr}_{tg}$ where $Input=\langle V',E',st',\tau'\rangle\in MODEL^{sr}$ and $Output=\langle
% V'',E'',st'',\tau''\rangle\in MODEL^{tg}$, then $trace(mam) = \big\langle V,E,st,\tau''',Input,Output\big\rangle$ where $E'\subseteq E$ and $v\rightarrow E'\setminus E$ 
\end{definition}
% 
% it builds traceability edges between vertices of the output model that were newly built by the $apply$ function (the vertices in ${mg}_{\Delta}$) and all the vertices from the input part of the input-output model. It is necessary to remove the indirect links from the rule because otherwise they would be created during the rewriting by the apply function.

In \cref{def:apply_function} $a_\Delta$ is an input-output model that contains an instance of the target metamodel. These instances are created by rule $rl$ and are used to extend the sub-models found by the match function. The $trace$ function builds traceability edges between newly created vertices of the output model in $a_{\Delta}$ and all the vertices from the input part of a model fragment found by the match function.

Note that, because we do not pose any constraints on $a_\Delta$ other than the fact that its union with the sub-model $m$ is isomorphic to $rl_{noind}$, the $a_\Delta$ variable can always be satisfied by an unlimited amount of input-output models. In order to avoid an infinite amounts of results when a transformation rule is executed, in what follows we will consider transformation results differ only up to typed graph isomorphism.

%  Note that because the vertices are not disjoint, when the graph produced by the apply function is united with the original match apply model the new vertices are created. It was necessary to remove the indirect links from the rule during rewriting, because otherwise those indirect links would also be rewritten.

Definitions~\ref{def:match_function} and~\ref{def:apply_function} are complementary: the former gathers all the fragments of an input-output model that are matched by a transformation rule; the latter glues on the output part of each of those fragments new objects and relations created by a transformation rule.

%  The $strip$ function is used to
%  enable matching over backward links but not elements to be created by the
%  transformation rule. The $back$ function connects all newly created vertices
%  to the elements of the source model that originated them.

%\begin{definition}{Transformation}

%Let $g\in TG$ be a typed graph. The graph transform function $transform : TR\times TG\rightarrow TG$ is recursively defined as:
%\begin{gather*}
%  transform_{(m,a)}(g) =
%  \begin{cases}
%    g  & \text{if } match_{m}(g)=\emptyset \\
%    instance(a)\cup transform_{(m,a)}(g') &\text{if } match_{m}(g)\neq \emptyset
%  \end{cases}
%\end{gather*}
%$$\text{where } g'=mark_{(m)}(g)$$
%\end{definition}

%\subsection{Transformation Semantics}
%\begin{definition} {Layer Semantics}

%Let $l\in Layer$ be a layer and $\{m,m'\} \in MAM^{s}_{t}$ be a match
%$$\frac{}
%{m,\emptyset \stackrel{layerstep}{\rightarrow}m}$$
%$$\frac{tr\in layer,\; transform_{tr}(m) = m''\; m'',layer\backslash\left\{tr\right\} \xrightarrow{layerstep}	 m'}{m,layer \xrightarrow{layerstep} m'}$$
%\end{definition}

\begin{definition} {Layer Step Semantics}
\label{def:layer_step_semantics}

Let $l\in \textsc{Layer}^{sr}_{tg}$ be a Layer. The \emph{layer step relation}
$\stackrel{layerstep}{\rightarrow}\subseteq \textsc{Iom}^{sr}_{tg}\times \textsc{Iom}^{sr}_{tg} \times \textsc{Layer}^{sr}_{tg}\times
\textsc{Iom}^{sr}_{tg}$ is defined as follows:

$$\frac{}
{\langle m_{in},m_{glue},\emptyset\rangle \xrightarrow{layerstep}m_{in} \sqcup m_{glue}}$$

\begin{multline*}
\frac{\begin{array}{ll} rl\in l,\; apply_{rl}&(m_{in}) = m_{rout},\\
&\langle m_{in},m_{glue}\sqcup m_{rout},l\backslash \{rl\}\rangle \xrightarrow{layerstep} m_{out}
\end{array}}
{\langle m_{in},m_{glue},l\rangle \xrightarrow{layerstep} m_{out}}$$
\end{multline*}

\begin{center}
where $m_{rout} \in\textsc{Iom}^{sr}_{tg}$ and $rl\in \textsc{Rule}^{sr}_{tg}$.
\end{center}

We impose that all input-output models that are part of $rout$ and have been generated by rule $rl$ are disjoint from input-output models accumulated in $m_{glue}$ that have been generated by other rules.

\end{definition}

In \cref{def:layer_step_semantics} we build the result of executing a layer of a DSLTrans transformation. The operational semantics-like rules in the definition execute each rule $rl$ in layer $l$, in a non-deterministic order, by using the $apply$ function. The result of executing each rule is accumulated in the temporary $m_{glue}$ input-output model. Finally, when the set of transformation rules in the layer has been exhausted, the result of executing all the rules in the layer (now contained in $m_{glue}$) is united with the input-output model $m_{in}$, the input to layer $l$. Note that this final union produces the result we expect because of the fact that the $m_{glue}$ input-output model is not disjoint from $m_{in}$. The common ``glue'' parts of $m_{glue}$ that have been built by the match function and extended by the apply function are now used to built the result of executing layer $l$.

\cref{def:layer_step_semantics} is the core of DSLTrans' semantics. Many model transformation languages are based on graph rewriting, where the result of each rule rewrite is immediately usable by all other rules. In DSLTrans the result of executing one layer in DSLTrans is totally produced before the input to the layer is changed. This is enforced in \cref{def:layer_step_semantics} by the fact that the apply function always executes over the same $m_{in}$ input-output model and all the results of rule execution in the same layer are added to the $m_{glue}$ structure that is write-only. Rules belonging to the same layer are thus forced to execute independently.

\begin{definition} {Transformation Step Semantics}
\label{def:transformation_step_semantics}

Let $[l::tr]\in \textsc{Transf}^{sr}_{tg}$ be a transformation, where $l\in
\textsc{Layer}^{sr}_{tg}$ is a Layer and $tr$ also a transformation. The \emph{transformation step relation}
$\stackrel{trstep}{\rightarrow}\subseteq \textsc{Iom}^{sr}_{tg}\times \textsc{Transf}^{sr}_{tg}\times
\textsc{Iom}^{sr}_{tg}$ is defined as follows: 
$$\frac{}{\langle m,[]\rangle \xrightarrow{trstep} m}$$

$$\frac{\big\langle m_{in},\epsilon,l^{\star}\big\rangle \xrightarrow{layerstep} m_{inter},\; \langle m_{inter},R\rangle \xrightarrow{trstep} m_{out}}{\langle m_{in},[l::T]\rangle \xrightarrow{trstep} m_{out}} \hspace{.3cm}$$

$$\text{where } l^{\star}=\bigcup_{rl\in l}rl^{\star}$$
\end{definition}

While the execution of the rules belonging to a layer happens in parallel, the execution of the layers of a transformation happens sequentially. As per \cref{def:transformation_step_semantics}, the input-output model $m_{inter}$ is the output of executing a given layer that is passed onto the next layer as input. Note that an empty input-output model ($\epsilon$) is passed as the second argument to the $layerstep$ relation in \cref{def:transformation_step_semantics}. This is because in \cref{def:layer_step_semantics} of layer step semantics, the second argument of the relation is used as an accumulator for the model fragments that are added to be added to the output of the previous layer once all the rules in the current layer have been executed. The transformation rules in a layer are expanded before execution ($l^{\star}$) such that polymorphism in the match elements can be handled (see \cref{def:transformation_rule_expansion_appendix}).

\begin{definition} {Model Transformation Execution}
\label{def:modeltransformation_appendix}

\CatchFileBetweenTags{\modeltransformation}{text/definitions}{modeltransformationappendix}{\modeltransformation}
\end{definition}

\CatchFileBetweenTags{\modeltransformationtextappendix}{text/definitions}{modeltransformationtextappendix}{\modeltransformationtextappendix}

% A transformation execution is formed from taking the input model, and executing
% the transformation on it to produce the output model. During this
% transformation, traceability links will be placed between match elements and the
% apply elements they create. Definition~\ref{def:transf_ex} expresses this
% formally. Note that we assume that transformation executions are built following
% the algorithm described in~\cite{DBLP:conf/sle/BarrocaLAFS10}.

% \begin{definition}{Transformation Execution}
% \label{def:transf_ex}
% 
% Let $t$ be a DSLTrans transformation having source metamodel $s$ and target
% metamodel $t$. A Transformation Execution is a 6-tuple $\langle V,E\cup
% Tl,\tau,Match,Apply,Tl\rangle$, where $\langle V,E,\tau,Match,Apply\rangle \in
% MAP^{s}_{t}$ is a match-apply pattern. $Match=\langle V',E', \tau',s\rangle$,
% $Apply=\langle V'',E'', \tau'',t\rangle$ and the edges $Tl\subseteq V'\times
% V''$ are called \emph{traceability links}. The set of all transformation
% executions having source metamodel $s$ and target metamodel $t$ is called
% $Exec^{s}_{t}$.
% \end{definition}

We now prove two important properties about executions of transformations expressed in the subset of DSLTrans presented in this paper: \emph{confluence} and \emph{termination}. The proofs are provided at a high level, given the fact that DSLTrans essentially enforces both these properties by construction of the semantics of DSLTrans.

\begin{proposition}{Confluence}

Every model transformation execution is confluent up to typed graph isomorphism.
\end{proposition}
\begin{pf}
We want to prove that for every model transformation execution of a transformation $tr\in \textsc{Transf}^{sr}_{tg}$ having as input a model $input \in MODEL^{s}$, its output is always the same up to typed graph isomorphism.\\
If we assume an execution of the transformation is not confluent then this should happen because of non-determinism when the execution of a transformation is being built. Non-determinism happens during the construction of a transformation execution at two points: 
\begin{enumerate}
\item in definition~\ref{def:apply_function}, $a_{\Delta}$ is non-deterministic up to typed graph isomorphism, which does not contradict the proposition we are trying to prove.
\item in definition~\ref{def:layer_step_semantics} transformation rule $rl$ is chosen non-deterministically from layer $l$. Thus, the order in which the transformation rules are treated is non-deterministic. However, by \cref{def:layer_step_semantics} rules in a layer execute independently, which means no-side effects from rule ordering influence the execution of rules in the same layer. Also, the increments to the transformation by each rule of a layer in \cref{def:layer_step_semantics} are united using $\sqcup$ (see \cref{def:typed_graph_union_appendix}), which is an operation that is commutative by construction and thus renders the transformation result from each layer deterministic. 
\end{enumerate}
Given there are no other sources of non-determinism when building the execution of a transformation, every model transformation execution is confluent up to typed graph isomorphism.
\end{pf}

\begin{proposition}{Termination}

Every model transformation execution terminates.
\end{proposition}
\begin{pf}
Let us assume that there is a transformation execution which does not terminate. In order for this to happen there must exist a part in the construction of the execution of a transformation which induces an algorithm with an infinite amount of steps. We identify three moments when this can happen:
\begin{enumerate}
\item if definition~\ref{def:layer_step_semantics} of execution of a layer induces an infinite amount of steps. The only possibility for this to happen is if a layer has an infinite amount of transformation rules, which is a contradiction with definition~\ref{def:layer_transformation_appendix}.
\item if definition~\ref{def:transformation_step_semantics} of execution of a transformation induces an infinite amount of steps. Given layers are executed sequentially and no looping is allowed, the only possibility for this to happen is if the transformation has an infinite amount of layers, which contradicts definition~\ref{def:layer_transformation_appendix}.
\item if the result of the $match_{rl}(m_{in})$ function in definition~\ref{def:match_function} is an infinite set of match-apply graphs. The input-output model $m_{in}$ is by definition finite and the matching of each rule is independent from the execution of other rules in the same layer. As such, the number of subgraphs of $m_{in}$ isomorphic to $rl$'s matcher found during the execution of $rl$ is finite.
\end{enumerate}
Given there are no other constructs in the semantics of a transformation that can induce an infinite amount of steps, every model transformation execution terminates.
\end{pf}

\clearpage

### 2015-03-13 text/appendix_property_proof.tex
previous version of the paper

\section{Validity and Completeness of Property Verification}
\label{sec:val_complete_prop_verif}

\counterwithin{proposition}{section}
\counterwithin{lemma}{section}
\counterwithin{equation}{section}
\setcounter{proposition}{0}
\setcounter{lemma}{0}
\setcounter{equation}{0}

\onehalfspacing

\begin{definition}{Property of a Transformation\\}
\label{def:trans_prop_appendix}
\CatchFileBetweenTags{\propertydef}{text/definitions}{propertydef}{\propertydef}
\end{definition}

In \cref{def:trans_prop_appendix}, pre-conditions use the same pattern language as the match graph in
DSLTrans rules, allowing the possibility of including several instances of
the same metamodel element as well as indirect links in the property. Indirect links in properties have
the same meaning as in the rule match graph -- they
involve patterns over the transitive closure of containment links in pre-condition graphs.

Post-conditions also use the same pattern language as the
apply graphs of DSLTrans transformation rules, with the additional
possibility of expressing indirect links in post-conditions. Traceability links can
also be used in properties to impose traceability relations between pre-condition
and post-condition elements.

Note that \cref{def:trans_prop_appendix} includes a condition stating a surjective typed graph homomorphism needs to exist between the match part of at least one of the transformation's path condition, and the pre-condition of the property of interest. This condition makes sure that the property's pre-condition can be found at least in one execution of the transformation abstracted (the mathematical argument for this fact is given in the proof of \cref{prop:proof_validity}). This condition makes the checking the validity of a property in the transformation meaningful. If this condition would not be true then it could be that the input pattern required by the property would never be fully matched during transformation execution, making such a property not relevant\footnote{In~\cite{DBLP:conf/sle/BarrocaLAFS10} we have referred to these properties \emph{non-provable}. In the work presented here we explicitely disallow the construction of this class of properties.} for the transformation at hand.

\begin{definition}{Satisfaction of a Property by an Execution of a Transformation\\}
\label{def:sat_prop_ex_appendix}
\CatchFileBetweenTags{\satpropexe}{text/definitions}{satpropexe}{\satpropexe}
\end{definition}

\Cref{def:sat_prop_ex_appendix} states that, every time a graph that is isomorphic to the property's pre-condition is found in (the containment transitive closure of) the input model of the transformation's execution, a graph that is isomorphic to the complete property needs to be found in (the containment transitive closure of) the transformation execution. Note that the last part of the proposition in \cref{def:sat_prop_ex_appendix} ensures that the graph that is isomorphic to the property's pre-condition and the graph that is isomorphic to the complete property overlap on their pre-condition parts.

% resulting from the input pattern of $ex$ identified as pre-condition by $f$. Generally speaking, whenever the property's pre-condition is isomorphically found as a pattern of the input model of a execution, the property's post-condition must also be isomorphically found in the output generated from that input pattern. Note that, because a property may contain traceability links between the pre- and post-condition elements, it is necessary to check that those traceability links are also present in the execution. Because of this, the $g$ homomorphism in definition~\ref{def:sat_prop_ex} that checks the property's post-condition is such that it considers the same match pattern identified by the $f$ homomorphism when the pre-condition is checked.

\begin{definition}{Satisfaction of a Property by a Path Condition\\}
\label{def:sat_prop_pc_appendix}
\CatchFileBetweenTags{\satproppc}{text/definitions}{satproppc}{\satproppc}
\end{definition}

The principle behind the satisfaction relation in \cref{def:sat_prop_pc_appendix} is the same as the one behind the satisfaction relation between a property and an execution of a transformation in \cref{def:sat_prop_ex_appendix}: whenever the property's pre-condition is found in the path condition then so is the complete property. Also, those two graphs found in the path condition share the property's pre-condition part. This last condition enforces that the pre- and post-conditions of the property are correctly linked by traceability link in the path condition.

\begin{proposition}{(Validity) The result of proving a property on a set of path conditions generated for a transformation or an all executions of that transformation is the same.\\}
\label{prop:proof_validity_appendix}
Let $tr\in \textsc{Transf}^{sr}_{tg}$ be a transformation and $p \in \textsc{Property}(tr)$ be a
property of $tr$. This given, we have that transformation $tr$ satisfies property $p$ if and only if:
\begin{equation}
\label{eq:prop_proof_transf_appendix}
\bigwedge_{pc\in \textsc{Pathcond}(tr)} pc\vdash p \hspace{.3cm} \Longleftrightarrow \hspace{.3cm} \bigwedge_{ex\in \textsc{Exec}(tr)}ex\models p
\end{equation}
\end{proposition}
\begin{pf}
In order to prove the proposition in \cref{eq:prop_proof_transf_appendix} we will start by demonstrating that,
if property $p$ holds on a path condition $pc$ generated for $tr$,
then $p$ will necessarily hold on any execution $ex$ of $tr$ that is abstracted by $pc$. On the other hand, if $p$ does not hold on $pc$ then it will not hold for at least of one execution $ex$ of $tr$ abstracted by $pc$. This lemma can be stated as follows:
\begin{equation}
\label{eq:prop_proof_pc_appendix}
pc\vdash p \;\Longleftrightarrow \;\forall ex\in\{ex\in\textsc{Exec}(tr)\;|\;ex\Vvdash pc\}\;.\; ex\models p
\end{equation}
We thus need to demonstrate both directions of the equivalence in \cref{eq:prop_proof_pc_appendix}. On the one hand we need to prove of the left-to-right direction of the equivalence:
\begin{multline}
\label{eq:prop_proof_left_right_appendix}
pc\vdash p \;\Longrightarrow \;\forall ex\in\{ex\in\textsc{Exec}(tr)\;|\;ex\Vvdash pc\}\;.\; ex\models p
\end{multline}
Proposition~\ref{eq:prop_proof_left_right_appendix} is shown to be true in \cref{lemma:validity1_appendix}.
We then need to show the right-to-left direction of the equivalence:
\begin{multline}
\label{eq:prop_proof_right_left_appendix}
\forall ex\in\{ex\in\textsc{Exec}(tr)\;|\;ex\Vvdash pc\}\;.\; ex\models p \;\Longrightarrow \;pc\vdash p 
\end{multline}
Proposition~\ref{eq:prop_proof_right_left_appendix} is shown to be true in \cref{lemma:validity2_appendix}.
Once propositions~\ref{eq:prop_proof_left_right_appendix} and~\ref{eq:prop_proof_right_left_appendix} are proved, we know that all path conditions on which a property holds represent executions on which the property also holds. Thus, if the property holds on all path conditions then it necessarily holds on all executions. On the other hand, if a property does not hold on one path condition, making it such that the conjunction on the left side of the equivalence in \cref{eq:prop_proof_transf_appendix} is false, then according to \cref{eq:prop_proof_pc_appendix} an execution for which it also does not hold exists. This makes it such that the conjunction on the right side of the equivalence in \cref{eq:prop_proof_pc_appendix} is also false.
\end{pf}

\begin{lemma}{If a property holds for a path condition then the property holds for any transformation execution that path condition abstracts.\\}
\label{lemma:validity1_appendix}
Let $tr$ be a transformation, $pc\in \textsc{Pathcond}(tr)$ be a path condition of $tr$, $ex\in \textsc{Exec}(tr)$ be an execution of $tr$ and $p\in \textsc{Prop}(tr)$ be a property of $tr$. Then we have that:
\begin{equation}
\label{prop:validity1_appendix}
pc\vdash p \;\Longrightarrow \;\forall ex\in\{ex\in\textsc{Exec}(tr)\;|\;ex\Vvdash pc\}\;.\; ex\models p
\end{equation}
\end{lemma}
\begin{pf}
By \cref{def:sat_prop_pc_appendix} we know that $pc\vdash p$ is equivalent to proposition $\forall f\; \exists g\,.\,\big(in\stackrel{f}{\blacktriangleleft} Pre \implies out \stackrel{g}\blacktriangleleft p\big)$, where $Pre$ is $p$'s pre-condition, $in$ is a subgraph of the containment transitive closure of the match part of $pc$, and $out$ is a subgraph of the containment transitive closure of $pc$. Additionally, by \cref{def:sat_prop_ex_appendix} we also know that $ex\models p$ is equivalent $\forall f\; \exists g\,.\,\big(Pre \stackrel{f}{\vartriangleleft} Input^{*} \implies p \stackrel{g}{\vartriangleleft} ex^*\big)$, where $Input$ is the input part of $ex$.

We will show that the implication holds by analysing the three cases where, $pc\vdash p$, the left side of Proposition~\ref{prop:validity1_appendix} holds.  
\begin{enumerate}
  \item If the precondition of the property cannot be found in the match part of a path condition $pc$, then it cannot be found in the input part of an execution abstracted by $pc$. Formally, we have that, assuming $ex$ is abstracted by $pc$:
\begin{equation}
\label{item:nopre}
\nexists f\,.\,\big(in\stackrel{f}{\blacktriangleleft} Pre\big) \implies \nexists f'\,.\, Pre \stackrel{f'}{\vartriangleleft} Input^{*}
\end{equation}
where, as before, $Input^{*}$ is the containment transitive closure of of the input part of $ex$ and $in$ is a subgraph of the match part of $pc$. Proposition~\ref{item:nopre} holds because of the fact that the surjection between $in$ and $Pre$ is defined such that it is in fact a set of injective typed graph homomorphisms between subgraphs of $in$ belonging to different rule copies that compose $pc$, and $Pre$. We know such a set of injective typed graph homomorphisms cannot be found from $in$ into $Pre$. However, the abstraction relation in \cref{def:abstraction_pc_ex_appendix} states that an injective typed graph homomorphism exists between each rule copy in the match part of $pc$ and $Input^{*}$. We thus know that there cannot exist an injective typed graph homomorphism between $Pre$ and $Input^{*}$. 
  \item For certain executions, the property holds on the path condition but the property's pre-condition cannot be found in the execution.
\begin{equation}
\label{item:prepostnotholds}
\forall f\; \exists g\,.\,\big(in\stackrel{f}{\blacktriangleleft} Pre \land out \stackrel{g}\blacktriangleleft p\big) \implies \nexists f'\,.\,\big(Pre \stackrel{f'}{\vartriangleleft} Input^{*}\big)
\end{equation}
These are the executions where a set of injective typed graph homomorphisms can be found from $in$ into $Pre$, but not from $in$ into $Input^{*}$, as required by the abstraction relation. If this is the case then this means that at least two vertices of $in$ belonging to different rule copies that were mapped by $f$ into the same vertex of $Pre$, are mapped into different vertices of $Input^{*}$ by $f'$ (or vice-versa). 
  \item For the remaining set executions abstracted by $pc$, if the property holds on the path condition then the property holds on the execution. Formally, according to \cref{def:sat_prop_pc_appendix} and \cref{def:sat_prop_ex_appendix} we have that:
\begin{multline}
\label{item:prepostholds}
\forall f\; \exists g\,.\,\big(in\stackrel{f}{\blacktriangleleft} Pre \land out \stackrel{g}\blacktriangleleft p\big) \implies \\\forall f'\; \exists g'\,.\,\big(Pre \stackrel{f'}{\vartriangleleft} Input^{*} \land p \stackrel{g'}{\vartriangleleft} ex^*\big)\\
\text{where } Dom(g)\cap Match(pc^{*}) = Dom(f) \text{ and}\\
V(Input)\cap CoDom(g') = CoDom(f')
\end{multline}
This is the case where every two vertices of $in$ belonging to different rule copies that were mapped by $f$ into a common vertex of $Pre$, are also mapped into a common vertex of $Input^{*}$ by $f'$. We thus need to show that the fact that the post-condition of the property holds on the path condition implies that the post-condition of the property also holds on the execution, i.e. that $out \stackrel{g}\blacktriangleleft p \implies p \stackrel{g'}{\vartriangleleft} ex^*$. This proposition is true because we know by \cref{def:abstraction_pc_ex_appendix} of abstraction relation that a surjective typed graph homomorphism exists between the output part of $ex$ and the apply part of $pc$. By composing this surjection with the surjection between $out$ and $p$ we take as hypothesis, we know a surjective typed graph homomorphism exists between the output of $ex$ and $p$. The inverse of this composed homomorphism contains an injective typed graph homomorphism between $p$'s post-condition and $ex$. We are thus missing accounting for the traceability links between the pre- and post-condition of property $p$, if they exist.  According to Proposition~\ref{item:prepostholds} we know that $in$ and $out$ overlap on their subgraphs that are isomorphic to $p$'s precondition. By \cref{def:abstraction_pc_ex_appendix} of the abstraction relation, we know that an injective typed graph homomorphism can be found between each strongly connected component formed of symbolic traceability links of $pc$, and $ex$. We also know that a typed graph surjective homomorphism exists between $out$ and $p$. We thus know that the traceability links between the pre- and post-condition of $p$ can be injectively found in $ex$. Note that strongly disjoint connected symbolic traceability link components mapped from $pc$ to $ex$ may be mapped onto joined traceability link components in $ex$ when disjoint vertices of the match part of $pc$ are mapped onto the same input vertex in $ex$. 
\end{enumerate}

The three cases above cover all executions that can be abstracted by a path condition, and as such we know that if the property holds on a path condition, it will necessarily hold on all the executions that path condition abstracts.
\end{pf}
% 
\begin{lemma}{If a property holds for a transformation execution then the property holds for the path condition that abstracts it.\\}
\label{lemma:validity2_appendix}
Let $tr$ be a transformation, $pc\in \textsc{Pathcond}(tr)$ be a path condition of $tr$, $ex\in \textsc{Exec}(tr)$ be an execution of $tr$ and $p\in \textsc{Prop}(tr)$ be a property of $tr$. Then we have that:
\begin{equation}
\label{eq:prop_proof_right_left_appendix2}
\forall ex\in\{ex\in\textsc{Exec}(tr)\;|\;ex\Vvdash pc\}\;.\; ex\models p\;\Longrightarrow \; pc\vdash p
\end{equation}
\end{lemma}
\begin{pf}
We will demonstrate Proposition~\ref{eq:prop_proof_right_left_appendix2} holds by contraposing it:
% By using contraposition we can show that:
\begin{multline}
\label{eq:prop_proof_right_left_contra}
\neg(pc\vdash p) \;\Longrightarrow \;\\\exists ex\in\{ex\in\textsc{Exec}(tr)\;|\;ex\Vvdash pc\}\;.\; \neg(ex\models p)
\end{multline}

By \cref{def:sat_prop_pc_appendix} we know that $pc\vdash p$ is equivalent to proposition $\forall f\; \exists g\,.\,\big(in\stackrel{f}{\blacktriangleleft} Pre \implies out \stackrel{g}\blacktriangleleft p\big)$, where $Pre$ is $p$'s pre-condition, $in$ is a subgraph of the containment transitive closure of the match part of $pc$, and $out$ is a subgraph of the containment transitive closure of $pc$. We also know by \cref{def:sat_prop_ex_appendix} that $ex\models p$ is equivalent $\forall f\; \exists g\,.\,\big(Pre \stackrel{f}{\vartriangleleft} Input^{*} \implies p \stackrel{g}{\vartriangleleft} ex^*\big)$, where $Input$ is the input part of $ex$. After replacing the left and the right hand side of Proposition~\ref{eq:prop_proof_right_left_contra} by equivalent formulas and solving the negations we reach the conclusion we need to prove:
\begin{multline}
\label{eq:prop_proof_right_left_contra_neg}
\exists f\; \forall g\,.\,\big(in\stackrel{f}{\blacktriangleleft} Pre \land \neg(out \stackrel{g}\blacktriangleleft p)\big) \;\Longrightarrow \;\\\exists ex\in\{ex\in\textsc{Exec}(tr)\;|\;ex\Vvdash pc\}\;.\;\\ \exists f'\; \forall g'\,.\,\big(Pre \stackrel{f'}{\vartriangleleft} Input^{*} \implies \neg(p \stackrel{g'}{\vartriangleleft} ex^*)\big)
\end{multline}

We thus need to demonstrate that whenever the pre-condition of the property is found at least once in a path condition, but not its corresponding post-condition, then the same thing happens for at least one of the executions abstracted by that path condition. We know by Proposition~\ref{eq:prop_proof_right_left_contra_neg} that $in\stackrel{f}{\blacktriangleleft} Pre$, i.e. the precondition of the property is found at least once in the path condition. We thus know that there exists one execution for which $Pre \stackrel{f'}{\vartriangleleft} Input^{*}$ holds, which is the execution for which the surjective typed graph homomorphism $f$ maps vertices belonging to the match parts of different rule copies in the same fashion that the set of injective typed graph homomorphisms from the abstraction relation in \cref{def:abstraction_pc_ex_appendix} maps to the match part of $pc$ onto $input^{*}$.

In order to complete the proof we need to show that the fact that $\neg(out \stackrel{g}\blacktriangleleft p)$, i.e. if the complete property cannot be found in the path condition, then $\neg(p \stackrel{g'}{\vartriangleleft} ex^*)$, i.e. the complete property cannot be found in the execution. Note that, according to \cref{def:sat_prop_pc_appendix} and \cref{def:sat_prop_ex_appendix}, we know the considered complete property graphs both in $p$ and $ex$ found by $g$ and $g'$ are anchored on the pre-condition graphs of the property found by $f$ and $f'$. Because of the abstraction relation, we know a surjective typed graph homomorphism between the output of $ex^{*}$ and the apply part of $pc$ exists. Given a surjective typed graph homomorphism does not exist between $pc$ and $p$, we know certain vertices and/or edges that exist in $p$, either in its apply part or in its symbolic traceability links, do not exist in $pc$. If the missing vertices and/or edges are part of the $apply$ part of $p$ then we are sure an injective typed graph homomorphism cannot exist between $p$ and $ex$ because $ex$ also does not contain those vertices or edges. If the missing edges are symbolic traceability edges then, according to the condition on strongly connected components in the abstraction relation in \cref{def:abstraction_pc_ex_appendix}, we know that the traceability links in $ex$ can be surjectively mapped onto $pc$. Because some of those traceability links are missing in $p$, an injective typed graph homomorphism cannot exist between $p$ and $ex$.

% We only need to worry about the case where we have  $ex\Vvdash pc\land
% \neg(pc\vdash p)$ given our property proof results are meaningful only for
% transformation executions which are abstracted by the path condition being
% examined\levi{because by the unicity result one execution is only abstracted by one path condition}. In this
% case we have that, because $ex\Vvdash pc$, by \cref{def:abstraction_pc_ex} we know
% that there is a surjective typed graph homomorphism between between the apply
% pattern of transformation execution $ex$ and the apply pattern of the path
% condition $pc$. Given we know that $\neg(pc\vdash p)$, then we know that an
% injective type graph homomorphism does not exist between $p$ and $pc$. This
% means that, by the \cref{def:typed_graph_homomorphism} of injective typed graph
% homomorphism, there exists a type $T$ which is instantiated in $pc$ but not in
% $prop$. However, because of the fact that $ex\Vvdash pc$ we know that $T$ is
% instantiated in $ex$ because a surjective typed graph homomorphism exists
% between $ex$ and $pc$, implying type $T$ is instantiated at least once in $ex$.
% We thus can deduce that we have $\neg(ex\models p)$ given that $ex\models p$
% implies that there exists an injective type graph homomorphism between property
% $p$ and execution $ex$. This injective type graph homomorphism cannot exist
% given the fact an instance of $T$ exists in $ex$ but not in $p$.

\end{pf}

\begin{proposition}{(Completeness) Properties of a transformation can be shown to either hold for all transformation executions, or not hold for at least one transformation execution.}
\label{prop:proof_completeness_appendix}
\end{proposition}
\begin{proof}
This results follows from two previous results: \cref{prop:pc_completeness_appendix}, that tells us that every transformation execution is abstracted by one path condition; and \cref{prop:proof_validity_appendix} that shows us that every path condition is taken into consideration during property proof. Note that \cref{lem:uniqueness_appendix} guarantees consistency of our results, in the sense that the uniqueness of one path condition per transformation execution guarantees that a property cannot be proven to be both \emph{true} and \emph{false} for two path conditions representing the same transformation execution.
\end{proof}

\clearpage

### 2015-03-13 text/building_path_conditions.tex
previous version of the paper

\section{Building Path Conditions}
\label{sec:building_pcs}

This section will present how path conditions are structured to represent symbolic rule execution. As well, we present our approach to building a set of path conditions to represent all executions of a DSLTrans transformation.

\subsection{Symbolic Execution}

Our algorithm operates on the principle of symbolic execution to build up these
path conditions. In order to explain the concept of symbolic execution of a
transformation, let us make an analogy with program symbolic execution as
introduced by King in his seminal work ``\emph{Symbolic Execution and Program
Testing}''~\cite{DBLP:journals/cacm/King76}. According to King, a symbolic
execution of a program is a set of \emph{constraints} on that program's
\emph{input variables} called \emph{path conditions}. Each \emph{path condition}
describes a traversal of the conditional branching commands of that program. A
\emph{path condition} is symbolic in the sense it \emph{abstracts} as many
concrete executions as there are instantiations of the path condition's
variables that render the path condition's constraints true.

We can transpose this notion of symbolic execution to model transformations. The
analog of an input variable in the model transformation context are
\emph{metamodel classes, relations and attributes}. As program statements impose
constraints on input and output variables during symbolic execution,
transformation rules impose conditions on which metamodel elements are
instantiated during a concrete transformation execution, and how that
instantiation happens. As well, rules in a model transformation are implicitly
or explicitly scheduled. These control and/or data dependencies must be
taken into consideration during path condition construction.

As in program symbolic execution, each path condition in our approach \emph{abstracts} as many concrete executions as there are input/output models
that satisfy them. This is formulated as an \emph{abstraction relation} as further explained in \cref{sec:abstraction_relation}.

In what follows we will examine in more detail how these symbolic execution
principles can apply to the verification of model transformations.

\subsection{Path Conditions}
\label{sec:gen_path_conds}

In order to present the intuition of path conditions and symbolic executions, we
first discuss the idea of \textit{rule combinations}.

As seen in \cref{sec:dsltrans}, a layer in a DSLTrans transformation contains a
number of rules.  We can create a set of rule combinations for this layer by
taking the powerset of all rules in that layer. Each rule combination in this
set will represent all possible transformation executions where the rules in
that combination would execute.

For example, in \cref{fig:rule_combos2}, the rule combination marked `AC'
represents the set of transformation executions where the rules A and C would
execute and no others. Another rule combination marked `A' represents the
transformation executions where only rule A would execute.

\begin{figure}[h!] \centering \includegraphics[width=.40\textwidth]{./figures/overview/rule_combos.pdf}
	\caption{Rule combinations created for a transformation layer}
	\label{fig:rule_combos2}
\end{figure}

Note that within these rule combinations, the number of times a rule has
executed is abstracted. Either a rule has executed zero times, and the rule is
not represented in a rule combination, or the rule has executed some finite
number of times and it is represented. This abstraction is key to our approach,
as it allows us to create a finite set of path conditions to abstract over an
infinite set of transformation executions, as seen in
\cref{sec:abstraction_relation}.

We also note that rule \textit{combinations} are created, and not rule
\textit{permutations}. This follows from the semantics of \\DSLTrans as described
in \cref{sec:dsltrans}, as transformation rules in a layer will execute in a
non-deterministic order but produce a deterministic result, by construction of the semantics of DSLTrans. As a final
note, the transformation executions that these rule combinations represent always terminate, also by construction of the semantics of DSLTrans~\cite{DBLP:conf/sle/BarrocaLAFS10}.

We base our concept of path conditions on these rule combinations. However, as
DSLTrans allows for dependencies between rules, we cannot create path conditions
for the transformation by taking the powerset of all rules. Instead, our
approach must move layer-by-layer and resolve the dependencies between rules.
The next two sections will introduce the concepts of traceability and dependency, before we
briefly discuss the syntax and semantics of path conditions themselves.

\subsubsection{Traceability}
\label{subsubsec:traceability}

DSLTrans rules allow for dependencies to be specified on which elements of the
output model were created from specific elements of the input model. To resolve
these dependencies, traceability information for the transformation is created
during the execution of a DSLTrans model transformation~\cite{DBLP:conf/sle/BarrocaLAFS10}.
In our verification approach, we store this same information as symbolic \emph{traceability links}, in
order to record which elements belong to the same DSLTrans rule.

At a particular point in the path condition construction process, symbolic traceability links are built for
each rule as follows: for all match and apply elements of a rule, given a match
element belonging to the match graph of a rule and an apply element belonging to
the apply graph of the same rule, a symbolic traceability link is built between the two
if the apply element is not connected to a backward link (as explained below).
This is intuitive: traceability links are built between a newly generated
element in the output model, and the elements of the input model that originated
it.

\begin{figure}[htb]
        \centering
        \begin{subfigure}[b]{0.24\textwidth}
                \centering
                \includegraphics[width=1\textwidth]{./figures/building_path_conditions/traceability_links.pdf}
                \caption{Before symbolic traceability links added}
                \label{fig:traceability_links1}
        \end{subfigure}%
        ~~
        \begin{subfigure}[b]{0.235\textwidth}
                \centering
                \includegraphics[width=1\textwidth]{./figures/building_path_conditions/traceability_links2.pdf}
                \caption{After symbolic traceability links added}
                \label{fig:traceability_links2}
        \end{subfigure}%
        \caption{Symbolic traceability links created for an abstract DSLTrans rule}
        \label{fig:trace_links}
\end{figure}

An example of the symbolic traceability link creation process is shown in
\cref{fig:trace_links}. Note that symbolic traceability links are a solid line between
match and apply elements in our visual notation.

\subsubsection{Backward Links}
\label{subsubsec:backward_links}

The dependencies in a DSLTrans rule are specified using the \textit{backward link} construct, as further detailed in \cref{subsec:DSLTrans_constructs} and \cref{def:transformation_rule}. \cref{subsubsec:resolve_dependencies} will discuss how these dependencies are then resolved during our symbolic execution approach.

\cref{fig:backward_links} demonstrates how backward links are used \\within a rule. The rule shown contains a backward link, which defines the dependency that an element of type X was created from an element of type A, and an element of type Y was created from an element of type B. If this dependency is satisfied, then another element of type Z should be created. This element should be associated with the Y element.

\begin{figure}[htb]
        \centering
        \begin{subfigure}[b]{0.235\textwidth}
                \centering
                \includegraphics[width=1\textwidth]{./figures/building_path_conditions/backward_link.pdf}
                \caption{Rule with backward links (dashed lines)}
                \label{fig:backward_links}
        \end{subfigure}%
        ~~
        \begin{subfigure}[b]{0.235\textwidth}
                \centering
                \includegraphics[width=1\textwidth]{./figures/building_path_conditions/backward_link2.pdf}
                \caption{Traceability links added\\~}
                \label{fig:backward_links2}
        \end{subfigure}%
        \caption{Adding traceability links to an abstract DSLTrans rule with backward links}
        \label{fig:back_links}
\end{figure}

\cref{fig:backward_links2} shows the rule after symbolic traceability links have been added. Two symbolic traceability links are created from the Z element to the A and B elements in the match graph to store traceability information. Note that no symbolic traceability links are built between two elements connected by backward links, as these links have already been built in a previous layer.

% \begin{definition}{Rule Traceability}
% 
% Let $rl = \big\langle V,E,st,\tau \big\rangle$ be a transformation
% rule. $trace_{\langle
% V_{\Delta},E_{\Delta},st_{\Delta},\tau_{\Delta}\rangle}(\langle
% V,E,st,\tau\rangle) = \langle V,E',st',\tau'\rangle$ where we have that
% $E\subseteq E'$, $st\subseteq st'$, $\tau \subseteq \tau'$ and if
% $v_1\xrightarrow{e} v_2\in E'\setminus E$ then $v_1\in Output(V_{\Delta})$,
% $v_2\in Input(V)$ and $\tau'(e)=trace$.
% \end{definition}

\subsubsection{Syntax and Semantics}
\label{subsubsec:path_condition_creation}

A path condition represents the symbolic execution of a set of DSLTrans rules, similar to a rule combination as explained above. Again, we use an abstraction over the number of times a rule has symbolically executed. Each path condition will represent that a rule has not executed, or has executed one or more times.

The path condition generation algorithm will symbolically combine transformation rules into a path condition. Each path condition will then abstract a set of concrete transformation executions, as defined by our abstraction relation in \cref{sec:abstraction_relation}.

As seen in the rest of this section, the structure of path conditions is similar to that of DSLTrans rules. The match graph of a path condition represents a pattern that must be present in the input model of the transformation, while the apply graph is a pattern which will be instantiated in the output model of the transformation. Symbolic traceability links are also kept between elements in the match and apply graphs to retain traceability information.

The formal definition of a path condition is presented in \cref{def:path_condition}.

\begin{definition}{Path Condition\\}
\label{def:path_condition}
\CatchFileBetweenTags{\pcdef}{text/definitions}{pcdef}{\pcdef}
\end{definition}

\CatchFileBetweenTags{\pc}{text/definitions}{pc}{\pc}

\subsection{Path Condition Generation Algorithm}
\label{sec:gen_all_pcs}

 This section will describe how path conditions are constructed for a DSLTrans transformation using our approach.

\cref{fig:next_layer} outlines the path condition generation algorithm. The algorithm will examine each transformation layer in turn. Path conditions from the previous layer will be combined with rules from the current layer to create a new set of path conditions. This new set of path conditions will then be combined with the rules from the next layer to produce yet another set of path conditions, and so on. At the end of the algorithm, a complete set of path conditions for the entire transformation will have been produced.

\begin{figure*}[htb]
        \centering
        \begin{subfigure}[b]{0.34\textwidth}
                \centering
                \includegraphics[width=1\textwidth]{./figures/building_path_conditions/next_layer.pdf}
                \caption{Previous path conditions are combined\\ with rules}
                \label{fig:next_layer}
        \end{subfigure}%
        ~~
        \begin{subfigure}[b]{0.34\textwidth}
                \centering
                \includegraphics[width=1\textwidth]{./figures/overview/layers_pc.pdf}
                \caption{Combining a path condition with two rules\\~}
                \label{fig:layers_pc2}
        \end{subfigure}%

         \caption{Two components in the path condition creation process}
         \label{fig:combining_path_conditions}
\end{figure*}

We now define what is occurring in the `combination step' in \cref{fig:next_layer}. This step begins by selecting each path condition in the working set, one at a time. Note that at the beginning of the path condition creation process, this working set consists of an empty path condition.

A new set of path conditions will then be created by sequentially combining each rule in the layer with the path condition selected. Recall that a path condition represents a set of rules that have symbolically executed, thereby abstracting a set of transformation executions through our abstraction relation. Combining a path condition with a rule will produce one or more path conditions depending on how the rule combines with the rules already represented by the path condition. The pre- and post- conditions defined by the path condition will be modified according to the elements found in that rule.

Each of the new path conditions created from combining a rule with a path condition will then be combined with the next rule in the layer. A small example is shown in \cref{fig:layers_pc2}, where a path condition is combined with two rules. Note that a rule can combine with a path condition in multiple ways (differentiated by prime marks in the figure). \Cref{fig:all_pcs2} shows how path conditions from the previous layer are sequentially combined with all the rules from the current layer. All the path conditions for the layer are then collected to produce the final working set of path conditions for the layer.

        \begin{figure}[bht]
                 \centering
                  \includegraphics[width=.38\textwidth]{./figures/overview/all_pcs.pdf}
                 \caption{Creating all path conditions for a layer}
                 \label{fig:all_pcs2}
         \end{figure}

\subsection{Combining a Path Condition with a Rule}
We will now examine the combination step between one path condition and one rule, which produces a set of new path conditions. A formal and generic definition of this step will be presented first, before we explain the specialized combination possibilities with figures and informal text.

\begin{definition}{Combination of a Path Condition with a Rule}
\label{def:combine_pc_with_rule}

\CatchFileBetweenTags{\pccfrc}{text/definitions}{pccfrc}{\pccfrc}
\end{definition}

\CatchFileBetweenTags{\pccfrctext}{text/definitions}{pccfrctext}{\pccfrctext}

We will now discuss the combination step possibilities. Let PC be the path condition selected from layer n-1, and R the rule selected from layer n. When PC and R are combined, there are four possibilities based on the dependencies between PC and R:

\begin{enumerate}
\item R has \textbf{no} dependencies
\item R has dependencies and \textbf{cannot} execute
\item R has dependencies and \textbf{may} execute
\item R has dependencies and \textbf{will} execute
\end{enumerate}

These dependencies are defined by the backward links within R. As mentioned in \cref{subsubsec:traceability}, backward links enforce that the elements in the apply graph were created by the connected elements in the match graph. In the context of combining a rule and a path condition, these backward links define dependencies between the rule and the elements created by the rules represented by the path condition.

The below figures will demonstrate the four cases above. As a reminder of visual notation, the backward links are dashed lines between the match and apply graphs of the rule and path condition, while symbolic traceability links are solid lines between the two graphs.

\subsubsection{No Dependencies}
\label{enum:no_back}

The rule R has a match graph which represents its pre-conditions. For a particular transformation execution, it is possible that this match graph would not match a specific input model, and thus R would not execute in these transformation executions. To represent all such transformation executions where the rule R would not execute, PC is copied unchanged to the new set of path conditions.

To represent the transformation executions where the match graph of R would match, and therefore R would execute, a new path condition is produced which consists of the union between R and PC. This situation is
seen in \cref{fig:no_dependencies} and formally defined in \cref{def:rule_comb_no_dependencies}.

\begin{figure}[bt] \centering \includegraphics[width=0.44\textwidth]{./figures/building_path_conditions/no_dependencies.pdf}
	\caption{R has no dependencies}
	\label{fig:no_dependencies}
\end{figure}

% \begin{definition}{No Dependencies Between a Rule and a Path Condition}
% 
% \levi{FIX}
% Let $A=\langle V,E,st,\tau\rangle, B=\langle V',E',st',\tau'\rangle\in \textsc{Pathcond}^{sr}_{tg}$ be two path conditions. We say that $R$ does not depend on $PC$ if and only if $\nexists e\in E'\,.\,\tau'(e)=backward$.
% \end{definition}

\begin{definition}{Path Condition and Rule Combination -- No Dependencies\\}
\label{def:rule_comb_no_dependencies}
\CatchFileBetweenTags{\nodeps}{text/definitions}{nodeps}{\nodeps}

\end{definition}

\CatchFileBetweenTags{\nodepstext}{text/definitions}{nodepstext}{\nodepstext}

\subsubsection{Resolving Dependencies}
\label{subsubsec:resolve_dependencies}
If R contains backward links and thus R defines dependencies on PC, then we need to analyse whether PC can satisfy those dependencies. This is done by matching the backward links in R over the symbolic traceability links in PC. Note that symbolic traceability links in R are not required to be found in PC, and that only backward links define dependencies.

\paragraph{Unsatisfied Dependencies}

If the backward links in R cannot be matched to symbolic traceability links in PC, then in the transformation executions abstracted by PC, R cannot execute. Again, PC will be copied unchanged to the new set of path conditions, but no new path condition will be created. This case is shown in \cref{fig:non_satisfied_dependencies}, where the backward links between the two B elements in R cannot match over the symbolic traceability link in PC. \cref{def:rule_comb_unsatisfied} describes this case formally.

\begin{figure}[h!] \centering \includegraphics[width=0.44\textwidth]{./figures/building_path_conditions/non_satisfied_dependencies.pdf}
	\caption{R's dependencies are not satisfied by PC}
	\label{fig:non_satisfied_dependencies}
\end{figure}

\begin{definition}{Path Condition and Rule Combination -- Unsatisfied Dependencies\\} 
\label{def:rule_comb_unsatisfied}
\CatchFileBetweenTags{\unsatdeps}{text/definitions}{unsatdeps}{\unsatdeps}
\end{definition}

According to the pre-conditions of the equation presented in \cref{def:rule_comb_unsatisfied}, a path condition does not satisfy the dependencies present in a rule if there is no surjective typed graph homomorphism between the backward links of the rule and the symbolic traceability links of the path condition. Besides expressing the fact that all backward links must exist as symbolic traceability links the path condition, the surjective homomorphism allows modeling the case where dependencies expressed by two (or more) backward links between similarly typed elements can be satisfied one single symbolic traceability link in the path condition. This is the case, for example, of rule \emph{FemaleToFemale} in the \emph{Police Station} in \cref{fig:dsltransformation}. The two similarly typed backward links in this rule are satisfied by a path condition containing only the rule \emph{females} generated from the first layer of the transformation, holding one single symbolic traceability link.

\paragraph{Partially- and Totally- Satisfied Dependencies}

Consider the possibility that the backward links of R can be found in PC, and R's dependencies are met. The question then becomes whether the rule R \textbf{may} or \textbf{will} execute in the abstracted transformation executions.

To resolve this question, the match graph of R, along with R's backward links, is matched to PC's match graph and traceability links. If all of these elements are found, then we denote this as the `totally-satisfied case', where R \textbf{will} necessarily execute in the abstracted transformation executions. Otherwise, we denote the `partially-satisfied' case, where R \textbf{may} execute. Note that we break up these cases for ease of explanation only. Formally, both cases are encompassed by \cref{def:rul_comb_partial_total}.

In the totally-satisfied case, R will be ``glued'' overtop PC, as seen in \cref{fig:total_satisfied_dependencies}. This gluing operation is anchored where the backwards links in R match over the traceability links in PC. The purpose of this operation is to include any elements in R's apply graph that may not exist in PC. Thus, all elements and associations which exist in both PC and R are ignored. Note that if multiple total matches exist in PC, that R will be glued at multiple points as seen in \cref{fig:multiple_total_satisfied_dependencies}. This ``gluing'' operation is also defined formally in \cref{def:rul_comb_partial_total}, as the addition of a delta graph.

\begin{figure*}[htb]
        \centering
        \begin{subfigure}[b]{0.7\textwidth}
                \centering
                \includegraphics[width=1\textwidth]{./figures/building_path_conditions/total_satisfied_dependencies.pdf}
                \caption{Totally satisfied at one location}
                \label{fig:total_satisfied_dependencies}
        \end{subfigure}%
        \\
        \begin{subfigure}[b]{0.8\textwidth}
                \centering
                \includegraphics[width=1\textwidth]{./figures/building_path_conditions/multiple_total_satisfied_dependencies.pdf}
                \caption{Totally satisfied at multiple locations}
                \label{fig:multiple_total_satisfied_dependencies}
        \end{subfigure}%
        \caption{R's dependencies are totally satisfied by PC}
        \label{fig:totes_sat_deps}
\end{figure*}

In the partially-satisfied case, rule R may or may not execute. Note that in \cref{fig:partial_satisfied_dependencies}, PC does not have the association between the A and B elements in the match graph. This means that it is possible that the input model for the transformation does not have this association present. In these transformation executions R would not execute.
\cref{fig:partial_satisfied_dependencies} shows the two path conditions produced in this case. The first produced is a copy of PC, where R does not symbolically execute. The second is where R symbolically executes at the matched location. Therefore, R is glued onto PC, with the gluing step the same as in the totally-satisfied case above.

\begin{figure*}[tb] \centering \includegraphics[width=0.8\textwidth]{./figures/building_path_conditions/partial_satisfied_dependencies.pdf}
	\caption{R's dependencies are partially satisfied by PC}
	\label{fig:partial_satisfied_dependencies}
\end{figure*}

Note that this gluing procedure must consider all matching possibilities, for each location the rule might match over the input model. For example, in \cref{fig:multiple_partial_satisfied_dependencies}, rule R has a backward link that can be partially matched on two locations in PC: the left-hand and right-hand pairs of traceability links. Therefore, there are four possibilities for how R would match over PC: not at all, on the left-hand side of PC, on the right-hand side, or on both sides. These four possibilities define the four new path conditions created.

The first is a copy of PC, as R is assumed to not execute and will produce no new elements. The second is where R will be glued on top of the backward links on the left-hand side, to add the elements that do not exist in PC already. The third is where the gluing will occur on the right-hand side. The fourth path condition produced is the case where R will be glued at both locations.

\begin{figure*}[tb] \centering \includegraphics[width=0.8\textwidth]{./figures/building_path_conditions/multiple_partial_satisfied_dependencies.pdf}
	\caption{R's dependencies are partially satisfied by PC, and are glued at all possible matches}
	\label{fig:multiple_partial_satisfied_dependencies}
\end{figure*}

Note that rules may also contain transitive links in their match graphs. In this case, the partial or total matching of R onto PC must consider all transitive matches in order to produce all valid path conditions.

As we have done for the previous cases, let us now formally define the combination step when a rule has partially and/or totally defined dependencies. As these cases are more complex than the previous two, we will need to construct the mathematical model of this case incrementally. We will start by an auxiliary relation that partially or totally combines a set of path conditions with a rule.

\begin{definition}{Single Partial and Total Combination of a Set of Path Conditions with a Rule\\}
\label{def:comb_path_cond_rule_single}
\CatchFileBetweenTags{\satdeps}{text/definitions}{satdeps}{\satdeps}

\begin{align}
\label{eq:pcomb}
\CatchFileBetweenTags{\satdepseqone}{text/definitions}{satdepseqone}{\satdepseqone}
\end{align}

\begin{align}
\label{eq:tcomb}
\CatchFileBetweenTags{\satdepseqtwo}{text/definitions}{satdepseqtwo}{\satdepseqtwo}
\end{align}

\end{definition}

Let us start by introducing relation $\stackrel{p\_comb}{\rightarrow}$, presented in \cref{eq:pcomb} of \cref{def:comb_path_cond_rule_single}. The relation takes as arguments a set of path conditions being accumulated for the current layer, the rule to be combined, and an $rl_{glue}$ argument indicating the place in each of the input path conditions the rule should be anchored to during the combination step. The relation's output is a new set of path conditions. This new set includes all the original path conditions, as well as each path condition in the accumulator set ``glued'' to a copy of rule being examined. Note that the relation $\stackrel{t\_comb}{\rightarrow}$ in \cref{eq:tcomb} of \cref{def:comb_path_cond_rule_single} is similarly defined, except for the fact path conditions in the accumulator set are not preserved in the relation's output set.\\

\setcounter{equation}{0}

Let us now define how a rule is combined with a path condition, whenever its backward links can be found several times in that path condition. This situation is described in the examples in \cref{fig:multiple_total_satisfied_dependencies} and \cref{fig:multiple_partial_satisfied_dependencies}. We formalize it in \cref{def:comb_path_cond_rule_mul}, by means of relations $\stackrel{p\_step}{\rightarrow}$ and $\stackrel{t\_step}{\rightarrow}$. These two relations operationally describe the sequence of steps necessary to ``glue'' a rule at multiples places of a path condition. The set of places targeted in the path condition for receiving a copy of the rule is given by the sets $partialSet$ and $totalSet$ (found respectively in Equation~(2) and Equation~(4) of \cref{def:comb_path_cond_rule_mul}). As expected, these sets contain the set of traceability links in the path condition where copies of the rule need to be anchored to.

\begin{definition}{Multiple Partial and Total Combination of a Set of Path Conditions with a Rule\\}
\label{def:comb_path_cond_rule_mul}

\CatchFileBetweenTags{\satdepstwo}{text/definitions}{satdepstwo}{\satdepstwo}
\end{definition}

Having \cref{def:comb_path_cond_rule_single} and \cref{def:comb_path_cond_rule_mul} in mind, we can now proceed to define the complete combination relation of a rule with a path condition in the case of partially and totally satisfied dependencies.

\begin{definition}{Path Condition and Rule Combination -- Partially and Totally Satisfied Dependencies\\}
\label{def:rul_comb_partial_total}

\CatchFileBetweenTags{\satdepsthree}{text/definitions}{satdepsthree}{\satdepsthree}
\end{definition}

The top equation in \cref{def:rul_comb_partial_total} defines the $\stackrel{combine}{\rightarrow}$ relation for when rule $rl$ has dependencies that are satisfied by path condition $pc$. The pre-conditions in the equation state that the backward links in the rule are found in the path condition, as expected. Additionally, two sequential steps perform the gluing of the rule $rl$ on all path conditions in accumulator $AC$, wherever the rule is partially and/or totally found in each of those path conditions. Relations $\stackrel{p\_comb}{\rightarrow}$ and $\stackrel{t\_comb}{\rightarrow}$ presented in \cref{def:comb_path_cond_rule_mul} are used to model these two operational ``gluing'' steps. Functions $partialsat$ and $totalsat$, described in the latter part of \cref{def:rul_comb_partial_total}, are used to gather the places of path condition $pc$ where copies of the rule need to be anchored to.

\subsubsection{Considering Further Rules}
\label{sec:further_rules}
Thus far we have described how to create a set of path conditions that represent how one rule from a layer will add new elements to one path condition from the previous layer. These path conditions are then themselves combined with the next rule in the layer in the same manner. Note that in \cref{def:rul_comb_partial_total} the choice of next rule does not matter, due to the rule non-interference guaranteed by the semantics of DSLTrans. In order to represent this non-interference in the construction of path conditions, we specify that the matching of rule dependencies is against the path condition from the previous layer (variable $pc$ in the main equation of \cref{def:rul_comb_partial_total}), not the specific path condition the rule is to be combined with in the accumulator argument of the $\stackrel{combine}{\rightarrow}$ relation. This ensures that the result of combining one rule with a path condition will have no impact on how following rules will combine.

% Note that the choice of next rule does not matter, due to the rule non-interference guaranteed by the semantics of DSLTrans. In order to represent this non-interference in the construction of path conditions, we specify that the matching of rule dependencies is against the path condition from the previous layer, not the specific path condition the rule is to be combined with. This ensures that the result of combining one rule with a path condition will have no impact on how following rules will combine.

The combination of one path condition with all the rules in the layer will produce a new set of path conditions. This process is depicted in \cref{fig:all_pcs2} and formalized in \cref{def:path_cond_layer_comb} by the layer combination relation $\stackrel{combpclayer}{\rightarrow}$.

\begin{definition} {Combining a Path Condition with a Layer\\}
\label{def:path_cond_layer_comb}
\CatchFileBetweenTags{\combpclayer}{text/definitions}{combpclayer}{\combpclayer}
\end{definition}

\CatchFileBetweenTags{\combpclayertext}{text/definitions}{combpclayertext}{\combpclayertext}

\begin{definition} {Combining a Set of Path Conditions with a Layer\\}
\label{def:path_cond_set_layer_comb}
\CatchFileBetweenTags{\combpcsetlayer}{text/definitions}{combpcsetlayer}{\combpcsetlayer}
\end{definition}

\CatchFileBetweenTags{\combpcsetlayertext}{text/definitions}{combpcsetlayertext}{\combpcsetlayertext}

\begin{definition} {Path Condition Generation\\}
\label{def:path_cond_gen}
\CatchFileBetweenTags{\pathcondgen}{text/definitions}{pathcondgen}{\pathcondgen}
\end{definition}

\CatchFileBetweenTags{\pathcondgentext}{text/definitions}{pathcondgentext}{\pathcondgentext}

### 2015-03-13 text/conclusion.tex
previous version of the paper

\section{Conclusion}
\label{sec:conclusion}

In this paper we have adapted symbolic execution techniques to verify DSLTrans
model transformations. As well, we have presented an algorithm to prove model syntax
relation properties by building all possible path conditions for a
transformation.

The concrete contributions of our work are the following:

\begin{itemize}
\item An algorithm for constructing all path conditions representing all executions of a DSLTrans transformation.
\item A property-checking algorithm that proves model syntax relation properties over all path conditions, and therefore over all transformation executions.
\item Validity and completeness proofs for the path condition construction and property proof algorithms.
\item A discussion of optimisations and scalability concerns for our methods, along with results from an industrial application.
\end{itemize}

As is the case in general for exhaustive verification methods, we have encountered theoretical and practical limitations when developing our technique. From a theoretical standpoint, not all DSLTrans transformations are currently addressed by the technique presented here. In particular, DSLTrans transformations where rules overlap in the match part (as per \cref{def:layer_transformation} in \cref{sec:formal_background}) are not currently treated. Addressing overlapping rules theoretically implies some revisions to the formalisation presented here: on the one hand, rules that overlap imply rule dependency management during path condition construction; on the other hand, the uniqueness lemma in \cref{lem:uniqueness} of \cref{sec:abstraction_relation} needs to be re-analysed under looser constraints. Note however that we have already addressed this issue in practice in~\cite{gehan:13,conf/gg/SelimLCDO14} and that we expect the impact in the theory to be relatively small.

Another theoretical limitation has to do with the properties that can be proved using our technique. As expected, the chosen abstraction relation we use imposes limitations on which properties can be shown to hold or not hold on a DSLTrans transformation. In particular, because in general we only consider one rule copy in each path condition, we cannot prove properties where the pattern in the property implies the same rule matches on an input model more than once. We do not see this as a too strong limitation of our technique given that: on the one hand we are able to prove, for all executions of a DSLTrans transformation, a range of properties concerning the interaction between different rules in a DSLTrans transformation, which is where we expect most errors to occur; on the other hand we believe we can solve, at least partially, this property expressiveness problem and we have pointed some solutions to it in~\cref{subsec:expressiveness_prop}.

From a practical standpoint, we have shown with the two examples presented in \cref{sec:experiments} that there are good indicators that our technique can scale to transformations of practical interest. We have shown in \cref{sec:implementation} that the complexities of path condition generation and property proving are, as expected, exponential. Still, we are confident that we have not exhausted the set of possible optimizations in our tool and that our implementation (using typed graph manipulations in T-Core) can be made to scale well for reasonably-sized model transformations. This remains to be proved for larger model transformations. In this direction, we are currently implementing the analysis of a UML-RT to Kiltera transformation~\cite{posse:10} which includes more than twice the number of rules in the industrial case study we present in this paper. For the UML-RT to Kiltera case study we are also including element attributes in the generation of path conditions and property proof.

Additionally, we have recently completed a propositional logic extension to our property language, which has already been used to express and prove meaningful properties in our industrial case study~\cite{gehan:13,conf/gg/SelimLCDO14}. This extension has been implemented in our tool, but its full impact in the theory of property proving, as explained in \cref{sec:verif_dsltrans_props}, is yet to be fully understood. A further topic of interest is that of negative DSLTrans constructs, where elements and associations of given types are prevented from being matched by a rule, and their inclusion in the property language.

%We are also exploring the issue of multiplicity in our verification technique, such as determining for a path condition how many times a particular rule has executed. This information may arise from recording how many times the rule was ``glued" to the path condition in the combination step. This could enable properties to be proved over the number of times a rule has executed.

% As seen in the results section, we believe our algorithms to scale to
% industrial-sized case studies. In future work, we will apply our technique to transformations and case studies such as structural relations in
% Simulink~\cite{feherflattening}, as well as those of interest
% to our partners in the automotive domain.

%\newpage

%Another point that needs to be further developed in our approach is the property
%language. In this paper we have concentrated on building the algorithms that
%allow symbolic execution construction and property proof, but have left the
%property language in a relatively basic state. For the time being our property
%language allows essentially to express what is expressible in transformation
%rules (including statements about multiples of instances of elements of the same
%type) and transitive containment connections at the \emph{Postcondition} part of
%he property.  We believe that the current property language can already be very
%useful to prove many relevant properties of practical transformations, but have
%not studied its full range yet. Regarding the possible extensions of our
%property language, inspiration can be drawn from several proposals in the
%literature~\cite{Narayanan:07,DBLP:journals/eceasst/CariouBBD09,DBLP:conf/icst/AsztalosLL10,DBLP:journals/ase/GuerraLWKKRSS13}.
%One of the natural extensions of our property language is the possibility
%to express conditions over the attributes of the elements in the properties,
%which for the time being we do not address. During the symbolic execution
%construction such conditions will have to be addressed symbolically, which adds
%an additional challenge to DSLTrans' symbolic execution construction. Negative
%links (associations, indirect links and backward links) are also part of our
%uture tasks, both at the level of the symbolic execution construction and of
%the property language.

### 2015-03-13 text/definitions.tex
previous version of the paper

%=======================================================

%<*typedgraphtext>

We will start by introducing the notion of typed graph. A typed graph is the essential object we will use throughout our mathematical development. Typed graphs will be used to formalise all the important graph-like structures we will present in this paper. A typed graph is a directed multigraph (a graph allowing multiple edges between two vertices) where vertices and edges are typed.

%</typedgraphtext>

%<*typedgraph>

A typed graph is a 6-tuple $\langle V,E,(s,t),\tau, VT, ET\rangle$ where: $V$ is a finite set of vertices; $E$ is a finite set of directed edges connecting the vertices $V$; $(s,t)$ is a pair of functions $s: E\rightarrow V$ and $t: E\rightarrow V$ that respectively provide the source and target vertices for each edge in the graph; function $\tau:V\cup E\rightarrow VT \cup ET$ is a typing function for the elements of $V$ and $E$, where $VT$ and $ET$ are disjoint finite sets of vertex and edge type identifiers and $\tau(v)\in VT$ if $v\in V$ and $\tau(e)\in ET$ if $e\in E$. Edges $e\in E$ are noted $v\xrightarrow{e} v'$ if $s(e)=v$ and $t(e)=v'$, or simply \emph{e} if the context is unambiguous. The set of all typed graphs is called $\textsc{Tg}$.

%</typedgraph>

%=======================================================

%<*typedgraphuniontext>

We now define how two typed graphs are united. A union of two typed graphs is trivially the set union of all the components of those two typed graphs. Note that we do not require the components of the two graphs to be disjoint, as in the following joint unions will be used to merge typed graphs.

%</typedgraphuniontext>

%<*typedgraphunion>
Let $\langle V,E,(s,t),\tau,VT,ET\rangle,\langle V',E',(s',t'),\tau',VT',ET'\rangle\in \textsc{Tg}$ be typed graphs, where $VT$ and $ET'$ are disjoint sets, as well as $VT'$ and $ET$. The typed graph union is the function $\sqcup :\textsc{Tg}\times \textsc{Tg}\rightarrow \textsc{Tg}$ defined as:
\begin{multline*}
\big\langle V,E,(s,t),\tau,VT,ET\big\rangle\;\sqcup\;\big\langle V',E',(s',t'),\tau',VT',ET'\big\rangle=\\\big\langle V\cup V', E\cup
E',(s\cup s', t\cup t'), \tau\cup \tau', VT\cup VT', ET\cup ET'\big\rangle
\end{multline*}
%</typedgraphunion>

%=======================================================

%<*typedgraphhomomorphismtext>

For the formal development of our technique, we are interested in relations between typed graphs that are structure-preserving, i.e. homomorphisms. Homomorphisms between typed graphs preserve not only structure, but also the types of vertices and edges that are mapped.

%</typedgraphhomomorphismtext>

%<*typedgraphhomomorphism>
Let $\langle V,E,st,\tau,VT,ET\rangle=g$ and $\langle V',E',st',\tau',VT',ET'\rangle=g'\in \textsc{Tg}$ be typed graphs. A typed graph homomorphism between $g$ and $g'$ is a function $f:V\rightarrow V'$ such that for all $v_1 \xrightarrow{e} v_2\in E$ we have that $f(v_1) \xrightarrow{e'} f(v_2)\in E'$, where $\tau(v_1)=\tau'(f(v_1))$, $\tau(v_2)=\tau'(f(v_2))$ and also $\tau(e)=\tau(e')$. The domain of $f$ is noted $Dom(f)$ and the co-domain of $f$ is noted $CoDom(f)$. When an \emph{injective} typed graph homomorphism $f$ exists between $g$ and $g'$ we write $g \stackrel{f}{\vartriangleleft} g'$, or simply $g \vartriangleleft g'$ when the context is unambiguous. When a \emph{surjective} typed graph homomorphism $f$ exists between typed graphs $g$ and $g'$ we write $g \stackrel{f}{\blacktriangleleft} g'$, or also simply $g \blacktriangleleft g'$ in an unambiguous context. 
%</typedgraphhomomorphism>

%=======================================================

%<*typedsubgraphtext>

We now define the useful notion of typed subgraph. As expected, a typed subgraph is simply a restriction of a typed graph to some of its vertices and edges.

%</typedsubgraphtext>

%<*typedsubgraph>
Let $\langle V,E,st,\tau,VT,ET\rangle=g,\langle V',E',st,\tau',VT',ET'\rangle=g'\in \textsc{Tg}$ be typed graphs. $g'$ is a typed subgraph of $g$, written $g'\sqsubseteq g$, iff $V'\subseteq V$, $E'\subseteq E$ and $\tau'=\tau |_{V'\cup E'}$.
%</typedsubgraph>

%=======================================================

%<*typedgraphisomorphismtext>

Two typed graphs are said to be isomorphic if they have exactly the same shape and related vertices and edges have the same type.

%</typedgraphisomorphismtext>

%<*typedgraphisomorphism>
Let $\langle V,E,st,\tau,VT,ET\rangle=g,\langle V',E',st',\tau',VT',ET'\rangle=g'\in \textsc{Tg}$ be typed graphs. We say that $g$ and $g'$ are isomorphic, written $g\cong g'$, if and only if there exists a bijective typed graph homomorphism $f:V\rightarrow V'$ such that $f^{-1}:V'\rightarrow V$ is a typed graph homomorphism.
%</typedgraphisomorphism>

%=======================================================

%<*notationformalbackground>
In order to simplify our notation, when the context is unambiguous we will abbreviate a typed graph $\langle V,E,st,\\\tau,VT, ET\rangle$ as as 4-tuple $\langle V,E,st,\tau\rangle$. Also, given a typed graph $g\in \textsc{Tg}$, will use the notation $Components(g)$ to describe the set of strongly connected typed graphs in $g$. Finally, we will use the notation $g|_{t}$ to refer to the restriction of graph $g$ to its subgraph containing only edges of type $t$.
%</notationformalbackground>

%=======================================================

%<*metamodel>

A metamodel is a 5-tuple $\langle V,E,st,\tau,\leq\rangle$ where $\langle V,E,st,\tau\rangle\in \textsc{Tg}$ is a typed graph, $(V,\leq)$ is a partial order and $\tau$ is a bijective typing function. Additionally we also have that: if $v\in V$ then $\tau(v)\in VT\times \{abstract,concrete\}$, where $VT$ is the set of vertex type names; if $e\in E$ then $\tau(e)\in ET\times \{containment,\\reference\}$, where $ET$ is a set of edge type names. The set of all metamodels is called $\textsc{Meta}$.

%</metamodel>

%<*metamodeltext>

A formal metamodel is a particular kind of typed graph where vertices represent classes and edges represent relations between those classes. A typed graph representing a metamodel has two special characteristics: on the one hand, the typing function for vertices and edges is bijective. This means that each type occurs only once in the metamodel, as is to be expected. On the other hand, a metamodel is equipped with a partial order between vertices. This partial order is used to model inheritance at the level of the metamodel's classes. Note that here we have overridden the co-domain of the typing function in the original typed graph presented in \cref{def:typed_graph} in order to allow distinguishing between \emph{abstract} and \emph{concrete} classes, as well as between \emph{containment} and \emph{reference} edges in our metamodels. For simplification purposes, we do not model association cardinalities in our formal notion of metamodel as cardinalities are not strictly necessary in our development.

%\levi{Cardinalities are not treated, meaning all relations can be n-to-m. It may be necessary to treat cardinalities.}

%</metamodeltext>

%<*metamodeltextappendix>

A formal metamodel is a particular kind of typed graph where vertices represent classes and edges represent relations between those classes. A typed graph representing a metamodel has two special characteristics: on the one hand, the typing function for vertices and edges is bijective. This means that each type occurs only once in the metamodel, as is to be expected. On the other hand, a metamodel is equipped with a partial order between vertices. This partial order is used to model specialization at the level of the metamodel's classes. Note that here we have overridden the co-domain of the typing function in the original typed graph presented in \cref{def:typed_graph_appendix} in order to allow distinguishing between \emph{abstract} and \emph{concrete} classes, as well as between \emph{containment} and \emph{reference} edges in our metamodels. For simplification purposes, we do not model association cardinalities in our formal notion of metamodel as cardinalities are not strictly necessary in our development.

%\levi{Cardinalities are not treated, meaning all relations can be n-to-m. It may be necessary to treat cardinalities.}

%</metamodeltextappendix>

%=======================================================

%<*expandedmetamodel>

Let $mm = \langle V,E,st,\tau,\leq\rangle\in \textsc{Meta}$ be a metamodel. The expansion of $mm$, noted $mm^{\star}$, is a typed graph $\langle V',E',st',\tau'\rangle\in \textsc{Tg}$ built as follows:
\begin{itemize}
  \item $V' = V\setminus \{v\in V\,|\,\tau(v)=(\cdot\footnote{In our mathematical development we use a `dot' notation to represent that we do not care about the value of a particular variable in a given context.},abstract)\}$;
  \item $v_1\xrightarrow{e}v_2\in E'$ if $v_1\xrightarrow{e}v_2\in E$ and $\tau(v_1)=(\cdot,concrete)$ and $\tau(v_2)=(\cdot,concrete)$;
  \item if $v_1\xrightarrow{e}v_2\in E$ we have that $v'_1\xrightarrow{e'}v'_2\in E'$, where $v'_1\leq v_1$, $v'_2\leq v_2$ and $\tau'(e') = \tau(e)$;
  \item for all $v\in V'$ and $e\in E'$ we have that $\tau'(v)=\tau(v)$ and that $\tau'(e)=\tau(e)$.
%   \item $(v,(tv,concrete))\in \tau'$ if $(v,(tv,concrete))\in \tau$ (where $tv$ is a vertex type);
%   \item $(v_1\xrightarrow{e}v_2,(te,concrete))\in \tau'$ if $(v_1\xrightarrow{e}v_2,(te,concrete))\in \tau$ ($te$ ia an edge type).
\end{itemize}
%The set of all expanded metamodels in called $META^{\star}$.

%</expandedmetamodel>

%<*expandedmetamodeltext>

An expanded metamodel is an auxiliary construct where all the relations between types of a metamodel are made explicit, rather than remaining implicit in the specialization hierarchy. It is built by adding to the original metamodel typed graph a relation of type $t$ between two classes of the metamodel, whenever those classes specialize two classes that are also related by a relation of type $t$. Abstract classes and their relations do not carry over to the expanded metamodel. Expanded metamodels will be used in the subsequent text to facilitate formal the treatment of any structure involving polymorphism.

% Definition~\ref{def:metamodel} formally states that a metamodel corresponds to a graph of typed elements where only one element for each type is represented. This definition is then used to state definition~\ref{def:pattern}, where a typed graph is combined with its metamodel.

%</expandedmetamodeltext>

%=======================================================

%<*metamodelinstance>

An instance of a metamodel $mm = \langle V',E',st',\tau',\leq\rangle\in \textsc{Meta}$ is a typed graph $\langle V,E,st,\tau\rangle\in \textsc{Tg}$, where the co-domain of $\tau$ equals the co-domain of $\tau'$. Also, there is a typed graph homomorphism $f:V\rightarrow V'$ from $\langle V,E,st,\tau\rangle$ to the expanded metamodel $mm^{\star}$ and the graph $\big\langle V,\{e\in E\,|\,\tau(e)=(\cdot,containment)\}\big\rangle$ is acyclic. The set of all instances for a metamodel $mm$ is called $\textsc{Instance}^{mm}$.

%</metamodelinstance>

%<*metamodelinstancetext>

A metamodel instance is a useful intermediate formal notion that lies between metamodel and model. The injective typed graph homomorphism between a metamodel instance and metamodel models multiple ``instances'' of objects and relations being typed by one single class or relation of the metamodel. Metamodel instances do not allow cyclic containment relations, as enforced by EMF.

% It requires the acyclicity constraints as that is required through models, transformation rules, path conditions and properties. It can be used to define transformation rules and can be further refined to define models.

%</metamodelinstancetext>

%=======================================================

%<*instanceclosure>

The containment transitive closure of a metamodel instance $\langle V',E',st',\tau'\rangle\in \textsc{Instance}^{mm}$ is a typed graph $\langle V,E,st,\tau\rangle$ where we have that $V=V'$, $\tau' \supseteq \tau$ and $\tau$'s co-domain is the union of the co-domain of $\tau'$ and the set $\{indirect\}$. We also have that $E' = E\cup E^*_c$ where $E^*_c$ is the transitive closure of the set $\big\{v\xrightarrow{e}v'\,|\,\tau(v\xrightarrow{e}v') = (\cdot,containment)\big\}$ and if $e\in E\setminus E'$ then $\tau(e) = indirect$.  We denote $mi^{*}$ the containment closure of a metamodel instance $mi\in \textsc{Instance}^{mm}$.

%</instanceclosure>

%<*instanceclosuretext>

Given a metamodel instance, its containment transitive closure includes, besides the original graph, all the edges belonging to the transitive closure of containment links in that metamodel instance. The transitive edges are typed as \emph{indirect}. In the definitions that follow we will use the $*$ notation, as in \cref{def:instance_closure}, to denote the containment transitive closure of structures that directly or indirectly include metamodel instances. For example, $tg^{*}$ would represent the containment transitive closure of typed graph $tg$ wherever containment edges are found in the graph. Note that the $*$ notation is different from the $\star$ notation, introduced in \cref{def:expanded_metamodel} for an expanded metamodel.

% It requires the acyclicity constraints as that is required through models, transformation rules, path conditions and properties. It can be used to define transformation rules and can be further refined to define models.

%</instanceclosuretext>

%<*instanceclosuretextappendix>

Given a metamodel instance, its containment transitive closure includes, besides the original graph, all the edges belonging to the transitive closure of containment links in that metamodel instance. The transitive edges are typed as \emph{indirect}. In the definitions that follow we will use the $*$ notation, as in \cref{def:instance_closure_appendix}, to denote the containment transitive closure of structures that directly or indirectly include metamodel instances. For example, $tg^{*}$ would represent the containment transitive closure of typed graph $tg$ wherever containment edges are found in the graph. Note that the $*$ notation is different from the $\star$ notation, introduced in \cref{def:expanded_metamodel_appendix} for an expanded metamodel.

% It requires the acyclicity constraints as that is required through models, transformation rules, path conditions and properties. It can be used to define transformation rules and can be further refined to define models.

%</instanceclosuretextappendix>

%=======================================================

%<*model>

A model of a  metamodel $mm = \langle V',E',st',\tau',\leq\rangle\in \textsc{Meta}$ is a metamodel instance $\langle V,E,st,\tau\rangle\in \textsc{Instance}^{mm}$, such that: there exists an injective typed graph homomorphism $f:V\rightarrow V'$ from $\langle V,E,st,\tau\rangle$ to metamodel $mm^{\star}$ where, if $b\in V$ and there exists an edge $a'\xrightarrow{e'} f(b)\in E'$ where $\tau(e') = (t,containment)$, then we also have that $a\xrightarrow{e} b\in E$, $f(a) = a'$ and $\tau(e)=t$. The set of all models for a metamodel $mm$ is called $\textsc{Model}^{mm}$.

%</model>

%<*modeltext>

A model, as per \cref{def:model}, is a metamodel instance where all the containment relations are respected. This means that if an object that is contained (as per the metamodel) exists in the model, then the model will also contain an instance of that containment relation together with an object that contains it. Two models can be observed in \cref{fig:transformationexample}, which are respectively instances of the \emph{Organization language} and \emph{Gender language} found in \cref{fig:squadmetamodel}.\\
Note that the containment constraint does not necessarily lead to infinite models in the case of containment relations with the same source and target classes. In fact, if the cardinality of the target class is allowed to be zero, then it is not necessary that the containment relation is instantiated. For example, this is the case for the containment relation \emph{supervise} in the metamodel of~\cref{fig:OrganizationLanguage}.

%</modeltext>

%<*modeltextappendix>
A model, as per \cref{def:model_appendix}, is a metamodel instance where all the containment relations are respected. This means that if an object having a containment relation exists in the model, then the model will also contain an instance of that containment relation together with a contained object. Note that this constraint does not necessarily lead to infinite models in the case of containment relations with the same source and target classes. In fact, if the cardinality of the target class is allowed to be zero, then it is not necessary that the containment relation is instantiated. This is for example the case of the containment relation \emph{supervise} in the metamodel of~\cref{fig:OrganizationLanguage}.

% \begin{figure}[h!] \centering \includegraphics[scale=.5]{./figures/model_instance_example.pdf}
% 	\caption{An Example of a Model, instance of a Metamodel}
% 	\label{fig:model_instance_example}
% \end{figure}

%</modeltextappendix>

%=======================================================

%<*inputoutputmodel>

An input-output model is a 6-tuple $\big\langle V,E,(s,t),\tau,Input,\\Output\big\rangle$, where: $Input=\langle V',E',st',\tau'\rangle\in \textsc{Instance}^{sr}$ is a model; $Output=\langle V'',E'',st'',\tau''\rangle\in \textsc{Instance}^{tg}$ is a metamodel instance; $Input$ and $Output$ are disjoint. Additionally we have that $V=V'\cup V''$, $E\subseteq E'\cup E''$ and $\tau\subseteq \tau'\cup \tau''$, where the co-domain of $\tau$ is the union of the co-domains of $\tau'$ and $\tau''$ and the set $\{trace\}$. An edge $e \in E\setminus E'\cup E''$ is called a \emph{traceability link} and is such that $s(e)\in V''$, $t(e)\in V'$ and $\tau(e)=trace$. The set of all match-apply patterns for a source metamodel $sr$ and a target metamodel $tg$ is called $\textsc{Iom}^{sr}_{tg}$.

%</inputoutputmodel>

%<*inputoutputmodeltext>

% \levi{A match-apply model is the union of an input and output models, plus potential traceability links between those models. Traceability links are now typed using the typing function $\tau$. A traceability link it typed with type ``trace''. This makes the definitions shorter and more compact, which was mentioned by one of the reviewers of the paper. However, the fact that a match-apply model is now made up of models makes it that the input and the output of the transformation are always well-formed, meaning that the containment relation is always respected. It may be that a transformation does not always produce a model respecting the containment relation, which is a problem. This can be solved however if we only consider that well-formed transformations actually have semantics.}

An input-output model is an object we will use when defining the semantics of a DSLTrans model transformation in \cref{sec:DSLTrans_formal}. It is composed of two metamodel instances, one called the \emph{input} and the other one the \emph{output}. An input-output model allows the representation of intermediate operational states during the execution of a model transformation. It may include a particular type of edges called \emph{traceability links}, for keeping a history of which elements in the output model originated from which elements in the input model.

%</inputoutputmodeltext>

%<*inputoutputmodeltextappendix>

An input-output model is an object we will use when defining the semantics of a DSLTrans model transformation in \cref{sec:DSLTrans_formal}. It is composed of two metamodel instances, one called the \emph{input} and the other one the \emph{output}. An input-output model allows the representation of intermediate operational states during the execution of a model transformation. It may include a particular type of edges called \emph{traceability links}, for keeping a history of which elements in the output model originated from which elements in the input model.

%</inputoutputmodeltextappendix>

%=======================================================

%<*metamodelpattern>

A pattern of a metamodel $mm\in \textsc{Meta}$ is an instance of $mm$. Given a metamodel pattern $\langle V',E',st',\tau'\rangle\in \textsc{Instance}^{mm}$ we have that $\langle V,E,st,\tau\rangle$ is an \emph{indirect} pattern if $V=V'$, $E'\supseteq E$ and the co-domain of $\tau$ is the union of the co-domains of $\tau'$ and the set $\{indirect\}$. Also, if $e\in E\setminus E'$, then we have that $\tau(e)=indirect$. Given a metamodel $mm$, the set of all metamodel patterns for $mm$ is called $\textsc{Pattern}^{mm}$. The set of all indirect metamodel patterns for $mm$ is called $\textsc{Ipattern}^{mm}$.

%</metamodelpattern>

%<*metamodelpatterntext>

Metamodel patterns are introduced in \cref{def:metamodel_pattern} as an intermediate notion, formally equal to metamodel instances. An \emph{indirect} metamodel pattern is a metamodel pattern that includes edges typed as \emph{indirect}. Both structures will be used as building blocks in the construction of transformation-related structures in the upcoming text.

%</metamodelpatterntext>

%<*metamodelpatterntextappendix>

Metamodel patterns are introduced in \cref{def:metamodel_pattern_appendix} as an intermediate notion, formally equal to metamodel instances. An \emph{indirect} metamodel pattern is a metamodel pattern that includes edges typed as \emph{indirect}. Both structures will be used as building blocks in the construction of transformation-related structures in the upcoming text.

%</metamodelpatterntextappendix>

%=======================================================

%<*transformationrule>

A transformation rule is a 6-tuple $\big\langle V,E,(s,t),\tau, Match,\\Apply\big\rangle$, where: $Match=\langle V',E',st',\tau'\rangle\in \textsc{Ipattern}^{sr}$ such that: $Match\neq \epsilon$\footnote{We use the simplified $\epsilon$ notation to denote empty n-tuples structures.} is an non-empty indirect metamodel pattern; $Apply=\langle V'',E'',st'',\tau''\rangle\in \textsc{Pattern}^{tg}$ such that $Apply\neq\epsilon$ is a metamodel pattern; $Match$ and $Apply$ are disjoint. We also have that $V=V'\cup V''$, $E\subseteq E'\cup E''$ and $\tau\subseteq \tau'\cup \tau''$, where the co-domain of $\tau$ is the union of the co-domains of $\tau'$ and $\tau''$ and the set $\{trace\}$. An edge $e \in E\setminus E'\cup E''$ is called a \emph{backward link} and is such that $s(e)\in V''$, $t(e)\in V'$ and $\tau(e)=trace$. We additionally impose that there always exists a $v_1\in V''$ in the Apply part of the rule such that $\nexists e\,.\, v_1\xrightarrow{e} v_2$ and $\tau(e)=trace$, or that $E''$ is not empty. The set of all transformation rules for a source metamodel $sr$ and a target metamodel $tg$ is called $\textsc{Rule}^{sr}_{tg}$.

%</transformationrule>

%<*transformationruletext>

A transformation rule is the elemental block of a model transformation. Several transformation rules can be observed in the Police Station transformation in \cref{fig:dsltransformation}. A formal transformation rule includes a non-empty match pattern and a non-empty apply pattern (also known in the model transformation literature as a rule's \emph{left hand side} and \emph{right hand side}). The apply pattern of a rule always contains at least one apply element that is not connected to a backward link or an edge, meaning in practice that a rule will always produce something and not only match. A match pattern can include indirect links that are used to transitively match containment relations in a model. An apply pattern does not include indirect links as it is used only for the construction of parts of instances of a metamodel. A transformation rule includes backward links, as informally introduced in \cref{subsec:DSLTrans_constructs}. Backward links are formally typed as \emph{trace}.%This eases modeling matching rules on models  by the use of typed graph homomorphisms.

%</transformationruletext>

%=======================================================

%<*backmatchtransformationrule>

Let $rl = \big\langle V,E,st,\tau, Match,Apply\big\rangle$ be a transformation rule where $Match =\langle V_m, E_m, st_m, \tau_m\rangle$. We define $rl$'s matcher, noted $\Vert rl\Vert$, as the transformation rule $\big\langle V',E',st',\tau',Match,\\Apply'\big\rangle\sqsubseteq rl$ where $v_1\xrightarrow{e} v_2\in E'$ if and only if $v_1,v_2\in Match$ or $\tau(e)=trace$ and $V'=V_m\cup \big\{v_1\,|\,v_1\xrightarrow{e} v_2\in E \land \tau(e)=trace\big\}$.

%</backmatchtransformationrule>

%<*backmatchtransformationruletext>

%'transformation rule cref in text
introduces the notion of matcher for a transformation rule which consists solely of the match pattern of a rule and its backward links, if any. The matcher of a rule constitutes the complete pattern that a DSLTrans rule attempts to match over a input-output model during rule execution. Traceability links between input and output model elements generated during transformation execution are matched by transformation rules' backward links, as informally explained in \cref{subsec:DSLTrans_constructs}.

%</backmatchtransformationruletext>

%=======================================================

%<*expandedtransformationrule>

Let $rl = \big\langle V,E,st,\tau,Match,Apply\rangle \in \textsc{Rule}^{sr}_{tg}$ be a transformation rule where $Match = \langle V',E',st',\tau'\rangle$ and also we have that $sr =\langle V'',E'',st'',\tau'',\leq \rangle$. The expansion of $rl$, noted $rl^{\star}$ is a set of transformation rules built as follows:
\begin{itemize}
  \item $rl\in rl^{\star}$;
  \item $\big\langle V,E,st,\tau',Match,Apply\rangle\in rl^{\star}$ iff for all $v\in V'$ we have that $\tau'(v)\leq \tau(v)$.
\end{itemize}

%</expandedtransformationrule>

%<*expandedtransformationruletext>

The expansion of a transformation rule is a set of transformation rules. Each rule in that set includes a possible replacement of each of the classes in the match part of the original rule by one of its subtypes. Expanded transformation rules will be important such that polymorphism is correctly handled in the developments that follow.

%\levi{A transformation rule expansion simply takes a rule and builds all possible rules where a rule element is replaced by one of its subtypes.}

%</expandedtransformationruletext>

%=======================================================

%<*layertransformation>

A layer is a finite set of transformation rules $l\subseteq \textsc{Rule}^{sr}_{tg}$. The set of all layers for a source metamodel $sr$ and a target metamodel $tg$  is called $\textsc{Layer}^{sr}_{tg}$. A model transformation is a finite list of layers denoted $[l_{1}::l_{2}::\ldots::l_{n}]$ where $l_{k}\in \textsc{Layer}^{sr}_{tg}$ and $1\leq k \leq n$, $n\in \mathbb{N}$. We also impose that for any pair of rules $rl_1,rl_2\in \bigcup_{1\leq k\leq n}l_k$, if $\Vert rl_1\Vert \cong rl$ and $rl\sqsubseteq \Vert rl_2\Vert $ then $rl_2$ appears in a layer later than $rl_1$ and the apply parts of $rl_1$ and $rl_2$ are not isomorphic. The set of all transformations for a source metamodel $s$ and a target metamodel $t$ is called $\textsc{Transf}^{sr}_{tr}$.

% The expansion of a layer $l\subseteq RULE^{sr}_{tg}$, noted $l^{\star}$, is such that that if
% $rl\in l$ then $rl'\in l$ if $rl'\in rl^{\star}$. This means that if a
% transformation rule is present is the layer, its expansion is also part of the
% layer.

%</layertransformation>

%<*layertransformationtext>

\cref{def:layer_transformation} formalises the abstract syntax of a model transformation, introduced in \cref{sec:dsltrans}. An example of a model transformation can be observed in \cref{fig:dsltransformation}, the Police Station transformation. As expected, a formal DSLTrans transformation is composed of a sequence of layers where each layer is composed of a set of rules. The last condition of \cref{def:layer_transformation} imposes that, for any two pair of rules in the transformation, the matcher of the second rule never partially or totally subsumes (or contains) the matcher of the first rule, unless the second rule is in a subsequent layer and produces something more than the first rule. This condition avoids situations where the execution of a rule in a DSLTrans model transformation necessarily implies the execution of another rule (except for when rules having backward links necessarily execute because all their dependencies were created during the execution of previous layers).

%\levi{At the syntactic level we have not included the expansion of a layer in the transformation. This is done at the semantic level.}

%</layertransformationtext>

%<*layertransformationtextappendix>

\cref{def:layer_transformation_appendix} formalises the abstract syntax of a model transformation, introduced at the beginning of this section. An example of a model transformation can be observed in \cref{fig:dsltransformation}, the Police Station transformation. As expected, a formal DSLTrans transformation is composed of a sequence of layers where each layer is composed of a set of rules. The last condition of \cref{def:layer_transformation_appendix} imposes that, for any two pair of rules in the transformation, the matcher of the second rule never partially or totally subsumes (or contains) the matcher of the first rule, unless the second rule is in a subsequent layer and produces something more than the first rule. This condition avoids situations where the execution of a rule in a DSLTrans model transformation necessarily implies the execution of another rule (except for when rules having backward links necessarily execute because all their dependencies were created during the execution of previous layers).

%\levi{At the syntactic level we have not included the expansion of a layer in the transformation. This is done at the semantic level.}

%</layertransformationtextappendix>

%=======================================================

%<*notationtext>

% In order to achieve a uniform notation throughout the text that follows, when the context is unambiguous we refer to transformation rules (\cref{def:transformation_rule}) as normal typed graphs. We also naturally extend to transformation rules the typed graph operations introduced in \cref{sec:formal_background}. Finally, given a structure such as transformation rule $rl = \langle V,E,st,\tau,Match,Apply\rangle$, we will refer to the structure's components by using the component's name followed by the variable that holds the structure in between parenthesis. For example, we will write $V(rl)$ to designate the $V$ component of $rl$ or $Apply(rl)$ to designate $rl$'s $Apply$ component.

We naturally extend to transformation rules (\cref{def:transformation_rule}) and transformation executions (\cref{def:modeltransformation}) the typed graph operations introduced in \cref{sec:formal_background}.  Also, given a structure such as transformation rule $rl = \langle V,E,st,\tau,Match,\\Apply\rangle$, we will refer to the structure's components by using the component's name followed by the variable that holds the structure in between parenthesis. For example, we will write $V(rl)$ to designate the $V$ component of $rl$ or $Apply(rl)$ to designate $rl$'s $Apply$ component.

% We will also use the natural extensions of the notions of typed graph union `$\sqcup$' (\cref{def:typed_graph_union}), typed graph homomorphism `$\vartriangleleft, \blacktriangleleft$' (\cref{def:typed_graph_homomorphism}), typed subgraph `$\sqsubseteq$' (\cref{def:typedsubgraph}) and typed graph isomorphism `$\cong$' (\cref{def:typed_graph_isomorphism}) to transformation rules.

%</notationtext>

%<*notationtextappendix>

We naturally extend to input-output models~(\cref{def:input_output_model_appendix}), transformation rules (\cref{def:transformation_rule_appendix}) and transformation executions (\cref{def:modeltransformation_appendix}) the typed graph operations introduced in \cref{sec:formal_background}. Also, given a structure such as transformation rule $rl = \langle V,E,st,\tau,Match,Apply\rangle$, we will refer to the structure's components by using the component's name followed by the variable that holds the structure in between parenthesis. For example, we will write $V(rl)$ to designate the $V$ component of $rl$ or $Apply(rl)$ to designate $rl$'s $Apply$ component. We also use a simplified notation to refer to the components of the input/output or match/apply typed graphs of input-output models and transformation rules. We refer to these structures' vertices that belong to the match part of the graph as $Match(V)$, its edges as $Match(E)$ and so on for pair $st$ and function $\tau$. In a similar fashion we use the notation $Apply(V)$, $Apply(E)$, etc. to refer to the transformation rule's components that belong to the apply part of the graph.

% In order to achieve a uniform and understandable notation throughout the text that follows we will often abbreviate input-output models~(\cref{def:input_output_model_appendix}) and transformation rules (\cref{def:transformation_rule_appendix}) to their typed graph components. We will also use a simplified notation to refer to the components of the input/output or match/apply typed graphs of those structures. For example, we will abbreviate transformation rule $\big\langle V,E,st,\tau,Match,\\Apply\big\rangle\in \textsc{Rule}^{sr}_{tg}$ to only its typed graph $\big\langle V,E,st,\tau\rangle$. In this case we will refer to the transformation rule's vertices that belong to the match part of the graph as $Match(V)$, its edges as $Match(E)$ and so on for pair $st$ and function $\tau$. In a similar fashion we use the notation $Apply(V)$, $Apply(E)$, etc. to refer to the transformation rule's components that belong to the apply part of the graph. We will also use the natural extensions of the notions of typed graph union `$\sqcup$' (\cref{def:typed_graph_union}), typed graph homomorphism `$\vartriangleleft, \blacktriangleleft$' (\cref{def:typed_graph_homomorphism}), typed subgraph `$\sqsubseteq$' (\cref{def:typedsubgraph}) and typed graph isomorphism `$\cong$' (\cref{def:typed_graph_isomorphism}) to input-output models and transformation rules.

%</notationtextappendix>

%=======================================================

%<*modeltransformation>

Let  $tr\in \textsc{Transf}^{sr}_{tg}$ be a transformation and $input \in \textsc{Model}^{sr}$ be a model. Assume we also have that: $$\langle V,E,st,\tau,input,\epsilon\rangle,tr \xrightarrow{trstep} \langle V',E',st',\tau',input,output\rangle$$
A model transformation execution is the input-output model $\langle V',E',st',\tau',input,output\rangle\in \textsc{Iom}^{sr}_{tg}$, where $output\in \textsc{Iom}^{sr}_{tg}$ is an input-output model. The set of all model transformation executions for transformation $tr$ is written $\textsc{Exec}(tr)$. A model transformation with an empty input model is noted $\epsilon_{ex}$. Note that relation $trstep$ is formally defined in \cref{sec:DSLTrans_formal_appendix}.

%If $output\in\textsc{Model}^{sr}_{tg}$ we say the model transformation execution is \emph{final}.

%</modeltransformation>

%<*modeltransformationappendix>

Let  $tr\in \textsc{Transf}^{sr}_{tg}$ be a transformation and $input \in \textsc{Model}^{sr}$ be a model. Assume we also have that: $$\langle V,E,st,\tau,input,\epsilon\rangle,tr \xrightarrow{trstep} \langle V',E',st',\tau',input,output\rangle$$
A model transformation execution is the input-output model $\langle V',E',st',\tau',input,output\rangle\in \textsc{Iom}^{sr}_{tg}$, where $output\in \textsc{Instance}^{tg}$ is an instance of the output metamodel. The set of all model transformation executions for transformation $tr$ is written $\textsc{Exec}(tr)$. A model transformation with an empty input model is noted $\epsilon_{ex}$.

%If $output\in\textsc{Model}^{sr}_{tg}$ we say the model transformation execution is \emph{final}.

%</modeltransformationappendix>

%<*modeltransformationtext>

Finally, as stated in \cref{def:modeltransformation}, we consider a model transformation execution to be the input-output model (IOM) resulting from executing a set of rules on a starting IOM. This starting IOM includes the transformation's input in its input part and has an empty output part. The starting IOM represents the first step of the transformation when no rule has been executed yet. A transformation execution results from executing all the rules in a DSLTrans model transformation.

% It holds the original input model, the output model, and the traceability links between their elements that are generated during the transformation's execution.

% We will consider input-output model as the transformation result is interesting for the purposes of this paper, in practice a transformation developer is normally only interested in the output component of the input-output model in a transformation execution. Model transformations that are \emph{final} are the ones enforce that containment relations are respected in the transformation's output.

%</modeltransformationtext>

%<*modeltransformationtextappendix>

Finally, as stated in \cref{def:modeltransformation_appendix}, we consider a model transformation execution to be the input-output model (IOM) resulting from executing a set of rules on a starting IOM. This starting IOM includes the transformation's input in its input part and has an empty output part. The starting IOM represents the first step of the transformation when no rule has been executed yet. A transformation execution results from executing all the rules in a DSLTrans model transformation.
% It holds the original input model, the output model, and the traceability links between their elements that are generated during the transformation's execution. 
% Although considering the complete input-output model as the transformation result is interesting for the purposes of this paper, in practice a transformation developer is normally only interested in the output component of the input-output model in a transformation execution. 
% 
% Note that, because in \cref{def:modeltransformation_appendix} $output$ is a formal model, we enforce that containment relations are respected in the transformation's output.

%</modeltransformationtextappendix>

%=======================================================

%<*pcdef>
A path condition is a 7-tuple $\big\langle V,E,(s,t),\tau, Match,Apply,\\Rulecop\big\rangle$, where: $Match=\langle V',E',st',\tau'\rangle\in \textsc{Ipattern}^{sr}$ is an indirect pattern; $Apply=\langle V'',E'',st'',\tau''\rangle\in \textsc{Pattern}^{tg}$ is a pattern; $Match$ and $Apply$ are disjoint graphs. We also have that $V=V'\cup V''$, $E\subseteq E'\cup E''$ and $\tau\subseteq \tau'\cup \tau''$ where the co-domain of $\tau$ is the union of the co-domains of $\tau'$ and $\tau''$ and the set $\{trace\}$. An edge $e \in E\setminus E'\cup E''$, called a \emph{symbolic traceability link}, is such that $s(e)\in V''$ and $t(e)\in V'$ and $\tau(e)=trace$. Finally, the $Rulecop$ component in the 7-tuple contains the set of rule copies used in the construction of the path condition, where each rule copy is a subgraph of $\big\langle V,E,(s,t),\tau\big\rangle$. The set of all path conditions for a source metamodel $sr$ and a target metamodel $tg$ is called $\textsc{Pathcond}^{sr}_{tg}$ and the empty path condition is noted $\epsilon_{pc}$.
%</pcdef>

%<*pc>

Similarly to a transformation rule (see~\cref{def:transformation_rule}), a path condition is also a typed graph with a match and an apply part. As mentioned before, a path condition contains a combination of rules where \emph{symbolic traceability links} represent the concrete traceability links of a transformation execution (see \cref{def:modeltransformation}). The path condition structure also contains a $Rulcop$ set that allows identifying individually all the copies of rules that were used when building the path condition's typed graph. Note that we refer to \emph{copies} of rules as, despite the fact that a path condition normally only contains one copy of each rule, in certain situations a rule may be used multiple times in the construction of a path condition. This will be explained further ahead in this section.

\paragraph{\textbf{Notation:}}
Given a path condition $pc = \big\langle V,E,st,\tau,\\Match,Apply,Rulecop\big\rangle\in \textsc{Pathcond}^{sr}_{tg}$ we refer to the set of transformation rules in $pc$ identified by the $Rulecop$ relation as $Rulecop(pc)$. Also, because a path condition is a particular kind of a typed graph, we naturally extend the basic notation of operators and homomorphisms on typed graphs defined in \cref{sec:formal_background} to path conditions.

%</pc>

%<*pcappendix>

Similarly to a transformation rule (see~\cref{def:transformation_rule_appendix}), a path condition is also a typed graph with a match and an apply part. As mentioned before, a path condition contains a combination of rules where \emph{symbolic traceability links} represent the concrete traceability links of a transformation execution (see \cref{def:modeltransformation_appendix}). The path condition structure also contains a $Rulcop$ set that allows identifying individually all the copies of rules that were used when building the path condition's typed graph. Note that we refer to \emph{copies} of rules as, despite the fact that a path condition normally only contains one copy of each rule, in certain situations a rule may be used multiple times in the construction of a path condition. This will be explained further ahead in this section.

\paragraph{\textbf{Notation:}}
Given a path condition $pc = \big\langle V,E,st,\tau,\\Match,Apply,Rule\big\rangle\in \textsc{Pathcond}^{sr}_{tg}$ we refer to the set of transformation rules in $pc$ identified by the $Rule$ relation as $Rule(pc)$. Also, because a path condition is particular kind of a typed graph, we naturally extend the basic notation of operators and homomorphisms on typed graphs defined in \cref{sec:formal_background} to path conditions.

%</pcappendix>

%=======================================================

%<*pccfrc>

Let $pc = \langle V',E',st',\tau',Match',Apply',Rulecop'\rangle\in\\
\textsc{Pathcond}^{sr}_{tg}$ be a path condition and $rl = \langle
V'',E'',st'',\tau'',\\Match'',Apply''\rangle\in \textsc{Rule}^{sr}_{tg}$ be a transformation rule, where their respective typed graphs can be joint. The union of $pc$ with $rl$ is built using the operator $\stackrel{trace}{\sqcup}:
\textsc{Pathcond}^{sr}_{tg}\times \textsc{Rule}^{sr}_{tg}\rightarrow
\textsc{Pathcond}^{sr}_{tg}$, as follows:

$$pc\stackrel{trace}{\sqcup}rl = \langle V,E,st,\tau,Match,Apply,Rulecop\rangle$$

where we have that $V=V'\cup V''$, $E'\cup E''\subseteq E$, $st'\cup
st''\subseteq st$, $\tau'\cup \tau''\subseteq \tau$ and if $v_1\xrightarrow{e}
v_2\in E\setminus E'\cup E''$ then we have that $v_1\in Apply(V'')$, $v_1\notin Apply(V')$, $v_2\in Match(V'')$ and also that $\tau'(e)=trace$. Additionally, $Match = Match'\sqcup Match''$ and $Apply = Apply'\sqcup Apply''$. Finally, we have that: $Rulecop = Rulecop'\cup rl$.

% $$pc = \bigsqcup_{rl\in RC} trace(rl_{pathcond})$$
% 
% where $trace:\textsc{Pathcond}\rightarrow \textsc{Pathcond}$ is such that
% $trace(\langle V,E,st,\tau\rangle) = \langle V,E',st',\tau'\rangle$ and we have
% that: $E\subseteq E'$; $st\subseteq st'$; $\tau \subseteq \tau'$; if $\neg
% (v_1\xrightarrow{e'} v_2\in E)\;\land\;\tau(e')=backward$ then we have that if
% $v_1\in Apply(V)$ and $v_2\in Match(V)$ then $v_1\xrightarrow{e} v_2\in
% E'\setminus E$ and $\tau'(e)=trace$. Also, $rl_{pathcond}$ is rule $rl$ typed as
% a path condition by adding to the $rl$ tuple a $Rule$ function that identifies
% each vertex in $V$ as belonging to $rl$. Finally, we have used the
% $\sqcup$ notation to denote the extension of the typed graph union to the
% component-wise of tuples that represent path conditions.

%</pccfrc>

%<*pccfrctext>
\cref{def:combine_pc_with_rule} shows the formal definition of combining a path condition with a rule. When a path condition is combined with a rule their typed graphs are united. Additionally, symbolic traceability links will be built at this time between the newly added apply elements of the rule and all of the rule's match elements. As a reminder, the link creation algorithm and examples have been introduced in \cref{subsubsec:traceability}.\\

Note that the fact that the graphs are potentially joint allows us to overlap a rule with the path condition by anchoring the rule on traceability links shared by the path condition and the rule graph. In the mathematical development that follows we will often refer to the joint parts of two or more typed graphs using the term ``glue''.
%</pccfrctext>

%<*pccfrctextappendix>
\cref{def:combine_pc_with_rule_appendix} shows the formal definition of combining a path condition with a rule. When a path condition is combined with a rule their typed graphs are united. Additionally, symbolic traceability links will be built at this time between the newly added apply elements of the rule and all of the rule's match elements.

Note that the fact that the graphs are potentially joint allows us to overlap a rule with the path condition by anchoring the rule on traceability links shared by the path condition and the rule graph. In the mathematical development that follows we will often refer to the joint parts of two or more typed graphs using the term ``glue''.
%</pccfrctextappendix>

%=======================================================

%<*nodeps>

The combination of a path condition $pc$ and a rule $rl$, when $rl$ has no dependencies, is described by the relation $\stackrel{combine}{\rightarrow}\subseteq \textsc{Pathcond}^{sr}_{tg} \times \mathcal{P}(\textsc{Pathcond}^{sr}_{tg}) \times \textsc{Rule}^{sr}_{tg} \times \mathcal{P}(\textsc{Pathcond}^{sr}_{tg})$, defined as follows:

% $$\frac{\begin{array}{ll}&
% pc\in PC\;,\;\langle pc,rl\rangle \xrightarrow{combine} pc'\;,\;PC\setminus\{pc\},rl,\rangle \xrightarrow{combstep} PC'
% \end{array}}
% {\langle PC,rules\rangle \xrightarrow{rulestep} PC'}$$

$$\frac{\begin{array}{ll}rl=\langle V,E,st,\tau, Match,Apply\big\rangle\;,\;\nexists e\in E\,.\,\tau'(e)=trace
\end{array}}
{\langle pc,AC,rl\rangle \xrightarrow{combine} AC\;\;\cup\;\; \bigcup_{pc'\in AC} pc'\stackrel{trace}{\sqcup} rl}$$
%</nodeps>

%<*nodepstext>
Relation $\stackrel{combine}{\rightarrow}$ in \cref{def:rule_comb_no_dependencies} models the operational combination step shown in \cref{fig:layers_pc2} (the vertical black arrows between boxes). The relation has three input arguments: the first argument is the original path condition from the previous layer (shown as the topmost box in \cref{fig:layers_pc2} with label $PC$); the second argument is the set of path conditions accumulated thus far by combining other rules in the current layer with the original path condition; and the third argument is the rule from the current layer now being combined. The fourth argument of the relation, the relation's output, is the new set of path conditions resulting from this combination.

Briefly, the equation in \cref{def:rule_comb_no_dependencies} states that whenever a rule has no backward links typed as \emph{trace} (i.e. no dependencies), all path conditions in the accumulator set are kept, along with the result of combining all the path conditions in the accumulator set with the current rule. 
%</nodepstext>

%<*nodepstextappendix>
Relation $\stackrel{combine}{\rightarrow}$ in \cref{def:rule_comb_no_dependencies_appendix} models the operational combination step. The relation has three input arguments: the first argument is the original path condition from the previous layer; the second argument is the set of path conditions accumulated thus far by combining other rules in the current layer with the original path condition; and the third argument is the rule from the current layer now being combined. The fourth argument of the relation, the relation's output, is the new set of path conditions resulting from this combination.

Briefly, the equation in \cref{def:rule_comb_no_dependencies_appendix} states that whenever a rule has no backward links typed as \emph{trace} (i.e. no dependencies), all path conditions in the accumulator set are kept, along with the result of combining all the path conditions in the accumulator set with the current rule. 
%</nodepstextappendix>

%=======================================================

%<*unsatdeps>
% \begin{definition}{Unsatisfied Dependencies Between Two Path Conditions}
% 
% Let $A=\langle V_1,E_1,st_1,\tau_1\rangle, B=\langle V_2,E_2,st_2,\tau_2\rangle\in \textsc{Pathcond}^{sr}_{tg}$ be two path conditions. We say that $A$ does not satisfy $B$'s dependencies if and only if $\neg (B'\vartriangleleft A')$, where $A'=A_{|backward}$ and $B'=B_{|backward}$.
% \end{definition}

The combination of a path condition $pc$ and a rule $rl$, when $rl$ has dependencies that are not satisfied by $pc$, is described by the relation $\stackrel{combine}{\rightarrow}\subseteq \textsc{Pathcond}^{sr}_{tg}\times \mathcal{P}(\textsc{Pathcond}^{sr}_{tg})\\ \times \textsc{Rule}^{sr}_{tg} \times \mathcal{P}(\textsc{Pathcond}^{sr}_{tg})$, defined as follows:

% $$\frac{\begin{array}{ll}&
% pc\in PC\;,\;\langle pc,rl\rangle \xrightarrow{combine} pc'\;,\;PC\setminus\{pc\},rl,\rangle \xrightarrow{combstep} PC'
% \end{array}}
% {\langle PC,rules\rangle \xrightarrow{rulestep} PC'}$$

$$\frac{\begin{array}{ll}\neg\big(rl|_{trace}\blacktriangleleft pc|_{trace}\big) \end{array}}
{\langle pc,AC,rl\rangle \xrightarrow{combine} AC}$$

%</unsatdeps>

%=======================================================

%<*satdeps>
% \begin{definition}{Satisfied Dependencies Between Two Path Conditions}
% 
% Let $A=\langle V_A,E_A,st_A,\tau_A,Match_A,Apply_A\rangle, B=\langle V_B,E_B,st_B,\tau_B,Match_B,Apply_B\rangle\in \textsc{Pathcond}^{sr}_{tg}$ be two path conditions. We say that $A$ satisfies $B$'s dependencies if and only if  $B'\vartriangleleft A'$, where $A'=A_{|backward}$ and $B'=B_{|backward}$.
% \end{definition}
% 
% A path condition A partially satisfies the dependencies of a path condition B if
% there exists an injective typed graph homomorphism between the backward links of
% B and those of A. In other words, 
%
% Note that this works for both cases where more instances of
% the same backward link exist in A, or in B.

The single rule partial and total combination relations $\stackrel{p\_comb}{\rightarrow}$ and $\stackrel{t\_comb}{\rightarrow}$, both having having signature $\mathcal{P}(\textsc{Pathcond}^{sr}_{tg}) \times \textsc{Rule}^{sr}_{tg} \times \textsc{Rule}^{sr}_{tg} \times \mathcal{P}(\textsc{Pathcond}^{sr}_{tg})$ are defined as follows:

%</satdeps>

%=======================================================

%<*satdepseqone>

\frac{\begin{array}{ll}
rl\cong rl_{glue}\sqcup ma_{\Delta}
\end{array}}
{\langle AC,rl,rl_{glue}\rangle \xrightarrow{p\_comb} AC\;\;\cup\;\; \bigcup_{pc\in AC} pc\stackrel{trace}{\sqcup} (rl_{glue} \sqcup ma_{\Delta})}
%</satdepseqone>

%=======================================================

%<*satdepseqtwo>
\frac{\begin{array}{ll}&
rl\cong rl_{glue}\sqcup ma_{\Delta}
\end{array}}
{\langle AC,rl,rl_{glue}\rangle \xrightarrow{t\_comb} \bigcup_{pc\in AC} pc\stackrel{trace}{\sqcup} (rl_{glue} \sqcup ma_{\Delta})}
%</satdepseqtwo>

%=======================================================

%<*satdepstwo>

The multiple rule partial and total combination relations $\stackrel{p\_step}{\rightarrow}$ and $\stackrel{t_step}{\rightarrow}$, both having having signature $\mathcal{P}(\textsc{Pathcond}^{sr}_{tg}) \times \mathcal{P}(\textsc{Rule}^{sr}_{tg}) \times \textsc{Rule}^{sr}_{tg} \times \mathcal{P}(\textsc{Pathcond}^{sr}_{tg})$ are defined as follows:

\begin{align}
\label{eq:pstepbase}
\frac{\begin{array}{ll}
\end{array}}
{\langle AC,rl,\emptyset\rangle \xrightarrow{p\_step} AC}
\end{align}

\begin{align}
\label{eq:pstep}
\frac{\begin{array}{ll}
rl_{glue}\in &partialSet,\;\langle AC,rl,rl_{glue}\rangle \xrightarrow{p\_comb} AC''\;,\;\\
&\langle AC'',rl,partialSet\setminus\{rl_{glue}\}\rangle \xrightarrow{p\_step} AC'
\end{array}}
{\langle AC,rl,partialSet\rangle \xrightarrow{p\_step} AC'}
\end{align}

\begin{align}
\label{eq:tstepbase}
\frac{\begin{array}{ll}
\end{array}}
{\langle AC,rl,\emptyset\rangle \xrightarrow{t\_step} AC}
\end{align}

\begin{align}
\label{eq:tstep}
\frac{\begin{array}{ll}
rl_{glue}\in &totalSet\;,\;\langle AC,rl,rl_{glue}\rangle \xrightarrow{t\_comb} AC''\;,\;\\
&\langle AC'',rl,totalSet\setminus\{rl_{glue}\}\rangle \xrightarrow{t\_step} AC'
\end{array}}
{\langle AC,rl,totalSet\rangle \xrightarrow{t\_step} AC''} 
\end{align}

%</satdepstwo>

%=======================================================

%<*satdepsthree>

% When a rule is combined with a path condition, partial and total matches of the rule may exist in the path condition. A total match of the rule is found in the path condition if a surjective typed graph homomorphism exists between the backward matcher of the rule and a subgraph of the path condition ($rl_{glue}\sqsubseteq pc \,\land\, rl\blacktriangleleft rl_{glue}$). As a reminder, the backward matcher of a rule consists of the rule's match part, together with its backward links (see \cref{def:back_match_transformation_rule}). In this case, existing path conditions in the output set will combined with\ldots
% 
% These two cases are modeled by the two inference rules in \cref{def:comb_path_cond_rule}. 
% 
% For each case a delta is added to the glue graph found such that the necessary apply or match part of the rules are added to the original path condition. In case the match is partial, then the previous set of path conditions is kept where the delta has been added, as well a second possibility with the total rule. Note that the fact that we use a surjective homomorphism allows us to deal with both rules that are isomorphic or homomorphic to parts of PC. This allows us to have more backward links in the path condition than in a rule, but also the other way around.

The combination of a path condition $pc$ and a rule $rl$, when $rl$ has dependencies that are satisfied by $pc$, is described by the relation $\stackrel{combine}{\rightarrow}\subseteq \textsc{Pathcond}^{sr}_{tg} \times \mathcal{P}(\textsc{Pathcond}^{sr}_{tg}) \times \textsc{Rule}^{sr}_{tg} \times \mathcal{P}(\textsc{Pathcond}^{sr}_{tg})$, defined as follows:%\levi{surjection is used too loosely here}

% $$\frac{\begin{array}{ll}&
% pc\in PC\;,\;\langle pc,rl\rangle \xrightarrow{combine} pc'\;,\;PC\setminus\{pc\},rl,\rangle \xrightarrow{combstep} PC'
% \end{array}}
% {\langle PC,rules\rangle \xrightarrow{rulestep} PC'}$$

$$\frac{\begin{array}{ll}rl|_{trace}\blacktriangleleft &pc|_{trace}\;,\\&\langle AC,rl,partialsat(rl,pc)\rangle\xrightarrow{p\_step} AC''\;,\;\\&\langle AC'',rl,totalsat(rl,pc)\rangle \xrightarrow{t\_step} AC'
\end{array}} 
{\langle pc,AC,rl\rangle \xrightarrow{combine} AC'}$$
\begin{center}
\vspace{.3cm}
where
\begin{align*}
rl_{glue}\in&~partialsat(rl,pc) \;\Longleftrightarrow\; \\
&rl_{glue}\sqsubseteq pc^{*} \,\land\, rl|_{trace}\blacktriangleleft rl_{glue}\quad\land\, \\&\hspace{1cm}\nexists rl' \,.\, (rl_{glue}\sqsubseteq rl'\sqsubseteq pc^{*} \land \Vert rl\Vert \blacktriangleleft rl')
\end{align*}
and 
\begin{align*}
rl_{glue}\in totalsat(rl,pc) \;\Longleftrightarrow \;rl_{glue}\sqsubseteq pc^{*} \,\land\, \Vert rl\Vert\blacktriangleleft rl_{glue}
\end{align*}
\end{center}

%</satdepsthree>

% It is important to notice that these subgraphs are always searched for in the same path condition (from the previous layer) and that the rule is only ``'glued' over the path conditions in the accumulator ($AC$) argument of the $\stackrel{combine}{\rightarrow}$. This allows avoiding interference between two rules in the same layer, which would occur if rules were sequentially ``glued'' on top of the same path condition. It also replicates the semantics of DSLTrans, as in a DSLTrans transformation rules from same layer always execute independently.

% \levi{CAREFUL WITH THE SURJECTIVE HOMOMORPHISM IN THE MATCHING!!! The combstep relation starts with the first path condition in PC and checks each rule in the path condition to combine for partial or total matches. The multiplied path conditions resulting from the partial or total matches are added to PC along until no more rules exist. Note that this does not induce a problem with parallelism in rule application because either the rule is isomorphic to a part of the path condition, in which case backward links are not touched, or there is a surjective homomorphism between the rule and a part of the original path condition, in which case a new backward link can be added that can be used by other rules. That is fine because if more backward links exist on the second path condition that means that an assumption was made on the number of elements consumed by similar previous rules. The first argument of the relation holds the original path condition such subgraphs to be glued can be fetched inside. We assume that when we create path conditions their nodes and edges are not disjoint.}

%=======================================================

% \begin{definition} {Combining a Path Condition with a Layer}
% \label{def:path_cond_layer_comb}
% 
% The \emph{path condition rule step relation}
% $\stackrel{rulecomb}{\rightarrow}\subseteq \textsc{Pathcond}^{sr}_{tg} \times \mathcal{P}(\textsc{Pathcond}^{sr}_{tg}) \times \textsc{Rule}^{sr}_{tg}\times
% \textsc{Pathcond}^{sr}_{tg}$ relation is defined as follows:
% 
% $$\frac{}
% {\langle pc,AC,\emptyset\rangle \xrightarrow{combpclayer} AC}$$
% 
% $$\frac{\begin{array}{ll}& rl\in layer\;,\;\langle pc,AC,rl\rangle\xrightarrow{combine}AC''\;,\;\langle pc,AC'',layer\setminus\{rl\}\rangle\xrightarrow{combpclayer}AC'
% \end{array}}
% {\langle pc,AC,layer\rangle \xrightarrow{combpclayer} AC''}$$
% 
% \end{definition}
% 
% \begin{definition} {Combining a Set of Path Condition with a Layer}
% \label{def:path_cond_set_layer_comb}
% 
% The \emph{path condition layer step relation}
% $\stackrel{combpcsetlayer}{\rightarrow}\subseteq \textsc{Pathcond}^{sr}_{tg} \times \textsc{Transf}^{sr}_{tg}\times
% \mathcal{P}(\textsc{Pathcond}^{sr}_{tg})$ relation is defined as follows:
% 
% $$\frac{}
% {\langle \emptyset,layer\rangle \xrightarrow{combpcsetlayer} \emptyset}$$
% 
% $$\frac{\begin{array}{ll}& pc\in AC\;,\;\langle pc,\{pc\},layer\rangle\xrightarrow{combpclayer}AC'\;,\;\langle AC\setminus\{pc\},layer\rangle\xrightarrow{combpcsetlayer}AC''
% \end{array}}
% {\langle AC,layer\rangle \xrightarrow{combpcsetlayer} AC'\cup AC''}$$
% 
% \end{definition}
% 
% 
% \begin{definition} {Path Condition Generation}
% \label{def:path_cond_gen}
% 
% Let $[layer::tr]\in \textsc{Transf}^{sr}_{tg}$ be a transformation, where $layer\in
% \textsc{Layer}^{sr}_{tg}$ is a Layer and $tr$ also a transformation. The $\stackrel{pathcondgen}{\rightarrow}\subseteq \mathcal{P}(\textsc{Pathcond}^{sr}_{tg}) \times \textsc{Layer}^{sr}_{tg}\times
% \textsc{Pathcond}^{sr}_{tg}$ is defined as follows:
% 
% $$\frac{}
% {\langle AC,[\;]\rangle \xrightarrow{pathcondgen} AC}$$
% 
% $$\frac{\begin{array}{ll}\langle \epsilon_{pc},\{\epsilon_{pc}\},layer\rangle\xrightarrow{layercomb}AC''\;,\;\langle AC'',tr\rangle\xrightarrow{pathcondgen}AC'
% \end{array}}
% {\langle \epsilon_{pc},[layer::tr]\rangle \xrightarrow{pathcondgen} AC}$$
% 
% \end{definition}

%<*combpclayer>

The \emph{layer combination relation}
$\stackrel{combpclayer}{\rightarrow}\subseteq \textsc{Pathcond}^{sr}_{tg} \times\\ \mathcal{P}(\textsc{Pathcond}^{sr}_{tg}) \times \textsc{Layer}^{sr}_{tg}\times
\mathcal{P}(\textsc{Pathcond}^{sr}_{tg})$ relation is defined as follows:
$$\frac{}
{\langle pc,AC,\emptyset\rangle \xrightarrow{combpclayer} AC}$$

$$\frac{\begin{array}{ll}rl\in layer,\;\langle pc,AC&,rl\rangle\xrightarrow{combine}AC'',\\ &\langle pc,AC'',layer\setminus\{rl\}\rangle\xrightarrow{combpclayer}AC'
\end{array}}
{\langle pc,AC,layer\rangle \xrightarrow{combpclayer} AC''}$$

%</combpclayer>

%<*combpclayertext>
After the step in~\cref{def:path_cond_layer_comb} is repeated for all the path conditions in the previous layer, these new sets of path conditions are collected together to produce the working set of path conditions for the layer. This process is modeled by relation $\stackrel{combpcsetlayer}{\rightarrow}$ in \cref{def:path_cond_set_layer_comb}.
%</combpclayertext>

%<*combpclayertextappendix>
After the step in~\cref{def:path_cond_layer_comb_appendix} is repeated for all the path conditions in the previous layer, these new sets of path conditions are collected together to produce the working set of path conditions for the layer. This process is modeled by relation $\stackrel{combpcsetlayer}{\rightarrow}$ in \cref{def:path_cond_set_layer_comb_appendix}.
%</combpclayertextappendix>

%<*combpcsetlayer>
The \emph{path condition layer step relation}
$\stackrel{combpcsetlayer}{\rightarrow}\subseteq \\\mathcal{P}(\textsc{Pathcond}^{sr}_{tg}) \times \textsc{Layer}^{sr}_{tg}\times
\mathcal{P}(\textsc{Pathcond}^{sr}_{tg})$ relation is defined as follows:
$$\frac{}
{\langle \emptyset,layer\rangle \xrightarrow{combpcsetlayer} \emptyset}$$

$$\frac{\begin{array}{ll}pc\in AC,\;\langle pc,\{pc\}&,layer\rangle\xrightarrow{combpclayer}AC',\\ &\langle AC\setminus\{pc\},layer\rangle\xrightarrow{combpcsetlayer}AC''
\end{array}}
{\langle AC,layer\rangle \xrightarrow{combpcsetlayer} AC'\cup AC''}$$

%</combpcsetlayer>

%<*combpcsetlayertext>
This working set of path conditions obtained for each layer is then itself combined with the rules in the next layer as in the algorithm just described, to obtain yet another working set of path conditions. This process will then continue in this layer-by-layer fashion through the transformation and is formally described in \cref{def:path_cond_gen}.\\
After all layers have been processed, the working set of the last layer contains all the  possible path conditions of the transformation. Through our abstraction relation defined in \cref{sec:abstraction_relation}, the final set of created path conditions will represent every feasible transformation execution. \cref{sec:verif_dsltrans_props} will discuss how our algorithm proves properties on these path conditions, and thus on all executions of the
transformation.
%</combpcsetlayertext>

%<*combpcsetlayertextappendix>
This working set of path conditions obtained for each layer is then itself combined with the rules in the next layer as in the algorithm just described, to obtain yet another working set of path conditions. This process will then continue in this layer-by-layer fashion through the transformation and is formally described in \cref{def:path_cond_gen_appendix}.
%</combpcsetlayertextappendix>

%<*pathcondgen>

Let $[layer::tr]\in \textsc{Transf}^{sr}_{tg}$ be a transformation, where $layer\in
\textsc{Layer}^{sr}_{tg}$ is a Layer and $tr$ also a transformation. The $\stackrel{pathcondgen}{\rightarrow}\subseteq \mathcal{P}(\textsc{Pathcond}^{sr}_{tg}) \times \textsc{Transf}^{sr}_{tg}\times
\mathcal{P}(\textsc{Pathcond}^{sr}_{tg})$ is defined as follows:
$$\frac{}
{\langle AC,[\;]\rangle \xrightarrow{pathcondgen} AC}$$

$$\frac{\begin{array}{ll}\langle \epsilon_{pc},layer^{\star}\rangle\xrightarrow{combpcsetlayer}AC''\;,\;\langle AC'',tr\rangle\xrightarrow{pathcondgen}AC'
\end{array}}
{\langle \epsilon_{pc},[layer::tr]\rangle \xrightarrow{pathcondgen} AC}$$
$$\text{where } layer^{\star}=\bigcup_{rl\in l}rl^{\star}$$

%</pathcondgen>

%<*pathcondgentext>
Note that in \cref{def:path_cond_gen}, the recursive rule considers the expansion ($layer^{\star}$) of all the rules in a layer (see \cref{def:transformation_rule_expansion}). This allows us to deal with polymorphism during path condition generation. In particular, given one rule $rl$ of $layer$, we consider for path condition generation all rules containing possible of replacements of each match element in $rl$ of certain type by an element belonging to one of the type's subtypes, as defined in the source metamodel $sr$.\\
After all layers have been processed, the working set of the last layer contains all the possible path conditions of the transformation. Through our abstraction relation in \cref{def:abstraction_pc_ex_appendix}, the final set of created path conditions will represent every feasible transformation execution.

\paragraph{\textbf{Notation:}}

We will use the abbreviation $\textsc{Pathcond}(tr)$ to represent the set of path conditions $AC$ produced for a transformation $tr$, where $\langle \epsilon_{pc},tr\rangle \xrightarrow{pathcondgen} AC$.
%</pathcondgentext>

%<*pathcondgentextappendix>
Note that in \cref{def:path_cond_gen_appendix}, the recursive rule considers the expansion ($layer^{\star}$) of all the rules in a layer (see \cref{def:transformation_rule_expansion_appendix}). This allows us to deal with polymorphism during path condition generation. In particular, given one rule $rl$ of $layer$, we consider for path condition generation all rules containing possible of replacements of each match element in $rl$ of certain type by an element belonging to one of the type's subtypes, as defined in the source metamodel $sr$.

\paragraph{\textbf{Notation:}}

We will use the abbreviation $\textsc{Pathcond}(tr)$ to represent the set of path conditions $AC$ produced for a transformation $tr$, where $\langle \epsilon_{pc},[layer::tr]\rangle \xrightarrow{pathcondgen} AC$.
%</pathcondgentextappendix>

%<*abstractionrel>

Let $tr\in \textsc{Transf}^{sr}_{tg}$ be a DSLTrans transformation. Let also $pc = \langle V,E,st,\tau,Match,Apply,Rules\rangle \in
\textsc{Pathcond}(tr)$ of be a path condition of $tr$ and $ex = \langle V',E',st',\tau',Input,Output\rangle \in Exec(tr)$ be an execution of $tr$. We have that $ex$ is abstracted by $pc$, noted $ex\Vvdash pc$, if and only if the set of transformation rules of $tr$ combined in $pc$ and the set of transformation rules of $tr$ used to built $ex$ is the same, and:
\begin{multline}
\label{eq:abstr_input_output}
\big(\forall rl\in Rules\;.\;Match(rl) \vartriangleleft Input^*\big) \;\land\; Output \blacktriangleleft Apply
\end{multline}
\center{and}
\begin{multline}
\label{eq:abstr_trace}
\big(\forall trc\in Components(pc|_{trace})\;.\;trc\vartriangleleft ex\big)\;\land\;\\
\big(\forall e'\in E'\, \exists e\in E\;.\;\tau'(e')=trace \Longrightarrow \tau(e)=trace\big)
\end{multline} 
% where for all $rl,rl'\in Rules$, if $rl\cong rl'$ then there exist two injective typed graph homomorphisms $f$ and $g$ such that $Match(rl)\stackrel{f}{\vartriangleleft} Input^*$, $Match(rl)\stackrel{g}{\vartriangleleft} Input^*$ and $CoDom(f)\neq CoDom(g)$.
% $/pc/ = Apply_{ex}\cup ex|_{trace}$ is the output part of $pc$, including symbolic traceability links.

%</abstractionrel>

%<*abstractionrelappendix>

Let $tr\in \textsc{Transf}^{sr}_{tg}$ be a DSLTrans transformation. Let also $pc = \langle V,E,st,\tau,Match,Apply,Rules\rangle \in
\textsc{Pathcond}(tr)$ of be a path condition of $tr$ and $ex = \langle V',E',st',\tau',Input,Output\rangle \in Exec(tr)$ be an execution of $tr$. We have that $ex$ is abstracted by $pc$, noted $ex\Vvdash pc$, if and only if the set of transformation rules of $tr$ combined in $pc$ and the set of transformation rules of $tr$ used to built $ex$ is the same, and:
\begin{multline}
\label{eq:abstr_input_output_appendix}
\big(\forall rl\in Rules\;.\;Match(rl) \vartriangleleft Input^*\big) \;\land\; Output \blacktriangleleft Apply
\end{multline}
\center{and}
\begin{multline}
\label{eq:abstr_trace_appendix}
\big(\forall trc\in Components(pc|_{trace})\;.\;trc\vartriangleleft ex\big)\;\land\;\\
\big(\forall e'\in E'\, \exists e\in E\;.\;\tau'(e')=trace \Longrightarrow \tau(e)=trace\big)
\end{multline} 
% where for all $rl,rl'\in Rules$, if $rl\cong rl'$ then there exist two injective typed graph homomorphisms $f$ and $g$ such that $Match(rl)\stackrel{f}{\vartriangleleft} Input^*$, $Match(rl)\stackrel{g}{\vartriangleleft} Input^*$ and $CoDom(f)\neq CoDom(g)$.
% $/pc/ = Apply_{ex}\cup ex|_{trace}$ is the output part of $pc$, including symbolic traceability links.

%</abstractionrelappendix>

%<*propertydef>

Let $tr\in \textsc{Transf}^{sr}_{tg}$ be a DSLTrans transformation. A property of $tr$ is a 6-tuple $\langle V,E,(s,t),\tau,Pre,Post\rangle$, where $Pre = \langle V',E',st',\tau'\rangle \in \textsc{Ipattern}^{sr}$ and $Post = \langle V'',E'',st''\tau''\rangle \in\\ \textsc{Ipattern}^{tg}$ are indirect metamodel patterns. We also have that $V=V'\cup V''$, $E\subseteq E'\cup E''$ and $\tau\subseteq \tau'\cup \tau''$, where the co-domain of $\tau$ is the union of the co-domains of $\tau'$ and $\tau''$ and the set $\{trace\}$. An edge $e \in E\setminus E'\cup E''$ is called a \emph{traceability link} and is such that $s(e)\in V''$, $t(e)\in V'$ and $\tau(e)=trace$. Finally we have that there is at least one path condition $\big\langle V_{pc},E_{pc},st_{pc},\tau_{pc}, Match,Apply,Rule\big\rangle\in \textsc{Pathcond}(tr)$ for which a surjective typed graph homomorphism $m\stackrel{f}{\blacktriangleleft} Pre$ exists, where $m\sqsubseteq Match$ and $f(v)\neq f(v')$ if $v$ and $v'$ are elements of the path condition belonging to the same rule of set $Rule$. The set of all properties of transformation $tr$ is called $\textsc{Property}(tr)$.

%</propertydef>

%<*satpropexe>

Let $tr\in \textsc{Transf}^{sr}_{tg}$ be a transformation. Let also $p = \langle V,E,st,\tau,\\Pre,Post\rangle \in \textsc{Property}(tr)$ be a property of $tr$ and $ex = \langle V',E',\\st',\tau',Input,Output\rangle \in Exec(tr)$ be an execution of $tr$. Execution $ex$ satisfies property $p$,  written $ex \models p$, if and only if:
\begin{multline*}
\forall f\; \exists g\,.\,\big(Pre \stackrel{f}{\vartriangleleft} Input^{*} \implies p \stackrel{g}{\vartriangleleft} ex^*\big)\\\text{where } V(Input)\cap CoDom(g) = CoDom(f)
\end{multline*}
% \begin{enumerate}
% \item $\langle V_p,E_p\setminus Il_p,\tau_p\rangle$ is a typed subgraph of $\langle V_x, E_x,\tau_x\rangle$
% \item if $v_p\rightarrow v_{p}'\in Il_p$ then there exists $v_x\rightarrow
% v_{x}'\in E_{x}^{*}$ where $\tau(v_p)=\tau(v_x)$, $\tau(v_p')=\tau(v_x')$ and
% $E_{x}^{*}$ is obtained by the transitive closure of $E_{x}$.
% \end{enumerate}
%</satpropexe>

%<*satproppc>

Let $tr\in \textsc{Transf}^{sr}_{tg}$ be a transformation. Let also $p = \langle V,E,st,\tau,\\Pre,Post\rangle \in \textsc{Property}(tr)$ be a property of $tr$ and $pc=\big\langle V',E',\\st',\tau', Match,Apply,Rulecop\big\rangle\in \textsc{Pathcond}(tr)$ be a path condition of $tr$. Path condition $pc$ satisfies property $p$, written $pc \vdash p$, if and only if:
\begin{multline*}
\forall f\; \exists g\,.\,\big(in\stackrel{f}{\blacktriangleleft} Pre \implies out \stackrel{g}\blacktriangleleft p\big)\\
\text{where } in \sqsubseteq Match^{*} \;\land\; out \sqsubseteq pc^{*}
\end{multline*}
Additionally $Dom(g)\cap Match(pc^{*}) = Dom(f)$ and $f(v)\neq f(v')$, $g(v)\neq g(v')$ whenever $v$ and $v'$ are elements of the path condition belonging to the same rule copy of set $Rulecop$.

%</satproppc>

%=======================================================

%<*pathgencompletenessproof>
\begin{pf}
Let $tr\in \textsc{Transf}^{sr}_{tg}$ be a DSLTans transformation. We wish to demonstrate that, for all transformation executions $ex\in \textsc{Exec}(tr)$, a path condition $pc\in \textsc{Pathcond}(tr)$ exists such that $ex$ is abstracted by $pc$, as formally expressed in \cref{def:abstraction_pc_ex}.\\

Completeness can be shown as a corollary of \cref{prop:pc_validity} about the \emph{validity} of path condition generation. The complete set of executions $\textsc{Exec}(tr)$ (see \cref{def:modeltransformation}) can be split into two kinds of executions:
\begin{enumerate}
  \item\label{lab:input_not_matched} The \emph{empty execution} $\epsilon_{ex}$ or the \emph{execution where the input model was not matched by any rule}. As mentioned in \cref{prop:pc_validity}, these executions are abstracted by the empty path condition $\epsilon_{pc}$.\vspace{.3cm}
  \item The \emph{execution $ex$ where a number of rules of $tr$ have been applied to the input model}, where each transformation rule $rl$ of $tr$ may have been applied more than once. In this case we have that, because all possible and valid rule combinations are considered when building path conditions, a path combination $pc$ exists that contains one or more copies of each of the rules used when operationally building $ex$. \\

  Moreover, during the proof of \emph{validity} of path condition generation in \cref{prop:pc_validity} we demonstrate that, when we add a new rule $rl$ to the last layer of a transformation $tr$ (such that we have a new transformation $tr'$), the rule combination step explained in Definitions~\ref{def:rule_comb_no_dependencies},~\ref{def:rule_comb_unsatisfied} and~\ref{def:rul_comb_partial_total} produces a new set of path conditions where each path condition in that set still abstracts at least one transformation execution of $tr'$. This part of the proof (the second induction) is achieved by building for transformation $tr'$ an input model $m$ that can be matched by $rl$ (as well as by all the other rules of $tr$), and then building from $m$ a new transformation execution that is abstracted by a path condition built for $tr'$. Because in the proof of \cref{prop:pc_validity}, $m$ is such that it can be matched by $rl$ an arbitrary amount of times, we know that, independently of the number of times a rule is applied during the construction of a transformation execution, a path condition abstracting that transformation execution exists.\\

Additionally, input elements that are not matched by any rule do not affect the abstraction relation, as explained in case~\ref{lab:input_not_matched} above. This means we also know that executions involving input models that are only partially matched by the rules of $tr$ are also abstracted by one path condition. 
\end{enumerate} 
\end{pf} 
%</pathgencompletenessproof>

%<*pathgencompletenessproofappendix>
\begin{pf}
Let $tr\in \textsc{Transf}^{sr}_{tg}$ be a DSLTans transformation. We wish to demonstrate that, for all transformation executions $ex\in \textsc{Exec}(tr)$, a path condition $pc\in \textsc{Pathcond}(tr)$ exists such that $ex$ is abstracted by $pc$, as formally expressed in \cref{def:abstraction_pc_ex_appendix}.
Completeness can be shown as a corollary of \cref{prop:pc_validity_appendix} about the \emph{validity} of path condition generation. The complete set of executions $\textsc{Exec}(tr)$ (see \cref{def:modeltransformation_appendix}) can be split into two kinds of executions:
\begin{enumerate}
  \item\label{lab:input_not_matched} The \emph{empty execution} $\epsilon_{ex}$ or the \emph{execution where the input model was not matched by any rule}. As mentioned in \cref{prop:pc_validity_appendix}, these executions are abstracted by the empty path condition $\epsilon_{pc}$.\vspace{.3cm}
  \item The \emph{execution $ex$ where a number of rules of $tr$ have been applied to the input model}, where each transformation rule $rl$ of $tr$ may have been applied more than once. In this case we have that, because all possible and valid rule combinations are considered when building path conditions, a path combination $pc$ exists that contains one or more copies of each of the rules used when operationally building $ex$. 
Moreover, during the proof of \emph{validity} of path condition generation in \cref{prop:pc_validity_appendix} we demonstrate that, when we add a new rule $rl$ to the last layer of a transformation $tr$ (such that we have a new transformation $tr'$), the rule combination step explained in Definitions~\ref{def:rule_comb_no_dependencies_appendix},~\ref{def:rule_comb_unsatisfied_appendix} and~\ref{def:rul_comb_partial_total_appendix} produces a new set of path conditions where each path condition in that set still abstracts at least one transformation execution of $tr'$. This part of the proof (the second induction) is achieved by building for transformation $tr'$ an input model $m$ that can be matched by $rl$ (as well as by all the other rules of $tr$), and then building from $m$ a new transformation execution that is abstracted by a path condition built for $tr'$. Because in the proof of \cref{prop:pc_validity_appendix}, $m$ is such that it can be matched by $rl$ an arbitrary amount of times, we know that, independently of the number of times a rule is applied during the construction of a transformation execution, a path condition abstracting that transformation execution exists.\\

Additionally, input elements that are not matched by any rule do not affect the abstraction relation, as explained in case~\ref{lab:input_not_matched} above. This means we also know that executions involving input models that are only partially matched by the rules of $tr$ are also abstracted by one path condition. 
\end{enumerate} 
\end{pf} 
%</pathgencompletenessproofappendix>

\subsection{Disambiguating Path Conditions}
\label{sec:disambig_path_condition}

Each individual rule combination defines a possible execution of rules, which
our algorithm then transforms into a path condition. However, at this step in
the algorithm, the path conditions formed thus far may be ambiguous. Two
elements of the same type in a path condition may match over the same or
different instances in a concrete input model. Therefore, it becomes necessary
to disambiguate the path conditions in order to explicitly represent all
possible transformation executions.

This process of disambiguation is based upon the merging of pairs of elements.
Given a (possibly ambiguous) path condition, the disambiguation algorithm will
produce a new set of disambiguated path conditions by recursively merging pairs
of elements of the same type in the match graph.

There are two cases:

\begin{itemize}
\item the two elements in the path condition consume two different instances of that type in the input model
\item the two elements consume the same instance in the input model
\end{itemize}

There is also the requirement that the two elements were not previously part of
the same rule. This is to prevent merging in the case where a rule was matching two
distinct elements of the same type. Two different elements in the input model must be consumed, and therefore merging is not permitted.

\subsubsection{Example}
We now consider the path condition in \cref{fig:collapse_locate_match} as
an example. This path condition is ambiguous as the \emph{Station} elements in
the match graph may match over two distinct elements in an input model, or the
same element. This section will examine the process to transform this ambiguous
path condition into a set of unambiguous path conditions.

The first unambiguous path condition is created by assuming that no elements in
the path condition will match over the same elements in the input model. Thus,
we add the unchanged path condition into the set of unambiguous path conditions.

All other path conditions are created by considering whether to merge pairs of elements of the same type. Again, these pairs are in the match graph of the path condition and cannot have been part of the same rule. In \cref{fig:collapse_locate_match} the only available elements, the \emph{Station} elements, are highlighted by dashed ellipses. Merging these elements is done in a straightforward manner, with the result appearing in \cref{fig:collapse_merge_match}.\levi{fix such that repeated edges are merged too. Careful if only vertices need to have ids or edges too} \levi{Shouldn't we keep the explanation/diagram for the apply merging? Otherwise the backward link requirement just pops up in the definition}

\begin{figure}[htb]
        \centering
        \begin{subfigure}[b]{0.40\textwidth}
                \centering
                \includegraphics[width=1\textwidth]{./figures/building_path_conditions/collapse_locate_match.pdf}
                \caption{Locating two match elements}
                \label{fig:collapse_locate_match}
        \end{subfigure}%
        ~~
        \begin{subfigure}[b]{0.40\textwidth}
                \centering
                \includegraphics[width=1\textwidth]{./figures/building_path_conditions/collapse_merge_match.pdf}
                \caption{Merging the elements}
                \label{fig:collapse_merge_match}
        \end{subfigure}%
        \caption{Selecting and merging match elements of the same type}
        \label{fig:merge_match_elements}
\end{figure}

%\paragraph{Merging of Apply Elements}
%\label{para:backward_links}
%
%At this step in the disambiguation algorithm, there is a special case to be
%handled which will happen much less often than the standard step of merging only
%match elements. The algorithm must check if the two match elements just merged
%in the previous step are connected by backward links to apply elements. If there
%are two apply elements that have the same type as each other, then these apply
%elements need to be merged together in the same way as the match elements were.
%This is because the backward links indicate that these two elements were
%generated from the same match element and are therefore the same element. The
%two apply elements to be merged are highlighted in
%figure~\ref{fig:collapse_locate_apply}, and are shown merged in
%figure~\ref{fig:collapse_merge_apply}.
%

%
%  \begin{figure}[htb]
%        \centering
%        \begin{subfigure}[b]{0.40\textwidth}
%                \centering
%                \includegraphics[width=1\textwidth]{./figures/building_path_conditions/collapse_locate_apply.pdf}
%                \caption{Locating connected apply elements}
%                \label{fig:collapse_locate_apply}
%        \end{subfigure}%
%        ~~
%        \begin{subfigure}[b]{0.40\textwidth}
%                \centering
%                \includegraphics[width=1\textwidth]{./figures/building_path_conditions/collapse_merge_apply.pdf}
%                \caption{Merging the elements}
%                \label{fig:collapse_merge_apply}
%        \end{subfigure}%
%        \caption{Selecting and merging apply elements connected by backward links}
%        \label{fig:merge_apply_elements}
%\end{figure}

In what follows we formally define how the disambiguation of a path condition including several rules is achieved.

\begin{definition}{Path Condition Match and Apply Element Merge Function}
\label{def:merge_apply_merge_function}

Let $pc = \big\langle V,E,st,\tau\big\rangle\in \textsc{Pathcond}^{sr}_{tg}$ be a path condition. The match element merge function $\alpha : \textsc{Pathcond}^{sr}_{tg} \rightarrow \mathcal{P}(\textsc{Pathcond}^{sr}_{tg})$ of $pc$ is such that:
\begin{align*}
\alpha(pc)=\big\{&\langle V\backslash\{y\},E',st',\tau' \rangle \;|\;\\
&\exists x,y\in Match(V)\;.\; \tau(x)=\tau(y)\;\land\; Rule(x)\setminus Rule(y) = \emptyset \;\land\\
&E'=\{x\xrightarrow{e}z\;|\;y\xrightarrow{e}z\in E_2\}\;\cup\\
&\qquad\,\,\{z\xrightarrow{e}x\;|\;z\xrightarrow{e}y\in E_2\}\;\cup\\
&\qquad\,\,\{w\xrightarrow{e}z\;|\;w\xrightarrow{e}z\in E_2 \land w\neq y \land z\neq y\}\big\}
\end{align*}

The complementary apply element merge function $\mu : \textsc{Pathcond}^{sr}_{tg}\rightarrow \mathcal{P}(\textsc{Pathcond}^{sr}_{tg})$ of $pc$ is defined as follows:
\begin{align*}
\mu(pc)=\big\{&\langle V\backslash\{y\},E',st',\tau' \rangle \;|\;\\
&\exists x,y\in Apply(V)\;.\; \tau(x)=\tau(y)\;\land\; Rule(x)\setminus Rule(y) = \emptyset \;\land\\
&\exists v\in Match(V_1)\;.\; x\xrightarrow{e} v\in E \land y\xrightarrow{e'} v\in E\;\land\; \tau(e)=\tau(e')=backward\;\land\\
&E'=\{x\xrightarrow{e}z\;|\;y\xrightarrow{e}z\in E_2\}\;\cup\\
&\qquad\,\,\{z\xrightarrow{e}x\;|\;z\xrightarrow{e}y\in E_2\}\;\cup\\
&\qquad\,\,\{w\xrightarrow{e}z\;|\;w\xrightarrow{e}z\in E_2 \land w\neq y \land z\neq y\}\big\}
\end{align*}
For the two definitions above we impose that the types of the relations that are copied from the second match/apply element to the one are preserved by the typing function $\tau'$.\\
\end{definition}

The match and apply element merge functions are complementary: the match merge function merges two match elements and the apply element merge function merges two apply elements, but only if they are connected by a backward link to the same match element.

\paragraph{Recursive Disambiguation}

In the above example, we have shown the merging of one pair of elements with the
same type in the match graph. This example has produced one path condition where no elements were merged, and one where the
\emph{Station} elements were merged.

In the general case, these produced path conditions may still be ambiguous, as
there could be further pairs of elements to be disambiguated. Therefore the
disambiguation algorithm is recursive. Every path condition produced by this
algorithm must be recursively examined for other pairs of elements to merge.

At the conclusion of the disambiguation algorithm, the final set of path
conditions will contain every possibility of merged and unmerged pairs of
elements. We then define this set to contain the \emph{unambiguous path
conditions} for the layer in question.

\begin{definition}{Path Condition Disambiguation Function}
\label{def:path_cond_disamb_function}

Let $pc = \big\langle V,E,st,\tau\big\rangle\in \textsc{Pathcond}^{sr}_{tg}$ be a path condition. The disambiguation function $disamb : \textsc{Pathcond}^{sr}_{tg}\rightarrow \mathcal{P}(\textsc{Pathcond}^{sr}_{tg})$ is recursively defined as:
\begin{gather*}
  disamb(pc) =
  \begin{cases}
    pc  & \text{if } \alpha(pc)=\emptyset \\
    pc\;\cup\;step(pc) \;\cup\;
    \bigcup_{pc'\in step(pc)} disamb(pc') & \text{if } \alpha(pc)\neq \emptyset
  \end{cases}
\end{gather*}
\center{\text{where} $step(pc) = \bigcup_{pc'\in \alpha(pc)} \mu(pc')$}
\end{definition}

As shown in definition~\ref{def:path_cond_disamb_function}, the disambiguation function returns the unchanged path condition if no match elements can be merged by function $\alpha$ ($\alpha(pc)=\emptyset$). In case match elements can be merged, the disambiguation function will return the input path condition, united with all separate possibilities of merging pairs of match (and potentially apply) elements ($step(pc)$), united with the recursive disambiguation of all those possibilities in $step(pc)$.

\begin{proposition}{The result of disambiguating a path condition is a finite set of path conditions.}
\end{proposition}

### 2015-03-13 text/dsltrans.tex
previous version of the paper

\section{The DSLTrans Transformation Language}
\label{sec:dsltrans}

In this section we will introduce the DSLTrans transformation language and its
constructs from~\cite{DBLP:conf/sle/BarrocaLAFS10}. A formal treatment of the syntax and semantics of DSLTrans is found in \cref{sec:DSLTrans_formal_appendix}.

\begin{figure*}[t]
        \centering
        \begin{subfigure}[b]{0.40\textwidth}
                \centering
                \includegraphics[width=1\textwidth]{./figures/policestation_dsltrans/organization.pdf}
                \caption{Organization language}
                \label{fig:OrganizationLanguage}
        \end{subfigure}%
        ~~
        \begin{subfigure}[b]{0.40\textwidth}
                \centering
                \includegraphics[width=1\textwidth]{./figures/policestation_dsltrans/gender.pdf}
                \caption{Gender language}
                \label{fig:GenderLanguage}
        \end{subfigure}%
        \caption{Metamodels for the Police Station transformation}
        \label{fig:squadmetamodel}
\end{figure*}

A DSLTrans transformation has a source and a target metamodel, which are seen
in \cref{fig:squadmetamodel}. This \emph{Police Station}
transformation will be presented throughout
the rest of this paper as an example transformation. The metamodel
in \cref{fig:OrganizationLanguage} represents a language for describing
the chain of command in a police station, which includes the male (\emph{Male}
class) and female officers (\emph{Female} class). The metamodel in
\cref{fig:GenderLanguage} represents a language for describing a different
view over the chain of command, where the officers working at the police station
are classified by gender.

\begin{figure*}[bht]
	\centering
		\includegraphics[width=0.9\textwidth]{./figures/policestation_dsltrans/transformation.pdf}
	\caption{The \emph{Police Station} model transformation expressed in DSLTrans}
	\label{fig:dsltransformation}
\end{figure*}

In \cref{fig:dsltransformation} we present a DSLTrans transformation that involves both metamodels.  A description of relevant constructs as well as visual notation remarks are found in \cref{subsec:DSLTrans_constructs}. Note
that the transformation is formed
from layers where each layer is a set of transformation rules. The
transformation will execute layer-by-layer, where transformation rules in a layer will execute in a
non-deterministic order but must produce a deterministic result, due to the fact that DSLTrans is confluent by construction~\cite{DBLP:conf/sle/BarrocaLAFS10}.

Another important characteristic of DSLTrans transformations is that they are not Turing-complete. As discussed in~\cite{DBLP:conf/sle/BarrocaLAFS10}, non-completeness is required to make a transformation execution always terminate, but yet still allows for appropriate expressiveness.

Besides the fact that DSLTrans' transformations are free of constructs that
imply unbounded recursion or \\non-determinism, DSLTrans' transformations are strictly outplace, meaning no changes are allowed to the input model. However, the output metamodel for a DSLTrans transformation can
be the same as the input metamodel. Also, elements cannot be removed
from the output metamodel as the result of applying a DSLTrans rule.
This restriction is consistent with the usage of model transformations as
translations~\cite{AMT2012}, as no deletion of output elements is strictly required.

The purpose of this \emph{Police Station} transformation is to flatten a chain of command given
in the \emph{Organization language} into two independent sets of male
and female officers represented in the \emph{Gender language}. The command
relations will be kept during this transformation, i.e. a female officer will
have a direct association to all her female subordinates and likewise for male
officers. Note that differences in the gender classification metamodel means
some relations present in the input model will not be retained in the output
model.

\begin{figure*}[htb]
        \centering
        \begin{subfigure}[b]{0.30\textwidth}
                \centering
                \includegraphics[width=1\textwidth]{./figures/policestation_dsltrans/model_police_hierarchy.pdf}
                \caption{Original model}
                \label{fig:police_hierarchy}
        \end{subfigure}%
        ~~
        \begin{subfigure}[b]{0.40\textwidth}
                \centering
                \includegraphics[width=1\textwidth]{./figures/policestation_dsltrans/model_police_gender.pdf}
                \caption{Transformed model}
                \label{fig:police_gender}
        \end{subfigure}%
        \caption{Model before and after transformation}
        \label{fig:transformationexample}
\end{figure*}

An example of this transformation's execution can be observed in \cref{fig:transformationexample}, where the input model is on the left and the output model is on the right. Notice that the elements $s$, $m_k$  and $f_k$ in \cref{fig:OrganizationLanguage} are instances of the source \emph{Organization} metamodel elements $Station$, $Male$ and $Female$ respectively. The primed elements in \cref{fig:GenderLanguage} are their counterpart instances in the target \emph{Gender} metamodel.

Each individual transformation rule in the transformation is composed of two graphs. The first graph
is denoted as the \emph{match graph}, and is a pattern holding elements from the
source metamodel. Likewise, the \emph{apply graph} is a pattern containing
elements from the target metamodel. A formal definition of a transformation rule is found in~\cref{def:transformation_rule} in \cref{sec:formal_background}.

As an example, consider the transformation rule marked \emph{Stations} in the
first rule layer in \cref{fig:dsltransformation}. The match graph holds
one \emph{Station} element from the source metamodel, while the apply graph
holds one \emph{Station} element from the target metamodel. This means that for all
elements in the input model which are of type \emph{Station} in the
\emph{Organization Language}, a element of type
\emph{Station} in the \emph{Gender Language} will be created in the output model.

Note that in our approach, we require that the match graph of a rule is not a subset of the match graph of any other rule (as formally stated in \cref{def:layer_transformation} of model transformation, in \cref{sec:formal_background} of this paper). This requirement is to prevent the case where a rule could not execute independently of another rule, except for the cases when such dependency is explicitly defined by backward links. This is undesirable for the algorithm as presented here as we will explain later. However, as seen in~\cite{conf/gg/SelimLCDO14}, the expressiveness of the transformations our algorithm can examine is not restricted. In that work, we detail an operational rule processing step to handle overlapping rules.

\subsection{Properties to Prove}

The properties we aim to prove on the Police Station transformation are structural properties. These properties are composed of a pre-condition and a post-condition component, as seen in \cref{fig:properties}.

Informally, a property can be read as `if the pre-condition graph matches on the input model to the transformation, then the post-condition graph will match any output model produced'. Further details as well as formal validity and completeness of the property proving process are discussed in \cref{sec:verif_dsltrans_props}.

As a brief example of property syntax and semantics, consider the property in \cref{fig:dsltrans_prop1}. The pre-condition graph is composed of a Station element connected to a Female element and a Male element, where all elements are from the \emph{Organization language} metamodel. This structure is repeated in the post-condition graph, with the difference that the metamodel for these elements is the \emph{Gender language}. Thus, this property represents the statement ``\emph{a model which includes a
police station that has both male and female officers will be
transformed into a model where the male officer will exist in the male set
and the female officer will exist in the female set}''. We expect this property to always hold in our transformation.

\begin{figure*}[thb]
        \centering
        \begin{subfigure}[b]{0.45\textwidth}
                \centering
                \includegraphics[height=5.5cm]{./figures/policestation_dsltrans/property1.pdf}
                \caption{Property 1 -- Expected to hold}
                \label{fig:dsltrans_prop1}
        \end{subfigure}%
        ~~
        \begin{subfigure}[b]{0.45\textwidth}
                \centering
                \includegraphics[height=5.5cm]{./figures/policestation_dsltrans/property2.pdf}
                \caption{Property 2 -- Not expected to hold}
                \label{fig:dsltrans_prop2}
        \end{subfigure}%
        \caption{ Properties to be proved on the Police Station transformation}
        \label{fig:properties}
\end{figure*}

In contrast, we do not expect the property in \cref{fig:dsltrans_prop2} to always hold. This property represents the statement ``\emph{any model which includes a
female officer will be transformed into a model where that female officer will
always supervise another female officer}''. It is not difficult to construct an input model where the pre-condition holds, but the post-condition does not. This would be an input model that contains only one female officer, as there will only be one female officer in the output model.

We shall discuss our property proving technique in \cref{sec:verif_dsltrans_props}. Following this, experiments in \cref{sec:experiments} will present experimental results from proving these two properties on the Police Station transformation.

\subsection{DSLTrans Constructs}
\label{subsec:DSLTrans_constructs}
This section will describe all of the DSLTrans constructs involved in our
property-proving algorithm. These constructs are found in the transformation presented in \cref{fig:dsltransformation}. Formal details for the handled constructs are found in \cref{sec:DSLTrans_formal} and \cref{sec:DSLTrans_syntax}, while \cref{sec:DSLTrans_semantics} briefly introduces the formal semantics of this subset of DSLTrans. The visual syntax presented here is based on the DSLTrans Eclipse plug-in syntax~\cite{dsltrans_manual}.

\begin{itemize}
\item \textbf{Match Elements}: Match elements are variables typed by elements of the source
metamodel which will match over elements of that type (or subtype) in the
input model when the transformation is executed. Note that match
elements in a rule are searched for injectively in a model. This means that, for
example, if a match graph includes two elements of type \emph{Station}, then the rule will only match over models that include at least two instances of type \emph{Station}.

In the DSLTrans notation as seen throughout this paper, the match elements will be in a white box in the top half of a rule.\\

\item \textbf{Direct Match Links}: Direct match links are variables typed by labelled associations of the source metamodel, which will match over associations of the same type in the input model. A direct match link is always expressed between two match elements.\\

\item \textbf{Indirect Match Links}: Indirect match links are similar to direct
match links, but there may exist a path of containment associations between the
matched instances. Our notion of indirect links captures only
acyclic EMF containment associations.

% As such, it avoids cycles and infinite
% amounts of matches over the transitive closure of associations in the input
% models.

In \cref{fig:dsltransformation}, indirect match links are represented in all the
transformation rules in the last layer as dashed arrows between elements in the match graph.\\

\item \textbf{Backward Links}: Backward links connect elements of the match and
the apply patterns of a DSLTrans rule in order to represent dependencies on
element creation by previous layers of the transformation. When used in a rule,
backward links match over traceability links between elements of the
transformation's input and output models.
These traceability links are implicitly created when any rule is executed during the transformation. Backward
links thus make it possible to refer in a rule to output elements created by a
previous layer.

Backward links are found in \cref{fig:dsltransformation} in all transformation rules on the last layer and are depicted as dashed lines.\\

%\item \emph{Negative Conditions}: it is possible to express negative conditions over
%match elements, backward, direct and indirect match links.

\item \textbf{Apply Elements and Apply Links}: Apply elements and apply links are similar to match elements and
match links, but are instead typed by elements of the target metamodel. Apply elements in a given transformation rule that are not connected to backward links will create elements of the same type in the transformation's output. Apply links will always be created in the transformation's output. These output elements and links will be created as many times as the match graph of the rule is isomorphically found in the input model.\\

Consider the transformation rule denoted \emph{Station2Male} in the last rule
layer of \cref{fig:dsltransformation}. This rule takes \emph{Station} and
\emph{Male} elements of the \emph{Gender Language} metamodel, where these
elements were created in a previous layer from \emph{Station} and \emph{Male}
elements of the \emph{Organization Language} metamodel, and connects them using a
\emph{male} association.

%\item \emph{Apply Attributes}: DSLTrans includes a small language for
%building the values attributes of apply model elements from references to one or
%more match model element attributes.
\end{itemize}

### 2015-03-13 text/formal_background.tex
previous version of the paper

\section{Formal Background}
\label{sec:formal_background}

In this section we will introduce the formal concepts that will be used throughout all this paper. We start in \cref{sec:typed_graphs} by a few (typed) graph concepts that will be used as mathematical building blocks throughout this paper. In particular we will introduce the notion of typed graph, typed graph union and subset, and useful relations between typed graphs based on homomorphisms. Notice that these concepts are well known from graph theory and are only slightly customized for our purposes.

Armed with the fundamental notion of typed graph, we can then introduce other formal concepts in Sections~\ref{sec:DSLTrans_formal}, \ref{sec:DSLTrans_syntax} and \ref{sec:DSLTrans_semantics} which describe the artifacts from the modeling and transformation world that we require for our verification technique. Naturally, we start by introducing the central notion of \emph{metamodel}, allowing the description of the inputs and outputs of a model transformation. Other fundamental notions we will define in this section are \emph{model}, \emph{transformation rule}, \emph{transformation} and the semantic concept of \emph{model transformation execution}. Several auxiliary and intermediate notions for defining the syntax and semantics of our techniques will also be introduced here.

Note that this section presents a collection of formal tools that are used in the subsequent sections of this paper where the contributions of this paper are presented. It is meant as a formal reference for the upcoming formal development. This section can be safely skipped or skimmed by the reader, who can return to these definitions punctually to understand the detailed formal underpinning of our approach.

\subsection{Typed Graphs}
\label{sec:typed_graphs}
% We begin this section by formally defining the graphs and graph instances used in our algorithm as fundamental structures.
% \begin{definition}{Typed Graph}
% \label{def:typed_graph}
% A typed graph is a triple $\langle V,E,\tau\rangle$ where $V$ is a finite set of
% vertices, $E\subseteq V\times V$ is a set of edges connecting the vertices and
% $\tau:V\rightarrow Type$ is a typing function for the elements of V, where $Type$ is
% a set of type names. Edges $(v,v')\in E$ are noted $v\rightarrow v'$. The set of
% all typed graphs is called $TG$.
% \end{definition}
% 
% Definition~\ref{def:typed_graph_homomorphism} describes when a typed graph is said to be homomorphic to another.
% 
% \begin{definition}{Typed Graph Homomorphism  / Typed Graph Instance}
% \label{def:typed_graph_homomorphism}
% 
% Let $\langle V,E,\tau\rangle=g,\langle V',E',\tau'\rangle=g'\in TG$ be typed
% graphs. There is a typed graph homomorphism between $g'$ and $g$ if and only if
% there exists a graph homomorphism $f:V'\rightarrow V$ such that for all $v'\in
% V$ we have that $\tau'(v')=\tau(f(v'))$. Note that, trivially, a typed graph
% homomorphism is a formal homomorphism. We denote an \emph{injective} typed graph
% homomorphism between typed graphs $g$ and $g'$ by $g' \vartriangleleft g$.
% \end{definition}
% 
% In the text that follows we will use the terms \emph{typed graph homomorphism}
% and \emph{typed graph instance} interchangeably.
% 
% Definition~\ref{def:typed_graph_strict_instance} is stricter than
% definition~\ref{def:typed_graph_homomorphism}, and indicates that there is a
% strict many-to-one type correspondence of the nodes of two graphs and their
% relations.
% 
% \begin{definition}{Typed Graph Strict Instance}
% \label{def:typed_graph_strict_instance}
% 
% Let $\langle V,E,\tau\rangle=g,\langle V',E',\tau\rangle=g'\in TG$ be typed
% graphs. $g'$ is a typed graph strict instance of $g$, noted $g'
% \blacktriangleleft g$, if and only if there exists a surjective
% typed graph homomorphism between $g'$ and $g$.
% 
% \end{definition}
% 
% In order to simplify the subsequent development we will often
% use the notation ${\langle V,E,\tau\rangle}^*$ to represent the transitive
% closure of the edges of typed graph  $\langle V,E,\tau\rangle$.

% \begin{definition}{Typed Graph}
% \label{def:typed_graph}
% A typed graph is a triple $\langle V,E,\tau\rangle$ where $V$ is a finite set of
% vertices, $E\subseteq V\times V$ is a set of edges connecting the vertices and
% $\tau:V\rightarrow Type$ is a typing function for the elements of V, where $Type$ is
% a set of type names. Edges $(v,v')\in E$ are noted $v\rightarrow v'$. The set of
% all typed graphs is called $TG$.
% \end{definition}

% \begin{definition}{Typed Graph and Indirect Typed Graph}
% \label{def:typed_graph}
% 
% A typed graph is a triple $\langle V,E,\tau\rangle$ where $V$ is a finite set of
% vertices, $E\subseteq V\times V$ is a finite set of directed edges connecting the vertices and $\tau:V\cup E \rightarrow Type\cup\{containment,reference\}$ is a typing function for the elements of V and E. Typing function $\tau$ is such that $\tau(v)\in Type$ if $v\in V$ and $\tau(e)\in \{containment,reference\}$\footnote{By using \emph{containment} and \emph{reference} as types for edges we allow modeling the different types of associations between the elements of a metamodel or a model. In particular, the fact that the subgraph of containment relations in a typed graph is acyclic models EMF containment associations.} if $e\in E$.  Edges $(v,v')\in E$ are noted $v\rightarrow v'$. We furthermore impose that the graph $\langle V,\{v\rightarrow v'\in E|\tau(v\rightarrow v')=containment\}\rangle$ is acyclic. The set of all typed graphs is called $TG$.
% 
% An indirect typed graph is a 4-tuple $\langle V,E,T,Il\rangle$, where $\langle V,E,T\rangle$ is a typed graph and $Il\subseteq E$ is a set of edges called \emph{indirect links}. The set of all indirect typed graphs is called $ITG$.
% \end{definition}

\CatchFileBetweenTags{\typedgraphtext}{text/definitions}{typedgraphtext}{\typedgraphtext}

\begin{definition}{Typed Graph\\}
\label{def:typed_graph}
\CatchFileBetweenTags{\typedgraph}{text/definitions}{typedgraph}{\typedgraph}
\end{definition}

\CatchFileBetweenTags{\typedgraphuniontext}{text/definitions}{typedgraphuniontext}{\typedgraphuniontext}

\begin{definition}{Typed Graph Union\\}
\label{def:typed_graph_union}
\CatchFileBetweenTags{\typedgraphunion}{text/definitions}{typedgraphunion}{\typedgraphunion}
\end{definition}

\CatchFileBetweenTags{\typedgraphhomomorphismtext}{text/definitions}{typedgraphhomomorphismtext}{\typedgraphhomomorphismtext}

\begin{definition}{Typed Graph Homomorphism\\}
\label{def:typed_graph_homomorphism}
\CatchFileBetweenTags{\typedgraphhomomorphism}{text/definitions}{typedgraphhomomorphism}{\typedgraphhomomorphism}
\end{definition}

Note that, trivially, a typed graph homomorphism is a graph homomorphism.

% \begin{lemma}{The Union of Two Typed Graph Homomorphisms is a Typed Graph Homomorphism}
% 
% Let $g=\langle V_g,E_g,st_g,\tau_g\rangle$, $g'=\langle
% V'_g,E'_g,st'_g,\tau'_g\rangle$, $h=\langle V_h,E_h,st_h,\tau_h\rangle$,
% $h'=\langle V'_h,E'_h,st'_h,\tau'_h\rangle$ be typed graphs, where $g$ and $h$
% are potentially joint, as well as $g'$ and $h'$. Let also $f:V_g\rightarrow V'_g$ and
% $i:V_h\rightarrow V'_h$ be a typed graph homomorphisms between $g$ and $g'$ and
% $h$ and $h'$, respectively. Let us show that $f\cup i$ is a typed graph homomorphism between $g\sqcup g'$ and $h\sqcup h'$.
% The proof is trivial if $g$ and $h$, as well as $h$ and $h'$ are disjoint. If $g$ and $h$ are joint over a vertex $v$, we know that $(f\cup i)(v)$
% \end{lemma}
% 
% \begin{lemma}{The Union of Two Surjective Typed Graph Homomorphisms is a Surjective Typed Graph Homomorphism}
% \levi{finish this proof}
% \end{lemma}

% \levi{changed the direction of the triangle for the surjective type graph homomorphism. Typed graph homomorphisms make sure that ALL connections between two nodes in the first graph also exists in the two mapped nodes of the second graph.}

% \begin{definition}{Strict Typed Graph Homomorphism}
% \label{def:strict_typed_graph_homomorphism}
% 
% Let $\langle V,E,\tau\rangle=g,\langle V',E',\tau\rangle=g'\in TG$ be typed
% graphs. $g'$ is a typed graph strict instance of $g$, noted $g'
% \blacktriangleleft g$, if and only if there exists a surjective
% typed graph homomorphism between $g'$ and $g$.

\CatchFileBetweenTags{\typedsubgraphtext}{text/definitions}{typedsubgraphtext}{\typedsubgraphtext}

\begin{definition}{Typed Subgraph\\}
\label{def:typedsubgraph}
\CatchFileBetweenTags{\typedsubgraph}{text/definitions}{typedsubgraph}{\typedsubgraph}

% An indirect typed graph $\langle V',E',\tau',Il\rangle \in ITG$ is an indirect typed
% subgraph of a typed graph $\langle V,E,\tau\rangle \in TG$, written $\langle
% V',E',\tau',Il\rangle \lhd \langle V,E,\tau\rangle$ iff:
% 
% \begin{enumerate}
% \item $\langle V',E'\setminus Il,\tau'\rangle \blacktriangleleft \langle V, E,\tau\rangle$
% \item if $v_i\rightarrow v_i'\in Il$ then there exists $v\rightarrow v'\in E_{c}^{*}$ where $\tau(v_i)=\tau(v)$, $\tau(v_i')=\tau(v')$ and $E_{c}^{*}$ is obtained by the transitive closure of $E_{c}=\{v\rightarrow v'\in E|\tau(v\rightarrow v')=containment\}$.
% \end{enumerate}
\end{definition}

\CatchFileBetweenTags{\typedgraphisomorphismtext}{text/definitions}{typedgraphisomorphismtext}{\typedgraphisomorphismtext}

\begin{definition}{Typed Graph Isomorphism\\}
\label{def:typed_graph_isomorphism}
\CatchFileBetweenTags{\typedgraphisomorphism}{text/definitions}{typedgraphisomorphism}{\typedgraphisomorphism}
\end{definition}
% \begin{definition}{Typed Graph Connected Component}
% \label{def:typed_graph_connected_component}
% 
% Let $g \in \textsc{TG}$ be a typed graph. We say that $\langle V,E,st,\tau\rangle \sqsubseteq g$ is a connected component of $g$ if for any pair of vertices $v,v'\in V$ there exists a path between $v$ and $v'$, where the direction of the edges between $v$ and $v'$ does not matter.
% \end{definition}

% \begin{definition}{Typed Graph Instance}
% \label{def:typed_graph_instance}
% 
% Let $\langle V,E,\tau\rangle=g,\langle V',E',\tau\rangle=g'\in TG$ be typed graphs.
% $g'$ is a typed graph instance of $g$, written $g'\Vdash g$, iff for all $v_{1}'\rightarrow v_{2}'\in E'$ there is a $v_{1}\rightarrow v_{2}\in E$ such that $\tau(v_{1}')=\tau(v_1)$, $\tau(v_{2}')=\tau(v_2)$ and $\tau(v_{1}'\rightarrow v_{2}')=\tau(v_{1}\rightarrow v_{2})$.
% \end{definition}

%Notice that we only enforce that connections between vertices of $g'$ must exist also in $g$ and have the same type.

% \levi{Note that each edge between two nodes in the first graph has to exist in the second graph and vice-versa.}
% \levi{in the sections that follow we will formally elaborate on relevant concepts that are introduced.}

\paragraph{\textbf{Notation:}}
\CatchFileBetweenTags{\notationformalbackground}{text/definitions}{notationformalbackground}{\notationformalbackground}

\subsection{Metamodel and Model-Related Constructs}
\label{sec:DSLTrans_formal}

% This section will formally describe the DSLTrans constructs that are required in
% our property-proving algorithm. The definitions presented strongly rely on the concepts of typed graphs, just introduced in \cref{sec:typed_graphs}.

We will start by introducing the notion of \emph{metamodel}, which in DSLTrans is used to type the input and output models of a DSLTrans transformation.  Two metamodels, the \emph{organization language} and \emph{gender language} are depicted in \cref{fig:squadmetamodel}.

\begin{definition}{Metamodel\\}
\label{def:metamodel}
\CatchFileBetweenTags{\metamodel}{text/definitions}{metamodel}{\metamodel}
\end{definition}

\CatchFileBetweenTags{\metamodeltext}{text/definitions}{metamodeltext}{\metamodeltext}

\begin{definition}{Expanded Metamodel\\}
\label{def:expanded_metamodel}
\CatchFileBetweenTags{\expandedmetamodel}{text/definitions}{expandedmetamodel}{\expandedmetamodel}
\end{definition}

\CatchFileBetweenTags{\expandedmetamodeltext}{text/definitions}{expandedmetamodeltext}{\expandedmetamodeltext}

\begin{definition}{Metamodel Instance\\}
\label{def:metamodelinstance}
\CatchFileBetweenTags{\metamodelinstance}{text/definitions}{metamodelinstance}{\metamodelinstance}
\end{definition}

\CatchFileBetweenTags{\metamodelinstancetext}{text/definitions}{metamodelinstancetext}{\metamodelinstancetext}

\begin{definition}{Containment Transitive Closure\\}
\label{def:instance_closure}
\CatchFileBetweenTags{\instanceclosure}{text/definitions}{instanceclosure}{\instanceclosure}

\end{definition}

\CatchFileBetweenTags{\instanceclosuretext}{text/definitions}{instanceclosuretext}{\instanceclosuretext}

\begin{definition}{Model\\}
\label{def:model}
\CatchFileBetweenTags{\model}{text/definitions}{model}{\model}
\end{definition}

\CatchFileBetweenTags{\modeltext}{text/definitions}{modeltext}{\modeltext}

\begin{definition}{Input-Output Model\\}
\label{def:input_output_model}
\CatchFileBetweenTags{\inputoutputmodel}{text/definitions}{inputoutputmodel}{\inputoutputmodel}
\end{definition}

%Change if needed
\CatchFileBetweenTags{\inputoutputmodeltext}{text/definitions}{inputoutputmodeltext}{\inputoutputmodeltext}

\begin{definition}{Metamodel Pattern and Indirect Metamodel Pattern\\}
\label{def:metamodel_pattern} 
\CatchFileBetweenTags{\metamodelpattern}{text/definitions}{metamodelpattern}{\metamodelpattern}
\end{definition}

\CatchFileBetweenTags{\metamodelpatterntext}{text/definitions}{metamodelpatterntext}{\metamodelpatterntext}

\subsection{Syntactic Transformation Constructs}
\label{sec:DSLTrans_syntax}

This section will detail the abstract syntax of the constructs involved in a DSLTrans transformation.

\begin{definition}{Transformation Rule\\}
\label{def:transformation_rule}
\CatchFileBetweenTags{\transformationrule}{text/definitions}{transformationrule}{\transformationrule}
\end{definition}

\CatchFileBetweenTags{\transformationruletext}{text/definitions}{transformationruletext}{\transformationruletext}

\begin{definition}{Matcher of a Transformation Rule\\}
\label{def:back_match_transformation_rule}
\CatchFileBetweenTags{\backmatchtransformationrule}{text/definitions}{backmatchtransformationrule}{\backmatchtransformationrule}
\end{definition}

\Cref{def:back_match_transformation_rule} \CatchFileBetweenTags{\backmatchtransformationruletext}{text/definitions}{backmatchtransformationruletext}{\backmatchtransformationruletext}

\begin{definition}{Expanded Transformation Rule\\}
\label{def:transformation_rule_expansion}
\CatchFileBetweenTags{\expandedtransformationrule}{text/definitions}{expandedtransformationrule}{\expandedtransformationrule}
\end{definition}
\CatchFileBetweenTags{\expandedtransformationruletext}{text/definitions}{expandedtransformationruletext}{\expandedtransformationruletext}

\begin{definition}{Layer, Model Transformation\\}
\label{def:layer_transformation}
\CatchFileBetweenTags{\layertransformation}{text/definitions}{layertransformation}{\layertransformation}
\end{definition}

\CatchFileBetweenTags{\layertransformationtext}{text/definitions}{layertransformationtext}{\layertransformationtext}

\paragraph{\textbf{Notation:}}

\CatchFileBetweenTags{\notationtext}{text/definitions}{notationtext}{\notationtext}

% We naturally extend the notion of union (definition~\ref{def:typed_graph_union})
% to models (definition~\ref{def:metamodelmodel}), match-apply models
% (definition~\ref{def:match_apply_model}) and transformation rules 
% (definition~\ref{def:transformation_rule}). We also extend the notion of
% indirect typed subgraph (definition~\ref{def:indirecttypedsubgraph}) to
% transformation rules (definition~\ref{def:transformation_rule}) and match-apply
% models (definition~\ref{def:match_apply_model}). Finally, we extend the notion
% of typed graph equivalence (definition~\ref{def:typed_graph_equivalence}) to
% transformation rules (definition~\ref{def:transformation_rule}).

% \subsubsection{Transformation Language Semantics}
% 
% We will now address the semantics of the DSLTrans language. We will start by defining a match function that, given an input-output model and a transformation rule, returns all subgraphs of that input-output model where the rule's match pattern (including backward links) is found.  
% 
% % \levi{Need to be careful throughout all this section to understand whether we model the match and apply functions as functions or relations. Because of the non-deterministic generation of new vertices in the definition of apply function~\ref{def:apply_function}, relations are used in the definitions that follow to evaluate a transformation. This is an issue of the formalism used to model, should we make it explicit?}
% 
% \begin{definition}{Match Function}
% \label{def:match_function}
% 
% Let $m_{in}\in \textsc{Iom}^{sr}_{tr}$ be a input-output model and $rl\in \textsc{Rule}^{sr}_{tg}$ be a transformation
% rule. The $match : \textsc{Iom}^{sr}_{tg}\times \textsc{Rule}^{sr}_{tg}\rightarrow
% \mathcal{P}(\textsc{Iom}^{sr}_{tg})$ function is defined as follows: $$match_{rl}(m_{in})= \big\{ glue_{noind}\;|\;
% glue\sqsubseteq m_{in}^{*} \land glue \cong rl^{backM}_{trace}\big\}$$
% 
% where $glue$ is an input-output model. Additionally we have that: $rl_{trace}$ is a version of $rl$ where edges of type \emph{backward} have been replaced by edges of type \emph{trace}; $glue_{noind}$ is a version of $glue$ where the indirect links have been removed.
% % \levi{all backward links are replaced by trace links such that the matching using subgraph isomorphim can work.}
% % 
% % \levi{Where $rl^{-}$ is rule $rl$ without nodes in the apply part that are not connected to \emph{trace links}, as well as links adjacent to those nodes. $mg^{-ind}$ is the match graph without indirect links that come from the rule, given those cannot be part of the intermediate $MAM$ match result since that result is going to be united with the original $MAM$ (this was incorrect in the original SLE paper).}
% % \levi{The matcher function returns all graphs of the input MAM that are matched by a rule, including its backward links. Note that to find a subgraph ($\sqsubseteq$) of the input MAM that is isomorphic ($\cong$) to the rule being matched because the graph rewriting is achieved by joint union of the match graph extended by the new rewrite node, and the original MAM.}
% % \levi{TODO: review the $mg^{-ind}$ and the $rl^{-}$ notations.}
% 
% % Due to the fact that the $\cong$ relation is based on the notion of graph
% % isomorphism, permutations of the same match result may exist in the
% % $\big\{g\;|\; g\lhd m \land g \cong strip(tr)\big\}$ set. The --- undefined ---
% % $remove:\mathcal{P}(TR^{s}_{t})\rightarrow \mathcal{P}(TR^{s}_{t})$ function is such that
% % it removes such undesired permutations.
% 
% %The $strip:TR^{s}_{t}\rightarrow TR^{s}_{t}$ function is such that
% %$$strip(\big\langle V,E,T,\langle V_m,E_m,T_m,s\rangle,\langle V_a,E_a,T_a,t\rangle,Bl,Il\big\rangle) = \big\langle V',E',T,\langle V_m,E_m,T_m\rangle,\langle V'_a,E'_a,T_a\rangle,Bl,Il\big\rangle$$
% %\begin{center}
% %where $(v'\in V'_a\Rightarrow v\rightarrow v'\in Bl)\;\land\;(v\rightarrow v'\in E'_a\Rightarrow (v\rightarrow v'\in E_m \land \{v,v'\}\subseteq V'_a))$
% %\end{center}
% \end{definition}
% 
% The match function in definition~\ref{def:match_function} looks for subgraphs ($glue\sqsubseteq m_{in}^{*}$) of an input-output model that are isomorphic to the backward matcher of the given transformation rule ($glue \cong rl^{backM}_{trace}$). Note that the containment transitive closure of the input-output model ($m_{in}^{*}$) is considered such that indirect links in the rule can looked for. Additionally, indirect links need to be removed from the input-output models resulting from the match function ($glue_{noind}$). This is do because indirect links are not part of the original input model but rather an auxiliary structure.
% 
% Let us now turn our attention to the apply function in definition~\ref{def:apply_function}. Its role is extend each of the sub-models found by the a match function for a given input-output model and a given transformation rule, such that they become isomorphic to the complete rule (minus its backward links). This process effectively creates the new objects and relations specified in the apply part of the rule for each of the sub-models found by the match function.
% 
% \begin{definition}{Apply Function}
% \label{def:apply_function}
% 
% Let $m_{glue}\in \textsc{Iom}^{sr}_{tg}$ be a input-output model and $rl\in \textsc{Rule}^{sr}_{tg}$ a
% transformation rule. The $apply : \textsc{Iom}^{sr}_{tg}\times \textsc{Rule}^{sr}_{tg}\rightarrow \textsc{Iom}^{sr}_{tg}$ function
% is defined as follows: 
% 
% $$apply_{rl}(m_{in})=\bigsqcup_{m_{glue}\in match_{rl}(m_{in})}trace_{a_{\Delta}}(m_{glue}\sqcup a_{\Delta})$$
% 
% \begin{center}
% where $a_{\Delta}\in \textsc{Iom}^{sr}_{tg}$ is such that $m_{glue} \sqcup a_{\Delta}\cong rl_{noind}$.\\
% \end{center}
% 
% We impose that any instance of $a_\Delta$ is always disjoint from the $m_{in}$
% input-output model and also that any two instances of $a_\Delta$ used in the
% large union are always disjoint. Partial function $trace:\textsc{Iom}\times
% \textsc{Iom} \nrightarrow \textsc{Iom}$ is such that $trace_{\langle
% V_{\Delta},E_{\Delta},st_{\Delta},\tau_{\Delta}\rangle}(\langle
% V,E,st,\tau\rangle) = \langle V,E',st',\tau'\rangle$ where we have that
% $E\subseteq E'$, $st\subseteq st'$ (using a light notational abuse for the $s\subseteq s'$
% and $t\subseteq t'$), $\tau \subseteq \tau'$ and if $v_1\xrightarrow{e}
% v_2\in E'\setminus E$ then $v_1\in Output(V_{\Delta})$, $v_2\in Input(V)$ and
% $\tau'(e)=trace$. Finally, $rl_{noind}$ is a version of $rl$ where indirect
% links have been removed.
% % More concretely, if we have $mam = $\big\langle V,E,st,\tau, Input,Output\big\rangle\in MAM^{sr}_{tg}$ where $Input=\langle V',E',st',\tau'\rangle\in MODEL^{sr}$ and $Output=\langle
% % V'',E'',st'',\tau''\rangle\in MODEL^{tg}$, then $trace(mam) = \big\langle V,E,st,\tau''',Input,Output\big\rangle$ where $E'\subseteq E$ and $v\rightarrow E'\setminus E$ 
% \end{definition}
% % 
% % it builds traceability edges between vertices of the output model that were newly built by the $apply$ function (the vertices in ${mg}_{\Delta}$) and all the vertices from the input part of the input-output model. It is necessary to remove the indirect links from the rule because otherwise they would be created during the rewriting by the apply function. 
% 
% In definition~\ref{def:apply_function} $a_\Delta$ is an input-output model that contains an instance of the target metamodel. These instances are created by rule $rl$ and are used to extend the sub-models found by the match function. The $trace$ function builds traceability edges between vertices of the output model in $a_{\Delta}$ and all the vertices from the input part of a model fragment found by the match function.
% 
% Note that, because we do not pose any constraints on $a_\Delta$ other than the fact that its union with the sub-model $m$ is isomorphic to $rl_{noind}$, the $a_\Delta$ variable can always be satisfied by an unlimited amount of input-output models. In order to avoid an infinite amounts of results when a transformation rule is executed, in what follows we will consider transformation results differ only up to typed graph isomorphism.   
% 
% %  Note that because the vertices are not disjoint, when the graph produced by the apply function is united with the original match apply model the new vertices are created. It was necessary to remove the indirect links from the rule during rewriting, because otherwise those indirect links would also be rewritten.
% 
% Definitions~\ref{def:match_function} and~\ref{def:apply_function} are complementary: the former gathers all the sub-models of an input-output model that are matched by a transformation rule; the latter glues on the output part of each of those sub-models new objects and relations created by a transformation rule. 
% 
% %  The $strip$ function is used to
% %  enable matching over backward links but not elements to be created by the
% %  transformation rule. The $back$ function connects all newly created vertices
% %  to the elements of the source model that originated them.
% 
% %\begin{definition}{Transformation}
% 
% %Let $g\in TG$ be a typed graph. The graph transform function $transform : TR\times TG\rightarrow TG$ is recursively defined as:
% %\begin{gather*}
% %  transform_{(m,a)}(g) =
% %  \begin{cases}
% %    g  & \text{if } match_{m}(g)=\emptyset \\
% %    instance(a)\cup transform_{(m,a)}(g') &\text{if } match_{m}(g)\neq \emptyset
% %  \end{cases}
% %\end{gather*}
% %$$\text{where } g'=mark_{(m)}(g)$$
% %\end{definition}
% 
% %\subsection{Transformation Semantics}
% %\begin{definition} {Layer Semantics}
% 
% %Let $l\in Layer$ be a layer and $\{m,m'\} \in MAM^{s}_{t}$ be a match
% %$$\frac{}
% %{m,\emptyset \stackrel{layerstep}{\rightarrow}m}$$
% %$$\frac{tr\in layer,\; transform_{tr}(m) = m''\; m'',layer\backslash\left\{tr\right\} \xrightarrow{layerstep}	 m'}{m,layer \xrightarrow{layerstep} m'}$$
% %\end{definition}
% 
% \begin{definition} {Layer Step Semantics}
% \label{def:layer_step_semantics}
% 
% Let $l\in \textsc{Layer}^{sr}_{tg}$ be a Layer. The \emph{layer step relation}
% $\stackrel{layerstep}{\rightarrow}\subseteq \textsc{Iom}^{sr}_{tg}\times \textsc{Iom}^{sr}_{tg} \times \textsc{Layer}^{sr}_{tg}\times
% \textsc{Iom}^{sr}_{tg}$ is defined as follows:
% 
% $$\frac{}
% {\langle m_{in},m_{glue},\epsilon\rangle \xrightarrow{layerstep}m_{in} \sqcup m_{glue}}$$
% 
% $$\frac{\begin{array}{ll}& rl\in l,\; apply_{rl}(m_{in}) = m_{rout},
% \langle m_{in},m_{glue}\sqcup m_{rout},l\backslash \{rl\}\rangle \xrightarrow{layerstep} m_{out}
% \end{array}}
% {\langle m_{in},m_{glue},l\rangle \xrightarrow{layerstep} m_{out}}$$
% 
% \begin{center}
% where $m_{rout} \in\textsc{Iom}^{sr}_{tg}$ and $rl\in \textsc{Rule}^{sr}_{tg}$.
% \end{center}
%  
% We impose that all input-output models that are part of $rout$ and have been generated by rule $rl$ are disjoint from input-output models accumulated in $m_{glue}$ that have been generated by other rules.
% 
% \end{definition}
% 
% In definition~\ref{def:layer_step_semantics} we build the result of executing a layer of a DSLTrans transformation. The operational semantics-like rules in the definition execute each rule $rl$ in layer $l$, in a non-deterministic order, by using the $apply$ function. The result of executing each rule is accumulated in the temporary $m_{glue}$ input-output model. Finally, when the set of transformation rules in the layer has been exhausted, the result of executing all the rules in the layer (now contained in $m_{glue}$) is united with the input-output model $m_{in}$, the input to layer $l$. Note that this final union produces the result we expect because of the fact that the $m_{glue}$ input-output model is not disjoint from $m_{in}$. The common ``glue'' parts of $m_{glue}$ that have been built by the match function and extended by the apply function are now used to built the result of executing layer $l$.
% 
% Definition~\ref{def:layer_step_semantics} is the core of DSLTrans' semantics. Many model transformation languages are based on graph rewriting, where the result of each rule rewrite is immediately usable by all other rules. In DSLTrans the result of executing one layer in DSLTrans is totally produced before the input to the layer is changed. This is enforced in definition~\ref{def:layer_step_semantics} by the fact that the apply function always executes over the same $m_{in}$ input-output model and all the results of rule execution in the same layer are added to the $m_{glue}$ structure that is write-only. Rules belonging to the same layer are thus forced to execute independently.
% 
% 
% \begin{definition} {Transformation Step Semantics}
% \label{def:transformation_step_semantics}
% 
% Let $[l::tr]\in \textsc{Transf}^{sr}_{tg}$ be a transformation, where $l\in
% \textsc{Layer}^{sr}_{tg}$ is a Layer and $tr$ also a transformation. The \emph{transformation step relation}
% $\stackrel{trstep}{\rightarrow}\subseteq \textsc{Iom}^{sr}_{tg}\times \textsc{Transf}^{sr}_{tg}\times
% \textsc{Iom}^{sr}_{tg}$ is defined as follows: 
% $$\frac{}{\langle m,[]\rangle \xrightarrow{trstep} m}$$
% 
% $$\frac{\big\langle m_{in},\epsilon,l^{\star}\big\rangle \xrightarrow{layerstep} m_{inter},\; \langle m_{inter},R\rangle \xrightarrow{trstep} m_{out}}{\langle m_{in},[l::T]\rangle \xrightarrow{trstep} m_{out}} \hspace{.3cm} \text{where} \hspace{.3cm} l^{\star}=\bigcup_{rl\in l}rl^{\star}$$
% \end{definition}
% 
% While the execution of the rules belonging to a layer happens in parallel, the execution of the layers of a transformation happens sequentially. 
% As per definition~\ref{def:transformation_step_semantics}, the input-output model $m_{inter}$ is the output of executing a given layer that is passed onto the next layer as input. The transformation rules in a layer are expanded before execution ($l^{\star}$) such that polymorphism in match elements can be handled (see definition~\ref{def:transformation_rule_expansion}).

\subsection{Semantic Transformation Constructs}
\label{sec:DSLTrans_semantics}

In the definition that follows we introduce the notion of execution of a DSLTrans model transformation. For our purposes it is sufficient to introduce it as an input-output model (see \cref{def:input_output_model}), containing the input model for the transformation, the produced output, and the traceability links built during execution. Due to space limitations, we cannot introduce the semantics of DSLTrans in the main text of this paper. We thus refer the reader to \cref{sec:DSLTrans_formal} for a formal description on how DSLTrans transformation executions are built.

\begin{definition} {Model Transformation Execution\\}
\label{def:modeltransformation} 
\CatchFileBetweenTags{\modeltransformation}{text/definitions}{modeltransformation}{\modeltransformation}
\end{definition}

\CatchFileBetweenTags{\modeltransformationtext}{text/definitions}{modeltransformationtext}{\modeltransformationtext}

% A transformation execution is formed from taking the input model, and executing
% the transformation on it to produce the output model. During this
% transformation, traceability links will be placed between match elements and the
% apply elements they create. Definition~\ref{def:transf_ex} expresses this
% formally. Note that we assume that transformation executions are built following
% the algorithm described in~\cite{DBLP:conf/sle/BarrocaLAFS10}.

% \begin{definition}{Transformation Execution}
% \label{def:transf_ex}
% 
% Let $t$ be a DSLTrans transformation having source metamodel $s$ and target
% metamodel $t$. A Transformation Execution is a 6-tuple $\langle V,E\cup
% Tl,\tau,Match,Apply,Tl\rangle$, where $\langle V,E,\tau,Match,Apply\rangle \in
% MAP^{s}_{t}$ is a match-apply pattern. $Match=\langle V',E', \tau',s\rangle$,
% $Apply=\langle V'',E'', \tau'',t\rangle$ and the edges $Tl\subseteq V'\times
% V''$ are called \emph{traceability links}. The set of all transformation
% executions having source metamodel $s$ and target metamodel $t$ is called
% $Exec^{s}_{t}$.
% \end{definition}

% We now prove two important properties about DSLTrans' transformations.
% 
% \begin{proposition}{Confluence}
% 
% Every model transformation is confluent up to typed graph isomorphism.
% \end{proposition}
% 
% \begin{proposition}{Termination} 
% 
% Every model transformation terminates.
% \end{proposition}

% The language to describe properties is in fact very similar to the language to
% express transformations, with the additional possibility of expressing indirect
% links in the $apply$ pattern --- thus allowing more abstract patterns than the
% ones expressed in transformations. This is natural given that the properties of
% a transformation can be more abstract than the rules implementing them.

### 2015-03-13 text/implementation.tex
previous version of the paper

\section{Implementation Details}
\label{sec:implementation}

In this section we will briefly describe our implementation of the algorithms described in the above sections. In particular, we highlight optimizations made and provide results suggesting that our algorithms can feasibly scale to industrial-sized applications.

\subsection{Enabling Technology and Prototype}

In previous work we have reported on the usage of Prolog as a means to build a
proof-of-concept prototype for our technique~\cite{Lucio:10}. The experiments
performed using Prolog were inconclusive regarding the scalability of our
technique given that the path condition construction algorithm as now described in \cref{sec:building_pcs} lacked a formal understanding, as well as several other
imprecisions. As such, no performance optimisations were attempted.

Through our sponsorship by the NECSIS (Network for the Engineering of Complex
Software-Intensive Systems for Automotive Systems) project, we have the
opportunity to apply our verification technique in an industrial setting. In
order to achieve high performance in this setting, despite the complexities of
verification techniques, we were required to choose an underlying efficient
implementation framework. Our goals were to select a framework which: 1) allows
graph manipulation natively. This detaches us from the worries of building and
optimising our own subgraph isomorphism NP-complete algorithms, which are
constantly used during path condition construction; and 2) allows detailed control over
graph manipulation such that the implementation of complex optimizations is feasible. These optimizations are potentially required to apply our technique to large and complex model transformations.

We have chosen T-Core~\cite{DBLP:journals/eceasst/SyrianiV10,SyrianiVanghel:13}
as our graph manipulation framework.
Aside from satisfying our basic requirements described above, T-Core allows
for native rewriting of typed graphs, which considerably eases our implementation
effort. The algorithms described throughout this work have been implemented by scheduling T-Core
graph manipulation primitives using the Python programming language.

\subsection{Complexity}
\label{sec:complexity}

Let us motivate our discussions of optimisation and performance by providing an
approximate formula for the complexity of the path condition construction and
property proof algorithms presented in Sections~\ref{sec:building_pcs}
and~\ref{sec:verif_dsltrans_props}.

\subsubsection{Path Condition Generation}
Recall that a DSLTrans transformation is composed of rules arranged in layers.
The path condition generation algorithm described in \cref{sec:building_pcs}
moves through these layers and combines rules into viable path conditions.

Let the number of rules in the transformation be $r$.  Then, the maximum number
of path conditions that can be created is $2^{r}$. Each path condition will
either represent a rule or not, and therefore the $2^{r}$ path conditions
represent all possible rule combinations. Note that this case assumes that all
path conditions are viable. In practice, unsatisfied rule dependencies will
prevent some rule combinations, reducing the number of path conditions created.

As discussed in \cref{sec:building_pcs}, the path condition generation algorithm builds these path conditions by
considering all possibilities of how a rule can combine with a path condition. This
combination step is composed of two algorithmic components. The first is to
determine all positions where the rule matches over the path condition. Let this
matching step be $m$. Note that this matching step is dependent on the size of the rules.

The second step of the combination step is to "glue" the rule at all matching
positions. Let this step be termed $g$. This step is linear in the size of the rule to be glued multiplied by the number of times the rule has matched in step $m$.

Note that $m$ and $g$ could be quite expensive operations. However, in
our implementation, these steps are implemented using the efficient T-Core graph
manipulation framework.

\begin{equation}
\label{eq:complexity}
\mathcal{O}\big({2^r \cdot (m + g)}\big)
\end{equation}

\cref{eq:complexity} presents the time and space complexity for the path condition generation algorithm.

\subsubsection{Property Proof}
\label{subsubsec:complex_prop_proof}
For our property proving algorithm, recall that each path condition created is examined to see if the property in question holds.

As mentioned in \cref{sec:verif_dsltrans_props}, a path condition must be \\matched by the property. However, different elements in the path condition may overlap (have been matched on) on the same elements in the input model, as described in \cref{sec:prop_satisfaction}.
Therefore, an operational step is required to resolve this ambiguity. One solution is to produce all possible path conditions, where for each pair of overlapping elements in a path condition, one new path condition is produced where they are merged, and one new path condition where they are not.

The complexity of this ``disambiguation step'' will be proportional to the average number of overlapping elements in the path condition, and will be denoted by the term $d$ in this discussion. Practically, $d$ will be dependent on how rules are combined during the path condition generation algorithm. Future work will precisely detail how the characteristics of the transformation affect the algorithm's complexity.

The complexity of the matching step will then be linear in the size of the set of path conditions. The property matching step itself ($p$) will then be linear in the size of the property and path conditions. Again, in practice $p$ is implemented using the T-Core framework.

\begin{equation}
\label{eq:complexity2}
\mathcal{O}\big({2^r \cdot 2^d \cdot p}\big)
\end{equation}

\cref{eq:complexity2} shows time and space complexity for the property proving step.

\subsection{Optimisations}
\label{sec:optimization}

In order to tackle the time and space complexities of the path condition
construction and property proof algorithms we have employed several engineering 
strategies. In the following paragraphs we describe the most relevant of these
strategies.

\begin{itemize}
\item Path condition construction and property proof are very repetitive
processes since most individual rules are often composed and searched in 
the same manner. Since many similar situations have to be investigated during
path condition construction and property proof, memoisation was used whenever
possible to avoid isomorphic graph matching and rewrite operations.
As such caching is heavily used in both algorithms;

\item In \cref{subsubsec:complex_prop_proof} we detail how the overlap of elements in a path condition can be operationally handled by producing two new path conditions for each pair of overlapping elements. Given this procedure is recursive and presents exponential
time complexity, we have performed this ``disambiguation'' step only when strictly necessary:  when performing
property verification on a path condition that contains the elements in
the property.
Note that the fact that  disambiguation is performed in this lazy fashion
allows us to operationally keep path conditions as sets of individual rules.
This makes it possible to heavily reuse pointers to the original transformation
rules when building path conditions, thus reducing the algorithm's space complexity when compared to the explicit representation of
each generated path condition. This also means that, practically,
path condition disambiguation is mostly done on demand during property proving;

\item For property proof we have implemented a strategy to avoid checking path
conditions where the property is sure to hold. The strategy is based on the fact
that if a path condition $B$ contains the same elements as a path condition
$A$ where the property has already been checked successfully, and no additional
elements of the property exist in $B$, then the property also holds
for $B$.
\end{itemize}

\section{Experiments}
\label{sec:experiments}

This section will detail the experiments we performed in order to measure the performance of our technique. We present timing results for two experiments. The first is to obtain timing results for proving two properties on a synthetic transformation, while the second experiment is sourced from our industrial partners.

\subsection{Experimental Setup and Results}

The complexity of \cref{eq:complexity} and \cref{eq:complexity2} suggest that our property proving
approach is intractable in the general case. However, we have provided in \cref{sec:optimization} a number
of concrete optimisations to allow us to prove properties on transformations of
non-trivial size. This section will detail our experiments to determine the
effect of the number of rules in the transformation on the performance of our
implementation.

For our experiment we have used the Police Station transformation as described
in \cref{sec:dsltrans} as a sample transformation. However, in order to
determine the performance characteristics of our approach, we have replicated
the rules within the transformation.

This was achieved by synthetically augmenting the original metamodels by
replicating their elements twice, thus building source and
target metamodels that are three times larger. For example, in the source
metamodel we will now have \emph{Station1} (renamed from the original
\emph{Station} class) and its replicas \emph{Station2} and \emph{Station3}.
These three metamodel elements are distinct from each other and are formally
three different types. We have also added new rules that utilise these new
types, as seen in \cref{fig:replicated_police_station_trafo}.

\begin{figure}[bht] \centering \includegraphics[width=0.48\textwidth]{./figures/policestation_dsltrans/3runtransformation.pdf}
	\caption{Replicated Police Station transformation for performance tests}
	\label{fig:replicated_police_station_trafo}
\end{figure}

Note that for clarity reasons in
\cref{fig:replicated_police_station_trafo} we have abbreviated the element
names \emph{Station}, \emph{Male} and \emph{Female} to \emph{S}, \emph{M} and
\emph{F} respectively. Additionally, the numerical suffix denotes which
replicated metamodel element is represented, as described above.

\renewcommand{\baselinestretch}{1.0}

\subsubsection{Results}

\begin{table*}[htb]
\label{table:results}
\centering
\begin{tabular}{|c|r|r|r|r|r|}
\hline
\rowcolor{Gray}
\parbox[t]{0.04\textwidth}{\# of rules\\} & 
\parbox[t]{0.13\textwidth}{\# of path conds. created\\} 
& \parbox[t]{0.15\textwidth}{Path conds. build time (s)\\}
& \parbox[t]{0.11\textwidth}{Memory used (KB)\\}  
& \parbox[t]{0.18\textwidth}{Proof time for property that holds (s)\\} 
& \parbox[t]{0.20\textwidth}{Proof time for property that does not hold (s)\\}\\
\hline
3	&	8		&	$<$0.01		&	0.08	&	-		&	-		\\
5	&	16		&	0.13		&	0.09	&	0.19		&	0.003	\\
7	&	34		&	0.39		&	0.17	&	1.26		&	0.003	\\
10	&	272		&	1.87		&	1.24	&	2.40		&	0.003	\\
12	&	442		&	2.68		&	1.83	&	3.40		&	0.003	\\
14	&	1156	&	9.00		&	4.98	&	8.38		&	0.003	\\
17	&	9248	&	59.08		&	38.01	&	73.51	&	0.003	\\
19	&	15028	&	97.52		&	60.10	&	140.77	&	0.003	\\
21	&	39304	&	369.19		&	156.79	&	412.02	&	0.003	\\
\hline
\end{tabular}
\caption{Results for creating the set of all path conditions and proving two properties}
\label{tab:scalability_results}
\end{table*}

The results in \cref{tab:scalability_results} were obtained by verifying
the properties in Figures~\ref{fig:dsltrans_prop1}
and~\ref{fig:dsltrans_prop2} on the transformation seen in
\cref{fig:replicated_police_station_trafo}. The experimental platform was
a 2.2 GHz Intel Core i7 machine with 8GB of DDR3 memory running Ubuntu 11.10 and
Python 2.7. For each measurement involving time, we repeated the given
experiment three times and calculated the final result as the average of the
three experiment results. The code used to run our experiments can be found
at~\cite{DSLTransVerif:13}.

\subsubsection{Time Required to Produce Path Conditions}

An important metric for our work is measuring how long it takes to produce the final set of path conditions from a DSLTrans transformation. As seen in \cref{sec:complexity}, this metric depends on the composition of the rules in the transformation's layers.

The first column of \cref{tab:scalability_results} shows the number of
rules for each part of the experiment. In order to provide greater granularity in the data, and determine the precise effect of adding more rules to a layer versus adding another layer of rules, we examine subsets of rules taken from \cref{fig:replicated_police_station_trafo}. For example, the subset with five rules contains the three first rules of layer 1 plus the two first rules of layer 2; the subset with seven rules contains the first three rules of layer 1 plus the four first rules of layer 2; and so on.

\begin{figure*}[htb]
        \centering
        \begin{subfigure}[b]{0.26\textwidth}
                \centering
                \includegraphics[width=1\textwidth]{./figures/results/rules_vs_pcs.pdf}
                \caption{Number of rules vs. \\path conds. created}
                \label{fig:rules_vs_pcs}
        \end{subfigure}%
        ~~
        \begin{subfigure}[b]{0.26\textwidth}
                \centering
                \includegraphics[width=1\textwidth]{./figures/results/rules_vs_time.pdf}
                \caption{Number of rules vs. \\time taken}
                \label{fig:rules_vs_time}
        \end{subfigure}%
        ~~
        \begin{subfigure}[b]{0.26\textwidth}
                \centering
                \includegraphics[width=1\textwidth]{./figures/results/rules_vs_memory.pdf}
                \caption{Number of rules vs. \\memory used}
                \label{fig:rules_vs_memory}
        \end{subfigure}%

        \caption{Metrics for the path condition creation process}
        \label{fig:path_condition_building}
\end{figure*}

\cref{fig:rules_vs_pcs} presents the number of path conditions created for
a given number of rules, while \cref{fig:rules_vs_time} graphs the time
taken to create all the path conditions. Both the number of path conditions and
the time required to build them rise steeply with the number of rules, but it
is quite feasible to build path conditions and prove properties for up to 21
rules. As shown later in the section on industrial experimentation, 21 rules
exceeds the number of rules in our industrial case study. It also exceeds the
number of rules in several useful DSLTrans
transformations~\cite{febavava:10,dsltrans_manual,zhang:ACP_APN:11}.

\cref{tab:scalability_results} and \cref{fig:rules_vs_memory} demonstrate that memory consumption is very modest, remaining well under a megabyte for thousands of path conditions. This is due to the optimisations that we perform, such as only storing pointers to path conditions. We are encouraged that this algorithm can scale extremely well in terms of respecting memory constraints.

\subsubsection{Time Required to Prove Properties}

We now examine the time it takes to prove two properties on the transformation based on the number of path conditions created from that transformation. The two properties to be proven are shown in \cref{fig:properties} in \cref{sec:dsltrans}. The first, in \cref{fig:dsltrans_prop1}, is a property that we expect to hold for all path conditions. \cref{fig:dsltrans_prop2} shows a property that we expect to \emph{not} hold for all path conditions.

\begin{figure}[bht] \centering \includegraphics[width=0.35\textwidth]{./figures/results/pcs_vs_prop1.pdf}
	\caption{Time required to prove the property that holds on all path conditions}
	\label{fig:pcs_vs_prop1}
\end{figure}

\cref{fig:pcs_vs_prop1} shows the time in seconds required to prove the property that holds on all path conditions, as seen in the fifth column in \cref{tab:scalability_results}. Note that the time taken increases linearly with the number of path conditions to examine. This increase occurs as each property must be checked to ensure that the property will hold.

In contrast, the time required to disprove the property that does not hold is roughly constant. This can be seen in the sixth column in \cref{tab:scalability_results}, where this proof took 0.003 seconds regardless of the number of path conditions examined. This is due to the fact that, given the property does not hold, the proof algorithm can stop as soon as a counterexample is found. The very short time to disprove the property is due to the fact that path conditions are checked for the property sequentially, in the order they are produced. In our example, a counterexample can be found very early in the set of path conditions. Note that a more complex property that involves rules which would appear only much later in the generated set of path conditions would require a longer time to reach a counter-example.

Experiments were also undertaken to determine what effects the size of the property to be proved has on the running time of the algorithm. Preliminary results indicate that an increase of the property size results in a proportional increase in running time. This is to be expected, as the underlying graph matching algorithm has to match more elements to determine if the property holds or not.

% \levi{Do you want to keep this?}
% Finally, one last observation on the values in
% table~\ref{tab:scalability_results} is the fact that when rules without backward
% links are added to the transformation (the jumps from~7 to~10,~14 to~17 and~21
% to~24 rules) both the number of produced path conditions and the used memory
% increases around tenfold. Comparatively, when rules with backward links are
% added to the transformation (the jumps from ~3 to~7,~10 to~14 and 17 to~21
% rules) both the number of path conditions and the used memory only increases
% around five times. This is despite the fact that we increase rules with backward
% links by sets of 4 rules, whereas we increase rules without backward links by
% sets of 3 rules. The reason for this difference is the fact that, during
% symbolic execution construction, a rule with backward links may simply not be
% executable for a given path condition or can be merged with a previous path
% conditions. In both these cases no new path conditions are added to the symbolic
% execution. However, a rule without backward links necessarily generates twice
% the path conditions already generated by the symbolic execution algorithm as it
% may always be executed.
% 
% \levi{talk about better time optimizations here}It is however the case that, if
% no or few collapse operations are required for a path condition, the algorithm
% still attempts to collapse rules in this fashion. This may result in overhead as
% compared to merging all the rules and recursively collapsing.

\subsection{Industrial Experimentation}
\label{sec:industrial_exp}

Aside from the experiments with the police station transformation we have
reported in the previous section, we have applied our technique in the context
of the NECSIS project.
The experiment regards a DSLTrans transformation that maps between subsets of a
proprietary metamodel from General Motors, describing legacy automotive configuration data, and
the AUTOSAR metamodel, an open platform shared by car manufacturers. This DSLTrans mapping transformation includes seven rules, distributed among three
layers. Further details of this experiment can be found in~\cite{gehan:13} and a complete description of the transformation can be found in~\cite{SelimWCD12}.

Our path condition generation approach generates a set of three path conditions for the transformation in approximately 0.8 seconds. This low number of path conditions is
due to the fact that several rules overlap, as explained in \cref{def:layer_transformation} of \cref{sec:formal_background}. Such overlapping causes the number of formed path conditions to be smaller than in the case where no overlaps occur, as certain combinations need not be considered due to rule dependency.

In~\cite{conf/gg/SelimLCDO14} we also describe the proof of nine properties (multiplicity invariants, security invariants and pattern contracts) that demonstrate several aspects of the correctness of this mapping transformation. The proof of these properties on all executions of a migration transformation is of interest to our industrial partners in order to ensure that the migration does not add extraneous elements or delete any needed information.

All properties were proved in around 0.02 seconds by our approach, and were expressed using a propositional logic extension to the property language that we present in this paper. Note that, despite the fact that not all aspects of the case study (overlapping rules, propositional logic extension) are considered theoretically in this manuscript, the obtained results are nonetheless very interesting in terms of the experimental scalability of our approach. In particular, we note that our verification approach performs orders of magnitude faster compared to an ATL-based verification tool that verified the same transformation~\cite{conf/gg/SelimLCDO14}.

\subsection{Discussion}

The experimental results of verifying the test Police Station transformation
portrayed in the graphs of \cref{fig:path_condition_building} show that,
as predicted by complexity \cref{eq:complexity}, both the path condition
construction time and the number of created path conditions grow exponentially
with the number of rules.
In \cref{fig:pcs_vs_prop1} we can also see that property proving time for
properties that hold increases linearly with the number of path conditions, as
was also theoretically predicted in \cref{sec:complexity}. Despite the exponential time
and space complexities of the path condition construction algorithms, our
experiments suggest that real-world sized model transformations can be tractable
by employing the optimizations described in \cref{sec:optimization}.
We also believe further optimization opportunities of our algorithms exist and that the number of rules handled by our approach can be driven higher.

The industrial case study presented in \cref{sec:industrial_exp} suggests
that validation of practical model transformations is not always very
computationally expensive. In fact, the properties we have proved in this
industrial case study are of practical use for our partners, yet required only fractions of seconds to prove.

From the differences in the examples we have presented in this section and from
our experience with building DSLTrans transformations we believe that the
complexity of verifying real-word model transformations can vary within a wide
range. The complexity of \cref{eq:complexity} provides us a referential that can
be used when evaluating the theoretical and operational complexities of
verifying further case study transformations. This complexity is influenced by several parameters that describe the shape of a model
transformation, and we believe the study of those parameters in further case studies is very
important. Refining \cref{eq:complexity} will provide better precision in our theoretical estimations and also direct our
optimization efforts by understanding what transformation parameters have the
highest impact on performance.

%  Refine the complexity formula to get better precision. Our problem
% is tractable using some implemention tricks. 
% 
% It's the same order of complexity
% for PC construction and property proof (despite disambiguation being done during
% property proof). The graphs show that we still have exponential curves that are
% flatter than expected due to optimizations. Future work will involve finding the
% more likely values for the parameters in our formula and thus optmize for them.
% This will provide guidance for optimization implementation and provide better
% definition of the expected vs real performance.
% 
% It does match our expectations, but is does not look bad.
% 
%  However, as can be seen in graph\levi{add graph with
% expected vs real number of path conditions and times}, our results are much
% better than predicted which suggests the problem is manageable when the number
% of rules is kept to a reasonable size and given our optimizations introduced in
% section~\ref{sec:optimizations}.

% The expected complexity of building the path
% conditions our experiment involving 21 rules as shown in
% table~\ref{table:results} can be calculated by instantiating the parameters of
% formula~\ref{eq:complexity} as $r=5$ (average number of rules per layer), $l=4$
% (number of layers), $e=2$ (average number of elements having the same type as
% elements in other rules of the same layer), $c=2$ (average pairs of rules per
% layer share common elements). We reach a final value of 5.308.416

% The performance results are promising, but
% further experiments with other transformations need to be done in order to
% assess the real performance of our approach. In fact, depending on the size of
% the rules in the considered model transformation, on rule distribution among
% layers, if those rules involve many backward links or not and whether many
% elements of the same type exist scattered by different rules or not (implying
% many collapse operation), we expect that the number of rules that can be tackled
% by our approach may vary substantially.\levi{need to finish this}

### 2015-03-13 text/introduction.tex
previous version of the paper

\section{\uppercase{Introduction}}
\label{sec:intro}

Model transformations were described as the \emph{heart and soul} of model driven
software development (MDD) by Sendall and Kozaczynski in
2003~\cite{Sendall2003}.  Due to their practicality and appropriate level of
abstraction, model transformations are the current technique for performing
computations on models. In their well-known 2006 paper `A Taxonomy of Model
Transformations', the authors Mens, Czarnecki and Van Gorp call for the
development of verification, validation and testing techniques for model
transformations~\cite{Mens2006125}. Despite the many publications on this topic
since then, the field of analysis of model transformations seems to be still in
its (late) infancy, as evidenced by Amrani, L\'ucio \emph{et al.}~\cite{DBLP:conf/icst/AmraniLSCDVTC12}.

% Since then, many model transformation languages have emerged and are being
% used intensively not only in MDD, but also in other scopes, such as when
% software development processes benefit from formalised translators between
% languages.

In this paper, we present our work on verification of properties on model
transformations. Specifically, we discuss concrete algorithms that can prove
whether properties will hold or do not hold on all executions of a
transformation written in the DSLTrans transformation language. Properties are
proved through a process that constructs a set of path conditions, where
each path condition symbolically represents an infinite number of concrete transformation
executions through an \emph{abstraction relation}.

% The technique has been extended from its original presentation
% in~\cite{Lucio:10}, ~\cite{DBLP:conf/sle/BarrocaLAFS10}, and~\cite{LucVan:13}.

In our previous work~\cite{Lucio:10}, this property-proving algorithm was
presented as a proof-of-concept. In the present work, we significantly expand
that proof-of-concept by clarifying and offering discussions on validity and
completeness for the presented algorithms. We also provide an implementation
that we believe will scale to industrial applications, as validated by an
automotive case study~\cite{gehan:13}.

Our approach is feasible due to the use of the transformation language DSLTrans~\cite{DBLP:conf/sle/BarrocaLAFS10}. DSLTrans is
\emph{Turing incomplete}, as it avoids constructs which imply unbounded
recursion or non-determinism. Despite this \emph{expressiveness reduction}, we have shown via several
examples~\cite{febavava:10,dsltrans_manual,zhang:ACP_APN:11} that DSLTrans is
sufficiently expressive to tackle typical translation problems. This sacrifice
of Turing-completeness allows us to construct a\\provably-finite set of path
conditions~\cite{Lucio:10}. Our approach currently considers a core subset
of the DSLTrans language that does not include negative conditions in rules or attribute manipulation. These features of the language will be addressed in future work.

Informed by the structure of DSLTrans transformations, our approach defines an algorithm for the creation of path conditions. Each path condition that is created represents a set of concrete
transformation executions through an \emph{abstraction relation} that we formally define. Once the set of all path conditions has been created for the transformation, we can then prove structural \emph{model syntax
relations}~\cite{DBLP:conf/icst/AmraniLSCDVTC12} using this relation. Such properties are essentially pre-condition/post-condition axioms involving statements about whether certain elements of the input model have been correctly transformed into elements of the output model, and have been explored by several authors~\cite{Akehurst:02,Narayanan:07,DBLP:journals/eceasst/CariouBBD09,DBLP:journals/ase/GuerraLWKKRSS13}.
In our proof technique, the properties examined can be proven to hold for all executions of a given
model transformation, no matter the input model.
Therefore, our technique is \emph{transformation dependent} and \emph{input
independent}~\cite{DBLP:conf/icst/AmraniLSCDVTC12}.

Our methods differ from previous work in the transformation verification field
in that we require no intermediate representation for a specific proving
framework (as in~\cite{ButtnerEC12,ButtnerECG12,DBLP:conf/icst/AsztalosLL10}) but instead work on DSLTrans
transformations themselves. Along with DSLTrans rules, all of the constructs
involved in our algorithms are typed graphs. This intuitive representation allows
our property proving technique to be composed of relatively simple steps, as the
metamodels, models, and properties involved are all constructed using a similar
graphical representation.

% This also provided us with performance gains over our
% previous implementation, due to the use of efficient graph-matching libraries to manipulate the graph structures required.

A large difficulty in any exhaustive proof technique is the tendency for the
state space to explode, even when abstractions are performed to render the
search space finite.  A later section of this work will discuss optimisation
opportunities and performance results obtained from our implementation. The
scalability of our approach will also be analysed in order to infer the
algorithm's potential applicability to real-world problems. A real-world industrial case study will also be briefly presented.

Our specific contributions include:

\begin{itemize}
\item An algorithm for constructing all path conditions for a given DSLTrans transformation;
\item An algorithm that proves transformation properties over these path conditions;
\item Validity and completeness proofs of the path condition construction and property proving algorithms;
\item A discussion of performance and scalability results for our implementation.
\end{itemize}

This paper is organised as follows: \cref{sec:dsltrans} briefly introduces
the DSLTrans model transformation language and its formal semantics, while the formal background for this work is presented in \cref{sec:formal_background}. The algorithms to build the complete set of path
conditions for the transformation will be discussed in
\cref{sec:building_pcs}. \cref{sec:abstraction_relation} will present the abstraction relation found in our technique, along with examples, while \cref{sec:verif_dsltrans_props} will examine how this abstraction relation is used in our process for proving properties. In \cref{sec:implementation} and \cref{sec:experiments}, we introduce our implementation with sample scalability results; \cref{sec:related_work} presents the related work;
and finally in \cref{sec:conclusion} we conclude with remarks and future
work.

### 2015-03-13 text/overview.tex
previous version of the paper

\section{Technique Overview}
\label{sec:overview}

This section provides a brief overview of the transformation verification
technique with each component algorithm further elaborated in the sections that follow.

\subsection{Symbolic Execution}

Our algorithm operates on the principle of symbolic execution. In order to
explain the concept of symbolic execution of a transformation, let us make an
analogy with program symbolic execution as introduced by King in his seminal
work ``\emph{Symbolic Execution and Program
Testing}''~\cite{DBLP:journals/cacm/King76}. According to King, a symbolic
execution of a program is a set of \emph{constraints} on that program's
\emph{input variables} called \emph{path conditions}. Each \emph{path condition}
describes a traversal of the conditional branching commands of that program. A
\emph{path condition} is symbolic in the sense it \emph{abstracts} as many
concrete executions as there are instantiations of the path condition's
variables that render the path condition's constrains true.

We can transpose this notion of symbolic execution to model transformations. The
analog of an input variable in the model transformation context are
\emph{metamodel classes, relations and attributes}. As program statements impose
constraints on input and output variables during symbolic execution,
transformation rules impose conditions on which metamodel elements are
instantiated during a concrete transformation execution, and how that
instantiation happens.
Such conditions, built from transformation rules, can then be used as the analog
of program \emph{path conditions} in the context of model transformations. In
fact, throughout this paper we also use the term
\emph{path condition} in the context of model transformation verification.

As in program symbolic execution, each transformation path condition
\emph{abstracts} as many concrete executions as there are input/output models
that satisfy them. We must not forget that in a model transformation, rules are
implicitly or explicitly scheduled. Such control and/or data dependencies must
also be taken into consideration during path condition construction. In what
follows we will examine in more detail how these symbolic execution principles can
apply to the verification of model transformations.

% The correspondence between the symbolic
% execution concepts for \emph{programs} and \emph{model transformations} is shown
% in table~\ref{tab:symb_exec_analogy}.
% 
% \begin{table}[h] \centering
% \begin{tabular}{|l|l|l|}
% \hline
% \rowcolor{Gray}
%  & Program & Transformation\\
% \rowcolor{Gray}
%  & Symbolic Execution & Symbolic Execution\\
% \hline
% \hline
% Abstraction Over & Sets of data values & Sets of models (graphs)\\
% \hline	
% Input Variables & Programming language variables & Metamodel classes,\\
%  & (int, string, float, etc) & relations and attributes\\
% \hline
% Constraints & Predicates on variables imposed by & Metamodel patterns imposed\\
% & assignment or conditional statements & by transformation rules\\
% \hline
% Path Conditions & Conjunction of predicates & Conjunction of metamodel\\
% & on variables & patterns\\
% \hline
% \end{tabular}
% \caption{\emph{Program} Symbolic Execution vs. \emph{Model Transformation} Symbolic Execution}
% \label{tab:symb_exec_analogy}
% \end{table}
% 
% \renewcommand{\baselinestretch}{1.5}
% 
% \bentley{Need to connect these section}

\subsection{DSLTrans Rule Sequences}

Transformations are composed of a set of rules, which are applied to the input
model in some sequence depending on the elements present in that model.

However, the sequence of rules that define a transformation execution may be infinite, as rules in an arbitrary transformation language may be repeatedly applied. The transformation language used here,
DSLTrans, is constructed so that each sequence of rules is finite. DSLTrans is explored more fully in the next section. One
important property of DSLTrans to highlight here is that rules are executed layer-by-layer, so that
rules from layer 2 cannot execute before rules from layer 1. Figure~\ref{fig:dsltransformation2} shows an abstract transformation with four rules arranged on two layers.

\begin{figure}[h!]
	\centering
		\includegraphics[width=0.4\textwidth]{./figures/overview/abstractDSLTransTransformation.pdf}
	\caption{An abstract DSLTrans model transformation}
	\label{fig:dsltransformation2}
\end{figure}

Each layer is composed of a set of transformation rules which execute in a
non-deterministic order but must produce a deterministic result. As the order of rule execution within a
layer will not matter, this drastically reduces the number of transformation rule sequences we need to examine. As well, DSLTrans transformations are not Turing-complete. As discussed in~\cite{DBLP:conf/sle/BarrocaLAFS10},
non-completeness is required to make rule sequences finite, but yet still allows
for appropriate expressiveness.

Besides the fact that DSLTrans' transformations are free of constructs that
imply unbounded recursion or non-determinism, DSLTrans' transformations are strictly outplace, meaning no changes are allowed
to the input model. However, the output metamodel for a DSLTrans transformation can
be the same as the input metamodel. Also, elements cannot be removed
from the output metamodel as the result of applying a DSLTrans rule.
This restriction is consistent with the usage of model transformations as
translations~\cite{AMT2012}, as no deletion of output elements is strictly required.

% This is
% however not the case when transformations are used to encode operational
% semantics (simulations) of systems~\levi{cite intent paper}. This illustrates the boundaries of the
% applicability of DSLTrans and that expressiveness reduction entails a compromise
% with the class of problems that can be tackled.

\subsection{Building Path Conditions}

As mentioned, a DSLTrans transformation will have rules structured in a number of layers, so as to define rule ordering. In order to create all possible path conditions defined by this transformation, path conditions will be created in a corresponding layer-by-layer manner. We provide a brief overview here, while a precise explanation follows in section~\ref{sec:building_pcs}.

\subsubsection{Rule Combinations}

In order to provide an intuition for path condition creation, we shall discuss a method to create path conditions for the first layer only. The technique is to produce all possible combinations of rules, using the powerset of rules. These rule combinations can then be made into path conditions by taking the union of all rules in that rule combination. Figure~\ref{fig:first_layer} shows how the possible rule combinations for a layer.

\begin{figure}[h!]
	\centering
		\includegraphics[width=0.5\textwidth]{./figures/overview/rule_combos.pdf}
	\caption{Creating path conditions through rule combinations}
	\label{fig:first_layer}
\end{figure}

An important characteristic of our technique is that the number of times a rule executes is abstracted over in a rule combination or path condition. If a rule's execution is represented by a path condition, the rule has executed one or more times. For instance, the example path condition \emph{AC} in Figure~\ref{fig:first_layer} represents all transformation executions where both rules A and C have executed one or more times, and rule B has not executed.

\subsubsection{Generating All Path Conditions}

The creation of path conditions through rule combinations is only a valid process for the first layer, and is mentioned solely as an explanatory tool. In further layers, DSLTrans rules may contain dependencies which prevent them from executing. Therefore, the algorithm must move layer-by-layer through the transformation, and determine which rules may execute.

The path condition generation algorithm begins by creating a set containing an empty path condition. This path condition represents all transformation executions where no rules have yet executed. The path condition is then sequentially combined with each rule from the first layer of the transformation. The combination process determines all the possible ways that the rule can be added to the path condition, as detailed in section~\ref{sec:building_pcs}.

For example, consider figure~\ref{fig:layers_pc}. The path condition PC at the top of the figure may either be the empty path condition, or a path condition from the previous layer. PC is then combined with a rule R1 
selected from the current layer. The path condition and rule may combine in different ways. For instance, the rule might not execute, in which case the path condition is left unchanged. Another case is that the rule does execute, in which case the rule is ``glued'' to the path condition in a particular way. This combination process will create one or more path conditions which take into account all possibilities of how the rule might execute.

In ~\ref{fig:layers_pc}, combining the path condition PC with rule 1 has produced three path conditions: the unchanged PC, and two different possibilities of how the rule combined with PC (denoted PC + R1 and PC + R1'). These three path conditions are then combined with rule 2. Three further path conditions are produced, representing transformation executions where rule 2 does not execute. As well, three path conditions are produced that represent transformation executions where rule 2 does execute.

\begin{figure}[htb]
\centering
        \begin{subfigure}[b]{0.480\textwidth}
                \centering
                \includegraphics[width=1\textwidth]{./figures/overview/layers_pc.pdf}
                 \caption{Combining a path condition with two rules}
                 \label{fig:layers_pc}
        \end{subfigure}%
        ~~
        \begin{subfigure}[b]{0.48\textwidth}
                        \centering

             \includegraphics[width=1\textwidth]{./figures/overview/all_pcs.pdf}
             \caption{Creating all path conditions for a layer}
             \label{fig:all_pcs}
	    \end{subfigure}%
           \caption{Creating path conditions}
         \label{fig:creating_pcs}
\end{figure}

Rules in DSLTrans may also define dependencies in which elements have been previously created in the transformation. These dependencies further restrict the ways that a path condition and rule may combine. These dependencies are checked against the path condition from the previous layer, not the path combination the rule is to be combined with. This is to preserve the semantics of DSLTrans rules, where a rule cannot depend upon the output of another rule in the same layer.

Figure~\ref{fig:next_layer2} diagrams how the path condition generation process occurs layer-by-layer. The path conditions from the previous layer are combined with all the rules in the current layer, to produce a set of path conditions for the layer. This set will then be combined with the rules from the next layer of rules. After this process repeats for all layers in the transformation, the final set of path conditions is then used for property proving.

\begin{figure}[htb]

        \centering
        \includegraphics[width=0.8\textwidth]{./figures/building_path_conditions/next_layer.pdf}
        \caption{Path conditions are combined with rules in next layer}
        \label{fig:next_layer2}
\end{figure}

\subsection{Proving Properties}

The final set of path conditions produced will define all possible concrete transformation executions through an abstraction relation that we define in Definition~\ref{def:instance_pc_ex}. This abstraction relation allows us to prove properties on each path condition in the final path condition set and have that result hold on all transformation executions abstracted by that path condition. An overview of the property proving process is presented here, while further details as well as formal validity and completeness proofs are found in section~\ref{sec:verif_dsltrans_props}.

Path conditions are represented in much the same way as DSLTrans rules. This is because path conditions represent the interaction of a set of DSLTrans rules. Path conditions contain two graphs. The first graph represents a pattern that must be present in the input model of the transformation. The second is a pattern which will be instantiated in the output model of the transformation, including traceability to elements matched by the rules. The formal definition of a path condition can be seen in definition~\ref{def:path_condition}.

The properties we are interested in have an implication form. They represent the
following statement: if this pattern is found in the input model, then
this other pattern must be found in the output model, possibly including traceability constraints. Therefore,
we encode properties as a set of pre-condition and post-condition patterns.

Property proving then becomes relatively simple. The path condition match graph
represents the elements which define that path condition. Likewise, the property
pre-condition pattern represents the prerequisite for the property. Thus,
property proving algorithm will attempt to isomorphically find the property's
pre-condition pattern in the path condition's match graph. If not found, then the
property will not be validated on this path condition, as the prerequisites do
not exist.

If the property's pre-condition pattern is found, then the property's
post-condition pattern is isomorphically searched for in the path condition. If
it is found, then the rule execution defined by that path condition will produce
the required elements for that property and the property will hold. If not
found, then the necessary elements will not be produced and the property check
will fail.

In this way, all path conditions are checked to see if the property specified is
applicable to that execution, and then if the execution satisfies that property.
Therefore, properties can be verified for all possible executions of a
transformation using the abstraction relation.

Future chapters will examine each step of this algorithm in detail. Following this, results will be provided to indicate that our algorithm is feasible and can scale to industry-sized transformations.

### 2015-03-13 text/property_verification.tex
previous version of the paper

\section{Verifying Properties of DSLTrans Transformations}
\label{sec:verif_dsltrans_props}

% As we have mentioned in section~\ref{sec:dsltrans}, all DSLTrans transformations
% are, by construction, \emph{terminating} and
% \emph{confluent}~\cite{DBLP:conf/sle/BarrocaLAFS10}. We have also abstractly
% shown in~\cite{Lucio:10} how to build a symbolic execution such that
% properties of DSLTrans transformations can be proved. We formally prove
% in~\cite{Lucio:10}, proposition 2, that such a symbolic execution is always
% finite. With our current work we aim at producing a tool that can prove or
% disprove such properties efficiently.

The algorithm presented in \cref{sec:building_pcs} will produce all possible
path conditions for a given DSLTrans transformation. This section will detail
our second contribution: a method to prove properties on the transformation by
examining the path conditions generated for it. We then rely on the abstraction
relation presented in~\cref{def:abstraction_pc_ex} to extrapolate the proof
result to all of the transformation's executions.

The properties we are interested in have an implication form. Similarly to rules and path conditions, properties are largely composed of two patterns. They represent the following statement: whenever this pattern is found in the input model, then
this other pattern must be found in the output model, possibly including traceability constraints.

The property proving algorithm is relatively simple. The match part of a path condition
includes a representation of all the elements and relations ``touched'' in the input model of a set of transformation executions. Likewise, the property's pre-condition pattern represents the prerequisite for the property. Thus, the
property proving algorithm will attempt to find the property's pre-condition pattern in the path condition's match graph. If not found, then the property will not be validated on this path condition, as the prerequisites do not exist.

Whenever the property's pre-condition pattern is found, the property's
post-condition pattern is searched for in the path condition. If
also found, then the rule execution(s) defined by that path condition will produce
the required elements for that property and the property will hold. If not
found, then the necessary elements will not be produced and the property check
will fail for that path condition.

Path conditions for a transformation are checked to understand whether the property of interest holds on all of them. If it does, then by making use of the abstraction relation in \cref{sec:abstraction_relation} between path conditions and transformation executions, we can deduce that the property then holds for all transformation executions. If not, we deduce the property does not hold for at least one transformation execution. Later in this section we will develop a formal argument for why this is true.

\subsection{Structure of a Property}
We will now elaborate on the structures and the relations required for the property proving algorithm. Let us start by precisely defining what a property of a transformation is.

\begin{definition}{Property of a Transformation\\}
\label{def:trans_prop}
\CatchFileBetweenTags{\propertydef}{text/definitions}{propertydef}{\propertydef}
\end{definition}

%Property proving then becomes relatively simple. The path condition match graph
%represents the elements which define that path condition. Likewise, the property
%pre-condition pattern represents the prerequisite for the property. Thus,
%property proving algorithm will attempt to isomorphically find the property's
%pre-condition pattern in the path condition's match graph. If not found, then the
%property will not be validated on this path condition, as the prerequisites do
%not exist.

%If the property's pre-condition pattern is found, then the property's
%post-condition pattern is isomorphically searched for in the path condition. If
%it is found, then the rule execution defined by that path condition will produce
%the required elements for that property and the property will hold. If not
%found, then the necessary elements will not be produced and the property check
%will fail.

%In this way, all path conditions are checked to see if the property specified is
%applicable to that execution, and then if the execution satisfies that property.
%Therefore, properties can be verified for all possible executions of a
%ransformation using the abstraction relation.

%These constraints are expressed as graph patterns on the
%\emph{match graph} and \emph{apply graph} of DSLTrans transformation rules.

% \levi{Needs revision}
% In~\cite{Narayanan:07} Narayanan and Karsai describe the nature of these model
% syntax relation properties, which they call called \emph{structural
% correspondence rules} in their work. As the authors claim, ``\emph{given the
% correspondence conditions are specified in terms of simple queries on the model
% around previously chosen context (metamodel) nodes, we expect that they will be
% easier to specify, and thus more reliable than the transformation itself}''.

In \cref{def:trans_prop}, pre-conditions use the same pattern language as the match graph in
DSLTrans rules, allowing the possibility of including several instances of
the same metamodel element as well as indirect links in the property. Indirect links in properties have
the same meaning as in the rule match graph -- they
involve patterns over the transitive closure of containment links in pre-condition graphs.

Post-conditions also use the same pattern language as the
apply graphs of DSLTrans transformation rules, with the additional
possibility of expressing indirect links in post-conditions. Traceability links can
also be used in properties to impose traceability relations between pre-condition
and post-condition elements.

Note that \cref{def:trans_prop} includes a condition stating a surjective typed graph homomorphism needs to exist between the match part of at least one of the transformation's path condition, and the pre-condition of the property of interest. This condition makes sure that the property's pre-condition can be found at least in one execution of the transformation abstracted (the mathematical argument for this fact is given in the proof of \cref{prop:proof_validity}). This condition makes the checking the validity of a property in the transformation meaningful. If this condition would not be true then it could be that the input pattern required by the property would never be fully matched during transformation execution, making such a property not relevant\footnote{In~\cite{DBLP:conf/sle/BarrocaLAFS10} we have referred to these properties \emph{non-provable}. In the work presented here we explicitely disallow the construction of this class of properties.} for the transformation at hand.

% This last condition explicitly restricts the approach to properties which pre-condition can be \emph{found} in the match part of a path condition of the considered transformation.

% Note that by \emph{found}, we mean that these pre-conditions need to be isomorphically matched on at least one graph resulting from overlapping elements of the same type (but belonging to different rules) in the match part of at least one path condition of the transformation. This overlapping between rules allows taking into account the fact that different rules in a path condition may have match elements of the same type that can potentially match (overlap) over the same class instance in an input model, when a transformation execution is built. This potential overlapping of rules in the path condition is modeled by the surjection in definition~\ref{def:trans_prop} between (a subgraph of) the path condition and the property.

%This would required considering more than one copy of the same rule in the same path condition.  
% \levi{unclear}
% In our symbolic execution algorithm, path conditions are built relying only on
% the rules of the DSLTrans transformation we are analysing. These path conditions
% are built to symbolically represent all possible executions of that DSLTrans
% transformation. Thus, if our prover algorithm replies \emph{yes}, then the
% precondition-postcondition implication expressed in the property will hold for
% all executions of the DSLTrans transformation under analysis. On the other hand,
% if the prover replies \emph{no}, then that will mean that there exists at least
% one exception to the implication in the property. In other words, there exists
% at least one model in which the precondition of the property holds, but the
% postcondition does not. A counter-example can be provided in this case,
% consisting of the set of rules that, when executed, lead to the property being
% violated.

%In figures~\ref{fig:dsltrans_prop1} and~\ref{fig:dsltrans_prop2} we present two
%properties we wish to prove or disprove regarding all executions of the
%transformation presented in \cref{fig:dsltransformation}.\levi{remove figures of properties} The property in
%\cref{fig:dsltrans_prop1} represents the statement ``\emph{a model which includes a
%police station that has both male and female officers will be
%transformed into a model where the male officer will exist in the male set
%and the female officer will exist in the female set}''. This is something
%we expect will always hold in our transformation. The property in
%\cref{fig:dsltrans_prop2} represents the statement ``\emph{any model which includes a
%female officer will be transformed into a model where that female officer will
%always supervise another female officer}'', which is something that we expect
%will hold for our transformation sometimes, but not always.
%
%\begin{figure}[htb]
%        \centering
%        \begin{subfigure}[b]{0.45\textwidth}
%                \centering
%                \includegraphics[width=1\textwidth]{./figures/policestation_dsltrans/satisfied.pdf}
%                \caption{Property 1 -- Expected to hold}
%                \label{fig:dsltrans_prop1}
%        \end{subfigure}%
%        ~~
%        \begin{subfigure}[b]{0.45\textwidth}
%                \centering
%                \includegraphics[width=1\textwidth]{./figures/policestation_dsltrans/unsatisfied.pdf}
%                \caption{Property 2 -- Not expected to hold}
%                \label{fig:dsltrans_prop2}
%        \end{subfigure}%
%        \caption{ Properties to be proved on the Police Station transformation}
%        \label{fig:properties}
%\end{figure}

\subsection{Satisfaction of a Property}
\label{sec:prop_satisfaction}

Let us now detail how a transformation execution is said to satisfy a property. Due to the common structure between properties and transformation executions, this satisfaction is based on whether the property can be isomorphically found in the transformation execution.

\begin{definition}{Satisfaction of a Property by an Execution of a Transformation\\}
\label{def:sat_prop_ex}
\CatchFileBetweenTags{\satpropexe}{text/definitions}{satpropexe}{\satpropexe}
\end{definition}

\Cref{def:sat_prop_ex} states that, every time a graph that is isomorphic to the property's pre-condition is found in (the containment transitive closure of) the input model of the transformation's execution, a graph that is isomorphic to the complete property needs to be found in (the containment transitive closure of) the transformation execution. Note that the last part of the proposition in \cref{def:sat_prop_ex} ensures that the graph that is isomorphic to the property's pre-condition and the graph that is isomorphic to the complete property overlap on their pre-condition parts.

% resulting from the input pattern of $ex$ identified as pre-condition by $f$. Generally speaking, whenever the property's pre-condition is isomorphically found as a pattern of the input model of a execution, the property's post-condition must also be isomorphically found in the output generated from that input pattern. Note that, because a property may contain traceability links between the pre- and post-condition elements, it is necessary to check that those traceability links are also present in the execution. Because of this, the $g$ homomorphism in definition~\ref{def:sat_prop_ex} that checks the property's post-condition is such that it considers the same match pattern identified by the $f$ homomorphism when the pre-condition is checked.

\Cref{fig:prop_execution_without_traceability} demonstrates how a property holds on a transformation execution. Note that the lack of traceability links in the property means no element creation dependencies have been specified. In contrast, the traceability links in the property in \cref{fig:prop_execution_with_traceability} specify that the 'x:X' element must have been created from the 'b:B' element in the transformation execution. This is not the case (as highlighted by the dashed red circle), and therefore the property does not hold on the transformation execution.

\begin{figure}[htb]
        \centering
        \begin{subfigure}[b]{0.40\textwidth}
                \centering
                \includegraphics[width=1\textwidth]{./figures/property_proving/prop_execution_without_traceability.pdf}
                \caption{Property holds}
                \label{fig:prop_execution_without_traceability}
        \end{subfigure}%
        \\
        \begin{subfigure}[b]{0.40\textwidth}
                \centering
                \includegraphics[width=1\textwidth]{./figures/property_proving/prop_execution_with_traceability.pdf}
                \caption{Property does not hold}
                \label{fig:prop_execution_with_traceability}
        \end{subfigure}%
        \caption{Matching property to a transformation execution}
        \label{fig:prop_execution}
\end{figure}

% \cref{def:sat_prop_pc} is similar to that of property satisfaction for a transformation execution, but for path conditions. Again, the property is isomorphically matched onto the respective graphs in the path condition.
% 
% Properties are proved on each path condition created by the path condition generation algorithm. Each path condition is examined to determine whether the pre-condition graph of the property can be isomorphically matched onto the match graph of the path condition, as in \cref{def:sat_prop_pc}. If this match cannot be performed, then the pre-condition does not hold for this path condition and this path condition will not be checked for this property.
% 
% If the pre-condition can be isomorphically matched to the match graph of the path condition, then we check if the entire
% property can be isomorphically matched to the path condition. Again, this is described in \cref{def:sat_prop_pc}. If this matching cannot be performed, then the property does not hold. The path condition itself then serves as a
% counter-example for the transformation property. The path condition also contains information on which rules were executed in the counter-example to permit further examination.
% 
% This property-checking can be done relatively quickly with efficient graph-matching techniques. \cref{sec:implementation} will briefly describe implementation and optimization details, while \cref{sec:experiments} presents two property-proving experiments.

Due to the fact an infinite amount of transformation executions exists, proving the property directly on the set of transformation executions is not possible. We thus rely on the finite set of path conditions to prove properties about the set of all transformation execution. Let us then define what it means for a property to hold on, or be satisfied by, a path condition.

\begin{definition}{Satisfaction of a Property by a Path Condition\\}
\label{def:sat_prop_pc}
\CatchFileBetweenTags{\satproppc}{text/definitions}{satproppc}{\satproppc}
\end{definition}

The principle behind the satisfaction relation in \cref{def:sat_prop_pc} is the same as the one behind the satisfaction relation between a property and an execution of a transformation in \cref{def:sat_prop_ex}: whenever the property's pre-condition is found in the path condition then so is the complete property. Also, those two graphs found in the path condition share the property's pre-condition part. This last condition enforces that the pre- and post-conditions of the property are correctly linked by symbolic traceability links in the path condition.

Note that, despite their semantic similarity, the relations are expressed differently in \cref{def:sat_prop_ex} and \cref{def:sat_prop_pc}. In \cref{def:sat_prop_ex} -- \emph{satisfaction of a property by an execution of a transformation}, typed graph injective homomorphisms are defined from the property into the execution. However, in \cref{def:sat_prop_pc} the direction of the typed graph surjective homomorphisms is from the path condition into the property. This can be explained by the fact, mentioned previously in this text, that different rules in a path condition may have elements that match over the same concrete instances of a transformation's input model. As such, we need to consider the case where match elements of a path condition, originating from different rules, overlap.

\begin{figure}[htb]
 \centering
                \includegraphics[width=0.44\textwidth]{./figures/property_proving/prop_pc.pdf}
                \caption{Property satisfied by a path condition}
                \label{fig:prop_pc}
\end{figure}

This overlapping is modeled by the surjective typed graph homomorphisms of \cref{def:sat_prop_pc} having the property as co-domain. The surjections allow ``forgetting'' that two match elements of the path condition belong to different rules. Note however that these surjections are special, as two elements belonging to the same copy of a rule have to be mapped injectively onto the property. This situation is depicted in \cref{fig:prop_pc}. Note that element $B$ is successfully matched even though it appears in two different rules in the path condition.

\subsection{Expressiveness of the Property Language}
\label{subsec:expressiveness_prop}

As a result of taking rule combination (\cref{sec:building_pcs}) and overlap (\cref{sec:prop_satisfaction}) into consideration, our technique allows proving properties of transformation executions that are matched and built by multiple rules. This is the main goal of our work, as the properties we are interested in regard all possible interactions of rules in a DSLTrans transformation. However, an expressiveness limitation of the property language exists: we cannot prove properties having pre-conditions that can be found by executing the exact same rule more than once. This is natural, as by definition our abstraction only considers one exemplar of each rule per path condition having the exact same type for each match element.

In order to illustrate this limitation, consider a transformation having one single rule that matches an elements of type A and that produces an element of type X. Our technique will create a path condition as seen in \cref{fig:unprovable_pc}, abstracting over the number of times this rule has executed. According to the definition of satisfaction of a property by a path condition in \cref{def:sat_prop_pc}, the property in \cref{fig:unprovable_prop} does not hold on the path condition in \cref{fig:unprovable_pc}, although intuitively it should.

\begin{figure}[htb]
        \centering
        \begin{subfigure}[b]{0.144\textwidth}
                \centering
                \includegraphics[width=1\textwidth]{./figures/property_proving/unprovable_rule.pdf}
                \caption{Path condition created from a rule}
                \label{fig:unprovable_pc}
        \end{subfigure}%
        ~~
        \begin{subfigure}[b]{0.20\textwidth}
                \centering
                \includegraphics[width=1\textwidth]{./figures/property_proving/unprovable_prop.pdf}
                \caption{Property with multiple instances of rule elements}
                \label{fig:unprovable_prop}
        \end{subfigure}%
        \caption{Example of unprovable property}
        \label{fig:unprovable_property}
\end{figure}

Although this might be seen as a limitation of our technique, our case studies thus far indicate that very interesting properties exist that exclusively regard interactions between different rules. Nonetheless, our technique could be extended in order to consider more than one exemplar of the same rule per path condition. In fact, theoretically we already consider more than one exemplar of each rule when we treat polymorphism during path condition generation (see \cref{def:path_cond_gen}). However, in this case all expanded rules for a rule $rl$ are considered to be different as they differ by at least the type of one match element. Although this extension seems to fit relatively simply in our current theory, additional steps would need to be taken to understand which rules would be interesting to replicate to prove a particular property, and how many exemplars of each rule should be considered. In operational terms the number of replicas to consider of each rule is very important, given the exponential complexity of our path condition generation and property proof algorithms. Another possibility to tackle this issue would be to investigate and implement a more powerful satisfaction relation between path conditions and properties than the one we now present in \cref{def:sat_prop_pc}.
%Note however that the homomorphisms are injective when two elements belong to the same rule.\levi{explain this better}

% Given these definitions, we see that as all transformation executions are represented by path conditions, proving properties on the path conditions allows us to conclude property validity on all transformation executions.

\subsection{Validity and Completeness}
\label{subsec:prop_proving_valid_complete}

As for the path condition building algorithm, \emph{validity} and
\emph{completeness} need to be examined regarding our property verification
algorithm. In this context \emph{validity} means that if a property is satisfied by all path conditions generated for a transformation $tr$, then that property is satisfied by all executions of that transformation. On the other hand, if the property is not satisfied by at least one path condition, then it will not be satisfied by at least one transformation execution. In other words, we wish to show that no false positive or false negative proof results are induced by the abstraction relation.

% The proof of of these statements necessarily relies on the abstraction relation formally presented in\cref{def:instance_pc_ex}.

On the other hand, \emph{completeness} means that we are sure that all properties that can be expressed about a transformation can be shown to hold or not hold in all transformation executions.

As with the proofs for the validity and completeness of the abstraction relation, we present only proof sketches in this section in the interest of readability. Full proofs are shown in \cref{sec:val_complete_prop_verif} as \cref{prop:proof_validity_appendix} and \cref{prop:proof_completeness_appendix}.

% \subsubsection{Validity}
% 
% In this section we prove the validity of our property proof algorithm by considering what it
% formally means when a property is proved on a path condition. We argue that if
% the property holds or not on the path condition, then it will accordingly hold
% or not on for any transformation executions that the path condition represents.

% As well, we also prove that if a property holds or not on a specific transformation
% execution, it will hold or not on the path condition that represents that
% transformation execution.
\begin{proposition}{(Validity) The result of proving a property on a set of path conditions generated for a transformation or an all executions of that transformation is the same.}
\label{prop:proof_validity}

Let $tr\in \textsc{Transf}^{sr}_{tg}$ be a transformation and $p \in \textsc{Property}(tr)$ be a
property of $tr$. Given this, we have that transformation $tr$ satisfies property $p$ if and only if:
\begin{equation}
\label{eq:prop_proof_transf}
\bigwedge_{pc\in \textsc{Pathcond}(tr)} pc\vdash p \hspace{.3cm} \Longleftrightarrow \hspace{.3cm} \bigwedge_{ex\in \textsc{Exec}(tr)}ex\models p
\end{equation}
\end{proposition}
\begin{ps}
In order to prove the proposition in \cref{eq:prop_proof_transf} we will start by demonstrating that,
if property $p$ holds on a path condition $pc$ generated for $tr$,
then $p$ will necessarily hold on all execution $ex$ of $tr$ that is abstracted by $pc$. On the other hand, if $p$ does not hold on $pc$ then it will not hold for at least one execution $ex$ of $tr$ abstracted by $pc$. This lemma can be stated as follows:
\begin{equation}
\label{eq:prop_proof_pc}
pc\vdash p \;\Longleftrightarrow \;\forall ex\in\{ex\in\textsc{Exec}(tr)\;|\;ex\Vvdash pc\}\;.\; ex\models p
\end{equation}
We thus need to demonstrate both directions of the equivalence in \cref{eq:prop_proof_pc}, as follows:
\begin{itemize}
\item The proof of the left-to-right direction of the equivalence is split
into two cases as $pc\vdash p$ is true (as stated
in~\cref{def:sat_prop_pc}) when either: (1) the pre-condition of property $p$ cannot
be found in path condition $pc$; or (2) the pre-condition of $p$ is
found in $pc$ and the post-condition of $p$ also.
We need to demonstrate that both these cases entail that $ex\models p$ holds.
\item The proof of the right-to-left direction of the equivalence, which can be shown on its contrapositive:
\begin{multline}
\label{eq:prop_proof_right_left}
\neg(pc\vdash p) \;\Longrightarrow \;\\\exists ex\in\{ex\in\textsc{Exec}(tr)\;|\;ex\Vvdash pc\}\;.\; \neg(ex\models p)
\end{multline}
According to \cref{def:sat_prop_pc}, to demonstrate \cref{eq:prop_proof_right_left} we will need to show that when the pre-condition of property $p$ is found in $pc$ but the post-condition of $p$ is not, the property does not hold on at least one execution of $tr$ abstracted by $pc$.
\end{itemize}

Once \cref{eq:prop_proof_pc} is proved, we know that all path conditions on which a property holds represent executions on which the property also holds. Thus, if the property holds on all path conditions then it necessarily holds on all executions. On the other hand, if a property does not hold on one path condition, making it such that the conjunction on the left side of the equivalence in \cref{eq:prop_proof_transf} is false, then according to \cref{eq:prop_proof_pc} an execution for which it also does not hold exists. This makes it such that the conjunction on the right side of the equivalence in \cref{eq:prop_proof_pc} is also false.
\end{ps}

\begin{proposition}{(Completeness) Properties of a transformation can be shown to either hold for all transformation executions, or not hold for at least one transformation execution.}
\label{prop:proof_completeness}
\end{proposition}
\begin{pf}
This results follows from two previous results: \cref{prop:pc_completeness}, that tells us that every transformation execution is abstracted by one path condition; and \cref{prop:proof_validity} that shows us that every path condition is taken into consideration during property proof. Note that \cref{lem:uniqueness} guarantees consistency of our results, in the sense that the uniqueness of one path condition per transformation execution guarantees that a property cannot be proven to be both \emph{true} and \emph{false} for two path conditions representing the same transformation execution.
\end{pf}

% ~\newline\noindent{\bfseries Proposition \ref{prop:validity}.}
% \textit{ (Validity) The result of checking a property on a path condition and of checking the property on all the transformation executions that path
% condition abstracts will be the same.}\newline
% 
% We place the formal proof of this statement in appendix~\ref{sec:proofs}, proposition~\ref{prop:validity} as this result requires significant mathematical development.
% This result provides the guarantee that path conditions accurately represent their corresponding concrete transformation executions when performing property validation.
% 
% ~\newline\noindent{\bfseries Proposition \ref{prop:PP_completeness}.}
% \textit{ (Completeness) Proving that a property holds or does not
% hold by examining all path conditions of the transformation is equivalent to
% proving the property holds or does not hold for all transformation executions.}\newline
% 
% %\begin{proposition}{(Completeness) Proving that a property holds or does not
% %hold by examining all path conditions of the transformation is equivalent to
% %proving the property holds or does not hold for all transformation executions.}
% %\end{proposition}
% 
% The \emph{completeness} of our property verification algorithm is shown in appendix~\ref{sec:proofs}, proposition~\ref{prop:PP_completeness}.

### 2015-03-13 text/related_work.tex
previous version of the paper

\section{Related Work}
\label{sec:related_work}

In order to analyse the work in the literature that is close to our proposal, we will make use of the study on the formal verification of model transformations proposed in~\cite{DBLP:conf/icst/AmraniLSCDVTC12}. The study uses three dimensions to classify the analysis of model transformations. The dimensions are: 1) the \emph{kind of transformations} considered; 2) the \emph{properties} of transformations that can be verified; and 3) the \emph{verification technique} used.

\subsubsection*{Kind of Transformations Considered} DSLTrans is a graph based transformation language and as such shares its principles with languages such as AGG~\cite{Taentzer2000}, AToM$^3$~\cite{atom3:2002}, VIATRA2~\cite{DBLP:conf/uml/VarroP04}, ATL~\cite{Jouault200831} or VTMS~\cite{DBLP:journals/entcs/LevendovszkyLMC05}. As mentioned previously, DSLTrans' transformation are \emph{terminating} and \emph{confluent} by construction. This is achieved by expressiveness reduction which means that constructs which imply unbounded recursion or non-determinism are avoided. DSLTrans is, to the best of our knowledge, the only graph based transformation language where these properties are enforced by construction.

It is recognized in the literature that \emph{termination} and \emph{confluence} are important properties of model transformations, as these transformations have properties that are easier to understand and analyse. However, termination is undecidable for graph based transformation languages~\cite{DBLP:journals/fuin/Plump98}. This problem has led to a number of proposed termination criteria, as well as criteria analysis techniques, for transformations written in graph based transformation languages~\cite{deLara:2010:ATA:1815798.1815800,EhrigEhrigTaentzerdeLaraVarroVarro2005,Varro-Varro-etAl-2006,J:Bruggink2008,J:Kuster2006}. Confluence is also undecidable for graph based transformation languages~\cite{Plump2005}. As for termination, several confluence criteria and corresponding analysis techniques have been proposed in the literature~\cite{HeckelKT02,J:Kuster2006,J:Lambers-etAl-2006,J:Biermann2011}.

\subsubsection*{Verifiable Properties of Transformations} According to the classification in~\cite{DBLP:conf/icst/AmraniLSCDVTC12} the technique presented in this paper deals with properties that can be regarded as \emph{model syntax relations}. Such properties of a model transformation have to do with the fact that certain elements, or structures, of the input model are necessarily transformed into other elements, or structures, of the output model.

As early as 2002, Akehurst and Kent have introduced a set of structural relations between the metamodels of the abstract syntax, concrete syntax and semantics domain of a fragment of the UML~\cite{Akehurst:02}. Although they do not use such relations as properties of model transformations, their text introduces the notion of structural relations between a source and a target metamodel for a transformation. In 2007, Narayanan and Karsai propose verifying model transformations by structural correspondence~\cite{Narayanan:07}. In their approach, structural correspondences are defined as pre-condition/post-condition axioms. As the axioms provide an additional level of specification of the transformation, they are written independently from the transformation rules and are predicate logic formulas relying solely on a pair of the transformation's input and output model objects and attributes. The verification of whether such predicates hold is achieved by relying on so-called cross links (also named \emph{traceability links} in~\cite{DBLP:conf/icst/AmraniLSCDVTC12}) that are built between the elements of the input and output transformation model during the transformation's execution.

Although our proposal follows the same basic idea as the work of Narayanan and Karsai, there is one essential difference. Narayanan and Karsai's technique is focused on showing that pre-condition/post-condition axioms hold for one execution of a model transformation, involving one input and its corresponding output model. Thus, according to~\cite{DBLP:conf/icst/AmraniLSCDVTC12} the technique is \emph{transformation dependent} and \emph{input dependent}. In our proposal, we aim at proving structural correspondences for all executions of a model transformation, and base the construction of the properties (or pre-condition/post-condition axioms, using the vocabulary in~\cite{Narayanan:07}) on the source and target metamodel structures. Our approach is thus \emph{transformation dependent} but \emph{input independent} and aims at achieving the proof of the same kind of properties as Narayanan and Karsai propose, but one meta-level above.

In 2009~\cite{DBLP:journals/eceasst/CariouBBD09} Cariou \emph{et al.} study the use of OCL contracts in the verification of model transformations. The approach is also \emph{transformation dependent} and \emph{input dependent} in the sense that it requires an input model and an output model of the transformation. However, the authors provide a good account how to build OCL contracts for model transformations and show how to verify those contracts for endogenous transformations.

Aztalos, Lengyel and Levendovszky have published in 2010 their approach to the verification of model transformations~\cite{DBLP:conf/icst/AsztalosLL10}. They propose an assertion language that allows making structural statements about models at a given point of the execution of the transformation and also statements about the transformation steps themselves. The authors' technique applies to transformations written in the VTMS transformation language~\cite{DBLP:journals/entcs/LevendovszkyLMC05}. The technique consists of transforming VTMS transformation rules and verification assertions into Prolog predicates such that deduction rules encoding VTMS's and the assertion language's semantics can be used on automated Prolog proofs to check whether those assertions hold or not.

The approach resembles ours in the sense that the technique is also \emph{transformation dependent} but \emph{input independent} (the authors call their technique \emph{offline}). The artifacts used in the proofs are also generated from the transformation and the properties to be proved. While it is foreseeable that our \emph{model syntax relations} properties might be expressed by the assertion language proposed by Aztalos \emph{et al.}, the authors provide no account of the scalability of their approach. They mention however that since their approach is based on the generic SWI-Prolog inference engine, there could be a performance bottleneck or the possibility of non-terminating computations. They foresee that a specialised reasoning system might be necessary for their approach to scale.

More recently in 2012 and 2013, Guerra \emph{et al.}~\cite{guerra2013specification} have proposed techniques for the automated verification of model transformations based on visual contracts. Their work describes a rich and well-studied language for describing syntactic relations between input and output models. These pre- and post- condition graphs then are transformed into OCL expressions, which are fed into a constraint-solver to generate test input models for the transformation. Their framework algorithm can then test a transformation on a number of these input models, and verify them by the OCL expressions. The approach is \emph{transformation dependent} and \emph{input independent} and is independent of the transformation language used, which is a feature that we have not found elsewhere in the literature. However the verification technique used by Guerra \emph{et al.} differs fundamentally from ours. Our abstraction over the number of elements of the same type present in the model enables our approach to be exhaustive and allows for correctness proofs, while the approach by Guerra \emph{et al.} is aimed at increasing the level of confidence in a transformation through coverage of test cases. A similar white-box generation approach is also seen in recent work by Gonz{\'a}lez and Cabot~\cite{gonzalez2012atltest}.

Also in 2012 B\"uttner \emph{at al.} have published their work on the verification of ATL transformations~\cite{ButtnerECG12,ButtnerEC12}. In~\cite{ButtnerECG12} the authors translate ATL transformations and their semantics into transformation models in Alloy. They then use Alloy's model finder to search for the negation of a given property that should hold, where the property is expressed as an OCL constraint. As the authors mention, Alloy performs bounded verification and as such it does not guarantee that a counterexample is found if it exists. In~\cite{ButtnerEC12} B\"uttner \emph{at al.} aim at proving model syntax relation properties of ATL transformations expressed as pre-condition/post-condition OCL constraints. In order to do so, the authors provide and use an axiomatisation of ATL's semantics in first order logic. Verification of a given model transformation is achieved by using a HOT to transform the transformation under analysis into additional first order logic axioms. Off-the-shelf SMT solvers such as Z3 and Yices are then used to check whether the pre-condition/post-condition OCL constraints hold.

The approach in~\cite{ButtnerEC12} comes very close to ours as the authors aim at proving the same type of properties in a model independent fashion and can do so exhaustively by using mathematical proofs at an appropriate level of abstraction, which can be seen as symbolic. However, there are several differences with our approach. First, the authors' proofs may require human assistance, depending on the used SAT solver. Also, despite the fact that B\"uttner \emph{at al.} do treat constraints on object attributes, which we do not do, their results are presented for a small (6 rule) transformation and no scalability data, even preliminary, is presented. Finally, contrarily to DSLTrans, ATL does not have explicit formal semantics and because of that B\"uttner \emph{et al.}'s axiomatization of ATL's semantics is tentative. More generally, while the authors' approach requires an intermediate logic representation of the transformation under analysis, our symbolic approach deals directly with transformation rules. This feature can ease the interpretation of analysis of results such as counterexamples and could be in general less error-prone due to the absence of an indirection layer which maps transformation concepts to concepts in the chosen logic. It is interesting to notice that, similarly to our approach, B\"uttner \emph{et al.} have chosen \emph{expressiveness reduction} as a means to work with subset of ATL that is verifiable.

Assertional reasoning in graph transformations has been studied by Habel and colleagues, who have introduced nested conditions as properties of graphs in~\cite{Habel:2009}. The authors formally prove these nested conditions have the expressiveness of first-order graph formulas. Poskitt and Plump later propose in~\cite{PoskittP12} a Hoare-style verification calculus which is anchored on their experimental graph programming language GP. Using this calculus they then go on to prove nested condition properties of a graph-colouring GP program. Our approach shares some resemblances with assertional reasoning in that we also propose a pre-condition/post-condition language and a calculus for proving such properties in DSLTrans. We remark however that the theoretical work in assertional reasoning described above is larger in scope than what we present here and that assertional reasoning results require lifting to more usable graph transformation languages than GP before they can be used in practice.

\subsubsection*{Verification Technique Used} A different possibility for our work would have been to utilise the GROOVE tool, which can specify, play, and analyse graph transformations~\cite{rensink:2012}. In particular, GROOVE assumes that the states of the systems to be analysed are expressed as graphs and that the system's behaviour is simulated by graph transformation rules that manipulate those graphs.

In~\cite{DBLP:conf/gg/RensinkSV04} Rensink, Schmidt and Varr\'o test whether safety and reachability properties that are expressed as constraints over graphs can be efficiently checked by building the state space for a transformation. The answer is positive, but the authors found state space explosion problems as we did. In order to tackle those issues the tool relies on exploiting the symmetric nature of a problem by investigating isomorphic situations only once. This is very similar to optimisations we have made in our implementation of our approach by maintaining caches throughout path condition construction and property proof. Those caches allow us to avoid rerunning the expensive subgraph isomorphism algorithm as much as possible. It is foreseeable that our approach could make use of the advanced state space construction and recent CTL property checking capabilities of GROOVE. This could be achieved by using GROOVE as the transformation framework for our approach instead of T-CORE. However, at the time of the construction of our tool, fine-grained control of GROOVE transformations via an API as we do with T-CORE did not exist. It was thus infeasible to implement our approach by relying solely on GROOVE's graphical interface.

Still in the context of GROOVE, several studies~\cite{RensinkD06,BauerBKR08,RensinkZ12} have been performed on abstractions that allow coping with the state space explosion when performing model checking of state-based systems modeled as graph transformations. The authors present various abstractions on state graphs that allow reducing their size during model checking while allowing equivalent (or approximate versions of) proofs of temporal logic properties using the abstracted state graphs. Although our technique is also based on abstraction, our main purpose is not to execute concrete graphs in order to examine the state they represent. We rather symbolically represent all transformation executions (resulting from the application of all rules in a DSLTrans transformation to any input model) which are in an abstraction relation with the path conditions, such that we are able to symbolically examine the relations between all of the transformation's inputs and outputs.

Also from the \emph{verification technique} viewpoint, Becker \emph{et al.} propose a technique for checking a dynamic system where state is encoded as a graph~\cite{Becker:2006:SIV:1134285.1134297}. They also use model transformations to simulate the system's progression and aim at verifying that no unsafe states are reached as part of the system's behavior. In this sense Becker \emph{et al.}'s approach is \emph{transformation dependent} and \emph{input independent}, as an infinite amount of initial graphs needs to be considered. However, instead of generating the exhaustive state space as is done with GROOVE, the authors follow a different strategy by checking that no unsafe states of the system can be reached. They do so by searching for unsafe states as counterexamples of invariants encoded in the transformation rules. The analysis is performed symbolically on the application transformation rules and as such resembles our symbolic execution technique.
However, rather than being generically applicable to model transformations, possibly exogenous, the approach is geared towards the mechatronic domain and graph transformations are used as a means to encode the dynamic structural adaptation of such systems.
The applicability or efficiency of Becker \emph{et al.}'s technique when applied to the verification of model syntax relations in model transformations remains to be studied.

### 2015-03-13 figures/convert.py

@@ -1,0 +1,39 @@
+#!/bin/python
+
+import os
+import subprocess
+
+def convert(fname):
+    if not fname.endswith(".png") and not fname.endswith(".svg"):
+        return
+    
+    fname_as_pdf = fname.replace(".svg", ".pdf")
+    
+    try:
+        pdf_mod_time = os.path.getmtime(fname_as_pdf)
+    except Exception:
+        pdf_mod_time = 0
+        
+    
+    if os.path.getmtime(fname) <= pdf_mod_time:
+        return
+     
+    print("Converting: " + fname)
+        
+    if fname.endswith(".png"):
+        subprocess.call("convert -density 1024x1024 " + fname + " " + fname_as_pdf, shell=True)
+    elif fname.endswith(".svg"):
+        #subprocess.call("inkscape --verb=FitCanvasToDrawing --verb=FileSave --verb=FileClose " + fname, shell=True)
+        subprocess.call("inkscape " + fname + " --export-pdf=" + fname_as_pdf, shell=True)
+
+def convert_dir(d):
+    print(d)
+    for fname in os.listdir(d):
+        path = d + "/" + fname
+        print(path)
+        if os.path.isdir(fname):
+            convert_dir(path)
+        else:
+            convert(path)
+
+convert_dir(".")

### 2015-03-13 figures/results/convert.py

@@ -1,0 +1,34 @@
+#!/bin/python
+
+import os
+import subprocess
+
+def convert(fname):
+    if not fname.endswith(".png") and not fname.endswith(".svg"):
+        return
+    
+    fname_as_pdf = fname.replace(".svg", ".pdf")
+    
+    try:
+        pdf_mod_time = os.path.getmtime(fname_as_pdf)
+    except Exception:
+        pdf_mod_time = 0
+        
+    
+    if os.path.getmtime(fname) <= pdf_mod_time:
+        return
+     
+    print("Converting: " + fname)
+        
+    if fname.endswith(".png"):
+        subprocess.call("convert -density 1024x1024 " + fname + " " + fname_as_pdf, shell=True)
+    elif fname.endswith(".svg"):
+        #subprocess.call("inkscape --verb=FitCanvasToDrawing --verb=FileSave --verb=FileClose " + fname, shell=True)
+        subprocess.call("inkscape " + fname + " --export-pdf=" + fname_as_pdf, shell=True)
+
+if __name__ == "__main__":
+    dirs = ["./results"]
+
+    for d in dirs:
+        for fname in os.listdir(d):
+            convert(d + "/" + fname)

### 2015-03-13 figures/results/results.py

@@ -1,0 +1,90 @@
+#!/usr/bin/env python
+
+import pickle
+import numpy
+import pylab
+import sys
+
+from convert import *
+
+number_of_rules = [3, 5, 7, 10, 12, 14, 17, 19, 21]
+number_of_pcs = [8,	16,	34,	272,	442,	1156,	9248,	15028,	39304]
+time_to_build = [0.003,	0.13,	0.39,	1.87,	2.68,	9.00,	59.08,	97.52,	369.19]
+mem_size = [0.08,	0.09,	0.17,	1.24,	1.83,	4.98,	38.01,	60.10,	156.79]
+time_prop_1= [	0, 0.19,	1.26,	2.4,	3.4,	8.38,	73.51,	140.77,	412.02]
+
+time_prop_2=[	0.003, 0.003,	0.003,	0.003,	0.003,	0.003,	0.003,	0.003,	0.003]
+
+
+
+from matplotlib.ticker import ScalarFormatter 
+majorFormatter = ScalarFormatter()
+
+
+font = {'family' : 'normal',
+        'weight' : 'normal',
+        'size'   : 22}
+
+pylab.rc('font', **font)
+
+#num rules vs num PCs
+#fig, ax = pylab.subplots()
+
+pylab.hold('off')
+#fig=pylab.semilogy(number_of_rules, number_of_pcs)
+##pylab.hold('on')
+
+
+
+
+pylab.clf()
+
+# NUM RULES VS MEMORY
+
+
+
+fig=pylab.loglog(number_of_pcs, time_prop_1)
+#pylab.hold('on')
+
+pylab.xlim([8,39304])
+
+#pylab.hold('on')
+#title = "Num rules vs PCs"
+filename = "pcs_vs_prop1"+ ".svg"
+
+pylab.xlabel("Number of path conditions")
+pylab.ylabel("Time (s)")
+#pylab.legend(legend, 'upper left')
+#pylab.title(title)
+
+
+pylab.savefig(filename, format = "svg", transparent=True, bbox_inches='tight', pad_inches=0.05)
+
+convert("./" + filename)
+
+
+
+pylab.clf()
+
+# NUM RULES VS MEMORY
+
+
+
+#fig=pylab.semilogx(number_of_pcs, time_prop_2)
+##pylab.hold('on')
+
+#pylab.xlim([8,39304])
+
+##pylab.hold('on')
+##title = "Num rules vs PCs"
+#filename = "pcs_vs_prop2"+ ".svg"
+
[…]

### 2015-03-13 text/appendix_denotational.tex
Some basic definition for the denotaional semantics.

\section{Definitions and Proofs of the Denotational Semantics}
\label{sec:proofs_denotation}

\subsection{Basic Definitions}

\begin{definition}{Pattern}
\label{def:rule_pattern}
A \emph{pattern} over a given set of vertex types $\mathcal{VT} = \{ \mathcal{V}_1,\dots,\mathcal{V}_k\}$ and a given set of edge types $\mathcal{ET} = \{ \mathcal{E}_1, \dots,\mathcal{E}_l \} \subseteq \{ (\mathcal{V}_i \times \mathcal{V}_j \mid 1 \leq i,j \leq k \}$ is defined as the structure $(V_1,\dots,V_m,E_1,\dots E_n)$ with $V_i \in \mathcal{VT}$ for $i \leq i \leq m$ and $E_i \in \mathcal{ET}$ for $1 \leq i \leq n$. It represents an instantiable graph pattern by describing the vertices and edges used in that rule and the set of their possible instances.
\end{definition}

\begin{example}{Pattern}
For the Stations Rule in Figure \ref{fig:dsltransformation}, the corresponding pattern is
$(S_O,S_G,T)$ where $S_O$ represents the set of all station instances from the orgnization language, $S_G$, the corresponding set of the gender language, and $T$ the set of a trace links between a pair of instances.
\end{example}

\begin{definition}{Instance Graph}
\label{def:instance_graph}
An \emph{instance graph} $(V,E)$ with $V = \{ v_1,\dots,v_m \}$ where $v_i \in V_i$ for $1 \leq i \leq m$ and $v_i \neq v_j$ for $i \neq j$, and $E = \{ e_1,\dots, e_n \}$ where $e_i \in E_i$ for $1 \leq i \leq n$ and $e_i \neq e_j$ for $i \neq j$ of a \emph{pattern} represents a graph instantiating a pattern, respecting the type constraints and multiplicities defined by that pattern.
\end{definition}

\begin{example}{Instance Graph}
Again, using the Stations Rule in Figure \ref{fig:dsltransformation}, the set of corresponding instance graphs is $\{ ( \{ s_o, s_t \} , \{  \} ) \mid \}$
$(S_O,S_G,T)$ where $S_O$ represents the set of all station instances from the orgnization language, $S_G$, the corresponding set of the gender language, and $T$ the set of a trace links between a pair of instances.
\end{example}

### 2015-03-18 text/appendix_denotational.tex
added the definition we got at the board of rules execution, as well as support for comments (use \levi, \bentley and \bernhard for your comments)

@@
 \end{example}
+
+\subsection{Semantics of DSLTrans}
+
+\begin{definition}{Set of Rule Executions.}
+\label{def:rule_exec_set}
+Let $rl\in \textsc{Rule}^{sr}_{tg}$ be a DSLTrans rule. The set of all rule
+executions is a set of graph relations, built inductively as follows:
+\begin{align*}
+\llbracket rl\rrbracket =\big\{&(i',o')\in \textsc{Gr}\;\lvert\;(i',o') =
+(rl_{\emptyset},rl_{\emptyset})\;\lor \\&(i',o') = (o,o\sqcup rl_{pat})\land
+(i,o)\in \llbracket rl\rrbracket \big\}\\\\
+&\text{where } rl_{pat} \text{ is a pattern over } rl
+\end{align*} 
+\end{definition}
+
+A set of rule executions is the set of all input/output pairs resulting from
+executing one rule over any input model. It is built inductively by starting
+from $rl_{\emptyset}$ (any graph where the rule does not match) and adding to it
+any patterns of $rl$, meaning $rl$ has executed any number of times.
+
+\levi{the definition of sets of rule executions is based on the notion of
+patterns over rules from Bernhard's text. I think I got the intent of
+\ref{def:rule_pattern}, but more discussion is needed such that I'm sure of my
+definition.}

### 2015-03-19 text/appendix_denotational.tex
added definitions of rule and layer executions

--- 2015-03-13 text/appendix_denotational.tex
@@
 \end{example}
+
+\subsection{Semantics of DSLTrans}
+
+\begin{definition}{Set of Rule Executions.}
+\label{def:rule_exec_set}
+Let $rl\in \textsc{Rule}^{sr}_{tg}$ be a DSLTrans rule. The set of all rule
+executions is a set of graph relations, built inductively as follows:
+\begin{align*}
+\llbracket rl\rrbracket =\big\{&(i',o')\in \textsc{Gr}\;\lvert\;(i',o') =
+(rl_{\emptyset},rl_{\emptyset})\;\lor \\&(i',o') = (o,o\sqcup rl_{pat})\land
+(i,o)\in \llbracket rl\rrbracket \big\}
+\end{align*} 
+where  $rl_{pat}$ is a pattern over $rl$ and $\sqcup$
+is the tracing graph union where at least one element of $\Vert rl_{pat}\Vert$
+is disjoint with $o$.
+\end{definition}
+
+A set of rule executions is the set of all input/output pairs resulting from
+executing one rule over any input model. It is built inductively by starting
+from $rl_{\emptyset}$ (any input graph where the rule does not match) and adding
+to it any traced patterns of $rl$, meaning $rl$ has executed any number of
+times. Note that the match part of $rl_{pat}$ (denoted $\Vert rl_{pat}\Vert$)
+can overlap with the graph in $o$ by at most $n-1$ elements, being that $n$ is the amount
+of match elements in $rl_{pat}$. Traceability adds trace links between any
+elements belonging all input elements and any output elements of $rl_{pat}$ that
+is not connected to a backward link.
+
+\levi{the definition of sets of rule executions is based on the notion of
+patterns over rules from Bernhard's text. I think I got the intent of
+\ref{def:rule_pattern}, but more discussion is needed such that I'm sure of my
+definition. In particular the notion of overlap between graphs is
+underspecified regarding types (is this solved in the basic definitions?)}
+
+\begin{definition}{Set of Rule Executions for a Layer.}
+\label{def:rule_exec_set}
+Let $l\in \textsc{Layer}^{sr}_{tg}$ be a DSLTrans layer. The set of all layer
+executions is a set of graph relations, built inductively as follows:
+\begin{gather*}
+  \llbracket l\rrbracket = 
+  \begin{cases}
+    \emptyset &\text{if}\; l=\emptyset\\
+    \llbracket l_1\rrbracket \otimes \llbracket rl\rrbracket &\text{if}\;
+    l = l_1\cup rl
+  \end{cases}
+\end{gather*}
+
+where $\textsc{Ex}_1 \otimes \textsc{Ex}_2$ is set of the unions of pairs of
+graphs $(ex_1,ex_2)\in \textsc{Ex}_1 \times \textsc{Ex}_2$ ($\textsc{Ex}_1$ and
+$\textsc{Ex}_2$ being sets of executions) where the match elements of $ex_1$ and
+the match elements of $ex_2$ may be joint.
+\end{definition}
+
+\levi{this is overapproximating the set of layer executions because of the
+corner case where the match of one rule contains the match of the other rule in
+the layer, in which case the smaller rule will necessarily execute where the
+larger one does (but not vice-versa). I think there is an elegant way to
+describe this\ldots or can we just use the overapproximation as is, given it
+does contain the set of all executions of the layer\ldots}

### 2015-03-26 text/appendix_denotational.tex
Definition of a transformation rule

--- 2015-03-13 text/appendix_denotational.tex
@@
+% !TEX root = ../main.tex
+
 \section{Definitions and Proofs of the Denotational Semantics}
@@
 
-\begin{example}{Pattern}
+\begin{example}
 For the Stations Rule in Figure \ref{fig:dsltransformation}, the corresponding pattern is
-$(S_O,S_G,T)$ where $S_O$ represents the set of all station instances from the orgnization language, $S_G$, the corresponding set of the gender language, and $T$ the set of a trace links between a pair of instances.
+$(S_O,S_G,T)$ where $S_O = \{ \mbox{station}_{o,1}, \mbox{station}_{o,2},\dots \}$ represents the set of all station instances from the organization language, $S_G = \{ \mbox{station}_{g,1}, \mbox{station}_{g,2},\dots \}$, the corresponding set of the gender language, and $T = S_O \times S_G$ the set of a trace links between a pair of instances.
 \end{example}
@@
 
-\begin{example}{Instance Graph}
-Again, using the Stations Rule in Figure \ref{fig:dsltransformation}, the set of corresponding instance graphs is $\{ ( \{ s_o, s_t \} , \{  \} ) \mid \}$
-$(S_O,S_G,T)$ where $S_O$ represents the set of all station instances from the orgnization language, $S_G$, the corresponding set of the gender language, and $T$ the set of a trace links between a pair of instances.
+\begin{example}
+Again, using the Stations Rule in Figure \ref{fig:dsltransformation}, the set of corresponding instance graphs is $\{ ( \{ s_o, s_g \} , \{ (s_o, s_t \} ) \mid s_o \in S_O \mbox{and} s_g \in S_G \}$.
+Thus, each instance graph consists of a pair of an organization station and a gender station, together with a single edge representing the trace link.
 \end{example}
+
+\begin{definition}{Transformation Rule}
+A \emph{transformation rule} over a given set of vertex types $\mathcal{VT} = \mathcal{VT}^i \cup \mathcal{VT}^o$ with $ \mathcal{VT}^i \cap \mathcal{VT}^o = emptyset$ and a given set of edge types $\mathcal{ET} =  \mathcal{ET}^i \cup \mathcal{ET}^o$ with $\mathcal{ET}^i \cap \mathcal{ET}^o = \emptyset$ is a partitioned pattern, describing a transformation by distinguishing between input elements $ \mathcal{VT}^i$ and  $\mathcal{ET}^i$ and output elements $ \mathcal{VT}^o$ and  $\mathcal{ET}^o$.
+\end{definition}

### 2015-03-26 text/appendix_denotational.tex

--- 2015-03-19 text/appendix_denotational.tex
@@
 
-\begin{example}{Instance Graph}
-Again, using the Stations Rule in Figure \ref{fig:dsltransformation}, the set of corresponding instance graphs is $\{ ( \{ s_o, s_t \} , \{  \} ) \mid \}$
-$(S_O,S_G,T)$ where $S_O$ represents the set of all station instances from the orgnization language, $S_G$, the corresponding set of the gender language, and $T$ the set of a trace links between a pair of instances.
+\begin{example}
+Again, using the Stations Rule in Figure \ref{fig:dsltransformation}, the set of corresponding instance graphs is $\{ ( \{ s_o, s_g \} , \{ (s_o, s_t \} ) \mid s_o \in S_O \mbox{and} s_g \in S_G \}$.
+Thus, each instance graph consists of a pair of an organization station and a gender station, together with a single edge representing the trace link.
 \end{example}
+
+\begin{definition}{Transformation Rule}
+A \emph{transformation rule} over a given set of vertex types $\mathcal{VT} = \mathcal{VT}^i \cup \mathcal{VT}^o$ with $ \mathcal{VT}^i \cap \mathcal{VT}^o = emptyset$ and a given set of edge types $\mathcal{ET} =  \mathcal{ET}^i \cup \mathcal{ET}^o$ with $\mathcal{ET}^i \cap \mathcal{ET}^o = \emptyset$ is a partitioned pattern, describing a transformation by distinguishing between input elements $ \mathcal{VT}^i$ and  $\mathcal{ET}^i$ and output elements $ \mathcal{VT}^o$ and  $\mathcal{ET}^o$.
+\end{definition}
 

### 2016-05-02 text/abstraction_relation.tex
Add reviewer's comments into the paper.

@@
 will abstract. \cref{fig:empty_pc} demonstrates two cases. In each, the path
-condition is on the left-hand side, and a transformation execution is on the
+condition \textit{pc} is on the left-hand side, and a transformation execution \textit{ex} is on the
 right-hand side. Note that in \cref{fig:empty_pc_success}, the path condition
@@
 
-The first condition for the abstraction relation is to determine whether a typed graph injective homomorphism can be found between the match graph of the path condition, and an transformation executions.
+The first condition for the abstraction relation is to determine whether a typed graph injective homomorphism can be found between the match graph of the path condition, and a transformation execution.
 Note that in both \cref{fig:empty_pc_success} and \cref{fig:empty_pc_fail}, an empty typed graph homomorphism satisfies this condition, highlighted by blue arrows.
@@
   output model).
-For any of these transformation executions, Proposition~\ref{eq:abstr_input_output} of the abstraction relation definition is satisfied, as no rule copy exists in the path condition and the output of the transformation execution is empty -- empty typed graph homomorphisms thus satisfy the all the conditions of the proposition.
+For any of these transformation executions, Proposition~\ref{eq:abstr_input_output} of the abstraction relation definition is satisfied, as no rule copy exists in the path condition and the output of the transformation execution is empty -- empty typed graph homomorphisms thus satisfy all the conditions of the proposition.
 Proposition~\ref{eq:abstr_trace} of the abstraction relation definition also trivially holds because no traceability links exist either in the path condition or in any of the considered executions.
@@
   \item \emph{Inductive case:} for the inductive case (transitive case of \cref{def:path_cond_layer_comb}) we need to show that, assuming the property holds for all path conditions generated for a transformation $tr$, then the property will also hold for a transformation $tr'$ -- where $tr'$ results from adding a new rule $rl$ to the last layer of $tr$.
-We will thus need to consider the four cases of rule combination:\vspace{.2cm}
+We will thus need to consider the four \reviewer{three?} cases of rule combination:\vspace{.2cm}
 
@@
 
-We will do so by attempting to to build an $ex\in \textsc{Exec}(tr)$ such that $ex\Vvdash pc_1$ and $ex\Vvdash pc_2$ and demonstrating that it is always the case that such is not possible.
+We will do so by attempting to build an $ex\in \textsc{Exec}(tr)$ such that $ex\Vvdash pc_1$ and $ex\Vvdash pc_2$ and demonstrating that it is always the case that such is not possible.
 In order to structure our argumentation, we will consider two cases:

### 2016-05-02 text/appendix_property_proof.tex
Add reviewer's comments into the paper.

@@
 \begin{proof}
-This results follows from two previous results: \cref{prop:pc_completeness_appendix}, that tells us that every transformation execution is abstracted by one path condition; and \cref{prop:proof_validity_appendix} that shows us that every path condition is taken into consideration during property proof.
+This result follows from two previous results: \cref{prop:pc_completeness_appendix}, that tells us that every transformation execution is abstracted by one path condition; and \cref{prop:proof_validity_appendix} that shows us that every path condition is taken into consideration during property proof.
 Note that \cref{lem:uniqueness_appendix} guarantees consistency of our results, in the sense that the uniqueness of one path condition per transformation execution guarantees that a property cannot be proven to be both \emph{true} and \emph{false} for two path conditions representing the same transformation execution.

### 2016-05-02 text/building_path_conditions.tex
Add reviewer's comments into the paper.

@@
 According to the pre-conditions of the equation presented in \cref{def:rule_comb_unsatisfied}, a path condition does not satisfy the dependencies present in a rule if there is no surjective typed graph homomorphism between the backward links of the rule and the symbolic traceability links of the path condition.
-Besides expressing the fact that all backward links must exist as symbolic traceability links the path condition, the surjective homomorphism allows modeling the case where dependencies expressed by two (or more) backward links between similarly typed elements can be satisfied one single symbolic traceability link in the path condition.
+Besides expressing the fact that all backward links must exist as symbolic traceability links the path condition, the surjective homomorphism allows modeling the case where dependencies expressed by two (or more) backward links between similarly typed elements can be satisfied by one single symbolic traceability link in the path condition \reviewer{Could be rephrased}.
 This is the case, for example, of rule \emph{FemaleToFemale} in the \emph{Police Station} in \cref{fig:dsltransformation}.
@@
                 \includegraphics[width=1\textwidth]{./figures/building_path_conditions/multiple_total_satisfied_dependencies.pdf}
-                \caption{Totally satisfied at multiple locations}
+                \caption{Totally satisfied at multiple locations - \reviewer{Fix boxes on bottom}}
                 \label{fig:multiple_total_satisfied_dependencies}

### 2016-05-02 text/definitions.tex
Add reviewer's comments into the paper.

@@
 %<*typedgraphtext>
-
 We will start by introducing the notion of typed graph.
@@
 %<*typedgraphhomomorphismtext>
-
+\reviewer{definition 3:''st'' was before ''(s, t)'' in the previous definitions (idem with ''stâ''), in the same way in definition 13, both notations are used.}
 For the formal development of our technique, we are interested in relations between typed graphs that are structure-preserving, i.e.
@@
 We define $rl$'s matcher, noted $\Vert rl\Vert$, as the transformation rule $\big\langle V',E',st',\tau',Match,\\Apply'\big\rangle\sqsubseteq rl$ where $v_1\xrightarrow{e} v_2\in E'$ if and only if $v_1,v_2\in Match$ or $\tau(e)=trace$ and $V'=V_m\cup \big\{v_1\,|\,v_1\xrightarrow{e} v_2\in E \land \tau(e)=trace\big\}$.
-
+\reviewer{Definition 14: where does Apply' come from? Should v1 be 
+  an element of Apply?}
 %</backmatchtransformationrule>
@@
 %'transformation rule cref in text
+
 introduces the notion of matcher for a transformation rule which consists solely of the match pattern of a rule and its backward links, if any.
@@
 %<*layertransformationtext>
-
+\reviewer{I do not fully understand why this
+  could be problematic. Do you have an example?}
 \cref{def:layer_transformation} formalises the abstract syntax of a model transformation, introduced in \cref{sec:dsltrans}.
@@
 %<*layertransformationtextappendix>
-
+\reviewer{I do not fully understand why this
+  could be problematic. Do you have an example?}
 \cref{def:layer_transformation_appendix} formalises the abstract syntax of a model transformation, introduced at the beginning of this section.
@@
 %<*pccfrc>
+\reviewer{A $\sqcup$ operator is defined in Def. 19, but there you seem to
+  rely on the fact that some elements of the path conditions and
+  some elements of the rule are identical, otherwise the union
+  would not work. Since a rule can be applied in several places,
+  this would require a renaming of the rule elements, which - as 
+  far as I could see - is not specified.}
 
@@
 Additionally, $Match = Match'\sqcup Match''$ and $Apply = Apply'\sqcup Apply''$.
-Finally, we have that: $Rulecop = Rulecop'\cup rl$.
+Finally, we have that: $Rulecop = Rulecop'\cup$ {rl}.
 
@@
 Through our abstraction relation defined in \cref{sec:abstraction_relation}, the final set of created path conditions will represent every feasible transformation execution.
-\cref{sec:verif_dsltrans_props} will discuss how our algorithm proves properties on these path conditions, and thus on all executions of the
-transformation.
+\cref{sec:verif_dsltrans_props} will discuss how our algorithm proves properties on these path conditions, and thus on all executions of the transformation.
 %</combpcsetlayertext>
@@
 Let $tr\in \textsc{Transf}^{sr}_{tg}$ be a DSLTrans transformation. Let also $pc = \langle V,E,st,\tau,Match,Apply,Rules\rangle \in
-\textsc{Pathcond}(tr)$ of be a path condition of $tr$ and $ex = \langle V',E',st',\tau',Input,Output\rangle \in Exec(tr)$ be an execution of $tr$.
+\textsc{Pathcond}(tr)$ of a path condition of $tr$ and $ex = \langle V',E',st',\tau',Input,Output\rangle \in Exec(tr)$ be an execution of $tr$.
 We have that $ex$ is abstracted by $pc$, noted $ex\Vvdash pc$, if and only if the set of transformation rules of $tr$ combined in $pc$ and the set of transformation rules of $tr$ used to built $ex$ is the same, and:
@@
 
+\reviewer{Definition 28: When you define the abstraction of an execution
+  by a path condition, you require two conditions (separated by an
+  and). I have the feeling that the morphisms used in both
+  conditions should somehow be compatible to each other. However,
+  your triangle notation indicates only the existence of a morphisms.
+  How do you express that these morphisms somehow "agree"?}
 %</abstractionrel>
@@
 
+\reviewer{You can not handle properties with multiple instances of rule 
+    element (example in Fig. 25). On p.14 in the answer section,
+    you claim that such conditions are excluded in Def. 29. However,
+    the last lines of this definition are not easy to understand.
+    Please explain why there are there (by giving the explanation
+    from the answers section.)
+
+    Without these explanations it is very hard to see why 
+    Prop. 3 should hold.}
+
 Let $tr\in \textsc{Transf}^{sr}_{tg}$ be a DSLTrans transformation.
@@
 
+\reviewer{In Def. 31 I found it confusing that - different from previous
+    definitions - the two morphisms now go in the same direction.
+    There is an explanation after Definition 31, but I found it 
+    very hard to understand.}
 Let $tr\in \textsc{Transf}^{sr}_{tg}$ be a transformation.

### 2016-05-02 text/property_verification.tex
Add reviewer's comments into the paper.

@@
 
-In order to illustrate this limitation, consider a transformation having one single rule that matches an elements of type A and that produces an element of type X.
+In order to illustrate this limitation, consider a transformation having one single rule that matches an element of type A and that produces an element of type X.
 Our technique will create a path condition as seen in \cref{fig:unprovable_pc}, abstracting over the number of times this rule has executed.
@@
 % transformation execution.
-\begin{proposition}{(Validity) The result of proving a property on a set of path conditions generated for a transformation or an all executions of that transformation is the same.}
+\begin{proposition}{(Validity) The result of proving a property on a set of path conditions generated for a transformation or all executions of that transformation is the same.}
 \label{prop:proof_validity}
@@
 \begin{pf}
-This results follows from two previous results: \cref{prop:pc_completeness}, that tells us that every transformation execution is abstracted by one path condition; and \cref{prop:proof_validity} that shows us that every path condition is taken into consideration during property proof.
+This result follows from two previous results: \cref{prop:pc_completeness}, that tells us that every transformation execution is abstracted by one path condition; and \cref{prop:proof_validity} that shows us that every path condition is taken into consideration during property proof.
 Note that \cref{lem:uniqueness} guarantees consistency of our results, in the sense that the uniqueness of one path condition per transformation execution guarantees that a property cannot be proven to be both \emph{true} and \emph{false} for two path conditions representing the same transformation execution.

### 2016-05-03 text/building_path_conditions.tex
Add all comments from third reviewer.

@@
 As well, we present our approach to building a set of path conditions to represent all executions of a DSLTrans transformation.
+
+\reviewer{In Section 4, rather than starting first with the construction of path conditions
+(Section 4,4), it would be more natural to discuss their meaning in terms of
+transformation executions (Section 5), since that should be the justification for
+your constructions.}
 
@@
 
-DSLTrans rules allow for dependencies to be specified on which elements of the
+DSLTrans rules specify which elements of the
 output model were created from specific elements of the input model. To resolve
@@
 An example of the symbolic traceability link creation process is shown in
-\cref{fig:trace_links}. Note that symbolic traceability links are a solid line between
+\cref{fig:trace_links}. Note that symbolic traceability links are solid lines between
 match and apply elements in our visual notation.
@@
 
+\reviewer{Def. 18 suddenly introduces the notion of “rule copies” (Rulecop) which are re-
+ally not explained properly. My guess is that they are isomorphic representatives
+of rules. On page 13 under ”notation” this component is called Rule and referred
+to as a relation, and in Def. 28 it is denoted as Rules and required to be “the same
+as the set of transformation rules used to build [the transformation execution]”
+(and hence not isomorphic copies?). Needless to say, these inconsistencies do
+not help understandability.}
+
 \CatchFileBetweenTags{\pc}{text/definitions}{pc}{\pc}
@@
 Recall that a path condition represents a set of rules that have symbolically executed, thereby abstracting a set of transformation executions through our abstraction relation.
-Combining a path condition with a rule will produce one or more path conditions depending on how the rule combines with the rules already represented by the path condition.
+\reviewer{we cannot recall this as it has not yet been presented;
+the abstraction relation is defined only in Section 5}Combining a path condition with a rule will produce one or more path conditions depending on how the rule combines with the rules already represented by the path condition.
 The pre- and post- conditions defined by the path condition will be modified according to the elements found in that rule.
@@
 
+\reviewer{Def. 19 can only be understood once it is clear that your plan is to make iso-
+morphic rule copies in which the node and edge identities already coincide with
+those in your path condition. (Hence my “isomorphic representative” guess in
+the previous remark.) This is very much nonstandard in the works of graph trans-
+formation, and poorly motivated. What are the advantages of this approach over
+relying on partial morphisms from a fixed rule to the path condition? As it is,
+you have to explain that you only consider isomorphic copies that differ in rele-
+vant parts, where “relevant” is determined by the question whether a node/edge
+identity coincides with one on the path condition. (You actually do not explain
+that at all: strictly following your definitions one would have to consider an in-
+finite number of isomorphic rule copies, assuming that the number of available
+identities is infinite (as it must be).)}
+
+\reviewer{“joint” $\rightarrow$ “overlapping”? Why are the type graphs actually not
+identical, as pc and rl are defined over the same source and target metamodels?
+Last line: “rl” $\rightarrow$ “$\{rl\}$”}
+
 \CatchFileBetweenTags{\pccfrctext}{text/definitions}{pccfrctext}{\pccfrctext}
@@
 
-The below figures will demonstrate the four cases above.
+The figures below will demonstrate the four cases above.
 As a reminder of visual notation, the backward links are dashed lines between the match and apply graphs of the rule and path condition, while symbolic traceability links are solid lines between the two graphs.
@@
 \end{definition}
+
+\reviewer{The right hand side of the conclusion of Def. 20 has $AC ∪ pc ′ ∈AC pc ′ $, of which
+I cannot make sense. AC is a set, but the pc' are not, so what operation is
+$\cup_{pc'\in AC} pc'$?}
+
+\reviewer{Do you mean $\sqcup_{pc'\in AC}$instead? But then the result is not a set, so how can you take its union with AC? The same problem reoccurs in other definitions.}
 
@@
 
+\reviewer{Def. 24: it seems to me you are making a very serious mistake here. $\blacktriangleright$ tests
+for the existence of a morphism, but in Def.
+19 you are relying on coincident node/edge identities.
+(See my remark 15.) Maybe you mean that the morphism from rl to pc should drive the choice of node/edge identity in the isomorphic copy of rl to be combined with pc, but this is really a wild guess on my part.}
+
 The top equation in \cref{def:rul_comb_partial_total} defines the $\stackrel{combine}{\rightarrow}$ relation for when rule $rl$ has dependencies that are satisfied by path condition $pc$.

### 2016-05-03 text/dsltrans.tex
Add reviewer's comments into the paper.
Add all comments from third reviewer.

@@
 constructs from~\cite{DBLP:conf/sle/BarrocaLAFS10}. A formal treatment of the syntax and semantics of DSLTrans is found in \cref{sec:DSLTrans_formal_appendix}.
+
+\reviewer{Please explain, earlier on, the meaning of indirect edges.}
 
@@
 transformation will execute layer-by-layer, where transformation rules in a layer will execute in a
-non-deterministic order but must produce a deterministic result, due to the fact that DSLTrans is confluent by construction~\cite{DBLP:conf/sle/BarrocaLAFS10}.
+non-deterministic order but will always produce a deterministic result, due to the fact that DSLTrans is confluent by construction~\cite{DBLP:conf/sle/BarrocaLAFS10}.
 
@@
 have a direct association to all her female subordinates and likewise for male
-officers. Note that differences in the gender classification metamodel means
+officers. Note that differences in the gender classification metamodel mean
 some relations present in the input model will not be retained in the output
@@
 \end{figure*}
+\reviewer{Figure 3: I would suggest adding another node f5 superviesd by f4, to illustrate the
+notion of transitive closure. Another remark: I think the black containment dia-
+monds do not belong in an instance model, as they are a metamodel notion.}
 
@@
 Notice that the elements $s$, $m_k$  and $f_k$ in \cref{fig:OrganizationLanguage} are instances of the source \emph{Organization} metamodel elements $Station$, $Male$ and $Female$ respectively.
-The primed elements in \cref{fig:GenderLanguage} are their counterpart instances in the target \emph{Gender} metamodel.
+The primed elements in \cref{fig:police_gender} are their counterpart instances in the target \emph{Gender} metamodel.
 
@@
 elements in the input model which are of type \emph{Station} in the
-\emph{Organization Language}, a element of type
+\emph{Organization Language}, an element of type
 \emph{Station} in the \emph{Gender Language} will be created in the output model.
 
-Note that in our approach, we require that the match graph of a rule is not a subset of the match graph of any other rule (as formally stated in \cref{def:layer_transformation} of model transformation, in \cref{sec:formal_background} of this paper).
+Note that in our approach, we require that the match graph of a rule is not a subgraph of the match graph of any other rule (as formally stated in \cref{def:layer_transformation} of model transformation, in \cref{sec:formal_background} of this paper).
 This requirement is to prevent the case where a rule could not execute independently of another rule, except for the cases when such dependency is explicitly defined by backward links.
 This is undesirable for the algorithm as presented here as we will explain later.
-However, as seen in~\cite{conf/gg/SelimLCDO14}, the expressiveness of the transformations our algorithm can examine is not restricted.
+However, as seen in~\cite{conf/gg/SelimLCDO14}, the expressiveness of the transformations our algorithm can examine \reviewer{Perhaps 'the expressiveness of the rule language'?}  is not restricted.
 In that work, we detail an operational rule processing step to handle overlapping rules.
+
+\reviewer{In page 10 (answer to the reviewers) you say that rules for a 
+  layer must be parallel independent. Is this written anywhere in the
+  main part of the paper? }
 
@@
 
+\reviewer{Figure 4: The two “agent” edges in the postcondition should be marked “female”
+resp. “male”} 
 In contrast, we do not expect the property in \cref{fig:dsltrans_prop2} to always hold. This property represents the statement ``\emph{any model which includes a
@@
 It is not difficult to construct an input model where the pre-condition holds, but the post-condition does not.
-This would be an input model that contains only one female officer, as there will only be one female officer in the output model.
+For example, take an input model that contains only one female officer, as there will only be one female officer in the output model.
 
@@
 matched instances. Our notion of indirect links captures only
-acyclic EMF containment associations.
+acyclic EMF containment associations. \reviewer{“a nonempty path” (I think). The label of the
+containment associations does not matter?}
 

### 2016-05-03 text/formal_background.tex
Add reviewer's comments into the paper.
Add all comments from third reviewer.

@@
 Notice that these concepts are well known from graph theory and are only slightly customized for our purposes.
+
+\reviewer{Your formalisation is made more complex than necessary because you do not put
+the power of typed graphs to work. If your source and target metamodels have
+disjoint types (which you can assume without loss of generality) you can take
+their union, augmented with trace edge types from all target node types to all
+source node types, and take that to be the meta-model for an input-output model,
+and also for a rule. Suddenly you don’t have to carry arround the distinction be-
+tween Match and Apply (for rules) or input and output (for input-output models)
+around any more: they are just the projections onto the types of the source, resp.
+target metamodel.}
+
+\reviewer{Section 4.2.2 on “backward links” is rather confusing, as it speaks about trace-
+ability links being added to rules. You never add traceability links to rules any-
+where in the paper; in fact, your formal notion of a rule has no room for them.
+Instead, rules only have backward links. (The clarity of the the situation is not
+improved by the fact that those are labelled trace.)}
+
+\reviewer{I think the situation would be much more clear if you were to use the common
+notion of left hand side and right hand side of a rule. The left hand side of a
+rule rl is precisely your ||rl||, the right hand side is the entire rl augmented by
+your traceability links, as in 4.2.2. Suddenly we are back in a well-known graph
+transformation formalism, and your rules are in fact so simple (no deletion, no
+negative application conditions, injective matching) that I believe it makes no
+difference whether you use algebraic graph transformation or some more con-
+structively defined variant. The only standard notion is that of transitive closure
+in the left hand side; but there are plenty of GT tools that offer this extension.
+In fact I see no reason why your rule application would then not precisely coin-
+cide with the the same notion in, say, SPO graph rewriting; and I hope you agree
+that this would help you no end in explaining what you are doing. (If, on the
+other hand, there is after all some difference then this, too, would be interesting
+to know, as that difference is not at all apparent right now.)}
+
+\reviewer{Viewed like this, moreover, I think your notions of input-output models and rules
+are very close to triple graph transformation as used by Schrr at all. I think the
+main (if not only) difference lies in the fact that your "glue graph" is actually not
+a graph; instead, you use traceability edges directly from the target model to the
+source model. If you were to turn those edges into nodes with source and target
+edges, even that difference would disappear.}
 
@@
 This section can be safely skipped or skimmed by the reader, who can return to these definitions punctually to understand the detailed formal underpinning of our approach.
+
+\reviewer{There is a lot of redundancy in the paper, on several levels:
+  - input-output-models, path conditions and properties are
+    all defined very similarly. Couldn't you subsume them in a single
+    definition and spell out the differences?}
 
@@
 
+\reviewer{Graph union (Def. 2) is only well-defined if s $\cup$ s' , t $\cup$t' and $\tau \cup \tau$' yield functions,
+which in turn is only the case if s and s' etc. coincide on common elements.
+This is neither stated as a requirement here nor checked anywhere in the use of
+the union. (Since I sure you implicitly had this restriction in mind, this may be
+counted as sloppiness; you can probably assume this to hold w.l.o.g. wherever
+you need it.)}
 \CatchFileBetweenTags{\typedgraphhomomorphismtext}{text/definitions}{typedgraphhomomorphismtext}{\typedgraphhomomorphismtext}
@@
 
+\reviewer{Def. 3 and all subsequent definitions g has an unexplained component st, is
+this the same as the pair (s,t) in Def. 1?}
+
+\reviewer{For multigraphs (which may contain parallel edges with the same label) it is
+necessary to include edge images in the homomorphism (Def. 3). If you do not
+do that, then graphs with parallel edges are isomorphic to graphs with single
+edges (between the same nodes), so you might as well restrict to simple graphs.}
+
+\reviewer{
+Concretely: Right now the def. requires the existence of \textit{some} unfixed e' in the
+target graph, but in fact \textit{f} should be extended to E $\rightarrow$ E' so that e' = f (e) in this
+situation.
+}
 % \begin{lemma}{The Union of Two Typed Graph Homomorphisms is a Typed Graph Homomorphism}
@@
 
+\reviewer{Def. 6: Since the codomain of $\tau$ is apparently VT $\times$ {abstract, concrete} $\cup$ ET $\times$
+{containment, reference}, it is certainly never going to be a bijection. However,
+I think you actually only want injectiveness; what purpose is served by requiring
+it to be surjective?}
+
+\reviewer{As an aside, I think this is quite an awkward way to encode the notions of ab-
+stractness and containment: I think it would be much more elegant to define
+abstractness as a predicate over VT and containment as a predicate over ET. To
+go even one step further, it would in fact be more natural (and more standard)
+to equate the notions of type graph and metamodel altogether, as they certainly
+serve the same conceptual purpose.
+}
+
+\reviewer{Def. 6 and many other places A word in math mode should always
+be in $\backslash$mathit to avoid ugly spacing as in $re f erence$.}
+
 \CatchFileBetweenTags{\metamodeltext}{text/definitions}{metamodeltext}{\metamodeltext}
@@
 \end{definition}
+
+\reviewer{Def. 7: is an expanded metamodel a metamodel? Then you need to allow non-
+injective $\tau$, at least on edges, as the requirement $\tau$(e) = $\tau$(e') in the third bullet
+will certainly cause non-inejectiveness.}
 
@@
 \CatchFileBetweenTags{\metamodelinstancetext}{text/definitions}{metamodelinstancetext}{\metamodelinstancetext}
+
+\reviewer{Def. 9 and many thereafter incrementally define new concepts on top of graphs.
+Most of these definitions are rather hard to read, for several reasons:}
+
+\begin{itemize}
+\item It is common to use an unprimed version for the original object and a
+primed version for the constructed object, whereas you turn this conven-
+tion around
+%\item Sloppiness: in more than 50 percent of all cases the ⊆ should be ⊇ or vice versa;
+%in this particular instance, you really mean τ ′ ⊇ τ and E = E ′ ∪ E c ∗ .
+
+\item It is really more readable to define the construction as a construction, so:
+\begin{itemize}
+\item  $E = E'\cup E^{\star}_{C}$
+\item $s = s' \cup {(e_1 \dots e_n , s(e_1)) | e_1 \dots e_n \in E^{\star}_{C} }$
+\item $t = t' \cup {(e_1 \dots e_n ,t(e_n )) | e_1 \dots e_n \in E^{\star}_{C} }$
+\item $\tau = \tau' \cup {(e, indirect) | e \in E^{\star}_{C} }$
+\end{itemize}
+\begin{itemize}
+\item Note that this also takes care of s and t, which you have forgotten here
+and in all other similar definitions.)
+\item The set $E^{\star}_{C}$should itself also be more carefully defined, as $E^{\star}_{C} = {e_1 \dots e_n \in E^∗' | \forall1 \leq i < n.t(e_i ) = s(e_{i+1} )}$
+\end{itemize}
+\end{itemize}
 
@@
 
+\reviewer{Def. 10: Here you certainly should not require $\tau$ to be injective. The requirement
+on containment relations is mathematical nonsense: if $e' \in E'$ has target b then
+$b \in V'$ and so f (b) does not exist, and most certainly is not the source of any edge
+in E. In fact, I really have no idea what you want to say here; the explanation
+below the definition does not clarify matters at all.}
+
 \CatchFileBetweenTags{\modeltext}{text/definitions}{modeltext}{\modeltext}
+
+\reviewer{“the cardinality of the target class”?? Are you talking about edge multiplicities here (which you have chosen not to include)?}
 
@@
 
+\reviewer{Why is Input a model and Output a metamodel instance? Note that
+$\backslash$ binds stronger than $\cup$ in $E \backslash E' \cup E$ ′′ . “match-apply patterns” → “input-output
+patterns”}
 %Change if needed
@@
 
+\reviewer{e in $\tau(e) = \mathit{trace}$ is still bound by $\nexists$, so “and” should be $\land$}
+
 \CatchFileBetweenTags{\transformationruletext}{text/definitions}{transformationruletext}{\transformationruletext}
@@
 
+\reviewer{Def. 14: According to your own criterion, $||rl||$ is not a rule as it does not create
+either a node or an edge.}
+
 \Cref{def:back_match_transformation_rule} \CatchFileBetweenTags{\backmatchtransformationruletext}{text/definitions}{backmatchtransformationruletext}{\backmatchtransformationruletext}
@@
 \end{definition}
+
+\reviewer{The first bullet is a special case of the second, so can be removed.}
 \CatchFileBetweenTags{\expandedtransformationruletext}{text/definitions}{expandedtransformationruletext}{\expandedtransformationruletext}
@@
 \end{definition}
+
+\reviewer{The second half of Def. 16 raises a lot of questions. First of all, neither $\sqsubseteq 	 $ nor
+$\cong$ are defined over rules; however, I can imagine what it would look like. Then,
+what role does rl play; do you just mean $||rl_1 || \sqsubseteq ||rl_2 ||$ given a suitable isomorphic copy? What if $||rl_1 || \cong ||rl_2 || $(as a special case); is your condition then not
+unsatisfiable as $rl_1$ must appear after $rl_2$ as well as vice versa?}
+
+\reviewer{More importantly though, why do you care? Your semantics makes sure that
+rules are never considered for application twice in succession, so dependencies
+are always resolved unambiguously.}
 
@@
 
+\reviewer{Definition 17: There is a copy/paste error w.r.t. Def. B.16 (which only serves to
+strengthen my point, made above, about avoiding this kind of duplication). In
+this case, B.16 is the correct version.}
+
 \begin{definition} {Model Transformation Execution\\}

### 2016-05-03 text/introduction.tex
Add all comments from third reviewer.

@@
 
+\reviewer{On Page 2, left, mid-page you state that your methods “differ from previous
+work”in that you require no intermediate representation. It is certainly not preor-
+dained that this difference is an advantage: such an intermediate representation
+typically gives access to general-purpose tooling, e.g., for model checking or
+theorem proving. You would have to argue that you are better off reimplement-
+ing that functionality in your own setting. Is there any evidence for that? An
+example of such evidence (in a somewhat different context) is for instance pro-
+vided by Pennemann in his PhD thesis, where by experiments he shows that his
+dedicated, graph-based resolution outperforms that of a general-purpose theorem
+prover.}
 % This also provided us with performance gains over our
@@
 work.
+
+\reviewer{The paper is inconsistent (maybe due to the iteration of submitted versions) in
+what it says about its own structure. Page 2, right, 1st paragraph states that the
+language and its samentcs are presented in Section 2; Section 2 (1st paragraph)
+says that the semantics is presented in Appendix B; most of it is actually in
+Section 3. Appendix B indeed repeats essentially all of Section 3, a duplication
+which certainly serves no purpose and should be remedied. In fact, I would much
+prefer all of Appendix B to be moved to the main text, as the only part that is
+now only in the appendix, the definition of a transformation, is absolutely central
+to the paper.
+As it currently stands, Section 3.4 has a similar problem: first it states that the
+semantics is not in the main text, then it refers to Section 3.2 for the semantics
+(which the reader has just gone through), then in Def. 17 jumps to a notation that
+is only introduced in Appendix B.}

### 2016-05-03 text/new_formal_background.tex
Started to revise formal background based on reviewer's comments.

\section{Formal Background}
\label{sec:formal_background}

In this section we will introduce the formal concepts that will be used throughout all this paper. We start in \cref{sec:typed_graphs} by a few (typed) graph concepts that will be used as mathematical building blocks throughout this paper. In particular we will introduce the notion of typed graph, typed graph union and subset, and useful relations between typed graphs based on homomorphisms. Notice that these concepts are well known from graph theory and are only slightly customized for our purposes.

\reviewer{Your formalisation is made more complex than necessary because you do not put
the power of typed graphs to work. If your source and target metamodels have
disjoint types (which you can assume without loss of generality) you can take
their union, augmented with trace edge types from all target node types to all
source node types, and take that to be the meta-model for an input-output model,
and also for a rule. Suddenly you don’t have to carry arround the distinction be-
tween Match and Apply (for rules) or input and output (for input-output models)
around any more: they are just the projections onto the types of the source, resp.
target metamodel.}

\reviewer{Section 4.2.2 on “backward links” is rather confusing, as it speaks about trace-
ability links being added to rules. You never add traceability links to rules any-
where in the paper; in fact, your formal notion of a rule has no room for them.
Instead, rules only have backward links. (The clarity of the the situation is not
improved by the fact that those are labelled trace.)}

\reviewer{I think the situation would be much more clear if you were to use the common
notion of left hand side and right hand side of a rule. The left hand side of a
rule rl is precisely your ||rl||, the right hand side is the entire rl augmented by
your traceability links, as in 4.2.2. Suddenly we are back in a well-known graph
transformation formalism, and your rules are in fact so simple (no deletion, no
negative application conditions, injective matching) that I believe it makes no
difference whether you use algebraic graph transformation or some more con-
structively defined variant. The only standard notion is that of transitive closure
in the left hand side; but there are plenty of GT tools that offer this extension.
In fact I see no reason why your rule application would then not precisely coin-
cide with the the same notion in, say, SPO graph rewriting; and I hope you agree
that this would help you no end in explaining what you are doing. (If, on the
other hand, there is after all some difference then this, too, would be interesting
to know, as that difference is not at all apparent right now.)}

\reviewer{Viewed like this, moreover, I think your notions of input-output models and rules
are very close to triple graph transformation as used by Schrr at all. I think the
main (if not only) difference lies in the fact that your "glue graph" is actually not
a graph; instead, you use traceability edges directly from the target model to the
source model. If you were to turn those edges into nodes with source and target
edges, even that difference would disappear.}

Armed with the fundamental notion of typed graph, we can then introduce other formal concepts in Sections~\ref{sec:DSLTrans_formal}, \ref{sec:DSLTrans_syntax} and \ref{sec:DSLTrans_semantics} which describe the artifacts from the modeling and transformation world that we require for our verification technique. Naturally, we start by introducing the central notion of \emph{metamodel}, allowing the description of the inputs and outputs of a model transformation. Other fundamental notions we will define in this section are \emph{model}, \emph{transformation rule}, \emph{transformation} and the semantic concept of \emph{model transformation execution}. Several auxiliary and intermediate notions for defining the syntax and semantics of our techniques will also be introduced here.

Note that this section presents a collection of formal tools that are used in the subsequent sections of this paper where the contributions of this paper are presented. It is meant as a formal reference for the upcoming formal development. This section can be safely skipped or skimmed by the reader, who can return to these definitions punctually to understand the detailed formal underpinning of our approach.

\reviewer{There is a lot of redundancy in the paper, on several levels:
  - input-output-models, path conditions and properties are
    all defined very similarly. Couldn't you subsume them in a single
    definition and spell out the differences?}

\subsection{Graph Structures and Functions}
\label{sec:typed_graphs}

\subsubsection*{Typed Graph}
We will start by introducing the notion of typed graph. A typed graph is the essential object we will use throughout our mathematical development. Typed graphs will be used to formalise all the important graph-like structures we will present in this paper. A typed graph is a directed multigraph (a graph allowing multiple edges between two vertices) where vertices and edges are typed.

\begin{definition}{Typed Graph\\}
\label{def:typed_graph}
A typed graph is a 6-tuple $\langle V,E,(s,t),\tau, VT, ET\rangle$ where:
\begin{itemize}
\item $V$ is a finite set of vertices
\item $E$ is a finite set of directed edges connecting the vertices $V$
\item $(s,t)$ is a pair of functions $s: E\rightarrow V$ and $t: E\rightarrow V$ that respectively provide the source and target vertices for each edge in the graph
\item Function $\tau:V\cup E\rightarrow VT \cup ET$ is a typing function for the elements of $V$ and $E$, where $VT$ and $ET$ are disjoint finite sets of vertex and edge type identifiers and $\tau(v)\in VT$ if $v\in V$ and $\tau(e)\in ET$ if $e\in E$
\item Edges $e\in E$ are noted $v\xrightarrow{e} v'$ if $s(e)=v$ and $t(e)=v'$, or simply e if the context is unambiguous
\item The set of all typed graphs is called $\textsc{Tg}$. 
\end{itemize}
\end{definition}

We also assume the existence of a function $\mathit{matchesOver}$ to handle polymorphism issues for $VT$ and $ET$. Let $\mathit{matchesOver}: \{VT \cup ET\} \times \{VT \cup ET\} \rightarrow \{\mathit{true}, \mathit{false}\}$ where:
\begin{itemize}
\item $\mathit{matchesOver}(T, T) \rightarrow \mathit{true}$
\item $\mathit{matchesOver}(T, T') \rightarrow \mathit{true}$ if T' is a subclass of T in some defined partial ordering $\leq$
\item Otherwise, $\mathit{false}$
\end{itemize}

\bentley{$\leq$ might need more explanation}

\subsubsection*{Typed Graph Union}
We now define how two typed graphs are united. A union of two typed graphs is trivially the set union of all the components of those two typed graphs. Note that we do not require the components of the two graphs to be disjoint, as in the following joint unions will be used to merge typed graphs.

\begin{definition}{Typed Graph Union\\}
\label{def:typed_graph_union}

Let $\langle V,E,(s,t),\tau,VT,ET\rangle = g$, and\\ $\langle V',E',(s',t'),\tau',VT',ET'\rangle = g'$, where $g, g' \in \textsc{Tg}$.%where $VT$ and $ET'$ are disjoint sets, as well as $VT'$ and $ET$.
The typed graph union is the function $\sqcup :\textsc{Tg}\times \textsc{Tg}\rightarrow \textsc{Tg}$ defined as:

\begin{multline*}
\big\langle V,E,(s,t),\tau,VT,ET\big\rangle\;\sqcup\;\big\langle V',E',(s',t'),\tau',VT',ET'\big\rangle=\\\big\langle V\cup V', E\cup
E',(s\cup s', t\cup t'), \tau\cup \tau', VT\cup VT', ET\cup ET'\big\rangle
\end{multline*}
\end{definition}

Note: as a reviewer helpfully pointed out, we require that s $\cup$ s', t $\cup$t', and $\tau \cup \tau$' coincide on common elements. However, this can be assumed w. l. o. g..

\subsubsection*{Typed Graph Homomorphism}
For the formal development of our technique, we are interested in relations between typed graphs that are structure-preserving, i.e. homomorphisms. Homomorphisms between typed graphs preserve not only structure, but also the types of vertices and edges that are mapped. Note that, trivially, a typed graph homomorphism is a graph homomorphism.

\begin{definition}{Typed Graph Homomorphism\\}
\label{def:typed_graph_homomorphism}
Let $\langle V,E,(s,t),\tau,VT,ET\rangle=g$ and $\langle V',E',(s',t'),\tau',VT',ET'\rangle=g'\in \textsc{Tg}$, where $g, g' \in \textsc{Tg}$. A typed graph homomorphism between $g$ and $g'$ is a function $f: f_v \cup f_e$ such that:
\begin{itemize}
\item $f_v: V\rightarrow V'$
\item $f_e: E\rightarrow E'$
\item $\forall v_1 \xrightarrow{e} v_2\in E$:
\begin{itemize}
\item $f_v(v_1), f_v(v_2) \in V'$
\item $f_e(e) \in E'$
\item $s'(f_e(e)) = f_v(v_1)$, $t'(f_e(e)) = f_v(v_2)$
\item $\mathit{matchesOver}(\tau(v_1), \tau(f_v(v_1)) \land \\\mathit{matchesOver}(\tau(v_2), \tau(f_v(v_2)) \land\\ \mathit{matchesOver}(\tau(e), \tau(f_e(e))$
\end{itemize}

%, where $\tau(v_1)=\tau'(f(v_1))$, $\tau(v_2)=\tau'(f(v_2))$ and also $\tau(e)=\tau(e')$.
\item The domain of $f$ is noted $Dom(f)$ and the co-domain of $f$ is noted $CoDom(f)$.
\end{itemize}  
\end{definition}

 When an \emph{injective} typed graph homomorphism $f$ exists between $g$ and $g'$ we write $g \stackrel{f}{\vartriangleleft} g'$, or simply $g \vartriangleleft g'$ when the context is unambiguous. When a \emph{surjective} typed graph homomorphism $f$ exists between typed graphs $g$ and $g'$ we write $g \stackrel{f}{\blacktriangleleft} g'$, or also simply $g \blacktriangleleft g'$ in an unambiguous context.

\reviewer{For multigraphs (which may contain parallel edges with the same label) it is
necessary to include edge images in the homomorphism (Def. 3). If you do not
do that, then graphs with parallel edges are isomorphic to graphs with single
edges (between the same nodes), so you might as well restrict to simple graphs.}

\reviewer{
Concretely: Right now the def. requires the existence of \textit{some} unfixed e' in the
target graph, but in fact \textit{f} should be extended to E $\rightarrow$ E' so that e' = f (e) in this
situation.
}

We now define the useful notion of typed subgraph. As expected, a typed subgraph is simply a restriction of a typed graph to some of its vertices and edges.

\begin{definition}{Typed Subgraph\\}
\label{def:typedsubgraph}
Let $\langle V,E,st,\tau,VT,ET\rangle=g,\langle V',E',st,\tau',VT',ET'\rangle=g'\in \textsc{Tg}$ be typed graphs. $g'$ is a typed subgraph of $g$, written $g'\sqsubseteq g$, iff $V'\subseteq V$, $E'\subseteq E$ and $\tau'=\tau |_{V'\cup E'}$.

\end{definition}

Two typed graphs are said to be isomorphic if they have exactly the same shape and related vertices and edges have the same type.

\begin{definition}{Typed Graph Isomorphism\\}
\label{def:typed_graph_isomorphism}
Let $\langle V,E,st,\tau,VT,ET\rangle=g,\langle V',E',st',\tau',VT',ET'\rangle=g'\in \textsc{Tg}$ be typed graphs. We say that $g$ and $g'$ are isomorphic, written $g\cong g'$, if and only if there exists a bijective typed graph homomorphism $f:V\rightarrow V'$ such that $f^{-1}:V'\rightarrow V$ is a typed graph homomorphism.
\end{definition}

\paragraph{\textbf{Notation:}}
In order to simplify our notation, when the context is unambiguous we will abbreviate a typed graph $\langle V,E,st,\\\tau,VT, ET\rangle$ as as 4-tuple $\langle V,E,st,\tau\rangle$. Also, given a typed graph $g\in \textsc{Tg}$, will use the notation $Components(g)$ to describe the set of strongly connected typed graphs in $g$. Finally, we will use the notation $g|_{t}$ to refer to the restriction of graph $g$ to its subgraph containing only edges of type $t$.

\subsection{Metamodel and Model-Related Constructs}
\label{sec:DSLTrans_formal}

We will start by introducing the notion of \emph{metamodel}, which in DSLTrans is used to type the input and output models of a DSLTrans transformation.  Two metamodels, the \emph{organization language} and \emph{gender language} are depicted in \cref{fig:squadmetamodel}.

\begin{definition}{Metamodel\\}
\label{def:metamodel}
A metamodel is a 5-tuple $\langle V,E,st,\tau,\leq\rangle$ where $\langle V,E,st,\tau\rangle\in \textsc{Tg}$ is a typed graph, $(V,\leq)$ is a partial order and $\tau$ is a bijective typing function. Additionally we also have that: if $v\in V$ then $\tau(v)\in VT\times \{abstract,concrete\}$, where $VT$ is the set of vertex type names; if $e\in E$ then $\tau(e)\in ET\times \{containment,\\reference\}$, where $ET$ is a set of edge type names. The set of all metamodels is called $\textsc{Meta}$.
\end{definition}

\reviewer{Def. 6: Since the codomain of $\tau$ is apparently VT $\times$ {abstract, concrete} $\cup$ ET $\times$
{containment, reference}, it is certainly never going to be a bijection. However,
I think you actually only want injectiveness; what purpose is served by requiring
it to be surjective?}

\reviewer{As an aside, I think this is quite an awkward way to encode the notions of ab-
stractness and containment: I think it would be much more elegant to define
abstractness as a predicate over VT and containment as a predicate over ET. To
go even one step further, it would in fact be more natural (and more standard)
to equate the notions of type graph and metamodel altogether, as they certainly
serve the same conceptual purpose.
}

\reviewer{Def. 6 and many other places A word in math mode should always
be in $\backslash$mathit to avoid ugly spacing as in $re f erence$.}

A formal metamodel is a particular kind of typed graph where vertices represent classes and edges represent relations between those classes. A typed graph representing a metamodel has two special characteristics: on the one hand, the typing function for vertices and edges is bijective. This means that each type occurs only once in the metamodel, as is to be expected. On the other hand, a metamodel is equipped with a partial order between vertices. This partial order is used to model inheritance at the level of the metamodel's classes. Note that here we have overridden the co-domain of the typing function in the original typed graph presented in \cref{def:typed_graph} in order to allow distinguishing between \emph{abstract} and \emph{concrete} classes, as well as between \emph{containment} and \emph{reference} edges in our metamodels. For simplification purposes, we do not model association cardinalities in our formal notion of metamodel as cardinalities are not strictly necessary in our development.

\begin{definition}{Expanded Metamodel\\}
\label{def:expanded_metamodel}
Let $mm = \langle V,E,st,\tau,\leq\rangle\in \textsc{Meta}$ be a metamodel. The expansion of $mm$, noted $mm^{\star}$, is a typed graph $\langle V',E',st',\tau'\rangle\in \textsc{Tg}$ built as follows:
\begin{itemize}
  \item $V' = V\setminus \{v\in V\,|\,\tau(v)=(\cdot\footnote{In our mathematical development we use a `dot' notation to represent that we do not care about the value of a particular variable in a given context.},abstract)\}$;
  \item $v_1\xrightarrow{e}v_2\in E'$ if $v_1\xrightarrow{e}v_2\in E$ and $\tau(v_1)=(\cdot,concrete)$ and $\tau(v_2)=(\cdot,concrete)$;
  \item if $v_1\xrightarrow{e}v_2\in E$ we have that $v'_1\xrightarrow{e'}v'_2\in E'$, where $v'_1\leq v_1$, $v'_2\leq v_2$ and $\tau'(e') = \tau(e)$;
  \item for all $v\in V'$ and $e\in E'$ we have that $\tau'(v)=\tau(v)$ and that $\tau'(e)=\tau(e)$.
%   \item $(v,(tv,concrete))\in \tau'$ if $(v,(tv,concrete))\in \tau$ (where $tv$ is a vertex type);
%   \item $(v_1\xrightarrow{e}v_2,(te,concrete))\in \tau'$ if $(v_1\xrightarrow{e}v_2,(te,concrete))\in \tau$ ($te$ ia an edge type).
\end{itemize}
%The set of all expanded metamodels in called $META^{\star}$.
\end{definition}

\reviewer{Def. 7: is an expanded metamodel a metamodel? Then you need to allow non-
injective $\tau$, at least on edges, as the requirement $\tau$(e) = $\tau$(e') in the third bullet
will certainly cause non-inejectiveness.}

An expanded metamodel is an auxiliary construct where all the relations between types of a metamodel are made explicit, rather than remaining implicit in the specialization hierarchy. It is built by adding to the original metamodel typed graph a relation of type $t$ between two classes of the metamodel, whenever those classes specialize two classes that are also related by a relation of type $t$. Abstract classes and their relations do not carry over to the expanded metamodel. Expanded metamodels will be used in the subsequent text to facilitate formal treatment of any structure involving polymorphism.

% Definition~\ref{def:metamodel} formally states that a metamodel corresponds to a graph of typed elements where only one element for each type is represented. This definition is then used to state definition~\ref{def:pattern}, where a typed graph is combined with its metamodel.

\begin{definition}{Metamodel Instance\\}
\label{def:metamodelinstance}

An instance of a metamodel $mm = \langle V',E',st',\tau',\leq\rangle\in \textsc{Meta}$ is a typed graph $\langle V,E,st,\tau\rangle\in \textsc{Tg}$, where the co-domain of $\tau$ equals the co-domain of $\tau'$. Also, there is a typed graph homomorphism $f:V\rightarrow V'$ from $\langle V,E,st,\tau\rangle$ to the expanded metamodel $mm^{\star}$ and the graph $\big\langle V,\{e\in E\,|\,\tau(e)=(\cdot,containment)\}\big\rangle$ is acyclic. The set of all instances for a metamodel $mm$ is called $\textsc{Instance}^{mm}$.

\end{definition}

A metamodel instance is a useful intermediate formal notion that lies between metamodel and model. The injective typed graph homomorphism between a metamodel instance and metamodel models multiple ``instances'' of objects and relations being typed by one single class or relation of the metamodel. Metamodel instances do not allow cyclic containment relations, as enforced by EMF. \reviewer{Do you mean “in analogy to EMF”?}

% It requires the acyclicity constraints as that is required through models, transformation rules, path conditions and properties. It can be used to define transformation rules and can be further refined to define models.

\reviewer{Def. 9 and many thereafter incrementally define new concepts on top of graphs.
Most of these definitions are rather hard to read, for several reasons:}

\begin{itemize}
\item It is common to use an unprimed version for the original object and a
primed version for the constructed object, whereas you turn this convention around
\item Sloppiness: in more than 50 percent of all cases the $\subseteq$ should be $\supseteq$ or vice versa
\begin{itemize}
\item In this particular instance, you really mean $\tau' \supseteq \tau$ and $E = E' \cup E_c^∗$ .
\end{itemize}

\item It is really more readable to define the construction as a construction, so:
\begin{itemize}
\item  $E = E'\cup E^{\star}_{C}$
\item $s = s' \cup {(e_1 \dots e_n , s(e_1)) | e_1 \dots e_n \in E^{\star}_{C} }$
\item $t = t' \cup {(e_1 \dots e_n ,t(e_n )) | e_1 \dots e_n \in E^{\star}_{C} }$
\item $\tau = \tau' \cup {(e, indirect) | e \in E^{\star}_{C} }$
\end{itemize}
\begin{itemize}
\item Note that this also takes care of s and t, which you have forgotten here
and in all other similar definitions.)
\item The set $E^{\star}_{C}$should itself also be more carefully defined, as $E^{\star}_{C} = {e_1 \dots e_n \in E^∗' | \forall1 \leq i < n.t(e_i ) = s(e_{i+1} )}$
\end{itemize}
\end{itemize}

\begin{definition}{Containment Transitive Closure\\}
\label{def:instance_closure}

The containment transitive closure of a metamodel instance $\langle V',E',st',\tau'\rangle\in \textsc{Instance}^{mm}$ is a typed graph $\langle V,E,st,\tau\rangle$ where we have that $V=V'$, $\tau' \supseteq \tau$ and $\tau$'s co-domain is the union of the co-domain of $\tau'$ and the set $\{indirect\}$. We also have that $E' = E\cup E^*_c$ where $E^*_c$ is the transitive closure of the set $\big\{v\xrightarrow{e}v'\,|\,\tau(v\xrightarrow{e}v') = (\cdot,containment)\big\}$ and if $e\in E\setminus E'$ then $\tau(e) = indirect$.  We denote $mi^{*}$ the containment closure of a metamodel instance $mi\in \textsc{Instance}^{mm}$.

\end{definition}

Given a metamodel instance, its containment transitive closure includes, besides the original graph, all the edges belonging to the transitive closure of containment links in that metamodel instance. The transitive edges are typed as \emph{indirect}. In the definitions that follow we will use the $*$ notation, as in \cref{def:instance_closure}, to denote the containment transitive closure of structures that directly or indirectly include metamodel instances. For example, $tg^{*}$ would represent the containment transitive closure of typed graph $tg$ wherever containment edges are found in the graph. Note that the $*$ notation is different from the $\star$ notation, introduced in \cref{def:expanded_metamodel} for an expanded metamodel. \reviewer{The difference between these stars is too small to be noticeable.}

% It requires the acyclicity constraints as that is required through models, transformation rules, path conditions and properties. It can be used to define transformation rules and can be further refined to define models.

\begin{definition}{Model\\}
\label{def:model}

A model of a  metamodel $mm = \langle V',E',st',\tau',\leq\rangle\in \textsc{Meta}$ is a metamodel instance $\langle V,E,st,\tau\rangle\in \textsc{Instance}^{mm}$, such that: there exists an injective typed graph homomorphism $f:V\rightarrow V'$ from $\langle V,E,st,\tau\rangle$ to metamodel $mm^{\star}$ where, if $b\in V$ and there exists an edge $a'\xrightarrow{e'} f(b)\in E'$ where $\tau(e') = (t,containment)$, then we also have that $a\xrightarrow{e} b\in E$, $f(a) = a'$ and $\tau(e)=t$. The set of all models for a metamodel $mm$ is called $\textsc{Model}^{mm}$.

\end{definition}

\reviewer{Def. 10: Here you certainly should not require $\tau$ to be injective. The requirement
on containment relations is mathematical nonsense: if $e' \in E'$ has target b then
$b \in V'$ and so f (b) does not exist, and most certainly is not the source of any edge
in E. In fact, I really have no idea what you want to say here; the explanation
below the definition does not clarify matters at all.}

A model, as per \cref{def:model}, is a metamodel instance where all the containment relations are respected. This means that if an object that is contained (as per the metamodel) exists in the model, then the model will also contain an instance of that containment relation together with an object that contains it. Two models can be observed in \cref{fig:transformationexample}, which are respectively instances of the \emph{Organization language} and \emph{Gender language} found in \cref{fig:squadmetamodel}.\\
Note that the containment constraint does not necessarily lead to infinite models in the case of containment relations with the same source and target classes. In fact, if the cardinality of the target class is allowed to be zero, then it is not necessary that the containment relation is instantiated. For example, this is the case for the containment relation \emph{supervise} in the metamodel of~\cref{fig:OrganizationLanguage}.

\reviewer{“the cardinality of the target class”?? Are you talking about edge multiplicities here (which you have chosen not to include)?}

\begin{definition}{Input-Output Model\\}
\label{def:input_output_model}
An input-output model is a 6-tuple $\big\langle V,E,(s,t),\tau,Input,\\Output\big\rangle$, where: $Input=\langle V',E',st',\tau'\rangle\in \textsc{Instance}^{sr}$ is a model; $Output=\langle V'',E'',st'',\tau''\rangle\in \textsc{Instance}^{tg}$ is a metamodel instance; $Input$ and $Output$ are disjoint. Additionally we have that $V=V'\cup V''$, $E\subseteq E'\cup E''$ and $\tau\subseteq \tau'\cup \tau''$, where the co-domain of $\tau$ is the union of the co-domains of $\tau'$ and $\tau''$ and the set $\{trace\}$. An edge $e \in E\setminus E'\cup E''$ is called a \emph{traceability link} and is such that $s(e)\in V''$, $t(e)\in V'$ and $\tau(e)=trace$. The set of all match-apply patterns for a source metamodel $sr$ and a target metamodel $tg$ is called $\textsc{Iom}^{sr}_{tg}$.
\end{definition}

\reviewer{Why is Input a model and Output a metamodel instance? Note that
$\backslash$ binds stronger than $\cup$ in $E \backslash E' \cup E$ ′′ . “match-apply patterns” → “input-output
patterns”}
%Change if needed
% \levi{A match-apply model is the union of an input and output models, plus potential traceability links between those models. Traceability links are now typed using the typing function $\tau$. A traceability link it typed with type ``trace''. This makes the definitions shorter and more compact, which was mentioned by one of the reviewers of the paper. However, the fact that a match-apply model is now made up of models makes it that the input and the output of the transformation are always well-formed, meaning that the containment relation is always respected. It may be that a transformation does not always produce a model respecting the containment relation, which is a problem. This can be solved however if we only consider that well-formed transformations actually have semantics.}

An input-output model is an object we will use when defining the semantics of a DSLTrans model transformation in \cref{sec:DSLTrans_formal}. It is composed of two metamodel instances, one called the \emph{input} and the other one the \emph{output}. An input-output model allows the representation of intermediate operational states during the execution of a model transformation. It may include a particular type of edges called \emph{traceability links}, for keeping a history of which elements in the output model originated from which elements in the input model.

\begin{definition}{Metamodel Pattern and Indirect Metamodel Pattern\\}
\label{def:metamodel_pattern}

A pattern of a metamodel $mm\in \textsc{Meta}$ is an instance of $mm$. Given a metamodel pattern $\langle V',E',st',\tau'\rangle\in \textsc{Instance}^{mm}$ we have that $\langle V,E,st,\tau\rangle$ is an \emph{indirect} pattern if $V=V'$, $E'\supseteq E$ and the co-domain of $\tau$ is the union of the co-domains of $\tau'$ and the set $\{indirect\}$. Also, if $e\in E\setminus E'$, then we have that $\tau(e)=indirect$. Given a metamodel $mm$, the set of all metamodel patterns for $mm$ is called $\textsc{Pattern}^{mm}$. The set of all indirect metamodel patterns for $mm$ is called $\textsc{Ipattern}^{mm}$.
\end{definition}

Metamodel patterns are introduced in \cref{def:metamodel_pattern} as an intermediate notion, formally equal to metamodel instances. An \emph{indirect} metamodel pattern is a metamodel pattern that includes edges typed as \emph{indirect}. Both structures will be used as building blocks in the construction of transformation-related structures in the upcoming text.

\subsection{Syntactic Transformation Constructs}
\label{sec:DSLTrans_syntax}

This section will detail the abstract syntax of the constructs involved in a DSLTrans transformation.

\begin{definition}{Transformation Rule\\}
\label{def:transformation_rule}

A transformation rule is a 6-tuple $\big\langle V,E,(s,t),\tau, Match,\\Apply\big\rangle$, where: $Match=\langle V',E',st',\tau'\rangle\in \textsc{Ipattern}^{sr}$ such that: $Match\neq \epsilon$\footnote{We use the simplified $\epsilon$ notation to denote empty n-tuples structures.} is an non-empty indirect metamodel pattern; $Apply=\langle V'',E'',st'',\tau''\rangle\in \textsc{Pattern}^{tg}$ such that $Apply\neq\epsilon$ is a metamodel pattern; $Match$ and $Apply$ are disjoint. We also have that $V=V'\cup V''$, $E\subseteq E'\cup E''$ and $\tau\subseteq \tau'\cup \tau''$, where the co-domain of $\tau$ is the union of the co-domains of $\tau'$ and $\tau''$ and the set $\{trace\}$. An edge $e \in E\setminus E'\cup E''$ is called a \emph{backward link} and is such that $s(e)\in V''$, $t(e)\in V'$ and $\tau(e)=trace$. We additionally impose that there always exists a $v_1\in V''$ in the Apply part of the rule such that $\nexists e\,.\, v_1\xrightarrow{e} v_2$ and $\tau(e)=trace$, or that $E''$ is not empty. The set of all transformation rules for a source metamodel $sr$ and a target metamodel $tg$ is called $\textsc{Rule}^{sr}_{tg}$.

\end{definition}

\reviewer{e in $\tau(e) = \mathit{trace}$ is still bound by $\nexists$, so “and” should be $\land$}

A transformation rule is the elemental block of a model transformation. Several transformation rules can be observed in the Police Station transformation in \cref{fig:dsltransformation}. A formal transformation rule includes a non-empty match pattern and a non-empty apply pattern (also known in the model transformation literature as a rule's \emph{left hand side} and \emph{right hand side}). The apply pattern of a rule always contains at least one apply element that is not connected to a backward link or an edge, meaning in practice that a rule will always produce something and not only match. A match pattern can include indirect links that are used to transitively match containment relations in a model. An apply pattern does not include indirect links as it is used only for the construction of parts of instances of a metamodel. A transformation rule includes backward links, as informally introduced in \cref{subsec:DSLTrans_constructs}. Backward links are formally typed as \emph{trace}.%This eases modeling matching rules on models  by the use of typed graph homomorphisms.

\begin{definition}{Matcher of a Transformation Rule\\}
\label{def:back_match_transformation_rule}

Let $rl = \big\langle V,E,st,\tau, Match,Apply\big\rangle$ be a transformation rule where $Match =\langle V_m, E_m, st_m, \tau_m\rangle$. We define $rl$'s matcher, noted $\Vert rl\Vert$, as the transformation rule $\big\langle V',E',st',\tau',Match,\\Apply'\big\rangle\sqsubseteq rl$ where $v_1\xrightarrow{e} v_2\in E'$ if and only if $v_1,v_2\in Match$ or $\tau(e)=trace$ and $V'=V_m\cup \big\{v_1\,|\,v_1\xrightarrow{e} v_2\in E \land \tau(e)=trace\big\}$.
\reviewer{Definition 14: where does Apply' come from? Should v1 be 
  an element of Apply?}
\end{definition}

\reviewer{Def. 14: According to your own criterion, $||rl||$ is not a rule as it does not create
either a node or an edge.}

\Cref{def:back_match_transformation_rule} introduces the notion of matcher for a transformation rule which consists solely of the match pattern of a rule and its backward links, if any. The matcher of a rule constitutes the complete pattern that a DSLTrans rule attempts to match over a input-output model during rule execution. Traceability links between input and output model elements generated during transformation execution are matched by transformation rules' backward links, as informally explained in \cref{subsec:DSLTrans_constructs}.

\begin{definition}{Expanded Transformation Rule\\}
\label{def:transformation_rule_expansion}

Let $rl = \big\langle V,E,st,\tau,Match,Apply\rangle \in \textsc{Rule}^{sr}_{tg}$ be a transformation rule where $Match = \langle V',E',st',\tau'\rangle$ and also we have that $sr =\langle V'',E'',st'',\tau'',\leq \rangle$. The expansion of $rl$, noted $rl^{\star}$ is a set of transformation rules built as follows:
\begin{itemize}
  \item $rl\in rl^{\star}$;
  \item $\big\langle V,E,st,\tau',Match,Apply\rangle\in rl^{\star}$ iff for all $v\in V'$ we have that $\tau'(v)\leq \tau(v)$.
\end{itemize}
\end{definition}

\reviewer{The first bullet is a special case of the second, so can be removed.}

The expansion of a transformation rule is a set of transformation rules. Each rule in that set includes a possible replacement of each of the classes in the match part of the original rule by one of its subtypes. Expanded transformation rules will be important to ensure that polymorphism is correctly handled in the developments that follow.

%\levi{A transformation rule expansion simply takes a rule and builds all possible rules where a rule element is replaced by one of its subtypes.}

\begin{definition}{Layer, Model Transformation\\}
\label{def:layer_transformation}

A layer is a finite set of transformation rules $l\subseteq \textsc{Rule}^{sr}_{tg}$. The set of all layers for a source metamodel $sr$ and a target metamodel $tg$  is called $\textsc{Layer}^{sr}_{tg}$. A model transformation is a finite list of layers denoted $[l_{1}::l_{2}::\ldots::l_{n}]$ where $l_{k}\in \textsc{Layer}^{sr}_{tg}$ and $1\leq k \leq n$, $n\in \mathbb{N}$. We also impose that for any pair of rules $rl_1,rl_2\in \bigcup_{1\leq k\leq n}l_k$, if $\Vert rl_1\Vert \cong rl$ and $rl\sqsubseteq \Vert rl_2\Vert $ then $rl_2$ appears in a layer later than $rl_1$ and the apply parts of $rl_1$ and $rl_2$ are not isomorphic. The set of all transformations for a source metamodel $s$ and a target metamodel $t$ is called $\textsc{Transf}^{sr}_{tr}$.

% The expansion of a layer $l\subseteq RULE^{sr}_{tg}$, noted $l^{\star}$, is such that that if
% $rl\in l$ then $rl'\in l$ if $rl'\in rl^{\star}$. This means that if a
% transformation rule is present is the layer, its expansion is also part of the
% layer.

\end{definition}

\reviewer{The second half of Def. 16 raises a lot of questions. First of all, neither $\sqsubseteq 	 $ nor
$\cong$ are defined over rules; however, I can imagine what it would look like. Then,
what role does rl play; do you just mean $||rl_1 || \sqsubseteq ||rl_2 ||$ given a suitable isomorphic copy? What if $||rl_1 || \cong ||rl_2 || $(as a special case); is your condition then not
unsatisfiable as $rl_1$ must appear after $rl_2$ as well as vice versa?}

\reviewer{More importantly though, why do you care? Your semantics makes sure that
rules are never considered for application twice in succession, so dependencies
are always resolved unambiguously.}

\cref{def:layer_transformation} formalises the abstract syntax of a model transformation, introduced in \cref{sec:dsltrans}. An example of a model transformation can be observed in \cref{fig:dsltransformation}, the Police Station transformation. As expected, a formal DSLTrans transformation is composed of a sequence of layers where each layer is composed of a set of rules. The last condition of \cref{def:layer_transformation} imposes that, for any two pair of rules in the transformation, the matcher of the second rule never partially or totally subsumes (or contains) the matcher of the first rule, unless the second rule is in a subsequent layer and produces something more than the first rule. This condition avoids situations where the execution of a rule in a DSLTrans model transformation necessarily implies the execution of another rule (except for when rules having backward links necessarily execute because all their dependencies were created during the execution of previous layers). \reviewer{I do not fully understand why this
  could be problematic. Do you have an example?}

\paragraph{\textbf{Notation:}}

We naturally extend to transformation rules (\cref{def:transformation_rule}) and transformation executions (\cref{def:modeltransformation}) the typed graph operations introduced in \cref{sec:formal_background}.  Also, given a structure such as transformation rule $rl = \langle V,E,st,\tau,Match,\\Apply\rangle$, we will refer to the structure's components by using the component's name followed by the variable that holds the structure in between parenthesis. For example, we will write $V(rl)$ to designate the $V$ component of $rl$ or $Apply(rl)$ to designate $rl$'s $Apply$ component.

\subsection{Semantic Transformation Constructs}
\label{sec:DSLTrans_semantics}

In the definition that follows we introduce the notion of execution of a DSLTrans model transformation. For our purposes it is sufficient to introduce it as an input-output model (see \cref{def:input_output_model}), containing the input model for the transformation, the produced output, and the traceability links built during execution. Due to space limitations, we cannot introduce the semantics of DSLTrans in the main text of this paper. We thus refer the reader to \cref{sec:DSLTrans_formal} for a formal description on how DSLTrans transformation executions are built.

\reviewer{Definition 17: There is a copy/paste error w.r.t. Def. B.16 (which only serves to
strengthen my point, made above, about avoiding this kind of duplication). In
this case, B.16 is the correct version.}

\begin{definition} {Model Transformation Execution\\}
\label{def:modeltransformation}

Let  $tr\in \textsc{Transf}^{sr}_{tg}$ be a transformation and $input \in \textsc{Model}^{sr}$ be a model. Assume we also have that: $$\langle V,E,st,\tau,input,\epsilon\rangle,tr \xrightarrow{trstep} \langle V',E',st',\tau',input,output\rangle$$
A model transformation execution is the input-output model $\langle V',E',st',\tau',input,output\rangle\in \textsc{Iom}^{sr}_{tg}$, where $output\in \textsc{Iom}^{sr}_{tg}$ is an input-output model. The set of all model transformation executions for transformation $tr$ is written $\textsc{Exec}(tr)$. A model transformation with an empty input model is noted $\epsilon_{ex}$. Note that relation $trstep$ is formally defined in \cref{sec:DSLTrans_formal_appendix}.
\end{definition}

Finally, as stated in \cref{def:modeltransformation}, we consider a model transformation execution to be the input-output model (IOM) resulting from executing a set of rules on a starting IOM. This starting IOM includes the transformation's input in its input part and has an empty output part. The starting IOM represents the first step of the transformation when no rule has been executed yet. A transformation execution results from executing all the rules in a DSLTrans model transformation.

\subsubsection{Transformation Language Semantics}

We will now address the semantics of the DSLTrans language. We will start by defining a match function that, given an input-output model and a transformation rule, returns all subgraphs of that input-output model where the rule's match pattern (including backward links) is found.

% \levi{Need to be careful throughout all this section to understand whether we model the match and apply functions as functions or relations. Because of the non-deterministic generation of new vertices in the definition of apply function~\ref{def:apply_function}, relations are used in the definitions that follow to evaluate a transformation. This is an issue of the formalism used to model, should we make it explicit?}

\begin{definition}{Match Function}
\label{def:match_function}

Let $m_{in}\in \textsc{Iom}^{sr}_{tr}$ be a input-output model and $rl\in \textsc{Rule}^{sr}_{tg}$ be a transformation
rule. The $match : \textsc{Iom}^{sr}_{tg}\times \textsc{Rule}^{sr}_{tg}\rightarrow
\mathcal{P}(\textsc{Iom}^{sr}_{tg})$ function is defined as follows: $$match_{rl}(m_{in})= \big\{ glue_{noind}\;|\;
glue\sqsubseteq m_{in}^{*} \land glue \cong \Vert rl\Vert\big\}$$

where $glue\in \textsc{Iom}^{sr}_{tr}$ is an input-output model and $glue_{noind}$ is a version of $glue$ where the indirect links have been removed.
% \levi{all backward links are replaced by trace links such that the matching using subgraph isomorphim can work.}
% 
% \levi{Where $rl^{-}$ is rule $rl$ without nodes in the apply part that are not connected to \emph{trace links}, as well as links adjacent to those nodes. $mg^{-ind}$ is the match graph without indirect links that come from the rule, given those cannot be part of the intermediate $MAM$ match result since that result is going to be united with the original $MAM$ (this was incorrect in the original SLE paper).}
% \levi{The matcher function returns all graphs of the input MAM that are matched by a rule, including its backward links. Note that to find a subgraph ($\sqsubseteq$) of the input MAM that is isomorphic ($\cong$) to the rule being matched because the graph rewriting is achieved by joint union of the match graph extended by the new rewrite node, and the original MAM.}
% \levi{TODO: review the $mg^{-ind}$ and the $rl^{-}$ notations.}

% Due to the fact that the $\cong$ relation is based on the notion of graph
% isomorphism, permutations of the same match result may exist in the
% $\big\{g\;|\; g\lhd m \land g \cong strip(tr)\big\}$ set. The --- undefined ---
% $remove:\mathcal{P}(TR^{s}_{t})\rightarrow \mathcal{P}(TR^{s}_{t})$ function is such that
% it removes such undesired permutations.

%The $strip:TR^{s}_{t}\rightarrow TR^{s}_{t}$ function is such that
%$$strip(\big\langle V,E,T,\langle V_m,E_m,T_m,s\rangle,\langle V_a,E_a,T_a,t\rangle,Bl,Il\big\rangle) = \big\langle V',E',T,\langle V_m,E_m,T_m\rangle,\langle V'_a,E'_a,T_a\rangle,Bl,Il\big\rangle$$
%\begin{center}
%where $(v'\in V'_a\Rightarrow v\rightarrow v'\in Bl)\;\land\;(v\rightarrow v'\in E'_a\Rightarrow (v\rightarrow v'\in E_m \land \{v,v'\}\subseteq V'_a))$
%\end{center}
\end{definition}

\reviewer{Def. B.12. First of all, I wonder why you define match as a function rather than
just defining what a match is; namely, an element of the set you are constructing
here. Secondly (and more importantly), why do you not use the notion of homo-
morphism instead of explicitly talking about ismorphic subgraphs? Your way,
you lose the connection between glue and $||rl||$: you know they are isomorphic,
but what if $||rl||$ has symmetries - then there is more than one match, which in
your definition cannot be distinguished.}

\reviewer{You pay the price for this loss of information in Def. B.13, which now is broken: $a\Delta$ depends on $m_{glue} $(which was called glue in Def. 12, why the change in
notation?) in a way depending on the mapping from $m_{glue}$ to $rl$ which you have
discarded.}

\reviewer{Def. B.13 is broken in another way as well: it does not properly define E' but
merely imposes some requirements that are also fulfilled by choosing E' = E.
Instead I know from the discussion about traceability links in Section 4 (which
logically comes after this appendix!) that you really want to add such links for
all newly created nodes of the output graph to all pre-existing nodes of the input
graph; but that is not made clear anywhere at this point.}

The match function in \cref{def:match_function} looks for subgraphs ($glue\sqsubseteq m_{in}^{*}$) of an input-output model that are isomorphic to the backward matcher of the given transformation rule ($glue \cong \Vert rl\Vert$). Note that the containment transitive closure of the input-output model ($m_{in}^{*}$) is considered such that indirect links in the rule can looked for in the input model. Additionally, indirect links need to be removed from the input-output models resulting from the match function ($glue_{noind}$). This is so because indirect links are not part of the original input model, but rather an auxiliary structure.

Let us now turn our attention to the apply function in \cref{def:apply_function}. Its role is to extend all model fragments found by matching a rule on a given input-output model, such that each of those fragments becomes isomorphic to the complete rule (minus its backward links). This process effectively creates the new objects and relations specified in the apply part of the rule, for each of the fragments found when matching the rule.

\begin{definition}{Apply Function}
\label{def:apply_function}

\reviewer{In your definitions you rely on the $\sqcup$ operator (sqare union),
  the definition of which is not entirely clear.  For instance, I do not know where the operator used in Def. B13
  is defined.}

Let $m_{glue}\in \textsc{Iom}^{sr}_{tg}$ be a input-output model and $rl\in \textsc{Rule}^{sr}_{tg}$ a
transformation rule. The $apply : \textsc{Iom}^{sr}_{tg}\times \textsc{Rule}^{sr}_{tg}\rightarrow \textsc{Iom}^{sr}_{tg}$ function
is defined as follows:

$$apply_{rl}(m_{in})=\bigsqcup_{m_{glue}\in match_{rl}(m_{in})}trace_{a_{\Delta}}(m_{glue}\sqcup a_{\Delta})$$

\begin{center}
where $a_{\Delta}\in \textsc{Iom}^{sr}_{tg}$ is such that $m_{glue} \sqcup a_{\Delta}\cong rl_{noind}$.\\
\end{center}

We impose that any instance of $a_\Delta$ is always disjoint from the $m_{in}$ input-output model and also that any two instances of $a_\Delta$ used in the large union are always disjoint.

Partial function $trace:\textsc{Iom}\times \textsc{Iom} \nrightarrow \textsc{Iom}$ is such that $trace_{\langle V_{\Delta},E_{\Delta},st_{\Delta},\tau_{\Delta}\rangle}(\langle V,E,st,\tau\rangle) = \langle V,E',st',\tau'\rangle$ where we have that $E\subseteq E'$, $st\subseteq st'$ (using a light notational abuse for the $s\subseteq s'$ and $t\subseteq t'$), $\tau \subseteq \tau'$ and if $v_1\xrightarrow{e} v_2\in E'\setminus E$ then $v_1\in Output(V_{\Delta})$, $v_2\in Input(V)$ and $\tau'(e)=trace$. Finally, $rl_{noind}$ is a version of $rl$ where indirect links have been removed.
% More concretely, if we have $mam = $\big\langle V,E,st,\tau, Input,Output\big\rangle\in MAM^{sr}_{tg}$ where $Input=\langle V',E',st',\tau'\rangle\in MODEL^{sr}$ and $Output=\langle
% V'',E'',st'',\tau''\rangle\in MODEL^{tg}$, then $trace(mam) = \big\langle V,E,st,\tau''',Input,Output\big\rangle$ where $E'\subseteq E$ and $v\rightarrow E'\setminus E$ 
\end{definition}
% 
% it builds traceability edges between vertices of the output model that were newly built by the $apply$ function (the vertices in ${mg}_{\Delta}$) and all the vertices from the input part of the input-output model. It is necessary to remove the indirect links from the rule because otherwise they would be created during the rewriting by the apply function.

In \cref{def:apply_function} $a_\Delta$ is an input-output model that contains an instance of the target metamodel. These instances are created by rule $rl$ and are used to extend the sub-models found by the match function. The $trace$ function builds traceability edges between newly created vertices of the output model in $a_{\Delta}$ and all the vertices from the input part of a model fragment found by the match function.

Note that, because we do not pose any constraints on $a_\Delta$ other than the fact that its union with the sub-model $m$ is isomorphic to $rl_{noind}$, the $a_\Delta$ variable can always be satisfied by an unlimited amount of input-output models. In order to avoid an infinite amounts of results when a transformation rule is executed, in what follows we will consider transformation results differ only up to typed graph isomorphism.

%  Note that because the vertices are not disjoint, when the graph produced by the apply function is united with the original match apply model the new vertices are created. It was necessary to remove the indirect links from the rule during rewriting, because otherwise those indirect links would also be rewritten.

Definitions~\ref{def:match_function} and~\ref{def:apply_function} are complementary: the former gathers all the fragments of an input-output model that are matched by a transformation rule; the latter glues on the output part of each of those fragments new objects and relations created by a transformation rule.

%  The $strip$ function is used to
%  enable matching over backward links but not elements to be created by the
%  transformation rule. The $back$ function connects all newly created vertices
%  to the elements of the source model that originated them.

%\begin{definition}{Transformation}

%Let $g\in TG$ be a typed graph. The graph transform function $transform : TR\times TG\rightarrow TG$ is recursively defined as:
%\begin{gather*}
%  transform_{(m,a)}(g) =
%  \begin{cases}
%    g  & \text{if } match_{m}(g)=\emptyset \\
%    instance(a)\cup transform_{(m,a)}(g') &\text{if } match_{m}(g)\neq \emptyset
%  \end{cases}
%\end{gather*}
%$$\text{where } g'=mark_{(m)}(g)$$
%\end{definition}

%\subsection{Transformation Semantics}
%\begin{definition} {Layer Semantics}

%Let $l\in Layer$ be a layer and $\{m,m'\} \in MAM^{s}_{t}$ be a match
%$$\frac{}
%{m,\emptyset \stackrel{layerstep}{\rightarrow}m}$$
%$$\frac{tr\in layer,\; transform_{tr}(m) = m''\; m'',layer\backslash\left\{tr\right\} \xrightarrow{layerstep}	 m'}{m,layer \xrightarrow{layerstep} m'}$$
%\end{definition}

\begin{definition} {Layer Step Semantics}
\label{def:layer_step_semantics}

Let $l\in \textsc{Layer}^{sr}_{tg}$ be a Layer. The \emph{layer step relation}
$\stackrel{layerstep}{\rightarrow}\subseteq \textsc{Iom}^{sr}_{tg}\times \textsc{Iom}^{sr}_{tg} \times \textsc{Layer}^{sr}_{tg}\times
\textsc{Iom}^{sr}_{tg}$ is defined as follows:

$$\frac{}
{\langle m_{in},m_{glue},\emptyset\rangle \xrightarrow{layerstep}m_{in} \sqcup m_{glue}}$$

\begin{multline*}
\frac{\begin{array}{ll} rl\in l,\; apply_{rl}&(m_{in}) = m_{rout},\\
&\langle m_{in},m_{glue}\sqcup m_{rout},l\backslash \{rl\}\rangle \xrightarrow{layerstep} m_{out}
\end{array}}
{\langle m_{in},m_{glue},l\rangle \xrightarrow{layerstep} m_{out}}$$
\end{multline*}

\begin{center}
where $m_{rout} \in\textsc{Iom}^{sr}_{tg}$ and $rl\in \textsc{Rule}^{sr}_{tg}$.
\end{center}

We impose that all input-output models that are part of $rout$ and have been generated by rule $rl$ are disjoint from input-output models accumulated in $m_{glue}$ that have been generated by other rules.

\end{definition}

\reviewer{Def. B.14: It seems to me rather roundabout to sequentially apply all rules, ac-
cumulate the results and then take their union. Why do you not take the union
simultaneously with the parallel application of all rules (of a given layer), as
you are essentially doing for all applications of a single rule in Def. B.13? This
would get rid of the artificial sequential ordering, and the whole issue of con-
fluence would not even have to be raised as there is no non-determinism any
more.}

\reviewer{In the constructions defined in Appendix B, you ignore the required absence of
containment cycles. Might such cycles not appear in the union of the graphs,
even if the individual graphs are cycle-free?}

In \cref{def:layer_step_semantics} we build the result of executing a layer of a DSLTrans transformation. The operational semantics-like rules in the definition execute each rule $rl$ in layer $l$, in a non-deterministic order, by using the $apply$ function. The result of executing each rule is accumulated in the temporary $m_{glue}$ input-output model. Finally, when the set of transformation rules in the layer has been exhausted, the result of executing all the rules in the layer (now contained in $m_{glue}$) is united with the input-output model $m_{in}$, the input to layer $l$. Note that this final union produces the result we expect because of the fact that the $m_{glue}$ input-output model is not disjoint from $m_{in}$. The common ``glue'' parts of $m_{glue}$ that have been built by the match function and extended by the apply function are now used to built the result of executing layer $l$.

\cref{def:layer_step_semantics} is the core of DSLTrans' semantics. Many model transformation languages are based on graph rewriting, where the result of each rule rewrite is immediately usable by all other rules. In DSLTrans the result of executing one layer in DSLTrans is totally produced before the input to the layer is changed. This is enforced in \cref{def:layer_step_semantics} by the fact that the apply function always executes over the same $m_{in}$ input-output model and all the results of rule execution in the same layer are added to the $m_{glue}$ structure that is write-only. Rules belonging to the same layer are thus forced to execute independently.

\begin{definition} {Transformation Step Semantics}
\label{def:transformation_step_semantics}

Let $[l::tr]\in \textsc{Transf}^{sr}_{tg}$ be a transformation, where $l\in
\textsc{Layer}^{sr}_{tg}$ is a Layer and $tr$ also a transformation. The \emph{transformation step relation}
$\stackrel{trstep}{\rightarrow}\subseteq \textsc{Iom}^{sr}_{tg}\times \textsc{Transf}^{sr}_{tg}\times
\textsc{Iom}^{sr}_{tg}$ is defined as follows: 
$$\frac{}{\langle m,[]\rangle \xrightarrow{trstep} m}$$

$$\frac{\big\langle m_{in},\epsilon,l^{\star}\big\rangle \xrightarrow{layerstep} m_{inter},\; \langle m_{inter},R\rangle \xrightarrow{trstep} m_{out}}{\langle m_{in},[l::T]\rangle \xrightarrow{trstep} m_{out}} \hspace{.3cm}$$

$$\text{where } l^{\star}=\bigcup_{rl\in l}rl^{\star}$$
\end{definition}

\reviewer{I believe tr, T and R in this definition stand for the same thing.}
While the execution of the rules belonging to a layer happens in parallel, the execution of the layers of a transformation happens sequentially. As per \cref{def:transformation_step_semantics}, the input-output model $m_{inter}$ is the output of executing a given layer that is passed onto the next layer as input. Note that an empty input-output model ($\epsilon$) is passed as the second argument to the $layerstep$ relation in \cref{def:transformation_step_semantics}. This is because in \cref{def:layer_step_semantics} of layer step semantics, the second argument of the relation is used as an accumulator for the model fragments that are added to be added to the output of the previous layer once all the rules in the current layer have been executed. The transformation rules in a layer are expanded before execution ($l^{\star}$) such that polymorphism in the match elements can be handled (see \cref{def:transformation_rule_expansion_appendix}).

\begin{definition} {Model Transformation Execution}
\label{def:modeltransformation_appendix}

\CatchFileBetweenTags{\modeltransformation}{text/definitions}{modeltransformationappendix}{\modeltransformation}
\end{definition}

\CatchFileBetweenTags{\modeltransformationtextappendix}{text/definitions}{modeltransformationtextappendix}{\modeltransformationtextappendix}

% A transformation execution is formed from taking the input model, and executing
% the transformation on it to produce the output model. During this
% transformation, traceability links will be placed between match elements and the
% apply elements they create. Definition~\ref{def:transf_ex} expresses this
% formally. Note that we assume that transformation executions are built following
% the algorithm described in~\cite{DBLP:conf/sle/BarrocaLAFS10}.

% \begin{definition}{Transformation Execution}
% \label{def:transf_ex}
% 
% Let $t$ be a DSLTrans transformation having source metamodel $s$ and target
% metamodel $t$. A Transformation Execution is a 6-tuple $\langle V,E\cup
% Tl,\tau,Match,Apply,Tl\rangle$, where $\langle V,E,\tau,Match,Apply\rangle \in
% MAP^{s}_{t}$ is a match-apply pattern. $Match=\langle V',E', \tau',s\rangle$,
% $Apply=\langle V'',E'', \tau'',t\rangle$ and the edges $Tl\subseteq V'\times
% V''$ are called \emph{traceability links}. The set of all transformation
% executions having source metamodel $s$ and target metamodel $t$ is called
% $Exec^{s}_{t}$.
% \end{definition}

We now prove two important properties about executions of transformations expressed in the subset of DSLTrans presented in this paper: \emph{confluence} and \emph{termination}. The proofs are provided at a high level, given the fact that DSLTrans essentially enforces both these properties by construction of the semantics of DSLTrans.

\begin{proposition}{Confluence}

Every model transformation execution is confluent up to typed graph isomorphism.
\end{proposition}
\begin{pf}
We want to prove that for every model transformation execution of a transformation $tr\in \textsc{Transf}^{sr}_{tg}$ having as input a model $input \in MODEL^{s}$, its output is always the same up to typed graph isomorphism.\\
If we assume an execution of the transformation is not confluent then this should happen because of non-determinism when the execution of a transformation is being built. Non-determinism happens during the construction of a transformation execution at two points: 
\begin{enumerate}
\item in definition~\ref{def:apply_function}, $a_{\Delta}$ is non-deterministic up to typed graph isomorphism, which does not contradict the proposition we are trying to prove.
\item in definition~\ref{def:layer_step_semantics} transformation rule $rl$ is chosen non-deterministically from layer $l$. Thus, the order in which the transformation rules are treated is non-deterministic. However, by \cref{def:layer_step_semantics} rules in a layer execute independently, which means no-side effects from rule ordering influence the execution of rules in the same layer. Also, the increments to the transformation by each rule of a layer in \cref{def:layer_step_semantics} are united using $\sqcup$ (see \cref{def:typed_graph_union_appendix}), which is an operation that is commutative by construction and thus renders the transformation result from each layer deterministic. 
\end{enumerate}
Given there are no other sources of non-determinism when building the execution of a transformation, every model transformation execution is confluent up to typed graph isomorphism.
\end{pf}

\begin{proposition}{Termination}

Every model transformation execution terminates.
\end{proposition}
\begin{pf}
Let us assume that there is a transformation execution which does not terminate. In order for this to happen there must exist a part in the construction of the execution of a transformation which induces an algorithm with an infinite amount of steps. We identify three moments when this can happen:
\begin{enumerate}
\item if definition~\ref{def:layer_step_semantics} of execution of a layer induces an infinite amount of steps. The only possibility for this to happen is if a layer has an infinite amount of transformation rules, which is a contradiction with definition~\ref{def:layer_transformation_appendix}.
\item if definition~\ref{def:transformation_step_semantics} of execution of a transformation induces an infinite amount of steps. Given layers are executed sequentially and no looping is allowed, the only possibility for this to happen is if the transformation has an infinite amount of layers, which contradicts definition~\ref{def:layer_transformation_appendix}.
\item if the result of the $match_{rl}(m_{in})$ function in definition~\ref{def:match_function} is an infinite set of match-apply graphs. The input-output model $m_{in}$ is by definition finite and the matching of each rule is independent from the execution of other rules in the same layer. As such, the number of subgraphs of $m_{in}$ isomorphic to $rl$'s matcher found during the execution of $rl$ is finite.
\end{enumerate}
Given there are no other constructs in the semantics of a transformation that can induce an infinite amount of steps, every model transformation execution terminates.
\end{pf}

### 2016-05-12 text/new_formal_background.tex
Add functions for typing in formalization.

@@
 
+\newcommand{\defineggprime}{Let $\langle V,E,(s,t),\tau,VT,ET\rangle = g$, and\\ $\langle V',E',(s',t'),\tau',VT',ET'\rangle = g'$, where $g, g' \in \textsc{Tg}$.}
+
+\newcommand{\ET}{\mathit{ET}}
+\newcommand{\VT}{\mathit{VT}}
+
 \subsection{Graph Structures and Functions}
@@
 \label{def:typed_graph}
-A typed graph is a 6-tuple $\langle V,E,(s,t),\tau, VT, ET\rangle$ where:
+A typed graph is a 6-tuple $\langle V,E,(s,t),\tau, \VT, \ET\rangle$ where:
 \begin{itemize}
@@
 \item $(s,t)$ is a pair of functions $s: E\rightarrow V$ and $t: E\rightarrow V$ that respectively provide the source and target vertices for each edge in the graph
-\item Function $\tau:V\cup E\rightarrow VT \cup ET$ is a typing function for the elements of $V$ and $E$, where $VT$ and $ET$ are disjoint finite sets of vertex and edge type identifiers and $\tau(v)\in VT$ if $v\in V$ and $\tau(e)\in ET$ if $e\in E$
+\item Function $\tau:V\cup E\rightarrow \VT \cup \ET$ is a typing function for the elements of $V$ and $E$, where $\VT$ and $\ET$ are disjoint finite sets of vertex and edge type identifiers and $\tau(v)\in \VT$ if $v\in V$ and $\tau(e)\in \ET$ if $e\in E$
 \item Edges $e\in E$ are noted $v\xrightarrow{e} v'$ if $s(e)=v$ and $t(e)=v'$, or simply e if the context is unambiguous
@@
 
-We also assume the existence of a function $\mathit{matchesOver}$ to handle polymorphism issues for $VT$ and $ET$.
-Let $\mathit{matchesOver}: \{VT \cup ET\} \times \{VT \cup ET\} \rightarrow \{\mathit{true}, \mathit{false}\}$ where:
+\subsubsection{Vertex and Edge Types}
+
+Note that the our verification technique is geared toward model-driven engineering. Therefore, the types of vertices and edges in our typed graphs will be drawn from a particular \textit{metamodel}.
+
+Sample metamodels for our running example can be seen in figures above \bentley{Fix.}.
+The particular metamodel for a typed graph will be indicated in the formalization when required, as (potentially) different metamodels will be present for the input and output graphs in DSLTrans rules.
+
+We assume the existence of a number of functions in our formalization to aid the treatment of typing:
+
 \begin{itemize}
+\item $\mathit{isAbstract}: \mathit{VT} \rightarrow \{\mathit{true}, \mathit{false}\}$
+\begin{itemize}
+\item Where $\mathit{isAbstract}(\VT)$ returns $\mathit{true}$ iff $\VT$ is denoted as abstract (not able to be instantiated) by the metamodel, else $\mathit{false}$
+\end{itemize}
+
+\item $\mathit{isIndirect}: \mathit{ET} \rightarrow \{\mathit{true}, \mathit{false}\}$
+\begin{itemize}
+\item Where $\mathit{isIndirect}(\ET)$ returns $\mathit{true}$ iff $\ET$ is denoted as an \textit{indirect edge}, else $\mathit{false}$.
+The \textit{indirect} classification allows matching over indirect paths between vertices.
+This will be further clarified when required for our constructs.
+\end{itemize}
+
+\item $\mathit{matchesOver}: \{\VT \cup \ET\} \times \{\VT \cup \ET\} \rightarrow \{\mathit{true}, \mathit{false}\}$
+
+\begin{itemize}
+\item The purpose of this function is to assist in handling polymorphism in the metamodel
 \item $\mathit{matchesOver}(T, T) \rightarrow \mathit{true}$
-\item $\mathit{matchesOver}(T, T') \rightarrow \mathit{true}$ if T' is a subclass of T in some defined partial ordering $\leq$
+\item $\mathit{matchesOver}(T, T') \rightarrow \mathit{true}$ iff T' is a subclass of T in some defined partial ordering $\leq$ given by the metamodel
 \item Otherwise, $\mathit{false}$
 \end{itemize}
+\end{itemize}
 
-\bentley{$\leq$ might need more explanation}
+For simplification purposes, we will not model the association cardinalities given by a metamodel in our notion of typed graph.
+Note that in fact require these cardinalities to be relaxed to perform our graph rewriting.
 
@@
 
-Let $\langle V,E,(s,t),\tau,VT,ET\rangle = g$, and\\ $\langle V',E',(s',t'),\tau',VT',ET'\rangle = g'$, where $g, g' \in \textsc{Tg}$.%where $VT$ and $ET'$ are disjoint sets, as well as $VT'$ and $ET$.
+\defineggprime
+
+%where $VT$ and $ET'$ are disjoint sets, as well as $VT'$ and $ET$.
 The typed graph union is the function $\sqcup :\textsc{Tg}\times \textsc{Tg}\rightarrow \textsc{Tg}$ defined as:
-
 \begin{multline*}
-\big\langle V,E,(s,t),\tau,VT,ET\big\rangle\;\sqcup\;\big\langle V',E',(s',t'),\tau',VT',ET'\big\rangle=\\\big\langle V\cup V', E\cup
-E',(s\cup s', t\cup t'), \tau\cup \tau', VT\cup VT', ET\cup ET'\big\rangle
+\big\langle V,E,(s,t),\tau,\VT,\ET\big\rangle\;\sqcup\;\big\langle V',E',(s',t'),\tau',\VT',\ET'\big\rangle=\\\big\langle V\cup V', E\cup
+E',(s\cup s', t\cup t'), \tau\cup \tau', \VT\cup \VT', \ET\cup ET'\big\rangle
 \end{multline*}
@@
 
-Note: as a reviewer helpfully pointed out, we require that s $\cup$ s', t $\cup$t', and $\tau \cup \tau$' coincide on common elements. However, this can be assumed w. l. o. g..
+Note: as a reviewer helpfully pointed out, we require that $s \cup s'$, $t \cup t'$, and $\tau \cup \tau'$ coincide on common elements. However, this can be assumed w. l. o. g..
 
@@
 \label{def:typed_graph_homomorphism}
-Let $\langle V,E,(s,t),\tau,VT,ET\rangle=g$ and $\langle V',E',(s',t'),\tau',VT',ET'\rangle=g'\in \textsc{Tg}$, where $g, g' \in \textsc{Tg}$.
+\defineggprime
+
 A typed graph homomorphism between $g$ and $g'$ is a function $f: f_v \cup f_e$ such that:
@@
 
+\bentley{Maybe put `inj' and `sub' in the homomorphism to reduce number of symbols}
  When an \emph{injective} typed graph homomorphism $f$ exists between $g$ and $g'$ we write $g \stackrel{f}{\vartriangleleft} g'$, or simply $g \vartriangleleft g'$ when the context is unambiguous.
@@
 
-\reviewer{For multigraphs (which may contain parallel edges with the same label) it is
-necessary to include edge images in the homomorphism (Def. 3). If you do not
-do that, then graphs with parallel edges are isomorphic to graphs with single
-edges (between the same nodes), so you might as well restrict to simple graphs.}
-
-\reviewer{
-Concretely: Right now the def. requires the existence of \textit{some} unfixed e' in the
-target graph, but in fact \textit{f} should be extended to E $\rightarrow$ E' so that e' = f (e) in this
-situation.
-}
-
+\subsubsection*{Typed Subgraph}
 We now define the useful notion of typed subgraph. As expected, a typed subgraph is simply a restriction of a typed graph to some of its vertices and edges.
@@
 \label{def:typedsubgraph}
-Let $\langle V,E,st,\tau,VT,ET\rangle=g,\langle V',E',st,\tau',VT',ET'\rangle=g'\in \textsc{Tg}$ be typed graphs.
+\defineggprime
+
 $g'$ is a typed subgraph of $g$, written $g'\sqsubseteq g$, iff $V'\subseteq V$, $E'\subseteq E$ and $\tau'=\tau |_{V'\cup E'}$.
@@
 
+\subsection*{Typed Graph Isomorphism}
 Two typed graphs are said to be isomorphic if they have exactly the same shape and related vertices and edges have the same type.
@@
 \label{def:typed_graph_isomorphism}
-Let $\langle V,E,st,\tau,VT,ET\rangle=g,\langle V',E',st',\tau',VT',ET'\rangle=g'\in \textsc{Tg}$ be typed graphs.
-We say that $g$ and $g'$ are isomorphic, written $g\cong g'$, if and only if there exists a bijective typed graph homomorphism $f:V\rightarrow V'$ such that $f^{-1}:V'\rightarrow V$ is a typed graph homomorphism.
+\defineggprime
+
+We say that $g$ and $g'$ are isomorphic, written $g\cong g'$, iff there exists a bijective typed graph homomorphism $f:V\rightarrow V'$ such that $f^{-1}:V'\rightarrow V$ is a typed graph homomorphism.
 \end{definition}
@@
 \paragraph{\textbf{Notation:}}
-In order to simplify our notation, when the context is unambiguous we will abbreviate a typed graph $\langle V,E,st,\\\tau,VT, ET\rangle$ as as 4-tuple $\langle V,E,st,\tau\rangle$.
+\bentley{Try to cut this as much as possible}
+In order to simplify our notation, when the context is unambiguous we will abbreviate a typed graph $\langle V,E,st,\\\tau,\VT, \ET\rangle$ as as 4-tuple $\langle V,E,st,\tau\rangle$.
 Also, given a typed graph $g\in \textsc{Tg}$, will use the notation $Components(g)$ to describe the set of strongly connected typed graphs in $g$.
@@
 
+\subsubsection*{Metamodel}
 We will start by introducing the notion of \emph{metamodel}, which in DSLTrans is used to type the input and output models of a DSLTrans transformation.
@@
 Note that here we have overridden the co-domain of the typing function in the original typed graph presented in \cref{def:typed_graph} in order to allow distinguishing between \emph{abstract} and \emph{concrete} classes, as well as between \emph{containment} and \emph{reference} edges in our metamodels.
-For simplification purposes, we do not model association cardinalities in our formal notion of metamodel as cardinalities are not strictly necessary in our development.
 

### 2016-05-12 text/new_formal_background.tex
Removed metamodel constructs, and worked on transformation rule.

@@
 
-For simplification purposes, we will not model the association cardinalities given by a metamodel in our notion of typed graph.
+For simplification purposes, we will not model the edge cardinalities given by a metamodel in our notion of typed graph.
 Note that in fact require these cardinalities to be relaxed to perform our graph rewriting.
@@
 
-\subsection{Metamodel and Model-Related Constructs}
-\label{sec:DSLTrans_formal}
+\subsubsection*{Transitive Closure}
 
-\subsubsection*{Metamodel}
-We will start by introducing the notion of \emph{metamodel}, which in DSLTrans is used to type the input and output models of a DSLTrans transformation.
-Two metamodels, the \emph{organization language} and \emph{gender language} are depicted in \cref{fig:squadmetamodel}.
+DSLTrans matching constructs allow for the matching of indirect links, where as described in Section~\ref{subsec:DSLTrans_constructs}, rules may match over indirect paths between elements.
+This transitive closure allow for the explicit creation of all edges implied by these indirect links to aid in matching.
 
-\begin{definition}{Metamodel\\}
-\label{def:metamodel}
-A metamodel is a 5-tuple $\langle V,E,st,\tau,\leq\rangle$ where $\langle V,E,st,\tau\rangle\in \textsc{Tg}$ is a typed graph, $(V,\leq)$ is a partial order and $\tau$ is a bijective typing function.
-Additionally we also have that: if $v\in V$ then $\tau(v)\in VT\times \{abstract,concrete\}$, where $VT$ is the set of vertex type names; if $e\in E$ then $\tau(e)\in ET\times \{containment,\\reference\}$, where $ET$ is a set of edge type names.
-The set of all metamodels is called $\textsc{Meta}$.
+\begin{definition}{Transitive Closure\\}
+\label{def:instance_closure}
+
+Let $g$ be a graph $\langle V,E,(s, t),\tau\rangle \in \textsc{Tg}$. Then the transitive closure $g^{*} = \langle V,E',(s, t),\tau\rangle \in \textsc{Tg}$, where:
+
+$E' = E \cup $ the transitive closure of $\big\{v\xrightarrow{e}v' | \mathit{isIndirect(e)} \big\}$
 \end{definition}
 
-\reviewer{Def. 6: Since the codomain of $\tau$ is apparently VT $\times$ {abstract, concrete} $\cup$ ET $\times$
-{containment, reference}, it is certainly never going to be a bijection. However,
-I think you actually only want injectiveness; what purpose is served by requiring
-it to be surjective?}
+Given a graph, its transitive closure includes, besides the original graph, all the edges belonging to the transitive closure of indirect links in that graph.
+Note that these transitive edges are given by the function $\mathit{isIndirect}$.
 
-\reviewer{As an aside, I think this is quite an awkward way to encode the notions of ab-
-stractness and containment: I think it would be much more elegant to define
-abstractness as a predicate over VT and containment as a predicate over ET. To
-go even one step further, it would in fact be more natural (and more standard)
-to equate the notions of type graph and metamodel altogether, as they certainly
-serve the same conceptual purpose.
-}
+In the definitions that follow we will use the $*$ notation, as in \cref{def:instance_closure}, to denote the transitive closure of our structures.
 
-\reviewer{Def. 6 and many other places A word in math mode should always
-be in $\backslash$mathit to avoid ugly spacing as in $re f erence$.}
-
-A formal metamodel is a particular kind of typed graph where vertices represent classes and edges represent relations between those classes.
-A typed graph representing a metamodel has two special characteristics: on the one hand, the typing function for vertices and edges is bijective.
-This means that each type occurs only once in the metamodel, as is to be expected.
-On the other hand, a metamodel is equipped with a partial order between vertices.
-This partial order is used to model inheritance at the level of the metamodel's classes.
-Note that here we have overridden the co-domain of the typing function in the original typed graph presented in \cref{def:typed_graph} in order to allow distinguishing between \emph{abstract} and \emph{concrete} classes, as well as between \emph{containment} and \emph{reference} edges in our metamodels.
-
-\begin{definition}{Expanded Metamodel\\}
-\label{def:expanded_metamodel}
-Let $mm = \langle V,E,st,\tau,\leq\rangle\in \textsc{Meta}$ be a metamodel.
-The expansion of $mm$, noted $mm^{\star}$, is a typed graph $\langle V',E',st',\tau'\rangle\in \textsc{Tg}$ built as follows:
-\begin{itemize}
-  \item $V' = V\setminus \{v\in V\,|\,\tau(v)=(\cdot\footnote{In our mathematical development we use a `dot' notation to represent that we do not care about the value of a particular variable in a given context.},abstract)\}$;
-  \item $v_1\xrightarrow{e}v_2\in E'$ if $v_1\xrightarrow{e}v_2\in E$ and $\tau(v_1)=(\cdot,concrete)$ and $\tau(v_2)=(\cdot,concrete)$;
-  \item if $v_1\xrightarrow{e}v_2\in E$ we have that $v'_1\xrightarrow{e'}v'_2\in E'$, where $v'_1\leq v_1$, $v'_2\leq v_2$ and $\tau'(e') = \tau(e)$;
-  \item for all $v\in V'$ and $e\in E'$ we have that $\tau'(v)=\tau(v)$ and that $\tau'(e)=\tau(e)$.
-%   \item $(v,(tv,concrete))\in \tau'$ if $(v,(tv,concrete))\in \tau$ (where $tv$ is a vertex type);
-%   \item $(v_1\xrightarrow{e}v_2,(te,concrete))\in \tau'$ if $(v_1\xrightarrow{e}v_2,(te,concrete))\in \tau$ ($te$ ia an edge type).
-\end{itemize}
-%The set of all expanded metamodels in called $META^{\star}$.
-\end{definition}
-
-\reviewer{Def. 7: is an expanded metamodel a metamodel? Then you need to allow non-
-injective $\tau$, at least on edges, as the requirement $\tau$(e) = $\tau$(e') in the third bullet
-will certainly cause non-inejectiveness.}
-
-An expanded metamodel is an auxiliary construct where all the relations between types of a metamodel are made explicit, rather than remaining implicit in the specialization hierarchy.
-It is built by adding to the original metamodel typed graph a relation of type $t$ between two classes of the metamodel, whenever those classes specialize two classes that are also related by a relation of type $t$.
-Abstract classes and their relations do not carry over to the expanded metamodel.
-Expanded metamodels will be used in the subsequent text to facilitate formal treatment of any structure involving polymorphism.
-
-% Definition~\ref{def:metamodel} formally states that a metamodel corresponds to a graph of typed elements where only one element for each type is represented.
-This definition is then used to state definition~\ref{def:pattern}, where a typed graph is combined with its metamodel.
-
-\begin{definition}{Metamodel Instance\\}
-\label{def:metamodelinstance}
-
-An instance of a metamodel $mm = \langle V',E',st',\tau',\leq\rangle\in \textsc{Meta}$ is a typed graph $\langle V,E,st,\tau\rangle\in \textsc{Tg}$, where the co-domain of $\tau$ equals the co-domain of $\tau'$.
-Also, there is a typed graph homomorphism $f:V\rightarrow V'$ from $\langle V,E,st,\tau\rangle$ to the expanded metamodel $mm^{\star}$ and the graph $\big\langle V,\{e\in E\,|\,\tau(e)=(\cdot,containment)\}\big\rangle$ is acyclic.
-The set of all instances for a metamodel $mm$ is called $\textsc{Instance}^{mm}$.
-
-\end{definition}
-
-A metamodel instance is a useful intermediate formal notion that lies between metamodel and model.
-The injective typed graph homomorphism between a metamodel instance and metamodel models multiple ``instances'' of objects and relations being typed by one single class or relation of the metamodel.
-Metamodel instances do not allow cyclic containment relations, as enforced by EMF.
-\reviewer{Do you mean “in analogy to EMF”?}
-
-% It requires the acyclicity constraints as that is required through models, transformation rules, path conditions and properties.
-It can be used to define transformation rules and can be further refined to define models.
-
-\reviewer{Def. 9 and many thereafter incrementally define new concepts on top of graphs.
-Most of these definitions are rather hard to read, for several reasons:}
-
-\begin{itemize}
-\item It is common to use an unprimed version for the original object and a
-primed version for the constructed object, whereas you turn this convention around
-\item Sloppiness: in more than 50 percent of all cases the $\subseteq$ should be $\supseteq$ or vice versa
-\begin{itemize}
-\item In this particular instance, you really mean $\tau' \supseteq \tau$ and $E = E' \cup E_c^∗$ .
-\end{itemize}
-
-\item It is really more readable to define the construction as a construction, so:
-\begin{itemize}
-\item  $E = E'\cup E^{\star}_{C}$
-\item $s = s' \cup {(e_1 \dots e_n , s(e_1)) | e_1 \dots e_n \in E^{\star}_{C} }$
-\item $t = t' \cup {(e_1 \dots e_n ,t(e_n )) | e_1 \dots e_n \in E^{\star}_{C} }$
-\item $\tau = \tau' \cup {(e, indirect) | e \in E^{\star}_{C} }$
-\end{itemize}
-\begin{itemize}
-\item Note that this also takes care of s and t, which you have forgotten here
-and in all other similar definitions.)
-\item The set $E^{\star}_{C}$should itself also be more carefully defined, as $E^{\star}_{C} = {e_1 \dots e_n \in E^∗' | \forall1 \leq i < n.t(e_i ) = s(e_{i+1} )}$
-\end{itemize}
-\end{itemize}
-
-\begin{definition}{Containment Transitive Closure\\}
-\label{def:instance_closure}
-
-The containment transitive closure of a metamodel instance $\langle V',E',st',\tau'\rangle\in \textsc{Instance}^{mm}$ is a typed graph $\langle V,E,st,\tau\rangle$ where we have that $V=V'$, $\tau' \supseteq \tau$ and $\tau$'s co-domain is the union of the co-domain of $\tau'$ and the set $\{indirect\}$.
-We also have that $E' = E\cup E^*_c$ where $E^*_c$ is the transitive closure of the set $\big\{v\xrightarrow{e}v'\,|\,\tau(v\xrightarrow{e}v') = (\cdot,containment)\big\}$ and if $e\in E\setminus E'$ then $\tau(e) = indirect$.
-We denote $mi^{*}$ the containment closure of a metamodel instance $mi\in \textsc{Instance}^{mm}$.
-
-\end{definition}
-
-Given a metamodel instance, its containment transitive closure includes, besides the original graph, all the edges belonging to the transitive closure of containment links in that metamodel instance.
-The transitive edges are typed as \emph{indirect}.
-In the definitions that follow we will use the $*$ notation, as in \cref{def:instance_closure}, to denote the containment transitive closure of structures that directly or indirectly include metamodel instances.
-For example, $tg^{*}$ would represent the containment transitive closure of typed graph $tg$ wherever containment edges are found in the graph.
-Note that the $*$ notation is different from the $\star$ notation, introduced in \cref{def:expanded_metamodel} for an expanded metamodel.
-\reviewer{The difference between these stars is too small to be noticeable.}
-
-% It requires the acyclicity constraints as that is required through models, transformation rules, path conditions and properties.
-It can be used to define transformation rules and can be further refined to define models.
-
-\begin{definition}{Model\\}
-\label{def:model}
-
-A model of a  metamodel $mm = \langle V',E',st',\tau',\leq\rangle\in \textsc{Meta}$ is a metamodel instance $\langle V,E,st,\tau\rangle\in \textsc{Instance}^{mm}$, such that: there exists an injective typed graph homomorphism $f:V\rightarrow V'$ from $\langle V,E,st,\tau\rangle$ to metamodel $mm^{\star}$ where, if $b\in V$ and there exists an edge $a'\xrightarrow{e'} f(b)\in E'$ where $\tau(e') = (t,containment)$, then we also have that $a\xrightarrow{e} b\in E$, $f(a) = a'$ and $\tau(e)=t$.
-The set of all models for a metamodel $mm$ is called $\textsc{Model}^{mm}$.
-
-\end{definition}
-
-\reviewer{Def. 10: Here you certainly should not require $\tau$ to be injective. The requirement
-on containment relations is mathematical nonsense: if $e' \in E'$ has target b then
-$b \in V'$ and so f (b) does not exist, and most certainly is not the source of any edge
-in E. In fact, I really have no idea what you want to say here; the explanation
-below the definition does not clarify matters at all.}
-
-A model, as per \cref{def:model}, is a metamodel instance where all the containment relations are respected.
-This means that if an object that is contained (as per the metamodel) exists in the model, then the model will also contain an instance of that containment relation together with an object that contains it.
-Two models can be observed in \cref{fig:transformationexample}, which are respectively instances of the \emph{Organization language} and \emph{Gender language} found in \cref{fig:squadmetamodel}.\\
-Note that the containment constraint does not necessarily lead to infinite models in the case of containment relations with the same source and target classes.
-In fact, if the cardinality of the target class is allowed to be zero, then it is not necessary that the containment relation is instantiated.
-For example, this is the case for the containment relation \emph{supervise} in the metamodel of~\cref{fig:OrganizationLanguage}.
-
-\reviewer{“the cardinality of the target class”?? Are you talking about edge multiplicities here (which you have chosen not to include)?}
-
-\begin{definition}{Input-Output Model\\}
-\label{def:input_output_model}
-An input-output model is a 6-tuple $\big\langle V,E,(s,t),\tau,Input,\\Output\big\rangle$, where: $Input=\langle V',E',st',\tau'\rangle\in \textsc{Instance}^{sr}$ is a model; $Output=\langle V'',E'',st'',\tau''\rangle\in \textsc{Instance}^{tg}$ is a metamodel instance; $Input$ and $Output$ are disjoint.
-Additionally we have that $V=V'\cup V''$, $E\subseteq E'\cup E''$ and $\tau\subseteq \tau'\cup \tau''$, where the co-domain of $\tau$ is the union of the co-domains of $\tau'$ and $\tau''$ and the set $\{trace\}$.
-An edge $e \in E\setminus E'\cup E''$ is called a \emph{traceability link} and is such that $s(e)\in V''$, $t(e)\in V'$ and $\tau(e)=trace$.
-The set of all match-apply patterns for a source metamodel $sr$ and a target metamodel $tg$ is called $\textsc{Iom}^{sr}_{tg}$.
-\end{definition}
-
-\reviewer{Why is Input a model and Output a metamodel instance? Note that
-$\backslash$ binds stronger than $\cup$ in $E \backslash E' \cup E$ ′′ . “match-apply patterns” → “input-output
-patterns”}
-%Change if needed
-% \levi{A match-apply model is the union of an input and output models, plus potential traceability links between those models.
-Traceability links are now typed using the typing function $\tau$.
-A traceability link it typed with type ``trace''.
-This makes the definitions shorter and more compact, which was mentioned by one of the reviewers of the paper.
-However, the fact that a match-apply model is now made up of models makes it that the input and the output of the transformation are always well-formed, meaning that the containment relation is always respected.
-It may be that a transformation does not always produce a model respecting the containment relation, which is a problem.
-This can be solved however if we only consider that well-formed transformations actually have semantics.}
-
-An input-output model is an object we will use when defining the semantics of a DSLTrans model transformation in \cref{sec:DSLTrans_formal}.
-It is composed of two metamodel instances, one called the \emph{input} and the other one the \emph{output}.
-An input-output model allows the representation of intermediate operational states during the execution of a model transformation.
-It may include a particular type of edges called \emph{traceability links}, for keeping a history of which elements in the output model originated from which elements in the input model.
-
-\begin{definition}{Metamodel Pattern and Indirect Metamodel Pattern\\}
-\label{def:metamodel_pattern}
-
-A pattern of a metamodel $mm\in \textsc{Meta}$ is an instance of $mm$.
-Given a metamodel pattern $\langle V',E',st',\tau'\rangle\in \textsc{Instance}^{mm}$ we have that $\langle V,E,st,\tau\rangle$ is an \emph{indirect} pattern if $V=V'$, $E'\supseteq E$ and the co-domain of $\tau$ is the union of the co-domains of $\tau'$ and the set $\{indirect\}$.
-Also, if $e\in E\setminus E'$, then we have that $\tau(e)=indirect$.
-Given a metamodel $mm$, the set of all metamodel patterns for $mm$ is called $\textsc{Pattern}^{mm}$.
-The set of all indirect metamodel patterns for $mm$ is called $\textsc{Ipattern}^{mm}$.
-\end{definition}
-
-Metamodel patterns are introduced in \cref{def:metamodel_pattern} as an intermediate notion, formally equal to metamodel instances.
-An \emph{indirect} metamodel pattern is a metamodel pattern that includes edges typed as \emph{indirect}.
-Both structures will be used as building blocks in the construction of transformation-related structures in the upcoming text.
-
-\subsection{Syntactic Transformation Constructs}
-\label{sec:DSLTrans_syntax}
+\subsection{DSLTrans Constructs}
+\label{sec:DSLTrans_formal}
 
@@
 
-\begin{definition}{Transformation Rule\\}
-\label{def:transformation_rule}
+As discussed in Section~\ref{subsec:DSLTrans_syntax}, DSLTrans transformations are composed of rules arranged in layers.
 
-A transformation rule is a 6-tuple $\big\langle V,E,(s,t),\tau, Match,\\Apply\big\rangle$, where: $Match=\langle V',E',st',\tau'\rangle\in \textsc{Ipattern}^{sr}$ such that: $Match\neq \epsilon$\footnote{We use the simplified $\epsilon$ notation to denote empty n-tuples structures.} is an non-empty indirect metamodel pattern; $Apply=\langle V'',E'',st'',\tau''\rangle\in \textsc{Pattern}^{tg}$ such that $Apply\neq\epsilon$ is a metamodel pattern; $Match$ and $Apply$ are disjoint.
-We also have that $V=V'\cup V''$, $E\subseteq E'\cup E''$ and $\tau\subseteq \tau'\cup \tau''$, where the co-domain of $\tau$ is the union of the co-domains of $\tau'$ and $\tau''$ and the set $\{trace\}$.
-An edge $e \in E\setminus E'\cup E''$ is called a \emph{backward link} and is such that $s(e)\in V''$, $t(e)\in V'$ and $\tau(e)=trace$.
-We additionally impose that there always exists a $v_1\in V''$ in the Apply part of the rule such that $\nexists e\,.\, v_1\xrightarrow{e} v_2$ and $\tau(e)=trace$, or that $E''$ is not empty.
-The set of all transformation rules for a source metamodel $sr$ and a target metamodel $tg$ is called $\textsc{Rule}^{sr}_{tg}$.
-
-\end{definition}
-
-\reviewer{e in $\tau(e) = \mathit{trace}$ is still bound by $\nexists$, so “and” should be $\land$}
+\subsubsection*{DSLTrans Transformation Rule}
 
@@
 A formal transformation rule includes a non-empty match pattern and a non-empty apply pattern (also known in the model transformation literature as a rule's \emph{left hand side} and \emph{right hand side}).
+
+A match pattern can include indirect links that are used to transitively match containment relations in a model.
 The apply pattern of a rule always contains at least one apply element that is not connected to a backward link or an edge, meaning in practice that a rule will always produce something and not only match.
-A match pattern can include indirect links that are used to transitively match containment relations in a model.
 An apply pattern does not include indirect links as it is used only for the construction of parts of instances of a metamodel.
-A transformation rule includes backward links, as informally introduced in \cref{subsec:DSLTrans_constructs}.
-Backward links are formally typed as \emph{trace}.%This eases modeling matching rules on models  by the use of typed graph homomorphisms.
+A transformation rule also includes backward links, as informally introduced in \cref{subsec:DSLTrans_constructs}.
+
+\begin{definition}{DSLTrans Transformation Rule\\}
+\label{def:transformation_rule}
+
+A DSLTrans transformation rule is a three-tuple $\big\langle \mathit{Match}, \mathit{Apply}, \mathit{backward}\big\rangle$, where:
+
+\begin{itemize}
+\item $\mathit{Match}, \mathit{Apply} \in \textsc{Tg}$
+\item $\mathit{Match}, \mathit{Apply} $ are non-empty and disjoint
+\item $\mathit{backward} = \{E_{back}, (s_{back}, t_{back})\}$
+\end{itemize}
+
+Note that when we require an element of the $\mathit{Match}$ or $\mathit{Apply}$ graphs, such as the vertices, we will index the required element.
+For example, the vertices for the \textit{Match} graph will be $V_{\textit{Match}}$.
+
+$\mathit{backward}$ contains backward links, which as discussed in Section~\ref{subsec:DSLTrans_constructs} allow for rules to define dependencies between elements.
+\begin{itemize}
+\item $E_{back}$ contains the links
+\item $(s_{back}, t_{back})$ is a pair of functions $s_{back}: E_{back}\rightarrow V_{\textit{Apply}}$ and $t_{back}: E_{back}\rightarrow V_{\textit{Match}}$ that respectively provide the source and target vertices for each backward edge
+\begin{itemize}
+\item Note the source of backward links is a vertex in $V_{\textit{Match}}$ while the target is a vertex in $V_{\textit{Apply}}$
+\end{itemize}
+\end{itemize}
+
+\end{definition}
+
+We additionally impose that for the well-formedness of a DSLTrans rule, there always exists an element or edge to be created in the \textit{Apply} graph to be created.
+That is, either there are edges to be created in the \textit{Apply} graph, or there is a vertex in the \textit{Apply} graph that is not the source of a backward link.
+
+\begin{itemize}
+\item $E_{\textit{Apply}} \neq \emptyset \lor$
+\item $ \exists v \in V_{\textit{Apply}} | \big\{\forall e \in E_{back}: s(e) \neq v \big\}$
+\end{itemize}
+
+a $v_1\in V''$ in the Apply part of the rule such that $\nexists e\,.\, v_1\xrightarrow{e} v_2$ and $\tau(e)=trace$, or that $E''$ is not empty.
 
@@
 For example, we will write $V(rl)$ to designate the $V$ component of $rl$ or $Apply(rl)$ to designate $rl$'s $Apply$ component.
+
+\begin{definition}{Input-Output Model\\}
+\label{def:input_output_model}
+An input-output model is a 6-tuple $\big\langle V,E,(s,t),\tau,Input,\\Output\big\rangle$, where:
+\begin{itemize}
+\item $Input=\langle V',E',st',\tau'\rangle\in \textsc{Instance}^{sr}$ is a model
+\item $Output=\langle V'',E'',st'',\tau''\rangle\in \textsc{Instance}^{tg}$ is a metamodel instance
+\item $Input$ and $Output$ are disjoint
+\end{itemize}
+
+Additionally we have that $V=V'\cup V''$, $E\subseteq E'\cup E''$ and $\tau\subseteq \tau'\cup \tau''$, where the co-domain of $\tau$ is the union of the co-domains of $\tau'$ and $\tau''$ and the set $\{trace\}$.
+
+An edge $e \in E\setminus E'\cup E''$ is called a \emph{traceability link} and is such that $s(e)\in V''$, $t(e)\in V'$ and $\tau(e)=trace$.
+
+ set of all match-apply patterns for a source metamodel $sr$ and a target metamodel $tg$ is called $\textsc{Iom}^{sr}_{tg}$.
+\end{definition}
+
+\reviewer{Why is Input a model and Output a metamodel instance? Note that
+$\backslash$ binds stronger than $\cup$ in $E \backslash E' \cup E$ ′′ . “match-apply patterns” → “input-output
+patterns”}
+%Change if needed
+% \levi{A match-apply model is the union of an input and output models, plus potential traceability links between those models.
+Traceability links are now typed using the typing function $\tau$.
+A traceability link it typed with type ``trace''.
+This makes the definitions shorter and more compact, which was mentioned by one of the reviewers of the paper.
+However, the fact that a match-apply model is now made up of models makes it that the input and the output of the transformation are always well-formed, meaning that the containment relation is always respected.
+It may be that a transformation does not always produce a model respecting the containment relation, which is a problem.
+This can be solved however if we only consider that well-formed transformations actually have semantics.}
+
+An input-output model is an object we will use when defining the semantics of a DSLTrans model transformation in \cref{sec:DSLTrans_formal}.
+It is composed of two metamodel instances, one called the \emph{input} and the other one the \emph{output}.
+An input-output model allows the representation of intermediate operational states during the execution of a model transformation.
+It may include a particular type of edges called \emph{traceability links}, for keeping a history of which elements in the output model originated from which elements in the input model.
 

### 2016-05-16 text/new_formal_background.tex
Initial version of disambiguated matching in formal notation.

@@
 
+\subsubsection*{Typed Subgraph}
+We now define the useful notion of typed subgraph. As expected, a typed subgraph is simply a restriction of a typed graph to some of its vertices and edges.
+
+\begin{definition}{Typed Subgraph\\}
+\label{def:typedsubgraph}
+\defineggprime
+
+$g'$ is a typed subgraph of $g$, written $g'\sqsubseteq g$, iff $V'\subseteq V$, $E'\subseteq E$ and $\tau'=\tau |_{V'\cup E'}$.
+
+\end{definition}
+
+Given a typed graph $g\in \textsc{Tg}$, will use the notation $Components(g)$ to describe the set of strongly connected typed graphs in $g$.
+Also, we will use the notation $g|_{t}$ to refer to the restriction of graph $g$ to its subgraph containing only edges of type $t$.
+
 \subsubsection*{Typed Graph Union}
@@
 \begin{itemize}
-\item $f_v(v_1), f_v(v_2) \in V'$
-\item $f_e(e) \in E'$
-\item $s'(f_e(e)) = f_v(v_1)$, $t'(f_e(e)) = f_v(v_2)$
-\item $\mathit{matchesOver}(\tau(v_1), \tau(f_v(v_1)) \land \\\mathit{matchesOver}(\tau(v_2), \tau(f_v(v_2)) \land\\ \mathit{matchesOver}(\tau(e), \tau(f_e(e))$
+\item Let $ v_1' = f_v(v_1), v_2' = f_v(v_2), e' = f_e(e)$
+\item $v_1', v_2' \in V', e' \in E'$
+\item $s'(e') = v_1'$, $t'(e') = v_2'$
+\item $\mathit{matchesOver}(\tau(v_1), \tau(v_1') \land \\\mathit{matchesOver}(\tau(v_2), \tau(v_2') \land\\ \mathit{matchesOver}(\tau(e), \tau(e')$
 \end{itemize}
@@
 
-\subsubsection*{Typed Subgraph}
-We now define the useful notion of typed subgraph. As expected, a typed subgraph is simply a restriction of a typed graph to some of its vertices and edges.
+\subsubsection*{Typed Graph Edge Homomorphism}
 
-\begin{definition}{Typed Subgraph\\}
-\label{def:typedsubgraph}
+Our technique also requires a form of graph homomorphism that primarily focuses on matching the edges of the graphs.
+This homomorphism will be an injective match regarding the edges, but note that a particular vertex in the source graph may match onto multiple vertices in the target graph.
+\bentley{What's the term for this?}
+
+\begin{definition}{Typed Graph Edge Homomorphism\\}
+\label{def:typed_graph_edge_homomorphism}
 \defineggprime
 
-$g'$ is a typed subgraph of $g$, written $g'\sqsubseteq g$, iff $V'\subseteq V$, $E'\subseteq E$ and $\tau'=\tau |_{V'\cup E'}$.
+A typed graph edge homomorphism between $g$ and $g'$ is a function $h: h_v \cup h_e$ such that:
+\begin{itemize}
+\item $h_v: V'\rightarrow V$
+\begin{itemize}
+\item Note that $h_v$ is an function from a vertex in the target graph onto a vertex in the source graph
+\end{itemize}
+\item $h_e: E\rightarrow E'$
+\item $\forall v_1 \xrightarrow{e} v_2\in E$:
+\begin{itemize}
+\item Let $e' = h_e(e), v_1' = s'(e'), v_2' = t'(e')$
+\item $v_1', v_2' \in V', e' \in E'$
+\item $v_1 = h_v(v_1'), v_2 = h_v(v_2')$
+\item $\mathit{matchesOver}(\tau(v_1), \tau(v_1') \land \\\mathit{matchesOver}(\tau(v_2), \tau(v_2') \land\\ \mathit{matchesOver}(\tau(e), \tau(e')$
+\end{itemize}
 
+\end{itemize}  
 \end{definition}
+
+Again, we wish to highlight the fact that this homomorphism is different than Definition~\ref{def:typed_graph_homomorphism} in the vertex matching function.
+A vertex in the target graph matches onto a vertex in the source graph, with multiple target vertices matching onto the same source vertex.
+\bentley{Create diagram to illustrate}
+
+This unusual homomorphism allows our technique to `split' our source graph over multiple locations in the target graph.
+This is critical, as our target graphs are constructed in a way that allows for vertex duplication.
+That is, two vertices of a particular type in the target graph may only represent one underlying element, as discussed in Section~\ref{sec:abstraction_relation}.
+Constructing this morphism allows our prover implementation to avoid explicitly creating `disambiguated' graphs where these duplications are resolved.
 
@@
 \end{definition}
-
-\paragraph{\textbf{Notation:}}
-\bentley{Try to cut this as much as possible}
-In order to simplify our notation, when the context is unambiguous we will abbreviate a typed graph $\langle V,E,st,\\\tau,\VT, \ET\rangle$ as as 4-tuple $\langle V,E,st,\tau\rangle$.
-Also, given a typed graph $g\in \textsc{Tg}$, will use the notation $Components(g)$ to describe the set of strongly connected typed graphs in $g$.
-Finally, we will use the notation $g|_{t}$ to refer to the restriction of graph $g$ to its subgraph containing only edges of type $t$.
 

### 2016-05-16 text/new_formal_background.tex
Formalize a rule matcher.

@@
 \newcommand{\VT}{\mathit{VT}}
+
+\bentley{Make sure that terminology is consistent between elements and vertices}
 
@@
 
-A transformation rule is the elemental block of a model transformation.
-Several transformation rules can be observed in the Police Station transformation in \cref{fig:dsltransformation}.
-A formal transformation rule includes a non-empty match pattern and a non-empty apply pattern (also known in the model transformation literature as a rule's \emph{left hand side} and \emph{right hand side}).
+A transformation rule is the elemental block of a DSLTrans transformation. Several transformation rules can be observed in the Police Station transformation in \cref{fig:dsltransformation}.
 
+A transformation rule includes a non-empty match pattern and a non-empty apply pattern.
+This is also known in the model transformation literature as a rule's \emph{left hand side} and \emph{right hand side}.
 A match pattern can include indirect links that are used to transitively match containment relations in a model.
@@
 An apply pattern does not include indirect links as it is used only for the construction of parts of instances of a metamodel.
+
 A transformation rule also includes backward links, as informally introduced in \cref{subsec:DSLTrans_constructs}.
+As described in that section, backward links define dependencies between rules.
+\bentley{Expand}
 
@@
 
-A DSLTrans transformation rule is a three-tuple $\big\langle \mathit{Match}, \mathit{Apply}, \mathit{backward}\big\rangle$, where:
+A DSLTrans transformation rule is a three-tuple $\big\langle \mathit{Match}$, $\mathit{Apply}$, $\mathit{backward}\big\rangle$, where:
 
@@
 \item $\mathit{backward} = \{E_{back}, (s_{back}, t_{back})\}$
+\begin{itemize}
+\item $E_{back} \cap E_{Match} = \emptyset, E_{back} \cap E_{Apply} = \emptyset$
+\end{itemize}
 \end{itemize}
@@
 
-$\mathit{backward}$ contains backward links, which as discussed in Section~\ref{subsec:DSLTrans_constructs} allow for rules to define dependencies between elements.
 \begin{itemize}
-\item $E_{back}$ contains the links
-\item $(s_{back}, t_{back})$ is a pair of functions $s_{back}: E_{back}\rightarrow V_{\textit{Apply}}$ and $t_{back}: E_{back}\rightarrow V_{\textit{Match}}$ that respectively provide the source and target vertices for each backward edge
+\item $E_{back}$ contains the backward links
+\item $(s_{back}, t_{back})$ is a pair of functions $s_{back}: E_{back}\rightarrow V_{\textit{Apply}}$ and $t_{back}: E_{back}\rightarrow V_{\textit{Match}}$ that respectively provide the source and target vertices for each backward link
 \begin{itemize}
@@
 \item $E_{\textit{Apply}} \neq \emptyset \lor$
-\item $ \exists v \in V_{\textit{Apply}} | \big\{\forall e \in E_{back}: s(e) \neq v \big\}$
+\item $ \exists v \in V_{\textit{Apply}} | \big\{\forall e \in E_{back}: s_{back}(e) \neq v \big\}$
 \end{itemize}
 
-a $v_1\in V''$ in the Apply part of the rule such that $\nexists e\,.\, v_1\xrightarrow{e} v_2$ and $\tau(e)=trace$, or that $E''$ is not empty.
+\subsubsection*{Matcher of a Transformation Rule}
+
+To create the actual matcher construct that will be matched during a rule's execution, the rule's \textit{Match} graph is combined with backward links, as well as any \textit{Apply} vertices connected to the backward links.
 
@@
 
-Let $rl = \big\langle V,E,st,\tau, Match,Apply\big\rangle$ be a transformation rule where $Match =\langle V_m, E_m, st_m, \tau_m\rangle$.
-We define $rl$'s matcher, noted $\Vert rl\Vert$, as the transformation rule $\big\langle V',E',st',\tau',Match,\\Apply'\big\rangle\sqsubseteq rl$ where $v_1\xrightarrow{e} v_2\in E'$ if and only if $v_1,v_2\in Match$ or $\tau(e)=trace$ and $V'=V_m\cup \big\{v_1\,|\,v_1\xrightarrow{e} v_2\in E \land \tau(e)=trace\big\}$.
-\reviewer{Definition 14: where does Apply' come from? Should v1 be 
-  an element of Apply?}
+Let the transformation rule $r = \big\langle \mathit{Match}$, $\mathit{Apply}$, $\mathit{backward}\big\rangle$.
+
+We define $r$'s matcher, noted $\Vert r \Vert$, to be a typed graph six-tuple $\langle V,E,(s,t),\tau, \VT, \ET\rangle$, where:
+\begin{itemize}
+\item $V = V_{Match} \cup \big\{v \in V_{Apply} | \{\exists e \in E_{back} | s_{back}(e) = v\}\big\}$
+\item $E = E_{Match} \cup E_{back}$
+\item $s = s_{Match} \cup s_{back}$, $t = t_{Match} \cup t_{back}$
+\item $\tau = \tau_{Match} \cup \tau_{Apply}$
+\item $\VT = \VT_{Match} \cup \VT_{Apply}$
+\item $\ET = \ET_{Match} \cup \ET_{Apply}$
+\end{itemize}
+
 \end{definition}
 
-\reviewer{Def. 14: According to your own criterion, $||rl||$ is not a rule as it does not create
-either a node or an edge.}
-
-\Cref{def:back_match_transformation_rule} introduces the notion of matcher for a transformation rule which consists solely of the match pattern of a rule and its backward links, if any.
-The matcher of a rule constitutes the complete pattern that a DSLTrans rule attempts to match over a input-output model during rule execution.
-Traceability links between input and output model elements generated during transformation execution are matched by transformation rules' backward links, as informally explained in \cref{subsec:DSLTrans_constructs}.
-
-\begin{definition}{Expanded Transformation Rule\\}
-\label{def:transformation_rule_expansion}
-
-Let $rl = \big\langle V,E,st,\tau,Match,Apply\rangle \in \textsc{Rule}^{sr}_{tg}$ be a transformation rule where $Match = \langle V',E',st',\tau'\rangle$ and also we have that $sr =\langle V'',E'',st'',\tau'',\leq \rangle$.
-The expansion of $rl$, noted $rl^{\star}$ is a set of transformation rules built as follows:
-\begin{itemize}
-  \item $rl\in rl^{\star}$;
-  \item $\big\langle V,E,st,\tau',Match,Apply\rangle\in rl^{\star}$ iff for all $v\in V'$ we have that $\tau'(v)\leq \tau(v)$.
-\end{itemize}
-\end{definition}
-
-\reviewer{The first bullet is a special case of the second, so can be removed.}
-
-The expansion of a transformation rule is a set of transformation rules.
-Each rule in that set includes a possible replacement of each of the classes in the match part of the original rule by one of its subtypes.
-Expanded transformation rules will be important to ensure that polymorphism is correctly handled in the developments that follow.
-
-%\levi{A transformation rule expansion simply takes a rule and builds all possible rules where a rule element is replaced by one of its subtypes.}
+\subsubsection*{DSLTrans Transformation Layer}
 

### 2016-05-17 text/new_formal_background.tex
Add NACs to rules. Shouldn't break everything too much...

@@
 It is meant as a formal reference for the upcoming formal development.
-This section can be safely skipped or skimmed by the reader, who can return to these definitions punctually to understand the detailed formal underpinning of our approach.
+%This section can be safely skipped or skimmed by the reader, who can return to these definitions punctually to understand the detailed formal underpinning of our approach.
 
@@
 \item Edges $e\in E$ are noted $v\xrightarrow{e} v'$ if $s(e)=v$ and $t(e)=v'$, or simply e if the context is unambiguous
-\item The set of all typed graphs is called $\textsc{Tg}$. 
+\item The set of all typed graphs is called $\textsc{Tg}$
+\item We define the empty graph to be a graph with $V = \emptyset$, $E = \emptyset$, and the other elements to be empty functions
 \end{itemize}
@@
 For simplification purposes, we will not model the edge cardinalities given by a metamodel in our notion of typed graph.
-Note that in fact require these cardinalities to be relaxed to perform our graph rewriting.
+Note that in fact we require these cardinalities to be relaxed to perform our graph rewriting.
 
@@
 
-\defineggprime
-
-%where $VT$ and $ET'$ are disjoint sets, as well as $VT'$ and $ET$.
 The typed graph union is the function $\sqcup :\textsc{Tg}\times \textsc{Tg}\rightarrow \textsc{Tg}$ defined as:
 \begin{multline*}
-\big\langle V,E,(s,t),\tau,\VT,\ET\big\rangle\;\sqcup\;\big\langle V',E',(s',t'),\tau',\VT',\ET'\big\rangle=\\\big\langle V\cup V', E\cup
-E',(s\cup s', t\cup t'), \tau\cup \tau', \VT\cup \VT', \ET\cup ET'\big\rangle
+\big\langle V,E,(s,t),\tau,\VT,\ET\big\rangle\;\sqcup\;\big\langle V',E',(s',t'),\tau',\VT',\ET'\big\rangle=\\
+\big\langle V\cup V', E\cup E',(s\cup s', t\cup t'), \tau\cup \tau', \VT\cup \VT', \ET\cup ET'\big\rangle
 \end{multline*}
@@
 Note: as a reviewer helpfully pointed out, we require that $s \cup s'$, $t \cup t'$, and $\tau \cup \tau'$ coincide on common elements. However, this can be assumed w. l. o. g..
+
+\bentley{Do we need typed graph intersection? Or to mention what it means for two graphs to be disjoint?}
 
@@
 
+A rule may also contain a negative application condition (NAC) which if matched over the source graph prevents application of the rule.
+
 A transformation rule also includes backward links, as informally introduced in \cref{subsec:DSLTrans_constructs}.
@@
 
-A DSLTrans transformation rule is a three-tuple $\big\langle \mathit{Match}$, $\mathit{Apply}$, $\mathit{backward}\big\rangle$, where:
+A DSLTrans transformation rule is a four-tuple $\big\langle \mathit{NAC}$, $\mathit{Match}$, $\mathit{Apply}$, $\mathit{backward}\big\rangle$, where:
 
@@
 \item $\mathit{Match}, \mathit{Apply} \in \textsc{Tg}$
-\item $\mathit{Match}, \mathit{Apply} $ are non-empty and disjoint
+\item $\mathit{Match}$ and $\mathit{Apply} $ are non-empty and are disjoint
+\item $\mathit{NAC}$ may be an empty graph, and is disjoint from both $\mathit{Match}$ and $\mathit{Apply} $
 \item $\mathit{backward} = \{E_{back}, (s_{back}, t_{back})\}$
-\begin{itemize}
-\item $E_{back} \cap E_{Match} = \emptyset, E_{back} \cap E_{Apply} = \emptyset$
-\end{itemize}
 \end{itemize}
 
-Note that when we require an element of the $\mathit{Match}$ or $\mathit{Apply}$ graphs, such as the vertices, we will index the required element.
+Note that when we require an element of the $\mathit{Match}$, $\mathit{Apply}$, or $\mathit{NAC}$ graphs, such as the vertices, we will index the required element.
 For example, the vertices for the \textit{Match} graph will be $V_{\textit{Match}}$.
@@
 \item $E_{back}$ contains the backward links
+\begin{itemize}
+\item $E_{back} $ is disjoint from $E_{Match}$,  $E_{Apply}$, and $E_{NAC}$
+\end{itemize}
 \item $(s_{back}, t_{back})$ is a pair of functions $s_{back}: E_{back}\rightarrow V_{\textit{Apply}}$ and $t_{back}: E_{back}\rightarrow V_{\textit{Match}}$ that respectively provide the source and target vertices for each backward link
@@
 
+Note that the rule's NAC will also be considered in the matching of a rule.
+
 \begin{definition}{Matcher of a Transformation Rule\\}
@@
 
-Let the transformation rule $r = \big\langle \mathit{Match}$, $\mathit{Apply}$, $\mathit{backward}\big\rangle$.
+Let the transformation rule $r = \big\langle \mathit{NAC}$, $\mathit{Match}$, $\mathit{Apply}$, $\mathit{backward}\big\rangle$.
 

### 2016-05-17 text/new_formal_background.tex
Fix up input-output model.

@@
 
-\newcommand{\defineggprime}{Let $\langle V,E,(s,t),\tau,VT,ET\rangle = g$, and\\ $\langle V',E',(s',t'),\tau',VT',ET'\rangle = g'$, where $g, g' \in \textsc{Tg}$.}
+\newcommand{\defineggprime}{Let $\langle V,E,(s,t),\tau,VT,ET\rangle = g$, and\\ $\langle V',E',(s',t'),\tau',VT',ET'\rangle = g'$, where $g, g' \in \textsc{TG}$.}
 
@@
 \item Edges $e\in E$ are noted $v\xrightarrow{e} v'$ if $s(e)=v$ and $t(e)=v'$, or simply e if the context is unambiguous
-\item The set of all typed graphs is called $\textsc{Tg}$
+\item The set of all typed graphs is called $\textsc{TG}$
 \item We define the empty graph to be a graph with $V = \emptyset$, $E = \emptyset$, and the other elements to be empty functions
@@
 
-Given a typed graph $g\in \textsc{Tg}$, will use the notation $Components(g)$ to describe the set of strongly connected typed graphs in $g$.
+Given a typed graph $g\in \textsc{TG}$, will use the notation $Components(g)$ to describe the set of strongly connected typed graphs in $g$.
 Also, we will use the notation $g|_{t}$ to refer to the restriction of graph $g$ to its subgraph containing only edges of type $t$.
@@
 
-The typed graph union is the function $\sqcup :\textsc{Tg}\times \textsc{Tg}\rightarrow \textsc{Tg}$ defined as:
+The typed graph union is the function $\sqcup :\textsc{TG}\times \textsc{TG}\rightarrow \textsc{TG}$ defined as:
 \begin{multline*}
@@
 
-Let $g$ be a graph $\langle V,E,(s, t),\tau\rangle \in \textsc{Tg}$. Then the transitive closure $g^{*} = \langle V,E',(s, t),\tau\rangle \in \textsc{Tg}$, where:
+Let $g$ be a graph $\langle V,E,(s, t),\tau\rangle \in \textsc{TG}$. Then the transitive closure $g^{*} = \langle V,E',(s, t),\tau\rangle \in \textsc{TG}$, where:
 
@@
 \begin{itemize}
-\item $\mathit{Match}, \mathit{Apply} \in \textsc{Tg}$
+\item $\mathit{Match}, \mathit{Apply} \in \textsc{TG}$
 \item $\mathit{Match}$ and $\mathit{Apply} $ are non-empty and are disjoint
@@
 \end{itemize}
+
+\textsc{Rules} is the set of all rules.
 
@@
 
-\subsubsection*{DSLTrans Transformation Layer}
+\subsubsection*{DSLTrans Layer and Transformation}
 
-\begin{definition}{Layer, Model Transformation\\}
-\label{def:layer_transformation}
+\cref{def:layer} and \cref{def:transformation} formalise the abstract syntax of a model transformation, introduced in \cref{sec:dsltrans}.
+An example of a model transformation can be observed in \cref{fig:dsltransformation}, the Police Station transformation.
+As expected, a DSLTrans transformation is composed of a sequence of layers where each layer is composed of a set of rules.
 
-A layer is a finite set of transformation rules $l\subseteq \textsc{Rule}^{sr}_{tg}$.
-The set of all layers for a source metamodel $sr$ and a target metamodel $tg$  is called $\textsc{Layer}^{sr}_{tg}$.
-A model transformation is a finite list of layers denoted $[l_{1}::l_{2}::\ldots::l_{n}]$ where $l_{k}\in \textsc{Layer}^{sr}_{tg}$ and $1\leq k \leq n$, $n\in \mathbb{N}$.
-We also impose that for any pair of rules $rl_1,rl_2\in \bigcup_{1\leq k\leq n}l_k$, if $\Vert rl_1\Vert \cong rl$ and $rl\sqsubseteq \Vert rl_2\Vert $ then $rl_2$ appears in a layer later than $rl_1$ and the apply parts of $rl_1$ and $rl_2$ are not isomorphic.
-The set of all transformations for a source metamodel $s$ and a target metamodel $t$ is called $\textsc{Transf}^{sr}_{tr}$.
+\begin{definition}{Layer\\}
+\label{def:layer}
+A layer is a finite set of transformation rules: $\{r_0, r_1, \dots, r_n | r_i \in \textsc{Rules}\}$.
 
-% The expansion of a layer $l\subseteq RULE^{sr}_{tg}$, noted $l^{\star}$, is such that that if
-% $rl\in l$ then $rl'\in l$ if $rl'\in rl^{\star}$. This means that if a
-% transformation rule is present is the layer, its expansion is also part of the
-% layer.
+The set of all layers is denoted $\textsc{Layers}$. 
+\end{definition}
+
+Note that the order of the rules within the layer does not matter, due to the semantics of DSLTrans rule execution.
+
+\begin{definition}{DSLTrans Transformation\\}
+\label{def:transformation}
+
+A DSLTrans transformation is a finite list of layers denoted $\{l_0, l_1, \dots, l_n | l \in \textsc{Layers}\}$. Note that the order of layers in a transformation is important.
+
+The set of all transformations is denoted $\textsc{Transforms}$.
 
@@
 
-\reviewer{The second half of Def. 16 raises a lot of questions. First of all, neither $\sqsubseteq 	 $ nor
-$\cong$ are defined over rules; however, I can imagine what it would look like. Then,
-what role does rl play; do you just mean $||rl_1 || \sqsubseteq ||rl_2 ||$ given a suitable isomorphic copy? What if $||rl_1 || \cong ||rl_2 || $(as a special case); is your condition then not
-unsatisfiable as $rl_1$ must appear after $rl_2$ as well as vice versa?}
+\subsubsection*{Input-Output Model}
 
-\reviewer{More importantly though, why do you care? Your semantics makes sure that
-rules are never considered for application twice in succession, so dependencies
-are always resolved unambiguously.}
+To describe the semantics of a DSLTrans model transformation, we must define an \textit{input-output model} construct.
+This input-output model allows the representation of intermediate operational states during the execution of a model transformation.
 
-\cref{def:layer_transformation} formalises the abstract syntax of a model transformation, introduced in \cref{sec:dsltrans}.
-An example of a model transformation can be observed in \cref{fig:dsltransformation}, the Police Station transformation.
-As expected, a formal DSLTrans transformation is composed of a sequence of layers where each layer is composed of a set of rules.
-The last condition of \cref{def:layer_transformation} imposes that, for any two pair of rules in the transformation, the matcher of the second rule never partially or totally subsumes (or contains) the matcher of the first rule, unless the second rule is in a subsequent layer and produces something more than the first rule.
-This condition avoids situations where the execution of a rule in a DSLTrans model transformation necessarily implies the execution of another rule (except for when rules having backward links necessarily execute because all their dependencies were created during the execution of previous layers).
-\reviewer{I do not fully understand why this
-  could be problematic. Do you have an example?}
-
-\paragraph{\textbf{Notation:}}
-
-We naturally extend to transformation rules (\cref{def:transformation_rule}) and transformation executions (\cref{def:modeltransformation}) the typed graph operations introduced in \cref{sec:formal_background}.
-Also, given a structure such as transformation rule $rl = \langle V,E,st,\tau,Match,\\Apply\rangle$, we will refer to the structure's components by using the component's name followed by the variable that holds the structure in between parenthesis.
-For example, we will write $V(rl)$ to designate the $V$ component of $rl$ or $Apply(rl)$ to designate $rl$'s $Apply$ component.
+The structure of a input-output model is intentionally very similar to a DSLTrans rule.
+There is one typed graph representing the input model and another typed graph representing the output model.
+As well, the construct contains a set of edges, named \emph{traceability links}, for keeping a history of which elements in the output model originated from which elements in the input model.
 
@@
 \label{def:input_output_model}
-An input-output model is a 6-tuple $\big\langle V,E,(s,t),\tau,Input,\\Output\big\rangle$, where:
+
+An input-output model rule is a three-tuple $\big\langle \mathit{Input}$, $\mathit{Output}$, $\mathit{trace}\big\rangle$, where:
+
 \begin{itemize}
-\item $Input=\langle V',E',st',\tau'\rangle\in \textsc{Instance}^{sr}$ is a model
-\item $Output=\langle V'',E'',st'',\tau''\rangle\in \textsc{Instance}^{tg}$ is a metamodel instance
-\item $Input$ and $Output$ are disjoint
+\item $\mathit{Input}, \mathit{Output} \in \textsc{TG}$
+\item $\mathit{Input}$ and $\mathit{Output} $ may be empty and are disjoint
+\item $\mathit{trace} = \{E_{trace}, (s_{trace}, t_{trace})\}$
+\begin{itemize}
+\item $E_{trace}$ contains the traceability links
+\item $E_{trace} $ is disjoint from $E_{Input}$ and $E_{Output}$
+\item $(s_{trace}, t_{trace})$ is a pair of functions $s_{trace}: E_{trace}\rightarrow V_{\textit{Output}}$ and $t_{trace}: E_{trace}\rightarrow V_{\textit{Input}}$ that respectively provide the source and target vertices for each traceability link
 \end{itemize}
 
+\end{itemize}
+
+\bentley{Do we need to have V and E for all vertices and edges in the IOM?}
 Additionally we have that $V=V'\cup V''$, $E\subseteq E'\cup E''$ and $\tau\subseteq \tau'\cup \tau''$, where the co-domain of $\tau$ is the union of the co-domains of $\tau'$ and $\tau''$ and the set $\{trace\}$.
 
-An edge $e \in E\setminus E'\cup E''$ is called a \emph{traceability link} and is such that $s(e)\in V''$, $t(e)\in V'$ and $\tau(e)=trace$.
-
- set of all match-apply patterns for a source metamodel $sr$ and a target metamodel $tg$ is called $\textsc{Iom}^{sr}_{tg}$.
 \end{definition}
-
-\reviewer{Why is Input a model and Output a metamodel instance? Note that
-$\backslash$ binds stronger than $\cup$ in $E \backslash E' \cup E$ ′′ . “match-apply patterns” → “input-output
-patterns”}
-%Change if needed
-% \levi{A match-apply model is the union of an input and output models, plus potential traceability links between those models.
-Traceability links are now typed using the typing function $\tau$.
-A traceability link it typed with type ``trace''.
-This makes the definitions shorter and more compact, which was mentioned by one of the reviewers of the paper.
-However, the fact that a match-apply model is now made up of models makes it that the input and the output of the transformation are always well-formed, meaning that the containment relation is always respected.
-It may be that a transformation does not always produce a model respecting the containment relation, which is a problem.
-This can be solved however if we only consider that well-formed transformations actually have semantics.}
-
-An input-output model is an object we will use when defining the semantics of a DSLTrans model transformation in \cref{sec:DSLTrans_formal}.
-It is composed of two metamodel instances, one called the \emph{input} and the other one the \emph{output}.
-An input-output model allows the representation of intermediate operational states during the execution of a model transformation.
-It may include a particular type of edges called \emph{traceability links}, for keeping a history of which elements in the output model originated from which elements in the input model.
 
@@
 \end{definition}
+
+Note that ${VT}_{Input}$ and ${ET}_{Input}$ will
 
@@
 
+\subsubsection*{Subsumption}
+
+ where $l_{k}\in \textsc{Layer}^{sr}_{tg}$ and $1\leq k \leq n$, $n\in \mathbb{N}$.
+We also impose that for any pair of rules $rl_1,rl_2\in \bigcup_{1\leq k\leq n}l_k$, if $\Vert rl_1\Vert \cong rl$ and $rl\sqsubseteq \Vert rl_2\Vert $ then $rl_2$ appears in a layer later than $rl_1$ and the apply parts of $rl_1$ and $rl_2$ are not isomorphic.
+% The expansion of a layer $l\subseteq RULE^{sr}_{tg}$, noted $l^{\star}$, is such that that if
+% $rl\in l$ then $rl'\in l$ if $rl'\in rl^{\star}$. This means that if a
+% transformation rule is present is the layer, its expansion is also part of the
+% layer.
+
+The last condition of \cref{def:layer_transformation} imposes that, for any two pair of rules in the transformation, the matcher of the second rule never partially or totally subsumes (or contains) the matcher of the first rule, unless the second rule is in a subsequent layer and produces something more than the first rule.
+This condition avoids situations where the execution of a rule in a DSLTrans model transformation necessarily implies the execution of another rule (except for when rules having backward links necessarily execute because all their dependencies were created during the execution of previous layers).
+\reviewer{I do not fully understand why this
+  could be problematic. Do you have an example?}
+
+\reviewer{The second half of Def. 16 raises a lot of questions. First of all, neither $\sqsubseteq 	 $ nor
+$\cong$ are defined over rules; however, I can imagine what it would look like. Then,
+what role does rl play; do you just mean $||rl_1 || \sqsubseteq ||rl_2 ||$ given a suitable isomorphic copy? What if $||rl_1 || \cong ||rl_2 || $(as a special case); is your condition then not
+unsatisfiable as $rl_1$ must appear after $rl_2$ as well as vice versa?}
+
+\reviewer{More importantly though, why do you care? Your semantics makes sure that
+rules are never considered for application twice in succession, so dependencies
+are always resolved unambiguously.}
+
+\subsection{Confluence and Termination Properties}
+
 We now prove two important properties about executions of transformations expressed in the subset of DSLTrans presented in this paper: \emph{confluence} and \emph{termination}.

### 2016-05-17 text/new_formal_background.tex
Work on defining model transformation.

@@
 
-For simplification purposes, we will not model the edge cardinalities given by a metamodel in our notion of typed graph.
-Note that in fact we require these cardinalities to be relaxed to perform our graph rewriting.
+For simplification purposes, we will not represent edge cardinalities or containment relationships given by a metamodel in our notion of typed graph.
+Note that in fact we require these conditions to be relaxed to perform our graph rewriting.
 
@@
 Note that these transitive edges are given by the function $\mathit{isIndirect}$.
+
+\bentley{I believe that the indirect edges don't need to be removed. There's a bit of handwaving, so maybe there are edge duplicates?}
 
@@
 
-\bentley{Do we need to have V and E for all vertices and edges in the IOM?}
+Let $\mathit{IOM}$ be the set of all input-output models.
+
+Note that the $\VT$ and $\ET$ for input and output will come from the input and output metamodels. \bentley{Expand}
+
+\bentley{Do we need to have V and E for all vertices and edges in the IOM? I don't think we talk about all vertices or edges in an IOM.}
 Additionally we have that $V=V'\cup V''$, $E\subseteq E'\cup E''$ and $\tau\subseteq \tau'\cup \tau''$, where the co-domain of $\tau$ is the union of the co-domains of $\tau'$ and $\tau''$ and the set $\{trace\}$.
@@
 
-\subsection{Semantic Transformation Constructs}
+\subsection{Transformation Semantics}
 \label{sec:DSLTrans_semantics}
 
-In the definition that follows we introduce the notion of execution of a DSLTrans model transformation.
-For our purposes it is sufficient to introduce it as an input-output model (see \cref{def:input_output_model}), containing the input model for the transformation, the produced output, and the traceability links built during execution.
-Due to space limitations, we cannot introduce the semantics of DSLTrans in the main text of this paper.
-We thus refer the reader to \cref{sec:DSLTrans_formal} for a formal description on how DSLTrans transformation executions are built.
+This section will discuss the semantics of a DSLTrans transformation. Given an input model and a transformation, an output layer will be produced through the repeated execution of rules and layers.
 
-\reviewer{Definition 17: There is a copy/paste error w.r.t. Def. B.16 (which only serves to
-strengthen my point, made above, about avoiding this kind of duplication). In
-this case, B.16 is the correct version.}
+\subsubsection*{Execution of a Model Transformation}
 
-\begin{definition} {Model Transformation Execution\\}
+We will begin our discussion of the semantics of model transformation by discussing the conditions for executing a model transformation on an input model.
+
+\begin{definition} {Execution of a Model Transformation\\}
 \label{def:modeltransformation}
 
-Let  $tr\in \textsc{Transf}^{sr}_{tg}$ be a transformation and $input \in \textsc{Model}^{sr}$ be a model.
-Assume we also have that: $$\langle V,E,st,\tau,input,\epsilon\rangle,tr \xrightarrow{trstep} \langle V',E',st',\tau',input,output\rangle$$
-A model transformation execution is the input-output model $\langle V',E',st',\tau',input,output\rangle\in \textsc{Iom}^{sr}_{tg}$, where $output\in \textsc{Iom}^{sr}_{tg}$ is an input-output model.
-The set of all model transformation executions for transformation $tr$ is written $\textsc{Exec}(tr)$.
-A model transformation with an empty input model is noted $\epsilon_{ex}$.
-Note that relation $trstep$ is formally defined in \cref{sec:DSLTrans_formal_appendix}.
+The execution of a model transformation is a function $transform: IOM \times Transforms \rightarrow IOM$.
+
 \end{definition}
 
-Note that ${VT}_{Input}$ and ${ET}_{Input}$ will
+For a well-formed transformation execution, we require the IOM in the domain to contain only an input graph.
+Recall that a IOM is the three-tuple $\big\langle \mathit{Input}$, $\mathit{Output}$, $\mathit{trace}\big\rangle$.
+We therefore require that both the $\mathit{Output}$ and $\mathit{trace}$ components in the domain IOM to be empty.
+This IOM thus represents the first step of the transformation, where no rule has been executed yet.
 
-Finally, as stated in \cref{def:modeltransformation}, we consider a model transformation execution to be the input-output model (IOM) resulting from executing a set of rules on a starting IOM.
-This starting IOM includes the transformation's input in its input part and has an empty output part.
-The starting IOM represents the first step of the transformation when no rule has been executed yet.
-A transformation execution results from executing all the rules in a DSLTrans model transformation.
+The IOM in the co-domain will contain this input graph, as well as the output graph and traceability links produced by the execution of the transformation.
+Let $\textsc{Execs}_t$ be the set of all well-formed IOMs produced for transformation $t$.
 
-\subsubsection{Transformation Language Semantics}
+\subsubsection*{Execution of a Rule}
 
-We will now address the semantics of the DSLTrans language.
+We will now address the execution of a rule to in the DSLTrans language.
+
 We will start by defining a match function that, given an input-output model and a transformation rule, returns all subgraphs of that input-output model where the rule's match pattern (including backward links) is found.
@@
 
+\subsubsection*{Layer Step Semantics}
+
 \begin{definition} {Layer Step Semantics}
@@
 The transformation rules in a layer are expanded before execution ($l^{\star}$) such that polymorphism in the match elements can be handled (see \cref{def:transformation_rule_expansion_appendix}).
-
-\begin{definition} {Model Transformation Execution}
-\label{def:modeltransformation_appendix}
-
-\CatchFileBetweenTags{\modeltransformation}{text/definitions}{modeltransformationappendix}{\modeltransformation}
-\end{definition}
-
-\CatchFileBetweenTags{\modeltransformationtextappendix}{text/definitions}{modeltransformationtextappendix}{\modeltransformationtextappendix}
 

### 2016-05-18 text/new_formal_background.tex
Try to make rule application fit double pushout theory.

@@
 
-\subsubsection*{Matcher of a Transformation Rule}
-
-To create the actual matcher construct that will be matched during a rule's execution, the rule's \textit{Match} graph is combined with backward links, as well as any \textit{Apply} vertices connected to the backward links.
-
-Note that the rule's NAC will also be considered in the matching of a rule.
-
-\begin{definition}{Matcher of a Transformation Rule\\}
-\label{def:back_match_transformation_rule}
-
-Let the transformation rule $r = \big\langle \mathit{NAC}$, $\mathit{Match}$, $\mathit{Apply}$, $\mathit{backward}\big\rangle$.
-
-We define $r$'s matcher, noted $\Vert r \Vert$, to be a typed graph six-tuple $\langle V,E,(s,t),\tau, \VT, \ET\rangle$, where:
-\begin{itemize}
-\item $V = V_{Match} \cup \big\{v \in V_{Apply} | \{\exists e \in E_{back} | s_{back}(e) = v\}\big\}$
-\item $E = E_{Match} \cup E_{back}$
-\item $s = s_{Match} \cup s_{back}$, $t = t_{Match} \cup t_{back}$
-\item $\tau = \tau_{Match} \cup \tau_{Apply}$
-\item $\VT = \VT_{Match} \cup \VT_{Apply}$
-\item $\ET = \ET_{Match} \cup \ET_{Apply}$
-\end{itemize}
-
-\end{definition}
-
 \subsubsection*{DSLTrans Layer and Transformation}
@@
 
-We will now address the execution of a rule to in the DSLTrans language.
+We will now address the execution of a rule in the DSLTrans language.
 
-We will start by defining a match function that, given an input-output model and a transformation rule, returns all subgraphs of that input-output model where the rule's match pattern (including backward links) is found.
+We will base our explanation on double pushouts, diagrammed in Figure~\ref{fig:dpo}.
+\bentley{Single or double pushouts?}
 
-% \levi{Need to be careful throughout all this section to understand whether we model the match and apply functions as functions or relations.
-Because of the non-deterministic generation of new vertices in the definition of apply function~\ref{def:apply_function}, relations are used in the definitions that follow to evaluate a transformation.
-This is an issue of the formalism used to model, should we make it explicit?}
+\begin{figure}[h!] \centering
+\includegraphics[width=0.44\textwidth]{figures/formal/dpo}
+	\caption{Double pushout approach}
+	\label{fig:dpo}
+\end{figure}
 
-\begin{definition}{Match Function}
-\label{def:match_function}
+We explain the components:
+\begin{itemize}
+\item L: The matcher of the rule
+\item R: The rewriter of the rule
+\item G: The graph to be matched/rewritten
+\item H: The rewritten graph
+\end{itemize}
 
-Let $m_{in}\in \textsc{Iom}^{sr}_{tr}$ be a input-output model and $rl\in \textsc{Rule}^{sr}_{tg}$ be a transformation
-rule. The $match : \textsc{Iom}^{sr}_{tg}\times \textsc{Rule}^{sr}_{tg}\rightarrow
-\mathcal{P}(\textsc{Iom}^{sr}_{tg})$ function is defined as follows: $$match_{rl}(m_{in})= \big\{ glue_{noind}\;|\;
-glue\sqsubseteq m_{in}^{*} \land glue \cong \Vert rl\Vert\big\}$$
+Note that as it is not possible in DSLTrans rules to remove elements, $L = K$, and $G = D$.
 
-where $glue\in \textsc{Iom}^{sr}_{tr}$ is an input-output model and $glue_{noind}$ is a version of $glue$ where the indirect links have been removed.
-% \levi{all backward links are replaced by trace links such that the matching using subgraph isomorphim can work.}
-% 
-% \levi{Where $rl^{-}$ is rule $rl$ without nodes in the apply part that are not connected to \emph{trace links}, as well as links adjacent to those nodes.
-$mg^{-ind}$ is the match graph without indirect links that come from the rule, given those cannot be part of the intermediate $MAM$ match result since that result is going to be united with the original $MAM$ (this was incorrect in the original SLE paper).}
-% \levi{The matcher function returns all graphs of the input MAM that are matched by a rule, including its backward links.
-Note that to find a subgraph ($\sqsubseteq$) of the input MAM that is isomorphic ($\cong$) to the rule being matched because the graph rewriting is achieved by joint union of the match graph extended by the new rewrite node, and the original MAM.}
-% \levi{TODO: review the $mg^{-ind}$ and the $rl^{-}$ notations.}
+\subsubsection*{Matcher of a Transformation Rule}
 
-% Due to the fact that the $\cong$ relation is based on the notion of graph
-% isomorphism, permutations of the same match result may exist in the
-% $\big\{g\;|\; g\lhd m \land g \cong strip(tr)\big\}$ set. The --- undefined ---
-% $remove:\mathcal{P}(TR^{s}_{t})\rightarrow \mathcal{P}(TR^{s}_{t})$ function is such that
-% it removes such undesired permutations.
+To create the actual matcher construct that will be matched during a rule's execution, the rule's \textit{Match} graph is combined with backward links, as well as any \textit{Apply} vertices connected to the backward links.
 
-%The $strip:TR^{s}_{t}\rightarrow TR^{s}_{t}$ function is such that
-%$$strip(\big\langle V,E,T,\langle V_m,E_m,T_m,s\rangle,\langle V_a,E_a,T_a,t\rangle,Bl,Il\big\rangle) = \big\langle V',E',T,\langle V_m,E_m,T_m\rangle,\langle V'_a,E'_a,T_a\rangle,Bl,Il\big\rangle$$
-%\begin{center}
-%where $(v'\in V'_a\Rightarrow v\rightarrow v'\in Bl)\;\land\;(v\rightarrow v'\in E'_a\Rightarrow (v\rightarrow v'\in E_m \land \{v,v'\}\subseteq V'_a))$
-%\end{center}
+Note that the rule's NAC will also be considered in the matching of a rule.
+
+\begin{definition}{Matcher of a Transformation Rule\\}
+\label{def:matcher_transformation_rule}
+
+Let the transformation rule $r = \big\langle \mathit{NAC}$, $\mathit{Match}$, $\mathit{Apply}$, $\mathit{backward}\big\rangle$.
+
+Let $\mathit{Match}\star$ be the transitive closure of $\mathit{Match}$ to handle any indirect edges.
+
+\bentley{Typed graph? Or copy the rule's structure? If it is a rule, note that it will not be well-formed. However, this is preferable to defining another structure to use}
+
+We define $r$'s matcher, noted $\lceil r \rceil$, to be a typed graph six-tuple $\langle V,E,(s,t),\tau, \VT, \ET\rangle$, where:
+\begin{itemize}
+\item $V = V_{Match\star} \cup \big\{v \in V_{Apply} | \{\exists e \in E_{back} | s_{back}(e) = v\}\big\}$
+\item $E = E_{Match\star} \cup E_{back}$
+\item $s = s_{Match\star} \cup s_{back}$, $t = t_{Match\star} \cup t_{back}$
+\item $\tau = \tau_{Match\star} \cup \tau_{Apply}$
+\item $\VT = \VT_{Match\star} \cup \VT_{Apply}$
+\item $\ET = \ET_{Match\star} \cup \ET_{Apply}$
+\end{itemize}
+
 \end{definition}
 
-\reviewer{Def. B.12. First of all, I wonder why you define match as a function rather than
-just defining what a match is; namely, an element of the set you are constructing
-here. Secondly (and more importantly), why do you not use the notion of homo-
-morphism instead of explicitly talking about ismorphic subgraphs? Your way,
-you lose the connection between glue and $||rl||$: you know they are isomorphic,
-but what if $||rl||$ has symmetries - then there is more than one match, which in
-your definition cannot be distinguished.}
+\subsubsection*{Rewriter of a Transformation Rule}
 
-\reviewer{You pay the price for this loss of information in Def. B.13, which now is broken: $a\Delta$ depends on $m_{glue} $(which was called glue in Def. 12, why the change in
-notation?) in a way depending on the mapping from $m_{glue}$ to $rl$ which you have
-discarded.}
+To continue with the double pushout approach, we must construct the rewriting (or replacement) graph.
 
-\reviewer{Def. B.13 is broken in another way as well: it does not properly define E' but
-merely imposes some requirements that are also fulfilled by choosing E' = E.
-Instead I know from the discussion about traceability links in Section 4 (which
-logically comes after this appendix!) that you really want to add such links for
-all newly created nodes of the output graph to all pre-existing nodes of the input
-graph; but that is not made clear anywhere at this point.}
+The construction of this rewriter is essentially the same as the underlying rule.
+However, to support traceability, we require that backward links in the rule be converted into traceability links, and that new traceability links be created between all match and apply vertices.
 
-The match function in \cref{def:match_function} looks for subgraphs ($glue\sqsubseteq m_{in}^{*}$) of an input-output model that are isomorphic to the backward matcher of the given transformation rule ($glue \cong \Vert rl\Vert$).
-Note that the containment transitive closure of the input-output model ($m_{in}^{*}$) is considered such that indirect links in the rule can looked for in the input model.
-Additionally, indirect links need to be removed from the input-output models resulting from the match function ($glue_{noind}$).
-This is so because indirect links are not part of the original input model, but rather an auxiliary structure.
+\begin{definition}{Rewriter of a Transformation Rule\\}
+\label{def:rewriter_transformation_rule}
 
-Let us now turn our attention to the apply function in \cref{def:apply_function}.
-Its role is to extend all model fragments found by matching a rule on a given input-output model, such that each of those fragments becomes isomorphic to the complete rule (minus its backward links).
-This process effectively creates the new objects and relations specified in the apply part of the rule, for each of the fragments found when matching the rule.
+Let the transformation rule $r = \big\langle \mathit{NAC}$, $\mathit{Match}$, $\mathit{Apply}$, $\mathit{backward}\big\rangle$.
 
-\begin{definition}{Apply Function}
-\label{def:apply_function}
+Let $\mathit{Match}\star$ be the transitive closure of $\mathit{Match}$ to handle any indirect edges.
 
-\reviewer{In your definitions you rely on the $\sqcup$ operator (sqare union),
-  the definition of which is not entirely clear.  For instance, I do not know where the operator used in Def. B13
-  is defined.}
+We define $r$'s rewriter, noted $\lfloor r \rfloor$, to be a input-output model $\big\langle \mathit{Input}$, $\mathit{Output}$, $\mathit{trace}\big\rangle$, where:
 
-Let $m_{glue}\in \textsc{Iom}^{sr}_{tg}$ be a input-output model and $rl\in \textsc{Rule}^{sr}_{tg}$ a
-transformation rule. The $apply : \textsc{Iom}^{sr}_{tg}\times \textsc{Rule}^{sr}_{tg}\rightarrow \textsc{Iom}^{sr}_{tg}$ function
-is defined as follows:
+\begin{itemize}
+\item $ \mathit{Input} = \mathit{Match\star}$
+\item $ \mathit{Output} = \mathit{Apply}$
+\end{itemize}
 
-$$apply_{rl}(m_{in})=\bigsqcup_{m_{glue}\in match_{rl}(m_{in})}trace_{a_{\Delta}}(m_{glue}\sqcup a_{\Delta})$$
+We then modify $\mathit{trace}$ to contain the appropriate traceability links
 
-\begin{center}
-where $a_{\Delta}\in \textsc{Iom}^{sr}_{tg}$ is such that $m_{glue} \sqcup a_{\Delta}\cong rl_{noind}$.\\
-\end{center}
+\begin{itemize}
+\item $E_{trace} = E_{backward} \cup \big\{e | \{$for each pair $m_i, a_j$ where $m_i \in V_{Match}, a_j \in V_{Apply}\}\big\}$
+\item $s_{trace} = s_{backward} \cup \big\{a_j | \{$for each pair $m_i, a_j$ where $m_i \in V_{Match}, a_j \in V_{Apply}\}\big\}$
+\item $t_{trace} = t_{backward} \cup \big\{m_i | \{$for each pair $m_i, a_j$ where $m_i \in V_{Match}, a_j \in V_{Apply}\}\big\}$
+\end{itemize}
 
-We impose that any instance of $a_\Delta$ is always disjoint from the $m_{in}$ input-output model and also that any two instances of $a_\Delta$ used in the large union are always disjoint.
+\bentley{Very messy}
+\end{definition}
 
-Partial function $trace:\textsc{Iom}\times \textsc{Iom} \nrightarrow \textsc{Iom}$ is such that $trace_{\langle V_{\Delta},E_{\Delta},st_{\Delta},\tau_{\Delta}\rangle}(\langle V,E,st,\tau\rangle) = \langle V,E',st',\tau'\rangle$ where we have that $E\subseteq E'$, $st\subseteq st'$ (using a light notational abuse for the $s\subseteq s'$ and $t\subseteq t'$), $\tau \subseteq \tau'$ and if $v_1\xrightarrow{e} v_2\in E'\setminus E$ then $v_1\in Output(V_{\Delta})$, $v_2\in Input(V)$ and $\tau'(e)=trace$.
-Finally, $rl_{noind}$ is a version of $rl$ where indirect links have been removed.
-% More concretely, if we have $mam = $\big\langle V,E,st,\tau, Input,Output\big\rangle\in MAM^{sr}_{tg}$ where $Input=\langle V',E',st',\tau'\rangle\in MODEL^{sr}$ and $Output=\langle
-% V'',E'',st'',\tau''\rangle\in MODEL^{tg}$, then $trace(mam) = \big\langle V,E,st,\tau''',Input,Output\big\rangle$ where $E'\subseteq E$ and $v\rightarrow E'\setminus E$ 
-\end{definition}
-% 
-% it builds traceability edges between vertices of the output model that were newly built by the $apply$ function (the vertices in ${mg}_{\Delta}$) and all the vertices from the input part of the input-output model.
-It is necessary to remove the indirect links from the rule because otherwise they would be created during the rewriting by the apply function.
+\subsubsection*{Application of a Transformation Rule (continued)}
 
-In \cref{def:apply_function} $a_\Delta$ is an input-output model that contains an instance of the target metamodel.
-These instances are created by rule $rl$ and are used to extend the sub-models found by the match function.
-The $trace$ function builds traceability edges between newly created vertices of the output model in $a_{\Delta}$ and all the vertices from the input part of a model fragment found by the match function.
+\bentley{Fix up morphism notation. Is $\rightarrow$ good enough?} 
+Recall that the interface graph K is equivalent to L due to the lack of deletion in DSLTrans rules.
+Thus, the l homomorphism is identity.
+As well, we consider r to be a trivial homomorphism between the $\mathit{Match} \rightarrow \mathit{Input}$, $\mathit{Apply} \rightarrow \mathit{Output}$, and $\mathit{backward} \rightarrow \mathit{trace}$.
 
-Note that, because we do not pose any constraints on $a_\Delta$ other than the fact that its union with the sub-model $m$ is isomorphic to $rl_{noind}$, the $a_\Delta$ variable can always be satisfied by an unlimited amount of input-output models.
-In order to avoid an infinite amounts of results when a transformation rule is executed, in what follows we will consider transformation results differ only up to typed graph isomorphism.
+We also note here that a rule cannot be applied if there is a homomorphism between the rule's NAC and G.
 
-%  Note that because the vertices are not disjoint, when the graph produced by the apply function is united with the original match apply model the new vertices are created.
-It was necessary to remove the indirect links from the rule during rewriting, because otherwise those indirect links would also be rewritten.
-
-Definitions~\ref{def:match_function} and~\ref{def:apply_function} are complementary: the former gathers all the fragments of an input-output model that are matched by a transformation rule; the latter glues on the output part of each of those fragments new objects and relations created by a transformation rule.
-
-%  The $strip$ function is used to
-%  enable matching over backward links but not elements to be created by the
-%  transformation rule. The $back$ function connects all newly created vertices
-%  to the elements of the source model that originated them.
-
-%\begin{definition}{Transformation}
-
-%Let $g\in TG$ be a typed graph. The graph transform function $transform : TR\times TG\rightarrow TG$ is recursively defined as:
-%\begin{gather*}
-%  transform_{(m,a)}(g) =
-%  \begin{cases}
-%    g  & \text{if } match_{m}(g)=\emptyset \\
-%    instance(a)\cup transform_{(m,a)}(g') &\text{if } match_{m}(g)\neq \emptyset
-%  \end{cases}
-%\end{gather*}
-%$$\text{where } g'=mark_{(m)}(g)$$
-%\end{definition}
-
-%\subsection{Transformation Semantics}
-%\begin{definition} {Layer Semantics}
-
-%Let $l\in Layer$ be a layer and $\{m,m'\} \in MAM^{s}_{t}$ be a match
-%$$\frac{}
-%{m,\emptyset \stackrel{layerstep}{\rightarrow}m}$$
-%$$\frac{tr\in layer,\; transform_{tr}(m) = m''\; m'',layer\backslash\left\{tr\right\} \xrightarrow{layerstep}	 m'}{m,layer \xrightarrow{layerstep} m'}$$
-%\end{definition}
+\begin{itemize}
+\item m is the typed graph homomorphism \bentley{Really is rule over input-output model}
+\item D = G
+\item n is homomorphism too
+\item g is too?
+\end{itemize}
 

### 2016-06-27 text/new_formal_background.tex
Rewrite layer and transformation execution. Very rough, but gives the flavour, and most detail has been handled by other constructs.

@@
 \begin{itemize}
-\item $\mathit{Match}, \mathit{Apply} \in \textsc{TG}$
+\item $\mathit{NAC}$, $\mathit{Match}, \mathit{Apply} \in \textsc{TG}$
 \item $\mathit{Match}$ and $\mathit{Apply} $ are non-empty and are disjoint
@@
 
-\subsubsection*{Execution of a Model Transformation}
-
-We will begin our discussion of the semantics of model transformation by discussing the conditions for executing a model transformation on an input model.
-
-\begin{definition} {Execution of a Model Transformation\\}
-\label{def:modeltransformation}
-
-The execution of a model transformation is a function $transform: IOM \times Transforms \rightarrow IOM$.
-
-\end{definition}
-
-For a well-formed transformation execution, we require the IOM in the domain to contain only an input graph.
-Recall that a IOM is the three-tuple $\big\langle \mathit{Input}$, $\mathit{Output}$, $\mathit{trace}\big\rangle$.
-We therefore require that both the $\mathit{Output}$ and $\mathit{trace}$ components in the domain IOM to be empty.
-This IOM thus represents the first step of the transformation, where no rule has been executed yet.
-
-The IOM in the co-domain will contain this input graph, as well as the output graph and traceability links produced by the execution of the transformation.
-Let $\textsc{Execs}_t$ be the set of all well-formed IOMs produced for transformation $t$.
-
-\subsubsection*{Execution of a Rule}
+\subsubsection*{Execution of a DSLTrans Rule}
 
@@
 \bentley{Fix up morphism notation. Is $\rightarrow$ good enough?} 
-Recall that the interface graph K is equivalent to L due to the lack of deletion in DSLTrans rules.
-Thus, the l homomorphism is identity.
+Recall that the interface graph K is equivalent to L due to the lack of deletion in DSLTrans rules. Thus, the l homomorphism is identity. The D correspondence graph will therefore be equivalent to G.
+
 As well, we consider r to be a trivial homomorphism between the $\mathit{Match} \rightarrow \mathit{Input}$, $\mathit{Apply} \rightarrow \mathit{Output}$, and $\mathit{backward} \rightarrow \mathit{trace}$.
@@
 
-\begin{itemize}
-\item m is the typed graph homomorphism \bentley{Really is rule over input-output model}
-\item D = G
-\item n is homomorphism too
-\item g is too?
-\end{itemize}
+m is the typed graph homomorphism between the matcher of a rule $r$, denoted ($\lceil r \rceil$) and the $\mathit{Input}$ component of the input-output model
 
-\subsubsection*{Layer Step Semantics}
+$n$ and $g$ will be homomorphism between their respective graphs. \bentley{This needs to be explained further. Maybe with a diagram?}
 
-\begin{definition} {Layer Step Semantics}
-\label{def:layer_step_semantics}
+\subsubsection*{DSLTrans Layer Execution}
 
-Let $l\in \textsc{Layer}^{sr}_{tg}$ be a Layer. The \emph{layer step relation}
-$\stackrel{layerstep}{\rightarrow}\subseteq \textsc{Iom}^{sr}_{tg}\times \textsc{Iom}^{sr}_{tg} \times \textsc{Layer}^{sr}_{tg}\times
-\textsc{Iom}^{sr}_{tg}$ is defined as follows:
+Now that the application of a particular DSLTrans rule has been explained, the next definition defines how a DSLTrans transformation layer executes.
+Recall that a layer is composed of rules, and acts on the current input-output model.
 
-$$\frac{}
-{\langle m_{in},m_{glue},\emptyset\rangle \xrightarrow{layerstep}m_{in} \sqcup m_{glue}}$$
+\begin{definition} {DSLTrans Layer Execution}
+\label{def:dsltrans_layer_execution}
 
-\begin{multline*}
-\frac{\begin{array}{ll} rl\in l,\; apply_{rl}&(m_{in}) = m_{rout},\\
-&\langle m_{in},m_{glue}\sqcup m_{rout},l\backslash \{rl\}\rangle \xrightarrow{layerstep} m_{out}
-\end{array}}
-{\langle m_{in},m_{glue},l\rangle \xrightarrow{layerstep} m_{out}}$$
-\end{multline*}
+The execution of a DSLTrans layer can be defined as a function as follows:
 
-\begin{center}
-where $m_{rout} \in\textsc{Iom}^{sr}_{tg}$ and $rl\in \textsc{Rule}^{sr}_{tg}$.
-\end{center}
+\textit{applyLayer} : IOM $\times$ Layer $\rightarrow$ IOM
 
-We impose that all input-output models that are part of $rout$ and have been generated by rule $rl$ are disjoint from input-output models accumulated in $m_{glue}$ that have been generated by other rules.
+Let $l = \{r_0, r_1, \dots, r_n | r_i \in \textsc{Rules}\}$, $l \in \textsc{Layers}$.
+
+Let $\mathit{IOM}_{Input}$, $\mathit{IOM}_{Output} \in \mathit{IOMs}$.
+
+$\mathit{IOM}_{Output}$ will be the result of applying the production \textit{p} as defined below to $\mathit{IOM}_{Input}.$
+
+We refer to the double-pushout literature for the creation of \textit{p}.
+Specifically, we refer to the definition of parallel direct derivations.
+This p will be created out of the productions created for each of the rules $r_i$ in the layer.
+
+$p = \big\langle (p_1, {in}^1), \dots, (p_k, {in}^k) \big\rangle : (L \xleftarrow{l} L \xrightarrow{r} R)$
+
+This collection of the rules is possible because DSLTrans rules do not delete anything. Essentially, a union is taken of the L, K, R components of each production.
+\end{definition}
+
+\begin{figure}[h!] \centering
+\includegraphics[width=0.44\textwidth]{figures/formal/parallel_productions}
+	\caption{Parallel production \textit{p}}
+	\label{fig:parallel_productions}
+\end{figure}
+
+\cref{def:dsltrans_layer_execution} is the core of DSLTrans' semantics, in which we build the result of executing a layer of a DSLTrans transformation.
+Many model transformation languages are based on graph rewriting, where the result of each rule rewrite is immediately usable by all other rules.
+In DSLTrans the result of executing one layer in DSLTrans is totally produced before the input to the layer is changed.
+This is enforced in \cref{def:dsltrans_layer_execution} by the fact that the layer production is composed of a union of the rule's productions.
+Rules belonging to the same layer are thus forced to execute independently, as described in the section on DSLTrans semantics.
+
+\bentley{How does this handle different multiplicity of rule application?}
+
+\subsubsection*{DSLTrans Transformation Execution}
+
+Our final definition is for the execution of a DSLTrans transformation. Essentially, it is the chaining of layer executions applied on an input-output model.
+
+We will begin by discussing the conditions for executing a model transformation on an input model.
+
+\paragraph{Input Conditions}
+
+For a well-formed transformation execution, we require the \textit{input-output model} in the domain to contain only an input graph.
+Recall that an \textit{input-output model} is the three-tuple $\big\langle \mathit{Input}$, $\mathit{Output}$, $\mathit{trace}\big\rangle$.
+We therefore require that both the $\mathit{Output}$ and $\mathit{trace}$ components in the domain IOM to be empty.
+This input-output model thus represents the first step of the transformation, where no rule has been executed yet.
+
+The input-output model in the co-domain will contain this input graph, as well as the output graph and traceability links produced by the execution of the transformation.
+Let $\textsc{Execs}_t$ be the set of all well-formed input-output models produced for transformation $t$.
+
+\begin{definition} {Execution of a DSLTrans Transformation\\}
+\label{def:DSLTrans_transformation_execution}
+
+The execution of a model transformation is a function $\mathit{applyTransformation}: \mathit{IOM} \times \mathit{Transforms} \rightarrow \mathit{IOM}$.
+
+This function is a chaining of executions of those layers within the transformation.
+
+Recall the application of a layer is \textit{applyLayer} : IOM $\times$ Layer $\rightarrow$ IOM. Therefore, the application of a transformation is:
+
+$\mathit{IOM}$ as input to chaining of layer executions \bentley{Is this function chaining? I don't know how to write the equation for this.}
 
@@
 
-\reviewer{Def. B.14: It seems to me rather roundabout to sequentially apply all rules, ac-
-cumulate the results and then take their union. Why do you not take the union
-simultaneously with the parallel application of all rules (of a given layer), as
-you are essentially doing for all applications of a single rule in Def. B.13? This
-would get rid of the artificial sequential ordering, and the whole issue of con-
-fluence would not even have to be raised as there is no non-determinism any
-more.}
-
-\reviewer{In the constructions defined in Appendix B, you ignore the required absence of
-containment cycles. Might such cycles not appear in the union of the graphs,
-even if the individual graphs are cycle-free?}
-
-In \cref{def:layer_step_semantics} we build the result of executing a layer of a DSLTrans transformation.
-The operational semantics-like rules in the definition execute each rule $rl$ in layer $l$, in a non-deterministic order, by using the $apply$ function.
-The result of executing each rule is accumulated in the temporary $m_{glue}$ input-output model.
-Finally, when the set of transformation rules in the layer has been exhausted, the result of executing all the rules in the layer (now contained in $m_{glue}$) is united with the input-output model $m_{in}$, the input to layer $l$.
-Note that this final union produces the result we expect because of the fact that the $m_{glue}$ input-output model is not disjoint from $m_{in}$.
-The common ``glue'' parts of $m_{glue}$ that have been built by the match function and extended by the apply function are now used to built the result of executing layer $l$.
-
-\cref{def:layer_step_semantics} is the core of DSLTrans' semantics.
-Many model transformation languages are based on graph rewriting, where the result of each rule rewrite is immediately usable by all other rules.
-In DSLTrans the result of executing one layer in DSLTrans is totally produced before the input to the layer is changed.
-This is enforced in \cref{def:layer_step_semantics} by the fact that the apply function always executes over the same $m_{in}$ input-output model and all the results of rule execution in the same layer are added to the $m_{glue}$ structure that is write-only.
-Rules belonging to the same layer are thus forced to execute independently.
-
-\begin{definition} {Transformation Step Semantics}
-\label{def:transformation_step_semantics}
-
-Let $[l::tr]\in \textsc{Transf}^{sr}_{tg}$ be a transformation, where $l\in
-\textsc{Layer}^{sr}_{tg}$ is a Layer and $tr$ also a transformation. The \emph{transformation step relation}
-$\stackrel{trstep}{\rightarrow}\subseteq \textsc{Iom}^{sr}_{tg}\times \textsc{Transf}^{sr}_{tg}\times
-\textsc{Iom}^{sr}_{tg}$ is defined as follows: 
-$$\frac{}{\langle m,[]\rangle \xrightarrow{trstep} m}$$
-
-$$\frac{\big\langle m_{in},\epsilon,l^{\star}\big\rangle \xrightarrow{layerstep} m_{inter},\; \langle m_{inter},R\rangle \xrightarrow{trstep} m_{out}}{\langle m_{in},[l::T]\rangle \xrightarrow{trstep} m_{out}} \hspace{.3cm}$$
-
-$$\text{where } l^{\star}=\bigcup_{rl\in l}rl^{\star}$$
-\end{definition}
-
-\reviewer{I believe tr, T and R in this definition stand for the same thing.}
 While the execution of the rules belonging to a layer happens in parallel, the execution of the layers of a transformation happens sequentially.
-As per \cref{def:transformation_step_semantics}, the input-output model $m_{inter}$ is the output of executing a given layer that is passed onto the next layer as input.
-Note that an empty input-output model ($\epsilon$) is passed as the second argument to the $layerstep$ relation in \cref{def:transformation_step_semantics}.
-This is because in \cref{def:layer_step_semantics} of layer step semantics, the second argument of the relation is used as an accumulator for the model fragments that are added to be added to the output of the previous layer once all the rules in the current layer have been executed.
-The transformation rules in a layer are expanded before execution ($l^{\star}$) such that polymorphism in the match elements can be handled (see \cref{def:transformation_rule_expansion_appendix}).
-
-% A transformation execution is formed from taking the input model, and executing
-% the transformation on it to produce the output model. During this
-% transformation, traceability links will be placed between match elements and the
-% apply elements they create. Definition~\ref{def:transf_ex} expresses this
-% formally. Note that we assume that transformation executions are built following
-% the algorithm described in~\cite{DBLP:conf/sle/BarrocaLAFS10}.
-
-% \begin{definition}{Transformation Execution}
-% \label{def:transf_ex}
-% 
-% Let $t$ be a DSLTrans transformation having source metamodel $s$ and target
-% metamodel $t$. A Transformation Execution is a 6-tuple $\langle V,E\cup
-% Tl,\tau,Match,Apply,Tl\rangle$, where $\langle V,E,\tau,Match,Apply\rangle \in
-% MAP^{s}_{t}$ is a match-apply pattern. $Match=\langle V',E', \tau',s\rangle$,
-% $Apply=\langle V'',E'', \tau'',t\rangle$ and the edges $Tl\subseteq V'\times
-% V''$ are called \emph{traceability links}. The set of all transformation
-% executions having source metamodel $s$ and target metamodel $t$ is called
-% $Exec^{s}_{t}$.
-% \end{definition}
-
-\subsubsection*{Subsumption}
-
- where $l_{k}\in \textsc{Layer}^{sr}_{tg}$ and $1\leq k \leq n$, $n\in \mathbb{N}$.
-We also impose that for any pair of rules $rl_1,rl_2\in \bigcup_{1\leq k\leq n}l_k$, if $\Vert rl_1\Vert \cong rl$ and $rl\sqsubseteq \Vert rl_2\Vert $ then $rl_2$ appears in a layer later than $rl_1$ and the apply parts of $rl_1$ and $rl_2$ are not isomorphic.
-% The expansion of a layer $l\subseteq RULE^{sr}_{tg}$, noted $l^{\star}$, is such that that if
-% $rl\in l$ then $rl'\in l$ if $rl'\in rl^{\star}$. This means that if a
-% transformation rule is present is the layer, its expansion is also part of the
-% layer.
-
-The last condition of \cref{def:layer_transformation} imposes that, for any two pair of rules in the transformation, the matcher of the second rule never partially or totally subsumes (or contains) the matcher of the first rule, unless the second rule is in a subsequent layer and produces something more than the first rule.
-This condition avoids situations where the execution of a rule in a DSLTrans model transformation necessarily implies the execution of another rule (except for when rules having backward links necessarily execute because all their dependencies were created during the execution of previous layers).
-\reviewer{I do not fully understand why this
-  could be problematic. Do you have an example?}
-
-\reviewer{The second half of Def. 16 raises a lot of questions. First of all, neither $\sqsubseteq 	 $ nor
-$\cong$ are defined over rules; however, I can imagine what it would look like. Then,
-what role does rl play; do you just mean $||rl_1 || \sqsubseteq ||rl_2 ||$ given a suitable isomorphic copy? What if $||rl_1 || \cong ||rl_2 || $(as a special case); is your condition then not
-unsatisfiable as $rl_1$ must appear after $rl_2$ as well as vice versa?}
-
-\reviewer{More importantly though, why do you care? Your semantics makes sure that
-rules are never considered for application twice in succession, so dependencies
-are always resolved unambiguously.}
+As per \cref{def:DSLTrans_transformation_execution}, the input-output model that is the output of executing a given layer is passed onto the next layer as input.
+The final input-output model created will be the result of the model transformation.
 

### 2016-06-29 text/abstraction_relation.tex
Place abstraction relation first.

@@
-\section{Abstraction Relation between Path Conditions and Transformation Executions}
+\section{Path Conditions and the Abstraction Relation}
 \label{sec:abstraction_relation}
 
-% More informally, there exists a typed graph injective homomorphism between
-% $Match_{pc}$ and $Input_x$  \textbf{and} $\langle V_x,E_x\setminus
-% E_{mx},\tau_x\rangle$ is a typed graph strict instance of $\langle
-% V_{pc},E_{pc}\setminus E_{pc},\tau_{pc}\rangle$.
-
-% \levi{It is possible that this relation is required to be a typed graph strict
-% instance between the path condition and the execution. Such a constraint would
-% oblige each set of executions for a path condition to be truly distinct from
-% the set of executions for another path condition. With the current definition
-% the executions for a path condition including rule 'A' will overlap with the
-% executions for a path condition including rules 'AB'. If this is so validity
-% needs to be proved again for the new relation between the match patterns of PC
-% and ex. Elements that do not show up in the match part of the rules are
-% excluded from the typed graph strict instance.}
-
-In this section we define the abstraction relation between the execution of a
-DSLTrans transformation and the path condition that represents it.
-This abstraction relation allows us to prove properties on a finite set of representative
-path conditions, as created by the path condition generation algorithm. As this set is finite, our technique is  guaranteed to be decidable.
+This section will present how path conditions are structured to represent symbolic rule execution.
+As well, we define the abstraction relation between the execution of a DSLTrans transformation and the path condition that represents it.
+This abstraction relation allows us to prove properties on a finite set of representative path conditions, as created by the path condition generation algorithm.
+As this set is finite, our technique is guaranteed to be decidable.
 
@@
 
+\subsection{Symbolic Execution}
+
+Our algorithm operates on the principle of symbolic execution to build up these
+path conditions. In order to explain the concept of symbolic execution of a
+transformation, let us make an analogy with program symbolic execution as
+introduced by King in his seminal work ``\emph{Symbolic Execution and Program
+Testing}''~\cite{DBLP:journals/cacm/King76}. According to King, a symbolic
+execution of a program is a set of \emph{constraints} on that program's
+\emph{input variables} called \emph{path conditions}. Each \emph{path condition}
+describes a traversal of the conditional branching commands of that program. A
+\emph{path condition} is symbolic in the sense it \emph{abstracts} as many
+concrete executions as there are instantiations of the path condition's
+variables that render the path condition's constraints true.
+
+We can transpose this notion of symbolic execution to model transformations. The
+analog of an input variable in the model transformation context are
+\emph{metamodel classes, relations and attributes}. As program statements impose
+constraints on input and output variables during symbolic execution,
+transformation rules impose conditions on which metamodel elements are
+instantiated during a concrete transformation execution, and how that
+instantiation happens. As well, rules in a model transformation are implicitly
+or explicitly scheduled. These control and/or data dependencies must be
+taken into consideration during path condition construction.
+
+As in program symbolic execution, each path condition in our approach \emph{abstracts} as many concrete executions as there are input/output models
+that satisfy them. This is formulated as an \emph{abstraction relation} as further explained in \cref{subsec:abstraction_relation}.
+
+In what follows we will examine in more detail how these symbolic execution
+principles can apply to the verification of model transformations.
+
+\subsection{Path Conditions}
+\label{sec:gen_path_conds}
+
+In order to present the intuition of path conditions and symbolic executions, we
+first discuss the idea of \textit{rule combinations}.
+
+As seen in \cref{sec:dsltrans}, a layer in a DSLTrans transformation contains a
+number of rules.  We can create a set of rule combinations for this layer by
+taking the powerset of all rules in that layer. Each rule combination in this
+set will represent all possible transformation executions where the rules in
+that combination would execute.
+
+For example, in \cref{fig:rule_combos2}, the rule combination marked `AC'
+represents the set of transformation executions where the rules A and C would
+execute and no others. Another rule combination marked `A' represents the
+transformation executions where only rule A would execute.
+
+\begin{figure}[h!] \centering \includegraphics[width=.40\textwidth]{./figures/overview/rule_combos.pdf}
+	\caption{Rule combinations created for a transformation layer}
+	\label{fig:rule_combos2}
+\end{figure}
+
+Note that within these rule combinations, the number of times a rule has
+executed is abstracted. Either a rule has executed zero times, and the rule is
+not represented in a rule combination, or the rule has executed some finite
+number of times and it is represented. This abstraction is key to our approach,
+as it allows us to create a finite set of path conditions to abstract over an
+infinite set of transformation executions, as seen in
+\cref{sec:abstraction_relation}.
+
+We also note that rule \textit{combinations} are created, and not rule
+\textit{permutations}. This follows from the semantics of \\DSLTrans as described
+in \cref{sec:dsltrans}, as transformation rules in a layer will execute in a
+non-deterministic order but produce a deterministic result, by construction of the semantics of DSLTrans. As a final
+note, the transformation executions that these rule combinations represent always terminate, also by construction of the semantics of DSLTrans~\cite{DBLP:conf/sle/BarrocaLAFS10}.
+
+We base our concept of path conditions on these rule combinations. However, as
+DSLTrans allows for dependencies between rules, we cannot create path conditions
+for the transformation by taking the powerset of all rules. Instead, our
+approach must move layer-by-layer and resolve the dependencies between rules.
+The next two sections will introduce the concepts of traceability and dependency, before we
+briefly discuss the syntax and semantics of path conditions themselves.
+
+\subsubsection{Traceability}
+\label{subsubsec:traceability}
+
+DSLTrans rules specify which elements of the
+output model were created from specific elements of the input model. To resolve
+these dependencies, traceability information for the transformation is created
+during the execution of a DSLTrans model transformation~\cite{DBLP:conf/sle/BarrocaLAFS10}.
+In our verification approach, we store this same information as symbolic \emph{traceability links}, in
+order to record which elements belong to the same DSLTrans rule.
+
+At a particular point in the path condition construction process, symbolic traceability links are built for
+each rule as follows: for all match and apply elements of a rule, given a match
+element belonging to the match graph of a rule and an apply element belonging to
+the apply graph of the same rule, a symbolic traceability link is built between the two
+if the apply element is not connected to a backward link (as explained below).
+This is intuitive: traceability links are built between a newly generated
+element in the output model, and the elements of the input model that originated
+it.
+
+\begin{figure}[htb]
+        \centering
+        \begin{subfigure}[b]{0.24\textwidth}
+                \centering
+                \includegraphics[width=1\textwidth]{./figures/building_path_conditions/traceability_links.pdf}
+                \caption{Before symbolic traceability links added}
+                \label{fig:traceability_links1}
+        \end{subfigure}%
+        ~~
+        \begin{subfigure}[b]{0.235\textwidth}
+                \centering
+                \includegraphics[width=1\textwidth]{./figures/building_path_conditions/traceability_links2.pdf}
+                \caption{After symbolic traceability links added}
+                \label{fig:traceability_links2}
+        \end{subfigure}%
+        \caption{Symbolic traceability links created for an abstract DSLTrans rule}
+        \label{fig:trace_links}
+\end{figure}
+
+An example of the symbolic traceability link creation process is shown in
+\cref{fig:trace_links}. Note that symbolic traceability links are solid lines between
+match and apply elements in our visual notation.
+
+\subsubsection{Backward Links}
+\label{subsubsec:backward_links}
+
+The dependencies in a DSLTrans rule are specified using the \textit{backward link} construct, as further detailed in \cref{subsec:DSLTrans_constructs} and \cref{def:transformation_rule}.
+\cref{subsubsec:resolve_dependencies} will discuss how these dependencies are then resolved during our symbolic execution approach.
+
+\cref{fig:backward_links} demonstrates how backward links are used \\within a rule.
+The rule shown contains a backward link, which defines the dependency that an element of type X was created from an element of type A, and an element of type Y was created from an element of type B.
+If this dependency is satisfied, then another element of type Z should be created.
+This element should be associated with the Y element.
+
+\begin{figure}[htb]
+        \centering
+        \begin{subfigure}[b]{0.235\textwidth}
+                \centering
+                \includegraphics[width=1\textwidth]{./figures/building_path_conditions/backward_link.pdf}
+                \caption{Rule with backward links (dashed lines)}
+                \label{fig:backward_links}
+        \end{subfigure}%
+        ~~
+        \begin{subfigure}[b]{0.235\textwidth}
+                \centering
+                \includegraphics[width=1\textwidth]{./figures/building_path_conditions/backward_link2.pdf}
+                \caption{Traceability links added\\~}
+                \label{fig:backward_links2}
+        \end{subfigure}%
+        \caption{Adding traceability links to an abstract DSLTrans rule with backward links}
+        \label{fig:back_links}
+\end{figure}
+
+\cref{fig:backward_links2} shows the rule after symbolic traceability links have been added.
+Two symbolic traceability links are created from the Z element to the A and B elements in the match graph to store traceability information.
+Note that no symbolic traceability links are built between two elements connected by backward links, as these links have already been built in a previous layer.
+
+\subsubsection{Syntax and Semantics}
+\label{subsubsec:path_condition_creation}
+
+A path condition represents the symbolic execution of a set of DSLTrans rules, similar to a rule combination as explained above.
+Each path condition represents the execution of a set of transformation rules, by containing the input and output elements which are produced by the execution of those transformation rules.
+
+Again, we use an abstraction over the number of times a rule has symbolically executed. Each path condition will represent that a rule has not executed, or has executed one or more times.
+
+   \begin{figure}[t]
+     \begin{center}
+       \includegraphics[width=0.45\textwidth]{figures/path_conditions/pc.pdf}
+       \caption{An example path condition representing the execution of three rules}
+       \label{fig:pc_first}
+     \end{center}
+     \vspace{-0.20in}
+   \end{figure}
+
+For example, the path condition in Figure~\ref{fig:pc_first} represents the execution of three rules in the transformation.
+This representation includes the input and output elements that will be present in the input and output models if these three rules execute.
+The set of path conditions produced by the prover will therefore partition the set of valid executions of the transformation, where each execution is an input/output model pair.
+This technique was first proposed in~\cite{Lucio2010} and further detailed in~\cite{Lucio2014}.
+
+%The path condition generation algorithm will symbolically combine transformation rules into a path condition.
+Each path condition will then abstract a set of concrete transformation executions, as defined by our abstraction relation in \cref{sec:abstraction_relation}.
+
+As seen in the rest of this section, the structure of path conditions is similar to that of DSLTrans rules.
+The match graph of a path condition represents a pattern that must be present in the input model of the transformation, while the apply graph is a pattern which will be instantiated in the output model of the transformation.
+Symbolic traceability links are also kept between elements in the match and apply graphs to retain traceability information.
+
+The formal definition of a path condition is presented in \cref{def:path_condition}.
+
+\begin{definition}{Path Condition\\}
+\label{def:path_condition}
+
+A path condition is a 7-tuple $\big\langle V,E,(s,t),\tau, Match,Apply,\\Rulecop\big\rangle$, where: $Match=\langle V',E',st',\tau'\rangle\in \textsc{Ipattern}^{sr}$ is an indirect pattern; $Apply=\langle V'',E'',st'',\tau''\rangle\in \textsc{Pattern}^{tg}$ is a pattern; $Match$ and $Apply$ are disjoint graphs.
+We also have that $V=V'\cup V''$, $E\subseteq E'\cup E''$ and $\tau\subseteq \tau'\cup \tau''$ where the co-domain of $\tau$ is the union of the co-domains of $\tau'$ and $\tau''$ and the set $\{trace\}$.
+An edge $e \in E\setminus E'\cup E''$, called a \emph{symbolic traceability link}, is such that $s(e)\in V''$ and $t(e)\in V'$ and $\tau(e)=trace$.
+Finally, the $Rulecop$ component in the 7-tuple contains the set of rule copies used in the construction of the path condition, where each rule copy is a subgraph of $\big\langle V,E,(s,t),\tau\big\rangle$.
+The set of all path conditions for a source metamodel $sr$ and a target metamodel $tg$ is called $\textsc{Pathcond}^{sr}_{tg}$ and the empty path condition is noted $\epsilon_{pc}$.
+
+\end{definition}
+
+\reviewer{Def. 18 suddenly introduces the notion of “rule copies” (Rulecop) which are re-
+ally not explained properly. My guess is that they are isomorphic representatives
+of rules. On page 13 under ”notation” this component is called Rule and referred
+to as a relation, and in Def. 28 it is denoted as Rules and required to be “the same
+as the set of transformation rules used to build [the transformation execution]”
+(and hence not isomorphic copies?). Needless to say, these inconsistencies do
+not help understandability.}
+
+Similarly to a transformation rule (see~\cref{def:transformation_rule}), a path condition is also a typed graph with a match and an apply part.
+As mentioned before, a path condition contains a combination of rules where \emph{symbolic traceability links} represent the concrete traceability links of a transformation execution (see \cref{def:modeltransformation}).
+The path condition structure also contains a $Rulcop$ set that allows identifying individually all the copies of rules that were used when building the path condition's typed graph.
+Note that we refer to \emph{copies} of rules as, despite the fact that a path condition normally only contains one copy of each rule, in certain situations a rule may be used multiple times in the construction of a path condition.
+This will be explained further ahead in this section.
+
+\paragraph{\textbf{Notation:}}
+Given a path condition $pc = \big\langle V,E,st,\tau,\\Match,Apply,Rulecop\big\rangle\in \textsc{Pathcond}^{sr}_{tg}$ we refer to the set of transformation rules in $pc$ identified by the $Rulecop$ relation as $Rulecop(pc)$.
+Also, because a path condition is a particular kind of a typed graph, we naturally extend the basic notation of operators and homomorphisms on typed graphs defined in \cref{sec:formal_background} to path conditions.
+
+\subsection{Abstraction Relation}
+\label{subsec:abstraction_relation}
+
 Let us start by formally defining the notion of abstraction of a transformation
@@
 \label{def:abstraction_pc_ex}
-\CatchFileBetweenTags{\abstractionrel}{text/definitions}{abstractionrel}{\abstractionrel}
+
+Let $tr\in \textsc{Transf}^{sr}_{tg}$ be a DSLTrans transformation. Let also $pc = \langle V,E,st,\tau,Match,Apply,Rules\rangle \in
+\textsc{Pathcond}(tr)$ of a path condition of $tr$ and $ex = \langle V',E',st',\tau',Input,Output\rangle \in Exec(tr)$ be an execution of $tr$.
+We have that $ex$ is abstracted by $pc$, noted $ex\Vvdash pc$, if and only if the set of transformation rules of $tr$ combined in $pc$ and the set of transformation rules of $tr$ used to build $ex$ is the same, and:
+\begin{multline}
+\label{eq:abstr_input_output}
+\big(\forall rl\in Rules\;.\;Match(rl) \vartriangleleft Input^*\big) \;\land\; Output \blacktriangleleft Apply
+\end{multline}
+\center{and}
+\begin{multline}
+\label{eq:abstr_trace}
+\big(\forall trc\in Components(pc|_{trace})\;.\;trc\vartriangleleft ex\big)\;\land\;\\
+\big(\forall e'\in E'\, \exists e\in E\;.\;\tau'(e')=trace \Longrightarrow \tau(e)=trace\big)
+\end{multline} 
+% where for all $rl,rl'\in Rules$, if $rl\cong rl'$ then there exist two injective typed graph homomorphisms $f$ and $g$ such that $Match(rl)\stackrel{f}{\vartriangleleft} Input^*$, $Match(rl)\stackrel{g}{\vartriangleleft} Input^*$ and $CoDom(f)\neq CoDom(g)$.
+% $/pc/ = Apply_{ex}\cup ex|_{trace}$ is the output part of $pc$, including symbolic traceability links.
+
+\reviewer{Definition 28: When you define the abstraction of an execution
+  by a path condition, you require two conditions (separated by an
+  and). I have the feeling that the morphisms used in both
+  conditions should somehow be compatible to each other. However,
+  your triangle notation indicates only the existence of a morphisms.
+  How do you express that these morphisms somehow "agree"?}
 \end{definition}
@@
 \end{proposition}
-\CatchFileBetweenTags{\pathgencompletenessproof}{text/definitions}{pathgencompletenessproof}{\pathgencompletenessproof}
+\begin{pf}
+Let $tr\in \textsc{Transf}^{sr}_{tg}$ be a DSLTans transformation.
+We wish to demonstrate that, for all transformation executions $ex\in \textsc{Exec}(tr)$, a path condition $pc\in \textsc{Pathcond}(tr)$ exists such that $ex$ is abstracted by $pc$, as formally expressed in \cref{def:abstraction_pc_ex}.\\
+
+Completeness can be shown as a corollary of \cref{prop:pc_validity} about the \emph{validity} of path condition generation.
+The complete set of executions $\textsc{Exec}(tr)$ (see \cref{def:modeltransformation}) can be split into two kinds of executions:
+\begin{enumerate}
+  \item\label{lab:input_not_matched} The \emph{empty execution} $\epsilon_{ex}$ or the \emph{execution where the input model was not matched by any rule}.
+As mentioned in \cref{prop:pc_validity}, these executions are abstracted by the empty path condition $\epsilon_{pc}$.\vspace{.3cm}
+  \item The \emph{execution $ex$ where a number of rules of $tr$ have been applied to the input model}, where each transformation rule $rl$ of $tr$ may have been applied more than once.
+In this case we have that, because all possible and valid rule combinations are considered when building path conditions, a path combination $pc$ exists that contains one or more copies of each of the rules used when operationally building $ex$.
+\\
+
+  Moreover, during the proof of \emph{validity} of path condition generation in \cref{prop:pc_validity} we demonstrate that, when we add a new rule $rl$ to the last layer of a transformation $tr$ (such that we have a new transformation $tr'$), the rule combination step explained in Definitions~\ref{def:rule_comb_no_dependencies},~\ref{def:rule_comb_unsatisfied} and~\ref{def:rul_comb_partial_total} produces a new set of path conditions where each path condition in that set still abstracts at least one transformation execution of $tr'$.
+This part of the proof (the second induction) is achieved by building for transformation $tr'$ an input model $m$ that can be matched by $rl$ (as well as by all the other rules of $tr$), and then building from $m$ a new transformation execution that is abstracted by a path condition built for $tr'$.
+Because in the proof of \cref{prop:pc_validity}, $m$ is such that it can be matched by $rl$ an arbitrary amount of times, we know that, independently of the number of times a rule is applied during the construction of a transformation execution, a path condition abstracting that transformation execution exists.\\
+
+Additionally, input elements that are not matched by any rule do not affect the abstraction relation, as explained in case~\ref{lab:input_not_matched} above.
+This means we also know that executions involving input models that are only partially matched by the rules of $tr$ are also abstracted by one path condition.
+\end{enumerate} 
+\end{pf}
 

### 2016-06-29 text/building_path_conditions.tex
Fix math in reviewer comment.
Place abstraction relation first.

@@
 
-This section will present how path conditions are structured to represent symbolic rule execution.
 As well, we present our approach to building a set of path conditions to represent all executions of a DSLTrans transformation.
-
-\reviewer{In Section 4, rather than starting first with the construction of path conditions
-(Section 4,4), it would be more natural to discuss their meaning in terms of
-transformation executions (Section 5), since that should be the justification for
-your constructions.}
-
-\subsection{Symbolic Execution}
-
-Our algorithm operates on the principle of symbolic execution to build up these
-path conditions. In order to explain the concept of symbolic execution of a
-transformation, let us make an analogy with program symbolic execution as
-introduced by King in his seminal work ``\emph{Symbolic Execution and Program
-Testing}''~\cite{DBLP:journals/cacm/King76}. According to King, a symbolic
-execution of a program is a set of \emph{constraints} on that program's
-\emph{input variables} called \emph{path conditions}. Each \emph{path condition}
-describes a traversal of the conditional branching commands of that program. A
-\emph{path condition} is symbolic in the sense it \emph{abstracts} as many
-concrete executions as there are instantiations of the path condition's
-variables that render the path condition's constraints true.
-
-We can transpose this notion of symbolic execution to model transformations. The
-analog of an input variable in the model transformation context are
-\emph{metamodel classes, relations and attributes}. As program statements impose
-constraints on input and output variables during symbolic execution,
-transformation rules impose conditions on which metamodel elements are
-instantiated during a concrete transformation execution, and how that
-instantiation happens. As well, rules in a model transformation are implicitly
-or explicitly scheduled. These control and/or data dependencies must be
-taken into consideration during path condition construction.
-
-As in program symbolic execution, each path condition in our approach \emph{abstracts} as many concrete executions as there are input/output models
-that satisfy them. This is formulated as an \emph{abstraction relation} as further explained in \cref{sec:abstraction_relation}.
-
-In what follows we will examine in more detail how these symbolic execution
-principles can apply to the verification of model transformations.
-
-\subsection{Path Conditions}
-\label{sec:gen_path_conds}
-
-In order to present the intuition of path conditions and symbolic executions, we
-first discuss the idea of \textit{rule combinations}.
-
-As seen in \cref{sec:dsltrans}, a layer in a DSLTrans transformation contains a
-number of rules.  We can create a set of rule combinations for this layer by
-taking the powerset of all rules in that layer. Each rule combination in this
-set will represent all possible transformation executions where the rules in
-that combination would execute.
-
-For example, in \cref{fig:rule_combos2}, the rule combination marked `AC'
-represents the set of transformation executions where the rules A and C would
-execute and no others. Another rule combination marked `A' represents the
-transformation executions where only rule A would execute.
-
-\begin{figure}[h!] \centering \includegraphics[width=.40\textwidth]{./figures/overview/rule_combos.pdf}
-	\caption{Rule combinations created for a transformation layer}
-	\label{fig:rule_combos2}
-\end{figure}
-
-Note that within these rule combinations, the number of times a rule has
-executed is abstracted. Either a rule has executed zero times, and the rule is
-not represented in a rule combination, or the rule has executed some finite
-number of times and it is represented. This abstraction is key to our approach,
-as it allows us to create a finite set of path conditions to abstract over an
-infinite set of transformation executions, as seen in
-\cref{sec:abstraction_relation}.
-
-We also note that rule \textit{combinations} are created, and not rule
-\textit{permutations}. This follows from the semantics of \\DSLTrans as described
-in \cref{sec:dsltrans}, as transformation rules in a layer will execute in a
-non-deterministic order but produce a deterministic result, by construction of the semantics of DSLTrans. As a final
-note, the transformation executions that these rule combinations represent always terminate, also by construction of the semantics of DSLTrans~\cite{DBLP:conf/sle/BarrocaLAFS10}.
-
-We base our concept of path conditions on these rule combinations. However, as
-DSLTrans allows for dependencies between rules, we cannot create path conditions
-for the transformation by taking the powerset of all rules. Instead, our
-approach must move layer-by-layer and resolve the dependencies between rules.
-The next two sections will introduce the concepts of traceability and dependency, before we
-briefly discuss the syntax and semantics of path conditions themselves.
-
-\subsubsection{Traceability}
-\label{subsubsec:traceability}
-
-DSLTrans rules specify which elements of the
-output model were created from specific elements of the input model. To resolve
-these dependencies, traceability information for the transformation is created
-during the execution of a DSLTrans model transformation~\cite{DBLP:conf/sle/BarrocaLAFS10}.
-In our verification approach, we store this same information as symbolic \emph{traceability links}, in
-order to record which elements belong to the same DSLTrans rule.
-
-At a particular point in the path condition construction process, symbolic traceability links are built for
-each rule as follows: for all match and apply elements of a rule, given a match
-element belonging to the match graph of a rule and an apply element belonging to
-the apply graph of the same rule, a symbolic traceability link is built between the two
-if the apply element is not connected to a backward link (as explained below).
-This is intuitive: traceability links are built between a newly generated
-element in the output model, and the elements of the input model that originated
-it.
-
-\begin{figure}[htb]
-        \centering
-        \begin{subfigure}[b]{0.24\textwidth}
-                \centering
-                \includegraphics[width=1\textwidth]{./figures/building_path_conditions/traceability_links.pdf}
-                \caption{Before symbolic traceability links added}
-                \label{fig:traceability_links1}
-        \end{subfigure}%
-        ~~
-        \begin{subfigure}[b]{0.235\textwidth}
-                \centering
-                \includegraphics[width=1\textwidth]{./figures/building_path_conditions/traceability_links2.pdf}
-                \caption{After symbolic traceability links added}
-                \label{fig:traceability_links2}
-        \end{subfigure}%
-        \caption{Symbolic traceability links created for an abstract DSLTrans rule}
-        \label{fig:trace_links}
-\end{figure}
-
-An example of the symbolic traceability link creation process is shown in
-\cref{fig:trace_links}. Note that symbolic traceability links are solid lines between
-match and apply elements in our visual notation.
-
-\subsubsection{Backward Links}
-\label{subsubsec:backward_links}
-
-The dependencies in a DSLTrans rule are specified using the \textit{backward link} construct, as further detailed in \cref{subsec:DSLTrans_constructs} and \cref{def:transformation_rule}.
-\cref{subsubsec:resolve_dependencies} will discuss how these dependencies are then resolved during our symbolic execution approach.
-
-\cref{fig:backward_links} demonstrates how backward links are used \\within a rule.
-The rule shown contains a backward link, which defines the dependency that an element of type X was created from an element of type A, and an element of type Y was created from an element of type B.
-If this dependency is satisfied, then another element of type Z should be created.
-This element should be associated with the Y element.
-
-\begin{figure}[htb]
-        \centering
-        \begin{subfigure}[b]{0.235\textwidth}
-                \centering
-                \includegraphics[width=1\textwidth]{./figures/building_path_conditions/backward_link.pdf}
-                \caption{Rule with backward links (dashed lines)}
-                \label{fig:backward_links}
-        \end{subfigure}%
-        ~~
-        \begin{subfigure}[b]{0.235\textwidth}
-                \centering
-                \includegraphics[width=1\textwidth]{./figures/building_path_conditions/backward_link2.pdf}
-                \caption{Traceability links added\\~}
-                \label{fig:backward_links2}
-        \end{subfigure}%
-        \caption{Adding traceability links to an abstract DSLTrans rule with backward links}
-        \label{fig:back_links}
-\end{figure}
-
-\cref{fig:backward_links2} shows the rule after symbolic traceability links have been added.
-Two symbolic traceability links are created from the Z element to the A and B elements in the match graph to store traceability information.
-Note that no symbolic traceability links are built between two elements connected by backward links, as these links have already been built in a previous layer.
-
-% \begin{definition}{Rule Traceability}
-% 
-% Let $rl = \big\langle V,E,st,\tau \big\rangle$ be a transformation
-% rule. $trace_{\langle
-% V_{\Delta},E_{\Delta},st_{\Delta},\tau_{\Delta}\rangle}(\langle
-% V,E,st,\tau\rangle) = \langle V,E',st',\tau'\rangle$ where we have that
-% $E\subseteq E'$, $st\subseteq st'$, $\tau \subseteq \tau'$ and if
-% $v_1\xrightarrow{e} v_2\in E'\setminus E$ then $v_1\in Output(V_{\Delta})$,
-% $v_2\in Input(V)$ and $\tau'(e)=trace$.
-% \end{definition}
-
-\subsubsection{Syntax and Semantics}
-\label{subsubsec:path_condition_creation}
-
-A path condition represents the symbolic execution of a set of DSLTrans rules, similar to a rule combination as explained above.
-Again, we use an abstraction over the number of times a rule has symbolically executed.
-Each path condition will represent that a rule has not executed, or has executed one or more times.
-
-The path condition generation algorithm will symbolically combine transformation rules into a path condition.
-Each path condition will then abstract a set of concrete transformation executions, as defined by our abstraction relation in \cref{sec:abstraction_relation}.
-
-As seen in the rest of this section, the structure of path conditions is similar to that of DSLTrans rules.
-The match graph of a path condition represents a pattern that must be present in the input model of the transformation, while the apply graph is a pattern which will be instantiated in the output model of the transformation.
-Symbolic traceability links are also kept between elements in the match and apply graphs to retain traceability information.
-
-The formal definition of a path condition is presented in \cref{def:path_condition}.
-
-\begin{definition}{Path Condition\\}
-\label{def:path_condition}
-\CatchFileBetweenTags{\pcdef}{text/definitions}{pcdef}{\pcdef}
-\end{definition}
-
-\reviewer{Def. 18 suddenly introduces the notion of “rule copies” (Rulecop) which are re-
-ally not explained properly. My guess is that they are isomorphic representatives
-of rules. On page 13 under ”notation” this component is called Rule and referred
-to as a relation, and in Def. 28 it is denoted as Rules and required to be “the same
-as the set of transformation rules used to build [the transformation execution]”
-(and hence not isomorphic copies?). Needless to say, these inconsistencies do
-not help understandability.}
-
-\CatchFileBetweenTags{\pc}{text/definitions}{pc}{\pc}
 
@@
 
-\reviewer{The right hand side of the conclusion of Def. 20 has $AC ∪ pc ′ ∈AC pc ′ $, of which
+\reviewer{The right hand side of the conclusion of Def. 20 has $AC \cup_{pc'\in AC} pc' $, of which
 I cannot make sense. AC is a set, but the pc' are not, so what operation is

### 2016-06-29 text/abstraction_proofs.tex
Move abstraction relation proofs into own section.

--- 2016-05-02 text/abstraction_relation.tex
@@
-\section{Abstraction Relation between Path Conditions and Transformation Executions}
-\label{sec:abstraction_relation}
-
-% More informally, there exists a typed graph injective homomorphism between
-% $Match_{pc}$ and $Input_x$  \textbf{and} $\langle V_x,E_x\setminus
-% E_{mx},\tau_x\rangle$ is a typed graph strict instance of $\langle
-% V_{pc},E_{pc}\setminus E_{pc},\tau_{pc}\rangle$.
-
-% \levi{It is possible that this relation is required to be a typed graph strict
-% instance between the path condition and the execution. Such a constraint would
-% oblige each set of executions for a path condition to be truly distinct from
-% the set of executions for another path condition. With the current definition
-% the executions for a path condition including rule 'A' will overlap with the
-% executions for a path condition including rules 'AB'. If this is so validity
-% needs to be proved again for the new relation between the match patterns of PC
-% and ex. Elements that do not show up in the match part of the rules are
-% excluded from the typed graph strict instance.}
-
-In this section we define the abstraction relation between the execution of a
-DSLTrans transformation and the path condition that represents it.
-This abstraction relation allows us to prove properties on a finite set of representative
-path conditions, as created by the path condition generation algorithm. As this set is finite, our technique is  guaranteed to be decidable.
-
-This section also presents our arguments that our path condition building
-algorithm is both \emph{valid} and \emph{complete}. In this context
-\emph{validity} means that for each path condition there exists at least one
-transformation execution that it abstracts.
-In other words, no path conditions are produced that lack a concrete
-transformation execution counterpart. \emph{Completeness} of the symbolic
-execution means that every transformation execution is abstracted by at least
-one path condition.
-
-Let us start by formally defining the notion of abstraction of a transformation
-execution by a path condition.
-
-\setcounter{equation}{0}
-
-\begin{definition}{Abstraction of a Transformation Execution by a Path Condition\\}
-\label{def:abstraction_pc_ex}
-\CatchFileBetweenTags{\abstractionrel}{text/definitions}{abstractionrel}{\abstractionrel}
-\end{definition}
-
-% \cref{def:abstraction_pc_ex} enforces the fact that the matcher part of
-% components of the path condition are found through an injective typed graph
-% homomorphism in the execution's input model. Additionally, the graph including
-% the output of the execution (including traceability information), can be
-% mapped by a surjective typed graph homomorphism onto the apply model (plus the
-% symbolic traceability links) of the path condition.
-
-To understand the abstraction relation in \cref{def:abstraction_pc_ex}, recall
-that during the construction of a transformation execution rules are matched
-injectively in the input model. This information is present in the first
-condition of the abstraction relation (Proposition~\ref{eq:abstr_input_output})
-via the injective typed graph homomorphism between the match part of the copies
-of rules ``glued'' onto the path condition and the containment transitive
-closure of the input part of the transformation execution. This relation
-enforces the fact that certain parts of the execution were found, or matched, by
-certain parts of the path condition.
-On the other hand, the surjection from the output of the execution towards the
-apply part of the path condition models the fact that the output of the
-execution has been completely built by instantiating the apply parts of the
-rules contained in the path condition.
-
-The second condition of the abstraction relation
-(Proposition~\ref{eq:abstr_trace}) checks for the fact that symbolic
-traceability links in the path condition and traceability links in the execution
-correctly correspond to each other. This is modeled by the fact that all
-strongly connected components in the path condition, composed only of symbolic
-traceability links, are injectively found on the execution. This injection
-models the fact that traceability graphs between individual or combined rules in
-the path condition are necessarily found in the execution. Note that components
-of the path condition are considered because of the fact that disconnected rules
-in the path condition may have matched over common elements of a particular
-execution. As such, a full injection between the complete traceability graph in
-the path condition and the execution would be incorrect. Additionally, in the
-second part of Proposition~\ref{eq:abstr_trace} we check the fact that every
-traceability link in the execution can be found in the path condition. This
-additional sanity check enforces that no spurious traceability links that could
-not have been created by the rules present in the path condition exist in the
-transformation execution.
-
-%Finally, the last clause of the abstraction relation states that rule copies
-%that are repeated a number of times in the path condition need to be found at
-%least a similar amount of times in the abstracted transformation execution.
-
-It is important to mention that another abstraction relation, weaker or
-stronger, could have been chosen. The abstraction relation presented in
-\cref{def:abstraction_pc_ex} suits our needs in the sense that it allows us to
-demonstrate the validity and completeness of our proof technique, as we will
-show in the text follows. Additionally, it is particularly interesting because
-it makes sure that, given a DSLTrans transformation, each of its transformation
-executions is abstracted by one and only one of its path conditions. This result
-adds to the consistency of our theory and is also exposed later in this section.
-
-\subsection{Examples}
-
-In this section, we provide a number of examples to demonstrate the workings of
-the abstraction relation we chose to use. \cref{fig:legend} presents the legend
-for the following figures.
-
-\begin{figure}[htb]
- \centering
-                \includegraphics[width=0.5\textwidth]{./figures/abstraction_relation/legend.pdf}
-                \caption{Legend for abstraction relation figures}
-                \label{fig:legend}
-\end{figure}
-
-\subsubsection{Example 1 -- Empty Path Condition}
-
-We begin by defining which transformation executions an empty path condition
-will abstract. \cref{fig:empty_pc} demonstrates two cases. In each, the path
-condition \textit{pc} is on the left-hand side, and a transformation execution \textit{ex} is on the
-right-hand side. Note that in \cref{fig:empty_pc_success}, the path condition
-abstracts the transformation execution, while in \cref{fig:empty_pc_fail}, the
-abstraction relation does not hold.
-
-\begin{figure}[htb]
-        \centering
-        \begin{subfigure}[b]{0.40\textwidth}
-                \centering
-                \includegraphics[width=1\textwidth]{./figures/abstraction_relation/empty_pc.pdf}
-               	\caption{Abstraction holds}
-               	\label{fig:empty_pc_success}
-        \end{subfigure}%
-        ~~\\
-        \begin{subfigure}[b]{0.40\textwidth}
-                \centering
-                \includegraphics[width=1\textwidth]{./figures/abstraction_relation/empty_pc2.pdf}
-                \caption{Abstraction does not hold}
-                \label{fig:empty_pc_fail}
-        \end{subfigure}%
-        \caption{Abstraction of transformation executions by the empty path condition}
-        \label{fig:empty_pc}
-\end{figure}
-
-The match part of the path condition represents the pre-conditions for the path condition to be true, depending on which rules have symbolically executed in the transformation.
-For example, if the match graph is empty, this represents all executions where no rules have executed.
-
-The first condition for the abstraction relation is to determine whether a typed graph injective homomorphism can be found between the match graph of the path condition, and a transformation execution.
-Note that in both \cref{fig:empty_pc_success} and \cref{fig:empty_pc_fail}, an empty typed graph homomorphism satisfies this condition, highlighted by blue arrows.
-
-The second condition for the abstraction relation is whether a typed graph surjective homomorphism can be found from the transformation execution's output model to the apply graph of the path condition.
-This is represented by orange arrows in \cref{fig:empty_pc_success} and \cref{fig:empty_pc_fail}.
-This relation is surjective as there may not be any elements in the output model that are not represented by the path condition's apply graph.
-Note that multiple elements in the output model may match to the same element in the apply graph of the path condition.
-This is expected, as the structure found in the apply graph may be found multiple times in the output model.
-
-The empty apply graph of the path condition defines no post-conditions on the output model, as no rules have executed.
-Note that there an empty surjective typed graph homomorphism can be found between the output model of the transformation execution in \cref{fig:empty_pc_success} and the path condition.
-This is intuitive, as the lack of elements in the output model means no rules have executed, which corresponds to the lack of post-conditions defined by the path condition.
-
-In contrast, there is no surjection between the elements of the output model in \cref{fig:empty_pc_fail} and the path condition.
-Note that the transformation execution has elements in the output model and thus at least one rule must have executed.
-However, the path condition does not represent that a rule has executed.
-Therefore, the path condition shown does not represent this execution.
-
-%The fact that this transformation execution will be abstracted by at least one path condition is demonstrated by \cref{prop:pc_completeness}.
-
-\subsubsection{Example 2 -- Non-overlapping Rule Components}
-
-This second example shows the abstraction relation between path conditions and transformation executions, when no match element of the same type appears in multiple rule components.
-
-For these examples, we will represent the abstraction relation with two figures.
-The first will demonstrate the matching performed on match and apply graphs, while the second figure focuses on traceability link matching.
-
-Let us first examine how the injection operates between the match elements in the path condition and the transformation executions in \cref{fig:non_overlapping_match_apply} and \cref{fig:non_overlapping2_match_apply}.
-Note that this injection can be found in both cases.
-
-\begin{figure}[htb]
-        \centering
-        \begin{subfigure}[b]{0.40\textwidth}
-                \centering
-                \includegraphics[width=1\textwidth]{./figures/abstraction_relation/non_overlapping.pdf}
-               	\caption{Abstraction holds on match and apply graphs}
-               	\label{fig:non_overlapping_match_apply}
-        \end{subfigure}%
-        ~~\\
-        \begin{subfigure}[b]{0.40\textwidth}
-                \centering
-                \includegraphics[width=1\textwidth]{./figures/abstraction_relation/non_overlapping_trace_links.pdf}
-                \caption{Abstraction holds on traceability links}
-                \label{fig:non_overlapping_trace_links}
-        \end{subfigure}%
-        \caption{Abstraction of transformation execution by non-overlapping rule components}
-        \label{fig:non_overlapping}
-\end{figure}
-
-\begin{figure}[htb]
-        \centering
-        \begin{subfigure}[b]{0.40\textwidth}
-                \centering
-                \includegraphics[width=1\textwidth]{./figures/abstraction_relation/non_overlapping2.pdf}
-               	\caption{Abstraction holds on match and apply graphs}
-               	\label{fig:non_overlapping2_match_apply}
-        \end{subfigure}%
-        ~~\\
-        \begin{subfigure}[b]{0.40\textwidth}
-                \centering
-                \includegraphics[width=1\textwidth]{./figures/abstraction_relation/non_overlapping2_trace_links.pdf}
-                \caption{Abstraction holds on traceability links}
-                \label{fig:non_overlapping2_trace_links}
-        \end{subfigure}%
-        \caption{Example of abstraction over multiple rule execution}
-        \label{fig:non_overlapping2}
-\end{figure}
-
-Similarly, there is a surjection between the elements of the output model for both transformation execution and the apply graph of the respective path condition.
-Note that this surjective match also holds in \cref{fig:non_overlapping2_match_apply}, where examination of the transformation execution shows that one rule has executed twice.
-As mentioned before, the abstraction relation abstracts over the number of times that a rule has executed.
-
-We also note that these matches must also match over associations between the elements, including association typing. This is not included in the figures for visual clarity.
-
-We now examine \cref{fig:non_overlapping_trace_links} and \cref{fig:non_overlapping2_trace_links} to resolve whether the traceability links in the path condition can be found in the transformation execution.
-This matching is represented by the arrows from each component highlighted in a bold outline and differentiated by colour.
-We note that each component in the path condition can be successfully found in the transformation execution.
-
-As well, there is a matching step from each individual traceability link in the transformation execution onto the path condition.
-Similar to the matching from the path condition, the bold components in the transformation execution figure are matched onto the path condition.
-We note that this matching is successful as well.
-
-\subsubsection{Example 3 -- Overlapping Rule Components}
-\label{subsubsec:abstraction_relation_overlapping}
-
-For these examples, the path conditions contain overlapping rule components, i.e.
-separate rules share match elements of the same type.
-Our goal is to illustrate the interaction of rule elements, where the elements of non-dependent rules may match over the same or different elements in the transformation execution.
-
-For example, the two rule components in \cref{fig:overlapping_match_apply} correctly match over the transformation execution shown.
-The abstraction relation holds due to the fact that, while match elements of the same component need to be found injectively in the execution, the injection constraint does not span multiple components.
-This allows the match elements from different rules to match to the same input model element.
-
-\begin{figure}[htb]
-        \centering
-        \begin{subfigure}[b]{0.40\textwidth}
-                \centering
-                \includegraphics[width=1\textwidth]{./figures/abstraction_relation/overlapping.pdf}
-               	\caption{Abstraction holds on match and apply graphs}
-               	\label{fig:overlapping_match_apply}
-        \end{subfigure}%
-        ~~\\
-        \begin{subfigure}[b]{0.40\textwidth}
-                \centering
-                \includegraphics[width=1\textwidth]{./figures/abstraction_relation/overlapping_trace_links.pdf}
-                \caption{Abstraction holds on traceability links}
-                \label{fig:overlapping_trace_links}
-        \end{subfigure}%
-        \caption{Abstraction of transformation execution by overlapping rule components}
-        \label{fig:overlapping}
-\end{figure}
-
-As well, \cref{fig:overlapping_trace_links} shows the mapping from the path condition to the transformation execution.
-However, note that the pattern composed of the A, B, and Y elements, along with the traceability links, is to be matched as a whole.
-This is to ensure that the traceability links are found in the proper configuration in the transformation execution.
-
-We also match the traceability links from the transformation execution back onto the path condition.
-Again, this is to ensure that no traceability links are found in the transformation execution that have not been represented in the path condition.
-Three matches are performed in this step, denoted by the three arrows in the bottom of \cref{fig:overlapping_trace_links}.
-Each match is composed of a traceability link as well as immediately connected elements.
-
-\begin{figure}[htb]
-        \centering
-        \begin{subfigure}[b]{0.40\textwidth}
-                \centering
-                \includegraphics[width=1\textwidth]{./figures/abstraction_relation/overlapping2.pdf}
-               	\caption{Elements cannot overlap within a component}
-               	\label{fig:overlapping2_match_apply}
-        \end{subfigure}%
-        ~~\\
-        \begin{subfigure}[b]{0.40\textwidth}
-                \centering
-                \includegraphics[width=1\textwidth]{./figures/abstraction_relation/overlapping2_trace_links.pdf}
-                \caption{Traceability links cannot be found}
-                \label{fig:overlapping2_trace_links}
-        \end{subfigure}%
-        \caption{Abstraction does not hold}
-        \label{fig:overlapping2}
-\end{figure}
-
-In contrast to \cref{fig:overlapping}, \cref{fig:overlapping2} shows an example where the abstraction relation does not hold.
-Consider \cref{fig:overlapping2_match_apply}.
-Note that a component in the match graph of the path condition contains two B elements.
-Both of these elements must be found in the transformation execution, and thus it is not correct for them to injectively match to the same element in the input model.
-
-As well, it is informative to examine \cref{fig:overlapping2_trace_links}.
-Note the one of the matches from the transformation execution attempts to match over 'a:A' and 'y:Y' elements, connected by a traceability link.
-Examination of the path condition shows that this traceability link is not present.
-Therefore, this path condition cannot accurately represent this transformation execution.
-
-\subsubsection{Example 4 -- Indirect Links}
-
-\begin{figure}[htb]
-        \centering
-        \begin{subfigure}[b]{0.40\textwidth}
-                \centering
-                \includegraphics[width=1\textwidth]{./figures/abstraction_relation/indirect.pdf}
-               	\caption{Matching over match and apply graphs}
-               	\label{fig:indirect_match_apply}
-        \end{subfigure}%
-        ~~\\
-        \begin{subfigure}[b]{0.40\textwidth}
-                \centering
-                \includegraphics[width=1\textwidth]{./figures/abstraction_relation/indirect_trace_links.pdf}
-                \caption{Matching over traceability links}
-                \label{fig:indirect_trace_links}
-        \end{subfigure}%
-        \caption{Abstraction with indirect links}
-        \label{fig:indirect}
-\end{figure}
-
-\begin{figure*}[htb]
-        \centering
-        \begin{subfigure}[b]{0.70\textwidth}
-                \centering
-                \includegraphics[width=1\textwidth]{./figures/abstraction_relation/combination.pdf}
-               	\caption{Abstraction holds on match and apply graphs}
-               	\label{fig:combination_match_apply}
-        \end{subfigure}%
-        ~~\\
-        \begin{subfigure}[b]{0.70\textwidth}
-                \centering
-                \includegraphics[width=1\textwidth]{./figures/abstraction_relation/combination_trace_links.pdf}
-                \caption{Abstraction holds on traceability links}
-                \label{fig:combination_trace_links}
-        \end{subfigure}%
-        \caption{Abstraction where path condition has combined rules}
-        \label{fig:combination}
-\end{figure*}
-
-We now present a path condition  in \cref{fig:indirect_match_apply} that includes indirect links.
-In this case, for the injective match to hold, the elements at both ends of the link must be found, and there must be an indirect link between the matched elements and between the elements in the transformation execution.
-
-Note that the indirect link between elements $a:A$ and $b:B$ in the transformation execution is added by the containment transitive closure $Input^{*}$ in proposition~\ref{eq:abstr_input_output} of \cref{def:abstraction_pc_ex} to allow matching indirect links.
-Note also that, for the sake of our example, we are assuming that the links between $a:A$, $b:b$ and $c:C$ are containment relations.
-
-\cref{fig:indirect_trace_links} highlights the structures involved in matching over traceability links.
-From the path condition, the structure contains the A, B, and Y elements with connected traceability links.
-From the transformation execution, there are two structures to be found in the path condition denoted in bold in the transformation execution.
-The matching of all structures can be successfully performed, and thus this abstraction relation holds.
-
-%Note that only the match graph of the path condition may include an indirect link.
-
-\subsubsection{Example 5 -- Combined Rules}
-
-\cref{fig:combination} shows a path condition that is composed of a multitude of rules which have been combined in the path condition generation algorithm.
-Each individual rule is surrounded by dashed lines.
-
-Note that the matching on the match and apply graphs in \cref{fig:combination_match_apply} is similar to other examples.
-The combined rules can be considered as a single graph for the abstraction relation.
-
-\cref{fig:combination_trace_links} shows the matching when multiple traceability links are present in the transformation execution.
-Note that each individual traceability link and the connected elements are matched onto the path condition.
-
-\subsection{Validity and Completeness}
+\section{Validity and Completeness}
 \label{subsec:abstraction_relation_validity_completeness}
@@
 
+   \item \emph{Base case:} the base case is the case when we have $tr=[\;]$, i.e.
+the empty transformation.
+In this case, according to \cref{def:path_cond_gen_appendix}, only the empty path condition $\epsilon_{pc}$ exists in the path condition set.
+The empty path condition abstracts the empty transformation execution $\epsilon_{ex}$ (see \cref{def:modeltransformation_appendix}), as well as any execution for which the input model is never matched by any rule (consequently having an empty output model).
+For any of these transformation executions, \cref{eq:abstr_input_output} of the abstraction relation definition is satisfied, as: a) no rule copy exists in the path condition and the output of the transformation execution is empty -- empty typed graph homomorphisms thus satisfy the all the conditions of the proposition; and b) \cref{eq:abstr_trace} of the abstraction relation definition trivially holds because no traceability links exist either in the path condition or in any of the considered executions.
+
 \item \emph{Inductive case:} assuming every path condition generated for a transformation $tr$ abstracts at least one transformation execution, we need to show that every path condition generated for a transformation $tr'$, resulting from adding a layer $l\in \textsc{Layer}^{sr}_{tg}$ to $tr$, will also abstract at least one transformation execution.
@@
 When a rule $rl$ is added to the last layer of $tr$ such that cases~\ref{lab:rule_case1} or~\ref{lab:rule_case3} occur, then the property can be shown to hold for $tr'$ as follows: 1) choose for a general path condition $pc$ generated for $tr$ an execution $ex$ such that $pc$ abstracts $ex$; 2) build an input model $m$ as the result of uniting the input model of $ex$ with a model that can be matched by $rl$; 3) execute $tr'$ having as input model $m$ to produce transformation execution $ex'$; and finally 4) demonstrate $ex'$ is abstracted by the path condition $pc'$ resulting from combining $pc$ with $rl$ whether rule $rl$ does not depend on $pc$ or rule $rl$ depends on $pc$ and may and/or will execute.
+
+The property trivially holds for case~\ref{lab:rule_case2_appendix}, given that no new path conditions are added to the path condition set generated for $tr$ and that the property holds for $tr$ by induction hypothesis.
+
+When a rule $rl$ is added to the last layer of $tr$ such that cases~\ref{lab:rule_case1_appendix} or~\ref{lab:rule_case3_appendix} occur, new path conditions are added to the path condition set.
+Both cases are based on combining a path condition with a rule, as laid out in \cref{def:combine_pc_with_rule_appendix}.
+In order to demonstrate this second inductive step we then need to show that, whenever the property holds for a path condition $pc$ generated for a transformation $tr$, the combination of $pc$ with a rule $rl$ results in a new path condition where the property is respected.
+
+We start by picking for $pc$ an execution $ex$ such that $pc$ abstracts $ex$.
+We know such a transformation execution exists by induction hypothesis.
+We can then build an input model $m$ as the result of uniting the input model of $ex$ with a model that can be matched by $rl$.
+If we execute $tr'$ having $m$ as input model we obtain transformation execution $ex'$.
+
+Let us now demonstrate $ex'$ is abstracted by the path condition $pc'=pc\stackrel{trace}{\sqcup} rl$, the combination of $pc$ with $rl$ as shown in \cref{def:combine_pc_with_rule_appendix}.
+We first recall the conditions of the abstraction relation in \cref{def:abstraction_pc_ex_appendix}:
+\begin{enumerate}
+	\item\label{item:abstr_rel1} a) injective typed graph homomorphisms must exist between the match parts of all the rule copies in the path condition and the input of the execution \emph{and} b) a surjective typed graph isomorphism must exist between the output of the execution and the apply part of the path condition.
+    \item\label{item:abstr_rel2} a) injective typed graph homomorphisms must exist between all strongly connected components of the path condition composed of only symbolic traceability links \emph{and} b) all isolated traceability links in the transformation execution must be found at least once in the path condition.\vspace{.3cm}
+\end{enumerate}
+
+Let us start by arguing for why condition~\ref{item:abstr_rel1} a) holds for $pc'$ and $ex'$.
+Because we know rule $rl$ has executed on the input model of $ex'$, we know by \cref{def:match_function} of the function matching a DSLTrans rule that an injective typed graph homomorphism exists between the match part of $rl$ and $ex'$.
+When $rl$ is combined with $pc$, its match part is preserved in $pc$ and as such an injective typed graph homomorphism must exist between it and $ex'$.
+By induction hypothesis and because the combination operator is additive (meaning nothing existing in $pc$ is deleted during combination) we know injective typed graph homomorphisms continue to exist between the match parts of all other rule copies in the path condition and the input of the $ex'$.\vspace{.3cm}
+
+In what concerns condition~\ref{item:abstr_rel1} b) above, we know by \cref{def:apply_function} and \cref{def:layer_step_semantics} that one or more copies of graphs isomorphic to the apply part of $rl$ are added to the output of $ex$.
+Also, by \cref{def:layer_step_semantics}, we know this addition preserves the output of $ex$ and we also know by hypothesis that a surjective typed graph isomorphism exists between the output of $ex$ and the apply part of $pc$.
+As mentioned before, the combination of $pc$ and $rl$ is additive and as such we can also deduce that a typed graph isomorphism exists between the apply part of any copy $rl$ added to $ex$ and the apply part of $rl$ added to $pc$.
+As such, all old and new edges and nodes in $ex'$ can be surjectively found in $pc'$.\vspace{.3cm}
+
+We will now discuss the reasons why condition~\ref{item:abstr_rel2} a) of the abstraction relation holds for $pc'$ and $ex'$.
+When $pc$ and $rl$ are combined, by \cref{def:combine_pc_with_rule_appendix} a copy of the rule is ``glued'' on top of $pc$.
+Symbolic traceability links are added between elements of the match part of the copy of the rule and of the apply copy of the rule, for those elements in the apply part of the copy of the rule not previously connected to backward links.
+We also know by \cref{def:apply_function} that traceability links are similarly added to the copy of $rl$ that is merged with $ex$.
+Because of the induction hypothesis we know that injective typed graph homomorphisms exists between all the strongly connected components composed of traceability links of $pc$ and $ex$.
+When $rl$ is combined with $pc$ two cases can occur: a) $rl$ has no backward links, in which the proposition trivially holds because isomorphic isolated strongly connected components are added both to $pc$ and $ex$; b) $rl$ has backward links, in which case the newly added components will connect to existing strongly connected components in $pc$ and $ex$, forming additional strongly connected components.
+In this case, an injective typed graph homomorphism exists between each of the newly formed strongly connected components in $pc$ and at least one newly formed strongly connected component in $ex$.
+This is so because, by \cref{def:abstraction_pc_ex_appendix} of the abstraction relation between a path condition and a transformation execution, the set of rules combined into $pc$ and the set of rules that have executed is the same.
+The combination of these rules in the path condition, according to \cref{def:combine_pc_with_rule_appendix}, replicates patterns that are produced in the execution by \cref{def:apply_function} and \cref{def:layer_step_semantics}.
+Note that condition~\ref{item:abstr_rel2} a) of the abstraction relation provides additional guarantee that, when multiple partially and/or totally satisfied dependencies occur during path condition combination (\cref{def:rul_comb_partial_total_appendix}), the corresponding executions are correctly abstracted given each place where the rules are ``glued'' corresponds to a different strongly connected component.\vspace{.3cm}
+
+Condition~\ref{item:abstr_rel2} b) of the abstraction relation trivially holds as each new traceability link added to $ex$ when $rl$ executes has at least one corresponding symbolic traceability link in $pc'$, resulting from the combination of $pc$ with $rl$.
+
 \end{itemize}
@@
 \end{proposition}
-\CatchFileBetweenTags{\pathgencompletenessproof}{text/definitions}{pathgencompletenessproof}{\pathgencompletenessproof}
+\begin{pf}
+Let $tr\in \textsc{Transf}^{sr}_{tg}$ be a DSLTans transformation.
+We wish to demonstrate that, for all transformation executions $ex\in \textsc{Exec}(tr)$, a path condition $pc\in \textsc{Pathcond}(tr)$ exists such that $ex$ is abstracted by $pc$, as formally expressed in \cref{def:abstraction_pc_ex}.\\
+
+Completeness can be shown as a corollary of \cref{prop:pc_validity} about the \emph{validity} of path condition generation.
+The complete set of executions $\textsc{Exec}(tr)$ (see \cref{def:modeltransformation}) can be split into two kinds of executions:
+\begin{enumerate}
+  \item\label{lab:input_not_matched} The \emph{empty execution} $\epsilon_{ex}$ or the \emph{execution where the input model was not matched by any rule}.
+As mentioned in \cref{prop:pc_validity}, these executions are abstracted by the empty path condition $\epsilon_{pc}$.\vspace{.3cm}
+  \item The \emph{execution $ex$ where a number of rules of $tr$ have been applied to the input model}, where each transformation rule $rl$ of $tr$ may have been applied more than once.
+In this case we have that, because all possible and valid rule combinations are considered when building path conditions, a path combination $pc$ exists that contains one or more copies of each of the rules used when operationally building $ex$.
+\\
+
+  Moreover, during the proof of \emph{validity} of path condition generation in \cref{prop:pc_validity} we demonstrate that, when we add a new rule $rl$ to the last layer of a transformation $tr$ (such that we have a new transformation $tr'$), the rule combination step explained in Definitions~\ref{def:rule_comb_no_dependencies},~\ref{def:rule_comb_unsatisfied} and~\ref{def:rul_comb_partial_total} produces a new set of path conditions where each path condition in that set still abstracts at least one transformation execution of $tr'$.
+This part of the proof (the second induction) is achieved by building for transformation $tr'$ an input model $m$ that can be matched by $rl$ (as well as by all the other rules of $tr$), and then building from $m$ a new transformation execution that is abstracted by a path condition built for $tr'$.
+Because in the proof of \cref{prop:pc_validity}, $m$ is such that it can be matched by $rl$ an arbitrary amount of times, we know that, independently of the number of times a rule is applied during the construction of a transformation execution, a path condition abstracting that transformation execution exists.\\
+
+Additionally, input elements that are not matched by any rule do not affect the abstraction relation, as explained in case~\ref{lab:input_not_matched} above.
+This means we also know that executions involving input models that are only partially matched by the rules of $tr$ are also abstracted by one path condition.
+\end{enumerate}
+
+Let $tr\in \textsc{Transf}^{sr}_{tg}$ be a DSLTans transformation.
+We wish to demonstrate that, for all transformation executions $ex\in \textsc{Exec}(tr)$, a path condition $pc\in \textsc{Pathcond}(tr)$ exists such that $ex$ is abstracted by $pc$, as formally expressed in \cref{def:abstraction_pc_ex_appendix}.
+Completeness can be shown as a corollary of \cref{prop:pc_validity_appendix} about the \emph{validity} of path condition generation.
+The complete set of executions $\textsc{Exec}(tr)$ (see \cref{def:modeltransformation_appendix}) can be split into two kinds of executions:
+\begin{enumerate}
+  \item\label{lab:input_not_matched} The \emph{empty execution} $\epsilon_{ex}$ or the \emph{execution where the input model was not matched by any rule}.
+As mentioned in \cref{prop:pc_validity_appendix}, these executions are abstracted by the empty path condition $\epsilon_{pc}$.\vspace{.3cm}
+  \item The \emph{execution $ex$ where a number of rules of $tr$ have been applied to the input model}, where each transformation rule $rl$ of $tr$ may have been applied more than once.
+In this case we have that, because all possible and valid rule combinations are considered when building path conditions, a path combination $pc$ exists that contains one or more copies of each of the rules used when operationally building $ex$.
+Moreover, during the proof of \emph{validity} of path condition generation in \cref{prop:pc_validity_appendix} we demonstrate that, when we add a new rule $rl$ to the last layer of a transformation $tr$ (such that we have a new transformation $tr'$), the rule combination step explained in Definitions~\ref{def:rule_comb_no_dependencies_appendix},~\ref{def:rule_comb_unsatisfied_appendix} and~\ref{def:rul_comb_partial_total_appendix} produces a new set of path conditions where each path condition in that set still abstracts at least one transformation execution of $tr'$.
+This part of the proof (the second induction) is achieved by building for transformation $tr'$ an input model $m$ that can be matched by $rl$ (as well as by all the other rules of $tr$), and then building from $m$ a new transformation execution that is abstracted by a path condition built for $tr'$.
+Because in the proof of \cref{prop:pc_validity_appendix}, $m$ is such that it can be matched by $rl$ an arbitrary amount of times, we know that, independently of the number of times a rule is applied during the construction of a transformation execution, a path condition abstracting that transformation execution exists.\\
+
+Additionally, input elements that are not matched by any rule do not affect the abstraction relation, as explained in case~\ref{lab:input_not_matched} above.
+This means we also know that executions involving input models that are only partially matched by the rules of $tr$ are also abstracted by one path condition.
+\end{enumerate} 
+\end{pf}
 
@@
 As such, we have that $ex_1$ is always different from $ex_2$ by at least one traceability link.
-Given that this traceability link is symbolically represented in either $pc_1$ or $pc_2$ (but not in both), according to condition \cref{eq:abstr_trace} in \cref{def:abstraction_pc_ex} it cannot be that either $pc_1$ or $pc_2$ abstract $ex_1$ and $ex_2$ simultaneously.\\\\
-We will now analyse the scenario where $tr$ falls into case~\ref{item:uniqueness_rule_has_dep} above, where some rules in $tr$ have dependencies.
+Given that this traceability link is symbolically represented in either $pc_1$ or $pc_2$ (but not in both), according to condition \cref{eq:abstr_trace} in \cref{def:abstraction_pc_ex_appendix} it cannot be that either $pc_1$ or $pc_2$ abstract $ex_1$ and $ex_2$ simultaneously.\\\\
+We will now analyse the scenario where $tr$ falls into case~\ref{item:uniqueness_rule_has_dep_appendix} above, where some rules in $tr$ have dependencies.
 For this case, assume we have a path condition $pc_1$ contained in the set of path conditions generated for $tr$, considering layers up to layer $l$ of $tr$ have executed.
 Assume also we have a rule $rl$ of layer $l+1$ of $tr$ that has dependencies and can be combined with $pc$.
-If rule $rl$ is totally combined with path condition $pc_1$, according to \cref{def:rul_comb_partial_total} and \cref{fig:multiple_total_satisfied_dependencies}, then nothing needs to be shown as $pc_1$ is not kept in the path condition set but rather replaced by its combination with $rl$.
-However, in case rule $rl$ is partially combined with $pc$, as defined in \cref{def:rul_comb_partial_total} and \cref{def:comb_path_cond_rule_single}, then multiple path conditions are generated and additionally $pc_1$ is kept in the path condition set.\\The proof will be complete once it is shown that for the path conditions that are generated when rules are partially combined, it is also the case that no two path conditions can abstract the same execution.
-This last part of the proof can be built in an analog fashion to the construction of the proof for point~\ref{item:uniqueness_rule_no_dep}.
-As previously mentioned, the complete proof can be found as \cref{lem:uniqueness_appendix} in \cref{sec:val_complete_path_cond_gen}.
+If rule $rl$ is totally combined with path condition $pc_1$, according to \cref{def:rul_comb_partial_total_appendix} and \cref{fig:multiple_total_satisfied_dependencies}, then nothing needs to be shown as $pc_1$ is not kept in the path condition set but rather replaced by its combination with $rl$.
+However, in case rule $rl$ is partially combined with $pc$, as defined in \cref{def:rul_comb_partial_total_appendix} and \cref{def:comb_path_cond_rule_single_appendix}, then multiple path conditions are generated and additionally $pc_1$ is kept in the path condition set.
+Consider $pc_2$ is one of the newly created path conditions.
+In this case we can find a model $m$ that can be injectively matched by the rule copies in both $pc_1$ and $pc_2$: $m$ is the union of the input model isomorphic to the match part of $pc_1$, united with the input model isomorphic to the match part of $pc_2$ (including symbolic traceability links).\vspace{.3cm}
+
+As before, we now consider that $ex_1$ and $ex_2$ are obtained by executing the rules used to build $pc_1$ and $pc_2$, respectively, having $m$ as input model.
+In this case, we have that either $rl$ was ``glued'' across different rule copies in $pc_2$, or over one single rule copy of $pc_2$.
+In the first case, by \cref{def:transformation_rule_appendix} of transformation rule we know either a new edge between output elements or a new output element have been produced in $ex_2$, but not in $ex_1$.
+According to the second part of Proposition~\ref{eq:abstr_trace} in \cref{def:abstraction_pc_ex_appendix} or the second part of Proposition~\ref{eq:abstr_trace} in \cref{eq:abstr_input_output}, this makes it such that it cannot be that either $pc_1$ or $pc_2$ abstract $ex_1$ and $ex_2$ simultaneously.\vspace{.3cm}
+
+Finally, let us consider an additional path condition $pc_3$, also obtained from the partial combination of $pc_1$ with $rl$ and where $pc_3$ is different of $pc_1$.
+In this case we have that $pc_2$ and $pc_3$ resulted from the combination of exactly the same rules, with the difference that certain rules have been ``glued'' at more locations of one path condition than of the other.
+We can thus build a model $m$ that can be injectively matched by the rule copies in both $pc_1$ and $pc_2$: the model is isomorphic to the the match part of the path condition (including symbolic traceability links) that has been ``glued'' more copies of $rl$ upon.
+When we now obtain $ex_2$ and $ex_3$ by executing the rules in $pc_2$ and $pc_3$, we will have that one of these executions will necessarily contain more copies of $rl$'s apply pattern than the other.
+Given the fact that these copies will necessarily have been ``glued'' over different strongly connected graphs of $pc_2$ and $pc_3$ (because rules having no dependencies do not overlap as explained for case~\ref{item:uniqueness_rule_no_dep_appendix}), there cannot be an injective typed graph homomorphisms between all the strongly connected components formed by the traceability graphs of at least one of path conditions $pc_2$ or $pc_3$, and $ex_1$ (likewise for $ex_2$).
+Given this is required by the first part of Proposition~\ref{eq:abstr_trace} in \cref{eq:abstr_input_output} of the abstraction relation, it cannot be that either $pc_1$ or $pc_2$ abstract $ex_1$ and $ex_2$ simultaneously.
 

### 2016-06-29 text/abstraction_relation.tex
Move abstraction relation proofs into own section.

@@
 
-%Finally, the last clause of the abstraction relation states that rule copies
-%that are repeated a number of times in the path condition need to be found at
-%least a similar amount of times in the abstracted transformation execution.
+Finally, the last clause of the abstraction relation states that rule copies
+that are repeated a number of times in the path condition need to be found at
+least a similar amount of times in the abstracted transformation execution.
 
@@
 Note that each individual traceability link and the connected elements are matched onto the path condition.
-
-\subsection{Validity and Completeness}
-\label{subsec:abstraction_relation_validity_completeness}
-
-In this section, we discuss the validity and completeness of the abstraction relation.
-Only proof sketches are presented here for \cref{prop:pc_validity} and \cref{lem:uniqueness}, while more complete proofs are found in \cref{sec:val_complete_path_cond_gen}, in the proofs of \cref{prop:pc_validity_appendix} and \cref{lem:uniqueness_appendix} respectively.
-
-%\newpage
-\begin{proposition}{(Validity)}
-\label{prop:pc_validity}
-Every path condition abstracts at least one transformation execution.
-\end{proposition}
-\begin{ps}
-
-Let $tr\in \textsc{Transf}^{sr}_{tg}$ be a DSLTans transformation.
-We wish to demonstrate that, for all path conditions $pc\in \textsc{Pathcond}(tr)$, there exists a transformation execution $ex\in \textsc{Exec}(tr)$ of the set of
-rules used to build $pc$ such that $pc$ abstracts $ex$, as formally expressed in \cref{def:abstraction_pc_ex}.
-We can prove this property by induction on the set of transformations $\textsc{Transf}^{sr}_{tg}$ (see \cref{def:layer_transformation}), as follows:
-
-\begin{itemize}
-  \item \emph{Base case:} the base case is when $tr$ is the empty
-  transformation. In this case, according to \cref{def:path_cond_gen} only the
-  empty path condition $\epsilon_{pc}$ exists in the path condition set. We thus
-  need to demonstrate that the empty path condition abstracts the empty
-  transformation execution $\epsilon_{ex}$, as well as any execution for which
-  the input model is never matched by any rule (consequently having an empty
-  output model).
-For any of these transformation executions, Proposition~\ref{eq:abstr_input_output} of the abstraction relation definition is satisfied, as no rule copy exists in the path condition and the output of the transformation execution is empty -- empty typed graph homomorphisms thus satisfy all the conditions of the proposition.
-Proposition~\ref{eq:abstr_trace} of the abstraction relation definition also trivially holds because no traceability links exist either in the path condition or in any of the considered executions.
-
-%   This is so because:
-%   1) the empty typed graph homomorphism is an injective typed graph homomorphism
-%   between the match part of the empty path condition and any input model; and 2) as
-%   well as a surjective typed graph homomorphism between an empty output model
-%   and the apply part of the empty path condition.
-
-\item \emph{Inductive case:} assuming every path condition generated for a transformation $tr$ abstracts at least one transformation execution, we need to show that every path condition generated for a transformation $tr'$, resulting from adding a layer $l\in \textsc{Layer}^{sr}_{tg}$ to $tr$, will also abstract at least one transformation execution.
-\end{itemize}
-
-In order to demonstrate the inductive case we need to show the property holds for all path conditions resulting from combining the rules of layer $l$ with any path condition generated for $tr$.
-These path conditions for transformation $tr'$ are built as expressed in \cref{def:path_cond_layer_comb}.
-According to this definition, path conditions for $tr'$ are built by incrementally combining the path conditions generated for $tr$ with a rule of layer $l$, until all the rules in $l$ have been treated.
-We can thus again use induction for this proof, this time on the set of possible layers $\textsc{Layer}^{sr}_{tg}$.
-
-\begin{itemize}
-  \item \emph{Base case:} this is the case where layer $l$ contains no rules.
-In this case, by the base case of \cref{def:path_cond_layer_comb}, no new path condition is added to the set of path conditions generated for the transformation $tr$.
-As such the $tr=tr'$ and by induction hypothesis the property trivially holds for all path conditions generated for $tr'$.
-
-  \item \emph{Inductive case:} for the inductive case (transitive case of \cref{def:path_cond_layer_comb}) we need to show that, assuming the property holds for all path conditions generated for a transformation $tr$, then the property will also hold for a transformation $tr'$ -- where $tr'$ results from adding a new rule $rl$ to the last layer of $tr$.
-We will thus need to consider the four \reviewer{three?} cases of rule combination:\vspace{.2cm}
-
-\begin{enumerate}
-\item\label{lab:rule_case1} Rule $rl$ has no dependencies (\cref{def:rule_comb_no_dependencies}).
-\item\label{lab:rule_case2} Rule $rl$ has dependencies and cannot execute (\cref{def:rule_comb_unsatisfied}).
-\item\label{lab:rule_case3} Rule $rl$ has dependencies and may and/or will execute (\cref{def:rul_comb_partial_total}).
-\end{enumerate}
-
-The property trivially holds for case~\ref{lab:rule_case2}, given that no new path conditions are added to the path condition set generated for $tr$ and that the property holds for $tr$ by induction hypothesis.
-When a rule $rl$ is added to the last layer of $tr$ such that cases~\ref{lab:rule_case1} or~\ref{lab:rule_case3} occur, then the property can be shown to hold for $tr'$ as follows: 1) choose for a general path condition $pc$ generated for $tr$ an execution $ex$ such that $pc$ abstracts $ex$; 2) build an input model $m$ as the result of uniting the input model of $ex$ with a model that can be matched by $rl$; 3) execute $tr'$ having as input model $m$ to produce transformation execution $ex'$; and finally 4) demonstrate $ex'$ is abstracted by the path condition $pc'$ resulting from combining $pc$ with $rl$ whether rule $rl$ does not depend on $pc$ or rule $rl$ depends on $pc$ and may and/or will execute.
-\end{itemize}
-\end{ps}
-
-\begin{proposition}{(Completeness)}
-\label{prop:pc_completeness}
-Every transformation execution is abstracted by one path condition.
-\end{proposition}
-\begin{pf}
-Let $tr\in \textsc{Transf}^{sr}_{tg}$ be a DSLTans transformation.
-We wish to demonstrate that, for all transformation executions $ex\in \textsc{Exec}(tr)$, a path condition $pc\in \textsc{Pathcond}(tr)$ exists such that $ex$ is abstracted by $pc$, as formally expressed in \cref{def:abstraction_pc_ex}.\\
-
-Completeness can be shown as a corollary of \cref{prop:pc_validity} about the \emph{validity} of path condition generation.
-The complete set of executions $\textsc{Exec}(tr)$ (see \cref{def:modeltransformation}) can be split into two kinds of executions:
-\begin{enumerate}
-  \item\label{lab:input_not_matched} The \emph{empty execution} $\epsilon_{ex}$ or the \emph{execution where the input model was not matched by any rule}.
-As mentioned in \cref{prop:pc_validity}, these executions are abstracted by the empty path condition $\epsilon_{pc}$.\vspace{.3cm}
-  \item The \emph{execution $ex$ where a number of rules of $tr$ have been applied to the input model}, where each transformation rule $rl$ of $tr$ may have been applied more than once.
-In this case we have that, because all possible and valid rule combinations are considered when building path conditions, a path combination $pc$ exists that contains one or more copies of each of the rules used when operationally building $ex$.
-\\
-
-  Moreover, during the proof of \emph{validity} of path condition generation in \cref{prop:pc_validity} we demonstrate that, when we add a new rule $rl$ to the last layer of a transformation $tr$ (such that we have a new transformation $tr'$), the rule combination step explained in Definitions~\ref{def:rule_comb_no_dependencies},~\ref{def:rule_comb_unsatisfied} and~\ref{def:rul_comb_partial_total} produces a new set of path conditions where each path condition in that set still abstracts at least one transformation execution of $tr'$.
-This part of the proof (the second induction) is achieved by building for transformation $tr'$ an input model $m$ that can be matched by $rl$ (as well as by all the other rules of $tr$), and then building from $m$ a new transformation execution that is abstracted by a path condition built for $tr'$.
-Because in the proof of \cref{prop:pc_validity}, $m$ is such that it can be matched by $rl$ an arbitrary amount of times, we know that, independently of the number of times a rule is applied during the construction of a transformation execution, a path condition abstracting that transformation execution exists.\\
-
-Additionally, input elements that are not matched by any rule do not affect the abstraction relation, as explained in case~\ref{lab:input_not_matched} above.
-This means we also know that executions involving input models that are only partially matched by the rules of $tr$ are also abstracted by one path condition.
-\end{enumerate} 
-\end{pf}
-
-% Our path conditions construction algorithm takes all possible interactions of rules
-% into account:
-% \begin{enumerate}
-% \item When building the path conditions for one layer we have built all
-% the combinations of rules, as well as their possible interactions through
-% \emph{disambiguation};
-% \item When combining the path conditions from different layer different layers
-% we have considered the cases where no interactions exist, or where backward
-% links in rules requires certain certain elements to already exist in the
-% generated model thus far.
-% \end{enumerate}
-% Note that, given the considered subset of DSLTrans constructs as described
-% in \cref{sec:dsltrans}, no additional cases of rule interaction other than the
-% ones described above need to be considered.
-% From \cref{prop:pc_validity} we also know that at least one concrete
-% transformation execution exists per path condition. In the proof of
-% \cref{prop:pc_validity} we have built, for each path condition $A$,
-% the simplest transformation execution that is abstracted by $A$. These
-% transformation executions produce one instance of each element and association
-% present in the apply pattern of $A$. However, given
-% \cref{def:instance_pc_ex} in \cref{sec:proofs} (abstraction of
-% a transformation execution by a path condition), an infinite amount of
-% transformation executions are always abstracted by $A$. These executions
-% correspond to the generation of output models holding an arbitrarily large
-% amount of the elements and associations while still being abstracted by $A$.
-% Because we have considered all possibilities of execution in our path condition
-% construction algorithm, the union of all transformation executions built for
-% each path condition constitutes the complete, infinite set of transformation
-% executions. We thus know that every possible transformation execution is
-% abstracted by at least one path condition.
-
-% Note that we do not provide any guarantee that the same transformation execution
-% cannot be abstracted by two different path conditions. Although given our symbolic
-% execution assumptions we believe this to be the case, further mathematical
-% exploration of this issue is necessary. Given however the fact that our
-% verification algorithm exhaustively explores all path conditions, it is
-% sufficient for the property proof to provide a \emph{no} result in the analysis
-% of a path condition for the result of the verification of the property to be
-% \emph{no}. \levi{keep this?}
-
-\begin{lemma} (Uniqueness) 
-\label{lem:uniqueness}
-A transformation execution is abstracted by exactly one path condition.
-\end{lemma}
-\begin{ps}
-Let $tr\in \textsc{Transf}^{sr}_{tr}$ be a model transformation.
-We will demonstrate that two different path conditions\\$pc_1, pc_2\in \textsc{Pathcond}(tr)$ cannot exist such that we have a transformation execution $ex\in \textsc{Exec}(tr)$ where $ex\Vvdash pc_1$ and $ex\Vvdash pc_2$.
-
-We will do so by attempting to build an $ex\in \textsc{Exec}(tr)$ such that $ex\Vvdash pc_1$ and $ex\Vvdash pc_2$ and demonstrating that it is always the case that such is not possible.
-In order to structure our argumentation, we will consider two cases:
-% \levi{the contradiction happens because at each point of the execution, when a rule is added any two path conditions from the existing set of path conditions will represent different executions.
-At all points if we consider two path conditions in the PC set we have that two different executions can be created if we find a model that satisfies the first condition of the abstraction relation, and it is always the case that none of those two executions can be abstracted by the two path conditions simultaneously.}
-\begin{enumerate}
-  \item\label{item:uniqueness_rule_no_dep} the case where no rules in $tr$ have dependencies.
-  \item\label{item:uniqueness_rule_has_dep} the case where some rules in $tr$ have dependencies.
-\end{enumerate}
-
-We start by considering that $tr$ falls into case~\ref{item:uniqueness_rule_no_dep} above.
-By~\cref{def:path_cond_gen} of path condition generation, each rule appears at most once in a path condition.
-Also, by construction, each path condition always contains a different combination of rules.
-We additionally know from \cref{def:layer_transformation} that the rules that compose $tr$ necessarily have non-overlapping matchers.
-We can nonetheless build a model $m$ as the typed graph union of two input models $m_1$ and $m_2$, where injective typed graph morphisms can be found between the match parts of the rule copies that form $pc_1$, and $m_1$.
-Injective typed graph morphisms can be found as well between the match parts of the rules that form $pc_2$, and $m_2$.
-We thus know that injective typed graph morphisms can be found between the rule copies that compose $pc_1$ and $pc_2$, and $m$.
-This satisfies the first condition of \cref{eq:abstr_input_output} in ~\cref{def:abstraction_pc_ex} of abstraction relation.
-
-Let us now consider that $ex_1$ and $ex_2$ are obtained by executing the transformations rules combined into $pc_1$ and $pc_2$, having $m$ as input model.
-As mentioned above, we know that the rules in $pc_1$ and $pc_2$ are not completely overlapping.
-This means that, due to the way in which $m$ is built (explained above), $m$ will always have at least one input that is matched by rules of $pc_1$ but not by rules of $pc_2$ (and vice-versa).
-Thus, when the transformation rules combined into $pc_1$ execute having $m$ as an input model, there will always exist a traceability link generated between an input and an output element of $m$ that is not generated when the transformation rules combined into $pc_2$ execute having $m$ as an input model (and vice-versa).
-As such, we have that $ex_1$ is always different from $ex_2$ by at least one traceability link.
-Given that this traceability link is symbolically represented in either $pc_1$ or $pc_2$ (but not in both), according to condition \cref{eq:abstr_trace} in \cref{def:abstraction_pc_ex} it cannot be that either $pc_1$ or $pc_2$ abstract $ex_1$ and $ex_2$ simultaneously.\\\\
-We will now analyse the scenario where $tr$ falls into case~\ref{item:uniqueness_rule_has_dep} above, where some rules in $tr$ have dependencies.
-For this case, assume we have a path condition $pc_1$ contained in the set of path conditions generated for $tr$, considering layers up to layer $l$ of $tr$ have executed.
-Assume also we have a rule $rl$ of layer $l+1$ of $tr$ that has dependencies and can be combined with $pc$.
-If rule $rl$ is totally combined with path condition $pc_1$, according to \cref{def:rul_comb_partial_total} and \cref{fig:multiple_total_satisfied_dependencies}, then nothing needs to be shown as $pc_1$ is not kept in the path condition set but rather replaced by its combination with $rl$.
-However, in case rule $rl$ is partially combined with $pc$, as defined in \cref{def:rul_comb_partial_total} and \cref{def:comb_path_cond_rule_single}, then multiple path conditions are generated and additionally $pc_1$ is kept in the path condition set.\\The proof will be complete once it is shown that for the path conditions that are generated when rules are partially combined, it is also the case that no two path conditions can abstract the same execution.
-This last part of the proof can be built in an analog fashion to the construction of the proof for point~\ref{item:uniqueness_rule_no_dep}.
-As previously mentioned, the complete proof can be found as \cref{lem:uniqueness_appendix} in \cref{sec:val_complete_path_cond_gen}.
-
-\end{ps}

### 2016-06-29 text/property_verification.tex
Move property verification proofs into own section.

@@
 % Given these definitions, we see that as all transformation executions are represented by path conditions, proving properties on the path conditions allows us to conclude property validity on all transformation executions.
-
-\subsection{Validity and Completeness}
-\label{subsec:prop_proving_valid_complete}
-
-As for the path condition building algorithm, \emph{validity} and
-\emph{completeness} need to be examined regarding our property verification
-algorithm.
-In this context \emph{validity} means that if a property is satisfied by all path conditions generated for a transformation $tr$, then that property is satisfied by all executions of that transformation.
-On the other hand, if the property is not satisfied by at least one path condition, then it will not be satisfied by at least one transformation execution.
-In other words, we wish to show that no false positive or false negative proof results are induced by the abstraction relation.
-
-% The proof of of these statements necessarily relies on the abstraction relation formally presented in\cref{def:instance_pc_ex}.
-
-On the other hand, \emph{completeness} means that we are sure that all properties that can be expressed about a transformation can be shown to hold or not hold in all transformation executions.
-
-As with the proofs for the validity and completeness of the abstraction relation, we present only proof sketches in this section in the interest of readability.
-Full proofs are shown in \cref{sec:val_complete_prop_verif} as \cref{prop:proof_validity_appendix} and \cref{prop:proof_completeness_appendix}.
-
-% \subsubsection{Validity}
-% 
-% In this section we prove the validity of our property proof algorithm by considering what it
-% formally means when a property is proved on a path condition. We argue that if
-% the property holds or not on the path condition, then it will accordingly hold
-% or not on for any transformation executions that the path condition represents.
-
-% As well, we also prove that if a property holds or not on a specific transformation
-% execution, it will hold or not on the path condition that represents that
-% transformation execution.
-\begin{proposition}{(Validity) The result of proving a property on a set of path conditions generated for a transformation or all executions of that transformation is the same.}
-\label{prop:proof_validity}
-
-Let $tr\in \textsc{Transf}^{sr}_{tg}$ be a transformation and $p \in \textsc{Property}(tr)$ be a
-property of $tr$. Given this, we have that transformation $tr$ satisfies property $p$ if and only if:
-\begin{equation}
-\label{eq:prop_proof_transf}
-\bigwedge_{pc\in \textsc{Pathcond}(tr)} pc\vdash p \hspace{.3cm} \Longleftrightarrow \hspace{.3cm} \bigwedge_{ex\in \textsc{Exec}(tr)}ex\models p
-\end{equation}
-\end{proposition}
-\begin{ps}
-In order to prove the proposition in \cref{eq:prop_proof_transf} we will start by demonstrating that,
-if property $p$ holds on a path condition $pc$ generated for $tr$,
-then $p$ will necessarily hold on all execution $ex$ of $tr$ that is abstracted by $pc$.
-On the other hand, if $p$ does not hold on $pc$ then it will not hold for at least one execution $ex$ of $tr$ abstracted by $pc$.
-This lemma can be stated as follows:
-\begin{equation}
-\label{eq:prop_proof_pc}
-pc\vdash p \;\Longleftrightarrow \;\forall ex\in\{ex\in\textsc{Exec}(tr)\;|\;ex\Vvdash pc\}\;.\; ex\models p
-\end{equation}
-We thus need to demonstrate both directions of the equivalence in \cref{eq:prop_proof_pc}, as follows:
-\begin{itemize}
-\item The proof of the left-to-right direction of the equivalence is split
-into two cases as $pc\vdash p$ is true (as stated
-in~\cref{def:sat_prop_pc}) when either: (1) the pre-condition of property $p$ cannot
-be found in path condition $pc$; or (2) the pre-condition of $p$ is
-found in $pc$ and the post-condition of $p$ also.
-We need to demonstrate that both these cases entail that $ex\models p$ holds.
-\item The proof of the right-to-left direction of the equivalence, which can be shown on its contrapositive:
-\begin{multline}
-\label{eq:prop_proof_right_left}
-\neg(pc\vdash p) \;\Longrightarrow \;\\\exists ex\in\{ex\in\textsc{Exec}(tr)\;|\;ex\Vvdash pc\}\;.\; \neg(ex\models p)
-\end{multline}
-According to \cref{def:sat_prop_pc}, to demonstrate \cref{eq:prop_proof_right_left} we will need to show that when the pre-condition of property $p$ is found in $pc$ but the post-condition of $p$ is not, the property does not hold on at least one execution of $tr$ abstracted by $pc$.
-\end{itemize}
-
-Once \cref{eq:prop_proof_pc} is proved, we know that all path conditions on which a property holds represent executions on which the property also holds.
-Thus, if the property holds on all path conditions then it necessarily holds on all executions.
-On the other hand, if a property does not hold on one path condition, making it such that the conjunction on the left side of the equivalence in \cref{eq:prop_proof_transf} is false, then according to \cref{eq:prop_proof_pc} an execution for which it also does not hold exists.
-This makes it such that the conjunction on the right side of the equivalence in \cref{eq:prop_proof_pc} is also false.
-\end{ps}
-
-\begin{proposition}{(Completeness) Properties of a transformation can be shown to either hold for all transformation executions, or not hold for at least one transformation execution.}
-\label{prop:proof_completeness}
-\end{proposition}
-\begin{pf}
-This result follows from two previous results: \cref{prop:pc_completeness}, that tells us that every transformation execution is abstracted by one path condition; and \cref{prop:proof_validity} that shows us that every path condition is taken into consideration during property proof.
-Note that \cref{lem:uniqueness} guarantees consistency of our results, in the sense that the uniqueness of one path condition per transformation execution guarantees that a property cannot be proven to be both \emph{true} and \emph{false} for two path conditions representing the same transformation execution.
-\end{pf}
-
-% ~\newline\noindent{\bfseries Proposition \ref{prop:validity}.}
-% \textit{ (Validity) The result of checking a property on a path condition and of checking the property on all the transformation executions that path
-% condition abstracts will be the same.}\newline
-% 
-% We place the formal proof of this statement in appendix~\ref{sec:proofs}, proposition~\ref{prop:validity} as this result requires significant mathematical development.
-% This result provides the guarantee that path conditions accurately represent their corresponding concrete transformation executions when performing property validation.
-% 
-% ~\newline\noindent{\bfseries Proposition \ref{prop:PP_completeness}.}
-% \textit{ (Completeness) Proving that a property holds or does not
-% hold by examining all path conditions of the transformation is equivalent to
-% proving the property holds or does not hold for all transformation executions.}\newline
-% 
-% %\begin{proposition}{(Completeness) Proving that a property holds or does not
-% %hold by examining all path conditions of the transformation is equivalent to
-% %proving the property holds or does not hold for all transformation executions.}
-% %\end{proposition}
-% 
-% The \emph{completeness} of our property verification algorithm is shown in appendix~\ref{sec:proofs}, proposition~\ref{prop:PP_completeness}.

### 2016-06-29 text/property_verification_proofs.tex
Move property verification proofs into own section.

--- 2016-05-02 text/appendix_property_proof.tex
@@
-\section{Validity and Completeness of Property Verification}
-\label{sec:val_complete_prop_verif}
+\subsection{Validity and Completeness}
+\label{subsec:prop_proving_valid_complete}
 
-\counterwithin{proposition}{section}
-\counterwithin{lemma}{section}
-\counterwithin{equation}{section}
-\setcounter{proposition}{0}
-\setcounter{lemma}{0}
-\setcounter{equation}{0}
+As for the path condition building algorithm, \emph{validity} and
+\emph{completeness} need to be examined regarding our property verification
+algorithm.
+In this context \emph{validity} means that if a property is satisfied by all path conditions generated for a transformation $tr$, then that property is satisfied by all executions of that transformation.
+On the other hand, if the property is not satisfied by at least one path condition, then it will not be satisfied by at least one transformation execution.
+In other words, we wish to show that no false positive or false negative proof results are induced by the abstraction relation.
 
-\onehalfspacing
+% The proof of of these statements necessarily relies on the abstraction relation formally presented in\cref{def:instance_pc_ex}.
 
-\begin{definition}{Property of a Transformation\\}
-\label{def:trans_prop_appendix}
-\CatchFileBetweenTags{\propertydef}{text/definitions}{propertydef}{\propertydef}
-\end{definition}
+On the other hand, \emph{completeness} means that we are sure that all properties that can be expressed about a transformation can be shown to hold or not hold in all transformation executions.
 
-In \cref{def:trans_prop_appendix}, pre-conditions use the same pattern language as the match graph in
-DSLTrans rules, allowing the possibility of including several instances of
-the same metamodel element as well as indirect links in the property. Indirect links in properties have
-the same meaning as in the rule match graph -- they
-involve patterns over the transitive closure of containment links in pre-condition graphs.
+As with the proofs for the validity and completeness of the abstraction relation, we present only proof sketches in this section in the interest of readability.
+Full proofs are shown in \cref{sec:val_complete_prop_verif} as \cref{prop:proof_validity_appendix} and \cref{prop:proof_completeness_appendix}.
 
-Post-conditions also use the same pattern language as the
-apply graphs of DSLTrans transformation rules, with the additional
-possibility of expressing indirect links in post-conditions. Traceability links can
-also be used in properties to impose traceability relations between pre-condition
-and post-condition elements.
+% \subsubsection{Validity}
+% 
+% In this section we prove the validity of our property proof algorithm by considering what it
+% formally means when a property is proved on a path condition. We argue that if
+% the property holds or not on the path condition, then it will accordingly hold
+% or not on for any transformation executions that the path condition represents.
 
-Note that \cref{def:trans_prop_appendix} includes a condition stating a surjective typed graph homomorphism needs to exist between the match part of at least one of the transformation's path condition, and the pre-condition of the property of interest.
-This condition makes sure that the property's pre-condition can be found at least in one execution of the transformation abstracted (the mathematical argument for this fact is given in the proof of \cref{prop:proof_validity}).
-This condition makes the checking the validity of a property in the transformation meaningful.
-If this condition would not be true then it could be that the input pattern required by the property would never be fully matched during transformation execution, making such a property not relevant\footnote{In~\cite{DBLP:conf/sle/BarrocaLAFS10} we have referred to these properties \emph{non-provable}.
-In the work presented here we explicitely disallow the construction of this class of properties.} for the transformation at hand.
-
-\begin{definition}{Satisfaction of a Property by an Execution of a Transformation\\}
-\label{def:sat_prop_ex_appendix}
-\CatchFileBetweenTags{\satpropexe}{text/definitions}{satpropexe}{\satpropexe}
-\end{definition}
-
-\Cref{def:sat_prop_ex_appendix} states that, every time a graph that is isomorphic to the property's pre-condition is found in (the containment transitive closure of) the input model of the transformation's execution, a graph that is isomorphic to the complete property needs to be found in (the containment transitive closure of) the transformation execution.
-Note that the last part of the proposition in \cref{def:sat_prop_ex_appendix} ensures that the graph that is isomorphic to the property's pre-condition and the graph that is isomorphic to the complete property overlap on their pre-condition parts.
-
-% resulting from the input pattern of $ex$ identified as pre-condition by $f$.
-Generally speaking, whenever the property's pre-condition is isomorphically found as a pattern of the input model of a execution, the property's post-condition must also be isomorphically found in the output generated from that input pattern.
-Note that, because a property may contain traceability links between the pre- and post-condition elements, it is necessary to check that those traceability links are also present in the execution.
-Because of this, the $g$ homomorphism in definition~\ref{def:sat_prop_ex} that checks the property's post-condition is such that it considers the same match pattern identified by the $f$ homomorphism when the pre-condition is checked.
-
-\begin{definition}{Satisfaction of a Property by a Path Condition\\}
-\label{def:sat_prop_pc_appendix}
-\CatchFileBetweenTags{\satproppc}{text/definitions}{satproppc}{\satproppc}
-\end{definition}
-
-The principle behind the satisfaction relation in \cref{def:sat_prop_pc_appendix} is the same as the one behind the satisfaction relation between a property and an execution of a transformation in \cref{def:sat_prop_ex_appendix}: whenever the property's pre-condition is found in the path condition then so is the complete property.
-Also, those two graphs found in the path condition share the property's pre-condition part.
-This last condition enforces that the pre- and post-conditions of the property are correctly linked by traceability link in the path condition.
-
+% As well, we also prove that if a property holds or not on a specific transformation
+% execution, it will hold or not on the path condition that represents that
+% transformation execution.
 \begin{proposition}{(Validity) The result of proving a property on a set of path conditions generated for a transformation or an all executions of that transformation is the same.\\}
@@
 \begin{proposition}{(Completeness) Properties of a transformation can be shown to either hold for all transformation executions, or not hold for at least one transformation execution.}
-\label{prop:proof_completeness_appendix}
+\label{prop:proof_completeness}
 \end{proposition}
-\begin{proof}
-This result follows from two previous results: \cref{prop:pc_completeness_appendix}, that tells us that every transformation execution is abstracted by one path condition; and \cref{prop:proof_validity_appendix} that shows us that every path condition is taken into consideration during property proof.
-Note that \cref{lem:uniqueness_appendix} guarantees consistency of our results, in the sense that the uniqueness of one path condition per transformation execution guarantees that a property cannot be proven to be both \emph{true} and \emph{false} for two path conditions representing the same transformation execution.
-\end{proof}
-
-\clearpage
+\begin{pf}
+This result follows from two previous results: \cref{prop:pc_completeness}, that tells us that every transformation execution is abstracted by one path condition; and \cref{prop:proof_validity} that shows us that every path condition is taken into consideration during property proof.
+Note that \cref{lem:uniqueness} guarantees consistency of our results, in the sense that the uniqueness of one path condition per transformation execution guarantees that a property cannot be proven to be both \emph{true} and \emph{false} for two path conditions representing the same transformation execution.
+\end{pf}

### 2016-06-29 text/building_path_conditions.tex
Move old sections. Divorce paper from definitions file.

@@
 
-\CatchFileBetweenTags{\pccfrc}{text/definitions}{pccfrc}{\pccfrc}
+\reviewer{A $\sqcup$ operator is defined in Def. 19, but there you seem to
+  rely on the fact that some elements of the path conditions and
+  some elements of the rule are identical, otherwise the union
+  would not work. Since a rule can be applied in several places,
+  this would require a renaming of the rule elements, which - as 
+  far as I could see - is not specified.}
+
+Let $pc = \langle V',E',st',\tau',Match',Apply',Rulecop'\rangle\in\\
+\textsc{Pathcond}^{sr}_{tg}$ be a path condition and $rl = \langle
+V'',E'',st'',\tau'',\\Match'',Apply''\rangle\in \textsc{Rule}^{sr}_{tg}$ be a transformation rule, where their respective typed graphs can be joint.
+The union of $pc$ with $rl$ is built using the operator $\stackrel{trace}{\sqcup}:
+\textsc{Pathcond}^{sr}_{tg}\times \textsc{Rule}^{sr}_{tg}\rightarrow
+\textsc{Pathcond}^{sr}_{tg}$, as follows:
+
+$$pc\stackrel{trace}{\sqcup}rl = \langle V,E,st,\tau,Match,Apply,Rulecop\rangle$$
+
+where we have that $V=V'\cup V''$, $E'\cup E''\subseteq E$, $st'\cup
+st''\subseteq st$, $\tau'\cup \tau''\subseteq \tau$ and if $v_1\xrightarrow{e}
+v_2\in E\setminus E'\cup E''$ then we have that $v_1\in Apply(V'')$, $v_1\notin Apply(V')$, $v_2\in Match(V'')$ and also that $\tau'(e)=trace$.
+Additionally, $Match = Match'\sqcup Match''$ and $Apply = Apply'\sqcup Apply''$.
+Finally, we have that: $Rulecop = Rulecop'\cup$ {rl}.
+
+% $$pc = \bigsqcup_{rl\in RC} trace(rl_{pathcond})$$
+% 
+% where $trace:\textsc{Pathcond}\rightarrow \textsc{Pathcond}$ is such that
+% $trace(\langle V,E,st,\tau\rangle) = \langle V,E',st',\tau'\rangle$ and we have
+% that: $E\subseteq E'$; $st\subseteq st'$; $\tau \subseteq \tau'$; if $\neg
+% (v_1\xrightarrow{e'} v_2\in E)\;\land\;\tau(e')=backward$ then we have that if
+% $v_1\in Apply(V)$ and $v_2\in Match(V)$ then $v_1\xrightarrow{e} v_2\in
+% E'\setminus E$ and $\tau'(e)=trace$. Also, $rl_{pathcond}$ is rule $rl$ typed as
+% a path condition by adding to the $rl$ tuple a $Rule$ function that identifies
+% each vertex in $V$ as belonging to $rl$. Finally, we have used the
+% $\sqcup$ notation to denote the extension of the typed graph union to the
+% component-wise of tuples that represent path conditions.
 \end{definition}
@@
 
-\CatchFileBetweenTags{\pccfrctext}{text/definitions}{pccfrctext}{\pccfrctext}
+\cref{def:combine_pc_with_rule} shows the formal definition of combining a path condition with a rule.
+When a path condition is combined with a rule their typed graphs are united.
+Additionally, symbolic traceability links will be built at this time between the newly added apply elements of the rule and all of the rule's match elements.
+As a reminder, the link creation algorithm and examples have been introduced in \cref{subsubsec:traceability}.\\
+
+Note that the fact that the graphs are potentially joint allows us to overlap a rule with the path condition by anchoring the rule on traceability links shared by the path condition and the rule graph.
+In the mathematical development that follows we will often refer to the joint parts of two or more typed graphs using the term ``glue''.
 
@@
 \label{def:rule_comb_no_dependencies}
-\CatchFileBetweenTags{\nodeps}{text/definitions}{nodeps}{\nodeps}
+
+The combination of a path condition $pc$ and a rule $rl$, when $rl$ has no dependencies, is described by the relation $\stackrel{combine}{\rightarrow}\subseteq \textsc{Pathcond}^{sr}_{tg} \times \mathcal{P}(\textsc{Pathcond}^{sr}_{tg}) \times \textsc{Rule}^{sr}_{tg} \times \mathcal{P}(\textsc{Pathcond}^{sr}_{tg})$, defined as follows:
+
+% $$\frac{\begin{array}{ll}&
+% pc\in PC\;,\;\langle pc,rl\rangle \xrightarrow{combine} pc'\;,\;PC\setminus\{pc\},rl,\rangle \xrightarrow{combstep} PC'
+% \end{array}}
+% {\langle PC,rules\rangle \xrightarrow{rulestep} PC'}$$
+
+$$\frac{\begin{array}{ll}rl=\langle V,E,st,\tau, Match,Apply\big\rangle\;,\;\nexists e\in E\,.\,\tau'(e)=trace
+\end{array}}
+{\langle pc,AC,rl\rangle \xrightarrow{combine} AC\;\;\cup\;\; \bigcup_{pc'\in AC} pc'\stackrel{trace}{\sqcup} rl}$$
 
@@
 
-\CatchFileBetweenTags{\nodepstext}{text/definitions}{nodepstext}{\nodepstext}
+Relation $\stackrel{combine}{\rightarrow}$ in \cref{def:rule_comb_no_dependencies} models the operational combination step shown in \cref{fig:layers_pc2} (the vertical black arrows between boxes).
+The relation has three input arguments: the first argument is the original path condition from the previous layer (shown as the topmost box in \cref{fig:layers_pc2} with label $PC$); the second argument is the set of path conditions accumulated thus far by combining other rules in the current layer with the original path condition; and the third argument is the rule from the current layer now being combined.
+The fourth argument of the relation, the relation's output, is the new set of path conditions resulting from this combination.
+
+Briefly, the equation in \cref{def:rule_comb_no_dependencies} states that whenever a rule has no backward links typed as \emph{trace} (i.e.
+no dependencies), all path conditions in the accumulator set are kept, along with the result of combining all the path conditions in the accumulator set with the current rule.
 
@@
 \label{def:rule_comb_unsatisfied}
-\CatchFileBetweenTags{\unsatdeps}{text/definitions}{unsatdeps}{\unsatdeps}
+% \begin{definition}{Unsatisfied Dependencies Between Two Path Conditions}
+% 
+% Let $A=\langle V_1,E_1,st_1,\tau_1\rangle, B=\langle V_2,E_2,st_2,\tau_2\rangle\in \textsc{Pathcond}^{sr}_{tg}$ be two path conditions.
+We say that $A$ does not satisfy $B$'s dependencies if and only if $\neg (B'\vartriangleleft A')$, where $A'=A_{|backward}$ and $B'=B_{|backward}$.
+% \end{definition}
+
+The combination of a path condition $pc$ and a rule $rl$, when $rl$ has dependencies that are not satisfied by $pc$, is described by the relation $\stackrel{combine}{\rightarrow}\subseteq \textsc{Pathcond}^{sr}_{tg}\times \mathcal{P}(\textsc{Pathcond}^{sr}_{tg})\\ \times \textsc{Rule}^{sr}_{tg} \times \mathcal{P}(\textsc{Pathcond}^{sr}_{tg})$, defined as follows:
+
+% $$\frac{\begin{array}{ll}&
+% pc\in PC\;,\;\langle pc,rl\rangle \xrightarrow{combine} pc'\;,\;PC\setminus\{pc\},rl,\rangle \xrightarrow{combstep} PC'
+% \end{array}}
+% {\langle PC,rules\rangle \xrightarrow{rulestep} PC'}$$
+
+$$\frac{\begin{array}{ll}\neg\big(rl|_{trace}\blacktriangleleft pc|_{trace}\big) \end{array}}
+{\langle pc,AC,rl\rangle \xrightarrow{combine} AC}$$
+
 \end{definition}
@@
 \label{def:comb_path_cond_rule_single}
-\CatchFileBetweenTags{\satdeps}{text/definitions}{satdeps}{\satdeps}
+% \begin{definition}{Satisfied Dependencies Between Two Path Conditions}
+% 
+% Let $A=\langle V_A,E_A,st_A,\tau_A,Match_A,Apply_A\rangle, B=\langle V_B,E_B,st_B,\tau_B,Match_B,Apply_B\rangle\in \textsc{Pathcond}^{sr}_{tg}$ be two path conditions.
+We say that $A$ satisfies $B$'s dependencies if and only if  $B'\vartriangleleft A'$, where $A'=A_{|backward}$ and $B'=B_{|backward}$.
+% \end{definition}
+% 
+% A path condition A partially satisfies the dependencies of a path condition B if
+% there exists an injective typed graph homomorphism between the backward links of
+% B and those of A. In other words, 
+%
+% Note that this works for both cases where more instances of
+% the same backward link exist in A, or in B.
+
+The single rule partial and total combination relations $\stackrel{p\_comb}{\rightarrow}$ and $\stackrel{t\_comb}{\rightarrow}$, both having having signature $\mathcal{P}(\textsc{Pathcond}^{sr}_{tg}) \times \textsc{Rule}^{sr}_{tg} \times \textsc{Rule}^{sr}_{tg} \times \mathcal{P}(\textsc{Pathcond}^{sr}_{tg})$ are defined as follows:
 
@@
 \label{eq:pcomb}
-\CatchFileBetweenTags{\satdepseqone}{text/definitions}{satdepseqone}{\satdepseqone}
+\frac{\begin{array}{ll}
+rl\cong rl_{glue}\sqcup ma_{\Delta}
+\end{array}}
+{\langle AC,rl,rl_{glue}\rangle \xrightarrow{p\_comb} AC\;\;\cup\;\; \bigcup_{pc\in AC} pc\stackrel{trace}{\sqcup} (rl_{glue} \sqcup ma_{\Delta})}
 \end{align}
@@
 \label{eq:tcomb}
-\CatchFileBetweenTags{\satdepseqtwo}{text/definitions}{satdepseqtwo}{\satdepseqtwo}
+\frac{\begin{array}{ll}&
+rl\cong rl_{glue}\sqcup ma_{\Delta}
+\end{array}}
+{\langle AC,rl,rl_{glue}\rangle \xrightarrow{t\_comb} \bigcup_{pc\in AC} pc\stackrel{trace}{\sqcup} (rl_{glue} \sqcup ma_{\Delta})}
 \end{align}
@@
 
-\CatchFileBetweenTags{\satdepstwo}{text/definitions}{satdepstwo}{\satdepstwo}
+The multiple rule partial and total combination relations $\stackrel{p\_step}{\rightarrow}$ and $\stackrel{t_step}{\rightarrow}$, both having having signature $\mathcal{P}(\textsc{Pathcond}^{sr}_{tg}) \times \mathcal{P}(\textsc{Rule}^{sr}_{tg}) \times \textsc{Rule}^{sr}_{tg} \times \mathcal{P}(\textsc{Pathcond}^{sr}_{tg})$ are defined as follows:
+
+\begin{align}
+\label{eq:pstepbase}
+\frac{\begin{array}{ll}
+\end{array}}
+{\langle AC,rl,\emptyset\rangle \xrightarrow{p\_step} AC}
+\end{align}
+
+\begin{align}
+\label{eq:pstep}
+\frac{\begin{array}{ll}
+rl_{glue}\in &partialSet,\;\langle AC,rl,rl_{glue}\rangle \xrightarrow{p\_comb} AC''\;,\;\\
+&\langle AC'',rl,partialSet\setminus\{rl_{glue}\}\rangle \xrightarrow{p\_step} AC'
+\end{array}}
+{\langle AC,rl,partialSet\rangle \xrightarrow{p\_step} AC'}
+\end{align}
+
+\begin{align}
+\label{eq:tstepbase}
+\frac{\begin{array}{ll}
+\end{array}}
+{\langle AC,rl,\emptyset\rangle \xrightarrow{t\_step} AC}
+\end{align}
+
+\begin{align}
+\label{eq:tstep}
+\frac{\begin{array}{ll}
+rl_{glue}\in &totalSet\;,\;\langle AC,rl,rl_{glue}\rangle \xrightarrow{t\_comb} AC''\;,\;\\
+&\langle AC'',rl,totalSet\setminus\{rl_{glue}\}\rangle \xrightarrow{t\_step} AC'
+\end{array}}
+{\langle AC,rl,totalSet\rangle \xrightarrow{t\_step} AC''} 
+\end{align}
 \end{definition}
@@
 
-\CatchFileBetweenTags{\satdepsthree}{text/definitions}{satdepsthree}{\satdepsthree}
+% When a rule is combined with a path condition, partial and total matches of the rule may exist in the path condition.
+A total match of the rule is found in the path condition if a surjective typed graph homomorphism exists between the backward matcher of the rule and a subgraph of the path condition ($rl_{glue}\sqsubseteq pc \,\land\, rl\blacktriangleleft rl_{glue}$).
+As a reminder, the backward matcher of a rule consists of the rule's match part, together with its backward links (see \cref{def:back_match_transformation_rule}).
+In this case, existing path conditions in the output set will combined with\ldots
+% 
+% These two cases are modeled by the two inference rules in \cref{def:comb_path_cond_rule}. 
+% 
+% For each case a delta is added to the glue graph found such that the necessary apply or match part of the rules are added to the original path condition.
+In case the match is partial, then the previous set of path conditions is kept where the delta has been added, as well a second possibility with the total rule.
+Note that the fact that we use a surjective homomorphism allows us to deal with both rules that are isomorphic or homomorphic to parts of PC.
+This allows us to have more backward links in the path condition than in a rule, but also the other way around.
+
+The combination of a path condition $pc$ and a rule $rl$, when $rl$ has dependencies that are satisfied by $pc$, is described by the relation $\stackrel{combine}{\rightarrow}\subseteq \textsc{Pathcond}^{sr}_{tg} \times \mathcal{P}(\textsc{Pathcond}^{sr}_{tg}) \times \textsc{Rule}^{sr}_{tg} \times \mathcal{P}(\textsc{Pathcond}^{sr}_{tg})$, defined as follows:%\levi{surjection is used too loosely here}
+
+% $$\frac{\begin{array}{ll}&
+% pc\in PC\;,\;\langle pc,rl\rangle \xrightarrow{combine} pc'\;,\;PC\setminus\{pc\},rl,\rangle \xrightarrow{combstep} PC'
+% \end{array}}
+% {\langle PC,rules\rangle \xrightarrow{rulestep} PC'}$$
+
+$$\frac{\begin{array}{ll}rl|_{trace}\blacktriangleleft &pc|_{trace}\;,\\&\langle AC,rl,partialsat(rl,pc)\rangle\xrightarrow{p\_step} AC''\;,\;\\&\langle AC'',rl,totalsat(rl,pc)\rangle \xrightarrow{t\_step} AC'
+\end{array}} 
+{\langle pc,AC,rl\rangle \xrightarrow{combine} AC'}$$
+\begin{center}
+\vspace{.3cm}
+where
+\begin{align*}
+rl_{glue}\in&~partialsat(rl,pc) \;\Longleftrightarrow\; \\
+&rl_{glue}\sqsubseteq pc^{*} \,\land\, rl|_{trace}\blacktriangleleft rl_{glue}\quad\land\, \\&\hspace{1cm}\nexists rl' \,.\, (rl_{glue}\sqsubseteq rl'\sqsubseteq pc^{*} \land \Vert rl\Vert \blacktriangleleft rl')
+\end{align*}
+and 
+\begin{align*}
+rl_{glue}\in totalsat(rl,pc) \;\Longleftrightarrow \;rl_{glue}\sqsubseteq pc^{*} \,\land\, \Vert rl\Vert\blacktriangleleft rl_{glue}
+\end{align*}
+\end{center}
+
 \end{definition}
@@
 \label{def:path_cond_layer_comb}
-\CatchFileBetweenTags{\combpclayer}{text/definitions}{combpclayer}{\combpclayer}
+
+The \emph{layer combination relation}
+$\stackrel{combpclayer}{\rightarrow}\subseteq \textsc{Pathcond}^{sr}_{tg} \times\\ \mathcal{P}(\textsc{Pathcond}^{sr}_{tg}) \times \textsc{Layer}^{sr}_{tg}\times
+\mathcal{P}(\textsc{Pathcond}^{sr}_{tg})$ relation is defined as follows:
+$$\frac{}
+{\langle pc,AC,\emptyset\rangle \xrightarrow{combpclayer} AC}$$
+
+$$\frac{\begin{array}{ll}rl\in layer,\;\langle pc,AC&,rl\rangle\xrightarrow{combine}AC'',\\ &\langle pc,AC'',layer\setminus\{rl\}\rangle\xrightarrow{combpclayer}AC'
+\end{array}}
+{\langle pc,AC,layer\rangle \xrightarrow{combpclayer} AC''}$$
+
 \end{definition}
 
-\CatchFileBetweenTags{\combpclayertext}{text/definitions}{combpclayertext}{\combpclayertext}
+After the step in~\cref{def:path_cond_layer_comb} is repeated for all the path conditions in the previous layer, these new sets of path conditions are collected together to produce the working set of path conditions for the layer.
+This process is modeled by relation $\stackrel{combpcsetlayer}{\rightarrow}$ in \cref{def:path_cond_set_layer_comb}.
 
@@
 \label{def:path_cond_set_layer_comb}
-\CatchFileBetweenTags{\combpcsetlayer}{text/definitions}{combpcsetlayer}{\combpcsetlayer}
+The \emph{path condition layer step relation}
+$\stackrel{combpcsetlayer}{\rightarrow}\subseteq \\\mathcal{P}(\textsc{Pathcond}^{sr}_{tg}) \times \textsc{Layer}^{sr}_{tg}\times
+\mathcal{P}(\textsc{Pathcond}^{sr}_{tg})$ relation is defined as follows:
+$$\frac{}
+{\langle \emptyset,layer\rangle \xrightarrow{combpcsetlayer} \emptyset}$$
+
+$$\frac{\begin{array}{ll}pc\in AC,\;\langle pc,\{pc\}&,layer\rangle\xrightarrow{combpclayer}AC',\\ &\langle AC\setminus\{pc\},layer\rangle\xrightarrow{combpcsetlayer}AC''
+\end{array}}
+{\langle AC,layer\rangle \xrightarrow{combpcsetlayer} AC'\cup AC''}$$
 \end{definition}
 
-\CatchFileBetweenTags{\combpcsetlayertext}{text/definitions}{combpcsetlayertext}{\combpcsetlayertext}
+This working set of path conditions obtained for each layer is then itself combined with the rules in the next layer as in the algorithm just described, to obtain yet another working set of path conditions.
+This process will then continue in this layer-by-layer fashion through the transformation and is formally described in \cref{def:path_cond_gen}.\\
+After all layers have been processed, the working set of the last layer contains all the  possible path conditions of the transformation.
+Through our abstraction relation defined in \cref{sec:abstraction_relation}, the final set of created path conditions will represent every feasible transformation execution.
+\cref{sec:verif_dsltrans_props} will discuss how our algorithm proves properties on these path conditions, and thus on all executions of the transformation.
 
@@
 \label{def:path_cond_gen}
-\CatchFileBetweenTags{\pathcondgen}{text/definitions}{pathcondgen}{\pathcondgen}
+Let $[layer::tr]\in \textsc{Transf}^{sr}_{tg}$ be a transformation, where $layer\in
+\textsc{Layer}^{sr}_{tg}$ is a Layer and $tr$ also a transformation. The $\stackrel{pathcondgen}{\rightarrow}\subseteq \mathcal{P}(\textsc{Pathcond}^{sr}_{tg}) \times \textsc{Transf}^{sr}_{tg}\times
+\mathcal{P}(\textsc{Pathcond}^{sr}_{tg})$ is defined as follows:
+$$\frac{}
+{\langle AC,[\;]\rangle \xrightarrow{pathcondgen} AC}$$
+
+$$\frac{\begin{array}{ll}\langle \epsilon_{pc},layer^{\star}\rangle\xrightarrow{combpcsetlayer}AC''\;,\;\langle AC'',tr\rangle\xrightarrow{pathcondgen}AC'
+\end{array}}
+{\langle \epsilon_{pc},[layer::tr]\rangle \xrightarrow{pathcondgen} AC}$$
+$$\text{where } layer^{\star}=\bigcup_{rl\in l}rl^{\star}$$
+
 \end{definition}
 
-\CatchFileBetweenTags{\pathcondgentext}{text/definitions}{pathcondgentext}{\pathcondgentext}
+Note that in \cref{def:path_cond_gen}, the recursive rule considers the expansion ($layer^{\star}$) of all the rules in a layer (see \cref{def:transformation_rule_expansion}).
+This allows us to deal with polymorphism during path condition generation.
+In particular, given one rule $rl$ of $layer$, we consider for path condition generation all rules containing possible of replacements of each match element in $rl$ of certain type by an element belonging to one of the type's subtypes, as defined in the source metamodel $sr$.\\
+After all layers have been processed, the working set of the last layer contains all the possible path conditions of the transformation.
+Through our abstraction relation in \cref{def:abstraction_pc_ex_appendix}, the final set of created path conditions will represent every feasible transformation execution.
+
+\paragraph{\textbf{Notation:}}
+
+We will use the abbreviation $\textsc{Pathcond}(tr)$ to represent the set of path conditions $AC$ produced for a transformation $tr$, where $\langle \epsilon_{pc},tr\rangle \xrightarrow{pathcondgen} AC$.

### 2016-06-29 text/formal_background.tex
Move old sections. Divorce paper from definitions file.

--- 2015-03-13 text/formal_background.tex
@@
 
+\reviewer{Your formalisation is made more complex than necessary because you do not put
+the power of typed graphs to work. If your source and target metamodels have
+disjoint types (which you can assume without loss of generality) you can take
+their union, augmented with trace edge types from all target node types to all
+source node types, and take that to be the meta-model for an input-output model,
+and also for a rule. Suddenly you don’t have to carry arround the distinction be-
+tween Match and Apply (for rules) or input and output (for input-output models)
+around any more: they are just the projections onto the types of the source, resp.
+target metamodel.}
+
+\reviewer{Section 4.2.2 on “backward links” is rather confusing, as it speaks about trace-
+ability links being added to rules. You never add traceability links to rules any-
+where in the paper; in fact, your formal notion of a rule has no room for them.
+Instead, rules only have backward links. (The clarity of the the situation is not
+improved by the fact that those are labelled trace.)}
+
+\reviewer{I think the situation would be much more clear if you were to use the common
+notion of left hand side and right hand side of a rule. The left hand side of a
+rule rl is precisely your ||rl||, the right hand side is the entire rl augmented by
+your traceability links, as in 4.2.2. Suddenly we are back in a well-known graph
+transformation formalism, and your rules are in fact so simple (no deletion, no
+negative application conditions, injective matching) that I believe it makes no
+difference whether you use algebraic graph transformation or some more con-
+structively defined variant. The only standard notion is that of transitive closure
+in the left hand side; but there are plenty of GT tools that offer this extension.
+In fact I see no reason why your rule application would then not precisely coin-
+cide with the the same notion in, say, SPO graph rewriting; and I hope you agree
+that this would help you no end in explaining what you are doing. (If, on the
+other hand, there is after all some difference then this, too, would be interesting
+to know, as that difference is not at all apparent right now.)}
+
+\reviewer{Viewed like this, moreover, I think your notions of input-output models and rules
+are very close to triple graph transformation as used by Schrr at all. I think the
+main (if not only) difference lies in the fact that your "glue graph" is actually not
+a graph; instead, you use traceability edges directly from the target model to the
+source model. If you were to turn those edges into nodes with source and target
+edges, even that difference would disappear.}
+
 Armed with the fundamental notion of typed graph, we can then introduce other formal concepts in Sections~\ref{sec:DSLTrans_formal}, \ref{sec:DSLTrans_syntax} and \ref{sec:DSLTrans_semantics} which describe the artifacts from the modeling and transformation world that we require for our verification technique.
@@
 It is meant as a formal reference for the upcoming formal development.
-This section can be safely skipped or skimmed by the reader, who can return to these definitions punctually to understand the detailed formal underpinning of our approach.
+%This section can be safely skipped or skimmed by the reader, who can return to these definitions punctually to understand the detailed formal underpinning of our approach.
 
-\subsection{Typed Graphs}
+\reviewer{There is a lot of redundancy in the paper, on several levels:
+  - input-output-models, path conditions and properties are
+    all defined very similarly. Couldn't you subsume them in a single
+    definition and spell out the differences?}
+
+\newcommand{\defineggprime}{Let $\langle V,E,(s,t),\tau,VT,ET\rangle = g$, and\\ $\langle V',E',(s',t'),\tau',VT',ET'\rangle = g'$, where $g, g' \in \textsc{TG}$.}
+
+\newcommand{\ET}{\mathit{ET}}
+\newcommand{\VT}{\mathit{VT}}
+
+\bentley{Make sure that terminology is consistent between elements and vertices}
+
+\subsection{Graph Structures and Functions}
 \label{sec:typed_graphs}
-% We begin this section by formally defining the graphs and graph instances used in our algorithm as fundamental structures.
-% \begin{definition}{Typed Graph}
-% \label{def:typed_graph}
-% A typed graph is a triple $\langle V,E,\tau\rangle$ where $V$ is a finite set of
-% vertices, $E\subseteq V\times V$ is a set of edges connecting the vertices and
-% $\tau:V\rightarrow Type$ is a typing function for the elements of V, where $Type$ is
-% a set of type names. Edges $(v,v')\in E$ are noted $v\rightarrow v'$. The set of
-% all typed graphs is called $TG$.
-% \end{definition}
-% 
-% Definition~\ref{def:typed_graph_homomorphism} describes when a typed graph is said to be homomorphic to another.
-% 
-% \begin{definition}{Typed Graph Homomorphism  / Typed Graph Instance}
-% \label{def:typed_graph_homomorphism}
-% 
-% Let $\langle V,E,\tau\rangle=g,\langle V',E',\tau'\rangle=g'\in TG$ be typed
-% graphs. There is a typed graph homomorphism between $g'$ and $g$ if and only if
-% there exists a graph homomorphism $f:V'\rightarrow V$ such that for all $v'\in
-% V$ we have that $\tau'(v')=\tau(f(v'))$. Note that, trivially, a typed graph
-% homomorphism is a formal homomorphism. We denote an \emph{injective} typed graph
-% homomorphism between typed graphs $g$ and $g'$ by $g' \vartriangleleft g$.
-% \end{definition}
-% 
-% In the text that follows we will use the terms \emph{typed graph homomorphism}
-% and \emph{typed graph instance} interchangeably.
-% 
-% Definition~\ref{def:typed_graph_strict_instance} is stricter than
-% definition~\ref{def:typed_graph_homomorphism}, and indicates that there is a
-% strict many-to-one type correspondence of the nodes of two graphs and their
-% relations.
-% 
-% \begin{definition}{Typed Graph Strict Instance}
-% \label{def:typed_graph_strict_instance}
-% 
-% Let $\langle V,E,\tau\rangle=g,\langle V',E',\tau\rangle=g'\in TG$ be typed
-% graphs. $g'$ is a typed graph strict instance of $g$, noted $g'
-% \blacktriangleleft g$, if and only if there exists a surjective
-% typed graph homomorphism between $g'$ and $g$.
-% 
-% \end{definition}
-% 
-% In order to simplify the subsequent development we will often
-% use the notation ${\langle V,E,\tau\rangle}^*$ to represent the transitive
-% closure of the edges of typed graph  $\langle V,E,\tau\rangle$.
 
-% \begin{definition}{Typed Graph}
-% \label{def:typed_graph}
-% A typed graph is a triple $\langle V,E,\tau\rangle$ where $V$ is a finite set of
-% vertices, $E\subseteq V\times V$ is a set of edges connecting the vertices and
-% $\tau:V\rightarrow Type$ is a typing function for the elements of V, where $Type$ is
-% a set of type names. Edges $(v,v')\in E$ are noted $v\rightarrow v'$. The set of
-% all typed graphs is called $TG$.
-% \end{definition}
-
-% \begin{definition}{Typed Graph and Indirect Typed Graph}
-% \label{def:typed_graph}
-% 
-% A typed graph is a triple $\langle V,E,\tau\rangle$ where $V$ is a finite set of
-% vertices, $E\subseteq V\times V$ is a finite set of directed edges connecting the vertices and $\tau:V\cup E \rightarrow Type\cup\{containment,reference\}$ is a typing function for the elements of V and E.
-Typing function $\tau$ is such that $\tau(v)\in Type$ if $v\in V$ and $\tau(e)\in \{containment,reference\}$\footnote{By using \emph{containment} and \emph{reference} as types for edges we allow modeling the different types of associations between the elements of a metamodel or a model.
-In particular, the fact that the subgraph of containment relations in a typed graph is acyclic models EMF containment associations.} if $e\in E$.
-Edges $(v,v')\in E$ are noted $v\rightarrow v'$.
-We furthermore impose that the graph $\langle V,\{v\rightarrow v'\in E|\tau(v\rightarrow v')=containment\}\rangle$ is acyclic.
-The set of all typed graphs is called $TG$.
-% 
-% An indirect typed graph is a 4-tuple $\langle V,E,T,Il\rangle$, where $\langle V,E,T\rangle$ is a typed graph and $Il\subseteq E$ is a set of edges called \emph{indirect links}.
-The set of all indirect typed graphs is called $ITG$.
-% \end{definition}
-
-\CatchFileBetweenTags{\typedgraphtext}{text/definitions}{typedgraphtext}{\typedgraphtext}
+\subsubsection*{Typed Graph}
+We will start by introducing the notion of typed graph.
+A typed graph is the essential object we will use throughout our mathematical development.
+Typed graphs will be used to formalise all the important graph-like structures we will present in this paper.
+A typed graph is a directed multigraph (a graph allowing multiple edges between two vertices) where vertices and edges are typed.
 
@@
 \label{def:typed_graph}
-\CatchFileBetweenTags{\typedgraph}{text/definitions}{typedgraph}{\typedgraph}
+A typed graph is a 6-tuple $\langle V,E,(s,t),\tau, \VT, \ET\rangle$ where:
+\begin{itemize}
+\item $V$ is a finite set of vertices
+\item $E$ is a finite set of directed edges connecting the vertices $V$
+\item $(s,t)$ is a pair of functions $s: E\rightarrow V$ and $t: E\rightarrow V$ that respectively provide the source and target vertices for each edge in the graph
+\item Function $\tau:V\cup E\rightarrow \VT \cup \ET$ is a typing function for the elements of $V$ and $E$, where $\VT$ and $\ET$ are disjoint finite sets of vertex and edge type identifiers and $\tau(v)\in \VT$ if $v\in V$ and $\tau(e)\in \ET$ if $e\in E$
+\item Edges $e\in E$ are noted $v\xrightarrow{e} v'$ if $s(e)=v$ and $t(e)=v'$, or simply e if the context is unambiguous
+\item The set of all typed graphs is called $\textsc{TG}$
+\item We define the empty graph to be a graph with $V = \emptyset$, $E = \emptyset$, and the other elements to be empty functions
+\end{itemize}
 \end{definition}
 
-\CatchFileBetweenTags{\typedgraphuniontext}{text/definitions}{typedgraphuniontext}{\typedgraphuniontext}
+\subsubsection{Vertex and Edge Types}
+
+Note that the our verification technique is geared toward model-driven engineering. Therefore, the types of vertices and edges in our typed graphs will be drawn from a particular \textit{metamodel}.
+
+Sample metamodels for our running example can be seen in figures above \bentley{Fix.}.
+The particular metamodel for a typed graph will be indicated in the formalization when required, as (potentially) different metamodels will be present for the input and output graphs in DSLTrans rules.
+
+We assume the existence of a number of functions in our formalization to aid the treatment of typing:
+
+\begin{itemize}
+\item $\mathit{isAbstract}: \mathit{VT} \rightarrow \{\mathit{true}, \mathit{false}\}$
+\begin{itemize}
+\item Where $\mathit{isAbstract}(\VT)$ returns $\mathit{true}$ iff $\VT$ is denoted as abstract (not able to be instantiated) by the metamodel, else $\mathit{false}$
+\end{itemize}
+
+\item $\mathit{isIndirect}: \mathit{ET} \rightarrow \{\mathit{true}, \mathit{false}\}$
+\begin{itemize}
+\item Where $\mathit{isIndirect}(\ET)$ returns $\mathit{true}$ iff $\ET$ is denoted as an \textit{indirect edge}, else $\mathit{false}$.
+The \textit{indirect} classification allows matching over indirect paths between vertices.
+This will be further clarified when required for our constructs.
+\end{itemize}
+
+\item $\mathit{matchesOver}: \{\VT \cup \ET\} \times \{\VT \cup \ET\} \rightarrow \{\mathit{true}, \mathit{false}\}$
+
+\begin{itemize}
+\item The purpose of this function is to assist in handling polymorphism in the metamodel
+\item $\mathit{matchesOver}(T, T) \rightarrow \mathit{true}$
+\item $\mathit{matchesOver}(T, T') \rightarrow \mathit{true}$ iff T' is a subclass of T in some defined partial ordering $\leq$ given by the metamodel
+\item Otherwise, $\mathit{false}$
+\end{itemize}
+\end{itemize}
+
+For simplification purposes, we will not represent edge cardinalities or containment relationships given by a metamodel in our notion of typed graph.
+Note that in fact we require these conditions to be relaxed to perform our graph rewriting.
+
+\subsubsection*{Typed Subgraph}
+We now define the useful notion of typed subgraph. As expected, a typed subgraph is simply a restriction of a typed graph to some of its vertices and edges.
+
+\begin{definition}{Typed Subgraph\\}
+\label{def:typedsubgraph}
+\defineggprime
+
+$g'$ is a typed subgraph of $g$, written $g'\sqsubseteq g$, iff $V'\subseteq V$, $E'\subseteq E$ and $\tau'=\tau |_{V'\cup E'}$.
+
+\end{definition}
+
+Given a typed graph $g\in \textsc{TG}$, will use the notation $Components(g)$ to describe the set of strongly connected typed graphs in $g$.
+Also, we will use the notation $g|_{t}$ to refer to the restriction of graph $g$ to its subgraph containing only edges of type $t$.
+
+\subsubsection*{Typed Graph Union}
+We now define how two typed graphs are united.
+A union of two typed graphs is trivially the set union of all the components of those two typed graphs.
+Note that we do not require the components of the two graphs to be disjoint, as in the following joint unions will be used to merge typed graphs.
 
@@
 \label{def:typed_graph_union}
-\CatchFileBetweenTags{\typedgraphunion}{text/definitions}{typedgraphunion}{\typedgraphunion}
+
+The typed graph union is the function $\sqcup :\textsc{TG}\times \textsc{TG}\rightarrow \textsc{TG}$ defined as:
+\begin{multline*}
+\big\langle V,E,(s,t),\tau,\VT,\ET\big\rangle\;\sqcup\;\big\langle V',E',(s',t'),\tau',\VT',\ET'\big\rangle=\\
+\big\langle V\cup V', E\cup E',(s\cup s', t\cup t'), \tau\cup \tau', \VT\cup \VT', \ET\cup ET'\big\rangle
+\end{multline*}
 \end{definition}
 
-\CatchFileBetweenTags{\typedgraphhomomorphismtext}{text/definitions}{typedgraphhomomorphismtext}{\typedgraphhomomorphismtext}
+Note: as a reviewer helpfully pointed out, we require that $s \cup s'$, $t \cup t'$, and $\tau \cup \tau'$ coincide on common elements. However, this can be assumed w. l. o. g..
+
+\bentley{Do we need typed graph intersection? Or to mention what it means for two graphs to be disjoint?}
+
+\subsubsection*{Typed Graph Homomorphism}
+For the formal development of our technique, we are interested in relations between typed graphs that are structure-preserving, i.e.
+homomorphisms.
+Homomorphisms between typed graphs preserve not only structure, but also the types of vertices and edges that are mapped.
+Note that, trivially, a typed graph homomorphism is a graph homomorphism.
 
@@
 \label{def:typed_graph_homomorphism}
-\CatchFileBetweenTags{\typedgraphhomomorphism}{text/definitions}{typedgraphhomomorphism}{\typedgraphhomomorphism}
+\defineggprime
+
+A typed graph homomorphism between $g$ and $g'$ is a function $f: f_v \cup f_e$ such that:
+\begin{itemize}
+\item $f_v: V\rightarrow V'$
+\item $f_e: E\rightarrow E'$
+\item $\forall v_1 \xrightarrow{e} v_2\in E$:
+\begin{itemize}
+\item Let $ v_1' = f_v(v_1), v_2' = f_v(v_2), e' = f_e(e)$
+\item $v_1', v_2' \in V', e' \in E'$
+\item $s'(e') = v_1'$, $t'(e') = v_2'$
+\item $\mathit{matchesOver}(\tau(v_1), \tau(v_1') \land \\\mathit{matchesOver}(\tau(v_2), \tau(v_2') \land\\ \mathit{matchesOver}(\tau(e), \tau(e')$
+\end{itemize}
+
+%, where $\tau(v_1)=\tau'(f(v_1))$, $\tau(v_2)=\tau'(f(v_2))$ and also $\tau(e)=\tau(e')$.
+\item The domain of $f$ is noted $Dom(f)$ and the co-domain of $f$ is noted $CoDom(f)$.
+\end{itemize}  
 \end{definition}
 
-Note that, trivially, a typed graph homomorphism is a graph homomorphism.
+\bentley{Maybe put `inj' and `sub' in the homomorphism to reduce number of symbols}
+ When an \emph{injective} typed graph homomorphism $f$ exists between $g$ and $g'$ we write $g \stackrel{f}{\vartriangleleft} g'$, or simply $g \vartriangleleft g'$ when the context is unambiguous.
+When a \emph{surjective} typed graph homomorphism $f$ exists between typed graphs $g$ and $g'$ we write $g \stackrel{f}{\blacktriangleleft} g'$, or also simply $g \blacktriangleleft g'$ in an unambiguous context.
 
-% \begin{lemma}{The Union of Two Typed Graph Homomorphisms is a Typed Graph Homomorphism}
-% 
-% Let $g=\langle V_g,E_g,st_g,\tau_g\rangle$, $g'=\langle
-% V'_g,E'_g,st'_g,\tau'_g\rangle$, $h=\langle V_h,E_h,st_h,\tau_h\rangle$,
-% $h'=\langle V'_h,E'_h,st'_h,\tau'_h\rangle$ be typed graphs, where $g$ and $h$
-% are potentially joint, as well as $g'$ and $h'$. Let also $f:V_g\rightarrow V'_g$ and
-% $i:V_h\rightarrow V'_h$ be a typed graph homomorphisms between $g$ and $g'$ and
-% $h$ and $h'$, respectively. Let us show that $f\cup i$ is a typed graph homomorphism between $g\sqcup g'$ and $h\sqcup h'$.
-% The proof is trivial if $g$ and $h$, as well as $h$ and $h'$ are disjoint. If $g$ and $h$ are joint over a vertex $v$, we know that $(f\cup i)(v)$
-% \end{lemma}
-% 
-% \begin{lemma}{The Union of Two Surjective Typed Graph Homomorphisms is a Surjective Typed Graph Homomorphism}
-% \levi{finish this proof}
-% \end{lemma}
+\subsubsection*{Typed Graph Edge Homomorphism}
 
-% \levi{changed the direction of the triangle for the surjective type graph homomorphism.
-Typed graph homomorphisms make sure that ALL connections between two nodes in the first graph also exists in the two mapped nodes of the second graph.}
+Our technique also requires a form of graph homomorphism that primarily focuses on matching the edges of the graphs.
+This homomorphism will be an injective match regarding the edges, but note that a particular vertex in the source graph may match onto multiple vertices in the target graph.
+\bentley{What's the term for this?}
 
-% \begin{definition}{Strict Typed Graph Homomorphism}
-% \label{def:strict_typed_graph_homomorphism}
-% 
-% Let $\langle V,E,\tau\rangle=g,\langle V',E',\tau\rangle=g'\in TG$ be typed
-% graphs. $g'$ is a typed graph strict instance of $g$, noted $g'
-% \blacktriangleleft g$, if and only if there exists a surjective
-% typed graph homomorphism between $g'$ and $g$.
+\begin{definition}{Typed Graph Edge Homomorphism\\}
+\label{def:typed_graph_edge_homomorphism}
+\defineggprime
 
-\CatchFileBetweenTags{\typedsubgraphtext}{text/definitions}{typedsubgraphtext}{\typedsubgraphtext}
+A typed graph edge homomorphism between $g$ and $g'$ is a function $h: h_v \cup h_e$ such that:
+\begin{itemize}
+\item $h_v: V'\rightarrow V$
+\begin{itemize}
+\item Note that $h_v$ is an function from a vertex in the target graph onto a vertex in the source graph
+\end{itemize}
+\item $h_e: E\rightarrow E'$
+\item $\forall v_1 \xrightarrow{e} v_2\in E$:
+\begin{itemize}
+\item Let $e' = h_e(e), v_1' = s'(e'), v_2' = t'(e')$
+\item $v_1', v_2' \in V', e' \in E'$
+\item $v_1 = h_v(v_1'), v_2 = h_v(v_2')$
+\item $\mathit{matchesOver}(\tau(v_1), \tau(v_1') \land \\\mathit{matchesOver}(\tau(v_2), \tau(v_2') \land\\ \mathit{matchesOver}(\tau(e), \tau(e')$
+\end{itemize}
 
-\begin{definition}{Typed Subgraph\\}
-\label{def:typedsubgraph}
-\CatchFileBetweenTags{\typedsubgraph}{text/definitions}{typedsubgraph}{\typedsubgraph}
-
-% An indirect typed graph $\langle V',E',\tau',Il\rangle \in ITG$ is an indirect typed
-% subgraph of a typed graph $\langle V,E,\tau\rangle \in TG$, written $\langle
-% V',E',\tau',Il\rangle \lhd \langle V,E,\tau\rangle$ iff:
-% 
-% \begin{enumerate}
-% \item $\langle V',E'\setminus Il,\tau'\rangle \blacktriangleleft \langle V, E,\tau\rangle$
-% \item if $v_i\rightarrow v_i'\in Il$ then there exists $v\rightarrow v'\in E_{c}^{*}$ where $\tau(v_i)=\tau(v)$, $\tau(v_i')=\tau(v')$ and $E_{c}^{*}$ is obtained by the transitive closure of $E_{c}=\{v\rightarrow v'\in E|\tau(v\rightarrow v')=containment\}$.
-% \end{enumerate}
+\end{itemize}  
 \end{definition}
 
-\CatchFileBetweenTags{\typedgraphisomorphismtext}{text/definitions}{typedgraphisomorphismtext}{\typedgraphisomorphismtext}
+Again, we wish to highlight the fact that this homomorphism is different than Definition~\ref{def:typed_graph_homomorphism} in the vertex matching function.
+A vertex in the target graph matches onto a vertex in the source graph, with multiple target vertices matching onto the same source vertex.
+\bentley{Create diagram to illustrate}
+
+This unusual homomorphism allows our technique to `split' our source graph over multiple locations in the target graph.
+This is critical, as our target graphs are constructed in a way that allows for vertex duplication.
+That is, two vertices of a particular type in the target graph may only represent one underlying element, as discussed in Section~\ref{sec:abstraction_relation}.
+Constructing this morphism allows our prover implementation to avoid explicitly creating `disambiguated' graphs where these duplications are resolved.
+
+\subsection*{Typed Graph Isomorphism}
+Two typed graphs are said to be isomorphic if they have exactly the same shape and related vertices and edges have the same type.
 
@@
 \label{def:typed_graph_isomorphism}
-\CatchFileBetweenTags{\typedgraphisomorphism}{text/definitions}{typedgraphisomorphism}{\typedgraphisomorphism}
+\defineggprime
+
+We say that $g$ and $g'$ are isomorphic, written $g\cong g'$, iff there exists a bijective typed graph homomorphism $f:V\rightarrow V'$ such that $f^{-1}:V'\rightarrow V$ is a typed graph homomorphism.
 \end{definition}
-% \begin{definition}{Typed Graph Connected Component}
-% \label{def:typed_graph_connected_component}
-% 
-% Let $g \in \textsc{TG}$ be a typed graph.
-We say that $\langle V,E,st,\tau\rangle \sqsubseteq g$ is a connected component of $g$ if for any pair of vertices $v,v'\in V$ there exists a path between $v$ and $v'$, where the direction of the edges between $v$ and $v'$ does not matter.
-% \end{definition}
 
-% \begin{definition}{Typed Graph Instance}
-% \label{def:typed_graph_instance}
-% 
-% Let $\langle V,E,\tau\rangle=g,\langle V',E',\tau\rangle=g'\in TG$ be typed graphs.
-% $g'$ is a typed graph instance of $g$, written $g'\Vdash g$, iff for all $v_{1}'\rightarrow v_{2}'\in E'$ there is a $v_{1}\rightarrow v_{2}\in E$ such that $\tau(v_{1}')=\tau(v_1)$, $\tau(v_{2}')=\tau(v_2)$ and $\tau(v_{1}'\rightarrow v_{2}')=\tau(v_{1}\rightarrow v_{2})$.
-% \end{definition}
+\subsubsection*{Transitive Closure}
 
-%Notice that we only enforce that connections between vertices of $g'$ must exist also in $g$ and have the same type.
+DSLTrans matching constructs allow for the matching of indirect links, where as described in Section~\ref{subsec:DSLTrans_constructs}, rules may match over indirect paths between elements.
+This transitive closure allow for the explicit creation of all edges implied by these indirect links to aid in matching.
 
-% \levi{Note that each edge between two nodes in the first graph has to exist in the second graph and vice-versa.}
-% \levi{in the sections that follow we will formally elaborate on relevant concepts that are introduced.}
+\begin{definition}{Transitive Closure\\}
+\label{def:instance_closure}
 
-\paragraph{\textbf{Notation:}}
-\CatchFileBetweenTags{\notationformalbackground}{text/definitions}{notationformalbackground}{\notationformalbackground}
+Let $g$ be a graph $\langle V,E,(s, t),\tau\rangle \in \textsc{TG}$. Then the transitive closure $g^{*} = \langle V,E',(s, t),\tau\rangle \in \textsc{TG}$, where:
 
-\subsection{Metamodel and Model-Related Constructs}
+$E' = E \cup $ the transitive closure of $\big\{v\xrightarrow{e}v' | \mathit{isIndirect(e)} \big\}$
+\end{definition}
+
+Given a graph, its transitive closure includes, besides the original graph, all the edges belonging to the transitive closure of indirect links in that graph.
+Note that these transitive edges are given by the function $\mathit{isIndirect}$.
+
+\bentley{I believe that the indirect edges don't need to be removed. There's a bit of handwaving, so maybe there are edge duplicates?}
+
+In the definitions that follow we will use the $*$ notation, as in \cref{def:instance_closure}, to denote the transitive closure of our structures.
+
+\subsection{DSLTrans Constructs}
 \label{sec:DSLTrans_formal}
 
-% This section will formally describe the DSLTrans constructs that are required in
-% our property-proving algorithm. The definitions presented strongly rely on the concepts of typed graphs, just introduced in \cref{sec:typed_graphs}.
+This section will detail the abstract syntax of the constructs involved in a DSLTrans transformation.
 
-We will start by introducing the notion of \emph{metamodel}, which in DSLTrans is used to type the input and output models of a DSLTrans transformation.
-Two metamodels, the \emph{organization language} and \emph{gender language} are depicted in \cref{fig:squadmetamodel}.
+As discussed in Section~\ref{subsec:DSLTrans_syntax}, DSLTrans transformations are composed of rules arranged in layers.
 
-\begin{definition}{Metamodel\\}
-\label{def:metamodel}
-\CatchFileBetweenTags{\metamodel}{text/definitions}{metamodel}{\metamodel}
-\end{definition}
+\subsubsection*{DSLTrans Transformation Rule}
 
-\CatchFileBetweenTags{\metamodeltext}{text/definitions}{metamodeltext}{\metamodeltext}
+A transformation rule is the elemental block of a DSLTrans transformation. Several transformation rules can be observed in the Police Station transformation in \cref{fig:dsltransformation}.
 
-\begin{definition}{Expanded Metamodel\\}
-\label{def:expanded_metamodel}
-\CatchFileBetweenTags{\expandedmetamodel}{text/definitions}{expandedmetamodel}{\expandedmetamodel}
-\end{definition}
+A transformation rule includes a non-empty match pattern and a non-empty apply pattern.
+This is also known in the model transformation literature as a rule's \emph{left hand side} and \emph{right hand side}.
+A match pattern can include indirect links that are used to transitively match containment relations in a model.
+The apply pattern of a rule always contains at least one apply element that is not connected to a backward link or an edge, meaning in practice that a rule will always produce something and not only match.
+An apply pattern does not include indirect links as it is used only for the construction of parts of instances of a metamodel.
 
-\CatchFileBetweenTags{\expandedmetamodeltext}{text/definitions}{expandedmetamodeltext}{\expandedmetamodeltext}
+A rule may also contain a negative application condition (NAC) which if matched over the source graph prevents application of the rule.
 
-\begin{definition}{Metamodel Instance\\}
-\label{def:metamodelinstance}
-\CatchFileBetweenTags{\metamodelinstance}{text/definitions}{metamodelinstance}{\metamodelinstance}
-\end{definition}
+A transformation rule also includes backward links, as informally introduced in \cref{subsec:DSLTrans_constructs}.
+As described in that section, backward links define dependencies between rules.
+\bentley{Expand}
 
-\CatchFileBetweenTags{\metamodelinstancetext}{text/definitions}{metamodelinstancetext}{\metamodelinstancetext}
+\begin{definition}{DSLTrans Transformation Rule\\}
+\label{def:transformation_rule}
 
-\begin{definition}{Containment Transitive Closure\\}
-\label{def:instance_closure}
-\CatchFileBetweenTags{\instanceclosure}{text/definitions}{instanceclosure}{\instanceclosure}
+A DSLTrans transformation rule is a four-tuple $\big\langle \mathit{NAC}$, $\mathit{Match}$, $\mathit{Apply}$, $\mathit{backward}\big\rangle$, where:
+
+\begin{itemize}
+\item $\mathit{NAC}$, $\mathit{Match}, \mathit{Apply} \in \textsc{TG}$
+\item $\mathit{Match}$ and $\mathit{Apply} $ are non-empty and are disjoint
+\item $\mathit{NAC}$ may be an empty graph, and is disjoint from both $\mathit{Match}$ and $\mathit{Apply} $
+\item $\mathit{backward} = \{E_{back}, (s_{back}, t_{back})\}$
+\end{itemize}
+
+Note that when we require an element of the $\mathit{Match}$, $\mathit{Apply}$, or $\mathit{NAC}$ graphs, such as the vertices, we will index the required element.
+For example, the vertices for the \textit{Match} graph will be $V_{\textit{Match}}$.
+
+\begin{itemize}
+\item $E_{back}$ contains the backward links
+\begin{itemize}
+\item $E_{back} $ is disjoint from $E_{Match}$,  $E_{Apply}$, and $E_{NAC}$
+\end{itemize}
+\item $(s_{back}, t_{back})$ is a pair of functions $s_{back}: E_{back}\rightarrow V_{\textit{Apply}}$ and $t_{back}: E_{back}\rightarrow V_{\textit{Match}}$ that respectively provide the source and target vertices for each backward link
+\begin{itemize}
+\item Note the source of backward links is a vertex in $V_{\textit{Match}}$ while the target is a vertex in $V_{\textit{Apply}}$
+\end{itemize}
+\end{itemize}
+
+\textsc{Rules} is the set of all rules.
 
@@
 
-\CatchFileBetweenTags{\instanceclosuretext}{text/definitions}{instanceclosuretext}{\instanceclosuretext}
+We additionally impose that for the well-formedness of a DSLTrans rule, there always exists an element or edge to be created in the \textit{Apply} graph to be created.
+That is, either there are edges to be created in the \textit{Apply} graph, or there is a vertex in the \textit{Apply} graph that is not the source of a backward link.
 
-\begin{definition}{Model\\}
-\label{def:model}
-\CatchFileBetweenTags{\model}{text/definitions}{model}{\model}
+\begin{itemize}
+\item $E_{\textit{Apply}} \neq \emptyset \lor$
+\item $ \exists v \in V_{\textit{Apply}} | \big\{\forall e \in E_{back}: s_{back}(e) \neq v \big\}$
+\end{itemize}
+
+\subsubsection*{DSLTrans Layer and Transformation}
+
+\cref{def:layer} and \cref{def:transformation} formalise the abstract syntax of a model transformation, introduced in \cref{sec:dsltrans}.
+An example of a model transformation can be observed in \cref{fig:dsltransformation}, the Police Station transformation.
+As expected, a DSLTrans transformation is composed of a sequence of layers where each layer is composed of a set of rules.
+
+\begin{definition}{Layer\\}
+\label{def:layer}
+A layer is a finite set of transformation rules: $\{r_0, r_1, \dots, r_n | r_i \in \textsc{Rules}\}$.
+
+The set of all layers is denoted $\textsc{Layers}$. 
 \end{definition}
 
-\CatchFileBetweenTags{\modeltext}{text/definitions}{modeltext}{\modeltext}
+Note that the order of the rules within the layer does not matter, due to the semantics of DSLTrans rule execution.
+
+\begin{definition}{DSLTrans Transformation\\}
+\label{def:transformation}
+
+A DSLTrans transformation is a finite list of layers denoted $\{l_0, l_1, \dots, l_n | l \in \textsc{Layers}\}$. Note that the order of layers in a transformation is important.
+
+The set of all transformations is denoted $\textsc{Transforms}$.
+
+\end{definition}
+
+\subsubsection*{Input-Output Model}
+
+To describe the semantics of a DSLTrans model transformation, we must define an \textit{input-output model} construct.
+This input-output model allows the representation of intermediate operational states during the execution of a model transformation.
+
+The structure of a input-output model is intentionally very similar to a DSLTrans rule.
+There is one typed graph representing the input model and another typed graph representing the output model.
+As well, the construct contains a set of edges, named \emph{traceability links}, for keeping a history of which elements in the output model originated from which elements in the input model.
 
@@
 \label{def:input_output_model}
-\CatchFileBetweenTags{\inputoutputmodel}{text/definitions}{inputoutputmodel}{\inputoutputmodel}
+
+An input-output model rule is a three-tuple $\big\langle \mathit{Input}$, $\mathit{Output}$, $\mathit{trace}\big\rangle$, where:
+
+\begin{itemize}
+\item $\mathit{Input}, \mathit{Output} \in \textsc{TG}$
+\item $\mathit{Input}$ and $\mathit{Output} $ may be empty and are disjoint
+\item $\mathit{trace} = \{E_{trace}, (s_{trace}, t_{trace})\}$
+\begin{itemize}
+\item $E_{trace}$ contains the traceability links
+\item $E_{trace} $ is disjoint from $E_{Input}$ and $E_{Output}$
+\item $(s_{trace}, t_{trace})$ is a pair of functions $s_{trace}: E_{trace}\rightarrow V_{\textit{Output}}$ and $t_{trace}: E_{trace}\rightarrow V_{\textit{Input}}$ that respectively provide the source and target vertices for each traceability link
+\end{itemize}
+
+\end{itemize}
+
+Let $\mathit{IOM}$ be the set of all input-output models.
+
+Note that the $\VT$ and $\ET$ for input and output will come from the input and output metamodels. \bentley{Expand}
+
+\bentley{Do we need to have V and E for all vertices and edges in the IOM? I don't think we talk about all vertices or edges in an IOM.}
+Additionally we have that $V=V'\cup V''$, $E\subseteq E'\cup E''$ and $\tau\subseteq \tau'\cup \tau''$, where the co-domain of $\tau$ is the union of the co-domains of $\tau'$ and $\tau''$ and the set $\{trace\}$.
+
 \end{definition}
 
-%Change if needed
-\CatchFileBetweenTags{\inputoutputmodeltext}{text/definitions}{inputoutputmodeltext}{\inputoutputmodeltext}
+\subsection{Transformation Semantics}
+\label{sec:DSLTrans_semantics}
 
-\begin{definition}{Metamodel Pattern and Indirect Metamodel Pattern\\}
-\label{def:metamodel_pattern} 
-\CatchFileBetweenTags{\metamodelpattern}{text/definitions}{metamodelpattern}{\metamodelpattern}
+This section will discuss the semantics of a DSLTrans transformation. Given an input model and a transformation, an output layer will be produced through the repeated execution of rules and layers.
+
+\subsubsection*{Execution of a DSLTrans Rule}
+
+We will now address the execution of a rule in the DSLTrans language.
+
+We will base our explanation on double pushouts, diagrammed in Figure~\ref{fig:dpo}.
+\bentley{Single or double pushouts?}
+
+\begin{figure}[h!] \centering
+\includegraphics[width=0.44\textwidth]{figures/formal/dpo}
+	\caption{Double pushout approach}
+	\label{fig:dpo}
+\end{figure}
+
+We explain the components:
+\begin{itemize}
+\item L: The matcher of the rule
+\item R: The rewriter of the rule
+\item G: The graph to be matched/rewritten
+\item H: The rewritten graph
+\end{itemize}
+
+Note that as it is not possible in DSLTrans rules to remove elements, $L = K$, and $G = D$.
+
+\subsubsection*{Matcher of a Transformation Rule}
+
+To create the actual matcher construct that will be matched during a rule's execution, the rule's \textit{Match} graph is combined with backward links, as well as any \textit{Apply} vertices connected to the backward links.
+
+Note that the rule's NAC will also be considered in the matching of a rule.
+
+\begin{definition}{Matcher of a Transformation Rule\\}
+\label{def:matcher_transformation_rule}
+
+Let the transformation rule $r = \big\langle \mathit{NAC}$, $\mathit{Match}$, $\mathit{Apply}$, $\mathit{backward}\big\rangle$.
+
+Let $\mathit{Match}\star$ be the transitive closure of $\mathit{Match}$ to handle any indirect edges.
+
+\bentley{Typed graph? Or copy the rule's structure? If it is a rule, note that it will not be well-formed. However, this is preferable to defining another structure to use}
+
+We define $r$'s matcher, noted $\lceil r \rceil$, to be a typed graph six-tuple $\langle V,E,(s,t),\tau, \VT, \ET\rangle$, where:
+\begin{itemize}
+\item $V = V_{Match\star} \cup \big\{v \in V_{Apply} | \{\exists e \in E_{back} | s_{back}(e) = v\}\big\}$
+\item $E = E_{Match\star} \cup E_{back}$
+\item $s = s_{Match\star} \cup s_{back}$, $t = t_{Match\star} \cup t_{back}$
+\item $\tau = \tau_{Match\star} \cup \tau_{Apply}$
+\item $\VT = \VT_{Match\star} \cup \VT_{Apply}$
+\item $\ET = \ET_{Match\star} \cup \ET_{Apply}$
+\end{itemize}
+
 \end{definition}
 
-\CatchFileBetweenTags{\metamodelpatterntext}{text/definitions}{metamodelpatterntext}{\metamodelpatterntext}
+\subsubsection*{Rewriter of a Transformation Rule}
 
-\subsection{Syntactic Transformation Constructs}
-\label{sec:DSLTrans_syntax}
+To continue with the double pushout approach, we must construct the rewriting (or replacement) graph.
 
-This section will detail the abstract syntax of the constructs involved in a DSLTrans transformation.
+The construction of this rewriter is essentially the same as the underlying rule.
+However, to support traceability, we require that backward links in the rule be converted into traceability links, and that new traceability links be created between all match and apply vertices.
 
-\begin{definition}{Transformation Rule\\}
-\label{def:transformation_rule}
-\CatchFileBetweenTags{\transformationrule}{text/definitions}{transformationrule}{\transformationrule}
+\begin{definition}{Rewriter of a Transformation Rule\\}
+\label{def:rewriter_transformation_rule}
+
+Let the transformation rule $r = \big\langle \mathit{NAC}$, $\mathit{Match}$, $\mathit{Apply}$, $\mathit{backward}\big\rangle$.
+
+Let $\mathit{Match}\star$ be the transitive closure of $\mathit{Match}$ to handle any indirect edges.
+
+We define $r$'s rewriter, noted $\lfloor r \rfloor$, to be a input-output model $\big\langle \mathit{Input}$, $\mathit{Output}$, $\mathit{trace}\big\rangle$, where:
+
+\begin{itemize}
+\item $ \mathit{Input} = \mathit{Match\star}$
+\item $ \mathit{Output} = \mathit{Apply}$
+\end{itemize}
+
+We then modify $\mathit{trace}$ to contain the appropriate traceability links
+
+\begin{itemize}
+\item $E_{trace} = E_{backward} \cup \big\{e | \{$for each pair $m_i, a_j$ where $m_i \in V_{Match}, a_j \in V_{Apply}\}\big\}$
+\item $s_{trace} = s_{backward} \cup \big\{a_j | \{$for each pair $m_i, a_j$ where $m_i \in V_{Match}, a_j \in V_{Apply}\}\big\}$
+\item $t_{trace} = t_{backward} \cup \big\{m_i | \{$for each pair $m_i, a_j$ where $m_i \in V_{Match}, a_j \in V_{Apply}\}\big\}$
+\end{itemize}
+
+\bentley{Very messy}
 \end{definition}
 
-\CatchFileBetweenTags{\transformationruletext}{text/definitions}{transformationruletext}{\transformationruletext}
+\subsubsection*{Application of a Transformation Rule (continued)}
 
-\begin{definition}{Matcher of a Transformation Rule\\}
-\label{def:back_match_transformation_rule}
-\CatchFileBetweenTags{\backmatchtransformationrule}{text/definitions}{backmatchtransformationrule}{\backmatchtransformationrule}
+\bentley{Fix up morphism notation. Is $\rightarrow$ good enough?} 
+Recall that the interface graph K is equivalent to L due to the lack of deletion in DSLTrans rules. Thus, the l homomorphism is identity. The D correspondence graph will therefore be equivalent to G.
+
+As well, we consider r to be a trivial homomorphism between the $\mathit{Match} \rightarrow \mathit{Input}$, $\mathit{Apply} \rightarrow \mathit{Output}$, and $\mathit{backward} \rightarrow \mathit{trace}$.
+
+We also note here that a rule cannot be applied if there is a homomorphism between the rule's NAC and G.
+
+m is the typed graph homomorphism between the matcher of a rule $r$, denoted ($\lceil r \rceil$) and the $\mathit{Input}$ component of the input-output model
+
+$n$ and $g$ will be homomorphism between their respective graphs. \bentley{This needs to be explained further. Maybe with a diagram?}
+
+\subsubsection*{DSLTrans Layer Execution}
+
+Now that the application of a particular DSLTrans rule has been explained, the next definition defines how a DSLTrans transformation layer executes.
+Recall that a layer is composed of rules, and acts on the current input-output model.
+
+\begin{definition} {DSLTrans Layer Execution}
+\label{def:dsltrans_layer_execution}
+
+The execution of a DSLTrans layer can be defined as a function as follows:
+
+\textit{applyLayer} : IOM $\times$ Layer $\rightarrow$ IOM
+
+Let $l = \{r_0, r_1, \dots, r_n | r_i \in \textsc{Rules}\}$, $l \in \textsc{Layers}$.
+
+Let $\mathit{IOM}_{Input}$, $\mathit{IOM}_{Output} \in \mathit{IOMs}$.
+
+$\mathit{IOM}_{Output}$ will be the result of applying the production \textit{p} as defined below to $\mathit{IOM}_{Input}.$
+
+We refer to the double-pushout literature for the creation of \textit{p}.
+Specifically, we refer to the definition of parallel direct derivations.
+This p will be created out of the productions created for each of the rules $r_i$ in the layer.
+
+$p = \big\langle (p_1, {in}^1), \dots, (p_k, {in}^k) \big\rangle : (L \xleftarrow{l} L \xrightarrow{r} R)$
+
+This collection of the rules is possible because DSLTrans rules do not delete anything. Essentially, a union is taken of the L, K, R components of each production.
 \end{definition}
 
-\Cref{def:back_match_transformation_rule} \CatchFileBetweenTags{\backmatchtransformationruletext}{text/definitions}{backmatchtransformationruletext}{\backmatchtransformationruletext}
+\begin{figure}[h!] \centering
+\includegraphics[width=0.44\textwidth]{figures/formal/parallel_productions}
+	\caption{Parallel production \textit{p}}
+	\label{fig:parallel_productions}
+\end{figure}
 
-\begin{definition}{Expanded Transformation Rule\\}
-\label{def:transformation_rule_expansion}
-\CatchFileBetweenTags{\expandedtransformationrule}{text/definitions}{expandedtransformationrule}{\expandedtransformationrule}
-\end{definition}
-\CatchFileBetweenTags{\expandedtransformationruletext}{text/definitions}{expandedtransformationruletext}{\expandedtransformationruletext}
+\cref{def:dsltrans_layer_execution} is the core of DSLTrans' semantics, in which we build the result of executing a layer of a DSLTrans transformation.
+Many model transformation languages are based on graph rewriting, where the result of each rule rewrite is immediately usable by all other rules.
+In DSLTrans the result of executing one layer in DSLTrans is totally produced before the input to the layer is changed.
+This is enforced in \cref{def:dsltrans_layer_execution} by the fact that the layer production is composed of a union of the rule's productions.
+Rules belonging to the same layer are thus forced to execute independently, as described in the section on DSLTrans semantics.
 
-\begin{definition}{Layer, Model Transformation\\}
-\label{def:layer_transformation}
-\CatchFileBetweenTags{\layertransformation}{text/definitions}{layertransformation}{\layertransformation}
+\bentley{How does this handle different multiplicity of rule application?}
+
+\subsubsection*{DSLTrans Transformation Execution}
+
+Our final definition is for the execution of a DSLTrans transformation. Essentially, it is the chaining of layer executions applied on an input-output model.
+
+We will begin by discussing the conditions for executing a model transformation on an input model.
+
+\paragraph{Input Conditions}
+
+For a well-formed transformation execution, we require the \textit{input-output model} in the domain to contain only an input graph.
+Recall that an \textit{input-output model} is the three-tuple $\big\langle \mathit{Input}$, $\mathit{Output}$, $\mathit{trace}\big\rangle$.
+We therefore require that both the $\mathit{Output}$ and $\mathit{trace}$ components in the domain IOM to be empty.
+This input-output model thus represents the first step of the transformation, where no rule has been executed yet.
+
+The input-output model in the co-domain will contain this input graph, as well as the output graph and traceability links produced by the execution of the transformation.
+Let $\textsc{Execs}_t$ be the set of all well-formed input-output models produced for transformation $t$.
+
+\begin{definition} {Execution of a DSLTrans Transformation\\}
+\label{def:DSLTrans_transformation_execution}
+
+The execution of a model transformation is a function $\mathit{applyTransformation}: \mathit{IOM} \times \mathit{Transforms} \rightarrow \mathit{IOM}$.
+
+This function is a chaining of executions of those layers within the transformation.
+
+Recall the application of a layer is \textit{applyLayer} : IOM $\times$ Layer $\rightarrow$ IOM. Therefore, the application of a transformation is:
+
+$\mathit{IOM}$ as input to chaining of layer executions \bentley{Is this function chaining? I don't know how to write the equation for this.}
+
 \end{definition}
 
-\CatchFileBetweenTags{\layertransformationtext}{text/definitions}{layertransformationtext}{\layertransformationtext}
+While the execution of the rules belonging to a layer happens in parallel, the execution of the layers of a transformation happens sequentially.
+As per \cref{def:DSLTrans_transformation_execution}, the input-output model that is the output of executing a given layer is passed onto the next layer as input.
+The final input-output model created will be the result of the model transformation.
 
-\paragraph{\textbf{Notation:}}
+\subsection{Confluence and Termination Properties}
 
-\CatchFileBetweenTags{\notationtext}{text/definitions}{notationtext}{\notationtext}
+We now prove two important properties about executions of transformations expressed in the subset of DSLTrans presented in this paper: \emph{confluence} and \emph{termination}.
+The proofs are provided at a high level, given the fact that DSLTrans essentially enforces both these properties by construction of the semantics of DSLTrans.
 
-% We naturally extend the notion of union (definition~\ref{def:typed_graph_union})
-% to models (definition~\ref{def:metamodelmodel}), match-apply models
-% (definition~\ref{def:match_apply_model}) and transformation rules 
-% (definition~\ref{def:transformation_rule}). We also extend the notion of
-% indirect typed subgraph (definition~\ref{def:indirecttypedsubgraph}) to
-% transformation rules (definition~\ref{def:transformation_rule}) and match-apply
-% models (definition~\ref{def:match_apply_model}). Finally, we extend the notion
-% of typed graph equivalence (definition~\ref{def:typed_graph_equivalence}) to
-% transformation rules (definition~\ref{def:transformation_rule}).
+\begin{proposition}{Confluence}
 
-% \subsubsection{Transformation Language Semantics}
-% 
-% We will now address the semantics of the DSLTrans language.
-We will start by defining a match function that, given an input-output model and a transformation rule, returns all subgraphs of that input-output model where the rule's match pattern (including backward links) is found.
-% 
-% % \levi{Need to be careful throughout all this section to understand whether we model the match and apply functions as functions or relations.
-Because of the non-deterministic generation of new vertices in the definition of apply function~\ref{def:apply_function}, relations are used in the definitions that follow to evaluate a transformation.
-This is an issue of the formalism used to model, should we make it explicit?}
-% 
-% \begin{definition}{Match Function}
-% \label{def:match_function}
-% 
-% Let $m_{in}\in \textsc{Iom}^{sr}_{tr}$ be a input-output model and $rl\in \textsc{Rule}^{sr}_{tg}$ be a transformation
-% rule. The $match : \textsc{Iom}^{sr}_{tg}\times \textsc{Rule}^{sr}_{tg}\rightarrow
-% \mathcal{P}(\textsc{Iom}^{sr}_{tg})$ function is defined as follows: $$match_{rl}(m_{in})= \big\{ glue_{noind}\;|\;
-% glue\sqsubseteq m_{in}^{*} \land glue \cong rl^{backM}_{trace}\big\}$$
-% 
-% where $glue$ is an input-output model.
-Additionally we have that: $rl_{trace}$ is a version of $rl$ where edges of type \emph{backward} have been replaced by edges of type \emph{trace}; $glue_{noind}$ is a version of $glue$ where the indirect links have been removed.
-% % \levi{all backward links are replaced by trace links such that the matching using subgraph isomorphim can work.}
-% % 
-% % \levi{Where $rl^{-}$ is rule $rl$ without nodes in the apply part that are not connected to \emph{trace links}, as well as links adjacent to those nodes.
-$mg^{-ind}$ is the match graph without indirect links that come from the rule, given those cannot be part of the intermediate $MAM$ match result since that result is going to be united with the original $MAM$ (this was incorrect in the original SLE paper).}
-% % \levi{The matcher function returns all graphs of the input MAM that are matched by a rule, including its backward links.
-Note that to find a subgraph ($\sqsubseteq$) of the input MAM that is isomorphic ($\cong$) to the rule being matched because the graph rewriting is achieved by joint union of the match graph extended by the new rewrite node, and the original MAM.}
-% % \levi{TODO: review the $mg^{-ind}$ and the $rl^{-}$ notations.}
-% 
-% % Due to the fact that the $\cong$ relation is based on the notion of graph
-% % isomorphism, permutations of the same match result may exist in the
-% % $\big\{g\;|\; g\lhd m \land g \cong strip(tr)\big\}$ set. The --- undefined ---
-% % $remove:\mathcal{P}(TR^{s}_{t})\rightarrow \mathcal{P}(TR^{s}_{t})$ function is such that
-% % it removes such undesired permutations.
-% 
-% %The $strip:TR^{s}_{t}\rightarrow TR^{s}_{t}$ function is such that
-% %$$strip(\big\langle V,E,T,\langle V_m,E_m,T_m,s\rangle,\langle V_a,E_a,T_a,t\rangle,Bl,Il\big\rangle) = \big\langle V',E',T,\langle V_m,E_m,T_m\rangle,\langle V'_a,E'_a,T_a\rangle,Bl,Il\big\rangle$$
-% %\begin{center}
-% %where $(v'\in V'_a\Rightarrow v\rightarrow v'\in Bl)\;\land\;(v\rightarrow v'\in E'_a\Rightarrow (v\rightarrow v'\in E_m \land \{v,v'\}\subseteq V'_a))$
-% %\end{center}
-% \end{definition}
-% 
-% The match function in definition~\ref{def:match_function} looks for subgraphs ($glue\sqsubseteq m_{in}^{*}$) of an input-output model that are isomorphic to the backward matcher of the given transformation rule ($glue \cong rl^{backM}_{trace}$).
-Note that the containment transitive closure of the input-output model ($m_{in}^{*}$) is considered such that indirect links in the rule can looked for.
-Additionally, indirect links need to be removed from the input-output models resulting from the match function ($glue_{noind}$).
-This is do because indirect links are not part of the original input model but rather an auxiliary structure.
-% 
-% Let us now turn our attention to the apply function in definition~\ref{def:apply_function}.
-Its role is extend each of the sub-models found by the a match function for a given input-output model and a given transformation rule, such that they become isomorphic to the complete rule (minus its backward links).
-This process effectively creates the new objects and relations specified in the apply part of the rule for each of the sub-models found by the match function.
-% 
-% \begin{definition}{Apply Function}
-% \label{def:apply_function}
-% 
-% Let $m_{glue}\in \textsc{Iom}^{sr}_{tg}$ be a input-output model and $rl\in \textsc{Rule}^{sr}_{tg}$ a
-% transformation rule. The $apply : \textsc{Iom}^{sr}_{tg}\times \textsc{Rule}^{sr}_{tg}\rightarrow \textsc{Iom}^{sr}_{tg}$ function
-% is defined as follows: 
-% 
-% $$apply_{rl}(m_{in})=\bigsqcup_{m_{glue}\in match_{rl}(m_{in})}trace_{a_{\Delta}}(m_{glue}\sqcup a_{\Delta})$$
-% 
-% \begin{center}
-% where $a_{\Delta}\in \textsc{Iom}^{sr}_{tg}$ is such that $m_{glue} \sqcup a_{\Delta}\cong rl_{noind}$.\\
-% \end{center}
-% 
-% We impose that any instance of $a_\Delta$ is always disjoint from the $m_{in}$
-% input-output model and also that any two instances of $a_\Delta$ used in the
-% large union are always disjoint. Partial function $trace:\textsc{Iom}\times
-% \textsc{Iom} \nrightarrow \textsc{Iom}$ is such that $trace_{\langle
-% V_{\Delta},E_{\Delta},st_{\Delta},\tau_{\Delta}\rangle}(\langle
-% V,E,st,\tau\rangle) = \langle V,E',st',\tau'\rangle$ where we have that
-% $E\subseteq E'$, $st\subseteq st'$ (using a light notational abuse for the $s\subseteq s'$
-% and $t\subseteq t'$), $\tau \subseteq \tau'$ and if $v_1\xrightarrow{e}
-% v_2\in E'\setminus E$ then $v_1\in Output(V_{\Delta})$, $v_2\in Input(V)$ and
-% $\tau'(e)=trace$. Finally, $rl_{noind}$ is a version of $rl$ where indirect
-% links have been removed.
-% % More concretely, if we have $mam = $\big\langle V,E,st,\tau, Input,Output\big\rangle\in MAM^{sr}_{tg}$ where $Input=\langle V',E',st',\tau'\rangle\in MODEL^{sr}$ and $Output=\langle
-% % V'',E'',st'',\tau''\rangle\in MODEL^{tg}$, then $trace(mam) = \big\langle V,E,st,\tau''',Input,Output\big\rangle$ where $E'\subseteq E$ and $v\rightarrow E'\setminus E$ 
-% \end{definition}
-% % 
-% % it builds traceability edges between vertices of the output model that were newly built by the $apply$ function (the vertices in ${mg}_{\Delta}$) and all the vertices from the input part of the input-output model.
-It is necessary to remove the indirect links from the rule because otherwise they would be created during the rewriting by the apply function.
-% 
-% In definition~\ref{def:apply_function} $a_\Delta$ is an input-output model that contains an instance of the target metamodel.
-These instances are created by rule $rl$ and are used to extend the sub-models found by the match function.
-The $trace$ function builds traceability edges between vertices of the output model in $a_{\Delta}$ and all the vertices from the input part of a model fragment found by the match function.
-% 
-% Note that, because we do not pose any constraints on $a_\Delta$ other than the fact that its union with the sub-model $m$ is isomorphic to $rl_{noind}$, the $a_\Delta$ variable can always be satisfied by an unlimited amount of input-output models.
-In order to avoid an infinite amounts of results when a transformation rule is executed, in what follows we will consider transformation results differ only up to typed graph isomorphism.
-% 
-% %  Note that because the vertices are not disjoint, when the graph produced by the apply function is united with the original match apply model the new vertices are created.
-It was necessary to remove the indirect links from the rule during rewriting, because otherwise those indirect links would also be rewritten.
-% 
-% Definitions~\ref{def:match_function} and~\ref{def:apply_function} are complementary: the former gathers all the sub-models of an input-output model that are matched by a transformation rule; the latter glues on the output part of each of those sub-models new objects and relations created by a transformation rule.
-% 
-% %  The $strip$ function is used to
-% %  enable matching over backward links but not elements to be created by the
-% %  transformation rule. The $back$ function connects all newly created vertices
-% %  to the elements of the source model that originated them.
-% 
-% %\begin{definition}{Transformation}
-% 
-% %Let $g\in TG$ be a typed graph. The graph transform function $transform : TR\times TG\rightarrow TG$ is recursively defined as:
-% %\begin{gather*}
-% %  transform_{(m,a)}(g) =
-% %  \begin{cases}
-% %    g  & \text{if } match_{m}(g)=\emptyset \\
-% %    instance(a)\cup transform_{(m,a)}(g') &\text{if } match_{m}(g)\neq \emptyset
-% %  \end{cases}
-% %\end{gather*}
-% %$$\text{where } g'=mark_{(m)}(g)$$
-% %\end{definition}
-% 
-% %\subsection{Transformation Semantics}
-% %\begin{definition} {Layer Semantics}
-% 
-% %Let $l\in Layer$ be a layer and $\{m,m'\} \in MAM^{s}_{t}$ be a match
-% %$$\frac{}
-% %{m,\emptyset \stackrel{layerstep}{\rightarrow}m}$$
-% %$$\frac{tr\in layer,\; transform_{tr}(m) = m''\; m'',layer\backslash\left\{tr\right\} \xrightarrow{layerstep}	 m'}{m,layer \xrightarrow{layerstep} m'}$$
-% %\end{definition}
-% 
-% \begin{definition} {Layer Step Semantics}
-% \label{def:layer_step_semantics}
-% 
-% Let $l\in \textsc{Layer}^{sr}_{tg}$ be a Layer. The \emph{layer step relation}
-% $\stackrel{layerstep}{\rightarrow}\subseteq \textsc{Iom}^{sr}_{tg}\times \textsc{Iom}^{sr}_{tg} \times \textsc{Layer}^{sr}_{tg}\times
-% \textsc{Iom}^{sr}_{tg}$ is defined as follows:
-% 
-% $$\frac{}
-% {\langle m_{in},m_{glue},\epsilon\rangle \xrightarrow{layerstep}m_{in} \sqcup m_{glue}}$$
-% 
-% $$\frac{\begin{array}{ll}& rl\in l,\; apply_{rl}(m_{in}) = m_{rout},
-% \langle m_{in},m_{glue}\sqcup m_{rout},l\backslash \{rl\}\rangle \xrightarrow{layerstep} m_{out}
-% \end{array}}
-% {\langle m_{in},m_{glue},l\rangle \xrightarrow{layerstep} m_{out}}$$
-% 
-% \begin{center}
-% where $m_{rout} \in\textsc{Iom}^{sr}_{tg}$ and $rl\in \textsc{Rule}^{sr}_{tg}$.
-% \end{center}
-%  
-% We impose that all input-output models that are part of $rout$ and have been generated by rule $rl$ are disjoint from input-output models accumulated in $m_{glue}$ that have been generated by other rules.
-% 
-% \end{definition}
-% 
-% In definition~\ref{def:layer_step_semantics} we build the result of executing a layer of a DSLTrans transformation.
-The operational semantics-like rules in the definition execute each rule $rl$ in layer $l$, in a non-deterministic order, by using the $apply$ function.
-The result of executing each rule is accumulated in the temporary $m_{glue}$ input-output model.
-Finally, when the set of transformation rules in the layer has been exhausted, the result of executing all the rules in the layer (now contained in $m_{glue}$) is united with the input-output model $m_{in}$, the input to layer $l$.
-Note that this final union produces the result we expect because of the fact that the $m_{glue}$ input-output model is not disjoint from $m_{in}$.
-The common ``glue'' parts of $m_{glue}$ that have been built by the match function and extended by the apply function are now used to built the result of executing layer $l$.
-% 
-% Definition~\ref{def:layer_step_semantics} is the core of DSLTrans' semantics.
-Many model transformation languages are based on graph rewriting, where the result of each rule rewrite is immediately usable by all other rules.
-In DSLTrans the result of executing one layer in DSLTrans is totally produced before the input to the layer is changed.
-This is enforced in definition~\ref{def:layer_step_semantics} by the fact that the apply function always executes over the same $m_{in}$ input-output model and all the results of rule execution in the same layer are added to the $m_{glue}$ structure that is write-only.
-Rules belonging to the same layer are thus forced to execute independently.
-% 
-% 
-% \begin{definition} {Transformation Step Semantics}
-% \label{def:transformation_step_semantics}
-% 
-% Let $[l::tr]\in \textsc{Transf}^{sr}_{tg}$ be a transformation, where $l\in
-% \textsc{Layer}^{sr}_{tg}$ is a Layer and $tr$ also a transformation. The \emph{transformation step relation}
-% $\stackrel{trstep}{\rightarrow}\subseteq \textsc{Iom}^{sr}_{tg}\times \textsc{Transf}^{sr}_{tg}\times
-% \textsc{Iom}^{sr}_{tg}$ is defined as follows: 
-% $$\frac{}{\langle m,[]\rangle \xrightarrow{trstep} m}$$
-% 
-% $$\frac{\big\langle m_{in},\epsilon,l^{\star}\big\rangle \xrightarrow{layerstep} m_{inter},\; \langle m_{inter},R\rangle \xrightarrow{trstep} m_{out}}{\langle m_{in},[l::T]\rangle \xrightarrow{trstep} m_{out}} \hspace{.3cm} \text{where} \hspace{.3cm} l^{\star}=\bigcup_{rl\in l}rl^{\star}$$
-% \end{definition}
-% 
-% While the execution of the rules belonging to a layer happens in parallel, the execution of the layers of a transformation happens sequentially. 
-% As per definition~\ref{def:transformation_step_semantics}, the input-output model $m_{inter}$ is the output of executing a given layer that is passed onto the next layer as input.
-The transformation rules in a layer are expanded before execution ($l^{\star}$) such that polymorphism in match elements can be handled (see definition~\ref{def:transformation_rule_expansion}).
+Every model transformation execution is confluent up to typed graph isomorphism.
 
-\subsection{Semantic Transformation Constructs}
-\label{sec:DSLTrans_semantics}
+\bentley{As we are relying on the double-pushout approach, I think we get this confluence for free.}
 
-In the definition that follows we introduce the notion of execution of a DSLTrans model transformation.
-For our purposes it is sufficient to introduce it as an input-output model (see \cref{def:input_output_model}), containing the input model for the transformation, the produced output, and the traceability links built during execution.
-Due to space limitations, we cannot introduce the semantics of DSLTrans in the main text of this paper.
-We thus refer the reader to \cref{sec:DSLTrans_formal} for a formal description on how DSLTrans transformation executions are built.
+\end{proposition}
 
-\begin{definition} {Model Transformation Execution\\}
-\label{def:modeltransformation} 
-\CatchFileBetweenTags{\modeltransformation}{text/definitions}{modeltransformation}{\modeltransformation}
-\end{definition}
+\begin{pf}
+We want to prove that for every model transformation execution of a transformation $\mathit{transform} \in \textsc{Transforms}$ having as input an input-output model $\mathit{input} \in \mathit{IOMs}$, its output is always the same up to typed graph isomorphism.\\
 
-\CatchFileBetweenTags{\modeltransformationtext}{text/definitions}{modeltransformationtext}{\modeltransformationtext}
+If we assume an execution of the transformation is not confluent then this should happen because of non-determinism when the execution of a transformation is being built.
+Non-determinism happens during the construction of a transformation execution at two points:
+\begin{enumerate}
+\item The section `Execution of a DSLTrans Rule' discusses the matching and rewriting components to rule application.
+Note that our technique relies on the double-pushout approach, which itself is non-deterministic up to typed graph isomorphism, which does not contradict the proposition we are trying to prove.
+\bentley{Revise this.}
 
-% A transformation execution is formed from taking the input model, and executing
-% the transformation on it to produce the output model. During this
-% transformation, traceability links will be placed between match elements and the
-% apply elements they create. Definition~\ref{def:transf_ex} expresses this
-% formally. Note that we assume that transformation executions are built following
-% the algorithm described in~\cite{DBLP:conf/sle/BarrocaLAFS10}.
+\item In Definition~\ref{def:dsltrans_layer_execution}, the rules composing a transformation layer are composed into a production on the input graph.
+Note that the order in which the transformation rules are composed may be non-deterministic.
+However, as these rules can be shown to be parallel independent \bentley{And they have to in that definition}, the production is by definition confluent up to typed graph isomorphism.
 
-% \begin{definition}{Transformation Execution}
-% \label{def:transf_ex}
-% 
-% Let $t$ be a DSLTrans transformation having source metamodel $s$ and target
-% metamodel $t$. A Transformation Execution is a 6-tuple $\langle V,E\cup
-% Tl,\tau,Match,Apply,Tl\rangle$, where $\langle V,E,\tau,Match,Apply\rangle \in
-% MAP^{s}_{t}$ is a match-apply pattern. $Match=\langle V',E', \tau',s\rangle$,
-% $Apply=\langle V'',E'', \tau'',t\rangle$ and the edges $Tl\subseteq V'\times
-% V''$ are called \emph{traceability links}. The set of all transformation
-% executions having source metamodel $s$ and target metamodel $t$ is called
-% $Exec^{s}_{t}$.
-% \end{definition}
+\end{enumerate}
+Given there are no other sources of non-determinism when building the execution of a transformation, every model transformation execution is confluent up to typed graph isomorphism.
+\end{pf}
 
-% We now prove two important properties about DSLTrans' transformations.
-% 
-% \begin{proposition}{Confluence}
-% 
-% Every model transformation is confluent up to typed graph isomorphism.
-% \end{proposition}
-% 
-% \begin{proposition}{Termination} 
-% 
-% Every model transformation terminates.
-% \end{proposition}
+\begin{proposition}{Termination}
 
-% The language to describe properties is in fact very similar to the language to
-% express transformations, with the additional possibility of expressing indirect
-% links in the $apply$ pattern --- thus allowing more abstract patterns than the
-% ones expressed in transformations. This is natural given that the properties of
-% a transformation can be more abstract than the rules implementing them.
+Every model transformation execution terminates.
+\end{proposition}
+\begin{pf}
+Let us assume that there is a transformation execution which does not terminate.
+In order for this to happen there must exist a part in the construction of the execution of a transformation which induces an algorithm with an infinite amount of steps.
+We identify three moments when this can happen:
+\begin{enumerate}
+
+\item If the matching and rewriting process of a rule is infinite.
+However, as the double-pushout approach uses a single production, then this must be a finite process.
+\bentley{Verify this, especially in the case of multiple applications of the rule.}
+
+% if the result of the $match_{rl}(m_{in})$ function in definition~\ref{def:match_function} is an infinite set of match-apply graphs.
+The input-output model $m_{in}$ is by definition finite and the matching of each rule is independent from the execution of other rules in the same layer.
+As such, the number of subgraphs of $m_{in}$ isomorphic to $rl$'s matcher found during the execution of $rl$ is finite.
+
+\item if Definition~\ref{def:dsltrans_layer_execution} of execution of a layer induces an infinite amount of steps.
+The only possibility for this to happen is if a layer has an infinite amount of transformation rules, which is a contradiction with Definition~\ref{def:layer}.
+
+\item if Definition~\ref{def:DSLTrans_transformation_execution} of execution of a transformation induces an infinite amount of steps.
+Given layers are executed sequentially and no looping is allowed, the only possibility for this to happen is if the transformation has an infinite amount of layers, which contradicts Definition~\ref{def:transformation}.
+
+\end{enumerate}
+Given there are no other constructs in the semantics of a transformation that can induce an infinite amount of steps, every model transformation execution terminates.
+\end{pf}

### 2016-06-29 text/old/appendix_DSLTrans.tex
Move old sections. Divorce paper from definitions file.

--- 2015-03-13 text/appendix_DSLTrans.tex
@@
 
+\reviewer{Def. B.12. First of all, I wonder why you define match as a function rather than
+just defining what a match is; namely, an element of the set you are constructing
+here. Secondly (and more importantly), why do you not use the notion of homo-
+morphism instead of explicitly talking about ismorphic subgraphs? Your way,
+you lose the connection between glue and $||rl||$: you know they are isomorphic,
+but what if $||rl||$ has symmetries - then there is more than one match, which in
+your definition cannot be distinguished.}
+
+\reviewer{You pay the price for this loss of information in Def. B.13, which now is broken: $a\Delta$ depends on $m_{glue} $(which was called glue in Def. 12, why the change in
+notation?) in a way depending on the mapping from $m_{glue}$ to $rl$ which you have
+discarded.}
+
+\reviewer{Def. B.13 is broken in another way as well: it does not properly define E' but
+merely imposes some requirements that are also fulfilled by choosing E' = E.
+Instead I know from the discussion about traceability links in Section 4 (which
+logically comes after this appendix!) that you really want to add such links for
+all newly created nodes of the output graph to all pre-existing nodes of the input
+graph; but that is not made clear anywhere at this point.}
+
 The match function in \cref{def:match_function} looks for subgraphs ($glue\sqsubseteq m_{in}^{*}$) of an input-output model that are isomorphic to the backward matcher of the given transformation rule ($glue \cong \Vert rl\Vert$).
@@
 Let us now turn our attention to the apply function in \cref{def:apply_function}.
-Its role is extend all model fragments found by matching a rule on a given input-output model, such that each of those fragments becomes isomorphic to the complete rule (minus its backward links).
+Its role is to extend all model fragments found by matching a rule on a given input-output model, such that each of those fragments becomes isomorphic to the complete rule (minus its backward links).
 This process effectively creates the new objects and relations specified in the apply part of the rule, for each of the fragments found when matching the rule.
@@
 \label{def:apply_function}
+
+\reviewer{In your definitions you rely on the $\sqcup$ operator (sqare union),
+  the definition of which is not entirely clear.  For instance, I do not know where the operator used in Def. B13
+  is defined.}
 
@@
 
+\reviewer{Def. B.14: It seems to me rather roundabout to sequentially apply all rules, ac-
+cumulate the results and then take their union. Why do you not take the union
+simultaneously with the parallel application of all rules (of a given layer), as
+you are essentially doing for all applications of a single rule in Def. B.13? This
+would get rid of the artificial sequential ordering, and the whole issue of con-
+fluence would not even have to be raised as there is no non-determinism any
+more.}
+
+\reviewer{In the constructions defined in Appendix B, you ignore the required absence of
+containment cycles. Might such cycles not appear in the union of the graphs,
+even if the individual graphs are cycle-free?}
+
 In \cref{def:layer_step_semantics} we build the result of executing a layer of a DSLTrans transformation.
@@
 
+\reviewer{I believe tr, T and R in this definition stand for the same thing.}
 While the execution of the rules belonging to a layer happens in parallel, the execution of the layers of a transformation happens sequentially.

### 2016-06-29 text/old/appendix_abstraction_relation.tex
Move old sections. Divorce paper from definitions file.

--- 2016-06-29 text/abstraction_proofs.tex
@@
-\section{Validity and Completeness}
-\label{subsec:abstraction_relation_validity_completeness}
+\section{Validity and Completeness of Path Condition Generation}
+\label{sec:val_complete_path_cond_gen}
+\counterwithin{proposition}{section}
+\counterwithin{lemma}{section}
+\counterwithin{equation}{section}
+\setcounter{proposition}{0}
+\setcounter{lemma}{0}
+\setcounter{equation}{0}
 
-In this section, we discuss the validity and completeness of the abstraction relation.
-Only proof sketches are presented here for \cref{prop:pc_validity} and \cref{lem:uniqueness}, while more complete proofs are found in \cref{sec:val_complete_path_cond_gen}, in the proofs of \cref{prop:pc_validity_appendix} and \cref{lem:uniqueness_appendix} respectively.
+\onehalfspacing
 
-%\newpage
+\begin{definition}{Path Condition\\}
+\label{def:path_condition_appendix}
+\CatchFileBetweenTags{\pcdef}{text/definitions}{pcdef}{\pcdef}
+\end{definition}
+
+\CatchFileBetweenTags{\pcappendix}{text/definitions}{pcappendix}{\pcappendix}
+
+\begin{definition}{Combination of a Path Condition with a Rule}
+\label{def:combine_pc_with_rule_appendix}
+\CatchFileBetweenTags{\pccfrc}{text/definitions}{pccfrc}{\pccfrc}
+\end{definition}
+
+\CatchFileBetweenTags{\pccfrctextappendix}{text/definitions}{pccfrctextappendix}{\pccfrctextappendix}
+
+\begin{definition}{Path Condition and Rule Combination -- No Dependencies\\}
+\label{def:rule_comb_no_dependencies_appendix}
+\CatchFileBetweenTags{\nodeps}{text/definitions}{nodeps}{\nodeps}
+\end{definition}
+
+\CatchFileBetweenTags{\nodepstextappendix}{text/definitions}{nodepstextappendix}{\nodepstextappendix}
+
+\begin{definition}{Path Condition and Rule Combination -- Unsatisfied Dependencies\\} 
+\label{def:rule_comb_unsatisfied_appendix}
+\CatchFileBetweenTags{\unsatdeps}{text/definitions}{unsatdeps}{\unsatdeps}
+\end{definition}
+
+According to the pre-conditions of the equation presented in \cref{def:rule_comb_unsatisfied_appendix}, a path condition does not satisfy the dependencies present in a rule if there is no surjective typed graph homomorphism between the backward links of the rule and the symbolic traceability links of the path condition.
+Besides expressing the fact that all backward links must exist as symbolic traceability links the path condition, the surjective homomorphism allows modeling the case where dependencies expressed by two (or more) backward links between similarly typed elements can be satisfied one single symbolic traceability link in the path condition.
+
+\begin{definition}{Single Partial and Total Combination of a Set of Path Conditions with a Rule\\}
+\label{def:comb_path_cond_rule_single_appendix}
+\CatchFileBetweenTags{\satdeps}{text/definitions}{satdeps}{\satdeps}
+
+\begin{align}
+\label{eq:pcomb_appendix}
+\CatchFileBetweenTags{\satdepseqone}{text/definitions}{satdepseqone}{\satdepseqone}
+\end{align}
+
+\begin{align}
+\label{eq:tcomb_appendix}
+\CatchFileBetweenTags{\satdepseqtwo}{text/definitions}{satdepseqtwo}{\satdepseqtwo}
+\end{align}
+
+\end{definition}
+
+Let us start by introducing relation $\stackrel{p\_comb}{\rightarrow}$, presented in \cref{eq:pcomb_appendix} of \cref{def:comb_path_cond_rule_single_appendix}.
+The relation takes as arguments a set of path conditions being accumulated for the current layer, the rule to be combined, and an $rl_{glue}$ argument indicating the place in each of the input path conditions the rule should be anchored to during the combination step.
+The relation's output is a new set of path conditions.
+This new set includes all the original path conditions, as well as each path condition in the accumulator set ``glued'' to a copy of rule being examined.
+Note that the relation $\stackrel{p\_total}{\rightarrow}$ in \cref{eq:tcomb_appendix} is similarly defined, except for the fact path conditions in the accumulator set are not preserved in the relation's output set.\\
+
+\setcounter{equation}{0}
+
+Let us now define how a rule is combined with a path condition, whenever its backward links can be found several times in that path condition.
+We formalize it in \cref{def:comb_path_cond_rule_mul_appendix}, by means of relations $\stackrel{p\_step}{\rightarrow}$ and $\stackrel{t\_step}{\rightarrow}$.
+These two relations operationally describe the sequence of steps necessary to ``glue'' a rule at multiples places of a path condition.
+The set of places targeted in the path condition for receiving a copy of the rule is given by the sets $partialSet$ and $totalSet$ (found respectively in \cref{eq:pstep} and \cref{eq:tstep}).
+As expected, these sets contain the set of traceability links in the path condition where copies of the rule need to be anchored to.
+
+\begin{definition}{Multiple Partial and Total Combination of a Set of Path Conditions with a Rule\\}
+\label{def:comb_path_cond_rule_mul_appendix}
+\CatchFileBetweenTags{\satdepstwo}{text/definitions}{satdepstwo}{\satdepstwo}
+\end{definition}
+
+Having \cref{def:comb_path_cond_rule_single_appendix} and \cref{def:comb_path_cond_rule_mul_appendix} in mind, we can now proceed to define the complete combination relation of a rule with a path condition in the case of partially and totally satisfied dependencies.
+
+\begin{definition}{Path Condition and Rule Combination -- Partially and Totally Satisfied Dependencies\\}
+\label{def:rul_comb_partial_total_appendix}
+\CatchFileBetweenTags{\satdepsthree}{text/definitions}{satdepsthree}{\satdepsthree}
+\end{definition}
+
+The top equation in \cref{def:rul_comb_partial_total_appendix} defines the $\stackrel{combine}{\rightarrow}$ relation for when rule $rl$ has dependencies that are satisfied by path condition $pc$.
+The pre-conditions in the equation state that the backward links in the rule are found in the path condition, as expected.
+Additionally, two sequential steps perform the gluing of the rule $rl$ on all path conditions in accumulator $AC$, wherever the rule is partially and/or totally found in each of those path conditions.
+Relations $\stackrel{p\_comb}{\rightarrow}$ and $\stackrel{t\_comb}{\rightarrow}$ presented in \cref{def:comb_path_cond_rule_mul_appendix} are used to model these two operational ``gluing'' steps.
+Functions $partialsat$ and $totalsat$, described in the latter part of \cref{def:rul_comb_partial_total_appendix}, are used to gather the places of path condition $pc$ where copies of the rule need to be anchored to.
+
+\begin{definition} {Combining a Path Condition with a Layer\\}
+\label{def:path_cond_layer_comb_appendix}
+\CatchFileBetweenTags{\combpclayer}{text/definitions}{combpclayer}{\combpclayer}
+\end{definition}
+
+\CatchFileBetweenTags{\combpclayertextappendix}{text/definitions}{combpclayertextappendix}{\combpclayertextappendix}
+
+\begin{definition} {Combining a Set of Path Conditions with a Layer\\}
+\label{def:path_cond_set_layer_comb_appendix}
+\CatchFileBetweenTags{\combpcsetlayer}{text/definitions}{combpcsetlayer}{\combpcsetlayer}
+\end{definition}
+
+\CatchFileBetweenTags{\combpcsetlayertextappendix}{text/definitions}{combpcsetlayertextappendix}{\combpcsetlayertextappendix}
+
+\begin{definition} {Path Condition Generation\\}
+\label{def:path_cond_gen_appendix}
+\CatchFileBetweenTags{\pathcondgen}{text/definitions}{pathcondgen}{\pathcondgen}
+\end{definition}
+
+\CatchFileBetweenTags{\pathcondgentextappendix}{text/definitions}{pathcondgentextappendix}{\pathcondgentextappendix}
+
+\begin{definition}{Abstraction of a Transformation Execution by a Path Condition\\}
+\label{def:abstraction_pc_ex_appendix}
+\CatchFileBetweenTags{\abstractionrelappendix}{text/definitions}{abstractionrelappendix}{\abstractionrelappendix}
+\end{definition}
+
+To understand the abstraction relation in \cref{def:abstraction_pc_ex_appendix}, recall
+that during the construction of a transformation execution rules are matched
+injectively in the input model. This information is present in the first
+condition of the abstraction relation (Proposition~\ref{eq:abstr_input_output_appendix})
+via the injective typed graph homomorphism between the match part of the copies
+of rules ``glued'' onto the path condition and the containment transitive
+closure of the input part of the transformation execution. This relation
+enforces the fact that certain parts of the execution were found, or matched, by
+certain parts of the path condition.
+On the other hand, the surjection from the output of the execution towards the
+apply part of the path condition models the fact that the output of the
+execution has been completely built by instantiating the apply parts of the
+rules contained in the path condition.
+
+The second condition of the abstraction relation
+(Proposition~\ref{eq:abstr_trace_appendix}) checks for the fact that symbolic
+traceability links in the path condition and traceability links in the execution
+correctly correspond to each other. This is modeled by the fact that all
+strongly connected components in the path condition, composed only of symbolic
+traceability links, are injectively found on the execution. This injection
+models the fact that traceability graphs between individual or combined rules in
+the path condition are necessarily found in the execution. Note that components
+of the path condition are considered because of the fact that disconnected rules
+in the path condition may have matched over common elements of a particular
+execution. As such, a full injection between the complete traceability graph in
+the path condition and the execution would be incorrect. Additionally, in the
+second part of Proposition~\ref{eq:abstr_trace_appendix} we check the fact that every
+traceability link in the execution can be found in the path condition. This
+additional sanity check enforces that no spurious traceability links that could
+not have been created by the rules present in the path condition exist in the
+transformation execution.
+
+Finally, the last clause of the abstraction relation states that rule copies
+that are repeated a number of times in the path condition need to be found at
+least a similar amount of times in the abstracted transformation execution.
+
+It is important to mention that another abstraction relation, weaker or
+stronger, could have been chosen. The abstraction relation presented in
+\cref{def:abstraction_pc_ex_appendix} suits our needs in the sense that it allows us to
+demonstrate the validity and completeness of our proof technique, as we will
+show in the text follows. Additionally, it is particularly interesting because
+it makes sure that, given a DSLTrans transformation, each of its transformation
+executions is abstracted by one and only one of its path conditions. This result
+adds to the consistency of our theory and is also exposed later in this section.
+
 \begin{proposition}{(Validity)}
-\label{prop:pc_validity}
+\label{prop:pc_validity_appendix}
 Every path condition abstracts at least one transformation execution.
 \end{proposition}
-\begin{ps}
+\begin{pf}
+Let $tr\in \textsc{Transf}^{sr}_{tg}$ be a DSLTrans transformation.
+We wish to demonstrate that, for all path conditions $pc\in \textsc{Pathcond}(tr)$, there exists a transformation execution $ex\in \textsc{Exec}(tr)$ of the set of rules used to build $pc$ such that $pc$ abstracts $ex$ (i.e.
+$ex\Vvdash pc$), as formally expressed in \cref{def:abstraction_pc_ex_appendix}.
+We can prove this property by induction on the set of transformations $\textsc{Transf}^{sr}_{tg}$ (see \cref{def:layer_transformation_appendix}), as follows:
 
-Let $tr\in \textsc{Transf}^{sr}_{tg}$ be a DSLTans transformation.
-We wish to demonstrate that, for all path conditions $pc\in \textsc{Pathcond}(tr)$, there exists a transformation execution $ex\in \textsc{Exec}(tr)$ of the set of
-rules used to build $pc$ such that $pc$ abstracts $ex$, as formally expressed in \cref{def:abstraction_pc_ex}.
-We can prove this property by induction on the set of transformations $\textsc{Transf}^{sr}_{tg}$ (see \cref{def:layer_transformation}), as follows:
+% We will start by showing that the property
+% holds for the path conditions that are generated for one single layer of a
+% DSLTrans transformation, which in what follows we will call lemma 1.
+% In this case the inductive proof will require the following steps:
 
 \begin{itemize}
-  \item \emph{Base case:} the base case is when $tr$ is the empty
-  transformation. In this case, according to \cref{def:path_cond_gen} only the
-  empty path condition $\epsilon_{pc}$ exists in the path condition set. We thus
-  need to demonstrate that the empty path condition abstracts the empty
-  transformation execution $\epsilon_{ex}$, as well as any execution for which
-  the input model is never matched by any rule (consequently having an empty
-  output model).
-For any of these transformation executions, Proposition~\ref{eq:abstr_input_output} of the abstraction relation definition is satisfied, as no rule copy exists in the path condition and the output of the transformation execution is empty -- empty typed graph homomorphisms thus satisfy all the conditions of the proposition.
-Proposition~\ref{eq:abstr_trace} of the abstraction relation definition also trivially holds because no traceability links exist either in the path condition or in any of the considered executions.
-
-%   This is so because:
-%   1) the empty typed graph homomorphism is an injective typed graph homomorphism
-%   between the match part of the empty path condition and any input model; and 2) as
-%   well as a surjective typed graph homomorphism between an empty output model
-%   and the apply part of the empty path condition.
-
   \item \emph{Base case:} the base case is the case when we have $tr=[\;]$, i.e.
@@
 In order to demonstrate the inductive case we need to show the property holds for all path conditions resulting from combining the rules of layer $l$ with any path condition generated for $tr$.
-These path conditions for transformation $tr'$ are built as expressed in \cref{def:path_cond_layer_comb}.
+These path conditions for transformation $tr'$ are built as expressed in \cref{def:path_cond_layer_comb_appendix}.
 According to this definition, path conditions for $tr'$ are built by incrementally combining the path conditions generated for $tr$ with a rule of layer $l$, until all the rules in $l$ have been treated.
-We can thus again use induction for this proof, this time on the set of possible layers $\textsc{Layer}^{sr}_{tg}$.
+We can thus again use induction for this proof, this time on the set of possible layers $\textsc{Layer}^{sr}_{tg}$ built as expressed in \cref{def:layer_transformation_appendix}.
 
@@
   \item \emph{Base case:} this is the case where layer $l$ contains no rules.
-In this case, by the base case of \cref{def:path_cond_layer_comb}, no new path condition is added to the set of path conditions generated for the transformation $tr$.
+In this case, by the base case of definition~\ref{def:path_cond_layer_comb_appendix}, no new path condition is added to the set of path conditions generated for the transformation $tr$.
 As such the $tr=tr'$ and by induction hypothesis the property trivially holds for all path conditions generated for $tr'$.
 
-  \item \emph{Inductive case:} for the inductive case (transitive case of \cref{def:path_cond_layer_comb}) we need to show that, assuming the property holds for all path conditions generated for a transformation $tr$, then the property will also hold for a transformation $tr'$ -- where $tr'$ results from adding a new rule $rl$ to the last layer of $tr$.
-We will thus need to consider the four \reviewer{three?} cases of rule combination:\vspace{.2cm}
+  \item \emph{Inductive case:} for the inductive case (transitive case of \cref{def:path_cond_layer_comb_appendix}) we need to show that, assuming the property holds for all path conditions generated for a transformation $tr$, then the property will also hold for a transformation $tr'$ -- where $tr'$ results from adding a new rule $rl$ to the last layer of $tr$.
+We will thus need to consider the four cases of rule combination:\vspace{.2cm}
 
 \begin{enumerate}
-\item\label{lab:rule_case1} Rule $rl$ has no dependencies (\cref{def:rule_comb_no_dependencies}).
-\item\label{lab:rule_case2} Rule $rl$ has dependencies and cannot execute (\cref{def:rule_comb_unsatisfied}).
-\item\label{lab:rule_case3} Rule $rl$ has dependencies and may and/or will execute (\cref{def:rul_comb_partial_total}).
+\item\label{lab:rule_case1_appendix} Rule $rl$ has no dependencies (\cref{def:rule_comb_no_dependencies_appendix}).
+\item\label{lab:rule_case2_appendix} Rule $rl$ has dependencies and cannot execute (\cref{def:rule_comb_unsatisfied_appendix}).
+\item\label{lab:rule_case3_appendix} Rule $rl$ has dependencies and may and/or will execute (\cref{def:rul_comb_partial_total_appendix}).
 \end{enumerate}
-
-The property trivially holds for case~\ref{lab:rule_case2}, given that no new path conditions are added to the path condition set generated for $tr$ and that the property holds for $tr$ by induction hypothesis.
-When a rule $rl$ is added to the last layer of $tr$ such that cases~\ref{lab:rule_case1} or~\ref{lab:rule_case3} occur, then the property can be shown to hold for $tr'$ as follows: 1) choose for a general path condition $pc$ generated for $tr$ an execution $ex$ such that $pc$ abstracts $ex$; 2) build an input model $m$ as the result of uniting the input model of $ex$ with a model that can be matched by $rl$; 3) execute $tr'$ having as input model $m$ to produce transformation execution $ex'$; and finally 4) demonstrate $ex'$ is abstracted by the path condition $pc'$ resulting from combining $pc$ with $rl$ whether rule $rl$ does not depend on $pc$ or rule $rl$ depends on $pc$ and may and/or will execute.
 
@@
 \end{itemize}
-\end{ps}
+\end{pf}
 
 \begin{proposition}{(Completeness)}
-\label{prop:pc_completeness}
+\label{prop:pc_completeness_appendix}
 Every transformation execution is abstracted by one path condition.
 \end{proposition}
-\begin{pf}
-Let $tr\in \textsc{Transf}^{sr}_{tg}$ be a DSLTans transformation.
-We wish to demonstrate that, for all transformation executions $ex\in \textsc{Exec}(tr)$, a path condition $pc\in \textsc{Pathcond}(tr)$ exists such that $ex$ is abstracted by $pc$, as formally expressed in \cref{def:abstraction_pc_ex}.\\
-
-Completeness can be shown as a corollary of \cref{prop:pc_validity} about the \emph{validity} of path condition generation.
-The complete set of executions $\textsc{Exec}(tr)$ (see \cref{def:modeltransformation}) can be split into two kinds of executions:
-\begin{enumerate}
-  \item\label{lab:input_not_matched} The \emph{empty execution} $\epsilon_{ex}$ or the \emph{execution where the input model was not matched by any rule}.
-As mentioned in \cref{prop:pc_validity}, these executions are abstracted by the empty path condition $\epsilon_{pc}$.\vspace{.3cm}
-  \item The \emph{execution $ex$ where a number of rules of $tr$ have been applied to the input model}, where each transformation rule $rl$ of $tr$ may have been applied more than once.
-In this case we have that, because all possible and valid rule combinations are considered when building path conditions, a path combination $pc$ exists that contains one or more copies of each of the rules used when operationally building $ex$.
-\\
-
-  Moreover, during the proof of \emph{validity} of path condition generation in \cref{prop:pc_validity} we demonstrate that, when we add a new rule $rl$ to the last layer of a transformation $tr$ (such that we have a new transformation $tr'$), the rule combination step explained in Definitions~\ref{def:rule_comb_no_dependencies},~\ref{def:rule_comb_unsatisfied} and~\ref{def:rul_comb_partial_total} produces a new set of path conditions where each path condition in that set still abstracts at least one transformation execution of $tr'$.
-This part of the proof (the second induction) is achieved by building for transformation $tr'$ an input model $m$ that can be matched by $rl$ (as well as by all the other rules of $tr$), and then building from $m$ a new transformation execution that is abstracted by a path condition built for $tr'$.
-Because in the proof of \cref{prop:pc_validity}, $m$ is such that it can be matched by $rl$ an arbitrary amount of times, we know that, independently of the number of times a rule is applied during the construction of a transformation execution, a path condition abstracting that transformation execution exists.\\
-
-Additionally, input elements that are not matched by any rule do not affect the abstraction relation, as explained in case~\ref{lab:input_not_matched} above.
-This means we also know that executions involving input models that are only partially matched by the rules of $tr$ are also abstracted by one path condition.
-\end{enumerate}
-
-Let $tr\in \textsc{Transf}^{sr}_{tg}$ be a DSLTans transformation.
-We wish to demonstrate that, for all transformation executions $ex\in \textsc{Exec}(tr)$, a path condition $pc\in \textsc{Pathcond}(tr)$ exists such that $ex$ is abstracted by $pc$, as formally expressed in \cref{def:abstraction_pc_ex_appendix}.
-Completeness can be shown as a corollary of \cref{prop:pc_validity_appendix} about the \emph{validity} of path condition generation.
-The complete set of executions $\textsc{Exec}(tr)$ (see \cref{def:modeltransformation_appendix}) can be split into two kinds of executions:
-\begin{enumerate}
-  \item\label{lab:input_not_matched} The \emph{empty execution} $\epsilon_{ex}$ or the \emph{execution where the input model was not matched by any rule}.
-As mentioned in \cref{prop:pc_validity_appendix}, these executions are abstracted by the empty path condition $\epsilon_{pc}$.\vspace{.3cm}
-  \item The \emph{execution $ex$ where a number of rules of $tr$ have been applied to the input model}, where each transformation rule $rl$ of $tr$ may have been applied more than once.
-In this case we have that, because all possible and valid rule combinations are considered when building path conditions, a path combination $pc$ exists that contains one or more copies of each of the rules used when operationally building $ex$.
-Moreover, during the proof of \emph{validity} of path condition generation in \cref{prop:pc_validity_appendix} we demonstrate that, when we add a new rule $rl$ to the last layer of a transformation $tr$ (such that we have a new transformation $tr'$), the rule combination step explained in Definitions~\ref{def:rule_comb_no_dependencies_appendix},~\ref{def:rule_comb_unsatisfied_appendix} and~\ref{def:rul_comb_partial_total_appendix} produces a new set of path conditions where each path condition in that set still abstracts at least one transformation execution of $tr'$.
-This part of the proof (the second induction) is achieved by building for transformation $tr'$ an input model $m$ that can be matched by $rl$ (as well as by all the other rules of $tr$), and then building from $m$ a new transformation execution that is abstracted by a path condition built for $tr'$.
-Because in the proof of \cref{prop:pc_validity_appendix}, $m$ is such that it can be matched by $rl$ an arbitrary amount of times, we know that, independently of the number of times a rule is applied during the construction of a transformation execution, a path condition abstracting that transformation execution exists.\\
-
-Additionally, input elements that are not matched by any rule do not affect the abstraction relation, as explained in case~\ref{lab:input_not_matched} above.
-This means we also know that executions involving input models that are only partially matched by the rules of $tr$ are also abstracted by one path condition.
-\end{enumerate} 
-\end{pf}
-
-% Our path conditions construction algorithm takes all possible interactions of rules
-% into account:
-% \begin{enumerate}
-% \item When building the path conditions for one layer we have built all
-% the combinations of rules, as well as their possible interactions through
-% \emph{disambiguation};
-% \item When combining the path conditions from different layer different layers
-% we have considered the cases where no interactions exist, or where backward
-% links in rules requires certain certain elements to already exist in the
-% generated model thus far.
-% \end{enumerate}
-% Note that, given the considered subset of DSLTrans constructs as described
-% in \cref{sec:dsltrans}, no additional cases of rule interaction other than the
-% ones described above need to be considered.
-% From \cref{prop:pc_validity} we also know that at least one concrete
-% transformation execution exists per path condition. In the proof of
-% \cref{prop:pc_validity} we have built, for each path condition $A$,
-% the simplest transformation execution that is abstracted by $A$. These
-% transformation executions produce one instance of each element and association
-% present in the apply pattern of $A$. However, given
-% \cref{def:instance_pc_ex} in \cref{sec:proofs} (abstraction of
-% a transformation execution by a path condition), an infinite amount of
-% transformation executions are always abstracted by $A$. These executions
-% correspond to the generation of output models holding an arbitrarily large
-% amount of the elements and associations while still being abstracted by $A$.
-% Because we have considered all possibilities of execution in our path condition
-% construction algorithm, the union of all transformation executions built for
-% each path condition constitutes the complete, infinite set of transformation
-% executions. We thus know that every possible transformation execution is
-% abstracted by at least one path condition.
-
-% Note that we do not provide any guarantee that the same transformation execution
-% cannot be abstracted by two different path conditions. Although given our symbolic
-% execution assumptions we believe this to be the case, further mathematical
-% exploration of this issue is necessary. Given however the fact that our
-% verification algorithm exhaustively explores all path conditions, it is
-% sufficient for the property proof to provide a \emph{no} result in the analysis
-% of a path condition for the result of the verification of the property to be
-% \emph{no}. \levi{keep this?}
+\CatchFileBetweenTags{\pathgencompletenessproofappendix}{text/definitions}{pathgencompletenessproofappendix}{\pathgencompletenessproofappendix}
 
 \begin{lemma} (Uniqueness)
-\label{lem:uniqueness}
+\label{lem:uniqueness_appendix}
 A transformation execution is abstracted by exactly one path condition.
 \end{lemma}
-\begin{ps}
+\begin{pf}
 Let $tr\in \textsc{Transf}^{sr}_{tr}$ be a model transformation.
-We will demonstrate that two different path conditions\\$pc_1, pc_2\in \textsc{Pathcond}(tr)$ cannot exist such that we have a transformation execution $ex\in \textsc{Exec}(tr)$ where $ex\Vvdash pc_1$ and $ex\Vvdash pc_2$.
+We will demonstrate that two different path conditions $pc_1, pc_2\in \textsc{Pathcond}(tr)$ cannot exist such that we have a transformation execution $ex\in \textsc{Exec}(tr)$ where $ex\Vvdash pc_1$ and $ex\Vvdash pc_2$.
 
-We will do so by attempting to build an $ex\in \textsc{Exec}(tr)$ such that $ex\Vvdash pc_1$ and $ex\Vvdash pc_2$ and demonstrating that it is always the case that such is not possible.
+We will do so by attempting to to build an $ex\in \textsc{Exec}(tr)$ such that $ex\Vvdash pc_1$ and $ex\Vvdash pc_2$ and demonstrating that it is always the case that such is not possible.
 In order to structure our argumentation, we will consider two cases:
@@
 \begin{enumerate}
-  \item\label{item:uniqueness_rule_no_dep} the case where no rules in $tr$ have dependencies.
-  \item\label{item:uniqueness_rule_has_dep} the case where some rules in $tr$ have dependencies.
+  \item\label{item:uniqueness_rule_no_dep_appendix} the case where no rules in $tr$ have dependencies.
+  \item\label{item:uniqueness_rule_has_dep_appendix} the case where some rules in $tr$ have dependencies.
 \end{enumerate}
 
-We start by considering that $tr$ falls into case~\ref{item:uniqueness_rule_no_dep} above.
-By~\cref{def:path_cond_gen} of path condition generation, each rule appears at most once in a path condition.
+We start by considering that $tr$ falls into case~\ref{item:uniqueness_rule_no_dep_appendix} above.
+By \cref{def:path_cond_gen_appendix} of path condition generation, each rule appears at most once in a path condition.
 Also, by construction, each path condition always contains a different combination of rules.
-We additionally know from \cref{def:layer_transformation} that the rules that compose $tr$ necessarily have non-overlapping matchers.
+We additionally know from \cref{def:layer_transformation_appendix} that the rules that compose $tr$ necessarily have non-overlapping matchers.
 We can nonetheless build a model $m$ as the typed graph union of two input models $m_1$ and $m_2$, where injective typed graph morphisms can be found between the match parts of the rule copies that form $pc_1$, and $m_1$.
-Injective typed graph morphisms can be found as well between the match parts of the rules that form $pc_2$, and $m_2$.
+Injective typed graph morphisms can be found as well between the match parts of the rules that form $pc_1$, and $m_2$.
 We thus know that injective typed graph morphisms can be found between the rule copies that compose $pc_1$ and $pc_2$, and $m$.
-This satisfies the first condition of \cref{eq:abstr_input_output} in ~\cref{def:abstraction_pc_ex} of abstraction relation.
+This satisfies the first condition of \cref{eq:abstr_input_output} in ~\cref{def:abstraction_pc_ex_appendix} of abstraction relation.
 
@@
 
-\end{ps}
+\end{pf}
+
+\clearpage

### 2016-06-29 text/old/architecture.tex
Move old sections. Divorce paper from definitions file.

\section{Property Prover Architecture\levi{remove this section?}}
\label{sec:architecture}

In this section we detail the operational artifacts used to implement the
algorithms described the previous sections. These artifacts are fully derived
from the source and target metamodels the DSLTrans transformation under
verification, and can be seen as a `compilation' step that need to be built for
each DSLTrans transformation in order to verify it. These artifacts can be
automatically built using higher-order transformations (HOTs), although for the
time being we have executed these HOTs as a manual step. We provide this section
in order to prove that the implementation of our verification technique follows
MDE principles: all artifacts are explicitly modelled by the appropriate
metamodels, and computations are performed using model transformations.

Figure~\ref{fig:verif_tool_arch} shows the higher order transformations needed
by our framework, along with other required artifacts such as metamodels and
models. Describing text will follow with a brief description of each HOT.

\begin{figure}[h!] \centering \includegraphics[scale=.45]{./figures/dsltrans_verification_architecture.pdf}
	\caption{Verification Tool Architecture}
	\label{fig:verif_tool_arch}
\end{figure}

\begin{enumerate}

  \item\label{label:abstract_rule_metamodel_hot} \textbf{Generate the
  \emph{abstract rule metamodel}}: this HOT takes as input the source and target
  metamodels of the transformation being analysed, and returns a metamodel in
  which an abstract form of the transformation rules can be written. Such a
  metamodel for the police station transformation can be observed in
  figure~\ref{fig:abstract_rule_metamodel_station_trafo}. The purpose of these
  abstract rules is to be the basic building blocks during symbolic execution
  construction.\\

  \item \textbf{Generate the \emph{property metamodel}}: this HOT also takes the
  source and target metamodels of the transformation. What is produced is the
  metamodel which will be used to express properties about the transformation.
  For example, the properties in  figures~\ref{fig:dsltrans_prop1}
  and~\ref{fig:dsltrans_prop2} are written using the \emph{property metamodel}
  generated for the Police Station transformation.\\

  \item \textbf{Generate the \emph{symbolic execution construction rules}}: this
  HOT builds a set that models that represent the transformation's rules, using
  the \emph{abstract rule metamodel}. These rules are then manipulated during
  path condition construction. As an example, the rules in the path
  conditions in the state space construction section
  (figures~\ref{fig:non_mergeable_symb_states},~\ref{fig:mergeable_symb_states}
  and~\ref{fig:mergeable_symb_states_repetition}) are \emph{symbolic execution
  construction rules}.\\

  \item \textbf{Generate the \emph{backward link match transformation}}: builds
  the query transformation responsible for checking whether a graph including
  backward links exists in a path condition. This is used in state space
  creation, where the existence of backward links between rule elements must be
  queried. The input metamodel of the backward link match transformation is the
  \emph{abstract rule metamodel}.\\

  \item \textbf{Generate the \emph{rule combination to path condition
  transformation}}: this HOT takes as input the \emph{abstract rule metamodel}
  generated in step~\ref{label:abstract_rule_metamodel_hot} and generates the
  transformation which transforms rules combinations into path conditions (which
  are both instances of the \emph{abstract rule metamodel}). This transformation
  has the \emph{abstract rule metamodel} as both source and target metamodel.\\

  \item \textbf{Generate the \emph{property verification transformation}}: this
  HOT generates three query transformations for each transformation property to
  be verified:
  \begin{itemize}
    \item a query transformation that checks whether the individual model
    elements present in the property are present in a path condition
    \item a second query transformation that checks whether the match part of the
    property is a subgraph of the match part of a path condition
    \item a third query transformation that checks whether the whole property is a subgraph of
    a path condition.
  \end{itemize}

  All query transformations have as source metamodel the \emph{abstract rule
  metamodel}. These transformations are then used in
  section~\ref{sec:verif_dsltrans_props}.
\end{enumerate}

\begin{figure}[h!] \centering \includegraphics[scale=.6]{./figures/abstract_rule_metamodel.pdf}
	\caption{Abstract Rule Metamodel for the Police Station Transformation}
	\label{fig:abstract_rule_metamodel_station_trafo}
\end{figure}

% \begin{figure}[h!] \centering \includegraphics[scale=.45]{./figures/hot_activity_diagram.pdf}
% 	\caption{Activity Diagram Repesentation of the HOTs Composing the Verification Tool Architecture}
% 	\label{fig:activity_diagram_verif_tool_arch}
% \end{figure}

### 2016-06-29 text/old/definitions.tex
Move old sections. Divorce paper from definitions file.

--- 2016-05-02 text/definitions.tex
@@
 Abstract classes and their relations do not carry over to the expanded metamodel.
-Expanded metamodels will be used in the subsequent text to facilitate formal the treatment of any structure involving polymorphism.
+Expanded metamodels will be used in the subsequent text to facilitate formal treatment of any structure involving polymorphism.
 
@@
 Metamodel instances do not allow cyclic containment relations, as enforced by EMF.
+\reviewer{Do you mean “in analogy to EMF”?}
 
@@
 Note that the $*$ notation is different from the $\star$ notation, introduced in \cref{def:expanded_metamodel} for an expanded metamodel.
+\reviewer{The difference between these stars is too small to be noticeable.}
 
@@
 Each rule in that set includes a possible replacement of each of the classes in the match part of the original rule by one of its subtypes.
-Expanded transformation rules will be important such that polymorphism is correctly handled in the developments that follow.
+Expanded transformation rules will be important to ensure that polymorphism is correctly handled in the developments that follow.
 
@@
 %<*layertransformationtext>
-\reviewer{I do not fully understand why this
-  could be problematic. Do you have an example?}
+
 \cref{def:layer_transformation} formalises the abstract syntax of a model transformation, introduced in \cref{sec:dsltrans}.
@@
 This condition avoids situations where the execution of a rule in a DSLTrans model transformation necessarily implies the execution of another rule (except for when rules having backward links necessarily execute because all their dependencies were created during the execution of previous layers).
+\reviewer{I do not fully understand why this
+  could be problematic. Do you have an example?}
 
@@
 \textsc{Pathcond}(tr)$ of a path condition of $tr$ and $ex = \langle V',E',st',\tau',Input,Output\rangle \in Exec(tr)$ be an execution of $tr$.
-We have that $ex$ is abstracted by $pc$, noted $ex\Vvdash pc$, if and only if the set of transformation rules of $tr$ combined in $pc$ and the set of transformation rules of $tr$ used to built $ex$ is the same, and:
+We have that $ex$ is abstracted by $pc$, noted $ex\Vvdash pc$, if and only if the set of transformation rules of $tr$ combined in $pc$ and the set of transformation rules of $tr$ used to build $ex$ is the same, and:
 \begin{multline}
@@
 \textsc{Pathcond}(tr)$ of be a path condition of $tr$ and $ex = \langle V',E',st',\tau',Input,Output\rangle \in Exec(tr)$ be an execution of $tr$.
-We have that $ex$ is abstracted by $pc$, noted $ex\Vvdash pc$, if and only if the set of transformation rules of $tr$ combined in $pc$ and the set of transformation rules of $tr$ used to built $ex$ is the same, and:
+We have that $ex$ is abstracted by $pc$, noted $ex\Vvdash pc$, if and only if the set of transformation rules of $tr$ combined in $pc$ and the set of transformation rules of $tr$ used to build $ex$ is the same, and:
 \begin{multline}

### 2016-06-29 text/old/formal_background.tex
Move old sections. Divorce paper from definitions file.

\section{Formal Background}
\label{sec:formal_background}

In this section we will introduce the formal concepts that will be used throughout all this paper. We start in \cref{sec:typed_graphs} by a few (typed) graph concepts that will be used as mathematical building blocks throughout this paper. In particular we will introduce the notion of typed graph, typed graph union and subset, and useful relations between typed graphs based on homomorphisms. Notice that these concepts are well known from graph theory and are only slightly customized for our purposes.

\reviewer{Your formalisation is made more complex than necessary because you do not put
the power of typed graphs to work. If your source and target metamodels have
disjoint types (which you can assume without loss of generality) you can take
their union, augmented with trace edge types from all target node types to all
source node types, and take that to be the meta-model for an input-output model,
and also for a rule. Suddenly you don’t have to carry arround the distinction be-
tween Match and Apply (for rules) or input and output (for input-output models)
around any more: they are just the projections onto the types of the source, resp.
target metamodel.}

\reviewer{Section 4.2.2 on “backward links” is rather confusing, as it speaks about trace-
ability links being added to rules. You never add traceability links to rules any-
where in the paper; in fact, your formal notion of a rule has no room for them.
Instead, rules only have backward links. (The clarity of the the situation is not
improved by the fact that those are labelled trace.)}

\reviewer{I think the situation would be much more clear if you were to use the common
notion of left hand side and right hand side of a rule. The left hand side of a
rule rl is precisely your ||rl||, the right hand side is the entire rl augmented by
your traceability links, as in 4.2.2. Suddenly we are back in a well-known graph
transformation formalism, and your rules are in fact so simple (no deletion, no
negative application conditions, injective matching) that I believe it makes no
difference whether you use algebraic graph transformation or some more con-
structively defined variant. The only standard notion is that of transitive closure
in the left hand side; but there are plenty of GT tools that offer this extension.
In fact I see no reason why your rule application would then not precisely coin-
cide with the the same notion in, say, SPO graph rewriting; and I hope you agree
that this would help you no end in explaining what you are doing. (If, on the
other hand, there is after all some difference then this, too, would be interesting
to know, as that difference is not at all apparent right now.)}

\reviewer{Viewed like this, moreover, I think your notions of input-output models and rules
are very close to triple graph transformation as used by Schrr at all. I think the
main (if not only) difference lies in the fact that your "glue graph" is actually not
a graph; instead, you use traceability edges directly from the target model to the
source model. If you were to turn those edges into nodes with source and target
edges, even that difference would disappear.}

Armed with the fundamental notion of typed graph, we can then introduce other formal concepts in Sections~\ref{sec:DSLTrans_formal}, \ref{sec:DSLTrans_syntax} and \ref{sec:DSLTrans_semantics} which describe the artifacts from the modeling and transformation world that we require for our verification technique. Naturally, we start by introducing the central notion of \emph{metamodel}, allowing the description of the inputs and outputs of a model transformation. Other fundamental notions we will define in this section are \emph{model}, \emph{transformation rule}, \emph{transformation} and the semantic concept of \emph{model transformation execution}. Several auxiliary and intermediate notions for defining the syntax and semantics of our techniques will also be introduced here.

Note that this section presents a collection of formal tools that are used in the subsequent sections of this paper where the contributions of this paper are presented. It is meant as a formal reference for the upcoming formal development. This section can be safely skipped or skimmed by the reader, who can return to these definitions punctually to understand the detailed formal underpinning of our approach.

\reviewer{There is a lot of redundancy in the paper, on several levels:
  - input-output-models, path conditions and properties are
    all defined very similarly. Couldn't you subsume them in a single
    definition and spell out the differences?}

\subsection{Typed Graphs}
\label{sec:typed_graphs}

We will start by introducing the notion of typed graph. A typed graph is the essential object we will use throughout our mathematical development. Typed graphs will be used to formalise all the important graph-like structures we will present in this paper. A typed graph is a directed multigraph (a graph allowing multiple edges between two vertices) where vertices and edges are typed.

\begin{definition}{Typed Graph\\}
\label{def:typed_graph}
A typed graph is a 6-tuple $\langle V,E,(s,t),\tau, VT, ET\rangle$ where: $V$ is a finite set of vertices; $E$ is a finite set of directed edges connecting the vertices $V$; $(s,t)$ is a pair of functions $s: E\rightarrow V$ and $t: E\rightarrow V$ that respectively provide the source and target vertices for each edge in the graph; function $\tau:V\cup E\rightarrow VT \cup ET$ is a typing function for the elements of $V$ and $E$, where $VT$ and $ET$ are disjoint finite sets of vertex and edge type identifiers and $\tau(v)\in VT$ if $v\in V$ and $\tau(e)\in ET$ if $e\in E$. Edges $e\in E$ are noted $v\xrightarrow{e} v'$ if $s(e)=v$ and $t(e)=v'$, or simply \emph{e} if the context is unambiguous. The set of all typed graphs is called $\textsc{Tg}$. 
\end{definition}

We now define how two typed graphs are united. A union of two typed graphs is trivially the set union of all the components of those two typed graphs. Note that we do not require the components of the two graphs to be disjoint, as in the following joint unions will be used to merge typed graphs.

\begin{definition}{Typed Graph Union\\}
\label{def:typed_graph_union}

Let $\langle V,E,(s,t),\tau,VT,ET\rangle,\langle V',E',(s',t'),\tau',VT',ET'\rangle\in \textsc{Tg}$ be typed graphs, where $VT$ and $ET'$ are disjoint sets, as well as $VT'$ and $ET$. The typed graph union is the function $\sqcup :\textsc{Tg}\times \textsc{Tg}\rightarrow \textsc{Tg}$ defined as:
\begin{multline*}
\big\langle V,E,(s,t),\tau,VT,ET\big\rangle\;\sqcup\;\big\langle V',E',(s',t'),\tau',VT',ET'\big\rangle=\\\big\langle V\cup V', E\cup
E',(s\cup s', t\cup t'), \tau\cup \tau', VT\cup VT', ET\cup ET'\big\rangle
\end{multline*}
\end{definition}

\reviewer{Graph union (Def. 2) is only well-defined if s $\cup$ s' , t $\cup$t' and $\tau \cup \tau$' yield functions,
which in turn is only the case if s and s' etc. coincide on common elements.
This is neither stated as a requirement here nor checked anywhere in the use of
the union. (Since I sure you implicitly had this restriction in mind, this may be
counted as sloppiness; you can probably assume this to hold w.l.o.g. wherever
you need it.)}

\reviewer{definition 3:''st'' was before ''(s, t)'' in the previous definitions (idem with ''stâ''), in the same way in definition 13, both notations are used.}
For the formal development of our technique, we are interested in relations between typed graphs that are structure-preserving, i.e. homomorphisms. Homomorphisms between typed graphs preserve not only structure, but also the types of vertices and edges that are mapped.

\begin{definition}{Typed Graph Homomorphism\\}
\label{def:typed_graph_homomorphism}
Let $\langle V,E,st,\tau,VT,ET\rangle=g$ and $\langle V',E',st',\tau',VT',ET'\rangle=g'\in \textsc{Tg}$ be typed graphs. A typed graph homomorphism between $g$ and $g'$ is a function $f:V\rightarrow V'$ such that for all $v_1 \xrightarrow{e} v_2\in E$ we have that $f(v_1) \xrightarrow{e'} f(v_2)\in E'$, where $\tau(v_1)=\tau'(f(v_1))$, $\tau(v_2)=\tau'(f(v_2))$ and also $\tau(e)=\tau(e')$. The domain of $f$ is noted $Dom(f)$ and the co-domain of $f$ is noted $CoDom(f)$. When an \emph{injective} typed graph homomorphism $f$ exists between $g$ and $g'$ we write $g \stackrel{f}{\vartriangleleft} g'$, or simply $g \vartriangleleft g'$ when the context is unambiguous. When a \emph{surjective} typed graph homomorphism $f$ exists between typed graphs $g$ and $g'$ we write $g \stackrel{f}{\blacktriangleleft} g'$, or also simply $g \blacktriangleleft g'$ in an unambiguous context. 
\end{definition}

Note that, trivially, a typed graph homomorphism is a graph homomorphism.

\reviewer{Def. 3 and all subsequent definitions g has an unexplained component st, is
this the same as the pair (s,t) in Def. 1?}

\reviewer{For multigraphs (which may contain parallel edges with the same label) it is
necessary to include edge images in the homomorphism (Def. 3). If you do not
do that, then graphs with parallel edges are isomorphic to graphs with single
edges (between the same nodes), so you might as well restrict to simple graphs.}

\reviewer{
Concretely: Right now the def. requires the existence of \textit{some} unfixed e' in the
target graph, but in fact \textit{f} should be extended to E $\rightarrow$ E' so that e' = f (e) in this
situation.
}

We now define the useful notion of typed subgraph. As expected, a typed subgraph is simply a restriction of a typed graph to some of its vertices and edges.

\begin{definition}{Typed Subgraph\\}
\label{def:typedsubgraph}
Let $\langle V,E,st,\tau,VT,ET\rangle=g,\langle V',E',st,\tau',VT',ET'\rangle=g'\in \textsc{Tg}$ be typed graphs. $g'$ is a typed subgraph of $g$, written $g'\sqsubseteq g$, iff $V'\subseteq V$, $E'\subseteq E$ and $\tau'=\tau |_{V'\cup E'}$.

\end{definition}

Two typed graphs are said to be isomorphic if they have exactly the same shape and related vertices and edges have the same type.

\begin{definition}{Typed Graph Isomorphism\\}
\label{def:typed_graph_isomorphism}
Let $\langle V,E,st,\tau,VT,ET\rangle=g,\langle V',E',st',\tau',VT',ET'\rangle=g'\in \textsc{Tg}$ be typed graphs. We say that $g$ and $g'$ are isomorphic, written $g\cong g'$, if and only if there exists a bijective typed graph homomorphism $f:V\rightarrow V'$ such that $f^{-1}:V'\rightarrow V$ is a typed graph homomorphism.
\end{definition}

\paragraph{\textbf{Notation:}}
In order to simplify our notation, when the context is unambiguous we will abbreviate a typed graph $\langle V,E,st,\\\tau,VT, ET\rangle$ as as 4-tuple $\langle V,E,st,\tau\rangle$. Also, given a typed graph $g\in \textsc{Tg}$, will use the notation $Components(g)$ to describe the set of strongly connected typed graphs in $g$. Finally, we will use the notation $g|_{t}$ to refer to the restriction of graph $g$ to its subgraph containing only edges of type $t$.

\subsection{Metamodel and Model-Related Constructs}
\label{sec:DSLTrans_formal}

We will start by introducing the notion of \emph{metamodel}, which in DSLTrans is used to type the input and output models of a DSLTrans transformation.  Two metamodels, the \emph{organization language} and \emph{gender language} are depicted in \cref{fig:squadmetamodel}.

\begin{definition}{Metamodel\\}
\label{def:metamodel}
A metamodel is a 5-tuple $\langle V,E,st,\tau,\leq\rangle$ where $\langle V,E,st,\tau\rangle\in \textsc{Tg}$ is a typed graph, $(V,\leq)$ is a partial order and $\tau$ is a bijective typing function. Additionally we also have that: if $v\in V$ then $\tau(v)\in VT\times \{abstract,concrete\}$, where $VT$ is the set of vertex type names; if $e\in E$ then $\tau(e)\in ET\times \{containment,\\reference\}$, where $ET$ is a set of edge type names. The set of all metamodels is called $\textsc{Meta}$.
\end{definition}

\reviewer{Def. 6: Since the codomain of $\tau$ is apparently VT $\times$ {abstract, concrete} $\cup$ ET $\times$
{containment, reference}, it is certainly never going to be a bijection. However,
I think you actually only want injectiveness; what purpose is served by requiring
it to be surjective?}

\reviewer{As an aside, I think this is quite an awkward way to encode the notions of ab-
stractness and containment: I think it would be much more elegant to define
abstractness as a predicate over VT and containment as a predicate over ET. To
go even one step further, it would in fact be more natural (and more standard)
to equate the notions of type graph and metamodel altogether, as they certainly
serve the same conceptual purpose.
}

\reviewer{Def. 6 and many other places A word in math mode should always
be in $\backslash$mathit to avoid ugly spacing as in $re f erence$.}

A formal metamodel is a particular kind of typed graph where vertices represent classes and edges represent relations between those classes. A typed graph representing a metamodel has two special characteristics: on the one hand, the typing function for vertices and edges is bijective. This means that each type occurs only once in the metamodel, as is to be expected. On the other hand, a metamodel is equipped with a partial order between vertices. This partial order is used to model inheritance at the level of the metamodel's classes. Note that here we have overridden the co-domain of the typing function in the original typed graph presented in \cref{def:typed_graph} in order to allow distinguishing between \emph{abstract} and \emph{concrete} classes, as well as between \emph{containment} and \emph{reference} edges in our metamodels. For simplification purposes, we do not model association cardinalities in our formal notion of metamodel as cardinalities are not strictly necessary in our development.

\begin{definition}{Expanded Metamodel\\}
\label{def:expanded_metamodel}
Let $mm = \langle V,E,st,\tau,\leq\rangle\in \textsc{Meta}$ be a metamodel. The expansion of $mm$, noted $mm^{\star}$, is a typed graph $\langle V',E',st',\tau'\rangle\in \textsc{Tg}$ built as follows:
\begin{itemize}
  \item $V' = V\setminus \{v\in V\,|\,\tau(v)=(\cdot\footnote{In our mathematical development we use a `dot' notation to represent that we do not care about the value of a particular variable in a given context.},abstract)\}$;
  \item $v_1\xrightarrow{e}v_2\in E'$ if $v_1\xrightarrow{e}v_2\in E$ and $\tau(v_1)=(\cdot,concrete)$ and $\tau(v_2)=(\cdot,concrete)$;
  \item if $v_1\xrightarrow{e}v_2\in E$ we have that $v'_1\xrightarrow{e'}v'_2\in E'$, where $v'_1\leq v_1$, $v'_2\leq v_2$ and $\tau'(e') = \tau(e)$;
  \item for all $v\in V'$ and $e\in E'$ we have that $\tau'(v)=\tau(v)$ and that $\tau'(e)=\tau(e)$.
%   \item $(v,(tv,concrete))\in \tau'$ if $(v,(tv,concrete))\in \tau$ (where $tv$ is a vertex type);
%   \item $(v_1\xrightarrow{e}v_2,(te,concrete))\in \tau'$ if $(v_1\xrightarrow{e}v_2,(te,concrete))\in \tau$ ($te$ ia an edge type).
\end{itemize}
%The set of all expanded metamodels in called $META^{\star}$.
\end{definition}

\reviewer{Def. 7: is an expanded metamodel a metamodel? Then you need to allow non-
injective $\tau$, at least on edges, as the requirement $\tau$(e) = $\tau$(e') in the third bullet
will certainly cause non-inejectiveness.}

An expanded metamodel is an auxiliary construct where all the relations between types of a metamodel are made explicit, rather than remaining implicit in the specialization hierarchy. It is built by adding to the original metamodel typed graph a relation of type $t$ between two classes of the metamodel, whenever those classes specialize two classes that are also related by a relation of type $t$. Abstract classes and their relations do not carry over to the expanded metamodel. Expanded metamodels will be used in the subsequent text to facilitate formal treatment of any structure involving polymorphism.

% Definition~\ref{def:metamodel} formally states that a metamodel corresponds to a graph of typed elements where only one element for each type is represented. This definition is then used to state definition~\ref{def:pattern}, where a typed graph is combined with its metamodel.

\begin{definition}{Metamodel Instance\\}
\label{def:metamodelinstance}

An instance of a metamodel $mm = \langle V',E',st',\tau',\leq\rangle\in \textsc{Meta}$ is a typed graph $\langle V,E,st,\tau\rangle\in \textsc{Tg}$, where the co-domain of $\tau$ equals the co-domain of $\tau'$. Also, there is a typed graph homomorphism $f:V\rightarrow V'$ from $\langle V,E,st,\tau\rangle$ to the expanded metamodel $mm^{\star}$ and the graph $\big\langle V,\{e\in E\,|\,\tau(e)=(\cdot,containment)\}\big\rangle$ is acyclic. The set of all instances for a metamodel $mm$ is called $\textsc{Instance}^{mm}$.

\end{definition}

A metamodel instance is a useful intermediate formal notion that lies between metamodel and model. The injective typed graph homomorphism between a metamodel instance and metamodel models multiple ``instances'' of objects and relations being typed by one single class or relation of the metamodel. Metamodel instances do not allow cyclic containment relations, as enforced by EMF. \reviewer{Do you mean “in analogy to EMF”?}

% It requires the acyclicity constraints as that is required through models, transformation rules, path conditions and properties. It can be used to define transformation rules and can be further refined to define models.

\reviewer{Def. 9 and many thereafter incrementally define new concepts on top of graphs.
Most of these definitions are rather hard to read, for several reasons:}

\begin{itemize}
\item It is common to use an unprimed version for the original object and a
primed version for the constructed object, whereas you turn this convention around
\item Sloppiness: in more than 50 percent of all cases the $\subseteq$ should be $\supseteq$ or vice versa
\begin{itemize}
\item In this particular instance, you really mean $\tau' \supseteq \tau$ and $E = E' \cup E_c^∗$ .
\end{itemize}

\item It is really more readable to define the construction as a construction, so:
\begin{itemize}
\item  $E = E'\cup E^{\star}_{C}$
\item $s = s' \cup {(e_1 \dots e_n , s(e_1)) | e_1 \dots e_n \in E^{\star}_{C} }$
\item $t = t' \cup {(e_1 \dots e_n ,t(e_n )) | e_1 \dots e_n \in E^{\star}_{C} }$
\item $\tau = \tau' \cup {(e, indirect) | e \in E^{\star}_{C} }$
\end{itemize}
\begin{itemize}
\item Note that this also takes care of s and t, which you have forgotten here
and in all other similar definitions.)
\item The set $E^{\star}_{C}$should itself also be more carefully defined, as $E^{\star}_{C} = {e_1 \dots e_n \in E^∗' | \forall1 \leq i < n.t(e_i ) = s(e_{i+1} )}$
\end{itemize}
\end{itemize}

\begin{definition}{Containment Transitive Closure\\}
\label{def:instance_closure}

The containment transitive closure of a metamodel instance $\langle V',E',st',\tau'\rangle\in \textsc{Instance}^{mm}$ is a typed graph $\langle V,E,st,\tau\rangle$ where we have that $V=V'$, $\tau' \supseteq \tau$ and $\tau$'s co-domain is the union of the co-domain of $\tau'$ and the set $\{indirect\}$. We also have that $E' = E\cup E^*_c$ where $E^*_c$ is the transitive closure of the set $\big\{v\xrightarrow{e}v'\,|\,\tau(v\xrightarrow{e}v') = (\cdot,containment)\big\}$ and if $e\in E\setminus E'$ then $\tau(e) = indirect$.  We denote $mi^{*}$ the containment closure of a metamodel instance $mi\in \textsc{Instance}^{mm}$.

\end{definition}

Given a metamodel instance, its containment transitive closure includes, besides the original graph, all the edges belonging to the transitive closure of containment links in that metamodel instance. The transitive edges are typed as \emph{indirect}. In the definitions that follow we will use the $*$ notation, as in \cref{def:instance_closure}, to denote the containment transitive closure of structures that directly or indirectly include metamodel instances. For example, $tg^{*}$ would represent the containment transitive closure of typed graph $tg$ wherever containment edges are found in the graph. Note that the $*$ notation is different from the $\star$ notation, introduced in \cref{def:expanded_metamodel} for an expanded metamodel. \reviewer{The difference between these stars is too small to be noticeable.}

% It requires the acyclicity constraints as that is required through models, transformation rules, path conditions and properties. It can be used to define transformation rules and can be further refined to define models.

\begin{definition}{Model\\}
\label{def:model}

A model of a  metamodel $mm = \langle V',E',st',\tau',\leq\rangle\in \textsc{Meta}$ is a metamodel instance $\langle V,E,st,\tau\rangle\in \textsc{Instance}^{mm}$, such that: there exists an injective typed graph homomorphism $f:V\rightarrow V'$ from $\langle V,E,st,\tau\rangle$ to metamodel $mm^{\star}$ where, if $b\in V$ and there exists an edge $a'\xrightarrow{e'} f(b)\in E'$ where $\tau(e') = (t,containment)$, then we also have that $a\xrightarrow{e} b\in E$, $f(a) = a'$ and $\tau(e)=t$. The set of all models for a metamodel $mm$ is called $\textsc{Model}^{mm}$.

\end{definition}

\reviewer{Def. 10: Here you certainly should not require $\tau$ to be injective. The requirement
on containment relations is mathematical nonsense: if $e' \in E'$ has target b then
$b \in V'$ and so f (b) does not exist, and most certainly is not the source of any edge
in E. In fact, I really have no idea what you want to say here; the explanation
below the definition does not clarify matters at all.}

A model, as per \cref{def:model}, is a metamodel instance where all the containment relations are respected. This means that if an object that is contained (as per the metamodel) exists in the model, then the model will also contain an instance of that containment relation together with an object that contains it. Two models can be observed in \cref{fig:transformationexample}, which are respectively instances of the \emph{Organization language} and \emph{Gender language} found in \cref{fig:squadmetamodel}.\\
Note that the containment constraint does not necessarily lead to infinite models in the case of containment relations with the same source and target classes. In fact, if the cardinality of the target class is allowed to be zero, then it is not necessary that the containment relation is instantiated. For example, this is the case for the containment relation \emph{supervise} in the metamodel of~\cref{fig:OrganizationLanguage}.

\reviewer{“the cardinality of the target class”?? Are you talking about edge multiplicities here (which you have chosen not to include)?}

\begin{definition}{Input-Output Model\\}
\label{def:input_output_model}
An input-output model is a 6-tuple $\big\langle V,E,(s,t),\tau,Input,\\Output\big\rangle$, where: $Input=\langle V',E',st',\tau'\rangle\in \textsc{Instance}^{sr}$ is a model; $Output=\langle V'',E'',st'',\tau''\rangle\in \textsc{Instance}^{tg}$ is a metamodel instance; $Input$ and $Output$ are disjoint. Additionally we have that $V=V'\cup V''$, $E\subseteq E'\cup E''$ and $\tau\subseteq \tau'\cup \tau''$, where the co-domain of $\tau$ is the union of the co-domains of $\tau'$ and $\tau''$ and the set $\{trace\}$. An edge $e \in E\setminus E'\cup E''$ is called a \emph{traceability link} and is such that $s(e)\in V''$, $t(e)\in V'$ and $\tau(e)=trace$. The set of all match-apply patterns for a source metamodel $sr$ and a target metamodel $tg$ is called $\textsc{Iom}^{sr}_{tg}$.
\end{definition}

\reviewer{Why is Input a model and Output a metamodel instance? Note that
$\backslash$ binds stronger than $\cup$ in $E \backslash E' \cup E$ ′′ . “match-apply patterns” → “input-output
patterns”}
%Change if needed
% \levi{A match-apply model is the union of an input and output models, plus potential traceability links between those models. Traceability links are now typed using the typing function $\tau$. A traceability link it typed with type ``trace''. This makes the definitions shorter and more compact, which was mentioned by one of the reviewers of the paper. However, the fact that a match-apply model is now made up of models makes it that the input and the output of the transformation are always well-formed, meaning that the containment relation is always respected. It may be that a transformation does not always produce a model respecting the containment relation, which is a problem. This can be solved however if we only consider that well-formed transformations actually have semantics.}

An input-output model is an object we will use when defining the semantics of a DSLTrans model transformation in \cref{sec:DSLTrans_formal}. It is composed of two metamodel instances, one called the \emph{input} and the other one the \emph{output}. An input-output model allows the representation of intermediate operational states during the execution of a model transformation. It may include a particular type of edges called \emph{traceability links}, for keeping a history of which elements in the output model originated from which elements in the input model.

\begin{definition}{Metamodel Pattern and Indirect Metamodel Pattern\\}
\label{def:metamodel_pattern}

A pattern of a metamodel $mm\in \textsc{Meta}$ is an instance of $mm$. Given a metamodel pattern $\langle V',E',st',\tau'\rangle\in \textsc{Instance}^{mm}$ we have that $\langle V,E,st,\tau\rangle$ is an \emph{indirect} pattern if $V=V'$, $E'\supseteq E$ and the co-domain of $\tau$ is the union of the co-domains of $\tau'$ and the set $\{indirect\}$. Also, if $e\in E\setminus E'$, then we have that $\tau(e)=indirect$. Given a metamodel $mm$, the set of all metamodel patterns for $mm$ is called $\textsc{Pattern}^{mm}$. The set of all indirect metamodel patterns for $mm$ is called $\textsc{Ipattern}^{mm}$.
\end{definition}

Metamodel patterns are introduced in \cref{def:metamodel_pattern} as an intermediate notion, formally equal to metamodel instances. An \emph{indirect} metamodel pattern is a metamodel pattern that includes edges typed as \emph{indirect}. Both structures will be used as building blocks in the construction of transformation-related structures in the upcoming text.

\subsection{Syntactic Transformation Constructs}
\label{sec:DSLTrans_syntax}

This section will detail the abstract syntax of the constructs involved in a DSLTrans transformation.

\begin{definition}{Transformation Rule\\}
\label{def:transformation_rule}

A transformation rule is a 6-tuple $\big\langle V,E,(s,t),\tau, Match,\\Apply\big\rangle$, where: $Match=\langle V',E',st',\tau'\rangle\in \textsc{Ipattern}^{sr}$ such that: $Match\neq \epsilon$\footnote{We use the simplified $\epsilon$ notation to denote empty n-tuples structures.} is an non-empty indirect metamodel pattern; $Apply=\langle V'',E'',st'',\tau''\rangle\in \textsc{Pattern}^{tg}$ such that $Apply\neq\epsilon$ is a metamodel pattern; $Match$ and $Apply$ are disjoint. We also have that $V=V'\cup V''$, $E\subseteq E'\cup E''$ and $\tau\subseteq \tau'\cup \tau''$, where the co-domain of $\tau$ is the union of the co-domains of $\tau'$ and $\tau''$ and the set $\{trace\}$. An edge $e \in E\setminus E'\cup E''$ is called a \emph{backward link} and is such that $s(e)\in V''$, $t(e)\in V'$ and $\tau(e)=trace$. We additionally impose that there always exists a $v_1\in V''$ in the Apply part of the rule such that $\nexists e\,.\, v_1\xrightarrow{e} v_2$ and $\tau(e)=trace$, or that $E''$ is not empty. The set of all transformation rules for a source metamodel $sr$ and a target metamodel $tg$ is called $\textsc{Rule}^{sr}_{tg}$.

\end{definition}

\reviewer{e in $\tau(e) = \mathit{trace}$ is still bound by $\nexists$, so “and” should be $\land$}

A transformation rule is the elemental block of a model transformation. Several transformation rules can be observed in the Police Station transformation in \cref{fig:dsltransformation}. A formal transformation rule includes a non-empty match pattern and a non-empty apply pattern (also known in the model transformation literature as a rule's \emph{left hand side} and \emph{right hand side}). The apply pattern of a rule always contains at least one apply element that is not connected to a backward link or an edge, meaning in practice that a rule will always produce something and not only match. A match pattern can include indirect links that are used to transitively match containment relations in a model. An apply pattern does not include indirect links as it is used only for the construction of parts of instances of a metamodel. A transformation rule includes backward links, as informally introduced in \cref{subsec:DSLTrans_constructs}. Backward links are formally typed as \emph{trace}.%This eases modeling matching rules on models  by the use of typed graph homomorphisms.

\begin{definition}{Matcher of a Transformation Rule\\}
\label{def:back_match_transformation_rule}

Let $rl = \big\langle V,E,st,\tau, Match,Apply\big\rangle$ be a transformation rule where $Match =\langle V_m, E_m, st_m, \tau_m\rangle$. We define $rl$'s matcher, noted $\Vert rl\Vert$, as the transformation rule $\big\langle V',E',st',\tau',Match,\\Apply'\big\rangle\sqsubseteq rl$ where $v_1\xrightarrow{e} v_2\in E'$ if and only if $v_1,v_2\in Match$ or $\tau(e)=trace$ and $V'=V_m\cup \big\{v_1\,|\,v_1\xrightarrow{e} v_2\in E \land \tau(e)=trace\big\}$.
\reviewer{Definition 14: where does Apply' come from? Should v1 be 
  an element of Apply?}
\end{definition}

\reviewer{Def. 14: According to your own criterion, $||rl||$ is not a rule as it does not create
either a node or an edge.}

\Cref{def:back_match_transformation_rule} introduces the notion of matcher for a transformation rule which consists solely of the match pattern of a rule and its backward links, if any. The matcher of a rule constitutes the complete pattern that a DSLTrans rule attempts to match over a input-output model during rule execution. Traceability links between input and output model elements generated during transformation execution are matched by transformation rules' backward links, as informally explained in \cref{subsec:DSLTrans_constructs}.

\begin{definition}{Expanded Transformation Rule\\}
\label{def:transformation_rule_expansion}

Let $rl = \big\langle V,E,st,\tau,Match,Apply\rangle \in \textsc{Rule}^{sr}_{tg}$ be a transformation rule where $Match = \langle V',E',st',\tau'\rangle$ and also we have that $sr =\langle V'',E'',st'',\tau'',\leq \rangle$. The expansion of $rl$, noted $rl^{\star}$ is a set of transformation rules built as follows:
\begin{itemize}
  \item $rl\in rl^{\star}$;
  \item $\big\langle V,E,st,\tau',Match,Apply\rangle\in rl^{\star}$ iff for all $v\in V'$ we have that $\tau'(v)\leq \tau(v)$.
\end{itemize}
\end{definition}

\reviewer{The first bullet is a special case of the second, so can be removed.}

The expansion of a transformation rule is a set of transformation rules. Each rule in that set includes a possible replacement of each of the classes in the match part of the original rule by one of its subtypes. Expanded transformation rules will be important to ensure that polymorphism is correctly handled in the developments that follow.

%\levi{A transformation rule expansion simply takes a rule and builds all possible rules where a rule element is replaced by one of its subtypes.}

\begin{definition}{Layer, Model Transformation\\}
\label{def:layer_transformation}

A layer is a finite set of transformation rules $l\subseteq \textsc{Rule}^{sr}_{tg}$. The set of all layers for a source metamodel $sr$ and a target metamodel $tg$  is called $\textsc{Layer}^{sr}_{tg}$. A model transformation is a finite list of layers denoted $[l_{1}::l_{2}::\ldots::l_{n}]$ where $l_{k}\in \textsc{Layer}^{sr}_{tg}$ and $1\leq k \leq n$, $n\in \mathbb{N}$. We also impose that for any pair of rules $rl_1,rl_2\in \bigcup_{1\leq k\leq n}l_k$, if $\Vert rl_1\Vert \cong rl$ and $rl\sqsubseteq \Vert rl_2\Vert $ then $rl_2$ appears in a layer later than $rl_1$ and the apply parts of $rl_1$ and $rl_2$ are not isomorphic. The set of all transformations for a source metamodel $s$ and a target metamodel $t$ is called $\textsc{Transf}^{sr}_{tr}$.

% The expansion of a layer $l\subseteq RULE^{sr}_{tg}$, noted $l^{\star}$, is such that that if
% $rl\in l$ then $rl'\in l$ if $rl'\in rl^{\star}$. This means that if a
% transformation rule is present is the layer, its expansion is also part of the
% layer.

\end{definition}

\reviewer{The second half of Def. 16 raises a lot of questions. First of all, neither $\sqsubseteq 	 $ nor
$\cong$ are defined over rules; however, I can imagine what it would look like. Then,
what role does rl play; do you just mean $||rl_1 || \sqsubseteq ||rl_2 ||$ given a suitable isomorphic copy? What if $||rl_1 || \cong ||rl_2 || $(as a special case); is your condition then not
unsatisfiable as $rl_1$ must appear after $rl_2$ as well as vice versa?}

\reviewer{More importantly though, why do you care? Your semantics makes sure that
rules are never considered for application twice in succession, so dependencies
are always resolved unambiguously.}

\cref{def:layer_transformation} formalises the abstract syntax of a model transformation, introduced in \cref{sec:dsltrans}. An example of a model transformation can be observed in \cref{fig:dsltransformation}, the Police Station transformation. As expected, a formal DSLTrans transformation is composed of a sequence of layers where each layer is composed of a set of rules. The last condition of \cref{def:layer_transformation} imposes that, for any two pair of rules in the transformation, the matcher of the second rule never partially or totally subsumes (or contains) the matcher of the first rule, unless the second rule is in a subsequent layer and produces something more than the first rule. This condition avoids situations where the execution of a rule in a DSLTrans model transformation necessarily implies the execution of another rule (except for when rules having backward links necessarily execute because all their dependencies were created during the execution of previous layers). \reviewer{I do not fully understand why this
  could be problematic. Do you have an example?}

\paragraph{\textbf{Notation:}}

We naturally extend to transformation rules (\cref{def:transformation_rule}) and transformation executions (\cref{def:modeltransformation}) the typed graph operations introduced in \cref{sec:formal_background}.  Also, given a structure such as transformation rule $rl = \langle V,E,st,\tau,Match,\\Apply\rangle$, we will refer to the structure's components by using the component's name followed by the variable that holds the structure in between parenthesis. For example, we will write $V(rl)$ to designate the $V$ component of $rl$ or $Apply(rl)$ to designate $rl$'s $Apply$ component.

\subsection{Semantic Transformation Constructs}
\label{sec:DSLTrans_semantics}

In the definition that follows we introduce the notion of execution of a DSLTrans model transformation. For our purposes it is sufficient to introduce it as an input-output model (see \cref{def:input_output_model}), containing the input model for the transformation, the produced output, and the traceability links built during execution. Due to space limitations, we cannot introduce the semantics of DSLTrans in the main text of this paper. We thus refer the reader to \cref{sec:DSLTrans_formal} for a formal description on how DSLTrans transformation executions are built.

\reviewer{Definition 17: There is a copy/paste error w.r.t. Def. B.16 (which only serves to
strengthen my point, made above, about avoiding this kind of duplication). In
this case, B.16 is the correct version.}

\begin{definition} {Model Transformation Execution\\}
\label{def:modeltransformation}

Let  $tr\in \textsc{Transf}^{sr}_{tg}$ be a transformation and $input \in \textsc{Model}^{sr}$ be a model. Assume we also have that: $$\langle V,E,st,\tau,input,\epsilon\rangle,tr \xrightarrow{trstep} \langle V',E',st',\tau',input,output\rangle$$
A model transformation execution is the input-output model $\langle V',E',st',\tau',input,output\rangle\in \textsc{Iom}^{sr}_{tg}$, where $output\in \textsc{Iom}^{sr}_{tg}$ is an input-output model. The set of all model transformation executions for transformation $tr$ is written $\textsc{Exec}(tr)$. A model transformation with an empty input model is noted $\epsilon_{ex}$. Note that relation $trstep$ is formally defined in \cref{sec:DSLTrans_formal_appendix}.
\end{definition}

Finally, as stated in \cref{def:modeltransformation}, we consider a model transformation execution to be the input-output model (IOM) resulting from executing a set of rules on a starting IOM. This starting IOM includes the transformation's input in its input part and has an empty output part. The starting IOM represents the first step of the transformation when no rule has been executed yet. A transformation execution results from executing all the rules in a DSLTrans model transformation.

\subsubsection{Transformation Language Semantics}

We will now address the semantics of the DSLTrans language. We will start by defining a match function that, given an input-output model and a transformation rule, returns all subgraphs of that input-output model where the rule's match pattern (including backward links) is found.

% \levi{Need to be careful throughout all this section to understand whether we model the match and apply functions as functions or relations. Because of the non-deterministic generation of new vertices in the definition of apply function~\ref{def:apply_function}, relations are used in the definitions that follow to evaluate a transformation. This is an issue of the formalism used to model, should we make it explicit?}

\begin{definition}{Match Function}
\label{def:match_function}

Let $m_{in}\in \textsc{Iom}^{sr}_{tr}$ be a input-output model and $rl\in \textsc{Rule}^{sr}_{tg}$ be a transformation
rule. The $match : \textsc{Iom}^{sr}_{tg}\times \textsc{Rule}^{sr}_{tg}\rightarrow
\mathcal{P}(\textsc{Iom}^{sr}_{tg})$ function is defined as follows: $$match_{rl}(m_{in})= \big\{ glue_{noind}\;|\;
glue\sqsubseteq m_{in}^{*} \land glue \cong \Vert rl\Vert\big\}$$

where $glue\in \textsc{Iom}^{sr}_{tr}$ is an input-output model and $glue_{noind}$ is a version of $glue$ where the indirect links have been removed.
% \levi{all backward links are replaced by trace links such that the matching using subgraph isomorphim can work.}
% 
% \levi{Where $rl^{-}$ is rule $rl$ without nodes in the apply part that are not connected to \emph{trace links}, as well as links adjacent to those nodes. $mg^{-ind}$ is the match graph without indirect links that come from the rule, given those cannot be part of the intermediate $MAM$ match result since that result is going to be united with the original $MAM$ (this was incorrect in the original SLE paper).}
% \levi{The matcher function returns all graphs of the input MAM that are matched by a rule, including its backward links. Note that to find a subgraph ($\sqsubseteq$) of the input MAM that is isomorphic ($\cong$) to the rule being matched because the graph rewriting is achieved by joint union of the match graph extended by the new rewrite node, and the original MAM.}
% \levi{TODO: review the $mg^{-ind}$ and the $rl^{-}$ notations.}

% Due to the fact that the $\cong$ relation is based on the notion of graph
% isomorphism, permutations of the same match result may exist in the
% $\big\{g\;|\; g\lhd m \land g \cong strip(tr)\big\}$ set. The --- undefined ---
% $remove:\mathcal{P}(TR^{s}_{t})\rightarrow \mathcal{P}(TR^{s}_{t})$ function is such that
% it removes such undesired permutations.

%The $strip:TR^{s}_{t}\rightarrow TR^{s}_{t}$ function is such that
%$$strip(\big\langle V,E,T,\langle V_m,E_m,T_m,s\rangle,\langle V_a,E_a,T_a,t\rangle,Bl,Il\big\rangle) = \big\langle V',E',T,\langle V_m,E_m,T_m\rangle,\langle V'_a,E'_a,T_a\rangle,Bl,Il\big\rangle$$
%\begin{center}
%where $(v'\in V'_a\Rightarrow v\rightarrow v'\in Bl)\;\land\;(v\rightarrow v'\in E'_a\Rightarrow (v\rightarrow v'\in E_m \land \{v,v'\}\subseteq V'_a))$
%\end{center}
\end{definition}

\reviewer{Def. B.12. First of all, I wonder why you define match as a function rather than
just defining what a match is; namely, an element of the set you are constructing
here. Secondly (and more importantly), why do you not use the notion of homo-
morphism instead of explicitly talking about ismorphic subgraphs? Your way,
you lose the connection between glue and $||rl||$: you know they are isomorphic,
but what if $||rl||$ has symmetries - then there is more than one match, which in
your definition cannot be distinguished.}

\reviewer{You pay the price for this loss of information in Def. B.13, which now is broken: $a\Delta$ depends on $m_{glue} $(which was called glue in Def. 12, why the change in
notation?) in a way depending on the mapping from $m_{glue}$ to $rl$ which you have
discarded.}

\reviewer{Def. B.13 is broken in another way as well: it does not properly define E' but
merely imposes some requirements that are also fulfilled by choosing E' = E.
Instead I know from the discussion about traceability links in Section 4 (which
logically comes after this appendix!) that you really want to add such links for
all newly created nodes of the output graph to all pre-existing nodes of the input
graph; but that is not made clear anywhere at this point.}

The match function in \cref{def:match_function} looks for subgraphs ($glue\sqsubseteq m_{in}^{*}$) of an input-output model that are isomorphic to the backward matcher of the given transformation rule ($glue \cong \Vert rl\Vert$). Note that the containment transitive closure of the input-output model ($m_{in}^{*}$) is considered such that indirect links in the rule can looked for in the input model. Additionally, indirect links need to be removed from the input-output models resulting from the match function ($glue_{noind}$). This is so because indirect links are not part of the original input model, but rather an auxiliary structure.

Let us now turn our attention to the apply function in \cref{def:apply_function}. Its role is to extend all model fragments found by matching a rule on a given input-output model, such that each of those fragments becomes isomorphic to the complete rule (minus its backward links). This process effectively creates the new objects and relations specified in the apply part of the rule, for each of the fragments found when matching the rule.

\begin{definition}{Apply Function}
\label{def:apply_function}

\reviewer{In your definitions you rely on the $\sqcup$ operator (sqare union),
  the definition of which is not entirely clear.  For instance, I do not know where the operator used in Def. B13
  is defined.}

Let $m_{glue}\in \textsc{Iom}^{sr}_{tg}$ be a input-output model and $rl\in \textsc{Rule}^{sr}_{tg}$ a
transformation rule. The $apply : \textsc{Iom}^{sr}_{tg}\times \textsc{Rule}^{sr}_{tg}\rightarrow \textsc{Iom}^{sr}_{tg}$ function
is defined as follows:

$$apply_{rl}(m_{in})=\bigsqcup_{m_{glue}\in match_{rl}(m_{in})}trace_{a_{\Delta}}(m_{glue}\sqcup a_{\Delta})$$

\begin{center}
where $a_{\Delta}\in \textsc{Iom}^{sr}_{tg}$ is such that $m_{glue} \sqcup a_{\Delta}\cong rl_{noind}$.\\
\end{center}

We impose that any instance of $a_\Delta$ is always disjoint from the $m_{in}$ input-output model and also that any two instances of $a_\Delta$ used in the large union are always disjoint.

Partial function $trace:\textsc{Iom}\times \textsc{Iom} \nrightarrow \textsc{Iom}$ is such that $trace_{\langle V_{\Delta},E_{\Delta},st_{\Delta},\tau_{\Delta}\rangle}(\langle V,E,st,\tau\rangle) = \langle V,E',st',\tau'\rangle$ where we have that $E\subseteq E'$, $st\subseteq st'$ (using a light notational abuse for the $s\subseteq s'$ and $t\subseteq t'$), $\tau \subseteq \tau'$ and if $v_1\xrightarrow{e} v_2\in E'\setminus E$ then $v_1\in Output(V_{\Delta})$, $v_2\in Input(V)$ and $\tau'(e)=trace$. Finally, $rl_{noind}$ is a version of $rl$ where indirect links have been removed.
% More concretely, if we have $mam = $\big\langle V,E,st,\tau, Input,Output\big\rangle\in MAM^{sr}_{tg}$ where $Input=\langle V',E',st',\tau'\rangle\in MODEL^{sr}$ and $Output=\langle
% V'',E'',st'',\tau''\rangle\in MODEL^{tg}$, then $trace(mam) = \big\langle V,E,st,\tau''',Input,Output\big\rangle$ where $E'\subseteq E$ and $v\rightarrow E'\setminus E$ 
\end{definition}
% 
% it builds traceability edges between vertices of the output model that were newly built by the $apply$ function (the vertices in ${mg}_{\Delta}$) and all the vertices from the input part of the input-output model. It is necessary to remove the indirect links from the rule because otherwise they would be created during the rewriting by the apply function.

In \cref{def:apply_function} $a_\Delta$ is an input-output model that contains an instance of the target metamodel. These instances are created by rule $rl$ and are used to extend the sub-models found by the match function. The $trace$ function builds traceability edges between newly created vertices of the output model in $a_{\Delta}$ and all the vertices from the input part of a model fragment found by the match function.

Note that, because we do not pose any constraints on $a_\Delta$ other than the fact that its union with the sub-model $m$ is isomorphic to $rl_{noind}$, the $a_\Delta$ variable can always be satisfied by an unlimited amount of input-output models. In order to avoid an infinite amounts of results when a transformation rule is executed, in what follows we will consider transformation results differ only up to typed graph isomorphism.

%  Note that because the vertices are not disjoint, when the graph produced by the apply function is united with the original match apply model the new vertices are created. It was necessary to remove the indirect links from the rule during rewriting, because otherwise those indirect links would also be rewritten.

Definitions~\ref{def:match_function} and~\ref{def:apply_function} are complementary: the former gathers all the fragments of an input-output model that are matched by a transformation rule; the latter glues on the output part of each of those fragments new objects and relations created by a transformation rule.

%  The $strip$ function is used to
%  enable matching over backward links but not elements to be created by the
%  transformation rule. The $back$ function connects all newly created vertices
%  to the elements of the source model that originated them.

%\begin{definition}{Transformation}

%Let $g\in TG$ be a typed graph. The graph transform function $transform : TR\times TG\rightarrow TG$ is recursively defined as:
%\begin{gather*}
%  transform_{(m,a)}(g) =
%  \begin{cases}
%    g  & \text{if } match_{m}(g)=\emptyset \\
%    instance(a)\cup transform_{(m,a)}(g') &\text{if } match_{m}(g)\neq \emptyset
%  \end{cases}
%\end{gather*}
%$$\text{where } g'=mark_{(m)}(g)$$
%\end{definition}

%\subsection{Transformation Semantics}
%\begin{definition} {Layer Semantics}

%Let $l\in Layer$ be a layer and $\{m,m'\} \in MAM^{s}_{t}$ be a match
%$$\frac{}
%{m,\emptyset \stackrel{layerstep}{\rightarrow}m}$$
%$$\frac{tr\in layer,\; transform_{tr}(m) = m''\; m'',layer\backslash\left\{tr\right\} \xrightarrow{layerstep}	 m'}{m,layer \xrightarrow{layerstep} m'}$$
%\end{definition}

\begin{definition} {Layer Step Semantics}
\label{def:layer_step_semantics}

Let $l\in \textsc{Layer}^{sr}_{tg}$ be a Layer. The \emph{layer step relation}
$\stackrel{layerstep}{\rightarrow}\subseteq \textsc{Iom}^{sr}_{tg}\times \textsc{Iom}^{sr}_{tg} \times \textsc{Layer}^{sr}_{tg}\times
\textsc{Iom}^{sr}_{tg}$ is defined as follows:

$$\frac{}
{\langle m_{in},m_{glue},\emptyset\rangle \xrightarrow{layerstep}m_{in} \sqcup m_{glue}}$$

\begin{multline*}
\frac{\begin{array}{ll} rl\in l,\; apply_{rl}&(m_{in}) = m_{rout},\\
&\langle m_{in},m_{glue}\sqcup m_{rout},l\backslash \{rl\}\rangle \xrightarrow{layerstep} m_{out}
\end{array}}
{\langle m_{in},m_{glue},l\rangle \xrightarrow{layerstep} m_{out}}$$
\end{multline*}

\begin{center}
where $m_{rout} \in\textsc{Iom}^{sr}_{tg}$ and $rl\in \textsc{Rule}^{sr}_{tg}$.
\end{center}

We impose that all input-output models that are part of $rout$ and have been generated by rule $rl$ are disjoint from input-output models accumulated in $m_{glue}$ that have been generated by other rules.

\end{definition}

\reviewer{Def. B.14: It seems to me rather roundabout to sequentially apply all rules, ac-
cumulate the results and then take their union. Why do you not take the union
simultaneously with the parallel application of all rules (of a given layer), as
you are essentially doing for all applications of a single rule in Def. B.13? This
would get rid of the artificial sequential ordering, and the whole issue of con-
fluence would not even have to be raised as there is no non-determinism any
more.}

\reviewer{In the constructions defined in Appendix B, you ignore the required absence of
containment cycles. Might such cycles not appear in the union of the graphs,
even if the individual graphs are cycle-free?}

In \cref{def:layer_step_semantics} we build the result of executing a layer of a DSLTrans transformation. The operational semantics-like rules in the definition execute each rule $rl$ in layer $l$, in a non-deterministic order, by using the $apply$ function. The result of executing each rule is accumulated in the temporary $m_{glue}$ input-output model. Finally, when the set of transformation rules in the layer has been exhausted, the result of executing all the rules in the layer (now contained in $m_{glue}$) is united with the input-output model $m_{in}$, the input to layer $l$. Note that this final union produces the result we expect because of the fact that the $m_{glue}$ input-output model is not disjoint from $m_{in}$. The common ``glue'' parts of $m_{glue}$ that have been built by the match function and extended by the apply function are now used to built the result of executing layer $l$.

\cref{def:layer_step_semantics} is the core of DSLTrans' semantics. Many model transformation languages are based on graph rewriting, where the result of each rule rewrite is immediately usable by all other rules. In DSLTrans the result of executing one layer in DSLTrans is totally produced before the input to the layer is changed. This is enforced in \cref{def:layer_step_semantics} by the fact that the apply function always executes over the same $m_{in}$ input-output model and all the results of rule execution in the same layer are added to the $m_{glue}$ structure that is write-only. Rules belonging to the same layer are thus forced to execute independently.

\begin{definition} {Transformation Step Semantics}
\label{def:transformation_step_semantics}

Let $[l::tr]\in \textsc{Transf}^{sr}_{tg}$ be a transformation, where $l\in
\textsc{Layer}^{sr}_{tg}$ is a Layer and $tr$ also a transformation. The \emph{transformation step relation}
$\stackrel{trstep}{\rightarrow}\subseteq \textsc{Iom}^{sr}_{tg}\times \textsc{Transf}^{sr}_{tg}\times
\textsc{Iom}^{sr}_{tg}$ is defined as follows: 
$$\frac{}{\langle m,[]\rangle \xrightarrow{trstep} m}$$

$$\frac{\big\langle m_{in},\epsilon,l^{\star}\big\rangle \xrightarrow{layerstep} m_{inter},\; \langle m_{inter},R\rangle \xrightarrow{trstep} m_{out}}{\langle m_{in},[l::T]\rangle \xrightarrow{trstep} m_{out}} \hspace{.3cm}$$

$$\text{where } l^{\star}=\bigcup_{rl\in l}rl^{\star}$$
\end{definition}

\reviewer{I believe tr, T and R in this definition stand for the same thing.}
While the execution of the rules belonging to a layer happens in parallel, the execution of the layers of a transformation happens sequentially. As per \cref{def:transformation_step_semantics}, the input-output model $m_{inter}$ is the output of executing a given layer that is passed onto the next layer as input. Note that an empty input-output model ($\epsilon$) is passed as the second argument to the $layerstep$ relation in \cref{def:transformation_step_semantics}. This is because in \cref{def:layer_step_semantics} of layer step semantics, the second argument of the relation is used as an accumulator for the model fragments that are added to be added to the output of the previous layer once all the rules in the current layer have been executed. The transformation rules in a layer are expanded before execution ($l^{\star}$) such that polymorphism in the match elements can be handled (see \cref{def:transformation_rule_expansion_appendix}).

\begin{definition} {Model Transformation Execution}
\label{def:modeltransformation_appendix}

\CatchFileBetweenTags{\modeltransformation}{text/definitions}{modeltransformationappendix}{\modeltransformation}
\end{definition}

\CatchFileBetweenTags{\modeltransformationtextappendix}{text/definitions}{modeltransformationtextappendix}{\modeltransformationtextappendix}

% A transformation execution is formed from taking the input model, and executing
% the transformation on it to produce the output model. During this
% transformation, traceability links will be placed between match elements and the
% apply elements they create. Definition~\ref{def:transf_ex} expresses this
% formally. Note that we assume that transformation executions are built following
% the algorithm described in~\cite{DBLP:conf/sle/BarrocaLAFS10}.

% \begin{definition}{Transformation Execution}
% \label{def:transf_ex}
% 
% Let $t$ be a DSLTrans transformation having source metamodel $s$ and target
% metamodel $t$. A Transformation Execution is a 6-tuple $\langle V,E\cup
% Tl,\tau,Match,Apply,Tl\rangle$, where $\langle V,E,\tau,Match,Apply\rangle \in
% MAP^{s}_{t}$ is a match-apply pattern. $Match=\langle V',E', \tau',s\rangle$,
% $Apply=\langle V'',E'', \tau'',t\rangle$ and the edges $Tl\subseteq V'\times
% V''$ are called \emph{traceability links}. The set of all transformation
% executions having source metamodel $s$ and target metamodel $t$ is called
% $Exec^{s}_{t}$.
% \end{definition}

We now prove two important properties about executions of transformations expressed in the subset of DSLTrans presented in this paper: \emph{confluence} and \emph{termination}. The proofs are provided at a high level, given the fact that DSLTrans essentially enforces both these properties by construction of the semantics of DSLTrans.

\begin{proposition}{Confluence}

Every model transformation execution is confluent up to typed graph isomorphism.
\end{proposition}
\begin{pf}
We want to prove that for every model transformation execution of a transformation $tr\in \textsc{Transf}^{sr}_{tg}$ having as input a model $input \in MODEL^{s}$, its output is always the same up to typed graph isomorphism.\\
If we assume an execution of the transformation is not confluent then this should happen because of non-determinism when the execution of a transformation is being built. Non-determinism happens during the construction of a transformation execution at two points: 
\begin{enumerate}
\item in definition~\ref{def:apply_function}, $a_{\Delta}$ is non-deterministic up to typed graph isomorphism, which does not contradict the proposition we are trying to prove.
\item in definition~\ref{def:layer_step_semantics} transformation rule $rl$ is chosen non-deterministically from layer $l$. Thus, the order in which the transformation rules are treated is non-deterministic. However, by \cref{def:layer_step_semantics} rules in a layer execute independently, which means no-side effects from rule ordering influence the execution of rules in the same layer. Also, the increments to the transformation by each rule of a layer in \cref{def:layer_step_semantics} are united using $\sqcup$ (see \cref{def:typed_graph_union_appendix}), which is an operation that is commutative by construction and thus renders the transformation result from each layer deterministic. 
\end{enumerate}
Given there are no other sources of non-determinism when building the execution of a transformation, every model transformation execution is confluent up to typed graph isomorphism.
\end{pf}

\begin{proposition}{Termination}

Every model transformation execution terminates.
\end{proposition}
\begin{pf}
Let us assume that there is a transformation execution which does not terminate. In order for this to happen there must exist a part in the construction of the execution of a transformation which induces an algorithm with an infinite amount of steps. We identify three moments when this can happen:
\begin{enumerate}
\item if definition~\ref{def:layer_step_semantics} of execution of a layer induces an infinite amount of steps. The only possibility for this to happen is if a layer has an infinite amount of transformation rules, which is a contradiction with definition~\ref{def:layer_transformation_appendix}.
\item if definition~\ref{def:transformation_step_semantics} of execution of a transformation induces an infinite amount of steps. Given layers are executed sequentially and no looping is allowed, the only possibility for this to happen is if the transformation has an infinite amount of layers, which contradicts definition~\ref{def:layer_transformation_appendix}.
\item if the result of the $match_{rl}(m_{in})$ function in definition~\ref{def:match_function} is an infinite set of match-apply graphs. The input-output model $m_{in}$ is by definition finite and the matching of each rule is independent from the execution of other rules in the same layer. As such, the number of subgraphs of $m_{in}$ isomorphic to $rl$'s matcher found during the execution of $rl$ is finite.
\end{enumerate}
Given there are no other constructs in the semantics of a transformation that can induce an infinite amount of steps, every model transformation execution terminates.
\end{pf}

### 2016-06-29 text/property_verification.tex
Move old sections. Divorce paper from definitions file.

@@
 \label{def:trans_prop}
-\CatchFileBetweenTags{\propertydef}{text/definitions}{propertydef}{\propertydef}
+
+\reviewer{You can not handle properties with multiple instances of rule 
+    element (example in Fig. 25). On p.14 in the answer section,
+    you claim that such conditions are excluded in Def. 29. However,
+    the last lines of this definition are not easy to understand.
+    Please explain why there are there (by giving the explanation
+    from the answers section.) Without these explanations it is very hard to see why 
+    Prop. 3 should hold.}
+
+Let $tr\in \textsc{Transf}^{sr}_{tg}$ be a DSLTrans transformation.
+A property of $tr$ is a 6-tuple $\langle V,E,(s,t),\tau,Pre,Post\rangle$, where $Pre = \langle V',E',st',\tau'\rangle \in \textsc{Ipattern}^{sr}$ and $Post = \langle V'',E'',st''\tau''\rangle \in\\ \textsc{Ipattern}^{tg}$ are indirect metamodel patterns.
+We also have that $V=V'\cup V''$, $E\subseteq E'\cup E''$ and $\tau\subseteq \tau'\cup \tau''$, where the co-domain of $\tau$ is the union of the co-domains of $\tau'$ and $\tau''$ and the set $\{trace\}$.
+An edge $e \in E\setminus E'\cup E''$ is called a \emph{traceability link} and is such that $s(e)\in V''$, $t(e)\in V'$ and $\tau(e)=trace$.
+Finally we have that there is at least one path condition $\big\langle V_{pc},E_{pc},st_{pc},\tau_{pc}, Match,Apply,Rule\big\rangle\in \textsc{Pathcond}(tr)$ for which a surjective typed graph homomorphism $m\stackrel{f}{\blacktriangleleft} Pre$ exists, where $m\sqsubseteq Match$ and $f(v)\neq f(v')$ if $v$ and $v'$ are elements of the path condition belonging to the same rule of set $Rule$.
+The set of all properties of transformation $tr$ is called $\textsc{Property}(tr)$.
+
 \end{definition}
@@
 \label{def:sat_prop_ex}
-\CatchFileBetweenTags{\satpropexe}{text/definitions}{satpropexe}{\satpropexe}
+
+Let $tr\in \textsc{Transf}^{sr}_{tg}$ be a transformation.
+Let also $p = \langle V,E,st,\tau,\\Pre,Post\rangle \in \textsc{Property}(tr)$ be a property of $tr$ and $ex = \langle V',E',\\st',\tau',Input,Output\rangle \in Exec(tr)$ be an execution of $tr$.
+Execution $ex$ satisfies property $p$,  written $ex \models p$, if and only if:
+\begin{multline*}
+\forall f\; \exists g\,.\,\big(Pre \stackrel{f}{\vartriangleleft} Input^{*} \implies p \stackrel{g}{\vartriangleleft} ex^*\big)\\\text{where } V(Input)\cap CoDom(g) = CoDom(f)
+\end{multline*}
+% \begin{enumerate}
+% \item $\langle V_p,E_p\setminus Il_p,\tau_p\rangle$ is a typed subgraph of $\langle V_x, E_x,\tau_x\rangle$
+% \item if $v_p\rightarrow v_{p}'\in Il_p$ then there exists $v_x\rightarrow
+% v_{x}'\in E_{x}^{*}$ where $\tau(v_p)=\tau(v_x)$, $\tau(v_p')=\tau(v_x')$ and
+% $E_{x}^{*}$ is obtained by the transitive closure of $E_{x}$.
+% \end{enumerate}
 \end{definition}
@@
 \label{def:sat_prop_pc}
-\CatchFileBetweenTags{\satproppc}{text/definitions}{satproppc}{\satproppc}
+
+\reviewer{In Def. 31 I found it confusing that - different from previous
+    definitions - the two morphisms now go in the same direction.
+    There is an explanation after Definition 31, but I found it 
+    very hard to understand.}
+Let $tr\in \textsc{Transf}^{sr}_{tg}$ be a transformation.
+Let also $p = \langle V,E,st,\tau,\\Pre,Post\rangle \in \textsc{Property}(tr)$ be a property of $tr$ and $pc=\big\langle V',E',\\st',\tau', Match,Apply,Rulecop\big\rangle\in \textsc{Pathcond}(tr)$ be a path condition of $tr$.
+Path condition $pc$ satisfies property $p$, written $pc \vdash p$, if and only if:
+\begin{multline*}
+\forall f\; \exists g\,.\,\big(in\stackrel{f}{\blacktriangleleft} Pre \implies out \stackrel{g}\blacktriangleleft p\big)\\
+\text{where } in \sqsubseteq Match^{*} \;\land\; out \sqsubseteq pc^{*}
+\end{multline*}
+Additionally $Dom(g)\cap Match(pc^{*}) = Dom(f)$ and $f(v)\neq f(v')$, $g(v)\neq g(v')$ whenever $v$ and $v'$ are elements of the path condition belonging to the same rule copy of set $Rulecop$.
+
 \end{definition}

### 2016-06-29 text/abstraction_proofs.tex
Update path condition. Start working on abstraction relation.

@@
 In this section, we discuss the validity and completeness of the abstraction relation.
-Only proof sketches are presented here for \cref{prop:pc_validity} and \cref{lem:uniqueness}, while more complete proofs are found in \cref{sec:val_complete_path_cond_gen}, in the proofs of \cref{prop:pc_validity_appendix} and \cref{lem:uniqueness_appendix} respectively.
+
+This section presents our arguments that our path condition building
+algorithm is both \emph{valid} and \emph{complete}. In this context
+\emph{validity} means that for each path condition there exists at least one
+transformation execution that it abstracts.
+In other words, no path conditions are produced that lack a concrete
+transformation execution counterpart. \emph{Completeness} of the symbolic
+execution means that every transformation execution is abstracted by at least
+one path condition.
 

### 2016-06-29 text/abstraction_relation.tex
Update path condition. Start working on abstraction relation.

@@
 As this set is finite, our technique is guaranteed to be decidable.
-
-This section also presents our arguments that our path condition building
-algorithm is both \emph{valid} and \emph{complete}. In this context
-\emph{validity} means that for each path condition there exists at least one
-transformation execution that it abstracts.
-In other words, no path conditions are produced that lack a concrete
-transformation execution counterpart. \emph{Completeness} of the symbolic
-execution means that every transformation execution is abstracted by at least
-one path condition.
 
@@
 infinite set of transformation executions, as seen in
-\cref{sec:abstraction_relation}.
+\cref{subsec:abstraction_relation}.
 
@@
 
-As seen in the rest of this section, the structure of path conditions is similar to that of DSLTrans rules.
-The match graph of a path condition represents a pattern that must be present in the input model of the transformation, while the apply graph is a pattern which will be instantiated in the output model of the transformation.
-Symbolic traceability links are also kept between elements in the match and apply graphs to retain traceability information.
-
-The formal definition of a path condition is presented in \cref{def:path_condition}.
-
 \begin{definition}{Path Condition\\}
@@
 
-A path condition is a 7-tuple $\big\langle V,E,(s,t),\tau, Match,Apply,\\Rulecop\big\rangle$, where: $Match=\langle V',E',st',\tau'\rangle\in \textsc{Ipattern}^{sr}$ is an indirect pattern; $Apply=\langle V'',E'',st'',\tau''\rangle\in \textsc{Pattern}^{tg}$ is a pattern; $Match$ and $Apply$ are disjoint graphs.
-We also have that $V=V'\cup V''$, $E\subseteq E'\cup E''$ and $\tau\subseteq \tau'\cup \tau''$ where the co-domain of $\tau$ is the union of the co-domains of $\tau'$ and $\tau''$ and the set $\{trace\}$.
-An edge $e \in E\setminus E'\cup E''$, called a \emph{symbolic traceability link}, is such that $s(e)\in V''$ and $t(e)\in V'$ and $\tau(e)=trace$.
-Finally, the $Rulecop$ component in the 7-tuple contains the set of rule copies used in the construction of the path condition, where each rule copy is a subgraph of $\big\langle V,E,(s,t),\tau\big\rangle$.
-The set of all path conditions for a source metamodel $sr$ and a target metamodel $tg$ is called $\textsc{Pathcond}^{sr}_{tg}$ and the empty path condition is noted $\epsilon_{pc}$.
+A path condition is a three-tuple $\big\langle \mathit{Input}$, $\mathit{Output}$, $\mathit{trace}\big\rangle$, where:
+
+\begin{itemize}
+\item $\mathit{Input}, \mathit{Output} \in \textsc{TG}$
+\item $\mathit{Input}$ and $\mathit{Output} $ may be empty and are disjoint
+\item $\mathit{trace} = \{E_{trace}, (s_{trace}, t_{trace})\}$
+\begin{itemize}
+\item $E_{trace}$ contains the traceability links
+\item $E_{trace} $ is disjoint from $E_{Input}$ and $E_{Output}$
+\item $(s_{trace}, t_{trace})$ is a pair of functions $s_{trace}: E_{trace}\rightarrow V_{\textit{Output}}$ and $t_{trace}: E_{trace}\rightarrow V_{\textit{Input}}$ that respectively provide the source and target vertices for each traceability link
+\end{itemize}
+
+\end{itemize}
+
+The empty path condition is defined as $\{\emptyset, \emptyset, \emptyset \}$.
+
+The set of all path conditions is noted as $\textsc{PathConds}$.
+
+We also define a utility function getRules: PC $\rightarrow$ Set(Rules).
+This function accepts a path condition and returns the rules which were involved in its construction.
+This information will be used to report the status of contract proof.
+Note that operationally, this is done by storing the names of the rules within the path condition.
+
+We define a utility function getTransformation: PC $\rightarrow$ Transforms.
+This function returns the Transform that the path condition was built for.
+The purpose of this function is to restrict path condition to only be applicable for the transformation they represent.
 
@@
 
-\reviewer{Def. 18 suddenly introduces the notion of “rule copies” (Rulecop) which are re-
-ally not explained properly. My guess is that they are isomorphic representatives
-of rules. On page 13 under ”notation” this component is called Rule and referred
-to as a relation, and in Def. 28 it is denoted as Rules and required to be “the same
-as the set of transformation rules used to build [the transformation execution]”
-(and hence not isomorphic copies?). Needless to say, these inconsistencies do
-not help understandability.}
+The formal definition of a path condition is presented in \cref{def:path_condition}.
+Note that the structure of path conditions is intentionally very similar to that of DSLTrans rules and input-output models.
 
-Similarly to a transformation rule (see~\cref{def:transformation_rule}), a path condition is also a typed graph with a match and an apply part.
-As mentioned before, a path condition contains a combination of rules where \emph{symbolic traceability links} represent the concrete traceability links of a transformation execution (see \cref{def:modeltransformation}).
-The path condition structure also contains a $Rulcop$ set that allows identifying individually all the copies of rules that were used when building the path condition's typed graph.
-Note that we refer to \emph{copies} of rules as, despite the fact that a path condition normally only contains one copy of each rule, in certain situations a rule may be used multiple times in the construction of a path condition.
-This will be explained further ahead in this section.
-
-\paragraph{\textbf{Notation:}}
-Given a path condition $pc = \big\langle V,E,st,\tau,\\Match,Apply,Rulecop\big\rangle\in \textsc{Pathcond}^{sr}_{tg}$ we refer to the set of transformation rules in $pc$ identified by the $Rulecop$ relation as $Rulecop(pc)$.
-Also, because a path condition is a particular kind of a typed graph, we naturally extend the basic notation of operators and homomorphisms on typed graphs defined in \cref{sec:formal_background} to path conditions.
+The \textit{input} graph of a path condition represents a pattern that must be present in the input model of the transformation, while the \textit{output} graph is a pattern which will be instantiated in the output model of the transformation.
+Symbolic traceability links are also kept between elements in the \textit{input} and \textit{output} graphs to retain traceability information.
 
@@
 
+Our abstraction relation is at the core of our technique.
+It allows us to represent an infinite set of transformation executions with a finite set of path conditions.
+As mentioned before, contract proof can then take place on this set of path conditions and hold on the set of transformation executions, as further explored in Section~\ref{sec:prop_satisfaction}.
+
 Let us start by formally defining the notion of abstraction of a transformation
 execution by a path condition.
-
-\setcounter{equation}{0}
 
@@
 
-Let $tr\in \textsc{Transf}^{sr}_{tg}$ be a DSLTrans transformation. Let also $pc = \langle V,E,st,\tau,Match,Apply,Rules\rangle \in
-\textsc{Pathcond}(tr)$ of a path condition of $tr$ and $ex = \langle V',E',st',\tau',Input,Output\rangle \in Exec(tr)$ be an execution of $tr$.
-We have that $ex$ is abstracted by $pc$, noted $ex\Vvdash pc$, if and only if the set of transformation rules of $tr$ combined in $pc$ and the set of transformation rules of $tr$ used to build $ex$ is the same, and:
+Let $iom = \big\langle \mathit{Input}$, $\mathit{Output}$, $\mathit{trace}\big\rangle \in \textsc{IOM}$ be an input-output model, also known as a transformation execution.
+\bentley{Make sure this terminology is consistent.}
+
+Let also $pc = \big\langle \mathit{Input}$, $\mathit{Output}$, $\mathit{trace}\big\rangle \in \textsc{IOM}  \in
+\textsc{PathConds}$ be a path condition.
+
+We also require that getTransformation(iom) and getTransformation(pc) return the same transformation.
+
+We have that $iom$ is abstracted by $pc$, noted $ex\Vvdash pc$, if and only if:
+
 \begin{multline}
@@
   How do you express that these morphisms somehow "agree"?}
+
 \end{definition}
@@
 least a similar amount of times in the abstracted transformation execution.
-
-It is important to mention that another abstraction relation, weaker or
-stronger, could have been chosen. The abstraction relation presented in
-\cref{def:abstraction_pc_ex} suits our needs in the sense that it allows us to
-demonstrate the validity and completeness of our proof technique, as we will
-show in the text follows. Additionally, it is particularly interesting because
-it makes sure that, given a DSLTrans transformation, each of its transformation
-executions is abstracted by one and only one of its path conditions. This result
-adds to the consistency of our theory and is also exposed later in this section.
 

### 2016-06-30 text/abstraction_relation.tex
Revisit homomorphisms.

@@
 
-A path condition is a three-tuple $\big\langle \mathit{Input}$, $\mathit{Output}$, $\mathit{trace}\big\rangle$, where:
+A path condition is a four-tuple $\big\langle \mathit{NAC}$, $\mathit{Input}$, $\mathit{Output}$, $\mathit{trace}\big\rangle$, where:
 
 \begin{itemize}
-\item $\mathit{Input}, \mathit{Output} \in \textsc{TG}$
+\item $\mathit{NAC}, \mathit{Input}, \mathit{Output} \in \textsc{TG}$
 \item $\mathit{Input}$ and $\mathit{Output} $ may be empty and are disjoint
+\item $\mathit{NAC}$ may be an empty graph, and is disjoint from both $\mathit{Input}$ and $\mathit{Output} $
 \item $\mathit{trace} = \{E_{trace}, (s_{trace}, t_{trace})\}$
@@
 \item $E_{trace}$ contains the traceability links
-\item $E_{trace} $ is disjoint from $E_{Input}$ and $E_{Output}$
+\item $E_{trace} $ is disjoint from $E_{Input}$, $E_{Output}$, and $E_{NAC}$
 \item $(s_{trace}, t_{trace})$ is a pair of functions $s_{trace}: E_{trace}\rightarrow V_{\textit{Output}}$ and $t_{trace}: E_{trace}\rightarrow V_{\textit{Input}}$ that respectively provide the source and target vertices for each traceability link
@@
 
-The empty path condition is defined as $\{\emptyset, \emptyset, \emptyset \}$.
+The empty path condition is defined as $\{\emptyset, \emptyset, \emptyset, \emptyset \}$.
 
@@
 
-Let $iom = \big\langle \mathit{Input}$, $\mathit{Output}$, $\mathit{trace}\big\rangle \in \textsc{IOM}$ be an input-output model, also known as a transformation execution.
+Let $iom = \big\langle \mathit{NAC}, \mathit{Input}$, $\mathit{Output}$, $\mathit{trace}\big\rangle \in \textsc{IOM}$ be an input-output model, also known as a transformation execution.
 \bentley{Make sure this terminology is consistent.}
@@
 We have that $iom$ is abstracted by $pc$, noted $ex\Vvdash pc$, if and only if:
+
+NAC doesn't edge match iom's Input
+Input edge matches iom's Input
 

### 2016-06-30 text/formal_background.tex
Update path condition. Start working on abstraction relation.
Revisit homomorphisms.

@@
 notion of left hand side and right hand side of a rule. The left hand side of a
-rule rl is precisely your ||rl||, the right hand side is the entire rl augmented by
+rule rl is precisely your rule matcher, the right hand side is the entire rl augmented by
 your traceability links, as in 4.2.2. Suddenly we are back in a well-known graph
@@
 
-\subsection{Graph Structures and Functions}
-\label{sec:typed_graphs}
+\subsection{Typed Graphs}
+\label{subsec:typed_graphs}
 
@@
 Sample metamodels for our running example can be seen in figures above \bentley{Fix.}.
-The particular metamodel for a typed graph will be indicated in the formalization when required, as (potentially) different metamodels will be present for the input and output graphs in DSLTrans rules.
+Note that DSLTrans rules (to be formally defined) will combine typed graphs with potentially different metamodels.
 
@@
 \begin{itemize}
-\item $\mathit{isAbstract}: \mathit{VT} \rightarrow \{\mathit{true}, \mathit{false}\}$
+\item $\mathit{isAbstract}: \mathit{VT} \rightarrow \{\mathit{true}$, $\mathit{false}\}$
 \begin{itemize}
@@
 
-\item $\mathit{isIndirect}: \mathit{ET} \rightarrow \{\mathit{true}, \mathit{false}\}$
+\item $\mathit{isIndirect}: \mathit{ET} \rightarrow \{\mathit{true}$, $\mathit{false}\}$
 \begin{itemize}
@@
 
-\item $\mathit{matchesOver}: \{\VT \cup \ET\} \times \{\VT \cup \ET\} \rightarrow \{\mathit{true}, \mathit{false}\}$
+\item $\mathit{matchesOver}: \{\VT \cup \ET\} \times \{\VT \cup \ET\} \rightarrow \{\mathit{true}$, $\mathit{false}\}$
 
@@
 \subsubsection*{Typed Subgraph}
-We now define the useful notion of typed subgraph. As expected, a typed subgraph is simply a restriction of a typed graph to some of its vertices and edges.
+We now define the useful notion of typed subgraph. As expected, a typed subgraph is simply a restriction of a typed graph to some of its vertices and edges. \bentley{Is this needed?}
 
@@
 
-Given a typed graph $g\in \textsc{TG}$, will use the notation $Components(g)$ to describe the set of strongly connected typed graphs in $g$.
+%Given a typed graph $g\in \textsc{TG}$, will use the notation $Components(g)$ to describe the set of strongly connected typed graphs in $g$.
 Also, we will use the notation $g|_{t}$ to refer to the restriction of graph $g$ to its subgraph containing only edges of type $t$.
@@
 
+\subsection{Homomorphisms}
+\label{subsec:typed_graphs}
+
+For the formal development of our technique, we are interested in relations between typed graphs that preserve some graph structure and vertex/edge type, i.e. homomorphisms.
+
+However, our technique of symbolic execution of transformation rules means that we cannot rely on a trivial homomorphism. Instead we must... \bentley{TODO: Finish this sentence.}
+
+Matching:
+
+\begin{itemize}
+
+\item Rules onto rules?
+\begin{itemize}
+\item If so, is isomorphic
+\end{itemize}
+
+\item Contracts onto path conditions
+\begin{itemize}
+\item Edges: Each edge of contract must exist injectively in PC
+\item Vertices: Each vertex match must be surjective from target to source. Note that this will be a partial function. This should also be a minimal function.
+\end{itemize}
+
+\item Rules onto path conditions
+
+\item Path conditions onto transformation executions
+
+\item Do we want to add slicing?
+\end{itemize}
+
 \subsubsection*{Typed Graph Homomorphism}
-For the formal development of our technique, we are interested in relations between typed graphs that are structure-preserving, i.e.
-homomorphisms.
-Homomorphisms between typed graphs preserve not only structure, but also the types of vertices and edges that are mapped.
-Note that, trivially, a typed graph homomorphism is a graph homomorphism.
+
+The first typed graph homomorphism we define is the trivial one. This is presented in order to
 
@@
 
-A typed graph homomorphism between $g$ and $g'$ is a function $f: f_v \cup f_e$ such that:
+A typed graph homomorphism from $g$ onto $g'$ is a function $f: f_v \cup f_e$ such that:
 \begin{itemize}
@@
 
-\bentley{Maybe put `inj' and `sub' in the homomorphism to reduce number of symbols}
- When an \emph{injective} typed graph homomorphism $f$ exists between $g$ and $g'$ we write $g \stackrel{f}{\vartriangleleft} g'$, or simply $g \vartriangleleft g'$ when the context is unambiguous.
-When a \emph{surjective} typed graph homomorphism $f$ exists between typed graphs $g$ and $g'$ we write $g \stackrel{f}{\blacktriangleleft} g'$, or also simply $g \blacktriangleleft g'$ in an unambiguous context.
+ When an \emph{injective} typed graph homomorphism $f$ exists from $g$ onto $g'$, we write $g \stackrel{inj}{\blacktriangleleft} g'$.
+When a \emph{surjective} typed graph homomorphism $f$ exists from $g$ onto $g'$ we write $g \stackrel{surj}{\blacktriangleleft} g'$.
 
@@
 Constructing this morphism allows our prover implementation to avoid explicitly creating `disambiguated' graphs where these duplications are resolved.
+
+ When an \emph{injective} typed graph edge homomorphism $f$ exists from $g$ onto $g'$, we write $g \stackrel{inj}{\vartriangleleft} g'$.
+When a \emph{surjective} typed graph edge homomorphism $f$ exists from $g$ onto $g'$ we write $g \stackrel{surj}{\vartriangleleft} g'$.
 
@@
 Let $\mathit{IOM}$ be the set of all input-output models.
+
+We define a utility function getTransformation: IOM $\rightarrow$ Transforms.
+This function returns the Transform that the input-output model was built for.
+The purpose of this function is to restrict input-output models to only be applicable for the transformation they represent.
 

### 2016-06-30 text/formal_background.tex
Add about unification of matching match/apply graphs and backward/trace links.

@@
 \item Vertices: Each vertex match must be surjective from target to source. Note that this will be a partial function. This should also be a minimal function.
+\item Backward links injectively match from contract into PC onto traceability links.
+Source and target of backward links on contract must match the source and target of the traceability link of the PC.
 \end{itemize}
@@
 \item Rules onto path conditions
+\begin{itemize}
+\item Same as above
+\end{itemize}
 
 \item Path conditions onto transformation executions
+\begin{itemize}
+\item Each component in graph:
+\begin{itemize}
+\item As in paper. Match component is injectively found. Apply component is surjectively found.
+\item For traceability links, PC links are found injectively.
+Transformation executions links are found surjectively.
+Not a bijection because multiple rule executions could be present in transformation execution.
+
+\item Question: How to handle unification of matching. Seems to be straightforward before/after must match.
+\end{itemize}
+\end{itemize}
 
 \item Do we want to add slicing?
+
+\item Explain backward links and trace links matching over trace links
 \end{itemize}

### 2016-07-04 text/formal_background.tex
Add more on homomorphisms.

--- 2016-06-30 text/formal_background.tex
@@
 \subsection{Homomorphisms}
-\label{subsec:typed_graphs}
+\label{subsec:homomorphisms}
 
@@
 
-However, our technique of symbolic execution of transformation rules means that we cannot rely on a trivial homomorphism. Instead we must... \bentley{TODO: Finish this sentence.}
-
-Matching:
-
-\begin{itemize}
-
-\item Rules onto rules?
-\begin{itemize}
-\item If so, is isomorphic
-\end{itemize}
-
-\item Contracts onto path conditions
-\begin{itemize}
-\item Edges: Each edge of contract must exist injectively in PC
-\item Vertices: Each vertex match must be surjective from target to source. Note that this will be a partial function. This should also be a minimal function.
-\end{itemize}
-
-\item Rules onto path conditions
-
-\item Path conditions onto transformation executions
-
-\item Do we want to add slicing?
-\end{itemize}
+However, our technique of symbolic execution of transformation rules means that we cannot rely on a trivial homomorphism.
+Instead our matching must be flexible about how vertices in the source graph may match over multiple vertices in the target graph.
 
@@
 
-The first typed graph homomorphism we define is the trivial one. This is presented in order to
+The first typed graph homomorphism we define is standard, where the structure of the source graph must be found in the target graph.
+This is presented for the reader to appreciate the complications required for the next homomorphisms.
 
@@
 Our technique also requires a form of graph homomorphism that primarily focuses on matching the edges of the graphs.
-This homomorphism will be an injective match regarding the edges, but note that a particular vertex in the source graph may match onto multiple vertices in the target graph.
+This homomorphism will be an injective\footnote{For a terminology refresh, \textit{injective matching} means there is a one-to-one correspondence between the source and target vertices.
+\textit{Surjective matching} means that all target vertices must be matched by at least one source vertex.
+}match regarding the edges, but note that a particular vertex in the source graph may match onto multiple vertices in the target graph.
 \bentley{What's the term for this?}
@@
 \begin{itemize}
-\item $h_v: V'\rightarrow V$
+\item $h_v: V'\rightarrow V$ Note that this function has unusual properties:
 \begin{itemize}
-\item Note that $h_v$ is an function from a vertex in the target graph onto a vertex in the source graph
+
+\item A function from a vertex in the target graph onto a vertex in the source graph
+\item A partial function, as not all vertices in the target graph will be reflected in the source graph
+\item Surjective, so that all source vertices are matched to target vertices
+\item A minimal function, such that the minimum number of target  vertices are matched to each source vertex \bentley{Not sure if this is needed.}
 \end{itemize}
-\item $h_e: E\rightarrow E'$
+\item $h_e: E\rightarrow E'$, and is injective
 \item $\forall v_1 \xrightarrow{e} v_2\in E$:

### 2016-07-05 text/abstraction_relation.tex
Add more on homomorphisms.
Update with some work-in-progress for abstraction relation.

@@
 
+%\item Path conditions onto transformation executions
+%\begin{itemize}
+%\item Each rule in graph:
+%\begin{itemize}
+%\item As in paper. Match component is injectively found. Apply component is surjectively found.
+%\item For traceability links, PC links are found injectively.
+Transformation executions links are found surjectively.
+Not a bijection because multiple rule executions could be present in transformation execution.
+%
+%\item Question: How to handle unification of matching. Seems to be straightforward before/after must match.
+%\end{itemize}
+%\end{itemize}
+
 The formal definition of a path condition is presented in \cref{def:path_condition}.
@@
 Symbolic traceability links are also kept between elements in the \textit{input} and \textit{output} graphs to retain traceability information.
+
+Finally, we define a function getComponents: PC $\rightarrow$ Set(Typed Graph).
+This function is used for the abstraction relation, as it returns a set of subsets from the match graph of the path condition.
+These components are collections of the match graph of rules that have been overlaid on top of each other.
+These components must be matched as a unit, as their elements must be injectively matched.
+However, elements from different components may match onto the same target vertex in the abstraction relation.
+More information on how the components are constructed will be provided in the section on path condition construction.
+\bentley{This is not elegant at all...}
 
@@
 
+LOOK AT COMPONENTS IN PC MATCH GRAPH AS ELEMENTS AND TRACE LINKS
+
+THEN LOOK AT OUTPUT ELEMENTS AND TRACE LINKS IN IOM
+
+REDO STRUCTURES TO ONLY BE ONE VERTEX AND EDGE SET, AND HAVE SUBSETS FOR MATCH, APPLY
+
 \begin{definition}{Abstraction of a Transformation Execution by a Path Condition\\}
@@
 
-Let $iom = \big\langle \mathit{NAC}, \mathit{Input}$, $\mathit{Output}$, $\mathit{trace}\big\rangle \in \textsc{IOM}$ be an input-output model, also known as a transformation execution.
+Let $iom = \big\langle \mathit{Input}$, $\mathit{Output}$, $\mathit{trace}\big\rangle \in \textsc{IOM}$ be an input-output model, also known as a transformation execution.
 \bentley{Make sure this terminology is consistent.}
 
-Let also $pc = \big\langle \mathit{Input}$, $\mathit{Output}$, $\mathit{trace}\big\rangle \in \textsc{IOM}  \in
-\textsc{PathConds}$ be a path condition.
+Let also $pc = \big\langle  \mathit{NAC}$, $\mathit{Input}$, $\mathit{Output}$, $\mathit{trace}\big\rangle  \in \textsc{PathConds}$ be a path condition.
 
@@
 
-We have that $iom$ is abstracted by $pc$, noted $ex\Vvdash pc$, if and only if:
+We have that $iom$ is abstracted by $pc$, noted $ex\Vvdash pc$, if and only if the following conditions are all true:
 
-NAC doesn't edge match iom's Input
-Input edge matches iom's Input
+\noindent\makebox[\linewidth]{\rule{\linewidth}{0.4pt}}
+The PC's NAC does not match onto the iom's Input
+
+\begin{multline}
+\label{eq:abstr_no_nac}
+\nexists \big(\mathit{PC}_{NAC} \stackrel{inj}{\vartriangleleft} \mathit{iom}_{Input}^*\big)
+\end{multline}
+
+\noindent\makebox[\linewidth]{\rule{\linewidth}{0.4pt}}
+
+Each component in the PC:
+Elements matches on the iom's Input
+Traceability links can be found as well
 
@@
 \label{eq:abstr_input_output}
-\big(\forall rl\in Rules\;.\;Match(rl) \vartriangleleft Input^*\big) \;\land\; Output \blacktriangleleft Apply
+\big(\forall \mathit{comp} \in \mathit{getComponents(PC)} \vartriangleleft Input^* : \mathit{comp} \stackrel{inj}{\vartriangleleft} \mathit{iom}_{Input}^*\big)
 \end{multline}
-\center{and}
+
+\noindent\makebox[\linewidth]{\rule{\linewidth}{0.4pt}}
+
+Elements in the iom's Output graph must surjectively match on the Output graph of the PC
+Trace links in iom can be found in PC
+
 \begin{multline}
-\label{eq:abstr_trace}
-\big(\forall trc\in Components(pc|_{trace})\;.\;trc\vartriangleleft ex\big)\;\land\;\\
-\big(\forall e'\in E'\, \exists e\in E\;.\;\tau'(e')=trace \Longrightarrow \tau(e)=trace\big)
+\label{eq:abstr_apply_surjective_match}
+\mathit{iom}_{Output} \stackrel{surj}{\vartriangleleft} \mathit{PC}_{Output}
 \end{multline}
+
+\noindent\makebox[\linewidth]{\rule{\linewidth}{0.4pt}}
+
+%\begin{multline}
+%\label{eq:abstr_trace}
+%
+%\forall \mathit{PCTrace} \in \mathit{PC}_{trace} :
+%
+%\exists \mathit{IOMTrace} | \mathit{IOMTrace}
+%
+%\big(\forall trc\in Components(pc|_{trace})\;.\;trc\vartriangleleft ex\big)\;\land\;\\
+%\big(\forall e'\in E'\, \exists e\in E\;.\;\tau'(e')=trace \Longrightarrow \tau(e)=trace\big)
+%\end{multline} 
 % where for all $rl,rl'\in Rules$, if $rl\cong rl'$ then there exist two injective typed graph homomorphisms $f$ and $g$ such that $Match(rl)\stackrel{f}{\vartriangleleft} Input^*$, $Match(rl)\stackrel{g}{\vartriangleleft} Input^*$ and $CoDom(f)\neq CoDom(g)$.

### 2016-07-05 text/formal_background.tex
Update with some work-in-progress for abstraction relation.

@@
 
-Note that the our verification technique is geared toward model-driven engineering. Therefore, the types of vertices and edges in our typed graphs will be drawn from a particular \textit{metamodel}.
-
+Note that the our verification technique is geared toward model-driven engineering.
+Therefore, the types of vertices and edges in our typed graphs will be drawn from a particular \textit{metamodel}.
 Sample metamodels for our running example can be seen in figures above \bentley{Fix.}.
@@
 
-We assume the existence of a number of functions in our formalization to aid the treatment of typing:
+We assume that this metamodel provides the necessary vertex and edge types, with any collisions resolved by adding the metamodel name.
+As well, we assume the presence of a partial ordering $\leq$ on the vertex and edge types for subtyping information.
+
+We require specially typed edges for some DSLTrans constructs.
+Thus, we include the edge types \{\textit{trace}, and \textit{backward}\} in ET, resolving collisions as needed.
+These edge types will be explained as needed.
+We also insert the ordering that \textit{trace} is a subtype of \textit{backward}.
+
+For convenience, we define some utility functions in our formalization to aid the treatment of typing:
 
@@
 \begin{itemize}
-\item The purpose of this function is to assist in handling polymorphism in the metamodel
+\item The purpose of this function is to assist in handling polymorphism in the metamodel, and to handle matching for the special DSLTrans types
 \item $\mathit{matchesOver}(T, T) \rightarrow \mathit{true}$
@@
 \subsubsection*{Typed Subgraph}
-We now define the useful notion of typed subgraph. As expected, a typed subgraph is simply a restriction of a typed graph to some of its vertices and edges. \bentley{Is this needed?}
+We now define the useful notion of typed subgraph.
+As expected, a typed subgraph is simply a restriction of a typed graph to some of its vertices and edges.
+This will be needed to partition DSLTrans constructs such as rules into logical components.
 
@@
 This homomorphism will be an injective\footnote{For a terminology refresh, \textit{injective matching} means there is a one-to-one correspondence between the source and target vertices.
-\textit{Surjective matching} means that all target vertices must be matched by at least one source vertex.
-}match regarding the edges, but note that a particular vertex in the source graph may match onto multiple vertices in the target graph.
+\textit{Surjective matching} means that all target vertices must be matched by at least one source vertex.} match regarding the edges, but note that a particular vertex in the source graph may match onto multiple vertices in the target graph.
 \bentley{What's the term for this?}
@@
 Given a graph, its transitive closure includes, besides the original graph, all the edges belonging to the transitive closure of indirect links in that graph.
-Note that these transitive edges are given by the function $\mathit{isIndirect}$.
+Note that these transitive edges are identified by the function $\mathit{isIndirect}$.
 
@@
 \end{itemize}
+
+\subsubsection{Link Homomorphism}
+
+\begin{itemize}
+\item Backward links injectively match from contract into PC onto traceability links.
+Source and target of backward links on contract must match the source and target of the traceability link of the PC.
+\end{itemize}
+
+We must also define a function to match the traceability and backward link structures.
+
+\begin{definition}{Link Homomorphism\\}
+\label{def:link_homomorphism}
+Consider a set of links $L = \{E_{L}, (s_{L}, t_{L})\}$. Note that these could be backward links or traceability links.
+
+We define a function $f$ to match $L$ over $T$, such that:
+
+\begin{itemize}
+\item $\forall \mathit{sourceLink} \in L:$
+\begin{itemize}
+\item $\exists \mathit{targetLink} \in T | f(s(\mathit{sourceLink})) = s(targetLink), $
+\end{itemize}
+\end{itemize} 
+\end{definition}
 

### 2016-07-05 text/abstraction_relation.tex
Sketching of abstraction relation formalities.

@@
 
-We also define a utility function getRules: PC $\rightarrow$ Set(Rules).
-This function accepts a path condition and returns the rules which were involved in its construction.
+We also define a utility function getRule: V $\rightarrow$ Rules.
+This function accepts a vertex from the path condition and returns the rule which created that vertex.
+This function will be used during the matching procedure of the abstraction relation, as elements from the same rule cannot match over the same target element.
+Note that this function allows production of the set of rules which are represented by the path condition.
 This information will be used to report the status of contract proof.
-Note that operationally, this is done by storing the names of the rules within the path condition.
 
@@
 
-Finally, we define a function getComponents: PC $\rightarrow$ Set(Typed Graph).
+%Finally, we define a function getComponents: PC $\rightarrow$ Set(Typed Graph).
 This function is used for the abstraction relation, as it returns a set of subsets from the match graph of the path condition.
@@
 
-Let us start by formally defining the notion of abstraction of a transformation
-execution by a path condition.
-
-LOOK AT COMPONENTS IN PC MATCH GRAPH AS ELEMENTS AND TRACE LINKS
-
-THEN LOOK AT OUTPUT ELEMENTS AND TRACE LINKS IN IOM
-
-REDO STRUCTURES TO ONLY BE ONE VERTEX AND EDGE SET, AND HAVE SUBSETS FOR MATCH, APPLY
+\subsubsection{Formal Definition}
+\label{subsubsec:abstraction_relation_formal}
+Let us start by formally defining the notion of abstraction of a transformation execution by a path condition.
 
@@
 
-\noindent\makebox[\linewidth]{\rule{\linewidth}{0.4pt}}
+\end{definition}
+
+\paragraph{Condition 1:}
 The PC's NAC does not match onto the iom's Input
 
-\begin{multline}
+\begin{equation}
 \label{eq:abstr_no_nac}
 \nexists \big(\mathit{PC}_{NAC} \stackrel{inj}{\vartriangleleft} \mathit{iom}_{Input}^*\big)
-\end{multline}
+\end{equation}
 
-\noindent\makebox[\linewidth]{\rule{\linewidth}{0.4pt}}
+\paragraph{Condition 2:}
 
-Each component in the PC:
-Elements matches on the iom's Input
-Traceability links can be found as well
+The PC's Input matches onto the iom's Input. However, elements from the same rule in the PC cannot match over the same iom element. Note that this match function \textit{f} is not injective.
 
-\begin{multline}
+\begin{equation}
 \label{eq:abstr_input_output}
-\big(\forall \mathit{comp} \in \mathit{getComponents(PC)} \vartriangleleft Input^* : \mathit{comp} \stackrel{inj}{\vartriangleleft} \mathit{iom}_{Input}^*\big)
-\end{multline}
+\exists \big(\mathit{PC}_{Input} \vartriangleleft \mathit{iom}_{Input}^*
+where \forall v_1, v_2 \in V_{PC-Input} :
+f(v_1) = f(v_2) \veebar getRule(v_1) = getRule(v_2) 
+\big)
+\end{equation}
 
-\noindent\makebox[\linewidth]{\rule{\linewidth}{0.4pt}}
+\paragraph{Condition 3:}
 
 Elements in the iom's Output graph must surjectively match on the Output graph of the PC
-Trace links in iom can be found in PC
 
-\begin{multline}
+\begin{equation}
 \label{eq:abstr_apply_surjective_match}
 \mathit{iom}_{Output} \stackrel{surj}{\vartriangleleft} \mathit{PC}_{Output}
-\end{multline}
+\end{equation}
 
-\noindent\makebox[\linewidth]{\rule{\linewidth}{0.4pt}}
+\paragraph{Condition 4:}
 
-%\begin{multline}
-%\label{eq:abstr_trace}
-%
-%\forall \mathit{PCTrace} \in \mathit{PC}_{trace} :
-%
-%\exists \mathit{IOMTrace} | \mathit{IOMTrace}
-%
-%\big(\forall trc\in Components(pc|_{trace})\;.\;trc\vartriangleleft ex\big)\;\land\;\\
-%\big(\forall e'\in E'\, \exists e\in E\;.\;\tau'(e')=trace \Longrightarrow \tau(e)=trace\big)
-%\end{multline} 
-% where for all $rl,rl'\in Rules$, if $rl\cong rl'$ then there exist two injective typed graph homomorphisms $f$ and $g$ such that $Match(rl)\stackrel{f}{\vartriangleleft} Input^*$, $Match(rl)\stackrel{g}{\vartriangleleft} Input^*$ and $CoDom(f)\neq CoDom(g)$.
-% $/pc/ = Apply_{ex}\cup ex|_{trace}$ is the output part of $pc$, including symbolic traceability links.
+Traceability links in the PC must injectively match onto links in the transformation execution.
+This is to ensure that element creation represented in the path condition reflects actual rule execution present in the transformation execution.
 
-\reviewer{Definition 28: When you define the abstraction of an execution
-  by a path condition, you require two conditions (separated by an
-  and). I have the feeling that the morphisms used in both
-  conditions should somehow be compatible to each other. However,
-  your triangle notation indicates only the existence of a morphisms.
-  How do you express that these morphisms somehow "agree"?}
+TODO: Define the InjMatch we need here
 
-\end{definition}
+\begin{equation}
+\label{eq:abstr_links_injective}
+\mathit{PC}_{trace} InjMatch \mathit{iom}_{trace}
+\end{equation}
+
+\paragraph{Condition 5:}
+
+Traceability links in the transformation execution must surjectively match onto links in the transformation execution.
+This case is to ensure that there is no element creation in the transformation execution that is not captured by the path condition.
+
+TODO: Define the SurjMatch we need here
+
+\begin{equation}
+\label{eq:abstr_links_surjective}
+\mathit{PC}_{trace} SurjMatch \mathit{iom}_{trace}
+\end{equation}
+
+\paragraph{Condition 6:}
+
+The morphisms used in the above conditions must match.
+
+TODO: Define how the morphisms agree.
+
+\begin{equation}
+\label{eq:abstr_morphisms_agree}
+\mathit{PC}_{trace}
+\end{equation}
 
@@
 
-To understand the abstraction relation in \cref{def:abstraction_pc_ex}, recall
+\subsubsection{Explanation}
+
+To understand the abstraction relation in \cref{subsubsec:abstraction_relation_formal}, recall
 that during the construction of a transformation execution rules are matched

### 2016-07-05 text/formal_background.tex
Sketching of abstraction relation formalities.

@@
 \item The set of all typed graphs is called $\textsc{TG}$
-\item We define the empty graph to be a graph with $V = \emptyset$, $E = \emptyset$, and the other elements to be empty functions
+\item We define the empty graph to be a graph with all elements to be empty functions or sets
 \end{itemize}
@@
 
+We also define a function getEdges: \textit{ET} $\rightarrow$ Set(Edges) to return all edges of a certain edge type.
+
 \subsubsection*{Typed Graph Union}
@@
 %, where $\tau(v_1)=\tau'(f(v_1))$, $\tau(v_2)=\tau'(f(v_2))$ and also $\tau(e)=\tau(e')$.
-\item The domain of $f$ is noted $Dom(f)$ and the co-domain of $f$ is noted $CoDom(f)$.
+%\item The domain of $f$ is noted $Dom(f)$ and the co-domain of $f$ is noted $CoDom(f)$.
 \end{itemize}  
@@
 
- When an \emph{injective} typed graph homomorphism $f$ exists from $g$ onto $g'$, we write $g \stackrel{inj}{\blacktriangleleft} g'$.
+ When an \emph{injective}\footnote{For a terminology refresh, \textit{injective matching} means there is a one-to-one correspondence between the source and target vertices.
+\textit{Surjective matching} means that all target vertices must be matched by at least one source vertex.} typed graph homomorphism $f$ exists from $g$ onto $g'$, we write $g \stackrel{inj}{\blacktriangleleft} g'$.
 When a \emph{surjective} typed graph homomorphism $f$ exists from $g$ onto $g'$ we write $g \stackrel{surj}{\blacktriangleleft} g'$.
@@
 Our technique also requires a form of graph homomorphism that primarily focuses on matching the edges of the graphs.
-This homomorphism will be an injective\footnote{For a terminology refresh, \textit{injective matching} means there is a one-to-one correspondence between the source and target vertices.
-\textit{Surjective matching} means that all target vertices must be matched by at least one source vertex.} match regarding the edges, but note that a particular vertex in the source graph may match onto multiple vertices in the target graph.
+This homomorphism will be an injective match regarding the edges, but note that a particular vertex in the source graph may match onto multiple vertices in the target graph.
 \bentley{What's the term for this?}

### 2016-07-05 text/abstraction_relation.tex
Integrated abstraction relation explanation into the different conditions. Haven't updated the examples yet.

@@
 
-%\item Path conditions onto transformation executions
-%\begin{itemize}
-%\item Each rule in graph:
-%\begin{itemize}
-%\item As in paper. Match component is injectively found. Apply component is surjectively found.
-%\item For traceability links, PC links are found injectively.
-Transformation executions links are found surjectively.
-Not a bijection because multiple rule executions could be present in transformation execution.
-%
-%\item Question: How to handle unification of matching. Seems to be straightforward before/after must match.
-%\end{itemize}
-%\end{itemize}
-
 The formal definition of a path condition is presented in \cref{def:path_condition}.
@@
 
+The purpose of this relation is to enforce that the path condition accurately represents the elements present in the input model of the transformation execution.
+Note that extra elements may be present but not matched.
+
 \begin{equation}
@@
 
-Elements in the iom's Output graph must surjectively match on the Output graph of the PC
+Elements in the iom's Output graph must surjectively match on the Output graph of the PC.
+
+This surjection relation enforces that all elements which have been created in the output of the transformation execution must be represented by the output of the path condition.
+Note that this matching is from the execution to the path condition.
+As there may be multiple copies of an element in the execution but the path condition only represents execution of rules once, these elements should match over the same element in the path condition.
 
@@
 
-Traceability links in the PC must injectively match onto links in the transformation execution.
-This is to ensure that element creation represented in the path condition reflects actual rule execution present in the transformation execution.
+Traceability links in the PC must injectively match onto links in the transformation execution using the match function \textit{f}.
+The purpose of this condition is to ensure that element creation represented in the path condition reflects actual rule execution present in the transformation execution.
+
+Note that again we cannot have two traceability links from the same rule in the path condition matching over a traceability link in the transformation execution.
+Instead, both traceability links must be uniquely found.
 
@@
 \end{equation}
-
-% \cref{def:abstraction_pc_ex} enforces the fact that the matcher part of
-% components of the path condition are found through an injective typed graph
-% homomorphism in the execution's input model. Additionally, the graph including
-% the output of the execution (including traceability information), can be
-% mapped by a surjective typed graph homomorphism onto the apply model (plus the
-% symbolic traceability links) of the path condition.
-
-\subsubsection{Explanation}
-
-To understand the abstraction relation in \cref{subsubsec:abstraction_relation_formal}, recall
-that during the construction of a transformation execution rules are matched
-injectively in the input model. This information is present in the first
-condition of the abstraction relation (Proposition~\ref{eq:abstr_input_output})
-via the injective typed graph homomorphism between the match part of the copies
-of rules ``glued'' onto the path condition and the containment transitive
-closure of the input part of the transformation execution. This relation
-enforces the fact that certain parts of the execution were found, or matched, by
-certain parts of the path condition.
-On the other hand, the surjection from the output of the execution towards the
-apply part of the path condition models the fact that the output of the
-execution has been completely built by instantiating the apply parts of the
-rules contained in the path condition.
-
-The second condition of the abstraction relation
-(Proposition~\ref{eq:abstr_trace}) checks for the fact that symbolic
-traceability links in the path condition and traceability links in the execution
-correctly correspond to each other. This is modeled by the fact that all
-strongly connected components in the path condition, composed only of symbolic
-traceability links, are injectively found on the execution. This injection
-models the fact that traceability graphs between individual or combined rules in
-the path condition are necessarily found in the execution. Note that components
-of the path condition are considered because of the fact that disconnected rules
-in the path condition may have matched over common elements of a particular
-execution. As such, a full injection between the complete traceability graph in
-the path condition and the execution would be incorrect. Additionally, in the
-second part of Proposition~\ref{eq:abstr_trace} we check the fact that every
-traceability link in the execution can be found in the path condition. This
-additional sanity check enforces that no spurious traceability links that could
-not have been created by the rules present in the path condition exist in the
-transformation execution.
-
-Finally, the last clause of the abstraction relation states that rule copies
-that are repeated a number of times in the path condition need to be found at
-least a similar amount of times in the abstracted transformation execution.
 

### 2016-07-06 text/building_path_conditions.tex
Updates to building PCs.

@@
 
-As well, we present our approach to building a set of path conditions to represent all executions of a DSLTrans transformation.
+In this section we present our approach to building a set of path conditions to represent all executions of a DSLTrans transformation.
 
@@
 \label{sec:gen_all_pcs}
-
- This section will describe how path conditions are constructed for a DSLTrans transformation using our approach.
 
@@
 A new set of path conditions will then be created by sequentially combining each rule in the layer with the path condition selected.
-Recall that a path condition represents a set of rules that have symbolically executed, thereby abstracting a set of transformation executions through our abstraction relation.
-\reviewer{we cannot recall this as it has not yet been presented;
-the abstraction relation is defined only in Section 5}Combining a path condition with a rule will produce one or more path conditions depending on how the rule combines with the rules already represented by the path condition.
+Recall from \cref{subsec:gen_path_conds} that a path condition represents a set of rules that have symbolically executed, thereby abstracting a set of transformation executions through our abstraction relation.
+Combining a path condition with a rule will produce one or more path conditions depending on how the rule combines with the rules already represented by the path condition.
 The pre- and post- conditions defined by the path condition will be modified according to the elements found in that rule.

### 2016-07-25 text/formal_background.tex
Do revisions from Friday.

@@
 
-We assume that this metamodel provides the necessary vertex and edge types, with any collisions resolved by adding the metamodel name.
+We assume that this metamodel provides the necessary vertex and edge types, with any collisions resolved by adding to the metamodel name.
 As well, we assume the presence of a partial ordering $\leq$ on the vertex and edge types for subtyping information.
-
-We require specially typed edges for some DSLTrans constructs.
-Thus, we include the edge types \{\textit{trace}, and \textit{backward}\} in ET, resolving collisions as needed.
-These edge types will be explained as needed.
-We also insert the ordering that \textit{trace} is a subtype of \textit{backward}.
 
@@
 
-For simplification purposes, we will not represent edge cardinalities or containment relationships given by a metamodel in our notion of typed graph.
-Note that in fact we require these conditions to be relaxed to perform our graph rewriting.
+This typing information is our implementation of a typed graph conforming to a metamodel.
+Note that for simplification purposes, we will not represent edge cardinalities or containment relationships given by a metamodel in our notion of typed graph.
+In fact we require these conditions to be relaxed to perform our graph rewriting.
 
@@
 
+\bentley{Where is this used?}
+
 \subsubsection*{Typed Graph Union}
@@
 \bentley{Do we need typed graph intersection? Or to mention what it means for two graphs to be disjoint?}
+\bentley{Is this used in the DPO approach?}
 
@@
 However, our technique of symbolic execution of transformation rules means that we cannot rely on a trivial homomorphism.
-Instead our matching must be flexible about how vertices in the source graph may match over multiple vertices in the target graph.
+Instead our matching must be flexible about how vertices in the pattern graph may match over multiple vertices in the target graph.
 
@@
 
-The first typed graph homomorphism we define is standard, where the structure of the source graph must be found in the target graph.
+The first typed graph homomorphism we define is standard, where the structure of the pattern graph must be found in the target graph.
 This is presented for the reader to appreciate the complications required for the next homomorphisms.
@@
 
- When an \emph{injective}\footnote{For a terminology refresh, \textit{injective matching} means there is a one-to-one correspondence between the source and target vertices.
-\textit{Surjective matching} means that all target vertices must be matched by at least one source vertex.} typed graph homomorphism $f$ exists from $g$ onto $g'$, we write $g \stackrel{inj}{\blacktriangleleft} g'$.
-When a \emph{surjective} typed graph homomorphism $f$ exists from $g$ onto $g'$ we write $g \stackrel{surj}{\blacktriangleleft} g'$.
+ When an \emph{injective}\footnote{For a terminology refresh, \textit{injective matching} means there is a one-to-one correspondence between the pattern and target vertices.
+\textit{Surjective matching} means that all target vertices must be matched by at least one pattern vertex.
+A \textit{bijection} is when the matching is both injective and surjective.} typed graph homomorphism $f$ exists from $g$ onto $g'$, we write $g \stackrel{inj}{\blacktriangleright} g'$.
+When a \emph{surjective} typed graph homomorphism $f$ exists from $g$ onto $g'$ we write $g \stackrel{surj}{\blacktriangleright} g'$.
 
@@
 Our technique also requires a form of graph homomorphism that primarily focuses on matching the edges of the graphs.
-This homomorphism will be an injective match regarding the edges, but note that a particular vertex in the source graph may match onto multiple vertices in the target graph.
+This homomorphism will be an injective match regarding the edges, but note that a particular vertex in the pattern graph may match onto multiple vertices in the target graph.
 \bentley{What's the term for this?}
@@
 
-\item A function from a vertex in the target graph onto a vertex in the source graph
-\item A partial function, as not all vertices in the target graph will be reflected in the source graph
-\item Surjective, so that all source vertices are matched to target vertices
-\item A minimal function, such that the minimum number of target vertices are matched to each source vertex \bentley{Not sure if this is needed.}
+\item A function from a vertex in the target graph onto a vertex in the pattern graph
+\item A partial function, as not all vertices in the target graph will be reflected in the pattern graph
+\item Surjective, so that all pattern vertices are matched to by target vertices
+\item A minimal function, such that the minimum number of target vertices are matched to each pattern vertex \bentley{Not sure if this is needed.}
 \end{itemize}
@@
 Again, we wish to highlight the fact that this homomorphism is different than Definition~\ref{def:typed_graph_homomorphism} in the vertex matching function.
-A vertex in the target graph matches onto a vertex in the source graph, with multiple target vertices matching onto the same source vertex.
+A vertex in the target graph matches onto a vertex in the pattern graph, with multiple target vertices matching onto the same pattern vertex.
 \bentley{Create diagram to illustrate}
 
-This unusual homomorphism allows our technique to `split' our source graph over multiple locations in the target graph.
+This unusual homomorphism allows our technique to `split' our pattern graph over multiple locations in the target graph.
 This is critical, as our target graphs are constructed in a way that allows for vertex duplication.
@@
 
- When an \emph{injective} typed graph edge homomorphism $f$ exists from $g$ onto $g'$, we write $g \stackrel{inj}{\vartriangleleft} g'$.
-When a \emph{surjective} typed graph edge homomorphism $f$ exists from $g$ onto $g'$ we write $g \stackrel{surj}{\vartriangleleft} g'$.
+ When an \emph{injective} typed graph edge homomorphism $f$ exists from $g$ onto $g'$, we write $g \stackrel{inj}{\vartriangleright} g'$.
+When a \emph{surjective} typed graph edge homomorphism $f$ exists from $g$ onto $g'$ we write $g \stackrel{surj}{\vartriangleright} g'$.
 
@@
 
-Let $g$ be a graph $\langle V,E,(s, t),\tau\rangle \in \textsc{TG}$. Then the transitive closure $g^{*} = \langle V,E',(s, t),\tau\rangle \in \textsc{TG}$, where:
+Let $g$ be a graph $\langle V,E,(s, t),\tau\rangle \in \textsc{TG}$. Then the transitive closure $g^{*} = \langle V,E',(s', t'),\tau', \VT, \ET' \rangle \in \textsc{TG}$, where:
 
-$E' = E \cup $ the transitive closure of $\big\{v\xrightarrow{e}v' | \mathit{isIndirect(e)} \big\}$
+\begin{itemize}
+\item $E' = E \cup v_1\xrightarrow{e''}v_2 | \exists v_1\xrightarrow{e_i}v_i, v_i\xrightarrow{e_j}v_j, \dots, v_k\xrightarrow{e_k}v_2 \land \mathit{isIndirect(e_i)}$
+\item \bentley{How to create s, t, and tau?}
+\end{itemize}
 \end{definition}
@@
 Note that these transitive edges are identified by the function $\mathit{isIndirect}$.
-
-\bentley{I believe that the indirect edges don't need to be removed. There's a bit of handwaving, so maybe there are edge duplicates?}
 
@@
 
+\subsubsection{Similar Structures}
+
+Note that in the definitions that follow, in this section and others, that we define structures which are very similar in composition, which each containing two or three typed graphs, as well as a link component.
+This similarity is essential to our technique, as we will define morphisms between the various components of these constructions, which vary depending on the particular structure under examination.
+
+While this composition of typed graph approach is preferred by the authors due to its reflection in our prover implementation, some readers may disagree.
+In this case, note that it is equivalent to think of these structures as one large typed graph annotated by the component it originates from, with appropriate projections for selecting relevant portions.
+
 \subsubsection*{DSLTrans Transformation Rule}
@@
 This is also known in the model transformation literature as a rule's \emph{left hand side} and \emph{right hand side}.
+As described in Section 3.4 or Section 2, which the match graph elements are found in the input model, then the apply graph elements are produced in the output model.
+
 A match pattern can include indirect links that are used to transitively match containment relations in a model.
@@
 
-A rule may also contain a negative application condition (NAC) which if matched over the source graph prevents application of the rule.
+A rule may also contain a negative application condition (NAC) which if matched over the target graph prevents application of the rule.
 
@@
 \item $\mathit{NAC}$ may be an empty graph, and is disjoint from both $\mathit{Match}$ and $\mathit{Apply} $
-\item $\mathit{backward} = \{E_{back}, (s_{back}, t_{back})\}$
+
 \end{itemize}
@@
 \begin{itemize}
+\item $\mathit{backward} = \{E_{back}, (s_{back}, t_{back})\}$
 \item $E_{back}$ contains the backward links
@@
 \end{itemize}
-\item $(s_{back}, t_{back})$ is a pair of functions $s_{back}: E_{back}\rightarrow V_{\textit{Apply}}$ and $t_{back}: E_{back}\rightarrow V_{\textit{Match}}$ that respectively provide the source and target vertices for each backward link
+\item $(s_{back}, t_{back})$ is a pair of functions $s_{back}: E_{back}\rightarrow V_{\textit{Apply}}$ and $t_{back}: E_{back}\rightarrow V_{\textit{Match}}$ that respectively provide the pattern and target vertices for each backward link
 \begin{itemize}
-\item Note the source of backward links is a vertex in $V_{\textit{Match}}$ while the target is a vertex in $V_{\textit{Apply}}$
+\item Note the source of backward links is a vertex in $V_{\textit{Apply}}$ while the target is a vertex in $V_{\textit{Match}}$
 \end{itemize}
@@
 
-We must also define a function to match the traceability and backward link structures.
+We must also define a function to match link structures, such as backward links and the traceability links (to be discussed).
 
@@
 \label{def:link_homomorphism}
-Consider a set of links $L = \{E_{L}, (s_{L}, t_{L})\}$. Note that these could be backward links or traceability links.
+Consider two sets of links $L = \{E_{L}, (s_{L}, t_{L})\}$, and $T = \{E_{T}, (s_{T}, t_{T})\}$. Note that these could be backward links or traceability links.
 
-We define a function $f$ to match $L$ over $T$, such that:
+We define an injective homomorphism function $f$ to match $L$ over $T$, such that:
 
 \begin{itemize}
-\item $\forall \mathit{sourceLink} \in L:$
+\item $\forall e_L \in E_L:$
 \begin{itemize}
-\item $\exists \mathit{targetLink} \in T | f(s(\mathit{sourceLink})) = s(targetLink), $
+\item $\forall e_T \in E_T | f(e_L) = e(T)$
 \end{itemize}
@@
 \end{definition}
+
+Note that the typing of these links will be given by the component in which they are found.
+
+We also note that we will resolve the binding of all morphisms at the appropriate time, and ensure that they are consistent. \bentley{FIXFIX}
 
@@
 Note that the order of the rules within the layer does not matter, due to the semantics of DSLTrans rule execution.
+As discussed in Section BLAH, these semantics ensure that rules are independent, and cannot act on the output of other rules within the same layer.
 
@@
 To describe the semantics of a DSLTrans model transformation, we must define an \textit{input-output model} construct.
-This input-output model allows the representation of intermediate operational states during the execution of a model transformation.
+This input-output model allows the representation of the initial state as well as intermediate operational states during the execution of a model transformation.
+These input-output models are thus the `path conditions' which represent all transformation executions through an abstraction relation.
+This is further discussed in Section ABSTRACTION RELATION.
 
@@
 
-\bentley{Do we need to have V and E for all vertices and edges in the IOM? I don't think we talk about all vertices or edges in an IOM.}
-Additionally we have that $V=V'\cup V''$, $E\subseteq E'\cup E''$ and $\tau\subseteq \tau'\cup \tau''$, where the co-domain of $\tau$ is the union of the co-domains of $\tau'$ and $\tau''$ and the set $\{trace\}$.
-
 \end{definition}
@@
 
-We will now address the execution of a rule in the DSLTrans language.
+We will now address the execution of a rule in the DSLTrans language, basing our explanation on double pushouts.
 
-We will base our explanation on double pushouts, diagrammed in Figure~\ref{fig:dpo}.
-\bentley{Single or double pushouts?}
+Note that we will apply rules differently, depending on whether they contain indirect links or not.
+We will begin with the simpler case where the rule does not contain indirect links, diagrammed in Figure~\ref{fig:dpo}.
 
-\begin{figure}[h!] \centering
-\includegraphics[width=0.44\textwidth]{figures/formal/dpo}
-	\caption{Double pushout approach}
-	\label{fig:dpo}
-\end{figure}
-
-We explain the components:
-\begin{itemize}
-\item L: The matcher of the rule
-\item R: The rewriter of the rule
-\item G: The graph to be matched/rewritten
-\item H: The rewritten graph
-\end{itemize}
-
-Note that as it is not possible in DSLTrans rules to remove elements, $L = K$, and $G = D$.
+We create first the matcher and the rewriter for a DSLTrans rule.
 
@@
 
-Let $\mathit{Match}\star$ be the transitive closure of $\mathit{Match}$ to handle any indirect edges.
-
-\bentley{Typed graph? Or copy the rule's structure? If it is a rule, note that it will not be well-formed. However, this is preferable to defining another structure to use}
-
 We define $r$'s matcher, noted $\lceil r \rceil$, to be a typed graph six-tuple $\langle V,E,(s,t),\tau, \VT, \ET\rangle$, where:
 \begin{itemize}
-\item $V = V_{Match\star} \cup \big\{v \in V_{Apply} | \{\exists e \in E_{back} | s_{back}(e) = v\}\big\}$
-\item $E = E_{Match\star} \cup E_{back}$
-\item $s = s_{Match\star} \cup s_{back}$, $t = t_{Match\star} \cup t_{back}$
-\item $\tau = \tau_{Match\star} \cup \tau_{Apply}$
-\item $\VT = \VT_{Match\star} \cup \VT_{Apply}$
-\item $\ET = \ET_{Match\star} \cup \ET_{Apply}$
+\item $V = V_{Match} \cup \big\{v \in V_{Apply} | \{\exists e \in E_{back} | s_{back}(e) = v\}\big\}$
+\item $E = E_{Match} \cup E_{back}$
+\item $s = s_{Match} \cup s_{back}$, $t = t_{Match} \cup t_{back}$
+\item $\tau = \tau_{Match} \cup \tau_{Apply}$
+\item $\VT = \VT_{Match} \cup \VT_{Apply}$
+\item $\ET = \ET_{Match} \cup \ET_{Apply}$
 \end{itemize}
@@
 
-Let $\mathit{Match}\star$ be the transitive closure of $\mathit{Match}$ to handle any indirect edges.
-
-We define $r$'s rewriter, noted $\lfloor r \rfloor$, to be a input-output model $\big\langle \mathit{Input}$, $\mathit{Output}$, $\mathit{trace}\big\rangle$, where:
+We define $r$'s rewriter, noted $\lfloor r \rfloor$, to be a structure $\big\langle \mathit{Input}$, $\mathit{Output}$, $\mathit{trace}\big\rangle$, where:
 
 \begin{itemize}
-\item $ \mathit{Input} = \mathit{Match\star}$
+\item $ \mathit{Input} = \mathit{Match}$
 \item $ \mathit{Output} = \mathit{Apply}$
@@
 \begin{itemize}
-\item $E_{trace} = E_{backward} \cup \big\{e | \{$for each pair $m_i, a_j$ where $m_i \in V_{Match}, a_j \in V_{Apply}\}\big\}$
-\item $s_{trace} = s_{backward} \cup \big\{a_j | \{$for each pair $m_i, a_j$ where $m_i \in V_{Match}, a_j \in V_{Apply}\}\big\}$
-\item $t_{trace} = t_{backward} \cup \big\{m_i | \{$for each pair $m_i, a_j$ where $m_i \in V_{Match}, a_j \in V_{Apply}\}\big\}$
+\item $E_{trace} = \exists e \in E_{trace} | $
+\begin{itemize}
+\item $ \forall m \in V_{Match}, a \in V_{Apply}:$$s_{trace}(e) = a$, $t_{trace}(e) = m$ 
+\end{itemize}
 \end{itemize}
 
-\bentley{Very messy}
 \end{definition}
 
-\subsubsection*{Application of a Transformation Rule (continued)}
+\subsubsection*{Double-pushout Approach}
+
+\begin{figure}[h!] \centering
+\includegraphics[width=0.44\textwidth]{figures/formal/dpo}
+	\caption{Double pushout approach}
+	\label{fig:dpo}
+\end{figure}
+
+We explain the components:
+\begin{itemize}
+\item L: The matcher of the rule
+\item R: The rewriter of the rule
+\item G: The graph to be matched/rewritten
+\item H: The rewritten graph
+\end{itemize}
+
+Note that as it is not possible in DSLTrans rules to remove elements, in this case $L = K$, and $G = D$.
 

### 2016-07-27 text/formal_background.tex
Put dpo approach with indirect links into the paper.

@@
 
-We also note that we will resolve the binding of all morphisms at the appropriate time, and ensure that they are consistent. \bentley{FIXFIX}
+\subsubsection{Homomorphisms Between Structures}
+\label{subsubsec:structure_homomorphism}
+
+In the developments that follow, we will need to find morphisms between different structures.
+
+For example, we may need to find a homomorphism between two structures $S_1 = \langle A, B, L\rangle$ and  $S_2 = \langle A', B', L'\rangle$, where $A, A', B, B' \in TG$ and $L, L'$ are links of the form $\{E, (s, t)\}$.
+
+This overall homomorphism will compose the homomorphism $f(A) = A'$, $g(B) = B'$, $h(L) = L'$, where f and g are the homomorphisms to be found, and h is the link homomorphism.
+
+A reviewer pointed out that we also require that these sub-homomorphisms are consistent with each other. That is, the link homomorphism h must agree with 
+the bindings given by f and g.
+
+\bentley{Complete this}
 
@@
 
-\subsubsection*{Double-pushout Approach}
+\subsubsection*{Double-Pushout Approach}
 
 \begin{figure}[h!] \centering
-\includegraphics[width=0.44\textwidth]{figures/formal/dpo}
+\includegraphics[width=0.44\textwidth]{figures/formal/dpo_with_NAC}
 	\caption{Double pushout approach}
@@
 
-We explain the components:
+If the reader is unfamiliar with the theory of double-pushouts, we refer to work on Fundamentals of Algebraic Graph Transformation.
+
+The core idea of double-pushouts is to present graph rewriting in terms of morphisms, which compartmentalize the actions of finding a match and adding/deleting edges.
+
+The application of a rule is called a \textit{production}.
+Below, we will first explain the production of a single DSLTrans rule, expand this production to handle indirect links in the rule, then discuss multiple productions.
+
+Let this production be for a rule $r = \langle \mathit{NAC}, \mathit{[Match}, \mathit{Apply}, \mathit{back} \rangle$.
+To simplify the explanation, we constrain our rules to not include any indirect links.
+That is, $\nexists E \in E_{\textit{Match}} \cup E_{\mathit{NAC}} | \mathit{isIndirect}(E)$.
+
+Figure~\ref{fig:dpo} presents a basic diagram for the approach.
+The major components are:
 \begin{itemize}
-\item L: The matcher of the rule
-\item R: The rewriter of the rule
+\item L: The matcher of the rule $\lceil r \rceil$
+\item R: The rewriter of the rule $\lfloor r \rfloor$
 \item G: The graph to be matched/rewritten
@@
 
-Note that as it is not possible in DSLTrans rules to remove elements, in this case $L = K$, and $G = D$.
+\paragraph{Rule Components}
+K is the interface graph between L and R.
+That is, elements in K appear in both L and R.
+Elements which are to be deleted by the production appear in L but not K, and elements to be created are present in R but not K.
 
-\bentley{Fix up morphism notation. Is $\rightarrow$ good enough?} 
-Recall that the interface graph K is equivalent to L due to the lack of deletion in DSLTrans rules. Thus, the l homomorphism is identity. The D correspondence graph will therefore be equivalent to G.
+In the absence of indirect links, we note that $K = L$ as elements cannot be deleted in DSLTrans rules. Therefore the morphism l is isomorphism.
 
-As well, we consider r to be a trivial homomorphism between the $\mathit{Match} \rightarrow \mathit{Input}$, $\mathit{Apply} \rightarrow \mathit{Output}$, and $\mathit{backward} \rightarrow \mathit{trace}$.
+The morphism r from K to R is an injective typed graph homomorphism.
+Note that as described in Section~\ref{subsubsec:structure_homomorphism}, this homomorphism will be composed of sub-homomorphisms between the $\mathit{Match} \rightarrow \mathit{Input}$, $\mathit{Apply} \rightarrow \mathit{Output}$, and $\mathit{backward} \rightarrow \mathit{trace}$.
 
-We also note here that a rule cannot be applied if there is a homomorphism between the rule's NAC and G.
+N represents the typed graph NAC of the rule. We note that this production cannot occur if there is a typed graph edge homomorphism c between the rule's NAC and G.
 
-m is the typed graph homomorphism between the matcher of a rule $r$, denoted ($\lceil r \rceil$) and the $\mathit{Input}$ component of the input-output model
+\paragraph{Graph Components}
 
-$n$ and $g$ will be homomorphism between their respective graphs. \bentley{This needs to be explained further. Maybe with a diagram?}
+G is the target graph which is to be matched on.
+
+m is the typed graph homomorphism between the matcher of a rule $r$, denoted ($\lceil r \rceil$) and the $\mathit{Input}$ component of the input-output model.
+
+D is the interface graph between the original graph G, and the rewritten graph H.
+Similar to K, D contains elements in both.
+Again, as DSLTrans rules do not delete elements, D will be isomorphic to G in this case, and f will be isomorphism.
+The morphism k will embed the K interface graph within the interface graph D.
+
+The morphism g from the interface graph D to the rewritten graph H gives the graph elements, while the morphism $r$ gives the embedding of the rewriter R with the elements to be produced.
+Thus, it is the pushout of R and D through K that gives H.
+g and n will be typed graph morphisms.
+
+\subsubsection{Double-Pushout Approach with Indirect Links}
+
+If the DSLTrans rule contains indirect links, we must modify the double-pushout approach described above to handle the transitive closure that the construct implies.
+The main change is to build the transitive edges, match on them, and then not build them in the final graph.
+
+Note that this second explanation is valid for rules containing indirect links, or not containing them.
+
+\begin{figure}[h!] \centering
+\includegraphics[width=0.44\textwidth]{figures/formal/dpo_with_NAC_and_indirect}
+	\caption{Double pushout approach with indirect links}
+	\label{fig:dpo_indirect}
+\end{figure}
+
+We add the component G' to the diagram, which represents the transitive closure of the Input graph of G. The homomorphism i gives the embedding of G into G'.
+
+The homomorphism m will try to map the indirect links onto the newly created edges in G'.
+
+K will not contain the indirect links. Therefore l will not longer be surjective, as nothing will map to the indirect links. R and r remain the same.
+
+The interface D does not contain the edges newly created in G'. Thus the morphism f does not map onto these edges in G'.
+
+n, g, and H remain the same.
+
+\paragraph{Example}
+
+\begin{figure}[h!] \centering
+\includegraphics[width=0.44\textwidth]{figures/formal/dpo_indirect_example}
+	\caption{DPO example with indirect links}
+	\label{fig:dpo_indirect_example}
+\end{figure}
+
+Figure~\ref{fig:dpo_indirect_example} shows an example of the DPO approach when the rule may contain indirect links.
+
+Note how the orange homomorphism in the bottom left does not match to the newly created edges, nor does the green homomorphism from D.
+
+The purple homomorphism from L does match the indirect link onto the created edges. But the red homomorphism does not match onto the indirect links in L, as they do not exist in K.
+
+\bentley{Fix this up a lot.}
 

### 2016-07-29 text/abstraction_relation.tex
Updates to building PCs.
Replace DSLTrans section with Frankenstein from different sources.

@@
 \subsection{Path Conditions}
-\label{sec:gen_path_conds}
+\label{subsec:gen_path_conds}
 
@@
 
-\begin{figure}[htb]
-        \centering
-        \begin{subfigure}[b]{0.24\textwidth}
-                \centering
-                \includegraphics[width=1\textwidth]{./figures/building_path_conditions/traceability_links.pdf}
-                \caption{Before symbolic traceability links added}
-                \label{fig:traceability_links1}
-        \end{subfigure}%
-        ~~
-        \begin{subfigure}[b]{0.235\textwidth}
-                \centering
-                \includegraphics[width=1\textwidth]{./figures/building_path_conditions/traceability_links2.pdf}
-                \caption{After symbolic traceability links added}
-                \label{fig:traceability_links2}
-        \end{subfigure}%
-        \caption{Symbolic traceability links created for an abstract DSLTrans rule}
-        \label{fig:trace_links}
-\end{figure}
+%\begin{figure}[htb]
+%        \centering
+%        \begin{subfigure}[b]{0.24\textwidth}
+%                \centering
+%                \includegraphics[width=1\textwidth]{./figures/building_path_conditions/traceability_links.pdf}
+%                \caption{Before symbolic traceability links added}
+%                \label{fig:traceability_links1}
+%        \end{subfigure}%
+%        ~~
+%        \begin{subfigure}[b]{0.235\textwidth}
+%                \centering
+%                \includegraphics[width=1\textwidth]{./figures/building_path_conditions/traceability_links2.pdf}
+%                \caption{After symbolic traceability links added}
+%                \label{fig:traceability_links2}
+%        \end{subfigure}%
+%        \caption{Symbolic traceability links created for an abstract DSLTrans rule}
+%        \label{fig:trace_links}
+%\end{figure}
 
@@
 
-\begin{figure}[htb]
-        \centering
-        \begin{subfigure}[b]{0.235\textwidth}
-                \centering
-                \includegraphics[width=1\textwidth]{./figures/building_path_conditions/backward_link.pdf}
-                \caption{Rule with backward links (dashed lines)}
-                \label{fig:backward_links}
-        \end{subfigure}%
-        ~~
-        \begin{subfigure}[b]{0.235\textwidth}
-                \centering
-                \includegraphics[width=1\textwidth]{./figures/building_path_conditions/backward_link2.pdf}
-                \caption{Traceability links added\\~}
-                \label{fig:backward_links2}
-        \end{subfigure}%
-        \caption{Adding traceability links to an abstract DSLTrans rule with backward links}
-        \label{fig:back_links}
-\end{figure}
+%\begin{figure}[htb]
+%        \centering
+%        \begin{subfigure}[b]{0.235\textwidth}
+%                \centering
+%                \includegraphics[width=1\textwidth]{./figures/building_path_conditions/backward_link.pdf}
+%                \caption{Rule with backward links (dashed lines)}
+%                \label{fig:backward_links}
+%        \end{subfigure}%
+%        ~~
+%        \begin{subfigure}[b]{0.235\textwidth}
+%                \centering
+%                \includegraphics[width=1\textwidth]{./figures/building_path_conditions/backward_link2.pdf}
+%                \caption{Traceability links added\\~}
+%                \label{fig:backward_links2}
+%        \end{subfigure}%
+%        \caption{Adding traceability links to an abstract DSLTrans rule with backward links}
+%        \label{fig:back_links}
+%\end{figure}
 
@@
 
-\begin{figure}[htb]
-        \centering
-        \begin{subfigure}[b]{0.40\textwidth}
-                \centering
-                \includegraphics[width=1\textwidth]{./figures/abstraction_relation/empty_pc.pdf}
-               	\caption{Abstraction holds}
-               	\label{fig:empty_pc_success}
-        \end{subfigure}%
-        ~~\\
-        \begin{subfigure}[b]{0.40\textwidth}
-                \centering
-                \includegraphics[width=1\textwidth]{./figures/abstraction_relation/empty_pc2.pdf}
-                \caption{Abstraction does not hold}
-                \label{fig:empty_pc_fail}
-        \end{subfigure}%
-        \caption{Abstraction of transformation executions by the empty path condition}
-        \label{fig:empty_pc}
-\end{figure}
+%\begin{figure}[htb]
+%        \centering
+%        \begin{subfigure}[b]{0.40\textwidth}
+%                \centering
+%                \includegraphics[width=1\textwidth]{./figures/abstraction_relation/empty_pc.pdf}
+%               	\caption{Abstraction holds}
+%               	\label{fig:empty_pc_success}
+%        \end{subfigure}%
+%        ~~\\
+%        \begin{subfigure}[b]{0.40\textwidth}
+%                \centering
+%                \includegraphics[width=1\textwidth]{./figures/abstraction_relation/empty_pc2.pdf}
+%                \caption{Abstraction does not hold}
+%                \label{fig:empty_pc_fail}
+%        \end{subfigure}%
+%        \caption{Abstraction of transformation executions by the empty path condition}
+%        \label{fig:empty_pc}
+%\end{figure}
 
@@
 
-\begin{figure}[htb]
-        \centering
-        \begin{subfigure}[b]{0.40\textwidth}
-                \centering
-                \includegraphics[width=1\textwidth]{./figures/abstraction_relation/non_overlapping.pdf}
-               	\caption{Abstraction holds on match and apply graphs}
-               	\label{fig:non_overlapping_match_apply}
-        \end{subfigure}%
-        ~~\\
-        \begin{subfigure}[b]{0.40\textwidth}
-                \centering
-                \includegraphics[width=1\textwidth]{./figures/abstraction_relation/non_overlapping_trace_links.pdf}
-                \caption{Abstraction holds on traceability links}
-                \label{fig:non_overlapping_trace_links}
-        \end{subfigure}%
-        \caption{Abstraction of transformation execution by non-overlapping rule components}
-        \label{fig:non_overlapping}
-\end{figure}
-
-\begin{figure}[htb]
-        \centering
-        \begin{subfigure}[b]{0.40\textwidth}
-                \centering
-                \includegraphics[width=1\textwidth]{./figures/abstraction_relation/non_overlapping2.pdf}
-               	\caption{Abstraction holds on match and apply graphs}
-               	\label{fig:non_overlapping2_match_apply}
-        \end{subfigure}%
-        ~~\\
-        \begin{subfigure}[b]{0.40\textwidth}
-                \centering
-                \includegraphics[width=1\textwidth]{./figures/abstraction_relation/non_overlapping2_trace_links.pdf}
-                \caption{Abstraction holds on traceability links}
-                \label{fig:non_overlapping2_trace_links}
-        \end{subfigure}%
-        \caption{Example of abstraction over multiple rule execution}
-        \label{fig:non_overlapping2}
-\end{figure}
+%\begin{figure}[htb]
+%        \centering
+%        \begin{subfigure}[b]{0.40\textwidth}
+%                \centering
+%                \includegraphics[width=1\textwidth]{./figures/abstraction_relation/non_overlapping.pdf}
+%               	\caption{Abstraction holds on match and apply graphs}
+%               	\label{fig:non_overlapping_match_apply}
+%        \end{subfigure}%
+%        ~~\\
+%        \begin{subfigure}[b]{0.40\textwidth}
+%                \centering
+%                \includegraphics[width=1\textwidth]{./figures/abstraction_relation/non_overlapping_trace_links.pdf}
+%                \caption{Abstraction holds on traceability links}
+%                \label{fig:non_overlapping_trace_links}
+%        \end{subfigure}%
+%        \caption{Abstraction of transformation execution by non-overlapping rule components}
+%        \label{fig:non_overlapping}
+%\end{figure}
+%
+%
+%\begin{figure}[htb]
+%        \centering
+%        \begin{subfigure}[b]{0.40\textwidth}
+%                \centering
+%                \includegraphics[width=1\textwidth]{./figures/abstraction_relation/non_overlapping2.pdf}
+%               	\caption{Abstraction holds on match and apply graphs}
+%               	\label{fig:non_overlapping2_match_apply}
+%        \end{subfigure}%
+%        ~~\\
+%        \begin{subfigure}[b]{0.40\textwidth}
+%                \centering
+%                \includegraphics[width=1\textwidth]{./figures/abstraction_relation/non_overlapping2_trace_links.pdf}
+%                \caption{Abstraction holds on traceability links}
+%                \label{fig:non_overlapping2_trace_links}
+%        \end{subfigure}%
+%        \caption{Example of abstraction over multiple rule execution}
+%        \label{fig:non_overlapping2}
+%\end{figure}
 
@@
 
-\begin{figure}[htb]
-        \centering
-        \begin{subfigure}[b]{0.40\textwidth}
-                \centering
-                \includegraphics[width=1\textwidth]{./figures/abstraction_relation/overlapping.pdf}
-               	\caption{Abstraction holds on match and apply graphs}
-               	\label{fig:overlapping_match_apply}
-        \end{subfigure}%
-        ~~\\
-        \begin{subfigure}[b]{0.40\textwidth}
-                \centering
-                \includegraphics[width=1\textwidth]{./figures/abstraction_relation/overlapping_trace_links.pdf}
-                \caption{Abstraction holds on traceability links}
-                \label{fig:overlapping_trace_links}
-        \end{subfigure}%
-        \caption{Abstraction of transformation execution by overlapping rule components}
-        \label{fig:overlapping}
-\end{figure}
+%\begin{figure}[htb]
+%        \centering
+%        \begin{subfigure}[b]{0.40\textwidth}
+%                \centering
+%                \includegraphics[width=1\textwidth]{./figures/abstraction_relation/overlapping.pdf}
+%               	\caption{Abstraction holds on match and apply graphs}
+%               	\label{fig:overlapping_match_apply}
+%        \end{subfigure}%
+%        ~~\\
+%        \begin{subfigure}[b]{0.40\textwidth}
+%                \centering
+%                \includegraphics[width=1\textwidth]{./figures/abstraction_relation/overlapping_trace_links.pdf}
+%                \caption{Abstraction holds on traceability links}
+%                \label{fig:overlapping_trace_links}
+%        \end{subfigure}%
+%        \caption{Abstraction of transformation execution by overlapping rule components}
+%        \label{fig:overlapping}
+%\end{figure}
 
@@
 
-\begin{figure}[htb]
-        \centering
-        \begin{subfigure}[b]{0.40\textwidth}
-                \centering
-                \includegraphics[width=1\textwidth]{./figures/abstraction_relation/overlapping2.pdf}
-               	\caption{Elements cannot overlap within a component}
-               	\label{fig:overlapping2_match_apply}
-        \end{subfigure}%
-        ~~\\
-        \begin{subfigure}[b]{0.40\textwidth}
-                \centering
-                \includegraphics[width=1\textwidth]{./figures/abstraction_relation/overlapping2_trace_links.pdf}
-                \caption{Traceability links cannot be found}
-                \label{fig:overlapping2_trace_links}
-        \end{subfigure}%
-        \caption{Abstraction does not hold}
-        \label{fig:overlapping2}
-\end{figure}
+%\begin{figure}[htb]
+%        \centering
+%        \begin{subfigure}[b]{0.40\textwidth}
+%                \centering
+%                \includegraphics[width=1\textwidth]{./figures/abstraction_relation/overlapping2.pdf}
+%               	\caption{Elements cannot overlap within a component}
+%               	\label{fig:overlapping2_match_apply}
+%        \end{subfigure}%
+%        ~~\\
+%        \begin{subfigure}[b]{0.40\textwidth}
+%                \centering
+%                \includegraphics[width=1\textwidth]{./figures/abstraction_relation/overlapping2_trace_links.pdf}
+%                \caption{Traceability links cannot be found}
+%                \label{fig:overlapping2_trace_links}
+%        \end{subfigure}%
+%        \caption{Abstraction does not hold}
+%        \label{fig:overlapping2}
+%\end{figure}
 
@@
 
-\begin{figure}[htb]
-        \centering
-        \begin{subfigure}[b]{0.40\textwidth}
-                \centering
-                \includegraphics[width=1\textwidth]{./figures/abstraction_relation/indirect.pdf}
-               	\caption{Matching over match and apply graphs}
-               	\label{fig:indirect_match_apply}
-        \end{subfigure}%
-        ~~\\
-        \begin{subfigure}[b]{0.40\textwidth}
-                \centering
-                \includegraphics[width=1\textwidth]{./figures/abstraction_relation/indirect_trace_links.pdf}
-                \caption{Matching over traceability links}
-                \label{fig:indirect_trace_links}
-        \end{subfigure}%
-        \caption{Abstraction with indirect links}
-        \label{fig:indirect}
-\end{figure}
-
-\begin{figure*}[htb]
-        \centering
-        \begin{subfigure}[b]{0.70\textwidth}
-                \centering
-                \includegraphics[width=1\textwidth]{./figures/abstraction_relation/combination.pdf}
-               	\caption{Abstraction holds on match and apply graphs}
-               	\label{fig:combination_match_apply}
-        \end{subfigure}%
-        ~~\\
-        \begin{subfigure}[b]{0.70\textwidth}
-                \centering
-                \includegraphics[width=1\textwidth]{./figures/abstraction_relation/combination_trace_links.pdf}
-                \caption{Abstraction holds on traceability links}
-                \label{fig:combination_trace_links}
-        \end{subfigure}%
-        \caption{Abstraction where path condition has combined rules}
-        \label{fig:combination}
-\end{figure*}
+%\begin{figure}[htb]
+%        \centering
+%        \begin{subfigure}[b]{0.40\textwidth}
+%                \centering
+%                \includegraphics[width=1\textwidth]{./figures/abstraction_relation/indirect.pdf}
+%               	\caption{Matching over match and apply graphs}
+%               	\label{fig:indirect_match_apply}
+%        \end{subfigure}%
+%        ~~\\
+%        \begin{subfigure}[b]{0.40\textwidth}
+%                \centering
+%                \includegraphics[width=1\textwidth]{./figures/abstraction_relation/indirect_trace_links.pdf}
+%                \caption{Matching over traceability links}
+%                \label{fig:indirect_trace_links}
+%        \end{subfigure}%
+%        \caption{Abstraction with indirect links}
+%        \label{fig:indirect}
+%\end{figure}
+%
+%\begin{figure*}[htb]
+%        \centering
+%        \begin{subfigure}[b]{0.70\textwidth}
+%                \centering
+%                \includegraphics[width=1\textwidth]{./figures/abstraction_relation/combination.pdf}
+%               	\caption{Abstraction holds on match and apply graphs}
+%               	\label{fig:combination_match_apply}
+%        \end{subfigure}%
+%        ~~\\
+%        \begin{subfigure}[b]{0.70\textwidth}
+%                \centering
+%                \includegraphics[width=1\textwidth]{./figures/abstraction_relation/combination_trace_links.pdf}
+%                \caption{Abstraction holds on traceability links}
+%                \label{fig:combination_trace_links}
+%        \end{subfigure}%
+%        \caption{Abstraction where path condition has combined rules}
+%        \label{fig:combination}
+%\end{figure*}
 

### 2016-07-29 text/building_path_conditions.tex
Replace DSLTrans section with Frankenstein from different sources.

@@
 
-\begin{figure*}[htb]
-        \centering
-        \begin{subfigure}[b]{0.34\textwidth}
-                \centering
-                \includegraphics[width=1\textwidth]{./figures/building_path_conditions/next_layer.pdf}
-                \caption{Previous path conditions are combined\\ with rules}
-                \label{fig:next_layer}
-        \end{subfigure}%
-        ~~
-        \begin{subfigure}[b]{0.34\textwidth}
-                \centering
-                \includegraphics[width=1\textwidth]{./figures/overview/layers_pc.pdf}
-                \caption{Combining a path condition with two rules\\~}
-                \label{fig:layers_pc2}
-        \end{subfigure}%
-
-         \caption{Two components in the path condition creation process}
-         \label{fig:combining_path_conditions}
-\end{figure*}
+%\begin{figure*}[htb]
+%        \centering
+%        \begin{subfigure}[b]{0.34\textwidth}
+%                \centering
+%                \includegraphics[width=1\textwidth]{./figures/building_path_conditions/next_layer.pdf}
+%                \caption{Previous path conditions are combined\\ with rules}
+%                \label{fig:next_layer}
+%        \end{subfigure}%
+%        ~~
+%        \begin{subfigure}[b]{0.34\textwidth}
+%                \centering
+%                \includegraphics[width=1\textwidth]{./figures/overview/layers_pc.pdf}
+%                \caption{Combining a path condition with two rules\\~}
+%                \label{fig:layers_pc2}
+%        \end{subfigure}%
+%        
+%       
+%         \caption{Two components in the path condition creation process}
+%         \label{fig:combining_path_conditions}
+%\end{figure*}
 
@@
 
-\begin{figure*}[htb]
-        \centering
-        \begin{subfigure}[b]{0.7\textwidth}
-                \centering
-                \includegraphics[width=1\textwidth]{./figures/building_path_conditions/total_satisfied_dependencies.pdf}
-                \caption{Totally satisfied at one location}
-                \label{fig:total_satisfied_dependencies}
-        \end{subfigure}%
-        \\
-        \begin{subfigure}[b]{0.8\textwidth}
-                \centering
-                \includegraphics[width=1\textwidth]{./figures/building_path_conditions/multiple_total_satisfied_dependencies.pdf}
-                \caption{Totally satisfied at multiple locations - \reviewer{Fix boxes on bottom}}
-                \label{fig:multiple_total_satisfied_dependencies}
-        \end{subfigure}%
-        \caption{R's dependencies are totally satisfied by PC}
-        \label{fig:totes_sat_deps}
-\end{figure*}
+%\begin{figure*}[htb]
+%        \centering
+%        \begin{subfigure}[b]{0.7\textwidth}
+%                \centering
+%                \includegraphics[width=1\textwidth]{./figures/building_path_conditions/total_satisfied_dependencies.pdf}
+%                \caption{Totally satisfied at one location}
+%                \label{fig:total_satisfied_dependencies}
+%        \end{subfigure}%
+%        \\
+%        \begin{subfigure}[b]{0.8\textwidth}
+%                \centering
+%                \includegraphics[width=1\textwidth]{./figures/building_path_conditions/multiple_total_satisfied_dependencies.pdf}
+%                \caption{Totally satisfied at multiple locations - \reviewer{Fix boxes on bottom}}
+%                \label{fig:multiple_total_satisfied_dependencies}
+%        \end{subfigure}%
+%        \caption{R's dependencies are totally satisfied by PC}
+%        \label{fig:totes_sat_deps}
+%\end{figure*}
 

### 2016-07-29 text/dsltrans.tex
Replace DSLTrans section with Frankenstein from different sources.

@@
-\section{The DSLTrans Transformation Language}
-\label{sec:dsltrans}
+\section{DSLTrans}
+\label{sec:DSLTrans_formal}
 
@@
 
+\subsection{DSLTrans Introduction}
+
+DSLTrans is a visual graph-based and rule-based model
+transformation engine that has two important properties enforced by construction: all its computations
+are both \emph{terminating} and \emph{confluent}~\cite{Barroca2011}.
+These properties stem from the fact that DSLTrans does not allow
+unbounded loops during execution, making it a Turing-incomplete
+computing language~\cite{Barroca2011}. Besides their obvious importance in
+practice, \emph{termination} and \emph{confluence} were instrumental in the
+implementation of our verification technique for pre-/post-condition contracts.
+
+Model transformations are expressed in DSLTrans as sets of graph rewriting
+rules, having an upper part (named \emph{MatchModel}), a lower part (\emph{ApplyModel}) and,
+optionally, negative application conditions. The main construction
+used in the scheduling of model transformation rules in DSLTrans is a
+\emph{layer}.
+Each model transformation rule in a layer cannot match over the output of any other rule in the same layer.
+As well, rules cannot modify the input graph during the rewriting phase (termed \emph{out-place} execution).
+Layers are organized sequentially and
+the output model that results from executing a given layer is passed as input to
+the next layer in the sequence.
+
+%Model transformations are expressed in DSLTrans as sets of graph rewriting
+%rules, having the classical left-hand side (LHS), right-hand side (RHS) and,
+%optionally, negative application conditions. The main construction
+%used in the scheduling of model transformation rules in DSLTrans is a
+%\emph{layer}: a layer contains a set of model transformation rules that execute
+%without being able to see each other's output. This forces the rules in a layer
+%to execute independently from each other. Layers are organized sequentially and
+%the output model that results from executing a given layer is passed as input to
+%the next layer in the sequence.
+
+A DSLTrans rule can match over the elements of the input model of the
+transformation and also over elements that have been generated so far in the
+output model. Matching over elements of the output model of a
+transformation is achieved using a DSLTrans construct called \emph{backward
+links}. Backward links allow matching over traces between elements in the input and
+the output models of the transformation. These traces are explicitly built by
+the DSLTrans transformation engine during rule execution.
+
+ \begin{figure}[t]
+   \begin{center}
+     \includegraphics[width=0.40\textwidth]{figures/DSLTrans/UWRule.pdf}
+     \caption{An example of a DSLTrans rule}
+     \label{fig:DSLTrans_rule}
+   \end{center}
+   \vspace{-0.25in}
+ \end{figure}
+
+For example, we depict in Figure~\ref{fig:DSLTrans_rule} a
+rule in the DSLTrans language.
+When a rule is executed, the graph in the \emph{MatchModel} of the rule is searched for in the transformation's input model, together with the classes in the \emph{ApplyModel} of the rule that are connected to \emph{backward
+links}. An example of a \emph{backward link} can be observed in
+Figure~\ref{fig:DSLTrans_rule} as a dotted line connecting the \emph{Country} and the
+\emph{Community} match classes. During the rewrite part of rule application,
+the instances of classes in the \emph{ApplyModel} of the rule that are not connected to
+backward links, together with their adjacent relations, are created in the
+output model.
+
+For example, the \emph{UnionWomanRule} rule in Figure~\ref{fig:DSLTrans_rule} will match over a \emph{Country} element connected to a \emph{Family} element connected to a \emph{Parent} element.
+If these elements are found in the input model along with the corresponding \emph{Community} and \emph{Woman} elements in the output model, then a \emph{persons} relation will be created between those output elements.
+
+Although not present in this rule,
+copying object attribute values from the \emph{MatchModel} to the \emph{ApplyModel} of the rules is
+also part of the DSLTrans language, as illustrated in Section~\ref{sec:DSLTransRepresentation}.
+
 \reviewer{Please explain, earlier on, the meaning of indirect edges.}
 
-\begin{figure*}[t]
-        \centering
-        \begin{subfigure}[b]{0.40\textwidth}
-                \centering
-                \includegraphics[width=1\textwidth]{./figures/policestation_dsltrans/organization.pdf}
-                \caption{Organization language}
-                \label{fig:OrganizationLanguage}
-        \end{subfigure}%
-        ~~
-        \begin{subfigure}[b]{0.40\textwidth}
-                \centering
-                \includegraphics[width=1\textwidth]{./figures/policestation_dsltrans/gender.pdf}
-                \caption{Gender language}
-                \label{fig:GenderLanguage}
-        \end{subfigure}%
-        \caption{Metamodels for the Police Station transformation}
-        \label{fig:squadmetamodel}
-\end{figure*}
+%\begin{figure*}[t]
+%        \centering
+%        \begin{subfigure}[b]{0.40\textwidth}
+%                \centering
+%                \includegraphics[width=1\textwidth]{./figures/policestation_dsltrans/organization.pdf}
+%                \caption{Organization language}
+%                \label{fig:OrganizationLanguage}
+%        \end{subfigure}%
+%        ~~
+%        \begin{subfigure}[b]{0.40\textwidth}
+%                \centering
+%                \includegraphics[width=1\textwidth]{./figures/policestation_dsltrans/gender.pdf}
+%                \caption{Gender language}
+%                \label{fig:GenderLanguage}
+%        \end{subfigure}%
+%        \caption{Metamodels for the Police Station transformation}
+%        \label{fig:squadmetamodel}
+%\end{figure*}
 
@@
 
-\begin{figure*}[htb]
-        \centering
-        \begin{subfigure}[b]{0.30\textwidth}
-                \centering
-                \includegraphics[width=1\textwidth]{./figures/policestation_dsltrans/model_police_hierarchy.pdf}
-                \caption{Original model}
-                \label{fig:police_hierarchy}
-        \end{subfigure}%
-        ~~
-        \begin{subfigure}[b]{0.40\textwidth}
-                \centering
-                \includegraphics[width=1\textwidth]{./figures/policestation_dsltrans/model_police_gender.pdf}
-                \caption{Transformed model}
-                \label{fig:police_gender}
-        \end{subfigure}%
-        \caption{Model before and after transformation}
-        \label{fig:transformationexample}
-\end{figure*}
+%\begin{figure*}[htb]
+%        \centering
+%        \begin{subfigure}[b]{0.30\textwidth}
+%                \centering
+%                \includegraphics[width=1\textwidth]{./figures/policestation_dsltrans/model_police_hierarchy.pdf}
+%                \caption{Original model}
+%                \label{fig:police_hierarchy}
+%        \end{subfigure}%
+%        ~~
+%        \begin{subfigure}[b]{0.40\textwidth}
+%                \centering
+%                \includegraphics[width=1\textwidth]{./figures/policestation_dsltrans/model_police_gender.pdf}
+%                \caption{Transformed model}
+%                \label{fig:police_gender}
+%        \end{subfigure}%
+%        \caption{Model before and after transformation}
+%        \label{fig:transformationexample}
+%\end{figure*}
 \reviewer{Figure 3: I would suggest adding another node f5 superviesd by f4, to illustrate the
@@
 
-\subsection{Properties to Prove}
-
-The properties we aim to prove on the Police Station transformation are structural properties.
-These properties are composed of a pre-condition and a post-condition component, as seen in \cref{fig:properties}.
-
-Informally, a property can be read as `if the pre-condition graph matches on the input model to the transformation, then the post-condition graph will match any output model produced'.
-Further details as well as formal validity and completeness of the property proving process are discussed in \cref{sec:verif_dsltrans_props}.
-
-As a brief example of property syntax and semantics, consider the property in \cref{fig:dsltrans_prop1}.
-The pre-condition graph is composed of a Station element connected to a Female element and a Male element, where all elements are from the \emph{Organization language} metamodel.
-This structure is repeated in the post-condition graph, with the difference that the metamodel for these elements is the \emph{Gender language}.
-Thus, this property represents the statement ``\emph{a model which includes a
-police station that has both male and female officers will be
-transformed into a model where the male officer will exist in the male set
-and the female officer will exist in the female set}''. We expect this property to always hold in our transformation.
-
-\begin{figure*}[thb]
-        \centering
-        \begin{subfigure}[b]{0.45\textwidth}
-                \centering
-                \includegraphics[height=5.5cm]{./figures/policestation_dsltrans/property1.pdf}
-                \caption{Property 1 -- Expected to hold}
-                \label{fig:dsltrans_prop1}
-        \end{subfigure}%
-        ~~
-        \begin{subfigure}[b]{0.45\textwidth}
-                \centering
-                \includegraphics[height=5.5cm]{./figures/policestation_dsltrans/property2.pdf}
-                \caption{Property 2 -- Not expected to hold}
-                \label{fig:dsltrans_prop2}
-        \end{subfigure}%
-        \caption{ Properties to be proved on the Police Station transformation}
-        \label{fig:properties}
-\end{figure*}
-
-\reviewer{Figure 4: The two “agent” edges in the postcondition should be marked “female”
-resp. “male”} 
-In contrast, we do not expect the property in \cref{fig:dsltrans_prop2} to always hold. This property represents the statement ``\emph{any model which includes a
-female officer will be transformed into a model where that female officer will
-always supervise another female officer}''.
-It is not difficult to construct an input model where the pre-condition holds, but the post-condition does not.
-For example, take an input model that contains only one female officer, as there will only be one female officer in the output model.
-
-We shall discuss our property proving technique in \cref{sec:verif_dsltrans_props}.
-Following this, experiments in \cref{sec:experiments} will present experimental results from proving these two properties on the Police Station transformation.
+In addition to the constructs presented in the example in
+Figure~\ref{fig:DSLTrans_rule}, DSLTrans has several others:
+\emph{existential matching} which allows selecting only one result when a match class of a rule
+matches an input model, \emph{indirect links} for transitive matching
+over containment relations in the input model, and \emph{negative application
+conditions} that allow the transformation designer to specify conditions under
+which a rule should not match. These constructs are not currently used in
+our verification approach, and the interested reader is referred to~\cite{Barroca2011} for further information.
 
@@
 \end{itemize}
+
+\subsection{Families To Persons}
+
+As our running example we present an extended version of the \FTP
+transformation described in \cite{Oakes2015}. The original \FTP transformation can be
+found in the ATL zoo~\cite{atlZoo}, and has also been discussed in a number of related works
+on verification and testing~\cite{Gogolla2011}.
+
+We chose this \FTP transformation as our running example for two reasons. First, it transforms domains whose concepts are easily understandable by anyone (cf. Section~\ref{sec:trans_domain}).
+Second, it has a certain degree of complexity since it uses many features available in the ATL language (cf. Section~\ref{sec:trans_explanation}).
+
+\subsection{Transformation Domains}\label{sec:trans_domain}
+
+\begin{figure}[bt]
+\centering
+\subfigure[\emph{Families\_Extended} metamodel]{
+\includegraphics[width=0.97\columnwidth]{figures/DSLTrans/Families_Extended.pdf}
+\label{fig:fams}
+}
+\subfigure[\emph{Persons\_Extended} metamodel]{
+\includegraphics[width=0.97\columnwidth]{figures/DSLTrans/Persons_Extended.pdf}
+\label{fig:pers}
+}
+\caption{Metamodels of the \emph{Families-to-Persons\_Extended} transformation}\label{fig:metamodels}
+\end{figure}
+
+The input and output metamodels of this transformation are shown in Figure~\ref{fig:metamodels}.
+Please note that abstract classes are depicted in grey color and with italic names, and inheritance relationships are depicted in grey.
+
+The input metamodel, the \emph{Families\_Extended} metamodel, has the \emph{Country} class as a root element.
+A \emph{Country} is made up of \emph{companies}, \emph{families} and \emph{cities}.
+A \emph{Family} has a \emph{lastName}, is \emph{registeredIn} a \emph{Neighborhood} and can have any number of \emph{mothers} and \emph{fathers}, who are \emph{Parent}s and may, in turn, work in (\emph{worksIn}) a \emph{Company}.
+It can also contain any number of \emph{sons} and \emph{daughters}, who are \emph{Child}ren, and every child \emph{goesTo} a \emph{School}.
+Both parents and children are \emph{Member}s that have a \emph{firstName}, belong to a \emph{family} and each of them \emph{livesIn} a \emph{City}.
+
+A \emph{City} may contain \emph{companies}, and a \emph{Company}, in turn, can be present in (relationship \emph{isIn}) several distinct cities.
+A \emph{City} is composed of \emph{neighborhoods}, and these can have \emph{schools}, where several \emph{students} are registered.
+Every \emph{School} has \emph{Service}s, and these may be \emph{special}, for students with special needs, or simply offer \emph{ordinary} services.
+Finally, countries, cities, companies, neighborhoods and schools have a \emph{name} attribute, which is inherited from the abstract \emph{NamedElement} class.
+
+The output metamodel, \emph{Persons\_Extended}, is shown in Figure~\ref{fig:pers}.
+The root class is \emph{Community}, which is made up of \emph{persons}, \emph{townHalls} and \emph{associations}.
+A \emph{Person} has a \emph{fullName} and can either be a \emph{Man} or a \emph{Woman}.
+An \emph{Association} has a \emph{Committee} that makes decisions.
+Every \emph{TownHall} has a roster of \emph{workers} (all the persons that are employed), hosts a \emph{Committee} to make decisions, and also governs several \emph{districts}.
+A \emph{District} may contain several \emph{facilities}, either of type \emph{SpecialFacility} for those with special needs, or \emph{OrdinaryFacility}.
+Each \emph{Facility} may have registered several persons as \emph{members}.
+Finally, associations, town halls, committees, districts and facilities have a \emph{name} attribute.
+
+\subsection{DSLTrans Representation}\label{sec:DSLTransRepresentation}
+
+\begin{figure*}
+  \begin{center}
+  \includegraphics[width=0.94\textwidth]{figures/DSLTrans/DSLTrans_Rules_javi.pdf}
+  \caption{DSLTrans version of the \emph{Families-to-Persons\_Extended} transformation}
+  \label{fig:DSLTrans_rules}
+  \end{center}
+\end{figure*}
+
+Figure~\ref{fig:DSLTrans_rules} displays the DSLTrans transformation which corresponds to the ATL \emph{Families-to-Persons\_Extended}\\
+transformation shown in Listing~\ref{lst:F2P_Extended}. Let us mention here that we have removed five rules from the figure to improve visual clarity.
+There is a vertical dotted blue line for each of these rules, located where the rules have been removed.
+The missed rules are similar to those that surround them, and can therefore be safely ignored in our explanation.
+
+%As explained in Section~\ref{sec:dsltrans}, each DSLTrans rule has a \emph{MatchModel} and an \emph{ApplyModel}.
+%As we can see, each of them has a left-hand side (LHS) and a right-hand side (RHS)\levi{not sure about the LHS/RHS reference here\ldots}.
+%There are elements of a specific type and associations among them displayed in
+% both the upper part (named \emph{MatchedModel} in DSLTrans) and lower part (called \emph{ApplyModel}) of the rules.
+%In each side, elements of a specific type and relationships among them are displayed.
+%Elements can also have attributes.
+%Rule \emph{Households2Community} corresponds to rule R1 in
+%Listing~\ref{lst:Families2Persons}, \emph{Father2Man} corresponds to rule R2 and \emph{UnionFather}
+%to binding B11.
+
+%They appear in the \emph{ApplyModel} when they have to be initialized, and
+%in the \emph{MatchedModel} when they are used to initialize attributes in the \emph{ApplyModel}.
+%For instance, attributes \emph{fullName} and \emph{lastName} of \emph{Member} and \emph{Family} in rule \emph{Father2Man} are used
+%to initialize attribute \emph{fullName} of \emph{Man}.
+%%
+%%Other than these attributes, there are others in the RHS which appear in an element
+%%whenever such element is going to be used in backward links, and they also appear when they are being used in such links. These attributes are always referred to with the name \emph{ApplyAttribute}.
+%%For instance, we can see in rule \emph{Households2Community}, which is created for R1 in Listing~\ref{lst:Families2Persons}, that a \emph{Community} is created for a \emph{Households}.
+%%Besides, since these two elements are going to be used later in a backward link, the \emph{ApplyAttribute} attribute is set in \emph{Community}.
+Note that the \emph{ApplyAttribute} attribute of an element must be created when the element is created, and the rule where it is created must be in the sequence of rules before the rule that uses the element in a backward link.
+%
+%Rule \emph{Households2Community} corresponds to R1 in
+%Listing~\ref{lst:Families2Persons}, although the reference \emph{has} is not
+%initialized in this rule.
+%
+%Rule \emph{Father2Man} corresponds to R2. We can see how the two elements in the
+%\emph{from} part of the ATL rule appear in the \emph{MatchedModel} of the DSLTrans rule, as well
+%as the filter used in ATL, which is represented as an association. The
+%element \emph{Man} appears in the \emph{ApplyModel}, and its \emph{fullName} attribute is
+%updated with the concatenation of the \emph{firstName} and \emph{lastName}
+%attributes. A similar rule, not shown here, creates elements of type
+%\emph{Woman}, which corresponds to R3 in Listing~\ref{lst:Families2Persons}.
+%
+%The last rule is used to set the \emph{has} relation between a \emph{Community}
+%and a \emph{Man}. Here, backward links are used. The first one is to indicate
+%that the \emph{Community} element has been created from a \emph{Households}
+%element (which was done in rule \emph{Households2Community}). The other two
+%backward links state that the \emph{Man} is created from a \emph{Family} and a
+%\emph{Member} that are linked with the relation \emph{father} (such as in rule
+%\emph{Father2Man}). This rule corresponds to binding B11 in
+%Listing~\ref{lst:Families2Persons}. There is a similar rule, not shown in the
+%figure, for initializing the \emph{has} association between \emph{Community} and
+%\emph{Woman}, what corresponds to B12.
+
+The process of constructing a DSLTrans transformation from an ATL one is described in the next section.
+For now, note that DSLTrans transformation obtained from ATL through the higher-order transformation includes only one rule per layer, meaning all rules execute sequentially.
+This is due to the sequential semantics of ATL that we replicate in DSLTrans.
+
+Also, note that attribute copies are represented with arrows from the \emph{ApplyModel} of the rule to the \emph{MatchModel}, such as in rule \emph{Neighborhood2District}, where the created \emph{District} gets the same name as the matched \emph{Neighborhood}.
+The string of an attribute of a created element can also be initialized with the concatenation of several strings.
+For instance, in rule \emph{Father2Man}, the full name of the created \emph{Man} comes from the concatenation of the first name of the matched \emph{Parent} and the last name of his/her \emph{Family}.
+Or it can be assigned the string of an attribute of an element in the \emph{MatchModel} concatenated with a given string, such as in rule \emph{City2TownHall}.
+
+%\emph{Father2Man} in Figure~\ref{fig:DSLTrans_transformation} performs an attribute
+%copy from the match to the apply part of the rule, graphically represented with arrows from the \emph{ApplyModel} to the \emph{MatchModel}. More precisely, it
+%fills in the \emph{fullName} attribute of the generated \emph{Man} instance by
+%concatenating the \emph{firstName} and \emph{lastName} attribute values found
+%respectively in the \emph{member} and \emph{family} instances identified by the
+%match part of the rule. This operation replicates the concatenation of the
+%\emph{firstName} and the \emph{lastName} occurring in the rule component marked
+%\emph{B2} in Listing~\ref{lst:Families2Persons}.
+
+%The first three rules in the DSLTrans transformation replicate the R1, R2, and R3
+% portions of the ATL transformation. Then, the last four rules in
+% Figure~\ref{fig:DSLTrans_transformation} connect those elements and set the values, replicating B11, B12, B2, and B3 in Listing~\ref{lst:Families2Persons}.
+%
+% This transformation has been built automatically from its ATL
+%counterpart using a higher-order transformation, which will be presented in Section~\ref{sec:mapping}. For now, we will briefly the differences and
+%similarities between the two transformations:
+%
+%\begin{itemize}
+%  \item The DSLTrans F2P transformation is completely graphical, while the ATL
+%  F2P transformation is fully textual.
+%  \item The DSLTrans F2P is composed of layers that are organized sequentially.
+%  A layer in DSLTrans is a box (e.g. the \emph{Root} rule in
+%  Figure~\ref{fig:DSLTrans_transformation}) that can contain one or more rules.
+%  In the transformation in Figure~\ref{fig:DSLTrans_transformation}) each layer
+%  contains only one rule, as that layout is automatically produced by the higher
+%  order transformation.
+%  \item The layers belonging to the transformation in
+%  Figure~\ref{fig:DSLTrans_transformation} execute one after the other, passing
+%  the result of one layer to the next one.
+%\end{itemize}
+
+\subsection{Formal Structures}
+
+This section will detail the abstract syntax of the constructs involved in a DSLTrans transformation.
+
+As discussed in Section~\ref{subsec:DSLTrans_syntax}, DSLTrans transformations are composed of rules arranged in layers.
+
+\subsubsection{Similar Structures}
+
+Note that in the definitions that follow, in this section and others, that we define structures which are very similar in composition, which each containing two or three typed graphs, as well as a link component.
+This similarity is essential to our technique, as we will define morphisms between the various components of these constructions, which vary depending on the particular structure under examination.
+
+While this composition of typed graph approach is preferred by the authors due to its reflection in our prover implementation, some readers may disagree.
+In this case, note that it is equivalent to think of these structures as one large typed graph annotated by the component it originates from, with appropriate projections for selecting relevant portions.
+
+\subsubsection*{DSLTrans Transformation Rule}
+
+A transformation rule is the elemental block of a DSLTrans transformation. Several transformation rules can be observed in the Police Station transformation in \cref{fig:dsltransformation}.
+
+A transformation rule includes a non-empty match pattern and a non-empty apply pattern.
+This is also known in the model transformation literature as a rule's \emph{left hand side} and \emph{right hand side}.
+As described in Section 3.4 or Section 2, which the match graph elements are found in the input model, then the apply graph elements are produced in the output model.
+
+A match pattern can include indirect links that are used to transitively match containment relations in a model.
+The apply pattern of a rule always contains at least one apply element that is not connected to a backward link or an edge, meaning in practice that a rule will always produce something and not only match.
+An apply pattern does not include indirect links as it is used only for the construction of parts of instances of a metamodel.
+
+A rule may also contain a negative application condition (NAC) which if matched over the target graph prevents application of the rule.
+
+A transformation rule also includes backward links, as informally introduced in \cref{subsec:DSLTrans_constructs}.
+As described in that section, backward links define dependencies between rules.
+\bentley{Expand}
+
+\begin{definition}{DSLTrans Transformation Rule\\}
+\label{def:transformation_rule}
+
+A DSLTrans transformation rule is a four-tuple $\big\langle \mathit{NAC}$, $\mathit{Match}$, $\mathit{Apply}$, $\mathit{backward}\big\rangle$, where:
+
+\begin{itemize}
+\item $\mathit{NAC}$, $\mathit{Match}, \mathit{Apply} \in \textsc{TG}$
+\item $\mathit{Match}$ and $\mathit{Apply} $ are non-empty and are disjoint
+\item $\mathit{NAC}$ may be an empty graph, and is disjoint from both $\mathit{Match}$ and $\mathit{Apply} $
+
+\end{itemize}
+
+Note that when we require an element of the $\mathit{Match}$, $\mathit{Apply}$, or $\mathit{NAC}$ graphs, such as the vertices, we will index the required element.
+For example, the vertices for the \textit{Match} graph will be $V_{\textit{Match}}$.
+
+\begin{itemize}
+\item $\mathit{backward} = \{E_{back}, (s_{back}, t_{back})\}$
+\item $E_{back}$ contains the backward links
+\begin{itemize}
+\item $E_{back} $ is disjoint from $E_{Match}$,  $E_{Apply}$, and $E_{NAC}$
+\end{itemize}
+\item $(s_{back}, t_{back})$ is a pair of functions $s_{back}: E_{back}\rightarrow V_{\textit{Apply}}$ and $t_{back}: E_{back}\rightarrow V_{\textit{Match}}$ that respectively provide the pattern and target vertices for each backward link
+\begin{itemize}
+\item Note the source of backward links is a vertex in $V_{\textit{Apply}}$ while the target is a vertex in $V_{\textit{Match}}$
+\end{itemize}
+\end{itemize}
+
+\textsc{Rules} is the set of all rules.
+
+\end{definition}
+
+We additionally impose that for the well-formedness of a DSLTrans rule, there always exists an element or edge to be created in the \textit{Apply} graph to be created.
+That is, either there are edges to be created in the \textit{Apply} graph, or there is a vertex in the \textit{Apply} graph that is not the source of a backward link.
+
+\begin{itemize}
+\item $E_{\textit{Apply}} \neq \emptyset \lor$
+\item $ \exists v \in V_{\textit{Apply}} | \big\{\forall e \in E_{back}: s_{back}(e) \neq v \big\}$
+\end{itemize}
+
+\subsubsection*{DSLTrans Layer and Transformation}
+
+\cref{def:layer} and \cref{def:transformation} formalise the abstract syntax of a model transformation, introduced in \cref{sec:dsltrans}.
+An example of a model transformation can be observed in \cref{fig:dsltransformation}, the Police Station transformation.
+As expected, a DSLTrans transformation is composed of a sequence of layers where each layer is composed of a set of rules.
+
+\begin{definition}{Layer\\}
+\label{def:layer}
+A layer is a finite set of transformation rules: $\{r_0, r_1, \dots, r_n | r_i \in \textsc{Rules}\}$.
+
+The set of all layers is denoted $\textsc{Layers}$. 
+\end{definition}
+
+Note that the order of the rules within the layer does not matter, due to the semantics of DSLTrans rule execution.
+As discussed in Section BLAH, these semantics ensure that rules are independent, and cannot act on the output of other rules within the same layer.
+
+\begin{definition}{DSLTrans Transformation\\}
+\label{def:transformation}
+
+A DSLTrans transformation is a finite list of layers denoted $\{l_0, l_1, \dots, l_n | l \in \textsc{Layers}\}$. Note that the order of layers in a transformation is important.
+
+The set of all transformations is denoted $\textsc{Transforms}$.
+
+\end{definition}
+
+\subsubsection*{Input-Output Model}
+
+To describe the semantics of a DSLTrans model transformation, we must define an \textit{input-output model} construct.
+This input-output model allows the representation of the initial state as well as intermediate operational states during the execution of a model transformation.
+These input-output models are thus the `path conditions' which represent all transformation executions through an abstraction relation.
+This is further discussed in Section ABSTRACTION RELATION.
+
+The structure of a input-output model is intentionally very similar to a DSLTrans rule.
+There is one typed graph representing the input model and another typed graph representing the output model.
+As well, the construct contains a set of edges, named \emph{traceability links}, for keeping a history of which elements in the output model originated from which elements in the input model.
+
+\begin{definition}{Input-Output Model\\}
+\label{def:input_output_model}
+
+An input-output model rule is a three-tuple $\big\langle \mathit{Input}$, $\mathit{Output}$, $\mathit{trace}\big\rangle$, where:
+
+\begin{itemize}
+\item $\mathit{Input}, \mathit{Output} \in \textsc{TG}$
+\item $\mathit{Input}$ and $\mathit{Output} $ may be empty and are disjoint
+\item $\mathit{trace} = \{E_{trace}, (s_{trace}, t_{trace})\}$
+\begin{itemize}
+\item $E_{trace}$ contains the traceability links
+\item $E_{trace} $ is disjoint from $E_{Input}$ and $E_{Output}$
+\item $(s_{trace}, t_{trace})$ is a pair of functions $s_{trace}: E_{trace}\rightarrow V_{\textit{Output}}$ and $t_{trace}: E_{trace}\rightarrow V_{\textit{Input}}$ that respectively provide the source and target vertices for each traceability link
+\end{itemize}
+
+\end{itemize}
+
+Let $\mathit{IOM}$ be the set of all input-output models.
+
+We define a utility function getTransformation: IOM $\rightarrow$ Transforms.
+This function returns the Transform that the input-output model was built for.
+The purpose of this function is to restrict input-output models to only be applicable for the transformation they represent.
+
+Note that the $\VT$ and $\ET$ for input and output will come from the input and output metamodels. \bentley{Expand}
+
+\end{definition}
+
+\subsection{Transformation Semantics}
+\label{sec:DSLTrans_semantics}
+
+This section will discuss the semantics of a DSLTrans transformation. Given an input model and a transformation, an output layer will be produced through the repeated execution of rules and layers.
+
+\subsubsection*{Execution of a DSLTrans Rule}
+
+We will now address the execution of a rule in the DSLTrans language, basing our explanation on double pushouts.
+
+Note that we will apply rules differently, depending on whether they contain indirect links or not.
+We will begin with the simpler case where the rule does not contain indirect links, diagrammed in Figure~\ref{fig:dpo}.
+
+We create first the matcher and the rewriter for a DSLTrans rule.
+
+\subsubsection*{Matcher of a Transformation Rule}
+
+To create the actual matcher construct that will be matched during a rule's execution, the rule's \textit{Match} graph is combined with backward links, as well as any \textit{Apply} vertices connected to the backward links.
+
+Note that the rule's NAC will also be considered in the matching of a rule.
+
+\begin{definition}{Matcher of a Transformation Rule\\}
+\label{def:matcher_transformation_rule}
+
+Let the transformation rule $r = \big\langle \mathit{NAC}$, $\mathit{Match}$, $\mathit{Apply}$, $\mathit{backward}\big\rangle$.
+
+We define $r$'s matcher, noted $\lceil r \rceil$, to be a typed graph six-tuple $\langle V,E,(s,t),\tau, \VT, \ET\rangle$, where:
+\begin{itemize}
+\item $V = V_{Match} \cup \big\{v \in V_{Apply} | \{\exists e \in E_{back} | s_{back}(e) = v\}\big\}$
+\item $E = E_{Match} \cup E_{back}$
+\item $s = s_{Match} \cup s_{back}$, $t = t_{Match} \cup t_{back}$
+\item $\tau = \tau_{Match} \cup \tau_{Apply}$
+\item $\VT = \VT_{Match} \cup \VT_{Apply}$
+\item $\ET = \ET_{Match} \cup \ET_{Apply}$
+\end{itemize}
+
+\end{definition}
+
+\subsubsection*{Rewriter of a Transformation Rule}
+
+To continue with the double pushout approach, we must construct the rewriting (or replacement) graph.
+
+The construction of this rewriter is essentially the same as the underlying rule.
+However, to support traceability, we require that backward links in the rule be converted into traceability links, and that new traceability links be created between all match and apply vertices.
+
+\begin{definition}{Rewriter of a Transformation Rule\\}
+\label{def:rewriter_transformation_rule}
+
+Let the transformation rule $r = \big\langle \mathit{NAC}$, $\mathit{Match}$, $\mathit{Apply}$, $\mathit{backward}\big\rangle$.
+
+We define $r$'s rewriter, noted $\lfloor r \rfloor$, to be a structure $\big\langle \mathit{Input}$, $\mathit{Output}$, $\mathit{trace}\big\rangle$, where:
+
+\begin{itemize}
+\item $ \mathit{Input} = \mathit{Match}$
+\item $ \mathit{Output} = \mathit{Apply}$
+\end{itemize}
+
+We then modify $\mathit{trace}$ to contain the appropriate traceability links
+
+\begin{itemize}
+\item $E_{trace} = \exists e \in E_{trace} | $
+\begin{itemize}
+\item $ \forall m \in V_{Match}, a \in V_{Apply}:$$s_{trace}(e) = a$, $t_{trace}(e) = m$ 
+\end{itemize}
+\end{itemize}
+
+\end{definition}
+
+\subsubsection*{Double-Pushout Approach}
+
+\begin{figure}[h!] \centering
+\includegraphics[width=0.44\textwidth]{figures/formal/dpo_with_NAC}
+	\caption{Double pushout approach}
+	\label{fig:dpo}
+\end{figure}
+
+If the reader is unfamiliar with the theory of double-pushouts, we refer to work on Fundamentals of Algebraic Graph Transformation.
+
+The core idea of double-pushouts is to present graph rewriting in terms of morphisms, which compartmentalize the actions of finding a match and adding/deleting edges.
+
+The application of a rule is called a \textit{production}.
+Below, we will first explain the production of a single DSLTrans rule, expand this production to handle indirect links in the rule, then discuss multiple productions.
+
+Let this production be for a rule $r = \langle \mathit{NAC}, \mathit{[Match}, \mathit{Apply}, \mathit{back} \rangle$.
+To simplify the explanation, we constrain our rules to not include any indirect links.
+That is, $\nexists E \in E_{\textit{Match}} \cup E_{\mathit{NAC}} | \mathit{isIndirect}(E)$.
+
+Figure~\ref{fig:dpo} presents a basic diagram for the approach.
+The major components are:
+\begin{itemize}
+\item L: The matcher of the rule $\lceil r \rceil$
+\item R: The rewriter of the rule $\lfloor r \rfloor$
+\item G: The graph to be matched/rewritten
+\item H: The rewritten graph
+\end{itemize}
+
+\paragraph{Rule Components}
+K is the interface graph between L and R.
+That is, elements in K appear in both L and R.
+Elements which are to be deleted by the production appear in L but not K, and elements to be created are present in R but not K.
+
+In the absence of indirect links, we note that $K = L$ as elements cannot be deleted in DSLTrans rules. Therefore the morphism l is isomorphism.
+
+The morphism r from K to R is an injective typed graph homomorphism.
+Note that as described in Section~\ref{subsubsec:structure_homomorphism}, this homomorphism will be composed of sub-homomorphisms between the $\mathit{Match} \rightarrow \mathit{Input}$, $\mathit{Apply} \rightarrow \mathit{Output}$, and $\mathit{backward} \rightarrow \mathit{trace}$.
+
+N represents the typed graph NAC of the rule. We note that this production cannot occur if there is a typed graph edge homomorphism c between the rule's NAC and G.
+
+\paragraph{Graph Components}
+
+G is the target graph which is to be matched on.
+
+m is the typed graph homomorphism between the matcher of a rule $r$, denoted ($\lceil r \rceil$) and the $\mathit{Input}$ component of the input-output model.
+
+D is the interface graph between the original graph G, and the rewritten graph H.
+Similar to K, D contains elements in both.
+Again, as DSLTrans rules do not delete elements, D will be isomorphic to G in this case, and f will be isomorphism.
+The morphism k will embed the K interface graph within the interface graph D.
+
+The morphism g from the interface graph D to the rewritten graph H gives the graph elements, while the morphism $r$ gives the embedding of the rewriter R with the elements to be produced.
+Thus, it is the pushout of R and D through K that gives H.
+g and n will be typed graph morphisms.
+
+\subsubsection{Double-Pushout Approach with Indirect Links}
+
+If the DSLTrans rule contains indirect links, we must modify the double-pushout approach described above to handle the transitive closure that the construct implies.
+The main change is to build the transitive edges, match on them, and then not build them in the final graph.
+
+Note that this second explanation is valid for rules containing indirect links, or not containing them.
+
+\begin{figure}[h!] \centering
+\includegraphics[width=0.44\textwidth]{figures/formal/dpo_with_NAC_and_indirect}
+	\caption{Double pushout approach with indirect links}
+	\label{fig:dpo_indirect}
+\end{figure}
+
+We add the component G' to the diagram, which represents the transitive closure of the Input graph of G. The homomorphism i gives the embedding of G into G'.
+
+The homomorphism m will try to map the indirect links onto the newly created edges in G'.
+
+K will not contain the indirect links. Therefore l will not longer be surjective, as nothing will map to the indirect links. R and r remain the same.
+
+The interface D does not contain the edges newly created in G'. Thus the morphism f does not map onto these edges in G'.
+
+n, g, and H remain the same.
+
+\paragraph{Example}
+
+\begin{figure}[h!] \centering
+\includegraphics[width=0.44\textwidth]{figures/formal/dpo_indirect_example}
+	\caption{DPO example with indirect links}
+	\label{fig:dpo_indirect_example}
+\end{figure}
+
+Figure~\ref{fig:dpo_indirect_example} shows an example of the DPO approach when the rule may contain indirect links.
+
+Note how the orange homomorphism in the bottom left does not match to the newly created edges, nor does the green homomorphism from D.
+
+The purple homomorphism from L does match the indirect link onto the created edges. But the red homomorphism does not match onto the indirect links in L, as they do not exist in K.
+
+\bentley{Fix this up a lot.}
+
+\subsubsection*{DSLTrans Layer Execution}
+
+Now that the production of a particular DSLTrans rule has been explained, the next definition defines how a DSLTrans transformation layer executes.
+Recall that a layer is composed of rules, and acts on the current input-output model.
+
+\begin{definition} {DSLTrans Layer Execution}
+\label{def:dsltrans_layer_execution}
+
+The execution of a DSLTrans layer can be defined as a function as follows:
+
+\textit{applyLayer} : IOM $\times$ Layer $\rightarrow$ IOM
+
+Let $l = \{r_0, r_1, \dots, r_n | r_i \in \textsc{Rules}\}$, $l \in \textsc{Layers}$.
+
+Let $\mathit{IOM}_{Input}$, $\mathit{IOM}_{Output} \in \mathit{IOMs}$.
+
+$\mathit{IOM}_{Output}$ will be the result of applying the production \textit{p} as defined below to $\mathit{IOM}_{Input}.$
+
+We refer to the double-pushout literature for the creation of \textit{p}.
+Specifically, we refer to the definition of parallel direct derivations.
+This p will be created out of the productions created for each of the rules $r_i$ in the layer.
+
+$p = \big\langle (p_1, {in}^1), \dots, (p_k, {in}^k) \big\rangle : (L \xleftarrow{l} L \xrightarrow{r} R)$
+
+This collection of the rules is possible because DSLTrans rules do not delete anything. Essentially, a union is taken of the L, K, R components of each production.
+\end{definition}
+
+\begin{figure}[h!] \centering
+\includegraphics[width=0.44\textwidth]{figures/formal/parallel_productions}
+	\caption{Parallel production \textit{p}}
+	\label{fig:parallel_productions}
+\end{figure}
+
+\cref{def:dsltrans_layer_execution} is the core of DSLTrans' semantics, in which we build the result of executing a layer of a DSLTrans transformation.
+Many model transformation languages are based on graph rewriting, where the result of each rule rewrite is immediately usable by all other rules.
+In DSLTrans the result of executing one layer in DSLTrans is totally produced before the input to the layer is changed.
+This is enforced in \cref{def:dsltrans_layer_execution} by the fact that the layer production is composed of a union of the rule's productions.
+Rules belonging to the same layer are thus forced to execute independently, as described in the section on DSLTrans semantics.
+
+\bentley{How does this handle different multiplicity of rule application?}
+
+\subsubsection*{DSLTrans Transformation Execution}
+
+Our final definition is for the execution of a DSLTrans transformation. Essentially, it is the chaining of layer executions applied on an input-output model.
+
+We will begin by discussing the conditions for executing a model transformation on an input model.
+
+\paragraph{Input Conditions}
+
+For a well-formed transformation execution, we require the \textit{input-output model} in the domain to contain only an input graph.
+Recall that an \textit{input-output model} is the three-tuple $\big\langle \mathit{Input}$, $\mathit{Output}$, $\mathit{trace}\big\rangle$.
+We therefore require that both the $\mathit{Output}$ and $\mathit{trace}$ components in the domain IOM to be empty.
+This input-output model thus represents the first step of the transformation, where no rule has been executed yet.
+
+The input-output model in the co-domain will contain this input graph, as well as the output graph and traceability links produced by the execution of the transformation.
+Let $\textsc{Execs}_t$ be the set of all well-formed input-output models produced for transformation $t$.
+
+\begin{definition} {Execution of a DSLTrans Transformation\\}
+\label{def:DSLTrans_transformation_execution}
+
+The execution of a model transformation is a function $\mathit{applyTransformation}: \mathit{IOM} \times \mathit{Transforms} \rightarrow \mathit{IOM}$.
+
+This function is a chaining of executions of those layers within the transformation.
+
+Recall the application of a layer is \textit{applyLayer} : IOM $\times$ Layer $\rightarrow$ IOM. Therefore, the application of a transformation is:
+
+$\mathit{IOM}$ as input to chaining of layer executions \bentley{Is this function chaining? I don't know how to write the equation for this.}
+
+\end{definition}
+
+While the execution of the rules belonging to a layer happens in parallel, the execution of the layers of a transformation happens sequentially.
+As per \cref{def:DSLTrans_transformation_execution}, the input-output model that is the output of executing a given layer is passed onto the next layer as input.
+The final input-output model created will be the result of the model transformation.
+
+\subsection{Confluence and Termination Properties}
+
+We now prove two important properties about executions of transformations expressed in the subset of DSLTrans presented in this paper: \emph{confluence} and \emph{termination}.
+The proofs are provided at a high level, given the fact that DSLTrans essentially enforces both these properties by construction of the semantics of DSLTrans.
+
+\begin{proposition}{Confluence}
+
+Every model transformation execution is confluent up to typed graph isomorphism.
+
+\bentley{As we are relying on the double-pushout approach, I think we get this confluence for free.}
+
+\end{proposition}
+
+\begin{pf}
+We want to prove that for every model transformation execution of a transformation $\mathit{transform} \in \textsc{Transforms}$ having as input an input-output model $\mathit{input} \in \mathit{IOMs}$, its output is always the same up to typed graph isomorphism.\\
+
+If we assume an execution of the transformation is not confluent then this should happen because of non-determinism when the execution of a transformation is being built.
+Non-determinism happens during the construction of a transformation execution at two points:
+\begin{enumerate}
+\item The section `Execution of a DSLTrans Rule' discusses the matching and rewriting components to rule application.
+Note that our technique relies on the double-pushout approach, which itself is non-deterministic up to typed graph isomorphism, which does not contradict the proposition we are trying to prove.
+\bentley{Revise this.}
+
+\item In Definition~\ref{def:dsltrans_layer_execution}, the rules composing a transformation layer are composed into a production on the input graph.
+Note that the order in which the transformation rules are composed may be non-deterministic.
+However, as these rules can be shown to be parallel independent \bentley{And they have to in that definition}, the production is by definition confluent up to typed graph isomorphism.
+
+\end{enumerate}
+Given there are no other sources of non-determinism when building the execution of a transformation, every model transformation execution is confluent up to typed graph isomorphism.
+\end{pf}
+
+\begin{proposition}{Termination}
+
+Every model transformation execution terminates.
+\end{proposition}
+\begin{pf}
+Let us assume that there is a transformation execution which does not terminate.
+In order for this to happen there must exist a part in the construction of the execution of a transformation which induces an algorithm with an infinite amount of steps.
+We identify three moments when this can happen:
+\begin{enumerate}
+
+\item If the matching and rewriting process of a rule is infinite.
+However, as the double-pushout approach uses a single production, then this must be a finite process.
+\bentley{Verify this, especially in the case of multiple applications of the rule.}
+
+% if the result of the $match_{rl}(m_{in})$ function in definition~\ref{def:match_function} is an infinite set of match-apply graphs.
+The input-output model $m_{in}$ is by definition finite and the matching of each rule is independent from the execution of other rules in the same layer.
+As such, the number of subgraphs of $m_{in}$ isomorphic to $rl$'s matcher found during the execution of $rl$ is finite.
+
+\item if Definition~\ref{def:dsltrans_layer_execution} of execution of a layer induces an infinite amount of steps.
+The only possibility for this to happen is if a layer has an infinite amount of transformation rules, which is a contradiction with Definition~\ref{def:layer}.
+
+\item if Definition~\ref{def:DSLTrans_transformation_execution} of execution of a transformation induces an infinite amount of steps.
+Given layers are executed sequentially and no looping is allowed, the only possibility for this to happen is if the transformation has an infinite amount of layers, which contradicts Definition~\ref{def:transformation}.
+
+\end{enumerate}
+Given there are no other constructs in the semantics of a transformation that can induce an infinite amount of steps, every model transformation execution terminates.
+\end{pf}

### 2016-07-29 text/formal_background.tex
Replace DSLTrans section with Frankenstein from different sources.

--- 2016-07-25 text/formal_background.tex
@@
 
-\subsection{DSLTrans Constructs}
-\label{sec:DSLTrans_formal}
-
-This section will detail the abstract syntax of the constructs involved in a DSLTrans transformation.
-
-As discussed in Section~\ref{subsec:DSLTrans_syntax}, DSLTrans transformations are composed of rules arranged in layers.
-
-\subsubsection{Similar Structures}
-
-Note that in the definitions that follow, in this section and others, that we define structures which are very similar in composition, which each containing two or three typed graphs, as well as a link component.
-This similarity is essential to our technique, as we will define morphisms between the various components of these constructions, which vary depending on the particular structure under examination.
-
-While this composition of typed graph approach is preferred by the authors due to its reflection in our prover implementation, some readers may disagree.
-In this case, note that it is equivalent to think of these structures as one large typed graph annotated by the component it originates from, with appropriate projections for selecting relevant portions.
-
-\subsubsection*{DSLTrans Transformation Rule}
-
-A transformation rule is the elemental block of a DSLTrans transformation. Several transformation rules can be observed in the Police Station transformation in \cref{fig:dsltransformation}.
-
-A transformation rule includes a non-empty match pattern and a non-empty apply pattern.
-This is also known in the model transformation literature as a rule's \emph{left hand side} and \emph{right hand side}.
-As described in Section 3.4 or Section 2, which the match graph elements are found in the input model, then the apply graph elements are produced in the output model.
-
-A match pattern can include indirect links that are used to transitively match containment relations in a model.
-The apply pattern of a rule always contains at least one apply element that is not connected to a backward link or an edge, meaning in practice that a rule will always produce something and not only match.
-An apply pattern does not include indirect links as it is used only for the construction of parts of instances of a metamodel.
-
-A rule may also contain a negative application condition (NAC) which if matched over the target graph prevents application of the rule.
-
-A transformation rule also includes backward links, as informally introduced in \cref{subsec:DSLTrans_constructs}.
-As described in that section, backward links define dependencies between rules.
-\bentley{Expand}
-
-\begin{definition}{DSLTrans Transformation Rule\\}
-\label{def:transformation_rule}
-
-A DSLTrans transformation rule is a four-tuple $\big\langle \mathit{NAC}$, $\mathit{Match}$, $\mathit{Apply}$, $\mathit{backward}\big\rangle$, where:
-
-\begin{itemize}
-\item $\mathit{NAC}$, $\mathit{Match}, \mathit{Apply} \in \textsc{TG}$
-\item $\mathit{Match}$ and $\mathit{Apply} $ are non-empty and are disjoint
-\item $\mathit{NAC}$ may be an empty graph, and is disjoint from both $\mathit{Match}$ and $\mathit{Apply} $
-
-\end{itemize}
-
-Note that when we require an element of the $\mathit{Match}$, $\mathit{Apply}$, or $\mathit{NAC}$ graphs, such as the vertices, we will index the required element.
-For example, the vertices for the \textit{Match} graph will be $V_{\textit{Match}}$.
-
-\begin{itemize}
-\item $\mathit{backward} = \{E_{back}, (s_{back}, t_{back})\}$
-\item $E_{back}$ contains the backward links
-\begin{itemize}
-\item $E_{back} $ is disjoint from $E_{Match}$,  $E_{Apply}$, and $E_{NAC}$
-\end{itemize}
-\item $(s_{back}, t_{back})$ is a pair of functions $s_{back}: E_{back}\rightarrow V_{\textit{Apply}}$ and $t_{back}: E_{back}\rightarrow V_{\textit{Match}}$ that respectively provide the pattern and target vertices for each backward link
-\begin{itemize}
-\item Note the source of backward links is a vertex in $V_{\textit{Apply}}$ while the target is a vertex in $V_{\textit{Match}}$
-\end{itemize}
-\end{itemize}
-
-\textsc{Rules} is the set of all rules.
-
-\end{definition}
-
-We additionally impose that for the well-formedness of a DSLTrans rule, there always exists an element or edge to be created in the \textit{Apply} graph to be created.
-That is, either there are edges to be created in the \textit{Apply} graph, or there is a vertex in the \textit{Apply} graph that is not the source of a backward link.
-
-\begin{itemize}
-\item $E_{\textit{Apply}} \neq \emptyset \lor$
-\item $ \exists v \in V_{\textit{Apply}} | \big\{\forall e \in E_{back}: s_{back}(e) \neq v \big\}$
-\end{itemize}
-
 \subsubsection{Link Homomorphism}
@@
 
-We also note that we will resolve the binding of all morphisms at the appropriate time, and ensure that they are consistent. \bentley{FIXFIX}
+\subsubsection{Homomorphisms Between Structures}
+\label{subsubsec:structure_homomorphism}
 
-\subsubsection*{DSLTrans Layer and Transformation}
+In the developments that follow, we will need to find morphisms between different structures.
 
-\cref{def:layer} and \cref{def:transformation} formalise the abstract syntax of a model transformation, introduced in \cref{sec:dsltrans}.
-An example of a model transformation can be observed in \cref{fig:dsltransformation}, the Police Station transformation.
-As expected, a DSLTrans transformation is composed of a sequence of layers where each layer is composed of a set of rules.
+For example, we may need to find a homomorphism between two structures $S_1 = \langle A, B, L\rangle$ and  $S_2 = \langle A', B', L'\rangle$, where $A, A', B, B' \in TG$ and $L, L'$ are links of the form $\{E, (s, t)\}$.
 
-\begin{definition}{Layer\\}
-\label{def:layer}
-A layer is a finite set of transformation rules: $\{r_0, r_1, \dots, r_n | r_i \in \textsc{Rules}\}$.
+This overall homomorphism will compose the homomorphism $f(A) = A'$, $g(B) = B'$, $h(L) = L'$, where f and g are the homomorphisms to be found, and h is the link homomorphism.
 
-The set of all layers is denoted $\textsc{Layers}$. 
-\end{definition}
+A reviewer pointed out that we also require that these sub-homomorphisms are consistent with each other. That is, the link homomorphism h must agree with 
+the bindings given by f and g.
 
-Note that the order of the rules within the layer does not matter, due to the semantics of DSLTrans rule execution.
-As discussed in Section BLAH, these semantics ensure that rules are independent, and cannot act on the output of other rules within the same layer.
-
-\begin{definition}{DSLTrans Transformation\\}
-\label{def:transformation}
-
-A DSLTrans transformation is a finite list of layers denoted $\{l_0, l_1, \dots, l_n | l \in \textsc{Layers}\}$. Note that the order of layers in a transformation is important.
-
-The set of all transformations is denoted $\textsc{Transforms}$.
-
-\end{definition}
-
-\subsubsection*{Input-Output Model}
-
-To describe the semantics of a DSLTrans model transformation, we must define an \textit{input-output model} construct.
-This input-output model allows the representation of the initial state as well as intermediate operational states during the execution of a model transformation.
-These input-output models are thus the `path conditions' which represent all transformation executions through an abstraction relation.
-This is further discussed in Section ABSTRACTION RELATION.
-
-The structure of a input-output model is intentionally very similar to a DSLTrans rule.
-There is one typed graph representing the input model and another typed graph representing the output model.
-As well, the construct contains a set of edges, named \emph{traceability links}, for keeping a history of which elements in the output model originated from which elements in the input model.
-
-\begin{definition}{Input-Output Model\\}
-\label{def:input_output_model}
-
-An input-output model rule is a three-tuple $\big\langle \mathit{Input}$, $\mathit{Output}$, $\mathit{trace}\big\rangle$, where:
-
-\begin{itemize}
-\item $\mathit{Input}, \mathit{Output} \in \textsc{TG}$
-\item $\mathit{Input}$ and $\mathit{Output} $ may be empty and are disjoint
-\item $\mathit{trace} = \{E_{trace}, (s_{trace}, t_{trace})\}$
-\begin{itemize}
-\item $E_{trace}$ contains the traceability links
-\item $E_{trace} $ is disjoint from $E_{Input}$ and $E_{Output}$
-\item $(s_{trace}, t_{trace})$ is a pair of functions $s_{trace}: E_{trace}\rightarrow V_{\textit{Output}}$ and $t_{trace}: E_{trace}\rightarrow V_{\textit{Input}}$ that respectively provide the source and target vertices for each traceability link
-\end{itemize}
-
-\end{itemize}
-
-Let $\mathit{IOM}$ be the set of all input-output models.
-
-We define a utility function getTransformation: IOM $\rightarrow$ Transforms.
-This function returns the Transform that the input-output model was built for.
-The purpose of this function is to restrict input-output models to only be applicable for the transformation they represent.
-
-Note that the $\VT$ and $\ET$ for input and output will come from the input and output metamodels. \bentley{Expand}
-
-\end{definition}
-
-\subsection{Transformation Semantics}
-\label{sec:DSLTrans_semantics}
-
-This section will discuss the semantics of a DSLTrans transformation. Given an input model and a transformation, an output layer will be produced through the repeated execution of rules and layers.
-
-\subsubsection*{Execution of a DSLTrans Rule}
-
-We will now address the execution of a rule in the DSLTrans language, basing our explanation on double pushouts.
-
-Note that we will apply rules differently, depending on whether they contain indirect links or not.
-We will begin with the simpler case where the rule does not contain indirect links, diagrammed in Figure~\ref{fig:dpo}.
-
-We create first the matcher and the rewriter for a DSLTrans rule.
-
-\subsubsection*{Matcher of a Transformation Rule}
-
-To create the actual matcher construct that will be matched during a rule's execution, the rule's \textit{Match} graph is combined with backward links, as well as any \textit{Apply} vertices connected to the backward links.
-
-Note that the rule's NAC will also be considered in the matching of a rule.
-
-\begin{definition}{Matcher of a Transformation Rule\\}
-\label{def:matcher_transformation_rule}
-
-Let the transformation rule $r = \big\langle \mathit{NAC}$, $\mathit{Match}$, $\mathit{Apply}$, $\mathit{backward}\big\rangle$.
-
-We define $r$'s matcher, noted $\lceil r \rceil$, to be a typed graph six-tuple $\langle V,E,(s,t),\tau, \VT, \ET\rangle$, where:
-\begin{itemize}
-\item $V = V_{Match} \cup \big\{v \in V_{Apply} | \{\exists e \in E_{back} | s_{back}(e) = v\}\big\}$
-\item $E = E_{Match} \cup E_{back}$
-\item $s = s_{Match} \cup s_{back}$, $t = t_{Match} \cup t_{back}$
-\item $\tau = \tau_{Match} \cup \tau_{Apply}$
-\item $\VT = \VT_{Match} \cup \VT_{Apply}$
-\item $\ET = \ET_{Match} \cup \ET_{Apply}$
-\end{itemize}
-
-\end{definition}
-
-\subsubsection*{Rewriter of a Transformation Rule}
-
-To continue with the double pushout approach, we must construct the rewriting (or replacement) graph.
-
-The construction of this rewriter is essentially the same as the underlying rule.
-However, to support traceability, we require that backward links in the rule be converted into traceability links, and that new traceability links be created between all match and apply vertices.
-
-\begin{definition}{Rewriter of a Transformation Rule\\}
-\label{def:rewriter_transformation_rule}
-
-Let the transformation rule $r = \big\langle \mathit{NAC}$, $\mathit{Match}$, $\mathit{Apply}$, $\mathit{backward}\big\rangle$.
-
-We define $r$'s rewriter, noted $\lfloor r \rfloor$, to be a structure $\big\langle \mathit{Input}$, $\mathit{Output}$, $\mathit{trace}\big\rangle$, where:
-
-\begin{itemize}
-\item $ \mathit{Input} = \mathit{Match}$
-\item $ \mathit{Output} = \mathit{Apply}$
-\end{itemize}
-
-We then modify $\mathit{trace}$ to contain the appropriate traceability links
-
-\begin{itemize}
-\item $E_{trace} = \exists e \in E_{trace} | $
-\begin{itemize}
-\item $ \forall m \in V_{Match}, a \in V_{Apply}:$$s_{trace}(e) = a$, $t_{trace}(e) = m$ 
-\end{itemize}
-\end{itemize}
-
-\end{definition}
-
-\subsubsection*{Double-pushout Approach}
-
-\begin{figure}[h!] \centering
-\includegraphics[width=0.44\textwidth]{figures/formal/dpo}
-	\caption{Double pushout approach}
-	\label{fig:dpo}
-\end{figure}
-
-We explain the components:
-\begin{itemize}
-\item L: The matcher of the rule
-\item R: The rewriter of the rule
-\item G: The graph to be matched/rewritten
-\item H: The rewritten graph
-\end{itemize}
-
-Note that as it is not possible in DSLTrans rules to remove elements, in this case $L = K$, and $G = D$.
-
-\bentley{Fix up morphism notation. Is $\rightarrow$ good enough?} 
-Recall that the interface graph K is equivalent to L due to the lack of deletion in DSLTrans rules. Thus, the l homomorphism is identity. The D correspondence graph will therefore be equivalent to G.
-
-As well, we consider r to be a trivial homomorphism between the $\mathit{Match} \rightarrow \mathit{Input}$, $\mathit{Apply} \rightarrow \mathit{Output}$, and $\mathit{backward} \rightarrow \mathit{trace}$.
-
-We also note here that a rule cannot be applied if there is a homomorphism between the rule's NAC and G.
-
-m is the typed graph homomorphism between the matcher of a rule $r$, denoted ($\lceil r \rceil$) and the $\mathit{Input}$ component of the input-output model
-
-$n$ and $g$ will be homomorphism between their respective graphs. \bentley{This needs to be explained further. Maybe with a diagram?}
-
-\subsubsection*{DSLTrans Layer Execution}
-
-Now that the application of a particular DSLTrans rule has been explained, the next definition defines how a DSLTrans transformation layer executes.
-Recall that a layer is composed of rules, and acts on the current input-output model.
-
-\begin{definition} {DSLTrans Layer Execution}
-\label{def:dsltrans_layer_execution}
-
-The execution of a DSLTrans layer can be defined as a function as follows:
-
-\textit{applyLayer} : IOM $\times$ Layer $\rightarrow$ IOM
-
-Let $l = \{r_0, r_1, \dots, r_n | r_i \in \textsc{Rules}\}$, $l \in \textsc{Layers}$.
-
-Let $\mathit{IOM}_{Input}$, $\mathit{IOM}_{Output} \in \mathit{IOMs}$.
-
-$\mathit{IOM}_{Output}$ will be the result of applying the production \textit{p} as defined below to $\mathit{IOM}_{Input}.$
-
-We refer to the double-pushout literature for the creation of \textit{p}.
-Specifically, we refer to the definition of parallel direct derivations.
-This p will be created out of the productions created for each of the rules $r_i$ in the layer.
-
-$p = \big\langle (p_1, {in}^1), \dots, (p_k, {in}^k) \big\rangle : (L \xleftarrow{l} L \xrightarrow{r} R)$
-
-This collection of the rules is possible because DSLTrans rules do not delete anything. Essentially, a union is taken of the L, K, R components of each production.
-\end{definition}
-
-\begin{figure}[h!] \centering
-\includegraphics[width=0.44\textwidth]{figures/formal/parallel_productions}
-	\caption{Parallel production \textit{p}}
-	\label{fig:parallel_productions}
-\end{figure}
-
-\cref{def:dsltrans_layer_execution} is the core of DSLTrans' semantics, in which we build the result of executing a layer of a DSLTrans transformation.
-Many model transformation languages are based on graph rewriting, where the result of each rule rewrite is immediately usable by all other rules.
-In DSLTrans the result of executing one layer in DSLTrans is totally produced before the input to the layer is changed.
-This is enforced in \cref{def:dsltrans_layer_execution} by the fact that the layer production is composed of a union of the rule's productions.
-Rules belonging to the same layer are thus forced to execute independently, as described in the section on DSLTrans semantics.
-
-\bentley{How does this handle different multiplicity of rule application?}
-
-\subsubsection*{DSLTrans Transformation Execution}
-
-Our final definition is for the execution of a DSLTrans transformation. Essentially, it is the chaining of layer executions applied on an input-output model.
-
-We will begin by discussing the conditions for executing a model transformation on an input model.
-
-\paragraph{Input Conditions}
-
-For a well-formed transformation execution, we require the \textit{input-output model} in the domain to contain only an input graph.
-Recall that an \textit{input-output model} is the three-tuple $\big\langle \mathit{Input}$, $\mathit{Output}$, $\mathit{trace}\big\rangle$.
-We therefore require that both the $\mathit{Output}$ and $\mathit{trace}$ components in the domain IOM to be empty.
-This input-output model thus represents the first step of the transformation, where no rule has been executed yet.
-
-The input-output model in the co-domain will contain this input graph, as well as the output graph and traceability links produced by the execution of the transformation.
-Let $\textsc{Execs}_t$ be the set of all well-formed input-output models produced for transformation $t$.
-
-\begin{definition} {Execution of a DSLTrans Transformation\\}
-\label{def:DSLTrans_transformation_execution}
-
-The execution of a model transformation is a function $\mathit{applyTransformation}: \mathit{IOM} \times \mathit{Transforms} \rightarrow \mathit{IOM}$.
-
-This function is a chaining of executions of those layers within the transformation.
-
-Recall the application of a layer is \textit{applyLayer} : IOM $\times$ Layer $\rightarrow$ IOM. Therefore, the application of a transformation is:
-
-$\mathit{IOM}$ as input to chaining of layer executions \bentley{Is this function chaining? I don't know how to write the equation for this.}
-
-\end{definition}
-
-While the execution of the rules belonging to a layer happens in parallel, the execution of the layers of a transformation happens sequentially.
-As per \cref{def:DSLTrans_transformation_execution}, the input-output model that is the output of executing a given layer is passed onto the next layer as input.
-The final input-output model created will be the result of the model transformation.
-
-\subsection{Confluence and Termination Properties}
-
-We now prove two important properties about executions of transformations expressed in the subset of DSLTrans presented in this paper: \emph{confluence} and \emph{termination}.
-The proofs are provided at a high level, given the fact that DSLTrans essentially enforces both these properties by construction of the semantics of DSLTrans.
-
-\begin{proposition}{Confluence}
-
-Every model transformation execution is confluent up to typed graph isomorphism.
-
-\bentley{As we are relying on the double-pushout approach, I think we get this confluence for free.}
-
-\end{proposition}
-
-\begin{pf}
-We want to prove that for every model transformation execution of a transformation $\mathit{transform} \in \textsc{Transforms}$ having as input an input-output model $\mathit{input} \in \mathit{IOMs}$, its output is always the same up to typed graph isomorphism.\\
-
-If we assume an execution of the transformation is not confluent then this should happen because of non-determinism when the execution of a transformation is being built.
-Non-determinism happens during the construction of a transformation execution at two points:
-\begin{enumerate}
-\item The section `Execution of a DSLTrans Rule' discusses the matching and rewriting components to rule application.
-Note that our technique relies on the double-pushout approach, which itself is non-deterministic up to typed graph isomorphism, which does not contradict the proposition we are trying to prove.
-\bentley{Revise this.}
-
-\item In Definition~\ref{def:dsltrans_layer_execution}, the rules composing a transformation layer are composed into a production on the input graph.
-Note that the order in which the transformation rules are composed may be non-deterministic.
-However, as these rules can be shown to be parallel independent \bentley{And they have to in that definition}, the production is by definition confluent up to typed graph isomorphism.
-
-\end{enumerate}
-Given there are no other sources of non-determinism when building the execution of a transformation, every model transformation execution is confluent up to typed graph isomorphism.
-\end{pf}
-
-\begin{proposition}{Termination}
-
-Every model transformation execution terminates.
-\end{proposition}
-\begin{pf}
-Let us assume that there is a transformation execution which does not terminate.
-In order for this to happen there must exist a part in the construction of the execution of a transformation which induces an algorithm with an infinite amount of steps.
-We identify three moments when this can happen:
-\begin{enumerate}
-
-\item If the matching and rewriting process of a rule is infinite.
-However, as the double-pushout approach uses a single production, then this must be a finite process.
-\bentley{Verify this, especially in the case of multiple applications of the rule.}
-
-% if the result of the $match_{rl}(m_{in})$ function in definition~\ref{def:match_function} is an infinite set of match-apply graphs.
-The input-output model $m_{in}$ is by definition finite and the matching of each rule is independent from the execution of other rules in the same layer.
-As such, the number of subgraphs of $m_{in}$ isomorphic to $rl$'s matcher found during the execution of $rl$ is finite.
-
-\item if Definition~\ref{def:dsltrans_layer_execution} of execution of a layer induces an infinite amount of steps.
-The only possibility for this to happen is if a layer has an infinite amount of transformation rules, which is a contradiction with Definition~\ref{def:layer}.
-
-\item if Definition~\ref{def:DSLTrans_transformation_execution} of execution of a transformation induces an infinite amount of steps.
-Given layers are executed sequentially and no looping is allowed, the only possibility for this to happen is if the transformation has an infinite amount of layers, which contradicts Definition~\ref{def:transformation}.
-
-\end{enumerate}
-Given there are no other constructs in the semantics of a transformation that can induce an infinite amount of steps, every model transformation execution terminates.
-\end{pf}
+\bentley{Complete this}

### 2016-07-29 text/implementation.tex
Add reviewer's comments into the paper.
Replace DSLTrans section with Frankenstein from different sources.

@@
 
+\reviewer{Table 1, even if it was not experimented to exceed 0.003 seconds when proving for property that does not hold, maybe it would be possible to estimate the maximum time based on that measurement and the complexity formula.}
+
 The results in \cref{tab:scalability_results} were obtained by verifying
@@
 
-\begin{figure*}[htb]
-        \centering
-        \begin{subfigure}[b]{0.26\textwidth}
-                \centering
-                \includegraphics[width=1\textwidth]{./figures/results/rules_vs_pcs.pdf}
-                \caption{Number of rules vs. \\path conds. created}
-                \label{fig:rules_vs_pcs}
-        \end{subfigure}%
-        ~~
-        \begin{subfigure}[b]{0.26\textwidth}
-                \centering
-                \includegraphics[width=1\textwidth]{./figures/results/rules_vs_time.pdf}
-                \caption{Number of rules vs. \\time taken}
-                \label{fig:rules_vs_time}
-        \end{subfigure}%
-        ~~
-        \begin{subfigure}[b]{0.26\textwidth}
-                \centering
-                \includegraphics[width=1\textwidth]{./figures/results/rules_vs_memory.pdf}
-                \caption{Number of rules vs. \\memory used}
-                \label{fig:rules_vs_memory}
-        \end{subfigure}%
-
-        \caption{Metrics for the path condition creation process}
-        \label{fig:path_condition_building}
-\end{figure*}
+%\begin{figure*}[htb]
+%        \centering
+%        \begin{subfigure}[b]{0.26\textwidth}
+%                \centering
+%                \includegraphics[width=1\textwidth]{./figures/results/rules_vs_pcs.pdf}
+%                \caption{Number of rules vs. \\path conds. created}
+%                \label{fig:rules_vs_pcs}
+%        \end{subfigure}%
+%        ~~
+%        \begin{subfigure}[b]{0.26\textwidth}
+%                \centering
+%                \includegraphics[width=1\textwidth]{./figures/results/rules_vs_time.pdf}
+%                \caption{Number of rules vs. \\time taken}
+%                \label{fig:rules_vs_time}
+%        \end{subfigure}%
+%        ~~
+%        \begin{subfigure}[b]{0.26\textwidth}
+%                \centering
+%                \includegraphics[width=1\textwidth]{./figures/results/rules_vs_memory.pdf}
+%                \caption{Number of rules vs. \\memory used}
+%                \label{fig:rules_vs_memory}
+%        \end{subfigure}%
+%
+%        \caption{Metrics for the path condition creation process}
+%        \label{fig:path_condition_building}
+%\end{figure*}
 

### 2016-07-29 text/property_verification.tex
Add more on homomorphisms.
Update with some work-in-progress for abstraction relation.
Replace DSLTrans section with Frankenstein from different sources.

@@
 
-\begin{figure}[htb]
-        \centering
-        \begin{subfigure}[b]{0.40\textwidth}
-                \centering
-                \includegraphics[width=1\textwidth]{./figures/property_proving/prop_execution_without_traceability.pdf}
-                \caption{Property holds}
-                \label{fig:prop_execution_without_traceability}
-        \end{subfigure}%
-        \\
-        \begin{subfigure}[b]{0.40\textwidth}
-                \centering
-                \includegraphics[width=1\textwidth]{./figures/property_proving/prop_execution_with_traceability.pdf}
-                \caption{Property does not hold}
-                \label{fig:prop_execution_with_traceability}
-        \end{subfigure}%
-        \caption{Matching property to a transformation execution}
-        \label{fig:prop_execution}
-\end{figure}
+%\begin{figure}[htb]
+%        \centering
+%        \begin{subfigure}[b]{0.40\textwidth}
+%                \centering
+%                \includegraphics[width=1\textwidth]{./figures/property_proving/prop_execution_without_traceability.pdf}
+%                \caption{Property holds}
+%                \label{fig:prop_execution_without_traceability}
+%        \end{subfigure}%
+%        \\
+%        \begin{subfigure}[b]{0.40\textwidth}
+%                \centering
+%                \includegraphics[width=1\textwidth]{./figures/property_proving/prop_execution_with_traceability.pdf}
+%                \caption{Property does not hold}
+%                \label{fig:prop_execution_with_traceability}
+%        \end{subfigure}%
+%        \caption{Matching property to a transformation execution}
+%        \label{fig:prop_execution}
+%\end{figure}
 
@@
 
-\begin{figure}[htb]
-        \centering
-        \begin{subfigure}[b]{0.144\textwidth}
-                \centering
-                \includegraphics[width=1\textwidth]{./figures/property_proving/unprovable_rule.pdf}
-                \caption{Path condition created from a rule}
-                \label{fig:unprovable_pc}
-        \end{subfigure}%
-        ~~
-        \begin{subfigure}[b]{0.20\textwidth}
-                \centering
-                \includegraphics[width=1\textwidth]{./figures/property_proving/unprovable_prop.pdf}
-                \caption{Property with multiple instances of rule elements}
-                \label{fig:unprovable_prop}
-        \end{subfigure}%
-        \caption{Example of unprovable property}
-        \label{fig:unprovable_property}
-\end{figure}
+%\begin{figure}[htb]
+%        \centering
+%        \begin{subfigure}[b]{0.144\textwidth}
+%                \centering
+%                \includegraphics[width=1\textwidth]{./figures/property_proving/unprovable_rule.pdf}
+%                \caption{Path condition created from a rule}
+%                \label{fig:unprovable_pc}
+%        \end{subfigure}%
+%        ~~
+%        \begin{subfigure}[b]{0.20\textwidth}
+%                \centering
+%                \includegraphics[width=1\textwidth]{./figures/property_proving/unprovable_prop.pdf}
+%                \caption{Property with multiple instances of rule elements}
+%                \label{fig:unprovable_prop}
+%        \end{subfigure}%
+%        \caption{Example of unprovable property}
+%        \label{fig:unprovable_property}
+%\end{figure}
 

### 2016-09-12 text/formal_background.tex
Add reference to Bruno's thesis.

@@
 
-In this section we will introduce the formal concepts that will be used throughout all this paper.
-We start in \cref{sec:typed_graphs} by a few (typed) graph concepts that will be used as mathematical building blocks throughout this paper.
-In particular we will introduce the notion of typed graph, typed graph union and subset, and useful relations between typed graphs based on homomorphisms.
-Notice that these concepts are well known from graph theory and are only slightly customized for our purposes.
+In this section we will introduce formal structures and concepts that will be used throughout this paper.
+We start in Section \ref{subsec:typed_graphs} by a few (typed) graph concepts that will be used as mathematical building blocks.
+In particular we introduce our representation of typed graphs and useful relations between typed graphs based on homomorphisms.
+Notice that these concepts are well known from graph theory and we attempt to make minimal customizations for our purposes.
 
-\reviewer{Your formalisation is made more complex than necessary because you do not put
-the power of typed graphs to work. If your source and target metamodels have
-disjoint types (which you can assume without loss of generality) you can take
-their union, augmented with trace edge types from all target node types to all
-source node types, and take that to be the meta-model for an input-output model,
-and also for a rule. Suddenly you don’t have to carry arround the distinction be-
-tween Match and Apply (for rules) or input and output (for input-output models)
-around any more: they are just the projections onto the types of the source, resp.
-target metamodel.}
+This section is based on that of \cite{Barroca2013},
 
-\reviewer{Section 4.2.2 on “backward links” is rather confusing, as it speaks about trace-
-ability links being added to rules. You never add traceability links to rules any-
-where in the paper; in fact, your formal notion of a rule has no room for them.
-Instead, rules only have backward links. (The clarity of the the situation is not
-improved by the fact that those are labelled trace.)}
-
-\reviewer{I think the situation would be much more clear if you were to use the common
-notion of left hand side and right hand side of a rule. The left hand side of a
-rule rl is precisely your rule matcher, the right hand side is the entire rl augmented by
-your traceability links, as in 4.2.2. Suddenly we are back in a well-known graph
-transformation formalism, and your rules are in fact so simple (no deletion, no
-negative application conditions, injective matching) that I believe it makes no
-difference whether you use algebraic graph transformation or some more con-
-structively defined variant. The only standard notion is that of transitive closure
-in the left hand side; but there are plenty of GT tools that offer this extension.
-In fact I see no reason why your rule application would then not precisely coin-
-cide with the the same notion in, say, SPO graph rewriting; and I hope you agree
-that this would help you no end in explaining what you are doing. (If, on the
-other hand, there is after all some difference then this, too, would be interesting
-to know, as that difference is not at all apparent right now.)}
-
-\reviewer{Viewed like this, moreover, I think your notions of input-output models and rules
-are very close to triple graph transformation as used by Schrr at all. I think the
-main (if not only) difference lies in the fact that your "glue graph" is actually not
-a graph; instead, you use traceability edges directly from the target model to the
-source model. If you were to turn those edges into nodes with source and target
-edges, even that difference would disappear.}
-
-Armed with the fundamental notion of typed graph, we can then introduce other formal concepts in Sections~\ref{sec:DSLTrans_formal}, \ref{sec:DSLTrans_syntax} and \ref{sec:DSLTrans_semantics} which describe the artifacts from the modeling and transformation world that we require for our verification technique.
-Naturally, we start by introducing the central notion of \emph{metamodel}, allowing the description of the inputs and outputs of a model transformation.
-Other fundamental notions we will define in this section are \emph{model}, \emph{transformation rule}, \emph{transformation} and the semantic concept of \emph{model transformation execution}.
-Several auxiliary and intermediate notions for defining the syntax and semantics of our techniques will also be introduced here.
-
-Note that this section presents a collection of formal tools that are used in the subsequent sections of this paper where the contributions of this paper are presented.
-It is meant as a formal reference for the upcoming formal development.
-%This section can be safely skipped or skimmed by the reader, who can return to these definitions punctually to understand the detailed formal underpinning of our approach.
-
-\reviewer{There is a lot of redundancy in the paper, on several levels:
-  - input-output-models, path conditions and properties are
-    all defined very similarly. Couldn't you subsume them in a single
-    definition and spell out the differences?}
+Armed with these fundamental definitions, we can then introduce the formal concepts of the DSLTrans transformation language in Section~\ref{subsec:dsltrans_formal_structures}, and the symbolic representation of transformation executions in Section~\ref{subsubsec:path_condition_creation}.
 
@@
 
-Note that the our verification technique is geared toward model-driven engineering.
+Note that our verification technique is geared toward model-driven engineering.
 Therefore, the types of vertices and edges in our typed graphs will be drawn from a particular \textit{metamodel}.
-Sample metamodels for our running example can be seen in figures above \bentley{Fix.}.
-Note that DSLTrans rules (to be formally defined) will combine typed graphs with potentially different metamodels.
+An example transformation with source and target metamodels is presented in Section~\ref{subsec:fam_to_persons_example}.
+Note that DSLTrans rules (to be formally defined) will consist of multiple typed graphs with potentially different metamodels.
 
-We assume that this metamodel provides the necessary vertex and edge types, with any collisions resolved by adding to the metamodel name.
-As well, we assume the presence of a partial ordering $\leq$ on the vertex and edge types for subtyping information.
+We assume that this metamodel provides the necessary vertex and edge types, and that all duplicate types have been resolved. 
+%As well, we assume that any name collisions arising from duplicate names in the source and target metamodel is resolved by appending the metamodel name to the vertex or edge type name.
+As well, we assume that the metamodel provides a partial ordering $\leq$ on the vertex and edge types for subtyping information.
 
@@
 Instead our matching must be flexible about how vertices in the pattern graph may match over multiple vertices in the target graph.
+The motivation for requiring this `one-to-many' matching is provided in Section~\ref{sec:abstraction_relation}.
 
@@
 
-The first typed graph homomorphism we define is standard, where the structure of the pattern graph must be found in the target graph.
-This is presented for the reader to appreciate the complications required for the next homomorphisms.
+The first typed graph homomorphism we define is standard in the literature, where the structure of the pattern graph is found one-to-one in the target graph.
+This homomorphism is presented for the reader to appreciate the complications required for the non-standard homomorphism.
 
@@
 This homomorphism will be an injective match regarding the edges, but note that a particular vertex in the pattern graph may match onto multiple vertices in the target graph.
-\bentley{What's the term for this?}
+Thus, this homomorphism is `one-to-many'.
 
@@
 \label{def:instance_closure}
+Let $g$ be a graph $\langle V,E,(s, t),\tau\rangle \in \textsc{TG}$.
+Then the transitive closure $g^{*} = \langle V,E',(s', t'),\tau', \VT, \ET' \rangle \in \textsc{TG}$ is created by $E' = E \cup E_{tc}$.
 
-Let $g$ be a graph $\langle V,E,(s, t),\tau\rangle \in \textsc{TG}$. Then the transitive closure $g^{*} = \langle V,E',(s', t'),\tau', \VT, \ET' \rangle \in \textsc{TG}$, where:
+The new edges $e_{tc} \in E_{tc}$ are created such that:
 
 \begin{itemize}
-\item $E' = E \cup v_1\xrightarrow{e''}v_2 | \exists v_1\xrightarrow{e_i}v_i, v_i\xrightarrow{e_j}v_j, \dots, v_k\xrightarrow{e_k}v_2 \land \mathit{isIndirect(e_i)}$
-\item \bentley{How to create s, t, and tau?}
+\item $v_1\xrightarrow{e_{tc}}v_2 | \exists v_1\xrightarrow{e_i}v_i, v_i\xrightarrow{e_j}v_j, \dots, v_k\xrightarrow{e_k}v_2$
+\item $s'(e_{tc}) = v_1$, $t'(e_{tc}) = v_2$
+\item $\tau(e_{tc}) = $
 \end{itemize}
+
 \end{definition}

### 2016-09-13 text/formal_background.tex
Do more on transitivity edges.

@@
 \label{def:instance_closure}
-Let $g$ be a graph $\langle V,E,(s, t),\tau\rangle \in \textsc{TG}$.
-Then the transitive closure $g^{*} = \langle V,E',(s', t'),\tau', \VT, \ET' \rangle \in \textsc{TG}$ is created by $E' = E \cup E_{tc}$.
-
-The new edges $e_{tc} \in E_{tc}$ are created such that:
+Let $g$ be a graph $\langle V,E,(s, t),\tau\rangle \in \textsc{TG}$. Then the transitive closure $g^{*} = \langle V,E',(s', t'),\tau', \VT, \ET' \rangle \in \textsc{TG}$ is created such that:
 
 \begin{itemize}
+\item $E' = E \cup E_{tc}$ where each new edge $e_{tc} \in E_{tc}$ is created between vertices $v_1$ and $v_2$ in $V$:
+\begin{itemize}
 \item $v_1\xrightarrow{e_{tc}}v_2 | \exists v_1\xrightarrow{e_i}v_i, v_i\xrightarrow{e_j}v_j, \dots, v_k\xrightarrow{e_k}v_2$
+\end{itemize}
+
 \item $s'(e_{tc}) = v_1$, $t'(e_{tc}) = v_2$
-\item $\tau(e_{tc}) = $
+\item $\tau(e_{tc}) = \mathit{indirect}$
 \end{itemize}
@@
 Given a graph, its transitive closure includes, besides the original graph, all the edges belonging to the transitive closure of indirect links in that graph.
-Note that these transitive edges are identified by the function $\mathit{isIndirect}$.
+Note that these transitive edges are given a special type \textit{indirect}, and will be identified by the function $\mathit{isIndirect}$.
 
@@
 
-\begin{itemize}
-\item Backward links injectively match from contract into PC onto traceability links.
-Source and target of backward links on contract must match the source and target of the traceability link of the PC.
-\end{itemize}
-
-We must also define a function to match link structures, such as backward links and the traceability links (to be discussed).
+For convenience, we also define a function to match components in our structures which contains only links.
+An example would be the backward link construct for DSLTrans rules, as discussed in Section~\ref{subsec:dsltrans_formal_structures}.
 
@@
 \label{def:link_homomorphism}
-Consider two sets of links $L = \{E_{L}, (s_{L}, t_{L})\}$, and $T = \{E_{T}, (s_{T}, t_{T})\}$. Note that these could be backward links or traceability links.
+Consider two sets of links $A = \{E_{A}, (s_{A}, t_{A})\}$, and $B = \{E_{B}, (s_{B}, t_{B})\}$. Note that these could be backward links or traceability links.
 
-We define an injective homomorphism function $f$ to match $L$ over $T$, such that:
+We define a (trivial) injective homomorphism function $f$ to match $A$ over $B$, such that:
 
 \begin{itemize}
-\item $\forall e_L \in E_L:$
+\item $\forall e_A \in E_A:$
 \begin{itemize}
-\item $\forall e_T \in E_T | f(e_L) = e(T)$
+\item $\exists e_b \in E_{B} | f(e_A) = e_b$
+\item $f(s_{A}(e_A)) = s_{B}(e_B)$
+\item $f(t_{A}(e_A)) = t_{B}(e_B)$
 \end{itemize}
@@
 
-For example, we may need to find a homomorphism between two structures $S_1 = \langle A, B, L\rangle$ and  $S_2 = \langle A', B', L'\rangle$, where $A, A', B, B' \in TG$ and $L, L'$ are links of the form $\{E, (s, t)\}$.
+For example, we may need to find a homomorphism $m$ between two structures $S_1 = \langle A, B, L\rangle$ and  $S_2 = \langle A', B', L'\rangle$, where $A, A', B, B' \in TG$ and $L, L'$ are links of the form $\{E, (s, t)\}$.
 
-This overall homomorphism will compose the homomorphism $f(A) = A'$, $g(B) = B'$, $h(L) = L'$, where f and g are the homomorphisms to be found, and h is the link homomorphism.
+This overall homomorphism be composed of the sub-homomorphisms $f(A) = A'$, $g(B) = B'$, $h(L) = L'$, where $f$ and $g$ are the homomorphisms to be found, and $h$ is the link homomorphism.
+Note that $f$ and $g$ may have differing properties, such as being injective or not, motivating our consideration of three sub-morphisms.
 
-A reviewer pointed out that we also require that these sub-homomorphisms are consistent with each other. That is, the link homomorphism h must agree with 
-the bindings given by f and g.
+Our definitions for these structures require that the components (such as $A$, $B$, and $L$) are all disjoint. Thus we consider the homomorphism to be trivially composed: $m = f \cup g \cup h$.
 
-\bentley{Complete this}
+However, a reviewer pointed out that we also require that these sub-homomorphisms are consistent with each other.
+Namely that the link homomorphism $h$ must agree with the bindings given by $f$ and $g$.
+That is:
+
+\bentley{Finish}
